### 2019-07-16 analysis/analysis.Rmd
Initial commit with analysis presented at IGWP2019

---
title: Ecological niche models of plant species exploited by Epipalaeolithic foragers
  in the *Badia* (c. 14500 BP)
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

## Dependencies

```{r packages, warning=FALSE, message=FALSE}
library("tidyverse")
library("magrittr")
library("sf")
library("raster")
library("broom")
library("dismo")
library("rnaturalearth")
library("gt")
library("BadiaPaleoFloraENM")
```

## Introduction

Species distribution models (SDM) or ecological niche models (ENM) use data on the observed occurrences of a taxon to construct a statistical model of its ecological niche.
These models are generally then used to predict the distribution of the species in another context, e.g. a region where empirical occurrence data is not available, or under past or future climate regimes.
We will attempt to model the ecological niches of six key plant taxa for prehistoric foragers.
We will use occurrence data from Israel and Palestine, which are well-studied, to predict the distribution of these taxa across the rest of the Southern Levant.

First, we define our regions of interest:

```{r regions, warning=FALSE, message=FALSE}
sw_asia <- extent(c(30, 50, 25, 40))
s_levant <- extent(c(34, 39, 29, 34))
palestine <- ne_countries(scale = 10, country = c("Israel", "Palestine"),
             returnclass = "sf") %>% st_union() %>% st_sf()

land <- read_sf("data/raw_data/ne/ne_10m_land.shp")[,"featurecla"] %>%
  st_crop(sw_asia)
rivers <- read_sf("data/raw_data/ne/ne_10m_rivers_lake_centerlines_scale_rank.shp") %>% 
  st_crop(s_levant)
water <- rbind(read_sf("data/raw_data/ne/ne_10m_ocean.shp")[,"featurecla"],
               read_sf("data/raw_data/ne/ne_10m_lakes.shp")[,"featurecla"]) %>% 
  st_crop(sw_asia)
```

```{r regions-map, echo=FALSE, results='hide'}
basemap <- ggplot() + 
  geom_sf(data = land, fill ="grey") +
  geom_sf(data = palestine, fill = "lightgrey") + 
  geom_sf(data = rivers, colour = "white") +
  geom_sf(data = water, fill = "white") +
  coord_sf(xlim = c(xmin(s_levant), xmax(s_levant)),
           ylim = c(ymin(s_levant), ymax(s_levant)))
basemap
```

## Occurrence data

First we download species occurrence data from GBIF (<https://gbif.org>).
This may change, so we cache the result used in the published analysis in `derived_data` and only redownload if the cached version is missing.

```{r occurrence-data, warning=FALSE, message=FALSE}
# Download or read data from GBIF
gbif_path <- here::here("analysis/data/raw_data/gbif")
occ <- bind_rows(
  fetch_gbif("Hordeum", "spontaneum*", ext = sw_asia, gbif_path),
  fetch_gbif("Bolboschoenus", "glaucus*", ext = sw_asia, gbif_path),    
  fetch_gbif("Bolboschoenus", "maritimus*", ext = sw_asia, gbif_path),    
  fetch_gbif("Scirpus", "maritimus*", ext = sw_asia, gbif_path),
  fetch_gbif("Triticum", "urartu*", ext = sw_asia, gbif_path),
  fetch_gbif("Triticum", "dicoccoides*", ext = sw_asia, gbif_path),
  fetch_gbif("Stipa", "*", ext = sw_asia, gbif_path),
  fetch_gbif("Zilla", "spinosa*", ext = sw_asia, gbif_path),
  fetch_gbif("Vitex", "*", ext = sw_asia, gbif_path),
  fetch_gbif("Fraxinus", "*", ext = sw_asia, gbif_path)
)

# Code taxonomic groups for analysis
occ %>% 
  mutate(taxon = case_when(genus == "Hordeum" ~ "Hordeum spontaneum",
                           genus == "Bolboschoenus" ~ "Bolboschoenus sp. (=Scirpus maritimus)",
                           species == "Triticum urartu" ~ "Triticum urartu",
                           species == "Triticum dicoccoides" ~ "Triticum dicoccoides",
                           genus == "Stipa" ~ "Stipa sp.",
                           genus == "Zilla" ~ "Zilla spinosa",
                           genus == "Vitex" ~ "Vitex sp.",
                           genus == "Fraxinus" ~ "Fraxinus sp.")) %T>%
  {if(any(is.na(.$taxon))) stop("Taxon contains missing values!")} ->
  occ

# Remove duplicate records within analytic groups
occ %>% 
  group_by(taxon) %>% 
  distinct(lon, lat, .keep_all = TRUE) %>% 
  ungroup() ->
  occ

# Convert to sf
occ %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) ->
  occ
```

It's interesting to map the current distribution of these plants.
```{r occurrence-map, echo=FALSE, results='hide'}
ggplot(occ, aes(x = lon, y = lat)) +
  geom_sf(data = occ, shape = 1, size=0.5) +
  geom_sf(data = water, inherit.aes = FALSE, fill = "white", colour = "darkgrey") +
  facet_wrap(~taxon)
```

## Pseudo-absences

Occurrence data only tells us where a species is present;
there is no definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of Barbet-Massin et al. (2012) for regression-based species distribution models and simply use a large (~10000) random sample of points.

```{r psabsences, warning=FALSE, message=FALSE}
occ <- as_tibble(occ)
occ$occurrence <- TRUE
for(taxon in unique(occ$taxon)) {
  occ <- bind_rows(occ, generate_psabsences(palestine, taxon = taxon,
                                            occurrence = FALSE, N = 10000))
}
occ <- sf::st_as_sf(occ, crs = 4326)
rm(taxon)
```

## Environmental data

### Bioclimatic variables

To construct the model, we'll use a standard set of "bioclimatic" predicator variables.
These sixteen variables are derived from monthly precipitation and temperate data:

```{r bioclim-table, echo=FALSE}
tibble(biovar = paste0("bio", 1:16),
       definition = c("Mean annual temperature",
                      "Mean diurnal range",
                      "Isothermality",
                      "Temperature seasonality",
                      "Maximum temperature of warmest month",
                      "Minimum temperature of coldest month",
                      "Annual temperature range",
                      "Mean temperature of wettest quarter",
                      "Mean temperature of driest quarter",
                      "Mean temperature of warmest quarter",
                      "Mean temperature of coldest quarter",
                      "Total annual precipitation",
                      "Precipitation of wettest month",
                      "Precipitation of driest month",
                      "Precipitation seasonality",
                      "Precipitation of wettest quarter"),
       formula = c("",
                   "",
                   "(bio2 / bio7) * 100",
                   "sd(temp) * 100",
                   "",
                   "",
                   "bio5 - bio6",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "")
) %>%
  gt()
```

Here the climate data is from WorldClim 1.4 (<http://www.worldclim.org>), a global dataset compiled from weather station records from the period 1960–1990.
WorldClim has been widely used in ecological niche modelling.
Other global climate datasets (e.g. CHELSA, MERRAClim) and a newer version of WorldClim are now available, so it may be a good idea to investigate using these instead.
The WorldClim data is large so again it is cached in the package's `derived_data` directory.

```{r bioclim-data, echo=FALSE}
# bioclim <- get_bioclim(s_levant, resolution = 0.5, cache_path = "data/derived_data")
# Workaround for memory problems in Safawi
load("data/derived_data/bioclim.RData")
plot(bioclim)
```

### Soil variables

Velazco et al. (2017) have shown that edaphic data, recently available on a global scale, is usefully combined with bioclimatic data in modelling plant niches.

We use data from SoilGrids, a global map of soil characteristics derived from remote sensing classification at 1km resolution.

```{r soil-data}
soilgrids <- raster::stack(list.files("data/raw_data/soilgrids", ".*tif", full.names = TRUE))
plot(soilgrids)
```

### Terrain

There are also relevent aspects of the terrain that can be easily derived from a digital elevation model.
We'll use NASA-JPL's SRTM v3 1 arc-second resolution DEM.

### Selection of predictor variables

We reviewed existing floras to identify specific variables from these datasets that are relevant to the niche of our target taxa. The results are summarised below.

```{r review-table, echo=FALSE, warning=FALSE, message=FALSE}
read_csv("data/raw_data/flora_review.csv") %>% 
  group_by(Taxon) %>% 
  gt()
```

Based on these results, we made the following initial selection of predictor variables to build the models.

```{r predictor-table, echo=FALSE}
pred <- bind_rows(
  tibble(group = rep("Climate", 5),
         variable = c("Minimum temperature of coldest month",
                      "Total annual precipitation", "Precipitation of wettest month",
                      "Precipitation of driest month", "Precipitation seasonality"),
         units = c("°C ×10", "mm", "mm", "mm", "σ/μ"),
         source = rep("WorldClim 1.4", 5),
         code = c("bio6", "bio12", "bio13", "bio14", "bio15"),
         raster = as.list(raster::subset(bioclim,
                                         c("bio6", "bio12", "bio13", "bio14", "bio15")))),
  tibble(group = rep("Soil", 5),
         variable = c("Organic carbon content", "Coarse fragment content",
                      "Sand content", "Acidity", "Depth to bedrock"),
         units = c("g per kg fine earth", "% volume", "% mass", "pH ×10 in H2O",
                   "cm (up to 200 cm)"),
         source = rep("SoilGrids250m", 5),
         code = c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"),
         raster = as.list(raster::subset(soilgrids,
                                         c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"))))
  # tibble(group = rep("Terrain", 3),
  #        variable = c("Elevation", "Slope", "Topographic Wetness Index"),
  #        units = c("m", "°", "ln(upslope area / tan(slope))"),
  #        source = rep("SRTM DEM v3", 3),
  #        code = c("srtm1v3_slevant", "srtm1v3_slevant_slope",
  #                         "srtm1v3_slevant_twi"))
)

pred %>%
  dplyr::select(-raster) %>% 
  group_by(group) %>% 
  gt()
```

One aspect that we cannot easily incorporate into the models is the interaction between these taxa and other plant and animal species: cereal grasses grow best when they have no competition from perennial plants, *Fraxinus* often grows in scrub or forest, and so on.

We also lack data for soil salinity so far.

### Combining predictor variables

These datasets come with different resolutions.
To use them all in the same analysis, we resample them all to the finest dataset (WorldClim, 30 arc-seconds).

```{r predictor-resample}
pred$raster <- lapply(pred$raster, raster::resample, y = bioclim, method = "ngb")
```

## Model construction

We will model the ecological niche of each species using a generalised linear regression.

First we extract the values of the predictor variables at each occurrence and pseudo-absence location:

```{r predictor-extraction}
# TODO: refactor to use pred
bioclim %>% 
  raster::extract(as_Spatial(occ)) %>% 
  {cbind(occ, .)} %>% 
  st_as_sf(crs = st_crs(occ)) ->
  occ

soilgrids %>% 
  raster::extract(as_Spatial(occ)) %>% 
  {cbind(occ, .)} %>% 
  st_as_sf(crs = st_crs(occ)) ->
  occ
```

### Species response curves

We can use species response curves to visualise the distribution of known occurrences in relation to these environmental gradients.
This will also help us parameterise the ecological niche models.

Species response curves are constructed with a simple logistic regression, fitting the probability of occurrence at a given location to 
The use of a linear regression assumes a unimodal response.
We fit first, second, and third order polynomial models to each species/predictor combination and calculate the Akaike Information Criterion (AIC).

```{r response-curves-fit, warning=FALSE, message=FALSE}
# Summarise occurrences on a grid
palestine %>% 
  as_Spatial() %>% 
  raster::rasterize(bioclim) ->
  palestine_raster

occ %>% 
  filter(occurrence == TRUE) %>% 
  group_by(taxon) %>% 
  do(occ_grid = raster_burn(., palestine_raster)) ->
  occ_grids

# Fit poly(1:3) to each taxon/predictor combination
expand.grid(taxon = unique(occ$taxon),
            predictor = pred$code,
            polyn = 1:3) %>% 
  group_by(taxon, predictor, polyn) %>% 
  do(
    # TODO: It works but I hate it.
    fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
                         unlist(pred[pred$code == .$predictor, "raster"])[[1]],
                         .$polyn)
  ) %>% 
  glance(fit) -> 
  response_glm
```

```{r response-curves-table, echo=FALSE}
# response_glm %>% 
#   spread(polyn, aic) %>% 
#   rename(aic_poly1 = `1`, aic_poly2 = `2`, aic_poly3 = `3`) %>% 
#   gt(rowname_col = "predictor", groupname_col = "taxon") %>% 
#   tab_stubhead_label("predictor")
```

The optimal model is the one with the lowest AIC.

```{r response-curves-optimal}
response_glm %>%
  group_by(taxon, predictor) %>%
  mutate(optimum = polyn == which.min(AIC)) ->
  response_glm

# TODO
# Re-fit and plot optimal 
# model_response(occ_grids$occ_grid[[1]], pred$raster[[1]], 2) %>% predict(data.frame(pred = -72:109, type = "response")) %>% plot()
# Need to dplyrise this – see broom::augment()
# response_glm %>% 
#   filter(optimum == TRUE) %>% 
#   slice(1:3) %>% 
#   do(
#     fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
#                          unlist(pred[pred$code == .$predictor,"raster"])[[1]],
#                          .$polyn)
#   ) 
```

### Model fitting

Now we can fit regression models combining all the predictor variables.

```{r model_fit, warning=FALSE, message=FALSE}
# TODO: use optimal polynomial order
occ %>% 
  group_by(taxon) %>% 
  remove_missing(vars = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", 
                          "bio7", "bio8", "bio9", "bio10", "bio11", "bio12", 
                          "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", 
                          "bio19", "BDRICM_M_250m", "CRFVOL_M_sl2_250m", 
                          "ORCDRC_M_sl2_250m", "PHIHOX_M_sl2_250m", 
                          "SNDPPT_M_sl2_250m")) %>% 
  do(model = glm(occurrence ~ 
                   poly(bio6, 2) +
                   poly(bio12, 2) +
                   poly(bio13, 2) +
                   # poly(bio14, 2) +
                   poly(bio15, 2) +
                   poly(BDRICM_M_250m, 2) +
                   poly(CRFVOL_M_sl2_250m, 2) +
                   poly(ORCDRC_M_sl2_250m, 2) + 
                   poly(PHIHOX_M_sl2_250m, 2) + 
                   poly(SNDPPT_M_sl2_250m, 2),
                 data = .)) ->
  models

```

## Predicting occurrences
### Current conditions

We can now use the fitted models to predict the occurrence of the different taxa according to the predictor variables.
Initially, we do this based on current conditions.
This allows us to extrapolate from regions with good occurrence data (i.e. Israel/Palestine) to those without.
It also hints at what the distribution of the plant might be under "ideal conditions", without agriculture, grazing, etc.

```{r pred_current}
models %>% 
  do(taxon = .$taxon,
     model = .$model,
     pred_current = raster::predict(object = raster::stack(pred$raster),
                                    model = .$model)) ->
  models
```

```{r pred_current_plot, echo=FALSE}
raster::stack(models$pred_current) %>%
  set_names(models$taxon) %>% 
  plot()
```

### Binning

To more easily interpret these predictions, we could convert the raw probability scores into confidence intervals, based on the proportion of known occurrences it successfully predicts.

```{r binning}
raster::stack(models$pred_current) %>%
  set_names(paste0("pred_current_", models$taxon)) %>% 
  raster::extract(as_Spatial(occ)) %>% 
  {cbind(occ, .)} %>% 
  st_as_sf(crs = st_crs(occ)) ->
  occ2
```

### 2019-07-16 analysis/analysis.tex
Initial commit with analysis presented at IGWP2019

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Ecological niche models of plant species exploited by Epipalaeolithic foragers in Badia (c.~14500 BP)},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Ecological niche models of plant species exploited by Epipalaeolithic
foragers in \emph{Badia} (c.~14500 BP)}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{}
    \preauthor{}\postauthor{}
    \date{}
    \predate{}\postdate{}

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}

\begin{document}
\maketitle

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"tidyverse"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Registered S3 methods overwritten by 'ggplot2':
##   method         from 
##   [.quosures     rlang
##   c.quosures     rlang
##   print.quosures rlang
\end{verbatim}

\begin{verbatim}
## -- Attaching packages ------------------------------------------------------------------------------------------ tidyverse 1.2.1 --
\end{verbatim}

\begin{verbatim}
## v ggplot2 3.1.1     v purrr   0.3.2
## v tibble  2.1.2     v dplyr   0.8.1
## v tidyr   0.8.3     v stringr 1.4.0
## v readr   1.3.1     v forcats 0.4.0
\end{verbatim}

\begin{verbatim}
## -- Conflicts --------------------------------------------------------------------------------------------- tidyverse_conflicts() --
## x dplyr::filter() masks stats::filter()
## x dplyr::lag()    masks stats::lag()
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"magrittr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 
## Attaching package: 'magrittr'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:purrr':
## 
##     set_names
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:tidyr':
## 
##     extract
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"sf"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Linking to GEOS 3.7.1, GDAL 2.3.2, PROJ 5.2.0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"raster"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: sp
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'raster'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:magrittr':
## 
##     extract
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:dplyr':
## 
##     select
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:tidyr':
## 
##     extract
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"broom"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"dismo"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"rnaturalearth"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"gt"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"BadiaPaleoFloraSDM"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Species distribution models (SDM) or ecological niche models (ENM) use
data on the observed occurrences of a taxon to construct a statistical
model of its ecological niche. These models are generally then used to
predict the distribution of the species in another context, e.g.~a
region where empirical occurrence data is not available, or under past
or future climate regimes. We will attempt to model the ecological
niches of six key plant taxa for prehistoric foragers. We will use
occurrence data from Israel and Palestine, which are well-studied, to
predict the distribution of these taxa across the rest of the Southern
Levant.

First, we define our regions of interest:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sw_asia <-}\StringTok{ }\KeywordTok{extent}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{30}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DecValTok{25}\NormalTok{, }\DecValTok{40}\NormalTok{))}
\NormalTok{s_levant <-}\StringTok{ }\KeywordTok{extent}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{34}\NormalTok{, }\DecValTok{39}\NormalTok{, }\DecValTok{29}\NormalTok{, }\DecValTok{34}\NormalTok{))}
\NormalTok{palestine <-}\StringTok{ }\KeywordTok{ne_countries}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{10}\NormalTok{, }\DataTypeTok{country =} \KeywordTok{c}\NormalTok{(}\StringTok{"Israel"}\NormalTok{, }\StringTok{"Palestine"}\NormalTok{),}
             \DataTypeTok{returnclass =} \StringTok{"sf"}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{st_union}\NormalTok{() }\OperatorTok{%>%}\StringTok{ }\KeywordTok{st_sf}\NormalTok{()}

\NormalTok{land <-}\StringTok{ }\KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_land.shp"}\NormalTok{)[,}\StringTok{"featurecla"}\NormalTok{] }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{st_crop}\NormalTok{(sw_asia)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersection assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant
## throughout all geometries
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{rivers <-}\StringTok{ }\KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_rivers_lake_centerlines_scale_rank.shp"}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{st_crop}\NormalTok{(s_levant)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersection assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant
## throughout all geometries
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{water <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_ocean.shp"}\NormalTok{)[,}\StringTok{"featurecla"}\NormalTok{],}
               \KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_lakes.shp"}\NormalTok{)[,}\StringTok{"featurecla"}\NormalTok{]) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{st_crop}\NormalTok{(sw_asia)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersection assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant
## throughout all geometries
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{basemap <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ land, }\DataTypeTok{fill =}\StringTok{"grey"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ palestine, }\DataTypeTok{fill =} \StringTok{"lightgrey"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ rivers, }\DataTypeTok{colour =} \StringTok{"white"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ water, }\DataTypeTok{fill =} \StringTok{"white"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{coord_sf}\NormalTok{(}\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{xmin}\NormalTok{(s_levant), }\KeywordTok{xmax}\NormalTok{(s_levant)),}
           \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{ymin}\NormalTok{(s_levant), }\KeywordTok{ymax}\NormalTok{(s_levant)))}
\NormalTok{basemap}
\end{Highlighting}
\end{Shaded}

\includegraphics{analysis_files/figure-latex/regions-1.pdf}

\hypertarget{occurrence-data}{%
\subsection{Occurrence data}\label{occurrence-data}}

First we download species occurrence data from GBIF
(\url{https://gbif.org}). This may change, so we cache the result used
in the published analysis in \texttt{derived\_data} and only redownload
if the cached version is missing.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Download or read data from GBIF}
\NormalTok{gbif_path <-}\StringTok{ }\NormalTok{here}\OperatorTok{::}\KeywordTok{here}\NormalTok{(}\StringTok{"analysis/data/raw_data/gbif"}\NormalTok{)}
\NormalTok{occ <-}\StringTok{ }\KeywordTok{bind_rows}\NormalTok{(}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Hordeum"}\NormalTok{, }\StringTok{"spontaneum*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Bolboschoenus"}\NormalTok{, }\StringTok{"glaucus*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),    }
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Bolboschoenus"}\NormalTok{, }\StringTok{"maritimus*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),    }
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Scirpus"}\NormalTok{, }\StringTok{"maritimus*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Triticum"}\NormalTok{, }\StringTok{"urartu*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Triticum"}\NormalTok{, }\StringTok{"dicoccoides*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Stipa"}\NormalTok{, }\StringTok{"*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Zilla"}\NormalTok{, }\StringTok{"spinosa*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Vitex"}\NormalTok{, }\StringTok{"*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Fraxinus"}\NormalTok{, }\StringTok{"*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path)}
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_hordeum_spontaneum.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_bolboschoenus_glaucus.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_bolboschoenus_maritimus.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_scirpus_maritimus.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_triticum_urartu.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_triticum_dicoccoides.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_stipa.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_zilla_spinosa.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_vitex.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_fraxinus.csv
\end{verbatim}

\begin{verbatim}
## Parsed with column specification:
## cols(
##   gbifID = col_double(),
##   genus = col_character(),
##   species = col_character(),
##   lon = col_double(),
##   lat = col_double(),
##   basisOfRecord = col_character()
## )
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Code taxonomic groups for analysis}
\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{taxon =} \KeywordTok{case_when}\NormalTok{(genus }\OperatorTok{==}\StringTok{ "Hordeum"} \OperatorTok{~}\StringTok{ "Hordeum spontaneum"}\NormalTok{,}
\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Bolboschoenus"} \OperatorTok{~}\StringTok{ "Bolboschoenus sp. (=Scirpus maritimus)"}\NormalTok{,}
\NormalTok{                           species }\OperatorTok{==}\StringTok{ "Triticum urartu"} \OperatorTok{~}\StringTok{ "Triticum urartu"}\NormalTok{,}
\NormalTok{                           species }\OperatorTok{==}\StringTok{ "Triticum dicoccoides"} \OperatorTok{~}\StringTok{ "Triticum dicoccoides"}\NormalTok{,}
\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Stipa"} \OperatorTok{~}\StringTok{ "Stipa sp."}\NormalTok{,}
\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Zilla"} \OperatorTok{~}\StringTok{ "Zilla spinosa"}\NormalTok{,}
\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Vitex"} \OperatorTok{~}\StringTok{ "Vitex sp."}\NormalTok{,}
\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Fraxinus"} \OperatorTok{~}\StringTok{ "Fraxinus sp."}\NormalTok{)) }\OperatorTok{%T>%}
\StringTok{  }\NormalTok{\{}\ControlFlowTok{if}\NormalTok{(}\KeywordTok{any}\NormalTok{(}\KeywordTok{is.na}\NormalTok{(.}\OperatorTok{$}\NormalTok{taxon))) }\KeywordTok{stop}\NormalTok{(}\StringTok{"Taxon contains missing values!"}\NormalTok{)\} ->}
\StringTok{  }\NormalTok{occ}

\CommentTok{# Remove duplicate records within analytic groups}
\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{distinct}\NormalTok{(lon, lat, }\DataTypeTok{.keep_all =} \OtherTok{TRUE}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{ungroup}\NormalTok{() ->}
\StringTok{  }\NormalTok{occ}

\CommentTok{# Convert to sf}
\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{coords =} \KeywordTok{c}\NormalTok{(}\StringTok{"lon"}\NormalTok{, }\StringTok{"lat"}\NormalTok{), }\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{, }\DataTypeTok{remove =} \OtherTok{FALSE}\NormalTok{) ->}
\StringTok{  }\NormalTok{occ}
\end{Highlighting}
\end{Shaded}

It's interesting to map the current distribution of these plants.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{(occ, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ lon, }\DataTypeTok{y =}\NormalTok{ lat)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ occ, }\DataTypeTok{shape =} \DecValTok{1}\NormalTok{, }\DataTypeTok{size=}\FloatTok{0.5}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ water, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"white"}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"darkgrey"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{facet_wrap}\NormalTok{(}\OperatorTok{~}\NormalTok{taxon)}
\end{Highlighting}
\end{Shaded}

\includegraphics{analysis_files/figure-latex/occurrence-map-1.pdf}

\hypertarget{pseudo-absences}{%
\subsection{Pseudo-absences}\label{pseudo-absences}}

Occurrence data only tells us where a species is present; there is no
definitive information on where the species is \emph{not} found. We
therefore need to generate random background points or
``pseudo-absences'' to feed to the model. There are several ways to do
this. We follow the advice of Barbet-Massin et al.~(2012) for
regression-based species distribution models and simply use a large
(\textasciitilde10000) random sample of points.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{occ <-}\StringTok{ }\KeywordTok{as_tibble}\NormalTok{(occ)}
\NormalTok{occ}\OperatorTok{$}\NormalTok{occurrence <-}\StringTok{ }\OtherTok{TRUE}
\ControlFlowTok{for}\NormalTok{(taxon }\ControlFlowTok{in} \KeywordTok{unique}\NormalTok{(occ}\OperatorTok{$}\NormalTok{taxon)) \{}
\NormalTok{  occ <-}\StringTok{ }\KeywordTok{bind_rows}\NormalTok{(occ, }\KeywordTok{generate_psabsences}\NormalTok{(palestine, }\DataTypeTok{taxon =}\NormalTok{ taxon,}
                                            \DataTypeTok{occurrence =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{N =} \DecValTok{10000}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes

## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{verbatim}
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
## although coordinates are longitude/latitude, st_intersects assumes that they are planar
\end{verbatim}

\begin{verbatim}
## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
## preserve their attributes
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{occ <-}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st_as_sf}\NormalTok{(occ, }\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{)}
\KeywordTok{rm}\NormalTok{(taxon)}
\end{Highlighting}
\end{Shaded}

\hypertarget{environmental-data}{%
\subsection{Environmental data}\label{environmental-data}}

\hypertarget{bioclimatic-variables}{%
\subsubsection{Bioclimatic variables}\label{bioclimatic-variables}}

To construct the model, we'll use a standard set of ``bioclimatic''
predicator variables. These sixteen variables are derived from monthly
precipitation and temperate data:

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{lll}
\toprule
biovar & definition & formula \\ 
\midrule
bio1 & Mean annual temperature &  \\ 
bio2 & Mean diurnal range &  \\ 
bio3 & Isothermality & (bio2 / bio7) * 100 \\ 
bio4 & Temperature seasonality & sd(temp) * 100 \\ 
bio5 & Maximum temperature of warmest month &  \\ 
bio6 & Minimum temperature of coldest month &  \\ 
bio7 & Annual temperature range & bio5 - bio6 \\ 
bio8 & Mean temperature of wettest quarter &  \\ 
bio9 & Mean temperature of driest quarter &  \\ 
bio10 & Mean temperature of warmest quarter &  \\ 
bio11 & Mean temperature of coldest quarter &  \\ 
bio12 & Total annual precipitation &  \\ 
bio13 & Precipitation of wettest month &  \\ 
bio14 & Precipitation of driest month &  \\ 
bio15 & Precipitation seasonality &  \\ 
bio16 & Precipitation of wettest quarter &  \\ 
\bottomrule
\end{longtable}

Here the climate data is from WorldClim 1.4
(\url{http://www.worldclim.org}), a global dataset compiled from weather
station records from the period 1960--1990. WorldClim has been widely
used in ecological niche modelling. Other global climate datasets
(e.g.~CHELSA, MERRAClim) and a newer version of WorldClim are now
available, so it may be a good idea to investigate using these instead.
The WorldClim data is large so again it is cached in the package's
\texttt{derived\_data} directory.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# bioclim <- get_bioclim(s_levant, resolution = 0.5, cache_path = "data/derived_data")}
\CommentTok{# Workaround for memory problems in Safawi}
\KeywordTok{load}\NormalTok{(}\StringTok{"data/derived_data/bioclim.RData"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(bioclim)}
\end{Highlighting}
\end{Shaded}

\includegraphics{analysis_files/figure-latex/bioclim-data-1.pdf}

\hypertarget{soil-variables}{%
\subsubsection{Soil variables}\label{soil-variables}}

Velazco et al.~(2017) have shown that edaphic data, recently available
on a global scale, is usefully combined with bioclimatic data in
modelling plant niches.

We use data from SoilGrids, a global map of soil characteristics derived
from remote sensing classification at 1km resolution.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{soilgrids <-}\StringTok{ }\NormalTok{raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(}\KeywordTok{list.files}\NormalTok{(}\StringTok{"data/raw_data/soilgrids"}\NormalTok{, }\StringTok{".*tif"}\NormalTok{, }\DataTypeTok{full.names =} \OtherTok{TRUE}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\hypertarget{terrain}{%
\subsubsection{Terrain}\label{terrain}}

There are also relevent aspects of the terrain that can be easily
derived from a digital elevation model. We'll use NASA-JPL's SRTM v3 1
arc-second resolution DEM.

\hypertarget{selection-of-predictor-variables}{%
\subsubsection{Selection of predictor
variables}\label{selection-of-predictor-variables}}

We reviewed existing floras to identify specific variables from these
datasets that are relevant to the niche of our target taxa. The results
are summarised below.

\begin{verbatim}
## Parsed with column specification:
## cols(
##   Taxon = col_character(),
##   Factor = col_character(),
##   Notes = col_character(),
##   Reference = col_logical()
## )
\end{verbatim}

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{llc}
\toprule
Factor & Notes & Reference \\ 
\midrule
\multicolumn{1}{l}{Triticum sp.} \\ 
\midrule
Precipitation & Adapted to dry conditions, at least 400 mm per annum for T. dicoccoides, or 300–350 mm for T. urartu and T. boeoticum. & NA \\ 
Soil type & T. urartu and T. boeoticum prefer basalt calcareous soils. & NA \\ 
Soil acidity & Found in acidic soils (limited distribution) & NA \\ 
Terrain & 100 m asl to 1600–1650 m asl. T. dicoccoides grows on mountain slopes. & NA \\ 
\midrule
\multicolumn{1}{l}{Hordeum spontaneum} \\ 
\midrule
Precipitation & Adapted to dry conditions, at least 200–250 mm rainfall per annum. & NA \\ 
Temperature & Not tolerant of cool conditions. & NA \\ 
Soil type & Tolerant of poor calcareous soils. & NA \\ 
Terrain & Grows on rocky limestone slopes, marly banks, and waste places, between 30 and 1650 m asl. & NA \\ 
\midrule
\multicolumn{1}{l}{Stipa sp.} \\ 
\midrule
Terrain & Grows on limestone mountain slopes between 400 and 2500 m asl. & NA \\ 
\midrule
\multicolumn{1}{l}{Vitex sp.} \\ 
\midrule
Water & Grows close to water resources, banks of streams and dry wadis & NA \\ 
Terrain & No higher than 750 m asl, mostly on sandy groumd, alluvial soils, rocky areas, dry river beds. & NA \\ 
\midrule
\multicolumn{1}{l}{Zilla spinosa} \\ 
\midrule
Soil type & Sandy deserts, dry soils, stony ground. & NA \\ 
\midrule
\multicolumn{1}{l}{Bolboschoenus glaucus} \\ 
\midrule
Water & A wetland plant, it grows in the shores of the lakes, often underwater. The subspecies maritimus (sea club-rush) grows in freshwater or saline marshes with Typha, Juncus and Phragmites, stagnant swamps, water meadows with Orchis palustris, by streams and rivers, dried up river beds, soda lakes, thermal springs, saline and alluvial flats, beaches, edge o/irrigation ditches and rice fields. & NA \\ 
Salinity & Some club rush like sea-club-rush can tolerate high salinity. & NA \\ 
Terrain & Grows anywhere between 0 and 2000 m asl. & NA \\ 
\midrule
\multicolumn{1}{l}{Fraxinus sp.} \\ 
\midrule
Water & Along with rivers and streams. 400-1900 small, by streams often in comparatively dry and rocky places & NA \\ 
Terrain & Mostly dryish, rocky places, in macchie, deciduous scrub or forest, Pinus brutia or P. nigra forests, or mixed forest, 650–1700 m asl. & NA \\ 
\bottomrule
\end{longtable}

Based on these results, we made the following initial selection of
predictor variables to build the models.

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{llll}
\toprule
variable & units & source & code \\ 
\midrule
\multicolumn{1}{l}{Climate} \\ 
\midrule
Minimum temperature of coldest month & °C ×10 & WorldClim 1.4 & bio6 \\ 
Total annual precipitation & mm & WorldClim 1.4 & bio12 \\ 
Precipitation of wettest month & mm & WorldClim 1.4 & bio13 \\ 
Precipitation of driest month & mm & WorldClim 1.4 & bio14 \\ 
Precipitation seasonality & σ/μ & WorldClim 1.4 & bio15 \\ 
\midrule
\multicolumn{1}{l}{Soil} \\ 
\midrule
Organic carbon content & g per kg fine earth & SoilGrids250m & ORCDRC\_M\_sl2\_250m \\ 
Coarse fragment content & \% volume & SoilGrids250m & CRFVOL\_M\_sl2\_250m \\ 
Sand content & \% mass & SoilGrids250m & SNDPPT\_M\_sl2\_250m \\ 
Acidity & pH ×10 in H2O & SoilGrids250m & PHIHOX\_M\_sl2\_250m \\ 
Depth to bedrock & cm (up to 200 cm) & SoilGrids250m & BDRICM\_M\_250m \\ 
\bottomrule
\end{longtable}

One aspect that we cannot easily incorporate into the models is the
interaction between these taxa and other plant and animal species:
cereal grasses grow best when they have no competition from perennial
plants, \emph{Fraxinus} often grows in scrub or forest, and so on.

We also lack data for soil salinity so far.

\hypertarget{combining-predictor-variables}{%
\subsubsection{Combining predictor
variables}\label{combining-predictor-variables}}

These datasets come with different resolutions. To use them all in the
same analysis, we resample them all to the finest dataset (WorldClim, 30
arc-seconds).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pred}\OperatorTok{$}\NormalTok{raster <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(pred}\OperatorTok{$}\NormalTok{raster, raster}\OperatorTok{::}\NormalTok{resample, }\DataTypeTok{y =}\NormalTok{ bioclim, }\DataTypeTok{method =} \StringTok{"ngb"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{model-construction}{%
\subsection{Model construction}\label{model-construction}}

We will model the ecological niche of each species using a generalised
linear regression.

First we extract the values of the predictor variables at each
occurrence and pseudo-absence location:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: refactor to use pred}
\NormalTok{bioclim }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{extract}\NormalTok{(}\KeywordTok{as_Spatial}\NormalTok{(occ)) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\NormalTok{\{}\KeywordTok{cbind}\NormalTok{(occ, .)\} }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(occ)) ->}
\StringTok{  }\NormalTok{occ}

\NormalTok{soilgrids }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{extract}\NormalTok{(}\KeywordTok{as_Spatial}\NormalTok{(occ)) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\NormalTok{\{}\KeywordTok{cbind}\NormalTok{(occ, .)\} }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(occ)) ->}
\StringTok{  }\NormalTok{occ}
\end{Highlighting}
\end{Shaded}

\hypertarget{species-response-curves}{%
\subsubsection{Species response curves}\label{species-response-curves}}

We can use species response curves to visualise the distribution of
known occurrences in relation to these environmental gradients. This
will also help us parameterise the ecological niche models.

Species response curves are constructed with a simple logistic
regression, fitting the probability of occurrence at a given location to
The use of a linear regression assumes a unimodal response. We fit
first, second, and third order polynomial models to each
species/predictor combination and calculate the Akaike Information
Criterion (AIC).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Summarise occurrences on a grid}
\NormalTok{palestine }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{as_Spatial}\NormalTok{() }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{rasterize}\NormalTok{(bioclim) ->}
\StringTok{  }\NormalTok{palestine_raster}

\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(occurrence }\OperatorTok{==}\StringTok{ }\OtherTok{TRUE}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{do}\NormalTok{(}\DataTypeTok{occ_grid =} \KeywordTok{raster_burn}\NormalTok{(., palestine_raster)) ->}
\StringTok{  }\NormalTok{occ_grids}

\CommentTok{# Fit poly(1:3) to each taxon/predictor combination}
\KeywordTok{expand.grid}\NormalTok{(}\DataTypeTok{taxon =} \KeywordTok{unique}\NormalTok{(occ}\OperatorTok{$}\NormalTok{taxon),}
            \DataTypeTok{predictor =}\NormalTok{ pred}\OperatorTok{$}\NormalTok{code,}
            \DataTypeTok{polyn =} \DecValTok{1}\OperatorTok{:}\DecValTok{3}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon, predictor, polyn) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{do}\NormalTok{(}
    \CommentTok{# }\AlertTok{TODO}\CommentTok{: It works but I hate it.}
    \DataTypeTok{fit =} \KeywordTok{model_response}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(occ_grids[occ_grids}\OperatorTok{$}\NormalTok{taxon }\OperatorTok{==}\StringTok{ }\NormalTok{.}\OperatorTok{$}\NormalTok{taxon, }\StringTok{"occ_grid"}\NormalTok{])[[}\DecValTok{1}\NormalTok{]],}
                         \KeywordTok{unlist}\NormalTok{(pred[pred}\OperatorTok{$}\NormalTok{code }\OperatorTok{==}\StringTok{ }\NormalTok{.}\OperatorTok{$}\NormalTok{predictor, }\StringTok{"raster"}\NormalTok{])[[}\DecValTok{1}\NormalTok{]],}
\NormalTok{                         .}\OperatorTok{$}\NormalTok{polyn)}
\NormalTok{  ) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{glance}\NormalTok{(fit) ->}\StringTok{ }
\StringTok{  }\NormalTok{response_glm}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred

## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
\end{verbatim}

The optimal model is the one with the lowest AIC.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{response_glm }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon, predictor) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{optimum =}\NormalTok{ polyn }\OperatorTok{==}\StringTok{ }\KeywordTok{which.min}\NormalTok{(AIC)) ->}
\StringTok{  }\NormalTok{response_glm}

\CommentTok{# }\AlertTok{TODO}
\CommentTok{# Re-fit and plot optimal }
\CommentTok{# model_response(occ_grids$occ_grid[[1]], pred$raster[[1]], 2) %>% predict(data.frame(pred = -72:109, type = "response")) %>% plot()}
\CommentTok{# Need to dplyrise this – see broom::augment()}
\CommentTok{# response_glm %>% }
\CommentTok{#   filter(optimum == TRUE) %>% }
\CommentTok{#   slice(1:3) %>% }
\CommentTok{#   do(}
\CommentTok{#     fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],}
\CommentTok{#                          unlist(pred[pred$code == .$predictor,"raster"])[[1]],}
\CommentTok{#                          .$polyn)}
\CommentTok{#   ) }
\end{Highlighting}
\end{Shaded}

\hypertarget{model-fitting}{%
\subsubsection{Model fitting}\label{model-fitting}}

Now we can fit regression models combining all the predictor variables.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: use optimal polynomial order}
\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{remove_missing}\NormalTok{(}\DataTypeTok{vars =} \KeywordTok{c}\NormalTok{(}\StringTok{"bio1"}\NormalTok{, }\StringTok{"bio2"}\NormalTok{, }\StringTok{"bio3"}\NormalTok{, }\StringTok{"bio4"}\NormalTok{, }\StringTok{"bio5"}\NormalTok{, }\StringTok{"bio6"}\NormalTok{, }
                          \StringTok{"bio7"}\NormalTok{, }\StringTok{"bio8"}\NormalTok{, }\StringTok{"bio9"}\NormalTok{, }\StringTok{"bio10"}\NormalTok{, }\StringTok{"bio11"}\NormalTok{, }\StringTok{"bio12"}\NormalTok{, }
                          \StringTok{"bio13"}\NormalTok{, }\StringTok{"bio14"}\NormalTok{, }\StringTok{"bio15"}\NormalTok{, }\StringTok{"bio16"}\NormalTok{, }\StringTok{"bio17"}\NormalTok{, }\StringTok{"bio18"}\NormalTok{, }
                          \StringTok{"bio19"}\NormalTok{, }\StringTok{"BDRICM_M_250m"}\NormalTok{, }\StringTok{"CRFVOL_M_sl2_250m"}\NormalTok{, }
                          \StringTok{"ORCDRC_M_sl2_250m"}\NormalTok{, }\StringTok{"PHIHOX_M_sl2_250m"}\NormalTok{, }
                          \StringTok{"SNDPPT_M_sl2_250m"}\NormalTok{)) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{do}\NormalTok{(}\DataTypeTok{model =} \KeywordTok{glm}\NormalTok{(occurrence }\OperatorTok{~}\StringTok{ }
\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio6, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio12, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio13, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
\StringTok{                   }\CommentTok{# poly(bio14, 2) +}
\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio15, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
\StringTok{                   }\KeywordTok{poly}\NormalTok{(BDRICM_M_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
\StringTok{                   }\KeywordTok{poly}\NormalTok{(CRFVOL_M_sl2_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
\StringTok{                   }\KeywordTok{poly}\NormalTok{(ORCDRC_M_sl2_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{                   }\KeywordTok{poly}\NormalTok{(PHIHOX_M_sl2_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{                   }\KeywordTok{poly}\NormalTok{(SNDPPT_M_sl2_250m, }\DecValTok{2}\NormalTok{),}
                 \DataTypeTok{data =}\NormalTok{ .)) ->}
\StringTok{  }\NormalTok{models}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: Removed 3337 rows containing missing values.
\end{verbatim}

\hypertarget{predicting-occurrences}{%
\subsection{Predicting occurrences}\label{predicting-occurrences}}

\hypertarget{current-conditions}{%
\subsubsection{Current conditions}\label{current-conditions}}

We can now use the fitted models to predict the occurrence of the
different taxa according to the predictor variables. Initially, we do
this based on current conditions. This allows us to extrapolate from
regions with good occurrence data (i.e.~Israel/Palestine) to those
without. It also hints at what the distribution of the plant might be
under ``ideal conditions'', without agriculture, grazing, etc.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{models }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{do}\NormalTok{(}\DataTypeTok{taxon =}\NormalTok{ .}\OperatorTok{$}\NormalTok{taxon,}
     \DataTypeTok{model =}\NormalTok{ .}\OperatorTok{$}\NormalTok{model,}
     \DataTypeTok{pred_current =}\NormalTok{ raster}\OperatorTok{::}\KeywordTok{predict}\NormalTok{(}\DataTypeTok{object =}\NormalTok{ raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(pred}\OperatorTok{$}\NormalTok{raster),}
                                    \DataTypeTok{model =}\NormalTok{ .}\OperatorTok{$}\NormalTok{model)) ->}
\StringTok{  }\NormalTok{models}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(models}\OperatorTok{$}\NormalTok{pred_current) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_names}\NormalTok{(models}\OperatorTok{$}\NormalTok{taxon) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{plot}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\includegraphics{analysis_files/figure-latex/pred_current_plot-1.pdf}

\hypertarget{binning}{%
\subsubsection{Binning}\label{binning}}

To more easily interpret these predictions, we could convert the raw
probability scores into confidence intervals, based on the proportion of
known occurrences it successfully predicts.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(models}\OperatorTok{$}\NormalTok{pred_current) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_names}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(}\StringTok{"pred_current_"}\NormalTok{, models}\OperatorTok{$}\NormalTok{taxon)) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{extract}\NormalTok{(}\KeywordTok{as_Spatial}\NormalTok{(occ)) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\NormalTok{\{}\KeywordTok{cbind}\NormalTok{(occ, .)\} }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(occ)) ->}
\StringTok{  }\NormalTok{occ2}
\end{Highlighting}
\end{Shaded}

\end{document}

### 2019-07-16 R/data.R

@@ -1,0 +1,136 @@
+# data.R
+
+
+# Occurrence data (GBIF) --------------------------------------------------
+
+#' Fetch GBIF data
+#'
+#' A simplified interface to [dismo::gbif()] with caching.
+#'
+#' @param genus       Character. Genus name.
+#' @param species     Character. Species name. Use '*' for fuzzy matching.
+#' @param ext         `extent` object. Geographic extent of records to return.
+#' @param cache_path  Character. Path to cache data. Downloaded data is cached as
+#'                    a CSV with an automatically generated filename.
+#' @param redownload  Logical. If `TRUE`, cached data will be ignored. Default: `FALSE`.
+#'
+#' @details
+#'
+#' The return value is always read from the cached file to ensure consistency.
+#' See [dismo::gbif()] for more details.
+#'
+#' @return  A `tibble` of downloaded GBIF records.
+#'
+#' @export
+fetch_gbif <- function(genus, species, ext, cache_path, redownload = FALSE) {
+  if(species != "*") {
+    cache_file <- file.path(cache_path, paste0("gbif_", tolower(genus), "_",
+                                               tolower(species), ".csv"))
+  }
+  else {
+    cache_file <- file.path(cache_path, paste0("gbif_", tolower(genus), ".csv"))
+  }
+  cache_file <- stringr::str_remove(cache_file, "\\*")
+
+  if (!file.exists(cache_file) | redownload == TRUE) {
+    message("Downloading '", genus, " ", species, "' from GBIF...")
+    dismo::gbif(genus, species, ext) %>%
+      tibble::as_tibble() %>%
+      dplyr::select(gbifID, genus, species, lon, lat, basisOfRecord) %>%
+      readr::write_csv(cache_file)
+  }
+
+  message("Reading ", cache_file)
+  return(readr::read_csv(cache_file))
+}
+
+# Bioclimatic data --------------------------------------------------------
+
+#' Regional Bioclimatic Variables
+#'
+#' Downloads WorldClim data and derives bioclimatic variables for an arbitrary
+#' region (or extent).
+#'
+#' @param extent      A [raster::Extent] object specifying the region of interest,
+#'                    or an object (e.g. matrix, Spatial*) that can be passed to
+#'                    [raster::extent()]
[…]

### 2019-07-16 R/edm.R

@@ -1,0 +1,41 @@
+# enm.R
+# Ecological niche modelling functions
+# Joe Roe <jwg983@hum.ku.dk>
+
+#' Generate pseudo-absence points
+#'
+#' @param region  object of class `sf` or `sfc`. The region in which to generate
+#'                pseudo-absence points.
+#' @param ...     Name-value pairs of expressions to be added as fixed attributes
+#'                of the pseudo-absences. See [dplyr::mutate()].
+#' @param N       integer. Number of pseudo-absence points to generate.
+#'
+#' @return
+#' @export
+#'
+#' @examples
+generate_psabsences <- function(region, ..., N = 10000) {
+  sf::st_sample(region, N) %>%
+    as_tibble() %>%
+    mutate(...) %>%
+    return()
+}
+
+#' @export
+model_response <- function(occ, pred, polyn) {
+  if(!raster::compareRaster(occ, pred)) {
+    stop("Occurrence and predictor rasters must have the same extent, resolution, and projection.")
+  }
+
+  occ <- as.vector(occ)
+  pred <- as.vector(pred)
+  # poly() can't handle missing values, so trim where predictor variable is NA
+  occ <- occ[!is.na(pred)]
+  pred <- pred[!is.na(pred)]
+  data <- data.frame(pred = pred, occ = occ)
+
+  mod <- glm(occ ~ poly(pred, degree = polyn, raw = TRUE), family = "binomial",
+             data = data, na.action = na.exclude, control = list(maxit = 100))
+
+  return(mod)
+}

### 2019-07-16 R/soilgrids.R

@@ -1,0 +1,4 @@
+# soilgrids.R
+# Functions for downloading data from SoilGrids
+
+

### 2019-07-16 R/utils-spatial.R

@@ -1,0 +1,16 @@
+#' Burn points into a raster
+#'
+#' @param points `sf`.
+#' @param raster
+#'
+#' @return
+#' @export
+#'
+#' @examples
+raster_burn <- function(points, raster) {
+  points %>%
+    sf::as_Spatial() %>%
+    raster::rasterize(raster, field = 1, background = 0) %>%
+    raster::mask(raster) %>%
+    return()
+}

### 2023-04-26 analysis/analysis.Rmd
Updates from Feb 2020

@@
 output:
+  pdf_document: 
+    latex_engine: xelatex
   html_document:
     df_print: paged
-  pdf_document:
-    latex_engine: xelatex
 ---
 
-## Dependencies
-
-```{r packages, warning=FALSE, message=FALSE}
+```{r packages, include=FALSE}
 library("tidyverse")
@@
 library("sf")
-library("raster")
+library("stars")
+library("parallel")
 library("broom")
@@
 library("gt")
+library("viridis")
 library("BadiaPaleoFloraENM")
+```
+
+```{r defs, include=FALSE}
+latlong <- 4326
+utm37n <- 32637
 ```
@@
 
-```{r regions, warning=FALSE, message=FALSE}
+```{r regions, include=FALSE}
 sw_asia <- extent(c(30, 50, 25, 40))
 s_levant <- extent(c(34, 39, 29, 34))
-palestine <- ne_countries(scale = 10, country = c("Israel", "Palestine"),
-             returnclass = "sf") %>% st_union() %>% st_sf()
 
-land <- read_sf("data/raw_data/ne/ne_10m_land.shp")[,"featurecla"] %>%
-  st_crop(sw_asia)
+ne_countries(scale = 10, country = c("Israel", "Palestine"), returnclass = "sf") %>% 
+  st_union() ->
+  palestine
+
+read_sf("data/raw_data/ne/ne_10m_land.shp")[,"featurecla"] %>%
+  st_crop(sw_asia) -> 
+  land
+
 rivers <- read_sf("data/raw_data/ne/ne_10m_rivers_lake_centerlines_scale_rank.shp") %>% 
   st_crop(s_levant)
+
 water <- rbind(read_sf("data/raw_data/ne/ne_10m_ocean.shp")[,"featurecla"],
-               read_sf("data/raw_data/ne/ne_10m_lakes.shp")[,"featurecla"]) %>% 
-  st_crop(sw_asia)
+               read_sf("data/raw_data/ne/ne_10m_lakes.shp")[,"featurecla"])
 ```
 
-```{r regions-map, echo=FALSE, results='hide'}
-basemap <- ggplot() + 
-  geom_sf(data = land, fill ="grey") +
-  geom_sf(data = palestine, fill = "lightgrey") + 
-  geom_sf(data = rivers, colour = "white") +
-  geom_sf(data = water, fill = "white") +
+```{r regions-map, echo=FALSE}
+ggplot() + 
+  geom_sf(data = palestine, fill = viridis(2)[2], colour = NA) +
+  geom_sf(data = rivers, colour = "#999999") +
+  geom_sf(data = water, fill = "#eeeeee", colour = "#999999") +
   coord_sf(xlim = c(xmin(s_levant), xmax(s_levant)),
-           ylim = c(ymin(s_levant), ymax(s_levant)))
-basemap
+           ylim = c(ymin(s_levant), ymax(s_levant))) +
+  theme_bw()
 ```
 
-## Occurrence data
+## Background
+
+### Plant exploitation in the Epipalaeolithic Southern Levant
+
+### Ecological niche modelling
+
+## Data
+
+### Occurrences
 
@@
 
-```{r occurrence-data, warning=FALSE, message=FALSE}
+```{r occurrence-data, include=FALSE}
 # Download or read data from GBIF
+# TODO: T. boeoticum ? (equate with T. urartu?)
 gbif_path <- here::here("analysis/data/raw_data/gbif")
@@
   mutate(taxon = case_when(genus == "Hordeum" ~ "Hordeum spontaneum",
-                           genus == "Bolboschoenus" ~ "Bolboschoenus sp. (=Scirpus maritimus)",
+                           genus == "Bolboschoenus" ~ "Bolboschoenus sp.",
                            species == "Triticum urartu" ~ "Triticum urartu",
@@
 It's interesting to map the current distribution of these plants.
-```{r occurrence-map, echo=FALSE, results='hide'}
+```{r occurrence-map, echo=FALSE}
 ggplot(occ, aes(x = lon, y = lat)) +
+  geom_sf(data = water, inherit.aes = FALSE, fill = "#eeeeee", colour = "#999999") +
   geom_sf(data = occ, shape = 1, size=0.5) +
-  geom_sf(data = water, inherit.aes = FALSE, fill = "white", colour = "darkgrey") +
-  facet_wrap(~taxon)
+  coord_sf(xlim = c(xmin(sw_asia), xmax(sw_asia)),
+           ylim = c(ymin(sw_asia), ymax(sw_asia))) +
+  facet_wrap(~taxon) +
+  theme_bw() +
+  theme(axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        axis.title = element_blank())
 ```
 
-## Pseudo-absences
+### Pseudo-absences
 
@@
 There are several ways to do this. 
-We follow the advice of Barbet-Massin et al. (2012) for regression-based species distribution models and simply use a large (~10000) random sample of points.
+We follow the advice of Barbet-Massin et al.
+(2012) for regression-based species distribution models and use a large (~10000) random sample of points, weighted equally against the presences in the regression.
 
-```{r psabsences, warning=FALSE, message=FALSE}
+```{r psabsences, include=FALSE}
+n_psab <- 10000
+
 occ <- as_tibble(occ)
@@
   occ <- bind_rows(occ, generate_psabsences(palestine, taxon = taxon,
-                                            occurrence = FALSE, N = 10000))
+                                            occurrence = FALSE, N = n_psab))
 }
@@
 rm(taxon)
+
+# Weights
+occ %>% 
+  group_by(taxon, occurrence) %>% 
+  mutate(weight = 1 + (n_psab / n() - 1)) ->
+  occ
 ```
 
-## Environmental data
+### Environmental data
 
-### Bioclimatic variables
+#### Current climate
 
@@
 ```{r bioclim-table, echo=FALSE}
-tibble(biovar = paste0("bio", 1:16),
+tibble(biovar = paste0("bio", 1:19),
        definition = c("Mean annual temperature",
@@
                       "Precipitation seasonality",
-                      "Precipitation of wettest quarter"),
+                      "Precipitation of wettest quarter",
+                      "Precipitation of driest quarter",
+                      "Precipitation of warmest quarter",
+                      "Precipitation of coldest quarter"),
        formula = c("",
@@
                    "",
+                   "",
+                   "",
+                   "",
                    "")
@@
 
-Here the climate data is from WorldClim 1.4 (<http://www.worldclim.org>), a global dataset compiled from weather station records from the period 1960–1990.
-WorldClim has been widely used in ecological niche modelling.
-Other global climate datasets (e.g. CHELSA, MERRAClim) and a newer version of WorldClim are now available, so it may be a good idea to investigate using these instead.
-The WorldClim data is large so again it is cached in the package's `derived_data` directory.
+Here the current bioclimatic data is from CHELSA (<http://chelsa-climate.org/bioclim/>).
 
-```{r bioclim-data, echo=FALSE}
-# bioclim <- get_bioclim(s_levant, resolution = 0.5, cache_path = "data/derived_data")
-# Workaround for memory problems in Safawi
-load("data/derived_data/bioclim.RData")
-plot(bioclim)
+```{r bioclim-data, include=FALSE}
+fetch_paleoclim("cur", "2_5m", s_levant,
+                "./data/derived_data/paleoclim/paleoclim_cur.grd") %>% 
+  st_as_stars() ->
+  bioclim
 ```
 
-### Soil variables
+#### Palaeoclimate
+
+```{r paleoclim-data, include=FALSE}
+tibble(
+  period_code = c("eh", "yds", "ba", "hs1", "lgm"),
+  period = c("Early Holocene", "Younger Dryas", "Bølling-Allerød", "Heinrich 1", "Last Glacial Maximum"),
+  raster = c(
+    fetch_paleoclim("eh", "2_5m", s_levant, 
+                    "./data/derived_data/paleoclim/paleoclim_eh.grd"),
+    fetch_paleoclim("yds", "2_5m", s_levant, 
+                    "./data/derived_data/paleoclim/paleoclim_yds.grd"),
+    fetch_paleoclim("ba", "2_5m", s_levant, 
+                    "./data/derived_data/paleoclim/paleoclim_ba.grd"),
+    fetch_paleoclim("hs1", "2_5m", s_levant, 
+                    "./data/derived_data/paleoclim/paleoclim_hs1.grd"),
+    fetch_paleoclim("lgm", "2_5m", s_levant, 
+                    "./data/derived_data/paleoclim/paleoclim_lgm.grd")
+  )
+) %>% 
+  mutate(raster = st_as_stars(raster)) ->
+  paleoclim
+```
+
+### Soil
 
@@
 
-```{r soil-data}
-soilgrids <- raster::stack(list.files("data/raw_data/soilgrids", ".*tif", full.names = TRUE))
-plot(soilgrids)
+```{r soil-data, include=FALSE}
+soilgrids <- raster::stack(list.files("data/raw_data/soilgrids", ".*tif",
+                                      full.names = TRUE))
 ```
@@
 There are also relevent aspects of the terrain that can be easily derived from a digital elevation model.
-We'll use NASA-JPL's SRTM v3 1 arc-second resolution DEM.
+We'll use NASA-JPL's SRTM v3 3 arc-second resolution DEM, smoothed with Sun's denoising algorithm [TODO: CITE].
+Slope was calculated from this surface using GDAL 3.0.0 and the topographic wetness index (TWI) using the r.topidx module of GRASS 7.8.
+The DEM was reprojected to a conformal projection (Jordan Transverse Mercator, EPSG:19995) for these calculations, using a bilinear interpolation, then back to geographic coordinates.
 
-### Selection of predictor variables
+```{r terrain-data, include=FALSE}
+raster::stack(paste0("./data/raw_data/srtm/",
+                     c("srtm3v3s_slevant.tif",
+                       "srtm3v3s_slevant_slope.tif",
+                       "srtm3v3s_slevant_twi.tif"))) %>% 
+  set_names(c("elev", "slope", "twi")) ->
+  terrain
+```
+
+## Model
+
+### Selection of predictors
 
@@
 
-```{r review-table, echo=FALSE, warning=FALSE, message=FALSE}
+```{r review-table, echo=FALSE}
 read_csv("data/raw_data/flora_review.csv") %>% 
   group_by(Taxon) %>% 
+  # TODO: add refs
+  dplyr::select(-Reference) %>% 
   gt()
@@
 pred <- bind_rows(
-  tibble(group = rep("Climate", 5),
+  tibble(group = rep("Climate", 4),
+         time = rep("now", 4),
          variable = c("Minimum temperature of coldest month",
                       "Total annual precipitation", "Precipitation of wettest month",
-                      "Precipitation of driest month", "Precipitation seasonality"),
-         units = c("°C ×10", "mm", "mm", "mm", "σ/μ"),
-         source = rep("WorldClim 1.4", 5),
-         code = c("bio6", "bio12", "bio13", "bio14", "bio15"),
+                      "Precipitation seasonality"),
+         units = c("°C ×10", "mm", "mm", "σ/μ"),
+         source = rep("PaleoClim 1.0", 4),
+         code = c("bio_6", "bio_12", "bio_13", "bio_15"),
          raster = as.list(raster::subset(bioclim,
-                                         c("bio6", "bio12", "bio13", "bio14", "bio15")))),
+                                         c("bio_6", "bio_12", "bio_13",
+                                           "bio_15")))),
   tibble(group = rep("Soil", 5),
+         time = rep(NA, 5),
          variable = c("Organic carbon content", "Coarse fragment content",
@@
                                          c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
-                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"))))
-  # tibble(group = rep("Terrain", 3),
-  #        variable = c("Elevation", "Slope", "Topographic Wetness Index"),
-  #        units = c("m", "°", "ln(upslope area / tan(slope))"),
-  #        source = rep("SRTM DEM v3", 3),
-  #        code = c("srtm1v3_slevant", "srtm1v3_slevant_slope",
-  #                         "srtm1v3_slevant_twi"))
+                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m")))),
+  tibble(group = rep("Terrain", 3),
+         time = rep(NA, 3),
+         variable = c("Elevation", "Slope", "Topographic Wetness Index"),
+         units = c("m", "°", "ln(upslope area / tan(slope))"),
+         source = rep("SRTM DEM v3", 3),
+         code = c("srtm1v3s_slevant", "srtm1v3s_slevant_slope",
+                  "srtm1v3s_slevant_twi"),
+         raster = as.list(terrain))
 )
 
+# PaleoClim
+paleoclim %>% 
+  transmute(group = "Climate", time = period_code, raster = raster) %>% 
+  mutate(raster = map(raster, ~raster::unstack(.))) %>% 
+  unnest() %>% 
+  mutate(code = unlist(map(raster, names))) %>% 
+  filter(code %in% c("bio_6", "bio_12", "bio_13", "bio_15")) %>% 
+  {bind_rows(pred, .)} ->
+  pred
+
 pred %>%
-  dplyr::select(-raster) %>% 
+  filter(time == "now" | is.na(time)) %>% 
+  dplyr::select(-raster, -time) %>% 
   group_by(group) %>% 
-  gt()
+  gt() %>% 
+  cols_label(variable = "Variable", units = "Units", source = "Dataset",
+             code = "Source code")
 ```
@@
 
-### Combining predictor variables
+These datasets come with different resolutions.
+To use them all in the same analysis, we resample them all to the finest dataset (the SRTM terrain variables, at 3 arc-second resolution).
 
-These datasets come with different resolutions.
-To use them all in the same analysis, we resample them all to the finest dataset (WorldClim, 30 arc-seconds).
-
-```{r predictor-resample}
-pred$raster <- lapply(pred$raster, raster::resample, y = bioclim, method = "ngb")
+```{r predictor-resample, include=FALSE}
+# Parallelisation is Linux and Mac only.
+# 
+pred$raster <- mclapply(pred$raster, raster::resample, y = terrain, method = "ngb",
+                        mc.cores = detectCores())
+# pred$raster <- lapply(pred$raster, raster::resample, y = terrain, method = "ngb")
 ```
-
-## Model construction
 
@@
 
-```{r predictor-extraction}
-# TODO: refactor to use pred
-bioclim %>% 
-  raster::extract(as_Spatial(occ)) %>% 
-  {cbind(occ, .)} %>% 
-  st_as_sf(crs = st_crs(occ)) ->
-  occ
-
-soilgrids %>% 
+```{r predictor-extraction, include=FALSE}
+pred %>% 
+  filter(time == "now" | is.na(time)) %>% 
+  magrittr::extract2("raster") %>% 
+  stack() %>% 
   raster::extract(as_Spatial(occ)) %>% 
@@
 
-```{r response-curves-fit, warning=FALSE, message=FALSE}
-# Summarise occurrences on a grid
-palestine %>% 
-  as_Spatial() %>% 
-  raster::rasterize(bioclim) ->
-  palestine_raster
+```{r response-curves-fit, include=FALSE}
+occ %>% 
+  gather("pred_var", "pred_val", bio_6:twi) %>% 
+  # NAs from gaps in the predictor data.
+  drop_na(pred_val) %>%
+  group_by(taxon, pred_var) %>% 
+  nest() %>% 
+  expand(taxon, pred_var, data, polyn = 1:3) %>%
+  mutate(
+    response_model = paste0("occurrence ~ poly(pred_val, ", polyn, ")"),
+    response_fit = map2(data, response_model,
+                        ~ glm(as.formula(.y),
+                              family = "binomial",
+                              data = .x,
+                              weights = weight)),
+    response_glance = map(response_fit, glance)
+  ) ->
+  response_models
 
-occ %>% 
-  filter(occurrence == TRUE) %>% 
-  group_by(taxon) %>% 
-  do(occ_grid = raster_burn(., palestine_raster)) ->
-  occ_grids
-
-# Fit poly(1:3) to each taxon/predictor combination
-expand.grid(taxon = unique(occ$taxon),
-            predictor = pred$code,
-            polyn = 1:3) %>% 
-  group_by(taxon, predictor, polyn) %>% 
-  do(
-    # TODO: It works but I hate it.
-    fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
-                         unlist(pred[pred$code == .$predictor, "raster"])[[1]],
-                         .$polyn)
-  ) %>% 
-  glance(fit) -> 
-  response_glm
+# Select model with lowest AIC for each taxon–predictor combination
+response_models %>%
+  unnest(response_glance) %>% 
+  group_by(taxon, pred_var) %>% 
+  filter(AIC == min(AIC)) ->
+  response_models
 ```
@@
 ```{r response-curves-table, echo=FALSE}
-# response_glm %>% 
+# response_models %>% 
 #   spread(polyn, aic) %>% 
@@
 
-```{r response-curves-optimal}
-response_glm %>%
-  group_by(taxon, predictor) %>%
-  mutate(optimum = polyn == which.min(AIC)) ->
-  response_glm
-
-# TODO
-# Re-fit and plot optimal 
-# model_response(occ_grids$occ_grid[[1]], pred$raster[[1]], 2) %>% predict(data.frame(pred = -72:109, type = "response")) %>% plot()
-# Need to dplyrise this – see broom::augment()
-# response_glm %>% 
-#   filter(optimum == TRUE) %>% 
-#   slice(1:3) %>% 
-#   do(
-#     fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
-#                          unlist(pred[pred$code == .$predictor,"raster"])[[1]],
-#                          .$polyn)
-#   ) 
-```
-
-### Model fitting
+### Logistic regression
 
@@
 
-```{r model_fit, warning=FALSE, message=FALSE}
-# TODO: use optimal polynomial order
+```{r model_fit, include=FALSE}
 occ %>% 
+  drop_na(bio_6:twi) %>% 
   group_by(taxon) %>% 
-  remove_missing(vars = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", 
-                          "bio7", "bio8", "bio9", "bio10", "bio11", "bio12", 
-                          "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", 
-                          "bio19", "BDRICM_M_250m", "CRFVOL_M_sl2_250m", 
-                          "ORCDRC_M_sl2_250m", "PHIHOX_M_sl2_250m", 
-                          "SNDPPT_M_sl2_250m")) %>% 
-  do(model = glm(occurrence ~ 
-                   poly(bio6, 2) +
-                   poly(bio12, 2) +
-                   poly(bio13, 2) +
-                   # poly(bio14, 2) +
-                   poly(bio15, 2) +
-                   poly(BDRICM_M_250m, 2) +
-                   poly(CRFVOL_M_sl2_250m, 2) +
-                   poly(ORCDRC_M_sl2_250m, 2) + 
-                   poly(PHIHOX_M_sl2_250m, 2) + 
-                   poly(SNDPPT_M_sl2_250m, 2),
-                 data = .)) ->
-  models
-
+  nest() %>% 
+  mutate(
+    model = map_chr(taxon, ~ paste0(
+      "occurrence ~ ",
+      str_c(paste0("poly(",
+                   filter(response_models, taxon == .x)$pred_var,
+                   ", ",
+                   filter(response_models, taxon == .x)$polyn,
+                   ")"),
+            collapse = " + "))),
+    fit = map2(model, data,
+               ~ glm(as.formula(.x),
+                     family = "binomial",
+                     data = .y,
+                     weights = weight)),
+    glance = map(fit, glance)
+  ) ->
+  enm
 ```
 
-## Predicting occurrences
+## Results
 ### Current conditions
@@
 
-```{r pred_current}
-models %>% 
-  do(taxon = .$taxon,
-     model = .$model,
-     pred_current = raster::predict(object = raster::stack(pred$raster),
-                                    model = .$model)) ->
-  models
+```{r predictions, include=FALSE}
+# Slow
+profvis::profvis({
+enm %>% 
+  mutate(pred_now = map(fit, 
+                        ~raster::predict(
+                          raster::stack(
+                            filter(pred, time == "now" | is.na(time))$raster),
+                          .))) %>% 
+  mutate(pred_eh = map(fit, 
+                        ~raster::predict(
+                          raster::stack(
+                            filter(pred, time == "eh" | is.na(time))$raster),
+                          .))) %>% 
+  mutate(pred_yds = map(fit, 
+                         ~raster::predict(
+                           raster::stack(
+                             filter(pred, time == "yds" | is.na(time))$raster),
+                           .))) %>% 
+  mutate(pred_ba = map(fit, 
+                        ~raster::predict(
+                          raster::stack(
+                            filter(pred, time == "ba" | is.na(time))$raster),
+                          .))) %>% 
+  mutate(pred_hs1 = map(fit, 
+                         ~raster::predict(
+                           raster::stack(
+                             filter(pred, time == "hs1" | is.na(time))$raster),
+                           .))) %>% 
+  mutate(pred_lgm = map(fit, 
+                         ~raster::predict(
+                           raster::stack(
+                             filter(pred, time == "lgm" | is.na(time))$raster),
+                           .))) %>% 
+  mutate(pred_now = map(pred_now, logit_to_p),
+         pred_eh = map(pred_eh, logit_to_p),
+         pred_yds = map(pred_yds, logit_to_p),
+         pred_ba = map(pred_ba, logit_to_p),
+         pred_hs1 = map(pred_hs1, logit_to_p),
+         pred_lgm = map(pred_yds, logit_to_p)) ->
+  enm
+})
 ```
@@
 ```{r pred_current_plot, echo=FALSE}
-raster::stack(models$pred_current) %>%
-  set_names(models$taxon) %>% 
+raster::stack(enm$pred_now) %>%
+  set_names(enm$taxon) %>% 
   plot()
@@
 
-### Binning
+```{r pred_yds_}
 
-To more easily interpret these predictions, we could convert the raw probability scores into confidence intervals, based on the proportion of known occurrences it successfully predicts.
-
-```{r binning}
-raster::stack(models$pred_current) %>%
-  set_names(paste0("pred_current_", models$taxon)) %>% 
-  raster::extract(as_Spatial(occ)) %>% 
-  {cbind(occ, .)} %>% 
-  st_as_sf(crs = st_crs(occ)) ->
-  occ2
-```
+## Discussion

### 2023-04-26 analysis/analysis.tex
Updates from Feb 2020

@@
 \hypersetup{unicode=true,
-            pdftitle={Ecological niche models of plant species exploited by Epipalaeolithic foragers in Badia (c.~14500 BP)},
+            pdftitle={Ecological niche models of plant species exploited by Epipalaeolithic foragers in the Badia (c.~14500 BP)},
             pdfborder={0 0 0},
@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \usepackage{graphicx,grffile}
@@
   \title{Ecological niche models of plant species exploited by Epipalaeolithic
-foragers in \emph{Badia} (c.~14500 BP)}
+foragers in the \emph{Badia} (c.~14500 BP)}
     \pretitle{\vspace{\droptitle}\centering\huge}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"tidyverse"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Registered S3 methods overwritten by 'ggplot2':
-##   method         from 
-##   [.quosures     rlang
-##   c.quosures     rlang
-##   print.quosures rlang
-\end{verbatim}
-
-\begin{verbatim}
-## -- Attaching packages ------------------------------------------------------------------------------------------ tidyverse 1.2.1 --
-\end{verbatim}
-
-\begin{verbatim}
-## v ggplot2 3.1.1     v purrr   0.3.2
-## v tibble  2.1.2     v dplyr   0.8.1
-## v tidyr   0.8.3     v stringr 1.4.0
-## v readr   1.3.1     v forcats 0.4.0
-\end{verbatim}
-
-\begin{verbatim}
-## -- Conflicts --------------------------------------------------------------------------------------------- tidyverse_conflicts() --
-## x dplyr::filter() masks stats::filter()
-## x dplyr::lag()    masks stats::lag()
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"magrittr"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Attaching package: 'magrittr'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:purrr':
-## 
-##     set_names
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:tidyr':
-## 
-##     extract
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"sf"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Linking to GEOS 3.7.1, GDAL 2.3.2, PROJ 5.2.0
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"raster"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Loading required package: sp
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'raster'
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:magrittr':
-## 
-##     extract
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:dplyr':
-## 
-##     select
-\end{verbatim}
-
-\begin{verbatim}
-## The following object is masked from 'package:tidyr':
-## 
-##     extract
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(}\StringTok{"broom"}\NormalTok{)}
-\KeywordTok{library}\NormalTok{(}\StringTok{"dismo"}\NormalTok{)}
-\KeywordTok{library}\NormalTok{(}\StringTok{"rnaturalearth"}\NormalTok{)}
-\KeywordTok{library}\NormalTok{(}\StringTok{"gt"}\NormalTok{)}
-\KeywordTok{library}\NormalTok{(}\StringTok{"BadiaPaleoFloraSDM"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{sw_asia <-}\StringTok{ }\KeywordTok{extent}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{30}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DecValTok{25}\NormalTok{, }\DecValTok{40}\NormalTok{))}
-\NormalTok{s_levant <-}\StringTok{ }\KeywordTok{extent}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\DecValTok{34}\NormalTok{, }\DecValTok{39}\NormalTok{, }\DecValTok{29}\NormalTok{, }\DecValTok{34}\NormalTok{))}
-\NormalTok{palestine <-}\StringTok{ }\KeywordTok{ne_countries}\NormalTok{(}\DataTypeTok{scale =} \DecValTok{10}\NormalTok{, }\DataTypeTok{country =} \KeywordTok{c}\NormalTok{(}\StringTok{"Israel"}\NormalTok{, }\StringTok{"Palestine"}\NormalTok{),}
-             \DataTypeTok{returnclass =} \StringTok{"sf"}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{st_union}\NormalTok{() }\OperatorTok{%>%}\StringTok{ }\KeywordTok{st_sf}\NormalTok{()}
+\includegraphics{analysis_files/figure-latex/regions-map-1.pdf}
 
-\NormalTok{land <-}\StringTok{ }\KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_land.shp"}\NormalTok{)[,}\StringTok{"featurecla"}\NormalTok{] }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{st_crop}\NormalTok{(sw_asia)}
-\end{Highlighting}
-\end{Shaded}
+\hypertarget{background}{%
+\subsection{Background}\label{background}}
 
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersection assumes that they are planar
-\end{verbatim}
+\hypertarget{plant-exploitation-in-the-epipalaeolithic-southern-levant}{%
+\subsubsection{Plant exploitation in the Epipalaeolithic Southern
+Levant}\label{plant-exploitation-in-the-epipalaeolithic-southern-levant}}
 
-\begin{verbatim}
-## Warning: attribute variables are assumed to be spatially constant
-## throughout all geometries
-\end{verbatim}
+\hypertarget{ecological-niche-modelling}{%
+\subsubsection{Ecological niche
+modelling}\label{ecological-niche-modelling}}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{rivers <-}\StringTok{ }\KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_rivers_lake_centerlines_scale_rank.shp"}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{st_crop}\NormalTok{(s_levant)}
-\end{Highlighting}
-\end{Shaded}
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
 
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersection assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: attribute variables are assumed to be spatially constant
-## throughout all geometries
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{water <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(}\KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_ocean.shp"}\NormalTok{)[,}\StringTok{"featurecla"}\NormalTok{],}
-               \KeywordTok{read_sf}\NormalTok{(}\StringTok{"data/raw_data/ne/ne_10m_lakes.shp"}\NormalTok{)[,}\StringTok{"featurecla"}\NormalTok{]) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{st_crop}\NormalTok{(sw_asia)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersection assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning: attribute variables are assumed to be spatially constant
-## throughout all geometries
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{basemap <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ land, }\DataTypeTok{fill =}\StringTok{"grey"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ palestine, }\DataTypeTok{fill =} \StringTok{"lightgrey"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
-\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ rivers, }\DataTypeTok{colour =} \StringTok{"white"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ water, }\DataTypeTok{fill =} \StringTok{"white"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{coord_sf}\NormalTok{(}\DataTypeTok{xlim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{xmin}\NormalTok{(s_levant), }\KeywordTok{xmax}\NormalTok{(s_levant)),}
-           \DataTypeTok{ylim =} \KeywordTok{c}\NormalTok{(}\KeywordTok{ymin}\NormalTok{(s_levant), }\KeywordTok{ymax}\NormalTok{(s_levant)))}
-\NormalTok{basemap}
-\end{Highlighting}
-\end{Shaded}
-
-\includegraphics{analysis_files/figure-latex/regions-1.pdf}
-
-\hypertarget{occurrence-data}{%
-\subsection{Occurrence data}\label{occurrence-data}}
+\hypertarget{occurrences}{%
+\subsubsection{Occurrences}\label{occurrences}}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Download or read data from GBIF}
-\NormalTok{gbif_path <-}\StringTok{ }\NormalTok{here}\OperatorTok{::}\KeywordTok{here}\NormalTok{(}\StringTok{"analysis/data/raw_data/gbif"}\NormalTok{)}
-\NormalTok{occ <-}\StringTok{ }\KeywordTok{bind_rows}\NormalTok{(}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Hordeum"}\NormalTok{, }\StringTok{"spontaneum*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Bolboschoenus"}\NormalTok{, }\StringTok{"glaucus*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),    }
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Bolboschoenus"}\NormalTok{, }\StringTok{"maritimus*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),    }
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Scirpus"}\NormalTok{, }\StringTok{"maritimus*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Triticum"}\NormalTok{, }\StringTok{"urartu*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Triticum"}\NormalTok{, }\StringTok{"dicoccoides*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Stipa"}\NormalTok{, }\StringTok{"*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Zilla"}\NormalTok{, }\StringTok{"spinosa*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Vitex"}\NormalTok{, }\StringTok{"*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path),}
-  \KeywordTok{fetch_gbif}\NormalTok{(}\StringTok{"Fraxinus"}\NormalTok{, }\StringTok{"*"}\NormalTok{, }\DataTypeTok{ext =}\NormalTok{ sw_asia, gbif_path)}
-\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_hordeum_spontaneum.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_bolboschoenus_glaucus.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_bolboschoenus_maritimus.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_scirpus_maritimus.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_triticum_urartu.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_triticum_dicoccoides.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_stipa.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_zilla_spinosa.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_vitex.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{verbatim}
-## Reading /home/joeroe/gdrive/Work/Lab/BadiaPaleoFloraSDM/analysis/data/raw_data/gbif/gbif_fraxinus.csv
-\end{verbatim}
-
-\begin{verbatim}
-## Parsed with column specification:
-## cols(
-##   gbifID = col_double(),
-##   genus = col_character(),
-##   species = col_character(),
-##   lon = col_double(),
-##   lat = col_double(),
-##   basisOfRecord = col_character()
-## )
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Code taxonomic groups for analysis}
-\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{taxon =} \KeywordTok{case_when}\NormalTok{(genus }\OperatorTok{==}\StringTok{ "Hordeum"} \OperatorTok{~}\StringTok{ "Hordeum spontaneum"}\NormalTok{,}
-\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Bolboschoenus"} \OperatorTok{~}\StringTok{ "Bolboschoenus sp. (=Scirpus maritimus)"}\NormalTok{,}
-\NormalTok{                           species }\OperatorTok{==}\StringTok{ "Triticum urartu"} \OperatorTok{~}\StringTok{ "Triticum urartu"}\NormalTok{,}
-\NormalTok{                           species }\OperatorTok{==}\StringTok{ "Triticum dicoccoides"} \OperatorTok{~}\StringTok{ "Triticum dicoccoides"}\NormalTok{,}
-\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Stipa"} \OperatorTok{~}\StringTok{ "Stipa sp."}\NormalTok{,}
-\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Zilla"} \OperatorTok{~}\StringTok{ "Zilla spinosa"}\NormalTok{,}
-\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Vitex"} \OperatorTok{~}\StringTok{ "Vitex sp."}\NormalTok{,}
-\NormalTok{                           genus }\OperatorTok{==}\StringTok{ "Fraxinus"} \OperatorTok{~}\StringTok{ "Fraxinus sp."}\NormalTok{)) }\OperatorTok{%T>%}
-\StringTok{  }\NormalTok{\{}\ControlFlowTok{if}\NormalTok{(}\KeywordTok{any}\NormalTok{(}\KeywordTok{is.na}\NormalTok{(.}\OperatorTok{$}\NormalTok{taxon))) }\KeywordTok{stop}\NormalTok{(}\StringTok{"Taxon contains missing values!"}\NormalTok{)\} ->}
-\StringTok{  }\NormalTok{occ}
-
-\CommentTok{# Remove duplicate records within analytic groups}
-\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{distinct}\NormalTok{(lon, lat, }\DataTypeTok{.keep_all =} \OtherTok{TRUE}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{ungroup}\NormalTok{() ->}
-\StringTok{  }\NormalTok{occ}
-
-\CommentTok{# Convert to sf}
-\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{coords =} \KeywordTok{c}\NormalTok{(}\StringTok{"lon"}\NormalTok{, }\StringTok{"lat"}\NormalTok{), }\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{, }\DataTypeTok{remove =} \OtherTok{FALSE}\NormalTok{) ->}
-\StringTok{  }\NormalTok{occ}
-\end{Highlighting}
-\end{Shaded}
-
 It's interesting to map the current distribution of these plants.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{ggplot}\NormalTok{(occ, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ lon, }\DataTypeTok{y =}\NormalTok{ lat)) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ occ, }\DataTypeTok{shape =} \DecValTok{1}\NormalTok{, }\DataTypeTok{size=}\FloatTok{0.5}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ water, }\DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"white"}\NormalTok{, }\DataTypeTok{colour =} \StringTok{"darkgrey"}\NormalTok{) }\OperatorTok{+}
-\StringTok{  }\KeywordTok{facet_wrap}\NormalTok{(}\OperatorTok{~}\NormalTok{taxon)}
-\end{Highlighting}
-\end{Shaded}
-
 \includegraphics{analysis_files/figure-latex/occurrence-map-1.pdf}
@@
 \hypertarget{pseudo-absences}{%
-\subsection{Pseudo-absences}\label{pseudo-absences}}
+\subsubsection{Pseudo-absences}\label{pseudo-absences}}
 
@@
 this. We follow the advice of Barbet-Massin et al.~(2012) for
-regression-based species distribution models and simply use a large
-(\textasciitilde10000) random sample of points.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{occ <-}\StringTok{ }\KeywordTok{as_tibble}\NormalTok{(occ)}
-\NormalTok{occ}\OperatorTok{$}\NormalTok{occurrence <-}\StringTok{ }\OtherTok{TRUE}
-\ControlFlowTok{for}\NormalTok{(taxon }\ControlFlowTok{in} \KeywordTok{unique}\NormalTok{(occ}\OperatorTok{$}\NormalTok{taxon)) \{}
-\NormalTok{  occ <-}\StringTok{ }\KeywordTok{bind_rows}\NormalTok{(occ, }\KeywordTok{generate_psabsences}\NormalTok{(palestine, }\DataTypeTok{taxon =}\NormalTok{ taxon,}
-                                            \DataTypeTok{occurrence =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{N =} \DecValTok{10000}\NormalTok{))}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{verbatim}
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-## although coordinates are longitude/latitude, st_intersects assumes that they are planar
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in bind_rows_(x, .id): Vectorizing 'sfc_POINT' elements may not
-## preserve their attributes
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{occ <-}\StringTok{ }\NormalTok{sf}\OperatorTok{::}\KeywordTok{st_as_sf}\NormalTok{(occ, }\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{)}
-\KeywordTok{rm}\NormalTok{(taxon)}
-\end{Highlighting}
-\end{Shaded}
+regression-based species distribution models and use a large
+(\textasciitilde10000) random sample of points, weighted equally against
+the presences in the regression.
 
 \hypertarget{environmental-data}{%
-\subsection{Environmental data}\label{environmental-data}}
+\subsubsection{Environmental data}\label{environmental-data}}
 
-\hypertarget{bioclimatic-variables}{%
-\subsubsection{Bioclimatic variables}\label{bioclimatic-variables}}
+\hypertarget{current-climate}{%
+\paragraph{Current climate}\label{current-climate}}
 
@@
 bio16 & Precipitation of wettest quarter &  \\ 
+bio17 & Precipitation of driest quarter &  \\ 
+bio18 & Precipitation of warmest quarter &  \\ 
+bio19 & Precipitation of coldest quarter &  \\ 
 \bottomrule
@@
 
-Here the climate data is from WorldClim 1.4
-(\url{http://www.worldclim.org}), a global dataset compiled from weather
-station records from the period 1960--1990. WorldClim has been widely
-used in ecological niche modelling. Other global climate datasets
-(e.g.~CHELSA, MERRAClim) and a newer version of WorldClim are now
-available, so it may be a good idea to investigate using these instead.
-The WorldClim data is large so again it is cached in the package's
-\texttt{derived\_data} directory.
+Here the current bioclimatic data is from CHELSA
+(\url{http://chelsa-climate.org/bioclim/}).
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# bioclim <- get_bioclim(s_levant, resolution = 0.5, cache_path = "data/derived_data")}
-\CommentTok{# Workaround for memory problems in Safawi}
-\KeywordTok{load}\NormalTok{(}\StringTok{"data/derived_data/bioclim.RData"}\NormalTok{)}
-\KeywordTok{plot}\NormalTok{(bioclim)}
-\end{Highlighting}
-\end{Shaded}
+\hypertarget{palaeoclimate}{%
+\paragraph{Palaeoclimate}\label{palaeoclimate}}
 
-\includegraphics{analysis_files/figure-latex/bioclim-data-1.pdf}
-
-\hypertarget{soil-variables}{%
-\subsubsection{Soil variables}\label{soil-variables}}
+\hypertarget{soil}{%
+\subsubsection{Soil}\label{soil}}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{soilgrids <-}\StringTok{ }\NormalTok{raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(}\KeywordTok{list.files}\NormalTok{(}\StringTok{"data/raw_data/soilgrids"}\NormalTok{, }\StringTok{".*tif"}\NormalTok{, }\DataTypeTok{full.names =} \OtherTok{TRUE}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
 \hypertarget{terrain}{%
@@
 There are also relevent aspects of the terrain that can be easily
-derived from a digital elevation model. We'll use NASA-JPL's SRTM v3 1
-arc-second resolution DEM.
+derived from a digital elevation model. We'll use NASA-JPL's SRTM v3 3
+arc-second resolution DEM, smoothed with Sun's denoising algorithm
+{[}TODO: CITE{]}. Slope was calculated from this surface using GDAL
+3.0.0 and the topographic wetness index (TWI) using the r.topidx module
+of GRASS 7.8. The DEM was reprojected to a conformal projection (Jordan
+Transverse Mercator, EPSG:19995) for these calculations, using a
+bilinear interpolation, then back to geographic coordinates.
 
-\hypertarget{selection-of-predictor-variables}{%
-\subsubsection{Selection of predictor
-variables}\label{selection-of-predictor-variables}}
+\hypertarget{model}{%
+\subsection{Model}\label{model}}
+
+\hypertarget{selection-of-predictors}{%
+\subsubsection{Selection of predictors}\label{selection-of-predictors}}
 
@@
 \toprule
-variable & units & source & code \\ 
+Variable & Units & Dataset & Source code \\ 
 \midrule
@@
 \midrule
-Minimum temperature of coldest month & °C ×10 & WorldClim 1.4 & bio6 \\ 
-Total annual precipitation & mm & WorldClim 1.4 & bio12 \\ 
-Precipitation of wettest month & mm & WorldClim 1.4 & bio13 \\ 
-Precipitation of driest month & mm & WorldClim 1.4 & bio14 \\ 
-Precipitation seasonality & σ/μ & WorldClim 1.4 & bio15 \\ 
+Minimum temperature of coldest month & °C ×10 & PaleoClim 1.0 & bio\_6 \\ 
+Total annual precipitation & mm & PaleoClim 1.0 & bio\_12 \\ 
+Precipitation of wettest month & mm & PaleoClim 1.0 & bio\_13 \\ 
+Precipitation of driest month & mm & PaleoClim 1.0 & bio\_14 \\ 
+Precipitation seasonality & σ/μ & PaleoClim 1.0 & bio\_15 \\ 
 \midrule
@@
 Depth to bedrock & cm (up to 200 cm) & SoilGrids250m & BDRICM\_M\_250m \\ 
+\midrule
+\multicolumn{1}{l}{Terrain} \\ 
+\midrule
+Elevation & m & SRTM DEM v3 & srtm1v3s\_slevant \\ 
+Slope & ° & SRTM DEM v3 & srtm1v3s\_slevant\_slope \\ 
+Topographic Wetness Index & ln(upslope area / tan(slope)) & SRTM DEM v3 & srtm1v3s\_slevant\_twi \\ 
 \bottomrule
@@
 
-\hypertarget{combining-predictor-variables}{%
-\subsubsection{Combining predictor
-variables}\label{combining-predictor-variables}}
-
 These datasets come with different resolutions. To use them all in the
-same analysis, we resample them all to the finest dataset (WorldClim, 30
-arc-seconds).
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{pred}\OperatorTok{$}\NormalTok{raster <-}\StringTok{ }\KeywordTok{lapply}\NormalTok{(pred}\OperatorTok{$}\NormalTok{raster, raster}\OperatorTok{::}\NormalTok{resample, }\DataTypeTok{y =}\NormalTok{ bioclim, }\DataTypeTok{method =} \StringTok{"ngb"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{model-construction}{%
-\subsection{Model construction}\label{model-construction}}
+same analysis, we resample them all to the finest dataset (the SRTM
+terrain variables, at 3 arc-second resolution).
 
@@
 occurrence and pseudo-absence location:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# }\AlertTok{TODO}\CommentTok{: refactor to use pred}
-\NormalTok{bioclim }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{extract}\NormalTok{(}\KeywordTok{as_Spatial}\NormalTok{(occ)) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\NormalTok{\{}\KeywordTok{cbind}\NormalTok{(occ, .)\} }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(occ)) ->}
-\StringTok{  }\NormalTok{occ}
-
-\NormalTok{soilgrids }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{extract}\NormalTok{(}\KeywordTok{as_Spatial}\NormalTok{(occ)) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\NormalTok{\{}\KeywordTok{cbind}\NormalTok{(occ, .)\} }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(occ)) ->}
-\StringTok{  }\NormalTok{occ}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# Summarise occurrences on a grid}
-\NormalTok{palestine }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{as_Spatial}\NormalTok{() }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{rasterize}\NormalTok{(bioclim) ->}
-\StringTok{  }\NormalTok{palestine_raster}
-
-\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{filter}\NormalTok{(occurrence }\OperatorTok{==}\StringTok{ }\OtherTok{TRUE}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{do}\NormalTok{(}\DataTypeTok{occ_grid =} \KeywordTok{raster_burn}\NormalTok{(., palestine_raster)) ->}
-\StringTok{  }\NormalTok{occ_grids}
-
-\CommentTok{# Fit poly(1:3) to each taxon/predictor combination}
-\KeywordTok{expand.grid}\NormalTok{(}\DataTypeTok{taxon =} \KeywordTok{unique}\NormalTok{(occ}\OperatorTok{$}\NormalTok{taxon),}
-            \DataTypeTok{predictor =}\NormalTok{ pred}\OperatorTok{$}\NormalTok{code,}
-            \DataTypeTok{polyn =} \DecValTok{1}\OperatorTok{:}\DecValTok{3}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon, predictor, polyn) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{do}\NormalTok{(}
-    \CommentTok{# }\AlertTok{TODO}\CommentTok{: It works but I hate it.}
-    \DataTypeTok{fit =} \KeywordTok{model_response}\NormalTok{(}\KeywordTok{unlist}\NormalTok{(occ_grids[occ_grids}\OperatorTok{$}\NormalTok{taxon }\OperatorTok{==}\StringTok{ }\NormalTok{.}\OperatorTok{$}\NormalTok{taxon, }\StringTok{"occ_grid"}\NormalTok{])[[}\DecValTok{1}\NormalTok{]],}
-                         \KeywordTok{unlist}\NormalTok{(pred[pred}\OperatorTok{$}\NormalTok{code }\OperatorTok{==}\StringTok{ }\NormalTok{.}\OperatorTok{$}\NormalTok{predictor, }\StringTok{"raster"}\NormalTok{])[[}\DecValTok{1}\NormalTok{]],}
-\NormalTok{                         .}\OperatorTok{$}\NormalTok{polyn)}
-\NormalTok{  ) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{glance}\NormalTok{(fit) ->}\StringTok{ }
-\StringTok{  }\NormalTok{response_glm}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-
-## Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred
-\end{verbatim}
-
 The optimal model is the one with the lowest AIC.
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{response_glm }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon, predictor) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{optimum =}\NormalTok{ polyn }\OperatorTok{==}\StringTok{ }\KeywordTok{which.min}\NormalTok{(AIC)) ->}
-\StringTok{  }\NormalTok{response_glm}
-
-\CommentTok{# }\AlertTok{TODO}
-\CommentTok{# Re-fit and plot optimal }
-\CommentTok{# model_response(occ_grids$occ_grid[[1]], pred$raster[[1]], 2) %>% predict(data.frame(pred = -72:109, type = "response")) %>% plot()}
-\CommentTok{# Need to dplyrise this – see broom::augment()}
-\CommentTok{# response_glm %>% }
-\CommentTok{#   filter(optimum == TRUE) %>% }
-\CommentTok{#   slice(1:3) %>% }
-\CommentTok{#   do(}
-\CommentTok{#     fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],}
-\CommentTok{#                          unlist(pred[pred$code == .$predictor,"raster"])[[1]],}
-\CommentTok{#                          .$polyn)}
-\CommentTok{#   ) }
-\end{Highlighting}
-\end{Shaded}
-
-\hypertarget{model-fitting}{%
-\subsubsection{Model fitting}\label{model-fitting}}
+\hypertarget{logistic-regression}{%
+\subsubsection{Logistic regression}\label{logistic-regression}}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{# }\AlertTok{TODO}\CommentTok{: use optimal polynomial order}
-\NormalTok{occ }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{group_by}\NormalTok{(taxon) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{remove_missing}\NormalTok{(}\DataTypeTok{vars =} \KeywordTok{c}\NormalTok{(}\StringTok{"bio1"}\NormalTok{, }\StringTok{"bio2"}\NormalTok{, }\StringTok{"bio3"}\NormalTok{, }\StringTok{"bio4"}\NormalTok{, }\StringTok{"bio5"}\NormalTok{, }\StringTok{"bio6"}\NormalTok{, }
-                          \StringTok{"bio7"}\NormalTok{, }\StringTok{"bio8"}\NormalTok{, }\StringTok{"bio9"}\NormalTok{, }\StringTok{"bio10"}\NormalTok{, }\StringTok{"bio11"}\NormalTok{, }\StringTok{"bio12"}\NormalTok{, }
-                          \StringTok{"bio13"}\NormalTok{, }\StringTok{"bio14"}\NormalTok{, }\StringTok{"bio15"}\NormalTok{, }\StringTok{"bio16"}\NormalTok{, }\StringTok{"bio17"}\NormalTok{, }\StringTok{"bio18"}\NormalTok{, }
-                          \StringTok{"bio19"}\NormalTok{, }\StringTok{"BDRICM_M_250m"}\NormalTok{, }\StringTok{"CRFVOL_M_sl2_250m"}\NormalTok{, }
-                          \StringTok{"ORCDRC_M_sl2_250m"}\NormalTok{, }\StringTok{"PHIHOX_M_sl2_250m"}\NormalTok{, }
-                          \StringTok{"SNDPPT_M_sl2_250m"}\NormalTok{)) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{do}\NormalTok{(}\DataTypeTok{model =} \KeywordTok{glm}\NormalTok{(occurrence }\OperatorTok{~}\StringTok{ }
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio6, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio12, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio13, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
-\StringTok{                   }\CommentTok{# poly(bio14, 2) +}
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(bio15, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(BDRICM_M_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(CRFVOL_M_sl2_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(ORCDRC_M_sl2_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(PHIHOX_M_sl2_250m, }\DecValTok{2}\NormalTok{) }\OperatorTok{+}\StringTok{ }
-\StringTok{                   }\KeywordTok{poly}\NormalTok{(SNDPPT_M_sl2_250m, }\DecValTok{2}\NormalTok{),}
-                 \DataTypeTok{data =}\NormalTok{ .)) ->}
-\StringTok{  }\NormalTok{models}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning: Removed 3337 rows containing missing values.
-\end{verbatim}
-
-\hypertarget{predicting-occurrences}{%
-\subsection{Predicting occurrences}\label{predicting-occurrences}}
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{models }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{do}\NormalTok{(}\DataTypeTok{taxon =}\NormalTok{ .}\OperatorTok{$}\NormalTok{taxon,}
-     \DataTypeTok{model =}\NormalTok{ .}\OperatorTok{$}\NormalTok{model,}
-     \DataTypeTok{pred_current =}\NormalTok{ raster}\OperatorTok{::}\KeywordTok{predict}\NormalTok{(}\DataTypeTok{object =}\NormalTok{ raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(pred}\OperatorTok{$}\NormalTok{raster),}
-                                    \DataTypeTok{model =}\NormalTok{ .}\OperatorTok{$}\NormalTok{model)) ->}
-\StringTok{  }\NormalTok{models}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(models}\OperatorTok{$}\NormalTok{pred_current) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{set_names}\NormalTok{(models}\OperatorTok{$}\NormalTok{taxon) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{plot}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
 \includegraphics{analysis_files/figure-latex/pred_current_plot-1.pdf}
 
-\hypertarget{binning}{%
-\subsubsection{Binning}\label{binning}}
-
-To more easily interpret these predictions, we could convert the raw
-probability scores into confidence intervals, based on the proportion of
-known occurrences it successfully predicts.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{raster}\OperatorTok{::}\KeywordTok{stack}\NormalTok{(models}\OperatorTok{$}\NormalTok{pred_current) }\OperatorTok{%>%}
-\StringTok{  }\KeywordTok{set_names}\NormalTok{(}\KeywordTok{paste0}\NormalTok{(}\StringTok{"pred_current_"}\NormalTok{, models}\OperatorTok{$}\NormalTok{taxon)) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\NormalTok{raster}\OperatorTok{::}\KeywordTok{extract}\NormalTok{(}\KeywordTok{as_Spatial}\NormalTok{(occ)) }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\NormalTok{\{}\KeywordTok{cbind}\NormalTok{(occ, .)\} }\OperatorTok{%>%}\StringTok{ }
-\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(occ)) ->}
-\StringTok{  }\NormalTok{occ2}
-\end{Highlighting}
-\end{Shaded}
+\hypertarget{discussion}{%
+\subsection{Discussion}\label{discussion}}
 

### 2023-04-26 R/data.R

@@ -117,6 +117,42 @@ get_bioclim <- function(extent, resolution = 0.5, cache_path = getwd(), quiet = 
   return(bio)
 }
 
+# PALEOCLIM DATA ----------------------------------------------------------
+
+#' Fetch PaleoClim data
+#'
+#' An interface to [rpaleoclim::paleoclim()] with caching.
+#'
+#' @param period      Character. Time period to fetch.
+#' @param resolution  Character. Resolution to fetch.
+#' @param ext         `Extent` object. Region to fetch.
+#' @param cache_file  Character. Path to cache data in .grd format.
+#' @param redownload  Logical. If `TRUE`, cached data will be ignored. Default: `FALSE`.
+#'
+#' @return
+#'
+#' `RasterStack` object.
+#'
+#' @details
+#'
+#' The raster is always read or re-read from the cached file to ensure consistency.
+#' See [rpaleoview::paleoview()] for more details.
+#'
+#' @export
+#'
+#' @examples
+fetch_paleoclim <- function(period, resolution, ext = NULL, cache_file, redownload = FALSE) {
+  if (!file.exists(cache_file) | redownload) {
+    message("Downloading ", period, ", ", resolution, ", from PaleoClim...")
+    rpaleoclim::paleoclim(period, resolution, region = ext) %>%
+      raster::writeRaster(cache_file, format = "raster")
+  }
+
+  message("Reading ", cache_file)
+  raster::stack(cache_file) %>%
+    return()
+}
+
 #' Merge Raster* objects (if necessary)
 #'
 #' Utility function for merging a list of Raster* objects. If passed a list of

### 2023-04-26 R/enm.R

@@ -39,3 +39,13 @@ model_response <- function(occ, pred, polyn) {
 
   return(mod)
 }
+
+#' Convert log-odds to probability
+#'
+#' @param x Numeric. Vector of log-odds, e.g. output of a logistic regression.
+#'
+#' @return Vector of normalised probabilities.
+#' @export
+logit_to_p <- function(x) {
+  return(exp(x) / (1+exp(x)))
+}

### 2023-07-05 analysis/analysis.Rmd
Fix unclosed chunk
Interim commit with rewrite

--- 2019-07-16 analysis/analysis.Rmd
@@
 ---
-title: Ecological niche models of plant species exploited by Epipalaeolithic foragers
-  in the *Badia* (c. 14500 BP)
-output:
-  html_document:
-    df_print: paged
-  pdf_document:
-    latex_engine: xelatex
+title: "Palaeodistribution of crop progenitors and wild foods in the Southern Levant"
+author: "Joe Roe"
+format: html
 ---
 
-## Dependencies
+```{r dependencies}
+library("dplyr")
+library("ggplot2")
+library("ggspatial")
+library("parsnip")
+library("purrr")
+library("ranger")
+library("rpaleoclim")
+library("sf")
+library("tibble")
+library("tidyr")
 
-```{r packages, warning=FALSE, message=FALSE}
-library("tidyverse")
-library("magrittr")
-library("sf")
-library("raster")
-library("broom")
-library("dismo")
-library("rnaturalearth")
-library("gt")
-library("BadiaPaleoFloraENM")
+library("BadiaPaleoFloraENM") # this package, install with devtools::install()
+```
+
+```{r definitions}
+latlong <- 4326
+utm37n <- 32637
+
+w_asia <- st_bbox(c(xmin = 30, xmax = 50, ymin = 25, ymax = 40), crs = 4326)
+s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
 ```
@@
 
-Species distribution models (SDM) or ecological niche models (ENM) use data on the observed occurrences of a taxon to construct a statistical model of its ecological niche.
-These models are generally then used to predict the distribution of the species in another context, e.g.
-a region where empirical occurrence data is not available, or under past or future climate regimes.
-We will attempt to model the ecological niches of six key plant taxa for prehistoric foragers.
-We will use occurrence data from Israel and Palestine, which are well-studied, to predict the distribution of these taxa across the rest of the Southern Levant.
+## Background
 
-First, we define our regions of interest:
+## Methods and materials
 
-```{r regions, warning=FALSE, message=FALSE}
-sw_asia <- extent(c(30, 50, 25, 40))
-s_levant <- extent(c(34, 39, 29, 34))
-palestine <- ne_countries(scale = 10, country = c("Israel", "Palestine"),
-             returnclass = "sf") %>% st_union() %>% st_sf()
+```{r data-basemap}
+# Turn off s2 while we deal with Natural Earth's dodgy geometries
+use_s2 <- sf_use_s2(FALSE)
 
-land <- read_sf("data/raw_data/ne/ne_10m_land.shp")[,"featurecla"] %>%
-  st_crop(sw_asia)
-rivers <- read_sf("data/raw_data/ne/ne_10m_rivers_lake_centerlines_scale_rank.shp") %>% 
-  st_crop(s_levant)
-water <- rbind(read_sf("data/raw_data/ne/ne_10m_ocean.shp")[,"featurecla"],
-               read_sf("data/raw_data/ne/ne_10m_lakes.shp")[,"featurecla"]) %>% 
-  st_crop(sw_asia)
+ne_countries <- read_sf(raw_data("ne"), "ne_10m_admin_0_countries")
+
+ne_lakes <- read_sf(raw_data("ne"), "ne_10m_lakes") |> 
+  st_geometry() |>
+  st_crop(buffer_bbox(w_asia, 10)) |>
+  st_union()
+
+ne_ocean <- read_sf(raw_data("ne"), "ne_10m_ocean") |>
+  st_geometry() |>
+  st_crop(buffer_bbox(w_asia, 10)) |>
+  st_union()
+
+ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
+  st_geometry() |>
+  st_crop(buffer_bbox(w_asia, 10)) |>
+  st_difference(ne_ocean) |>
+  st_difference(ne_lakes) |>
+  st_union()
+
+s_levant_land <- st_intersection(st_as_sfc(s_levant), ne_land)
+
+# Restore previous setting of s2
+sf_use_s2(use_s2)
 ```
 
-```{r regions-map, echo=FALSE, results='hide'}
-basemap <- ggplot() + 
-  geom_sf(data = land, fill ="grey") +
-  geom_sf(data = palestine, fill = "lightgrey") + 
-  geom_sf(data = rivers, colour = "white") +
-  geom_sf(data = water, fill = "white") +
-  coord_sf(xlim = c(xmin(s_levant), xmax(s_levant)),
-           ylim = c(ymin(s_levant), ymax(s_levant)))
-basemap
-```
+### Occurrence data
 
-## Occurrence data
+Occurrence data was obtained from GBIF.
 
-First we download species occurrence data from GBIF (<https://gbif.org>).
-This may change, so we cache the result used in the published analysis in `derived_data` and only redownload if the cached version is missing.
-
-```{r occurrence-data, warning=FALSE, message=FALSE}
+```{r data-occ}
 # Download or read data from GBIF
-gbif_path <- here::here("analysis/data/raw_data/gbif")
+# TODO: T. boeoticum ? (equate with T. urartu?)
+gbif_path <- here::here("analysis", "data", "derived_data", "gbif")
 occ <- bind_rows(
-  fetch_gbif("Hordeum", "spontaneum*", ext = sw_asia, gbif_path),
-  fetch_gbif("Bolboschoenus", "glaucus*", ext = sw_asia, gbif_path),    
-  fetch_gbif("Bolboschoenus", "maritimus*", ext = sw_asia, gbif_path),    
-  fetch_gbif("Scirpus", "maritimus*", ext = sw_asia, gbif_path),
-  fetch_gbif("Triticum", "urartu*", ext = sw_asia, gbif_path),
-  fetch_gbif("Triticum", "dicoccoides*", ext = sw_asia, gbif_path),
-  fetch_gbif("Stipa", "*", ext = sw_asia, gbif_path),
-  fetch_gbif("Zilla", "spinosa*", ext = sw_asia, gbif_path),
-  fetch_gbif("Vitex", "*", ext = sw_asia, gbif_path),
-  fetch_gbif("Fraxinus", "*", ext = sw_asia, gbif_path)
+  fetch_gbif("Hordeum", "spontaneum*", ext = w_asia, gbif_path),
+  #fetch_gbif("Bolboschoenus", "glaucus*", ext = w_asia, gbif_path),    
+  #fetch_gbif("Bolboschoenus", "maritimus*", ext = w_asia, gbif_path),    
+  #fetch_gbif("Scirpus", "maritimus*", ext = w_asia, gbif_path),
+  #fetch_gbif("Triticum", "urartu*", ext = w_asia, gbif_path),
+  #fetch_gbif("Triticum", "dicoccoides*", ext = w_asia, gbif_path),
+  #fetch_gbif("Stipa", "*", ext = w_asia, gbif_path),
+  #fetch_gbif("Zilla", "spinosa*", ext = w_asia, gbif_path),
+  #fetch_gbif("Vitex", "*", ext = w_asia, gbif_path),
+  #fetch_gbif("Fraxinus", "*", ext = w_asia, gbif_path)
 )
@@
 # Code taxonomic groups for analysis
-occ %>% 
+occ |>
   mutate(taxon = case_when(genus == "Hordeum" ~ "Hordeum spontaneum",
-                           genus == "Bolboschoenus" ~ "Bolboschoenus sp. (=Scirpus maritimus)",
+                           genus == "Bolboschoenus" ~ "Bolboschoenus sp.",
                            species == "Triticum urartu" ~ "Triticum urartu",
@@
                            genus == "Vitex" ~ "Vitex sp.",
-                           genus == "Fraxinus" ~ "Fraxinus sp.")) %T>%
-  {if(any(is.na(.$taxon))) stop("Taxon contains missing values!")} ->
+                           genus == "Fraxinus" ~ "Fraxinus sp.")) ->
   occ
 
+# TODO: convert to a test
+if(any(is.na(occ$taxon))) stop("Taxon contains missing values!")
+
 # Remove duplicate records within analytic groups
-occ %>% 
-  group_by(taxon) %>% 
-  distinct(lon, lat, .keep_all = TRUE) %>% 
+occ |>
+  group_by(taxon) |>
+  distinct(lon, lat, .keep_all = TRUE) |>
   ungroup() ->
@@
 # Convert to sf
-occ %>% 
+occ |>
   st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) ->
@@
 
-It's interesting to map the current distribution of these plants.
-```{r occurrence-map, echo=FALSE, results='hide'}
-ggplot(occ, aes(x = lon, y = lat)) +
-  geom_sf(data = occ, shape = 1, size=0.5) +
-  geom_sf(data = water, inherit.aes = FALSE, fill = "white", colour = "darkgrey") +
-  facet_wrap(~taxon)
+```{r plot-occ-distribution}
+ggplot() + 
+  layer_spatial(ne_land, fill = "#ffffff") +
+  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
+  layer_spatial(occ, alpha = 0.1, symbol = 16) +
+  # layer_spatial(ne_countries, fill = NA) +
+  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
+           ylim = w_asia[c("ymin", "ymax")],
+           crs = utm37n,
+           default_crs = latlong)
 ```
 
-## Pseudo-absences
-
+<!-- TODO: following @ValaviEtAl2022, more accurate to call pseudo-absences
+  "background samples" -->
 Occurrence data only tells us where a species is present;
@@
 There are several ways to do this. 
-We follow the advice of Barbet-Massin et al. (2012) for regression-based species distribution models and simply use a large (~10000) random sample of points.
+We follow the advice of Barbet-Massin et al.
+(2012) for regression-based species distribution models and use a large (~10000) random sample of points, weighted equally against the presences in the regression.
+@ValaviEtAl2022 recommends using a very large background sample for random forest models.
 
-```{r psabsences, warning=FALSE, message=FALSE}
-occ <- as_tibble(occ)
-occ$occurrence <- TRUE
-for(taxon in unique(occ$taxon)) {
-  occ <- bind_rows(occ, generate_psabsences(palestine, taxon = taxon,
-                                            occurrence = FALSE, N = 10000))
-}
-occ <- sf::st_as_sf(occ, crs = 4326)
-rm(taxon)
+```{r psabsences, include=FALSE}
+n_background <- 1000
+
+occ |>
+  mutate(present = TRUE) |>
+  bind_rows(background_sample(s_levant_land, 
+                              n_background,
+                              coord_x = "lon",
+                              coord_y = "lat", 
+                              taxon = "Hordeum spontaneum", 
+                              present = FALSE)) ->
+  occ
+
+# Weights
+#occ %>% 
+#  group_by(taxon, occurrence) %>% 
+#  mutate(weight = 1 + (n_psab / n() - 1)) ->
+#  occ
 ```
 
-## Environmental data
+### Climate data
 
-### Bioclimatic variables
+```{r data-paleoclim}
+tibble(period = c("lgm", "hs1", "ba", "yds", "eh", "cur")) |>
+  mutate(raster = map(period, paleoclim, "2_5m", 
+                      region = buffer_bbox(w_asia, 10),
+                      cache_path = here::here(".paleoclim_cache"))) -> # TODO: revert to default cache
+  climate
 
-To construct the model, we'll use a standard set of "bioclimatic" predicator variables.
-These sixteen variables are derived from monthly precipitation and temperate data:
+training_climate <- filter(climate, period == "cur")$raster[[1]]
 
-```{r bioclim-table, echo=FALSE}
-tibble(biovar = paste0("bio", 1:16),
-       definition = c("Mean annual temperature",
-                      "Mean diurnal range",
-                      "Isothermality",
-                      "Temperature seasonality",
-                      "Maximum temperature of warmest month",
-                      "Minimum temperature of coldest month",
-                      "Annual temperature range",
-                      "Mean temperature of wettest quarter",
-                      "Mean temperature of driest quarter",
-                      "Mean temperature of warmest quarter",
-                      "Mean temperature of coldest quarter",
-                      "Total annual precipitation",
-                      "Precipitation of wettest month",
-                      "Precipitation of driest month",
-                      "Precipitation seasonality",
-                      "Precipitation of wettest quarter"),
-       formula = c("",
-                   "",
-                   "(bio2 / bio7) * 100",
-                   "sd(temp) * 100",
-                   "",
-                   "",
-                   "bio5 - bio6",
-                   "",
-                   "",
-                   "",
-                   "",
-                   "",
-                   "",
-                   "",
-                   "",
-                   "")
-) %>%
-  gt()
-```
-
-Here the climate data is from WorldClim 1.4 (<http://www.worldclim.org>), a global dataset compiled from weather station records from the period 1960–1990.
-WorldClim has been widely used in ecological niche modelling.
-Other global climate datasets (e.g. CHELSA, MERRAClim) and a newer version of WorldClim are now available, so it may be a good idea to investigate using these instead.
-The WorldClim data is large so again it is cached in the package's `derived_data` directory.
-
-```{r bioclim-data, echo=FALSE}
-# bioclim <- get_bioclim(s_levant, resolution = 0.5, cache_path = "data/derived_data")
-# Workaround for memory problems in Safawi
-load("data/derived_data/bioclim.RData")
-plot(bioclim)
-```
-
-### Soil variables
-
-Velazco et al. (2017) have shown that edaphic data, recently available on a global scale, is usefully combined with bioclimatic data in modelling plant niches.
-
-We use data from SoilGrids, a global map of soil characteristics derived from remote sensing classification at 1km resolution.
-
-```{r soil-data}
-soilgrids <- raster::stack(list.files("data/raw_data/soilgrids", ".*tif", full.names = TRUE))
-plot(soilgrids)
-```
-
-### Terrain
-
-There are also relevent aspects of the terrain that can be easily derived from a digital elevation model.
-We'll use NASA-JPL's SRTM v3 1 arc-second resolution DEM.
-
-### Selection of predictor variables
-
-We reviewed existing floras to identify specific variables from these datasets that are relevant to the niche of our target taxa. The results are summarised below.
-
-```{r review-table, echo=FALSE, warning=FALSE, message=FALSE}
-read_csv("data/raw_data/flora_review.csv") %>% 
-  group_by(Taxon) %>% 
-  gt()
-```
-
-Based on these results, we made the following initial selection of predictor variables to build the models.
-
-```{r predictor-table, echo=FALSE}
-pred <- bind_rows(
-  tibble(group = rep("Climate", 5),
-         variable = c("Minimum temperature of coldest month",
-                      "Total annual precipitation", "Precipitation of wettest month",
-                      "Precipitation of driest month", "Precipitation seasonality"),
-         units = c("°C ×10", "mm", "mm", "mm", "σ/μ"),
-         source = rep("WorldClim 1.4", 5),
-         code = c("bio6", "bio12", "bio13", "bio14", "bio15"),
-         raster = as.list(raster::subset(bioclim,
-                                         c("bio6", "bio12", "bio13", "bio14", "bio15")))),
-  tibble(group = rep("Soil", 5),
-         variable = c("Organic carbon content", "Coarse fragment content",
-                      "Sand content", "Acidity", "Depth to bedrock"),
-         units = c("g per kg fine earth", "% volume", "% mass", "pH ×10 in H2O",
-                   "cm (up to 200 cm)"),
-         source = rep("SoilGrids250m", 5),
-         code = c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
-                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"),
-         raster = as.list(raster::subset(soilgrids,
-                                         c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
-                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"))))
-  # tibble(group = rep("Terrain", 3),
-  #        variable = c("Elevation", "Slope", "Topographic Wetness Index"),
-  #        units = c("m", "°", "ln(upslope area / tan(slope))"),
-  #        source = rep("SRTM DEM v3", 3),
-  #        code = c("srtm1v3_slevant", "srtm1v3_slevant_slope",
-  #                         "srtm1v3_slevant_twi"))
-)
-
-pred %>%
-  dplyr::select(-raster) %>% 
-  group_by(group) %>% 
-  gt()
-```
-
-One aspect that we cannot easily incorporate into the models is the interaction between these taxa and other plant and animal species: cereal grasses grow best when they have no competition from perennial plants, *Fraxinus* often grows in scrub or forest, and so on.
-
-We also lack data for soil salinity so far.
-
-### Combining predictor variables
-
-These datasets come with different resolutions.
-To use them all in the same analysis, we resample them all to the finest dataset (WorldClim, 30 arc-seconds).
-
-```{r predictor-resample}
-pred$raster <- lapply(pred$raster, raster::resample, y = bioclim, method = "ngb")
-```
-
-## Model construction
-
-We will model the ecological niche of each species using a generalised linear regression.
-
-First we extract the values of the predictor variables at each occurrence and pseudo-absence location:
-
-```{r predictor-extraction}
-# TODO: refactor to use pred
-bioclim %>% 
-  raster::extract(as_Spatial(occ)) %>% 
-  {cbind(occ, .)} %>% 
-  st_as_sf(crs = st_crs(occ)) ->
-  occ
-
-soilgrids %>% 
-  raster::extract(as_Spatial(occ)) %>% 
-  {cbind(occ, .)} %>% 
-  st_as_sf(crs = st_crs(occ)) ->
+occ |>
+  mutate(bio = terra::extract(training_climate, data.frame(lon, lat), 
+                              method = "bilinear", ID = FALSE)) |>
+  unpack(bio) |>
+  ungroup() ->
   occ
@@
 
-### Species response curves
+### Other predictor data
 
-We can use species response curves to visualise the distribution of known occurrences in relation to these environmental gradients.
-This will also help us parameterise the ecological niche models.
+Soil, terrain (incl. TWI, slope?)
 
-Species response curves are constructed with a simple logistic regression, fitting the probability of occurrence at a given location to 
-The use of a linear regression assumes a unimodal response.
-We fit first, second, and third order polynomial models to each species/predictor combination and calculate the Akaike Information Criterion (AIC).
+### Random Forest
 
-```{r response-curves-fit, warning=FALSE, message=FALSE}
-# Summarise occurrences on a grid
-palestine %>% 
-  as_Spatial() %>% 
-  raster::rasterize(bioclim) ->
-  palestine_raster
-
-occ %>% 
-  filter(occurrence == TRUE) %>% 
-  group_by(taxon) %>% 
-  do(occ_grid = raster_burn(., palestine_raster)) ->
-  occ_grids
-
-# Fit poly(1:3) to each taxon/predictor combination
-expand.grid(taxon = unique(occ$taxon),
-            predictor = pred$code,
-            polyn = 1:3) %>% 
-  group_by(taxon, predictor, polyn) %>% 
-  do(
-    # TODO: It works but I hate it.
-    fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
-                         unlist(pred[pred$code == .$predictor, "raster"])[[1]],
-                         .$polyn)
-  ) %>% 
-  glance(fit) -> 
-  response_glm
+```{r model}
+rand_forest(mode = "regression") |>
+  fit(as.numeric(present) ~ ., data = select(occ, present, starts_with("bio_"))) ->
+  rf_fit
 ```
 
-```{r response-curves-table, echo=FALSE}
-# response_glm %>% 
-#   spread(polyn, aic) %>% 
-#   rename(aic_poly1 = `1`, aic_poly2 = `2`, aic_poly3 = `3`) %>% 
-#   gt(rowname_col = "predictor", groupname_col = "taxon") %>% 
-#   tab_stubhead_label("predictor")
+## Results
+
+### Modelled ecological niches
+
+* On current data
+* Sensitivity to climate fluctuations
+
+### Hindcasting
+
+```{r predict}
+predict(rf_fit, as.data.frame(filter(climate, period == "cur")$raster), type = "raw")
 ```
 
-The optimal model is the one with the lowest AIC.
+### Verification
 
-```{r response-curves-optimal}
-response_glm %>%
-  group_by(taxon, predictor) %>%
-  mutate(optimum = polyn == which.min(AIC)) ->
-  response_glm
-
-# TODO
-# Re-fit and plot optimal 
-# model_response(occ_grids$occ_grid[[1]], pred$raster[[1]], 2) %>% predict(data.frame(pred = -72:109, type = "response")) %>% plot()
-# Need to dplyrise this – see broom::augment()
-# response_glm %>% 
-#   filter(optimum == TRUE) %>% 
-#   slice(1:3) %>% 
-#   do(
-#     fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
-#                          unlist(pred[pred$code == .$predictor,"raster"])[[1]],
-#                          .$polyn)
-#   ) 
-```
-
-### Model fitting
-
-Now we can fit regression models combining all the predictor variables.
-
-```{r model_fit, warning=FALSE, message=FALSE}
-# TODO: use optimal polynomial order
-occ %>% 
-  group_by(taxon) %>% 
-  remove_missing(vars = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", 
-                          "bio7", "bio8", "bio9", "bio10", "bio11", "bio12", 
-                          "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", 
-                          "bio19", "BDRICM_M_250m", "CRFVOL_M_sl2_250m", 
-                          "ORCDRC_M_sl2_250m", "PHIHOX_M_sl2_250m", 
-                          "SNDPPT_M_sl2_250m")) %>% 
-  do(model = glm(occurrence ~ 
-                   poly(bio6, 2) +
-                   poly(bio12, 2) +
-                   poly(bio13, 2) +
-                   # poly(bio14, 2) +
-                   poly(bio15, 2) +
-                   poly(BDRICM_M_250m, 2) +
-                   poly(CRFVOL_M_sl2_250m, 2) +
-                   poly(ORCDRC_M_sl2_250m, 2) + 
-                   poly(PHIHOX_M_sl2_250m, 2) + 
-                   poly(SNDPPT_M_sl2_250m, 2),
-                 data = .)) ->
-  models
-
-```
-
-## Predicting occurrences
-### Current conditions
-
-We can now use the fitted models to predict the occurrence of the different taxa according to the predictor variables.
-Initially, we do this based on current conditions.
-This allows us to extrapolate from regions with good occurrence data (i.e. Israel/Palestine) to those without.
-It also hints at what the distribution of the plant might be under "ideal conditions", without agriculture, grazing, etc.
-
-```{r pred_current}
-models %>% 
-  do(taxon = .$taxon,
-     model = .$model,
-     pred_current = raster::predict(object = raster::stack(pred$raster),
-                                    model = .$model)) ->
-  models
-```
-
-```{r pred_current_plot, echo=FALSE}
-raster::stack(models$pred_current) %>%
-  set_names(models$taxon) %>% 
-  plot()
-```
-
-### Binning
-
-To more easily interpret these predictions, we could convert the raw probability scores into confidence intervals, based on the proportion of known occurrences it successfully predicts.
-
-```{r binning}
-raster::stack(models$pred_current) %>%
-  set_names(paste0("pred_current_", models$taxon)) %>% 
-  raster::extract(as_Spatial(occ)) %>% 
-  {cbind(occ, .)} %>% 
-  st_as_sf(crs = st_crs(occ)) ->
-  occ2
-```
+## Discussion

### 2023-07-05 R/data.R

@@ -1,5 +1,28 @@
 # data.R
 
+#' Path helpers
+#'
+#' Helper functions to construct paths relative to the root of the research
+#' compendium.
+#'
+#' @param ... Additional path components passed to [here::here()]
+#'
+#' @rdname path_helpers
+#'
+#' @export
+#'
+#' @examples
+#' raw_data("ne")
+#' derived_data("gbif")
+raw_data <- function(...) {
+  here::here("analysis", "data", "raw_data", ...)
+}
+
+#' @rdname path_helpers
+#' @export
+derived_data <- function(...) {
+  here::here("analysis", "data", "derived_data", ...)
+}
 
 # Occurrence data (GBIF) --------------------------------------------------
 

### 2023-07-05 R/enm.R

@@ -4,21 +4,62 @@
 
 #' Generate pseudo-absence points
 #'
-#' @param region  object of class `sf` or `sfc`. The region in which to generate
-#'                pseudo-absence points.
+#' @param region  The region in which to generate pseudo-absence points. Must be
+#'                an object with an [sf::st_sample()] method, or an object
+#'                coercible with [sf::st_bbox] (for much faster sampling from
+#'                a rectangular bounding box).
 #' @param ...     Name-value pairs of expressions to be added as fixed attributes
 #'                of the pseudo-absences. See [dplyr::mutate()].
 #' @param N       integer. Number of pseudo-absence points to generate.
 #'
 #' @return
+#' An `sf` object containing `N` random points within `region`, with attributes
+#' specified in `...`.
+#'
 #' @export
 #'
 #' @examples
-generate_psabsences <- function(region, ..., N = 10000) {
-  sf::st_sample(region, N) %>%
-    as_tibble() %>%
-    mutate(...) %>%
-    return()
+background_sample <- function(region, N, coord_x = "x", coord_y = "y", ...) {
+  if (inherits(region, c("sf", "sfc", "sfg"))) {
+    bg_sample <- sample_sf(region, N, ...)
+  }
+  else {
+    bg_sample <- sample_bbox(region, N, ...)
+  }
+
+  if (!is.na(coord_x)) {
+    colnames(bg_sample)[colnames(bg_sample) == "X"] <- coord_x
+  }
+  if (!is.na(coord_y)) {
+    colnames(bg_sample)[colnames(bg_sample) == "Y"] <- coord_y
+  }
+
+  bg_sample
+}
+
+#' @noRd
+#' @keywords {internal}
+sample_bbox <- function(region, N, ...) {
+  region <- sf::st_bbox(region)
+  data.frame(
+    X = runif(N, region[["xmin"]], region[["xmax"]]),
+    Y =  runif(N, region[["ymin"]], region[["ymax"]])
+  ) |>
+    sf::st_as_sf(coords = c("X", "Y"),
+                 crs = sf::st_crs(region),
+                 remove = FALSE) |>
+    dplyr::mutate(...)
+}
+
+#' @noRd
+#' @keywords {internal}
+sample_sf <- function(region, N, ...) {
+  points <- sf::st_sample(region, N)
+  sf::st_sf(
+    as.data.frame(sf::st_coordinates(points)),
+    geometry = points
+    ) |>
+    dplyr::mutate(...)
 }
 
 #' @export

### 2023-07-05 R/utils-spatial.R

@@ -14,3 +14,21 @@ raster_burn <- function(points, raster) {
     raster::mask(raster) %>%
     return()
 }
+
+#' Buffer a bounding box
+#'
+#' @add Value to expand the bounding box by, in map units.
+#'
+#' @export
+#'
+#' @examples
+#' w_asia <- st_bbox(c(xmin = 30, xmax = 50, ymin = 25, ymax = 40))
+#' buffer_bbox(w_asia, 10)
+buffer_bbox <- function(bbox, add) {
+  sf::st_bbox(c(
+    xmin = bbox[["xmin"]] - add,
+    xmax = bbox[["xmax"]] + add,
+    ymin = bbox[["ymin"]] - add,
+    ymax = bbox[["ymax"]] + add
+  ))
+}

### 2023-07-31 analysis/analysis.Rmd
Use rgbif instead of dismo

@@
 library("ggspatial")
+library("here")
 library("parsnip")
@@
 ```{r data-occ}
-# Download or read data from GBIF
-# TODO: T. boeoticum ? (equate with T. urartu?)
+gbif_data <- function(scientificName, ...) {
+  gbif_data <- rgbif::occ_data(scientificName = scientificName, ...)
+  gbif_data$data
+}
+
 gbif_path <- here::here("analysis", "data", "derived_data", "gbif")
-occ <- bind_rows(
-  fetch_gbif("Hordeum", "spontaneum*", ext = w_asia, gbif_path),
-  #fetch_gbif("Bolboschoenus", "glaucus*", ext = w_asia, gbif_path),    
-  #fetch_gbif("Bolboschoenus", "maritimus*", ext = w_asia, gbif_path),    
-  #fetch_gbif("Scirpus", "maritimus*", ext = w_asia, gbif_path),
-  #fetch_gbif("Triticum", "urartu*", ext = w_asia, gbif_path),
-  #fetch_gbif("Triticum", "dicoccoides*", ext = w_asia, gbif_path),
-  #fetch_gbif("Stipa", "*", ext = w_asia, gbif_path),
-  #fetch_gbif("Zilla", "spinosa*", ext = w_asia, gbif_path),
-  #fetch_gbif("Vitex", "*", ext = w_asia, gbif_path),
-  #fetch_gbif("Fraxinus", "*", ext = w_asia, gbif_path)
-)
 
-# Code taxonomic groups for analysis
-occ |>
-  mutate(taxon = case_when(genus == "Hordeum" ~ "Hordeum spontaneum",
-                           genus == "Bolboschoenus" ~ "Bolboschoenus sp.",
-                           species == "Triticum urartu" ~ "Triticum urartu",
-                           species == "Triticum dicoccoides" ~ "Triticum dicoccoides",
-                           genus == "Stipa" ~ "Stipa sp.",
-                           genus == "Zilla" ~ "Zilla spinosa",
-                           genus == "Vitex" ~ "Vitex sp.",
-                           genus == "Fraxinus" ~ "Fraxinus sp.")) ->
+taxa <- c("Hordeum spontaneum" = "Hordeum spontaneum")
+          # "Bolboshoenus spp." = "Bolboschoenus")
+
+# Get occurrence data from GBIF
+# Including only records with coordinates and no recorded issues with geospatial
+#   data.
+taxa |>
+  map(gbif_data, hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
+      geometry = w_asia, limit = 100000) |>
+  list_rbind(names_to = "taxon") ->
   occ
 
-# TODO: convert to a test
-if(any(is.na(occ$taxon))) stop("Taxon contains missing values!")
+# Exclude:
+# * Coordinate uncertainty of 5 km or more (≈2.5 mins)
+# * Fossil occurrences
+occ |>
+  filter(coordinateUncertaintyInMeters < 5000 | is.na(coordinateUncertaintyInMeters)) |>
+  filter(basisOfRecord != c("FOSSIL_SPECIMEN")) ->
+  occ
 
-# Remove duplicate records within analytic groups
+# Remove duplicated coordinates (as recommended by CITE)
 occ |>
   group_by(taxon) |>
-  distinct(lon, lat, .keep_all = TRUE) |>
+  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE) |>
   ungroup() ->
@@
 occ |>
-  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) ->
+  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
+           crs = 4326, remove = FALSE) ->
   occ
@@
 ```{r plot-occ-distribution}
-ggplot() + 
+ggplot(occ) + 
+  facet_wrap(vars(taxon)) +
   layer_spatial(ne_land, fill = "#ffffff") +
@@
   layer_spatial(occ, alpha = 0.1, symbol = 16) +
-  # layer_spatial(ne_countries, fill = NA) +
+  layer_spatial(ne_countries, fill = NA) +
   coord_sf(xlim = w_asia[c("xmin", "xmax")], 
@@
                               n_background,
-                              coord_x = "lon",
-                              coord_y = "lat", 
+                              coord_x = "decimalLongitude",
+                              coord_y = "decimalLatitude", 
                               taxon = "Hordeum spontaneum", 
@@
                       region = buffer_bbox(w_asia, 10),
-                      cache_path = here::here(".paleoclim_cache"))) -> # TODO: revert to default cache
+                      cache_path = here(".paleoclim_cache"))) -> # TODO: revert to default cache
   climate
@@
 occ |>
-  mutate(bio = terra::extract(training_climate, data.frame(lon, lat), 
+  mutate(bio = terra::extract(training_climate, data.frame(decimalLongitude, decimalLatitude), 
                               method = "bilinear", ID = FALSE)) |>

### 2023-08-03 analysis/analysis.Rmd
Switch to stars from raster/terra

@@
 library("dplyr")
+library("fs")
 library("ggplot2")
@@
 library("sf")
+library("stars")
+library("stringr")
 library("tibble")
@@
 
-```{r data-paleoclim}
-tibble(period = c("lgm", "hs1", "ba", "yds", "eh", "cur")) |>
-  mutate(raster = map(period, paleoclim, "2_5m", 
-                      region = buffer_bbox(w_asia, 10),
-                      cache_path = here(".paleoclim_cache"))) -> # TODO: revert to default cache
+```{r data-climate}
+raw_data("paleoclim") |>
+  dir_ls() |>
+  map(dir_ls, glob = "*.tif") |>
+  map(read_stars) |>
+  map(\(x) set_names(x, path_ext_remove(path_file(names(x))))) ->
   climate
 
-training_climate <- filter(climate, period == "cur")$raster[[1]]
+# Tidy up names
+names(climate) |>
+  basename() |>
+  str_remove(regex("chelsa_", ignore_case = TRUE)) |>
+  str_split_i("_", 1) ->
+  names(climate)
+```
 
+```{r training-climate}
 occ |>
-  mutate(bio = terra::extract(training_climate, data.frame(decimalLongitude, decimalLatitude), 
-                              method = "bilinear", ID = FALSE)) |>
-  unpack(bio) |>
-  ungroup() ->
+  mutate(bio = extract_points(climate$cur, occ$geometry)) |>
+  unpack(bio) ->
   occ
@@
 ```{r model}
+# TODO: Investigate where these NAs are coming from
 rand_forest(mode = "regression") |>
@@
 ```{r predict}
-predict(rf_fit, as.data.frame(filter(climate, period == "cur")$raster), type = "raw")
+# TODO: deal with missing data in pred vars
+# TODO: turn back to something spatial? stars?
+predict(rf_fit, climate$cur)
 ```

### 2023-08-03 R/data.R

@@ -159,7 +159,7 @@ get_bioclim <- function(extent, resolution = 0.5, cache_path = getwd(), quiet = 
 #' @details
 #'
 #' The raster is always read or re-read from the cached file to ensure consistency.
-#' See [rpaleoview::paleoview()] for more details.
+#' See [rpaleoclim::paleoclim()] for more details.
 #'
 #' @export
 #'

### 2023-08-03 R/utils-spatial.R

@@ -32,3 +32,21 @@ buffer_bbox <- function(bbox, add) {
     ymax = bbox[["ymax"]] + add
   ))
 }
+
+#' Extract point values from a stars raster
+#'
+#' Analogous to raster::extract() / terra::extract(), but for stars format rasters.
+#'
+#' @param raster `stars` object
+#' @param points `sf` object with point geometry
+#'
+#' @return
+#' A data frame with the same number of rows as `points`.
+#'
+#' @export
+extract_points <- function(raster, points) {
+  raster |>
+    aggregate(points, \(x) x[1], as_points = FALSE) |>
+    as.data.frame() |>
+    dplyr::select(-geometry)
+}

### 2023-08-03 analysis/analysis.Rmd
Update TODOs

@@
 ```{r model}
-# TODO: Investigate where these NAs are coming from
+# TODO: Deal with missing values
 rand_forest(mode = "regression") |>
@@
 ```{r predict}
-# TODO: deal with missing data in pred vars
-# TODO: turn back to something spatial? stars?
 predict(rf_fit, climate$cur)

### 2023-08-03 R/utils-spatial.R
Simplify training data extraction

@@ -32,21 +32,3 @@ buffer_bbox <- function(bbox, add) {
     ymax = bbox[["ymax"]] + add
   ))
 }
-
-#' Extract point values from a stars raster
-#'
-#' Analogous to raster::extract() / terra::extract(), but for stars format rasters.
-#'
-#' @param raster `stars` object
-#' @param points `sf` object with point geometry
-#'
-#' @return
-#' A data frame with the same number of rows as `points`.
-#'
-#' @export
-extract_points <- function(raster, points) {
-  raster |>
-    aggregate(points, \(x) x[1], as_points = FALSE) |>
-    as.data.frame() |>
-    dplyr::select(-geometry)
-}

### 2023-08-03 analysis/analysis.Rmd
Simplify training data extraction
Fix model prediction

@@
 ```{r training-climate}
-occ |>
-  mutate(bio = extract_points(climate$cur, occ$geometry)) |>
-  unpack(bio) ->
-  occ
+occ <- cbind(occ, st_extract(climate$cur, occ))
 ```
@@
 ```{r model}
-# TODO: Deal with missing values
+# TODO: Where do the missing value(s) actually come from?
+occ |>
+  as_tibble() |>
+  select(present, starts_with("bio")) |>
+  drop_na() ->
+  rf_data
+
 rand_forest(mode = "regression") |>
-  fit(as.numeric(present) ~ ., data = select(occ, present, starts_with("bio_"))) ->
+  fit(as.numeric(present) ~ ., data = rf_data) ->
   rf_fit
@@
 ```{r predict}
-predict(rf_fit, climate$cur)
+rf_pred <- predict(climate$cur, rf_fit)
 ```

### 2023-08-04 analysis/analysis.Rmd
Tune RF model
Use a classification forest, probability (relative likelihood) output, down-sampling to deal with class imbalance

@@
 ```{r psabsences, include=FALSE}
-n_background <- 1000
+n_background <- 10000
 
@@
   occ
-
-# Weights
-#occ %>% 
-#  group_by(taxon, occurrence) %>% 
-#  mutate(weight = 1 + (n_psab / n() - 1)) ->
-#  occ
 ```
@@
 
+There are many methods for ENM.
+Random Forest is one of the best performing, providing it is appropriately parameterised to account for the class imbalance between presence and background samples (balanced random forest) [@ValaviEtAl2021a; @ValaviEtAl2021b].
+
+We used the R package ranger [@WrightZiegler2017].
+
 ```{r model}
 # TODO: Where do the missing value(s) actually come from?
+# Parameters adapted from the example of a down-sampled RF model given in
+#   https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
 occ |>
@@
 
-rand_forest(mode = "regression") |>
-  fit(as.numeric(present) ~ ., data = rf_data) ->
-  rf_fit
+# Model
+# TODO: tune?
+rand_forest(mode = "classification", trees = 1000) |>
+  set_engine(
+    "ranger",
+    probability = TRUE,
+    sample.fraction = c(2250 / n_background, 1), # TODO: parameterise
+  ) -> 
+  rf_model
+
+# Fit
+rf_fit <- fit(rf_model, factor(present) ~ ., data = rf_data)
 ```
@@
 ```{r predict}
-rf_pred <- predict(climate$cur, rf_fit)
+# TODO: predict on all periods (maybe combine stars and collapse into dimension first?)
+rf_pred_cur <- predict(climate$cur, rf_fit)
+rf_pred_eh <- predict(climate$EH, rf_fit)
 ```

### 2023-08-04 analysis/analysis.Rmd
Predict all climatologies

@@
 
-taxa <- c("Hordeum spontaneum" = "Hordeum spontaneum")
-          # "Bolboshoenus spp." = "Bolboschoenus")
+# taxa <- c("Hordeum spontaneum" = "Hordeum spontaneum")
+taxa <- c("Bolboshoenus spp." = "Bolboschoenus")
 
@@
     "ranger",
-    probability = TRUE,
-    sample.fraction = c(2250 / n_background, 1), # TODO: parameterise
+    # probability = TRUE,
+    sample.fraction = c(220 / n_background, 1), # TODO: parameterise
   ) -> 
@@
 # TODO: predict on all periods (maybe combine stars and collapse into dimension first?)
-rf_pred_cur <- predict(climate$cur, rf_fit)
-rf_pred_eh <- predict(climate$EH, rf_fit)
+rf_pred_cur <- predict(climate$cur, rf_fit, type = "prob")
+rf_pred_eh <- predict(climate$EH, rf_fit, type = "prob")
+rf_pred_yds <- predict(climate$YDS, rf_fit, type = "prob")
+rf_pred_ba <- predict(climate$BA, rf_fit, type = "prob")
+rf_pred_hs1 <- predict(climate$HS1, rf_fit, type = "prob")
+rf_pred_lgm <- predict(climate$LGM, rf_fit, type = "prob")
+
+rf_pred <- c(rf_pred_cur, rf_pred_eh, rf_pred_yds, rf_pred_ba, rf_pred_hs1,
+             rf_pred_lgm)
+names(rf_pred) <- c("cur", "EH", "YDS", "BA", "HS1", "LGM")
+
+ggplot() + 
+  geom_stars(data = st_redimension(st_crop(rf_pred, w_asia))) + 
+  facet_wrap(~new_dim)
 ```

### 2023-08-04 analysis/analysis.Rmd
Model taxon-by-taxon

--- 2023-08-03 analysis/analysis.Rmd
@@
 
-taxa <- c("Hordeum spontaneum" = "Hordeum spontaneum")
-          # "Bolboshoenus spp." = "Bolboschoenus")
+taxa <- c("Triticum dicoccoides" = "Triticum dicoccoides",
+          "Triticum urartu" = "Triticum urartu",
+          "Hordeum spontaneum" = "Hordeum spontaneum",
+          "Bolboshoenus spp." = "Bolboschoenus")
 
@@
 
-# Remove duplicated coordinates (as recommended by CITE)
+# Remove redundant observations of the same taxon in the same place
 occ |>
@@
 
-# Convert to sf
+# Drop unneeded columns, normalise column names and convert to sf
 occ |>
-  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), 
+  select(
+    taxon, gbif_key = key, genus, species,
+    longitude = decimalLongitude, latitude = decimalLatitude
+  ) |>
+  st_as_sf(coords = c("longitude", "latitude"), 
            crs = 4326, remove = FALSE) ->
@@
   annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
-  layer_spatial(occ, alpha = 0.1, symbol = 16) +
+  layer_spatial(occ, alpha = 0.2) +
   layer_spatial(ne_countries, fill = NA) +
@@
 
-```{r psabsences, include=FALSE}
-n_background <- 1000
+```{r data-background}
+n_background <- 10000
 
@@
   mutate(present = TRUE) |>
-  bind_rows(background_sample(s_levant_land, 
-                              n_background,
-                              coord_x = "decimalLongitude",
-                              coord_y = "decimalLatitude", 
-                              taxon = "Hordeum spontaneum", 
-                              present = FALSE)) ->
+  group_by(taxon) |>
+  nest(.key = "occ_data") |>
+  mutate(bg_data = map(1:n(), \(i) { 
+    background_sample(
+      region = s_levant_land, 
+      N = n_background,
+      coord_x = "longitude", coord_y = "latitude",
+      present = FALSE
+    )
+  })) |>
+  mutate(data = map2(occ_data, bg_data, bind_rows)) |>
+  select(taxon, data) |>
+  unnest(data) |>
+  st_sf() ->
   occ
-
-# Weights
-#occ %>% 
-#  group_by(taxon, occurrence) %>% 
-#  mutate(weight = 1 + (n_psab / n() - 1)) ->
-#  occ
 ```
@@
 ```{r training-climate}
-occ <- cbind(occ, st_extract(climate$cur, occ))
+occ <- cbind(occ, st_extract(climate$cur, occ)) |> select(-geometry.1)
 ```
@@
 
+There are many methods for ENM.
+Random Forest is one of the best performing, providing it is appropriately parameterised to account for the class imbalance between presence and background samples (balanced random forest) [@ValaviEtAl2021a; @ValaviEtAl2021b].
+
+We used the R package ranger [@WrightZiegler2017].
+
 ```{r model}
-# TODO: Where do the missing value(s) actually come from?
+rf_pred_vars <- function() {
+  c(any_of("present"), starts_with("bio"))
+}
+
+# Drop any rows with NA values in any of the predictor variables
+# TODO: Where do these actually come from? There's usually only a handful.
+occ <- drop_na(occ, rf_pred_vars())
+
+# Model
+# Parameters adapted from the example of a down-sampled RF model given in
+#   https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
+# TODO: tune?
+
+rand_forest(mode = "classification", trees = 1000) |>
+  set_engine(
+    "ranger",
+    probability = TRUE,
+    sample.fraction = c(sum(as.logical(.y())) / n_background, 1)
+  ) -> 
+  rf_model
+
+# Fit to each taxon
 occ |>
-  as_tibble() |>
-  select(present, starts_with("bio")) |>
-  drop_na() ->
-  rf_data
-
-rand_forest(mode = "regression") |>
-  fit(as.numeric(present) ~ ., data = rf_data) ->
-  rf_fit
+  nest(.by = taxon) |>
+  mutate(
+    data = map(data, as_tibble),
+    fit = map(data, \(data) {
+      fit(rf_model, factor(present) ~ ., data = select(data, rf_pred_vars()))
+    })) ->
+  rf_fits
 ```
@@
 ```{r predict}
-rf_pred <- predict(climate$cur, rf_fit)
+rf_fit <- rf_fits$fit[[1]]
+
+# TODO: predict on all periods (maybe combine stars and collapse into dimension first?)
+rf_pred_cur <- predict(climate$cur, rf_fit, type = "prob")
+rf_pred_eh <- predict(climate$EH, rf_fit, type = "prob")
+rf_pred_yds <- predict(climate$YDS, rf_fit, type = "prob")
+rf_pred_ba <- predict(climate$BA, rf_fit, type = "prob")
+rf_pred_hs1 <- predict(climate$HS1, rf_fit, type = "prob")
+rf_pred_lgm <- predict(climate$LGM, rf_fit, type = "prob")
+
+rf_pred <- c(rf_pred_cur, rf_pred_eh, rf_pred_yds, rf_pred_ba, rf_pred_hs1,
+             rf_pred_lgm)
+names(rf_pred) <- c("cur", "EH", "YDS", "BA", "HS1", "LGM")
+
+ggplot() + 
+  geom_stars(data = st_redimension(st_crop(rf_pred, w_asia))) + 
+  facet_wrap(~new_dim)
 ```

### 2023-08-15 analysis/analysis.Rmd
Add TODO
Add (lots) more taxa

@@
 format: html
+bibliography: references.bib
 ---
@@
 library("fs")
+library("furrr")
 library("ggplot2")
 library("ggspatial")
+library("gt")
 library("here")
@@
 library("ranger")
+library("readr")
+library("rgbif")
 library("rpaleoclim")
@@
 library("BadiaPaleoFloraENM") # this package, install with devtools::install()
+```
+
+```{r parallelise}
+# Several parts of this analysis are processing- and memory-intensive.
+# You may need to adjust the futures strategy specified here to fit your 
+# computational resources. See ?future::plan for options.
+plan("multisession")
 ```
@@
 
-### Occurrence data
+### Taxa and occurrence data
 
-Occurrence data was obtained from GBIF.
+To generate the list of taxa modelled in this study, we ...
+
+* Taxa present at more than three Neolithic sites, according to the database of our previous study [@ArranzOtaeguiRoe2023]
+* Identified at least to species (and matchable to the GBIF Backbone Taxonomy)
+* Have at least 100 occurrences in West Asia (meeting the criteria below) in the GBIF database
+* Excluded domestic species, replacing them with their wild progenitor, where known and present in the GBIF database
+
+We also considered the number modern occurrences available for each taxon.
+ENMs are known to have reasonable predictive power even with small training samples (i.e. less than a hundred observations, [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008]).
+
+```{r data-neolithic-flora}
+neolithic_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv"))
+
+# Rename some columns for clarity
+neolithic_flora <- rename(neolithic_flora,
+                          taxon_group = "taxon", taxon = "taxon_detail")
+
+# Assign to archaeological periods
+periods <- tribble(
+  ~period,             ~agriculture,                          ~start_bp, ~end_bp,
+  "Late Epipal.",      "Foraging",                            15000,     11700,
+  "PPNA",              "Pre-domestication cultivation",       11700,     10700,
+  "EPPNB",             "Cultivation of domesticated species", 10700,     10200,
+  "MPPNB",             "Cultivation of domesticated species", 10200,     9500,
+  "LPPNB/C",           "Agriculture",                         9500,      8500,
+  "Pottery Neolithic", "Agriculture",                         8500,      6500,
+  "Chalcolithic",      "Agriculture",                         6500,      5000,
+)
+
+neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")
+
+neolithic_flora <- mutate(neolithic_flora,
+                          age_mid = age_end + ((age_start - age_end) / 2),
+                          period = cut(-age_mid, 
+                                       breaks = -c(periods$start_bp, 0), 
+                                       labels = periods$period,
+                                       ordered_result = TRUE)
+)
+
+# Assemblage statistics
+neolithic_flora |> 
+  distinct(site_name, phase_code, period) |> 
+  filter(period %in% neolithic) |> 
+  nrow() ->
+  n_assemb_neolithic
+
+neolithic_flora |> 
+  filter(period %in% neolithic) |> 
+  distinct(site_name) |> 
+  nrow() ->
+  n_site_neolithic
+```
+
+```{r data-taxa}
+neolithic_flora |>
+  # group_by(taxon_group) |>
+  # mutate(taxon = list(unique(taxon))) |>
+  filter(period %in% neolithic) |>
+  group_by(taxon) |>
+  summarise(
+    n_present = length(unique(phase_code)),
+    p_present = n_present / n_assemb_neolithic,
+    .groups = "drop"
+  ) |>
+  filter(n_present > 2) ->
+  taxa
+```
+
+```{r taxa-names}
+# Normalise species names and match to the GBIF Backbone Taxonomy
+# Excluding taxa not identified to species
+taxa |>
+  filter(!str_ends(taxon, "sp.")) |>
+  mutate(
+    taxon = case_match(taxon,
+                       "Vicia narbonense" ~ "Vicia narbonensis", # N=3
+                       "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
+                       "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
+                       "Lens orientalis" ~ "Vicia orientalis", # N=2
+                       "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
+                       "Stipa capensis" ~ "Stipa dregeana", # N=2
+                       .default = taxon),
+    taxon_original = taxon,
+    taxon = map_chr(taxon, gbif_accepted_name)
+  ) ->
+  taxa
+
+# TODO: trim to wild types?
+# taxa |>
+#   drop_na(taxon) |>
+#   mutate(
+#     taxon = str_remove(taxon, " sp."),
+#     taxon_original = taxon,
+#     taxon = map_chr(taxon, gbif_accepted_name)
+#   ) |>
+#   drop_na(taxon) |> # Drop names that couldn't be matched to GBIF
+#   filter(!taxon %in% domesticates) |>
+#   group_by(taxon_group) |>
+#   mutate(
+#     genus = str_extract(taxon, "[^\\s]+"),
+#     is_genus = genus == taxon,
+#     genus_in_group = genus %in% taxon
+#   ) |>
+#   filter(is_genus | !genus_in_group) |> # Drop redundant taxons
+#   distinct(taxon_group, taxon, n_present, p_present) ->
+#   taxa
+```
+
+```{r taxa-domesticates}
+# Recode domesticates as their wild progenitors.
+# Add other wild progenitors of wheats
+taxa |> 
+  mutate(taxon = case_match(
+    taxon,
+    "Cicer arietinum" ~ "Cicer reticulatum",
+    "Hordeum vulgare" ~ "Hordeum spontaneum",
+    "Vicia lens" ~ "Vicia orientalis",
+    "Linum usitatissimum" ~ "Linum bienne",
+    "Triticum aestivum" ~ "Triticum turgidum dicoccum",
+    "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
+    "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
+    "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
+    .default = taxon
+  )) |>
+  bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
+                             "Aegilops tauschii"))) ->
+  taxa
+```
+
+```{r taxa-occ-count}
+taxa |>
+  group_by(taxon) |>
+  summarise(
+    taxon_original = list(c(taxon_original)),
+    n_present = max(n_present), # TODO: be more precise?
+    p_present = max(p_present)
+    ) |>
+  mutate(n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
+                                               hasCoordinate = TRUE, 
+                                               hasGeospatialIssue = FALSE,
+                                               geometry = bbox_wkt(w_asia)))) ->
+  taxa
+```
+
+```{r tbl-taxa-occ-count}
+#| tbl-cap: "Taxonomic groups modelled in this study."
+# TODO: add to caption:
+# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
+# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
+# TODO: add count of species included?
+taxa |>
+  mutate(
+    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
+    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
+    include = n_occ > 50
+  ) |>
+  select(-taxon_original) |>
+  group_by(include) |>
+  arrange(-n_present) |>
+  gt() |>
+  cols_merge(
+    c(taxon, taxon_detail),
+    rows = !is.na(taxon_detail),
+    pattern = "{1}<br><small>(incl. {2})</small>"
+  ) |>
+  fmt_percent(p_present, decimals = 0) |>
+  cols_merge_n_pct(n_present, p_present) |>
+  tab_style(cell_text(style = "italic"), cells_body(taxon))
+```
+
+Occurrence data was obtained from GBIF using the GBIF occurrence API.
+This deals with synonyms etc.
+(using the GBIF Backbone Taxonomy as a standard), so e.g.
+retrieving data for "Bolboschoenus" includes all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.
 
 ```{r data-occ}
+taxa <- filter(taxa, n_occ > 50)
 gbif_data <- function(scientificName, ...) {
@@
 
-taxa <- c("Triticum dicoccoides" = "Triticum dicoccoides",
-          "Triticum urartu" = "Triticum urartu",
-          "Hordeum spontaneum" = "Hordeum spontaneum",
-          "Bolboshoenus spp." = "Bolboschoenus")
+#taxa <- c("Triticum dicoccoides" = "Triticum dicoccoides",
+#          "Triticum urartu" = "Triticum urartu",
+#          "Hordeum spontaneum" = "Hordeum spontaneum",
+#          "Vicia orientalis" = "Vicia orientalis", # AKA Lens (culinaris) orientalis
+#          "Bolboshoenus spp." = "Bolboschoenus")
+
+# TODO: maybe first with occ_count first?
 
@@
 taxa |>
-  map(gbif_data, hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
+  select(taxon, taxon) |>
+  deframe() |> # Extract as named vector
+  future_map(gbif_data, hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
       geometry = w_asia, limit = 100000) |>
@@
 
+# TODO: temp dump -- remove
+save(occ, file = "big_occ.Rdata")
+
 # Drop unneeded columns, normalise column names and convert to sf
@@
   layer_spatial(ne_land, fill = "#ffffff") +
-  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
+  # annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
   layer_spatial(occ, alpha = 0.2) +
-  layer_spatial(ne_countries, fill = NA) +
+  # layer_spatial(ne_countries, fill = NA) +
   coord_sf(xlim = w_asia[c("xmin", "xmax")], 
@@
 ```{r data-background}
+# TODO: Make your mind up – what region are we modelling?
+# Currently presence data is taken from across W Asia, but background is
+#   generated only for the S. Levant.
+# TODO: maybe defer this to just before the model is actually fit 
+#   (and discard the background sample)
 n_background <- 10000
@@
 ```{r predict}
-rf_fit <- rf_fits$fit[[1]]
+# TODO: tidy
+# TODO: parallelise
+rf_fit <- filter(rf_fits, taxon == "Secale cereale") |>
+  pull(fit)[[1]]
 

### 2023-08-15 R/gbif.R

@@ -1,0 +1,21 @@
+# gbif.R
+# Functions for working with data from the GBIF API
+
+#' Accepted name of a taxon
+#'
+#' Resolves the accepted name of a given taxon according to the GBIF Backbone
+#' Taxonomy.
+#'
+#' @param name Character. Name of a taxon.
+#'
+#' @return
+#' If the GBIF Backbone Taxonomy considers `name` a synonym, the canonical name
+#' of the accepted usage. Otherwise, `name`
+#'
+#' @export
+gbif_accepted_name <- function(name) {
+  gbif <- rgbif::name_backbone(name)
+  if (gbif$matchType != "EXACT") NA_character_
+  else if (gbif$synonym) name_usage(gbif$acceptedUsageKey)$data$canonicalName
+  else gbif$canonicalName
+}

### 2023-08-15 R/utils-spatial.R

@@ -32,3 +32,13 @@ buffer_bbox <- function(bbox, add) {
     ymax = bbox[["ymax"]] + add
   ))
 }
+
+#' Well-Known Text (WKT) representation of a bounding box
+#'
+#' @param bbox [sf::st_bbox] object
+#'
+#' @return
+#' @export
+bbox_wkt <- function(bbox) {
+  sf::st_as_text(sf::st_as_sfc(bbox))
+}

### 2023-08-15 analysis/analysis.Rmd
Tidy predictions

@@
 # Fit to each taxon
+# TODO: parallelise
 occ |>
@@
 # TODO: tidy
-# TODO: parallelise
-rf_fit <- filter(rf_fits, taxon == "Secale cereale") |>
-  pull(fit)[[1]]
+rf_fit <- filter(rf_fits, taxon == "Triticum urartu") |>
+  pull(fit)
+rf_fit <- rf_fit[[1]]
 
-# TODO: predict on all periods (maybe combine stars and collapse into dimension first?)
-rf_pred_cur <- predict(climate$cur, rf_fit, type = "prob")
-rf_pred_eh <- predict(climate$EH, rf_fit, type = "prob")
-rf_pred_yds <- predict(climate$YDS, rf_fit, type = "prob")
-rf_pred_ba <- predict(climate$BA, rf_fit, type = "prob")
-rf_pred_hs1 <- predict(climate$HS1, rf_fit, type = "prob")
-rf_pred_lgm <- predict(climate$LGM, rf_fit, type = "prob")
-
-rf_pred <- c(rf_pred_cur, rf_pred_eh, rf_pred_yds, rf_pred_ba, rf_pred_hs1,
-             rf_pred_lgm)
-names(rf_pred) <- c("cur", "EH", "YDS", "BA", "HS1", "LGM")
+# TODO: Maybe combine stars and collapse period into dimension first?
+rf_pred <- map(climate, predict, model = rf_fit, type = "prob")
+rf_pred <- map(rf_pred, \(x) x[".pred_TRUE"])
+rf_pred <- do.call(c, rf_pred)
+rf_pred <- st_redimension(rf_pred, name = "period")
 

### 2023-08-16 analysis/analysis.Rmd
Refactor occurrence reading and tidying

--- 2023-08-04 analysis/analysis.Rmd
@@
 
-```{r definitions}
+```{r setup-parallel}
+# Several parts of this analysis are processing- and memory-intensive.
+# You may need to adjust the futures strategy specified here to fit your 
+# computational resources. See ?future::plan for options.
+plan("multisession")
+```
+
+```{r define-regions}
 latlong <- 4326
@@
 s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
+```
+
+```{r define-periods}
+archaeo_periods <- tribble(
+  ~period,             ~agriculture,                          ~start_bp, ~end_bp,
+  "Late Epipal.",      "Foraging",                            15000,     11700,
+  "PPNA",              "Pre-domestication cultivation",       11700,     10700,
+  "EPPNB",             "Cultivation of domesticated species", 10700,     10200,
+  "MPPNB",             "Cultivation of domesticated species", 10200,     9500,
+  "LPPNB/C",           "Agriculture",                         9500,      8500,
+  "Pottery Neolithic", "Agriculture",                         8500,      6500,
+  "Chalcolithic",      "Agriculture",                         6500,      5000,
+)
+
+neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")
 ```
@@
 
-```{r data-basemap}
+```{r read-basemap}
 # Turn off s2 while we deal with Natural Earth's dodgy geometries
@@
 s_levant_land <- st_intersection(st_as_sfc(s_levant), ne_land)
+w_asia_land <- st_intersection(st_as_sfc(w_asia), ne_land)
 
@@
 
-Occurrence data was obtained from GBIF.
+```{r read-archaeo_flora}
+archae_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv"))
+```
 
-```{r data-occ}
+```{r tidy-archaeo_flora}
+archaeo_flora |>
+  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
+  # Assign to archaeological periods
+  # TODO: refactor as function
+  # TODO: assign climate periods too
+  mutate(
+    age_mid = age_end + ((age_start - age_end) / 2),
+    period = cut(-age_mid, 
+                 breaks = -c(archaeo_periods$start_bp, 0), 
+                 labels = archaeo_periods$period,
+                 ordered_result = TRUE)
+  ) ->
+  archaeo_flora
+```
+
+```{r summarise-archaeo_flora}
+n_assemb_neolithic <- archaeo_flora |> 
+  distinct(site_name, phase_code, period) |> 
+  filter(period %in% neolithic) |> 
+  nrow()
+
+n_site_neolithic <- archaeo_flora |> 
+  filter(period %in% neolithic) |> 
+  distinct(site_name) |> 
+  nrow()
+```
+
+To generate the list of taxa modelled in this study, we ...
+
+* Taxa present at more than three Neolithic sites, according to the database of our previous study [@ArranzOtaeguiRoe2023]
+* Identified at least to species (and matchable to the GBIF Backbone Taxonomy)
+* Have at least 100 occurrences in West Asia (meeting the criteria below) in the GBIF database
+* Excluded domestic species, replacing them with their wild progenitor, where known and present in the GBIF database
+
+We also considered the number modern occurrences available for each taxon.
+ENMs are known to have reasonable predictive power even with small training samples (i.e. less than a hundred observations, [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008]).
+
+```{r generate-flora}
+archaeo_flora |>
+  filter(period %in% neolithic) |>
+  group_by(taxon) |>
+  summarise(
+    n_present = length(unique(phase_code)),
+    p_present = n_present / n_assemb_neolithic,
+    .groups = "drop"
+  ) |>
+  filter(n_present > 2) ->
+  flora
+```
+
+```{r tidy-flora-taxonomy}
+# Normalise species names and match to the GBIF Backbone Taxonomy
+# Excluding taxa not identified to species
+flora |>
+  filter(!str_ends(taxon, "sp.")) |>
+  mutate(
+    taxon = case_match(taxon,
+                       "Vicia narbonense" ~ "Vicia narbonensis", # N=3
+                       "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
+                       "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
+                       "Lens orientalis" ~ "Vicia orientalis", # N=2
+                       "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
+                       "Stipa capensis" ~ "Stipa dregeana", # N=2
+                       .default = taxon),
+    taxon_original = taxon,
+    taxon = map_chr(taxon, gbif_accepted_name)
+  ) ->
+  flora
+```
+
+```{r tidy-flora-domesticates}
+# Recode domesticates as their wild progenitors.
+# Add other wild progenitors of wheats
+flora |> 
+  mutate(taxon = case_match(
+    taxon,
+    "Cicer arietinum" ~ "Cicer reticulatum",
+    "Hordeum vulgare" ~ "Hordeum spontaneum",
+    "Vicia lens" ~ "Vicia orientalis",
+    "Linum usitatissimum" ~ "Linum bienne",
+    "Triticum aestivum" ~ "Triticum turgidum dicoccum",
+    "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
+    "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
+    "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
+    .default = taxon
+  )) |>
+  bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
+                             "Aegilops tauschii"))) ->
+  flora
+```
+
+```{r fetch-occ-count}
+flora |>
+  group_by(taxon) |>
+  summarise(
+    taxon_original = list(c(taxon_original)),
+    n_present = max(n_present), # TODO: be more precise?
+    p_present = max(p_present)
+    ) |>
+  mutate(n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
+                                               hasCoordinate = TRUE, 
+                                               hasGeospatialIssue = FALSE,
+                                               geometry = bbox_wkt(w_asia)))) ->
+  flora
+```
+
+```{r tbl-occ-count}
+#| tbl-cap: "Species modelled in this study."
+# TODO: add to caption:
+# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
+# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
+# TODO: add count of species included?
+flora |>
+  mutate(
+    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
+    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
+    include = n_occ > 50
+  ) |>
+  select(-taxon_original) |>
+  group_by(include) |>
+  arrange(-n_present) |>
+  gt() |>
+  cols_merge(
+    c(taxon, taxon_detail),
+    rows = !is.na(taxon_detail),
+    pattern = "{1}<br><small>(incl. {2})</small>"
+  ) |>
+  fmt_percent(p_present, decimals = 0) |>
+  cols_merge_n_pct(n_present, p_present) |>
+  tab_style(cell_text(style = "italic"), cells_body(taxon))
+```
+
+```{r filter-flora}
+flora <- filter(flora, n_occ > 50)
+```
+
+Occurrence data was obtained from GBIF using the GBIF occurrence API.
+This deals with synonyms etc.
+(using the GBIF Backbone Taxonomy as a standard), so e.g.
+retrieving data for "Bolboschoenus" includes all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.
+
+```{r fetch-occ}
+# TODO: Refactor
 gbif_data <- function(scientificName, ...) {
@@
 
-gbif_path <- here::here("analysis", "data", "derived_data", "gbif")
-
-taxa <- c("Triticum dicoccoides" = "Triticum dicoccoides",
-          "Triticum urartu" = "Triticum urartu",
-          "Hordeum spontaneum" = "Hordeum spontaneum",
-          "Bolboshoenus spp." = "Bolboschoenus")
-
-# Get occurrence data from GBIF
-# Including only records with coordinates and no recorded issues with geospatial
-#   data.
-taxa |>
-  map(gbif_data, hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
-      geometry = w_asia, limit = 100000) |>
-  list_rbind(names_to = "taxon") ->
-  occ
-
-# Exclude:
-# * Coordinate uncertainty of 5 km or more (≈2.5 mins)
-# * Fossil occurrences
-occ |>
-  filter(coordinateUncertaintyInMeters < 5000 | is.na(coordinateUncertaintyInMeters)) |>
-  filter(basisOfRecord != c("FOSSIL_SPECIMEN")) ->
-  occ
-
-# Remove redundant observations of the same taxon in the same place
-occ |>
-  group_by(taxon) |>
-  distinct(decimalLongitude, decimalLatitude, .keep_all = TRUE) |>
-  ungroup() ->
-  occ
-
-# Drop unneeded columns, normalise column names and convert to sf
-occ |>
-  select(
-    taxon, gbif_key = key, genus, species,
-    longitude = decimalLongitude, latitude = decimalLatitude
-  ) |>
-  st_as_sf(coords = c("longitude", "latitude"), 
-           crs = 4326, remove = FALSE) ->
-  occ
+flora |>
+  mutate(
+    occ = future_map(taxon, gbif_data, 
+                     hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
+                     geometry = w_asia, limit = 100000)
+  ) ->
+  flora
 ```
 
-```{r plot-occ-distribution}
-ggplot(occ) + 
+```{r tidy-occ}
+# TODO: further tidying. Ideas:
+# - https://data-blog.gbif.org/post/gbif-filtering-guide/
+# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
+# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
+flora |>
+  mutate(
+    occ = future_map(occ, gbif_drop_imprecise_coords, threshold = 5000),
+    occ = future_map(occ, gbif_drop_fossils),
+    occ = future_map(occ, gbif_drop_duplicate_coords),
+    occ = future_map(occ, gbif_standardise_names),
+    occ = future_map(occ, \(x) {
+      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
+             latitude)
+    })
+  ) ->
+  flora
+```
+
+```{r fig-occ-map}
+flora |>
+  unnest(occ) |>
+  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
+  ggplot() + 
   facet_wrap(vars(taxon)) +
   layer_spatial(ne_land, fill = "#ffffff") +
-  annotation_spatial(s_levant, fill = alpha("red", 0.15), colour = "red") +
-  layer_spatial(occ, alpha = 0.2) +
-  layer_spatial(ne_countries, fill = NA) +
+  geom_sf(shape = ".") +
   coord_sf(xlim = w_asia[c("xmin", "xmax")], 
@@
 
-```{r data-background}
+```{r generate-background}
+# TODO: Refactor to work with nested occ
+# TODO: maybe defer this to just before the model is actually fit 
+#   (and discard the background sample)
 n_background <- 10000
@@
     background_sample(
-      region = s_levant_land, 
+      region = w_asia_land, 
       N = n_background,
@@
 
-### Climate data
+### Predictor data
 
@@
 
-### Other predictor data
-
-Soil, terrain (incl. TWI, slope?)
+TODO: Soil, terrain (incl. TWI, slope?)
 
@@
 ```{r model}
-rf_pred_vars <- function() {
+rf_predictors <- function() {
   c(any_of("present"), starts_with("bio"))
@@
 # TODO: Where do these actually come from? There's usually only a handful.
-occ <- drop_na(occ, rf_pred_vars())
+occ <- drop_na(occ, rf_predictors())
 
@@
 
-rand_forest(mode = "classification", trees = 1000) |>
+rf_model <- rand_forest(mode = "classification", trees = 1000) |>
   set_engine(
@@
     sample.fraction = c(sum(as.logical(.y())) / n_background, 1)
-  ) -> 
-  rf_model
+  )
 
 # Fit to each taxon
+# TODO: parallelise
 occ |>
@@
   rf_fits
+
+# TODO: not this
+rf_fits$taxon <- taxa$taxon
+```
+
+The output of the model is probalistic.
+However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to habitat suitability.
+To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
+We select the threshold value for each model individually, using MaxSSS (as recommended by [@LuiEtAl2013])
+This also makes it possible to analyses the predictions together as an assemblage.
+
+```{r thresholding}
+# TODO: thresholding
+# - Pick a threshold
+# - Reclassify pred raster
+# - Extract predicted classification (for cur) for occ
+# - ...or predict back onto the occurrence data? do we actually need the rasters for this?
+# - Calculate sensitivity and with yardstick
+# - Sum (SSS)
+# - (Plot?)
+# - Pick max SSS
 ```
@@
 ```{r predict}
-rf_fit <- rf_fits$fit[[1]]
+# TODO: tidy
+rf_fit <- filter(rf_fits, taxon == "Triticum urartu") |>
+  pull(fit)
+rf_fit <- rf_fit[[1]]
 
-# TODO: predict on all periods (maybe combine stars and collapse into dimension first?)
-rf_pred_cur <- predict(climate$cur, rf_fit, type = "prob")
-rf_pred_eh <- predict(climate$EH, rf_fit, type = "prob")
-rf_pred_yds <- predict(climate$YDS, rf_fit, type = "prob")
-rf_pred_ba <- predict(climate$BA, rf_fit, type = "prob")
-rf_pred_hs1 <- predict(climate$HS1, rf_fit, type = "prob")
-rf_pred_lgm <- predict(climate$LGM, rf_fit, type = "prob")
-
-rf_pred <- c(rf_pred_cur, rf_pred_eh, rf_pred_yds, rf_pred_ba, rf_pred_hs1,
-             rf_pred_lgm)
-names(rf_pred) <- c("cur", "EH", "YDS", "BA", "HS1", "LGM")
+# TODO: Maybe combine stars and collapse period into dimension first?
+rf_pred <- map(climate, predict, model = rf_fit, type = "prob")
+rf_pred <- map(rf_pred, \(x) x[".pred_TRUE"])
+rf_pred <- do.call(c, rf_pred)
+rf_pred <- st_redimension(rf_pred, name = "period")
 
 ggplot() + 
-  geom_stars(data = st_redimension(st_crop(rf_pred, w_asia))) + 
-  facet_wrap(~new_dim)
+  geom_stars(data = st_crop(rf_pred, w_asia)) + 
+  facet_wrap(~period) +
+  scale_fill_batlow(reverse = TRUE)
 ```
 
-### Verification
+### Model assessment & verification
+
+```{r todo}
+# TODO: model performance assessment:
+# - R2
+# - Boyce index
+# - AUC, ROC, AUC-ROC, AUC-PRG ?
+# - TSS
+
+```
 

### 2023-08-16 R/gbif.R

@@ -10,7 +10,7 @@
 #'
 #' @return
 #' If the GBIF Backbone Taxonomy considers `name` a synonym, the canonical name
-#' of the accepted usage. Otherwise, `name`
+#' of the accepted usage. Otherwise, `name`.
 #'
 #' @export
 gbif_accepted_name <- function(name) {
@@ -19,3 +19,80 @@ gbif_accepted_name <- function(name) {
   else if (gbif$synonym) name_usage(gbif$acceptedUsageKey)$data$canonicalName
   else gbif$canonicalName
 }
+
+#' Clean GBIF data
+#'
+#' A collection of functions for cleaning occurrence data from GBIF.
+#'
+#' @param data Table of occurrence data retrieved with [rgbif::occ_data()]
+#' @param threshold For `gbif_drop_imprecise_coords()`, the minimum acceptable
+#'   coordinate uncertainty (in meters).
+#' @param shorten If `TRUE`, `gbif_standardise_names()` shorten long column
+#'   names in addition to converting them to snake case, e.g. `decimalLongitude`
+#'   to `longitude`.
+#'
+#' @return
+#' Modified version of `data`.
+#'
+#' @rdname gbif_clean
+#' @export
+gbif_drop_imprecise_coords <- function(data, threshold = 5000) {
+  if (!"coordinateUncertaintyInMeters" %in% names(data)) return(data)
+  dplyr::filter(data, .data$coordinateUncertaintyInMeters < threshold |
+                  is.na(.data$coordinateUncertaintyInMeters))
+}
+
+#' @rdname gbif_clean
+#' @export
+gbif_drop_duplicate_coords <- function(data) {
+  dplyr::distinct(data, .data$decimalLongitude, .data$decimalLatitude,
+                  .keep_all = TRUE)
+}
+
+#' @rdname gbif_clean
+#' @export
+gbif_drop_fossils <- function(data) {
+  dplyr::filter(data, .data$basisOfRecord != "FOSSIL_SPECIMEN")
+}
+
+#' @rdname gbif_clean
+#' @export
+gbif_standardise_names <- function(data, shorten = TRUE) {
+  colnames(data) <- snakecase::to_snake_case(colnames(data))
+  if (isTRUE(shorten)) data <- gbif_shorten_names(data)
+  return(data)
+}
+
+#' @noRd
+#' @keywords internal
+gbif_shorten_names <- function(data) {
+    dplyr::rename(
+      data,
+      longitude = "decimal_longitude",
[…]

### 2023-08-17 analysis/analysis.Rmd
Tidy fitting

@@
 library("parsnip")
+library("progressr")
 library("purrr")
@@
 ```{r read-archaeo_flora}
-archae_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv"))
+archaeo_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv"))
 ```
@@
 flora <- filter(flora, n_occ > 50)
+# TODO: Temporary measure.
+# Why are there 47k occurrences of Avena sterilis??
+flora <- filter(flora, taxon != "Avena sterilis")
 ```
@@
   mutate(
-    occ = future_map(taxon, gbif_data, 
+    occ = map(taxon, gbif_data, 
               hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
-                     geometry = w_asia, limit = 100000)
+              geometry = w_asia, limit = 100000,
+              .progress = TRUE)
   ) ->
@@
   mutate(
-    occ = future_map(occ, gbif_drop_imprecise_coords, threshold = 5000),
-    occ = future_map(occ, gbif_drop_fossils),
-    occ = future_map(occ, gbif_drop_duplicate_coords),
-    occ = future_map(occ, gbif_standardise_names),
-    occ = future_map(occ, \(x) {
+    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
+    occ = map(occ, gbif_drop_fossils),
+    occ = map(occ, gbif_drop_duplicate_coords),
+    occ = map(occ, gbif_standardise_names),
+    occ = map(occ, \(x) {
       select(x, gbif_key = key, scientific_name, genus, species, longitude, 
@@
 ```{r generate-background}
-# TODO: Refactor to work with nested occ
-# TODO: maybe defer this to just before the model is actually fit 
+# TODO: maybe defer this to just before the model is actually fit ?
 #   (and discard the background sample)
@@
 
-occ |>
-  mutate(present = TRUE) |>
-  group_by(taxon) |>
-  nest(.key = "occ_data") |>
-  mutate(bg_data = map(1:n(), \(i) { 
+flora |>
+  mutate(
+    occ = map(occ, mutate, present = TRUE),
+    bg_data = map(1:n(), \(i) { 
       background_sample(
@@
       )
-  })) |>
-  mutate(data = map2(occ_data, bg_data, bind_rows)) |>
-  select(taxon, data) |>
-  unnest(data) |>
-  st_sf() ->
-  occ
+    }),
+    occ = map2(occ, bg_data, bind_rows)
+  ) |>
+  select(-bg_data) ->
+  flora
 ```
@@
 ```{r training-climate}
-occ <- cbind(occ, st_extract(climate$cur, occ)) |> select(-geometry.1)
+flora |> 
+  mutate(
+    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x))))
+  ) ->
+  flora
 ```
@@
 # TODO: Where do these actually come from? There's usually only a handful.
-occ <- drop_na(occ, rf_predictors())
+# occ <- drop_na(occ, rf_predictors())
 
@@
 # TODO: parallelise
-occ |>
-  nest(.by = taxon) |>
+flora |>
   mutate(
-    data = map(data, as_tibble),
-    fit = map(data, \(data) {
-      fit(rf_model, factor(present) ~ ., data = select(data, rf_pred_vars()))
-    })) ->
-  rf_fits
-
-# TODO: not this
-rf_fits$taxon <- taxa$taxon
+    fit = map(occ, \(occ) {
+      fit(rf_model, factor(present) ~ ., data = select(occ, rf_predictors()))
+    }, .progress = TRUE)
+  ) ->
+  flora
 ```

### 2023-08-17 R/gbif.R

@@ -88,11 +88,6 @@ gbif_shorten_names <- function(data) {
 #'
 #' @export
 gbif_to_sf <- function(data) {
-  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)
+  sf::st_as_sf(data, coords = c("longitude", "latitude"),
+               crs = 4326, remove = FALSE)
 }
-
-# Drop unneeded columns, normalise column names and convert to sf
-#  select(
-#    taxon, gbif_key = key, genus, species,
-#    longitude = decimalLongitude, latitude = decimalLatitude
-#  )

### 2023-09-01 analysis/analysis.qmd
Outline introductory sections

--- 2023-08-17 analysis/analysis.Rmd
@@
 ---
-title: "Palaeodistribution of crop progenitors and wild foods in the Southern Levant"
-author: "Joe Roe"
+title: Palaeodistribution of crop progenitors and other plant resources over the Pleistocene–Holocene transition in Southwest Asia
+date: today
+author: 
+- name: Joe Roe
+  id: jr
+  orcid: 0000-0002-1011-1244
+  email: joeroe@hey.com
+  affiliation: University of Bern
+- name: Amaia Arranz-Otaegui
+  id: aao
+  orcid: 0000-0002-5091-6426
+  affiliation: University of the Basque Country
+abstract: |
+  This paper presents the first comprehensive reconstructions of the prehistoric 
+  ranges of [X] plant species known to have been gathered or cultivated by early
+  agricultural societies in Southwest Asia, including the progenitors of the 
+  first crops. We used machine learning to train an ecological niche model (ENM)
+  of each species based on its present-day distribution in relation to climate 
+  and environmental variables. Predictions of the potential ranges of these 
+  species at key stages of the Pleistocene–Holocene transition could then be
+  derived from these models using palaeoclimate 
+  Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
+  compared to today, indicating [...]. The modelled ranges predict the
+  observed occurrence of these species on archaeological sites with [low|medium|high]
+  accuracy. The regional ubiquity of species in the archaeological record is [not] 
+  correlated with the predicted size of its range and the diversity of
+  archaeobotanical assemblages is [not] correlated with the predicted diversity
+  of its environs. This indicates that trends in taxonomic composition of
+  the archaeobotanical record is [not] likely to be influenced by environmental 
+  change and species turnover, not just, as is often assumed, human economic 
+  choices.
 format: html
@@
 
+* The first farming societies had an ecological context
+* Subsistence is understood (largely) through archaeobotany; ecological context from archaeobotany, palynology, palaeoclimate, etc.
+  * But these have a variety of biases (human selection, taphonomy, etc.)
+  * And at the end of the day only represent specific places – interpolation to the entire region is not straightforward
+  * Overlap makes it difficult to see where human choices depart from the environmental background [cf. Martin et al. 2016, for zooarch]
+* Here we present an alternative approach using ENM
+  * Whole region, at multiple climate snapshots
+  * Independent of archaeobot. and pal. clim. data, so can verify and compare
+
 ## Background
+
+* Epipal./Neo. plant-based economies
+  * More than the "founder crops"; 
+  * More than food;
+  * Regionally and temporally diverse
+  * ...so we model lots of species!
+* Regional ecological reconstructions
+  * See CSEAS (AEA-prep) presentation
+  * Figure: comparisons
+* Archaeological applications of ENM
+  * Review: Franklin et al. 2015
+  * In SW Asia: Conolly et al. 2012, Collins et al. 2018
+  * Train with archaeo, or train with present-day?
+    * In ENM literature, many sceptical about fore- and hind-casting
+    * Archaeo data a golden opportunity here
 
@@
 ## Discussion
+
+## Acknowledgments {.appendix}

### 2024-02-08 analysis/analysis.qmd
Adjust title

@@
 ---
-title: Palaeodistribution of crop progenitors and other plant resources over the Pleistocene–Holocene transition in Southwest Asia
+title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia
 date: today
@@
 library("dplyr")
+library("forcats")
 library("fs")
@@
     taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
-    include = n_occ > 50
+    #include = n_occ > 50
   ) |>
   select(-taxon_original) |>
-  group_by(include) |>
+  # group_by(include) |>
+  filter(n_occ > 50) |>
   arrange(-n_present) |>
@@
 # TODO: tidy
-rf_fit <- filter(rf_fits, taxon == "Triticum urartu") |>
+rf_fit <- filter(flora, taxon == "Triticum turgidum dicoccum") |>
   pull(fit)
@@
 
+# Order and relabel periods
+climate_periods <- rf_pred |>
+  st_get_dimension_values("period") |>
+  ordered(levels = c("LGM", "HS1", "BA", "YDS", "EH", "cur")) |>
+  recode(
+    "LGM" = "LGM (21 ka)",
+    "HS1" = "HS1 (17.0-14.7 ka)",
+    "BA"  = "BA (14.7-12.9 ka)",
+    "YDS" = "YDS (12.9-11.7 ka)",
+    "EH"  = "EH (11.7-8.326 ka)",
+    "cur" = "current"
+  )
+
+rf_pred <- st_set_dimensions(rf_pred, "period", values = climate_periods)
+
 ggplot() + 
-  geom_stars(data = st_crop(rf_pred, w_asia)) + 
-  facet_wrap(~period) +
-  scale_fill_batlow(reverse = TRUE)
+  geom_stars(data = rf_pred) +
+  facet_wrap(~period, ncol = 2) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(crs = latlong) +
+  labs(
+    title = "Triticum turgidum dicoccum",
+    fill = "Predicted occurrence",
+    x = NULL,
+    y = NULL
+  ) +
+  theme(
+    plot.title = element_text(face = "italic")
+  )
 ```

### 2024-10-25 analysis/analysis.qmd
Add missing dep

@@
 library("here")
+library("khroma")
 library("parsnip")
@@
   geom_stars(data = rf_pred) +
-  facet_wrap(~period, ncol = 2) +
+  facet_wrap(~period, ncol = 3) +
   scale_fill_batlow(reverse = TRUE) +
@@
     title = "Triticum turgidum dicoccum",
-    fill = "Predicted occurrence",
+    fill = NULL,
     x = NULL,

### 2024-11-27 analysis/analysis.qmd
Tidify model fitting and prediction
Leaving tuning out for now due to performance cost

@@
 ```{r dependencies}
+library("cowplot")
 library("dplyr")
@@
 library("ggplot2")
+library("ggrepel")
 library("ggspatial")
@@
 library("khroma")
-library("parsnip")
 library("progressr")
@@
 library("readr")
+library("recipes")
 library("rgbif")
 library("rpaleoclim")
+library("scales")
 library("sf")
@@
 library("tibble")
+library("tidymodels")
 library("tidyr")
@@
   gt() |>
+  # TODO: move synonyms to footnotes?
   cols_merge(
@@
 #   (and discard the background sample)
+# using a custom recipe step even?
 n_background <- 10000
 
-flora |>
+flora <- flora |>
   mutate(
-    occ = map(occ, mutate, present = TRUE),
-    bg_data = map(1:n(), \(i) { 
+    occ = map(occ, mutate, present = TRUE, weight = 1),
+    bg_data = map(occ, function(x, region, n) { 
       background_sample(
-        region = w_asia_land, 
-        N = n_background,
+        region = region, 
+        N = n,
         coord_x = "longitude", coord_y = "latitude",
-        present = FALSE
+        present = FALSE,
+        weight = (1 / n) * nrow(x) # weight presences and background equally
       )
-    }),
+    }, region = w_asia_land, n = n_background),
     occ = map2(occ, bg_data, bind_rows)
   ) |>
-  select(-bg_data) ->
-  flora
+  select(-bg_data)
 ```
@@
   mutate(
-    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x))))
+    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x)))),
+    # TODO: where is the extra geometry column coming from?
+    occ = map(occ, \(x) as_tibble(select(x, -geometry))),
+    # TODO: can this happen earlier?
+    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE))))
   ) ->
@@
 
-```{r model}
-rf_predictors <- function() {
-  c(any_of("present"), starts_with("bio"))
-}
+```{r prepare-niche-model-data}
+# Split training, test (for model assessment) and cross-validation sets (for
+# tuning)
+flora <- flora |>
+  mutate(
+    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
+    occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
+    occ_split = map(occ, \(x) initial_split(x, strata = present)),
+    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
+  )
 
-# Drop any rows with NA values in any of the predictor variables
-# TODO: Where do these actually come from? There's usually only a handful.
-# occ <- drop_na(occ, rf_predictors())
+# Data preprocessing recipe
+# Following recommended preprocessing steps for RF classification from:
+#   https://www.tmwr.org/pre-proc-table.html
+niche_recipe <- recipe(training(flora$occ_split[[1]])) |>
+  update_role(gbif_key:latitude, new_role = "ID") |>
+  update_role(present, new_role = "outcome") |>
+  update_role(starts_with("bio"), new_role = "predictor") |>
+  step_nzv(all_predictors())
+```
 
-# Model
-# Parameters adapted from the example of a down-sampled RF model given in
-#   https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
-# TODO: tune?
-
-rf_model <- rand_forest(mode = "classification", trees = 1000) |>
+```{r niche-models}
+# Random Forest Niche Model
+# With hyperparameters adapted from the example of a down-sampled RF model given 
+# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
+niche_model <- rand_forest(mode = "classification", trees = 1000) |>
   set_engine(
     "ranger", 
-    probability = TRUE,
-    sample.fraction = c(sum(as.logical(.y())) / n_background, 1)
+    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
+    probability = TRUE
   )
 
-# Fit to each taxon
-# TODO: parallelise
-flora |>
-  mutate(
-    fit = map(occ, \(occ) {
-      fit(rf_model, factor(present) ~ ., data = select(occ, rf_predictors()))
-    }, .progress = TRUE)
-  ) ->
-  flora
+# TODO: Or... tune parameters? But this is very compute-intensive.
+# niche_model <- rand_forest(
+#   mode = "classification", 
+#   mtry = tune(),
+#   trees = tune(),
+#   min_n = tune()
+# ) |>
+#   set_engine("ranger", probability = TRUE)
+
+niche_workflow <- workflow(niche_recipe, niche_model)
 ```
 
-The output of the model is probalistic.
-However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to habitat suitability.
+We follow @ValaviEtAl2021a in using down-sampling and their recommended hyperparameters to avoid overfitting.
+
+<!-- TODO: OR:
+Hyperparameter tuning is important to avoid over-fitting [@CITE].
+We tuned each model separately to maximise ROC AUC.
+-->
+
+```{r tune-niche-models}
+# Tune model (separately for each taxon)
+# TODO: would be nice but is extremely slow
+# flora |>
+#   slice_sample(n = 10) |>
+#   mutate(
+#     niche_tuning_grid = map(occ_split, function(data, recipe) {
+#       grid_regular(
+#         finalize(mtry(), bake(prep(recipe), training(data), all_predictors())),
+#         trees(),
+#         min_n(),
+#         levels = 5
+#       )
+#     }, recipe = niche_recipe, .progress = TRUE),
+#     niche_tuning = map2(niche_tuning_grid, occ_folds, function(grid, folds, wflow) {
+#       tune_grid(wflow, folds, grid = grid)
+#     }, wflow = niche_workflow, .progress = TRUE),
+#     niche_hparams = map(niche_tuning, select_best, metric = "roc_auc"),
+#     niche_model = map(niche_hparams, function(params, wflow) {
+#       finalize_workflow(wflow, params)
+#     }, wflow = niche_workflow)
+#   )
+```
+
+```{r fit-niche-models}
+# Fit best model to training data
+flora <- flora |>
+  mutate(
+    niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
+  )
+```
+
+The output of the model is probabilistic.
+However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to an estimate of habitat suitability.
 To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
@@
 ```{r thresholding}
-# TODO: thresholding
+# TODO: thresholding?
 # - Pick a threshold
@@
 # - Pick max SSS
+#
+# Or just stick with probabilities?
 ```
@@
 ## Results
+
+### Model assessment
+
+```{r niche_model_test}
+# Predict model on test data and assess performance
+# TODO: model performance assessments:
+# - [ ] R2
+# - [ ] Boyce index
+# - [x] AUC, ROC, AUC-ROC, AUC-PRG ?
+# - [ ] TSS
+# flora <- flora |>
+#   mutate(
+#     niche_test = map2(niche, occ_split, \(x, y) augment(x, testing(y))),
+#     niche_test_roc_auc = map_dbl(niche_test, function(x) {
+#       roc_auc(x, present, .pred_TRUE)$.estimate
+#     }),
+#     niche_test_accuracy = map_dbl(niche_test, function(x) {
+#       accuracy(x, present, .pred_class)$.estimate
+#     })
+#   )
+
+# TODO: filter to minimum accuracy? 0.85? 0.9? 0.95?
+```
+
+```{r fig-model-accuracy-vs-n_occ}
+#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
+flora |>
+  mutate(niche_test_fit = map(niche, collect_metrics)) |>
+  unnest(niche_test_fit) |>
+  filter(.metric == "accuracy") |> # TODO: or roc_auc?
+  mutate(label = if_else(.estimate < 0.8, taxon, NA_character_)) |>
+  ggplot(aes(n_occ, .estimate)) + 
+  geom_smooth(method = "glm", formula = y ~ log(x)) +
+  geom_point() + 
+  geom_text_repel(aes(label = label), fontface = "italic") +
+  scale_x_log10(name = "Occurrences") +
+  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
+  theme_half_open() 
+```
 
@@
 ```{r predict}
-# TODO: tidy
-rf_fit <- filter(flora, taxon == "Triticum turgidum dicoccum") |>
-  pull(fit)
-rf_fit <- rf_fit[[1]]
+# TODO: structure this way from the beginning?
+# Thanks to https://stackoverflow.com/questions/65183306/stars-package-how-to-define-additional-dimensions-based-on-an-attribute-filena
+climate_pred <- do.call(c, map(climate, st_redimension, name = "variable")) |>
+  st_redimension(name = "period") |>
+  split("variable")
 
-# TODO: Maybe combine stars and collapse period into dimension first?
-rf_pred <- map(climate, predict, model = rf_fit, type = "prob")
-rf_pred <- map(rf_pred, \(x) x[".pred_TRUE"])
-rf_pred <- do.call(c, rf_pred)
-rf_pred <- st_redimension(rf_pred, name = "period")
+# TODO: very high memory usage
+flora <- flora |>
+  mutate(
+    prediction = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)), 
+                     obj = climate_pred)
+  )
+```
+
+```{r fig-niche-prediction}
+niche_prediction <- filter(flora, taxon == "Linum bienne")$prediction[[1]]
 
 # Order and relabel periods
-climate_periods <- rf_pred |>
+climate_periods <- niche_prediction |>
   st_get_dimension_values("period") |>
@@
     "YDS" = "YDS (12.9-11.7 ka)",
-    "EH"  = "EH (11.7-8.326 ka)",
+    "EH"  = "EH (11.7-8.3 ka)",
     "cur" = "current"
@@
 
-rf_pred <- st_set_dimensions(rf_pred, "period", values = climate_periods)
+niche_prediction <- st_set_dimensions(niche_prediction, "period", values = climate_periods)
 
 ggplot() + 
-  geom_stars(data = rf_pred) +
+  geom_stars(data = select(niche_prediction, prediction.TRUE.) > 0.5) +
   facet_wrap(~period, ncol = 3) +
-  scale_fill_batlow(reverse = TRUE) +
+  # scale_fill_batlow(reverse = TRUE) +
   coord_sf(crs = latlong) +
   labs(
-    title = "Triticum turgidum dicoccum",
+    title = "Linum bienne",
     fill = NULL,
@@
   ) +
+  theme_map() +
   theme(
@@
 
-### Model assessment & verification
+### Comparison to archaeological occurrences
 
-```{r todo}
-# TODO: model performance assessment:
-# - R2
-# - Boyce index
-# - AUC, ROC, AUC-ROC, AUC-PRG ?
-# - TSS
+## Discussion
+
+```{r sum-pred-eh}
+flora |>
+  mutate(pred_eh = map(fit, \(x) predict(climate$EH, model = x, type = "prob"))) ->
+  x
+
+pred_eh <- x$pred_eh
+names(pred_eh) <- x$taxon
+pred_eh <- do.call(c, map(pred_eh, select, .pred_TRUE))
+pred_eh <- merge(pred_eh, name = "taxon")
+sum_pred_eh <- st_apply(pred_eh, c("x", "y"), sum, na.rm = TRUE)
+```
+
+```{r fig-sum-pred-eh}
+#| fig-cap: predicted species richness for the early Holocene
+ggplot() + 
+  geom_stars(data = sum_pred_eh) +
+  scale_fill_batlow(reverse = FALSE) +
+  coord_sf(crs = latlong) +
+  labs(
+    title = "Predicted species richness",
+    subtitle = "Early Holocene (11.7–8.326 ka)",
+    fill = NULL,
+    x = NULL,
+    y = NULL
+  )
+
+saveRDS(pred_eh, "pred_eh.Rdata")
+```
+
+```{r fig-pred-observed-eh}
 
@@
 
-## Discussion
-
 ## Acknowledgments {.appendix}

### 2024-11-27 analysis/analysis.qmd
Remove redundant chunk

@@
 
-```{r niche_model_test}
-# Predict model on test data and assess performance
-# TODO: model performance assessments:
-# - [ ] R2
-# - [ ] Boyce index
-# - [x] AUC, ROC, AUC-ROC, AUC-PRG ?
-# - [ ] TSS
-# flora <- flora |>
-#   mutate(
-#     niche_test = map2(niche, occ_split, \(x, y) augment(x, testing(y))),
-#     niche_test_roc_auc = map_dbl(niche_test, function(x) {
-#       roc_auc(x, present, .pred_TRUE)$.estimate
-#     }),
-#     niche_test_accuracy = map_dbl(niche_test, function(x) {
-#       accuracy(x, present, .pred_class)$.estimate
-#     })
-#   )
-
-# TODO: filter to minimum accuracy? 0.85? 0.9? 0.95?
-```
-
 ```{r fig-model-accuracy-vs-n_occ}
@@
 
-# TODO: very high memory usage
+# NOTE: very high memory usage
 flora <- flora |>

### 2024-11-28 analysis/analysis.qmd
Update predictor datasets

@@
 library("dplyr")
+library("dotenv")
 library("forcats")
@@
 library("rgbif")
-library("rpaleoclim")
 library("scales")
@@
 
-w_asia <- st_bbox(c(xmin = 30, xmax = 50, ymin = 25, ymax = 40), crs = 4326)
+w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)
 s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
@@
 neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")
+
+climate_periods <- fct_inorder(c(
+  lgm = "LGM (21 ka)",
+  hs1 = "HS1 (17.0-14.7 ka)",
+  ba  = "BA (14.7-12.9 ka)",
+  yds = "YDS (12.9-11.7 ka)",
+  eh  = "EH (11.7-8.3 ka)",
+  cur = "current"
+))
 ```
@@
 
-```{r data-climate}
-raw_data("paleoclim") |>
-  dir_ls() |>
-  map(dir_ls, glob = "*.tif") |>
-  map(read_stars) |>
-  map(\(x) set_names(x, path_ext_remove(path_file(names(x))))) ->
-  climate
+Bioclimatic data – the standard, and what allows us to do hindcasting.
 
-# Tidy up names
-names(climate) |>
-  basename() |>
-  str_remove(regex("chelsa_", ignore_case = TRUE)) |>
-  str_split_i("_", 1) ->
-  names(climate)
+Topographic data derived from a DEM:
+
+* Aspect and slope – proxies for solar radiation, but only if high resolution [@AustinVanNiel2011; @LeempoelEtAl2015]
+* TWI – proxy for soil moisture, important for arid environments [@KopeckyCizkova2010; @CamposEtAl2016; DiVirgilioEtAl2018]
+
+Soil data – important for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017] but not all variables from SoilGrids are reliable: so we just use soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth) [@MillerEtAl2024].
+
+```{r data-terrain}
+# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
+if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
+  message("Redownloading DEM data...")
+  source(here("data", "data-dem.R"))
+}
+
+dem <- read_stars(derived_data("dem", "srtm_15plus_w_asia.tif"))
+names(dem) <- "elevation"
+
+# TODO: slope
+# TODO: aspect
+# TODO: twi
 ```
 
-```{r training-climate}
+```{r data-soil}
+# Reacquire derived_data/soilgrids if needed:
+if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
+  message("Redownloading SoilGrids data...")
+  source(here("data", "data-soilgrids.R"))
+}
+
+soil <- map(dir_ls(derived_data("soilgrids")), read_stars)
+names(soil) <- str_split_i(basename(names(soil)), "_", 2)
+soil <- do.call(c, soil)
+```
+
+```{r data-climate}
+# Reacquire derived_data/paleoclim if needed:
+if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
+  message("Redownloading Paleoclim data...")
+  source(here("data", "data-paleoclim.R"))
+}
+
+climate <- read_stars(
+  derived_data(
+    "paleoclim", 
+    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
+  ),
+  along = NA
+)
+
+names(climate) <- names(climate_periods)
+
+climate <- st_redimension(climate, name = "period") |>
+  split("band")
+```
+
+```{r data-environment}
+# TODO: resample to common res
+# TODO: reproject
+# TODO: combine
+# TODO: fill NAs?
+```
+
+```{r extract-predictor-data}
 flora |> 

### 2024-11-28 analysis/data/data-dem.R

@@ -1,0 +1,7 @@
+# data-dem.R
+# Retrieve DEM <http://www.paleoclim.org/> data for W. Asia from OpenTopography
+# NOTE: relies on an external API (last run 2024-11-28)
+library("elevatr")
+
+srtm_15plus <- get_elev_raster(st_sf(st_as_sfc(w_asia)), prj = 4326, src = "srtm15plus")
+writeRaster(srtm_15plus, derived_data("dem", "srtm_15plus_w_asia.tif"))

### 2024-11-28 analysis/data/data-paleoclim.R

@@ -1,0 +1,27 @@
+# data-paleoclim.R
+# Retrieve Paleoclim <http://www.paleoclim.org/> data for W. Asia
+# NOTE: relies on an external API (last run 2024-11-28)
+library(rpaleoclim)
+
+paleoclim <- map(names(climate_periods), function(period, region) {
+  paleoclim(
+    period = period,
+    resolution = "2_5m",
+    region = region,
+    cache_path = derived_data("paleoclim", "raw")
+  )
+}, region = w_asia)
+
+# Fix extent of CHELSA 'current' layer
+terra::ext(paleoclim[[6]]) <- terra::ext(paleoclim[[1]])
+
+names(paleoclim) <- names(climate_periods)
+
+# Write files
+walk2(
+  paleoclim,
+  derived_data("paleoclim", paste0("paleoclim_", names(paleoclim), "_2_5m_w_asia.tif")),
+  terra::writeRaster,
+  datatype = "FLT4S",
+  overwrite = TRUE
+)

### 2024-11-28 analysis/data/data-soilgrids.R

@@ -1,0 +1,29 @@
+# data-soilgrids.R
+# Retrieve SoilGrids <https://www.isric.org/explore/soilgrids> data for W. Asia
+# Based on https://git.wur.nl/isric/soilgrids/soilgrids.notebooks/-/blob/master/markdown/webdav_from_R_terra.md
+# NOTE: relies on an external API (last run 2024-11-28)
+library("terra")
+
+soilgrids_vars <- c("clay", "sand", "silt", "phh2o")
+
+soilgrids <- map(soilgrids_vars, function(var, region) {
+  base_url <- "https://files.isric.org/soilgrids/latest/data_aggregated/1000m/"
+  layer <- paste(var, "0-5cm", "mean", sep="_") # layer of interest
+
+  soilgrid <- rast(paste0("/vsicurl/", base_url, var, "/", layer, "_1000.tif"))
+  bbox <- project(ext(region), from = st_crs(region)$wkt, to = crs(soilgrid))
+  window(soilgrid) <- bbox
+
+  return(soilgrid)
+}, region = w_asia)
+
+names(soilgrids) <- soilgrids_vars
+
+walk2(
+  soilgrids,
+  derived_data("soilgrids", paste0("soilgrids_", names(soilgrids), "_0-5cm_mean_w_asia.tif")),
+  writeRaster,
+  datatype = "FLT4S",
+  overwrite = TRUE
+)
+

### 2024-12-23 analysis/analysis.qmd
Add topographic predictors

@@
 plan("multisession")
+```
+
+```{r setup-whitebox}
 ```
@@
 sf_use_s2(use_s2)
+```
+
+### Predictor data
+
+Bioclimatic data – the standard, and what allows us to do hindcasting.
+
+Topographic data derived from a DEM:
+
+* Aspect and slope – proxies for solar radiation, but only if high resolution [@AustinVanNiel2011; @LeempoelEtAl2015]
+* TWI – proxy for soil moisture, important for arid environments [@KopeckyCizkova2010; @CamposEtAl2016; DiVirgilioEtAl2018]
+
+Soil data – important for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017] but not all variables from SoilGrids are reliable: so we just use soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth) [@MillerEtAl2024].
+
+```{r data-terrain}
+# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
+if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
+  message("Redownloading DEM data...")
+  source(here("data", "data-dem.R"))
+}
+
+# TODO: derived terrain products are two big to commit
+# TODO: does TWI actually make sense at this resolution? Could start with (much)
+# higher res, then upscale?
+terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
+                                            "srtm_15plus_w_asia_slope.tif",
+                                            "srtm_15plus_w_asia_aspect.tif",
+                                            "srtm_15plus_w_asia_wetness.tif")))
+names(terrain) <- c("elevation", "slope", "aspect", "wetness")
+```
+
+```{r data-soil}
+# Reacquire derived_data/soilgrids if needed:
+if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
+  message("Redownloading SoilGrids data...")
+  source(here("data", "data-soilgrids.R"))
+}
+
+soil <- map(dir_ls(derived_data("soilgrids")), read_stars)
+names(soil) <- str_split_i(basename(names(soil)), "_", 2)
+soil <- do.call(c, soil)
+```
+
+```{r data-climate}
+# Reacquire derived_data/paleoclim if needed:
+if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
+  message("Redownloading Paleoclim data...")
+  source(here("data", "data-paleoclim.R"))
+}
+
+climate <- read_stars(
+  derived_data(
+    "paleoclim", 
+    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
+  ),
+  along = NA
+)
+
+names(climate) <- names(climate_periods)
+
+climate <- st_redimension(climate, name = "period") |>
+  split("band")
+```
+
+```{r data-environment}
+# For time-invariant predictors, collapse attributes into dimension and 
+# duplicate to create a dummy period dimension
+# TODO: maybe need to downsample first?
+terrain <- st_redimension(terrain, name = "variable")
+terrain <- do.call(c, rep(list(terrain), length(climate_periods)))
+names(terrain) <- names(climate_periods)
+terrain <- st_redimension(terrain, name = "period") # TODO: why *so* slow?
+
+soil <- st_redimension(soil, name = "variable")
+soil <- do.call(c, rep(list(soil), length(climate_periods)))
+names(soil) <- names(climate_periods)
+soil <- st_redimension(soil, name = "period")
+
+# Resample to common grid
+terrain <- st_warp(terrain, climate, use_gdal = TRUE, method = "cubicspline")
+soil <- st_warp(soil, climate, use_gdal = TRUE, method = "cubicspline")
+
+# TODO: combine and split variable
+# TODO: fill NAs?
 ```
@@
   select(-bg_data)
-```
-
-### Predictor data
-
-Bioclimatic data – the standard, and what allows us to do hindcasting.
-
-Topographic data derived from a DEM:
-
-* Aspect and slope – proxies for solar radiation, but only if high resolution [@AustinVanNiel2011; @LeempoelEtAl2015]
-* TWI – proxy for soil moisture, important for arid environments [@KopeckyCizkova2010; @CamposEtAl2016; DiVirgilioEtAl2018]
-
-Soil data – important for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017] but not all variables from SoilGrids are reliable: so we just use soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth) [@MillerEtAl2024].
-
-```{r data-terrain}
-# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
-if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
-  message("Redownloading DEM data...")
-  source(here("data", "data-dem.R"))
-}
-
-dem <- read_stars(derived_data("dem", "srtm_15plus_w_asia.tif"))
-names(dem) <- "elevation"
-
-# TODO: slope
-# TODO: aspect
-# TODO: twi
-```
-
-```{r data-soil}
-# Reacquire derived_data/soilgrids if needed:
-if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
-  message("Redownloading SoilGrids data...")
-  source(here("data", "data-soilgrids.R"))
-}
-
-soil <- map(dir_ls(derived_data("soilgrids")), read_stars)
-names(soil) <- str_split_i(basename(names(soil)), "_", 2)
-soil <- do.call(c, soil)
-```
-
-```{r data-climate}
-# Reacquire derived_data/paleoclim if needed:
-if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
-  message("Redownloading Paleoclim data...")
-  source(here("data", "data-paleoclim.R"))
-}
-
-climate <- read_stars(
-  derived_data(
-    "paleoclim", 
-    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
-  ),
-  along = NA
-)
-
-names(climate) <- names(climate_periods)
-
-climate <- st_redimension(climate, name = "period") |>
-  split("band")
-```
-
-```{r data-environment}
-# TODO: resample to common res
-# TODO: reproject
-# TODO: combine
-# TODO: fill NAs?
 ```

### 2024-12-23 analysis/data/data-dem.R

@@ -1,7 +1,36 @@
 # data-dem.R
-# Retrieve DEM <http://www.paleoclim.org/> data for W. Asia from OpenTopography
+# Retrieve DEM <http://www.paleoclim.org/> data for W. Asia
+# Requires OpenTopography <https://opentopography.org> API key set in the
+#   environment and system installation of WhiteBox Tools
+#   <https://www.whiteboxgeo.com/>
+#
 # NOTE: relies on an external API (last run 2024-11-28)
 library("elevatr")
+library("whitebox")
 
 srtm_15plus <- get_elev_raster(st_sf(st_as_sfc(w_asia)), prj = 4326, src = "srtm15plus")
 writeRaster(srtm_15plus, derived_data("dem", "srtm_15plus_w_asia.tif"))
+
+dem <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia.tif"))
+
+# Slope
+slope <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia_slope.tif"))
+wbt_slope(dem, slope, compress_rasters = TRUE)
+
+# Aspect
+aspect <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia_aspect.tif"))
+wbt_aspect(dem, aspect, compress_rasters = TRUE)
+
+# Topographic wetness
+sca <- wbt_file_path(fs::file_temp("sca", ex = "tif"))
+wbt_flow_accumulation_full_workflow(
+  dem,
+  out_dem = fs::file_temp("dem", ext = "tif"),
+  out_pntr = fs::file_temp("pntr", ext = "tif"),
+  out_accum = sca,
+  out_type = "Specific Contributing Area",
+  compress_rasters = TRUE
+)
+
+wetness <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia_wetness.tif"))
+wbt_wetness_index(sca, slope, wetness, compress_rasters = TRUE)

### 2025-01-03 analysis/paper.qmd
Refactor as manuscript

--- 2024-12-23 analysis/analysis.qmd
@@
 ---
-title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia
+title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21–8.3 ka
 date: today
@@
   email: joeroe@hey.com
-  affiliation: University of Bern
+  affiliations:
+    - name: University of Bern
+      country: Switzerland
+    - name: University of Copenhagen
+      country: Denmark
+  corresponding: true
 - name: Amaia Arranz-Otaegui
@@
   orcid: 0000-0002-5091-6426
-  affiliation: University of the Basque Country
+  affiliation: 
+    - name: University of the Basque Country
+      country: Spain
 abstract: |
-  This paper presents the first comprehensive reconstructions of the prehistoric 
-  ranges of [X] plant species known to have been gathered or cultivated by early
-  agricultural societies in Southwest Asia, including the progenitors of the 
-  first crops. We used machine learning to train an ecological niche model (ENM)
-  of each species based on its present-day distribution in relation to climate 
-  and environmental variables. Predictions of the potential ranges of these 
-  species at key stages of the Pleistocene–Holocene transition could then be
-  derived from these models using palaeoclimate 
+  This paper presents the first comprehensive reconstructions of the 
+  palaeodistributions of [X] plant species known to have been gathered or 
+  cultivated by early agricultural societies in Southwest Asia, including the 
+  progenitors of the first crops. We used machine learning to train an 
+  ecological niche model (ENM) of each species based on its present-day 
+  distribution in relation to climate and environmental variables. Predictions 
+  of the potential ranges of these species at key stages of the 
+  Pleistocene–Holocene transition could then be derived from these models using 
+  hindcast data from palaeoclimate simulations.
   Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
@@
   choices.
-format: html
+format: 
+  elsevier-pdf:
+    keep-tex: true
+    cite-method: citeproc
+    journal:
+      name: Open Quaternary
+      formatting: review
 bibliography: references.bib
+execute: 
+  echo: false
+  eval: false
 ---
 
-```{r dependencies}
+* [ ] Finish analysis:
+  - [ ] Add terrain, soil
+  - [ ] Find a resolution that's workable
+  - [ ] Do thresholding?
+- [ ] First draft:
+  - [ ] Introduction
+  - [ ] Background
+    - [ ] Biogeography
+    - [ ] ENM
+  - [x] Methods & Materials
+    - [x] Occurrence data
+    - [x] Predictor data
+    - [x] Random Forest model
+  - [ ] Results
+    - [ ] Model assessment
+    - [ ] Hindcasting
+  - [ ] Discussion
+    - [ ] General trends
+    - [ ] Case studies
+  - [ ] Conclusion
+- [ ] Figures
+  - [ ] tbl-predictors
+  - [ ] tbl-climate-periods
+* [ ] Appendix with all hindcast predictions
+- [ ] References & copyediting
+- [ ] Final proofread
+
+```{r dependencies, eval=TRUE, message=FALSE}
 library("cowplot")
@@
 library("dotenv")
+library("english")
 library("forcats")
@@
 
-```{r define-periods}
+```{r define-periods, eval=TRUE}
 archaeo_periods <- tribble(
@@
 
-## Introduction
+# Introduction
 
 * The first farming societies had an ecological context
-* Subsistence is understood (largely) through archaeobotany; ecological context from archaeobotany, palynology, palaeoclimate, etc.
+* Subsistence is understood (largely) through archaeobotany and zooarchaeology; ecological context from environmental archaeology, palynology, palaeoclimate records, etc.
   * But these have a variety of biases (human selection, taphonomy, etc.)
   * And at the end of the day only represent specific places – interpolation to the entire region is not straightforward
-  * Overlap makes it difficult to see where human choices depart from the environmental background [cf. Martin et al. 2016, for zooarch]
+  * Overlap makes it difficult to see where human choices depart from the environmental background [cf. @MartinEtAl2016]
+  * Or to fully contextualise subsistence strategies and shifts in strategies in response to environmental shifts [e.g. @YaworskyEtAl2023].
 * Here we present an alternative approach using ENM
@@
 
-## Background
+# Background
 
-* Epipal./Neo. plant-based economies
+* The transition to agriculture in West Asia was...
+
+## Biogeography and agricultural origins
+
+* Has always been important in study of agricultural origins
+  * Historically: Vavilov, Pumpelly & Childe
+  * Genetic studies tell us origin points, but not ranges
+* Epipal./Neo. plant-based economies were diverse
   * More than the "founder crops"; 
-  * More than food;
+  * More than food
+  * (In archaebot., not all intentionally collected)
   * Regionally and temporally diverse
   * ...so we model lots of species!
-* Regional ecological reconstructions
+* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
   * See CSEAS (AEA-prep) presentation
   * Figure: comparisons
-* Archaeological applications of ENM
-  * Review: Franklin et al. 2015
-  * In SW Asia: Conolly et al. 2012, Collins et al. 2018
-  * Train with archaeo, or train with present-day?
+
+## Ecological niche modelling in archaeology
+
+* Review: @FranklinEtAl2015
+* In SW Asia: @ConollyEtAl2012 (first use ever??), @CollinsEtAl2018
+* Varying uses: to reconstruct human niches (esp.
+in Palaeolithic), to reconstruct spread of (esp.
+domestic) species from archaeological record, or to reconstruct wild ranges from modern occurrences and palaeoclimate (our approach)
+* Caveats
   * In ENM literature, many sceptical about fore- and hind-casting
@@
 
-## Methods and materials
+# Methods and materials
 
@@
 
-### Predictor data
+## Occurrence data
 
-Bioclimatic data – the standard, and what allows us to do hindcasting.
-
-Topographic data derived from a DEM:
-
-* Aspect and slope – proxies for solar radiation, but only if high resolution [@AustinVanNiel2011; @LeempoelEtAl2015]
-* TWI – proxy for soil moisture, important for arid environments [@KopeckyCizkova2010; @CamposEtAl2016; DiVirgilioEtAl2018]
-
-Soil data – important for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017] but not all variables from SoilGrids are reliable: so we just use soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth) [@MillerEtAl2024].
-
-```{r data-terrain}
-# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
-if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
-  message("Redownloading DEM data...")
-  source(here("data", "data-dem.R"))
-}
-
-# TODO: derived terrain products are two big to commit
-# TODO: does TWI actually make sense at this resolution? Could start with (much)
-# higher res, then upscale?
-terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
-                                            "srtm_15plus_w_asia_slope.tif",
-                                            "srtm_15plus_w_asia_aspect.tif",
-                                            "srtm_15plus_w_asia_wetness.tif")))
-names(terrain) <- c("elevation", "slope", "aspect", "wetness")
-```
-
-```{r data-soil}
-# Reacquire derived_data/soilgrids if needed:
-if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
-  message("Redownloading SoilGrids data...")
-  source(here("data", "data-soilgrids.R"))
-}
-
-soil <- map(dir_ls(derived_data("soilgrids")), read_stars)
-names(soil) <- str_split_i(basename(names(soil)), "_", 2)
-soil <- do.call(c, soil)
-```
-
-```{r data-climate}
-# Reacquire derived_data/paleoclim if needed:
-if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
-  message("Redownloading Paleoclim data...")
-  source(here("data", "data-paleoclim.R"))
-}
-
-climate <- read_stars(
-  derived_data(
-    "paleoclim", 
-    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
-  ),
-  along = NA
-)
-
-names(climate) <- names(climate_periods)
-
-climate <- st_redimension(climate, name = "period") |>
-  split("band")
-```
-
-```{r data-environment}
-# For time-invariant predictors, collapse attributes into dimension and 
-# duplicate to create a dummy period dimension
-# TODO: maybe need to downsample first?
-terrain <- st_redimension(terrain, name = "variable")
-terrain <- do.call(c, rep(list(terrain), length(climate_periods)))
-names(terrain) <- names(climate_periods)
-terrain <- st_redimension(terrain, name = "period") # TODO: why *so* slow?
-
-soil <- st_redimension(soil, name = "variable")
-soil <- do.call(c, rep(list(soil), length(climate_periods)))
-names(soil) <- names(climate_periods)
-soil <- st_redimension(soil, name = "period")
-
-# Resample to common grid
-terrain <- st_warp(terrain, climate, use_gdal = TRUE, method = "cubicspline")
-soil <- st_warp(soil, climate, use_gdal = TRUE, method = "cubicspline")
-
-# TODO: combine and split variable
-# TODO: fill NAs?
-```
-
-### Occurrence data
-
-```{r read-archaeo_flora}
-archaeo_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv"))
-```
-
-```{r tidy-archaeo_flora}
-archaeo_flora |>
+```{r data-flora}
+# Flora present at archaeological sites
+archaeo_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv")) |>
   rename(taxon_group = "taxon", taxon = "taxon_detail") |>
@@
   archaeo_flora
-```
 
-```{r summarise-archaeo_flora}
 n_assemb_neolithic <- archaeo_flora |> 
@@
   nrow()
-```
 
-To generate the list of taxa modelled in this study, we ...
-
-* Taxa present at more than three Neolithic sites, according to the database of our previous study [@ArranzOtaeguiRoe2023]
-* Identified at least to species (and matchable to the GBIF Backbone Taxonomy)
-* Have at least 100 occurrences in West Asia (meeting the criteria below) in the GBIF database
-* Excluded domestic species, replacing them with their wild progenitor, where known and present in the GBIF database
-
-We also considered the number modern occurrences available for each taxon.
-ENMs are known to have reasonable predictive power even with small training samples (i.e. less than a hundred observations, [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008]).
-
-```{r generate-flora}
+# Flora to be modelled
 archaeo_flora |>
@@
   ) |>
-  filter(n_present > 2) ->
+  filter(n_present > 2) |>
+  # Exclude taxa not identified to species
+  filter(!str_ends(taxon, "sp.")) ->
   flora
@@
 
-```{r tidy-flora-taxonomy}
-# Normalise species names and match to the GBIF Backbone Taxonomy
-# Excluding taxa not identified to species
-flora |>
-  filter(!str_ends(taxon, "sp.")) |>
+We consider X<!-- TODO --> distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic sites in West Asia, according to our previous study [@ArranzOtaeguiRoe2023].
+<!-- TODO: something about not distinguishing between source of plant remains? -->
+
+```{r data-flora-taxonomy}
+# - Normalise species names and match to the GBIF Backbone Taxonomy
+# - Recode domesticates as their wild progenitors.
+# - Add other wild progenitors of wheats
+flora <- flora |>
   mutate(
@@
     taxon_original = taxon,
-    taxon = map_chr(taxon, gbif_accepted_name)
-  ) ->
-  flora
-```
-
-```{r tidy-flora-domesticates}
-# Recode domesticates as their wild progenitors.
-# Add other wild progenitors of wheats
-flora |> 
-  mutate(taxon = case_match(
+    taxon = map_chr(taxon, gbif_accepted_name),
+    taxon = case_match(
       taxon,
@@
       .default = taxon
-  )) |>
+    )
+  ) |>
   bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
-                             "Aegilops tauschii"))) ->
-  flora
+                             "Aegilops tauschii")))
 ```
 
-```{r fetch-occ-count}
+Taxonomic names were resolves to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
+<!-- TODO: still relevent?
+aren't we modelling species?
+-->So for example occurrences for "Bolboschoenus" included all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.
+Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.
+
+## Occurrence data
+
+```{r data-occ-count}
 flora |>
@@
 flora <- filter(flora, n_occ > 50)
-# TODO: Temporary measure.
 # Why are there 47k occurrences of Avena sterilis??
@@
 
-Occurrence data was obtained from GBIF using the GBIF occurrence API.
-This deals with synonyms etc.
-(using the GBIF Backbone Taxonomy as a standard), so e.g.
-retrieving data for "Bolboschoenus" includes all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.
+Occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package rgbif [@ChamberlainBoettiger2017; @rgbif].
+Although ENMs have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded X<!-- TODO --> taxa with less than fifty recorded occurrences in our study region.
+We also excluded one taxon (*Avena sterilis*) with over 47,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.
 
@@
 
-<!-- TODO: following @ValaviEtAl2022, more accurate to call pseudo-absences
-  "background samples" -->
-Occurrence data only tells us where a species is present;
-there is no definitive information on where the species is *not* found.
-We therefore need to generate random background points or "pseudo-absences" to feed to the model.
-There are several ways to do this. 
-We follow the advice of Barbet-Massin et al.
-(2012) for regression-based species distribution models and use a large (~10000) random sample of points, weighted equally against the presences in the regression.
-@ValaviEtAl2022 recommends using a very large background sample for random forest models.
+<!-- TODO: occurrence tidying -->
 
@@
 
+<!-- TODO: fix informal tone -->
+Occurrence data only tells us where a species is present;
+there is rarely definitive information on where the species is *not* found.
+We therefore need to generate random background points or "pseudo-absences" to feed to the model.
+There are several ways to do this. 
+We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (:10000) random sample of points, weighted equally against the presences in the regression.
+@ValaviEtAl2022 also recommend using a very large background sample for random forest models.
+
+## Predictor data
+
+We modelled the occurrence of species as a function of <!-- TODO -->X spatial predictor variables (@tbl-predictors).
+These included:
+
+```{r data-climate}
+# Reacquire derived_data/paleoclim if needed:
+if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
+  message("Redownloading Paleoclim data...")
+  source(here("data", "data-paleoclim.R"))
+}
+
+climate <- read_stars(
+  derived_data(
+    "paleoclim", 
+    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
+  ),
+  along = NA
+)
+
+names(climate) <- names(climate_periods)
+
+climate <- st_redimension(climate, name = "period") |>
+  split("band")
+```
+
+* Sixteen 'bioclimatic' variables derived from monthly temperature and precipitation values, following standard practice for species distribution models [@HijmansEtAl2005].
+Contemporary bioclimatic predictor data for West Asia was extracted from the global CHELSA dataset [@KargerEtAl2017], which predicts temperature and precipitation from downscaled general circulation model output at 1 km resolution.
+
+```{r data-terrain}
+# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
+if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
+  message("Redownloading DEM data...")
+  source(here("data", "data-dem.R"))
+}
+
+# TODO: derived terrain products are two big to commit
+# TODO: does TWI actually make sense at this resolution? Could start with (much)
+# higher res, then upscale?
+terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
+                                            "srtm_15plus_w_asia_slope.tif",
+                                            "srtm_15plus_w_asia_aspect.tif",
+                                            "srtm_15plus_w_asia_wetness.tif")))
+names(terrain) <- c("elevation", "slope", "aspect", "wetness")
+```
+
+* Terrain aspect and slope, which at high resolution perform well as proxies for solar radiation when modelling plant occurrence [@AustinVanNiel2011; @LeempoelEtAl2015]; and the topographic wetness index (TWI), which serves as a proxy for soil moisture and is particularly important in modelling arid environments [@KopeckyCizkova2010; @CamposEtAl2016; @DiVirgilioEtAl2018].
+All three were derived from the SRTM15+ digital elevation model using algorithms from WhiteboxTools [@Lindsay2016].
+
+```{r data-soil}
+# Reacquire derived_data/soilgrids if needed:
+if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
+  message("Redownloading SoilGrids data...")
+  source(here("data", "data-soilgrids.R"))
+}
+
+soil <- map(dir_ls(derived_data("soilgrids")), read_stars)
+names(soil) <- str_split_i(basename(names(soil)), "_", 2)
+soil <- do.call(c, soil)
+```
+
+* Edaphic data from SoilGrids [@HenglEtAl2014; @HenglEtAl2017], which improves model performance for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017].
+Based on a recent assessment of the reliability of SoilGrids data for species distribution modelling [@MillerEtAl2024], we used a subset of four variables relating to soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth).
+
+```{r data-environment}
+# For time-invariant predictors, collapse attributes into dimension and 
+# duplicate to create a dummy period dimension
+# TODO: maybe need to downsample first?
+terrain <- st_redimension(terrain, name = "variable")
+terrain <- do.call(c, rep(list(terrain), length(climate_periods)))
+names(terrain) <- names(climate_periods)
+terrain <- st_redimension(terrain, name = "period") # TODO: why *so* slow?
+
+soil <- st_redimension(soil, name = "variable")
+soil <- do.call(c, rep(list(soil), length(climate_periods)))
+names(soil) <- names(climate_periods)
+soil <- st_redimension(soil, name = "period")
+
+# Resample to common grid
+terrain <- st_warp(terrain, climate, use_gdal = TRUE, method = "cubicspline")
+soil <- st_warp(soil, climate, use_gdal = TRUE, method = "cubicspline")
+
+# TODO: combine and split variable
+# TODO: fill NAs?
+```
+
+Predictor data was transformed to the same projection system (WGS84 / UTM 37 N) and a common resolution of X<!-- TODO --> km.
+
+For hindcasting, we used reconstructed bioclimatic data for `r english(nrow(climate_periods) - 1)` key periods (@tbl-climate-periods) generated from downscaled paleoclimate simulations from the HadCM3 general circulation model [@BrownEtAl2018].
+Terrain and soil predictors were held constant, since reconstructions of these variables in the past are not available at sufficient scale.
+It is not likely that either macroscale topography or soil characteristics have altered significantly over the period of time considered here, so we assume that this does not degrade model performance, and may in fact benefit it by providing 'anchoring' predictors that are independent of climate change.
+
 ```{r extract-predictor-data}
@@
 
-TODO: Soil, terrain (incl. TWI, slope?)
+## Random Forest
 
-### Random Forest
-
-There are many methods for ENM.
-Random Forest is one of the best performing, providing it is appropriately parameterised to account for the class imbalance between presence and background samples (balanced random forest) [@ValaviEtAl2021a; @ValaviEtAl2021b].
-
-We used the R package ranger [@WrightZiegler2017].
-
-```{r prepare-niche-model-data}
+```{r niche-data}
 # Split training, test (for model assessment) and cross-validation sets (for
@@
 
-```{r niche-models}
+Ecological niche modelling is a classification problem that can be approached with a wide range of statistical methods.
+A substantial literature exists on the relatively performance of these approaches and their respective parameterisations [reviewed in @ValaviEtAl2022].
+Random Forest, a widely-used machine learning algorithm, is amongst the best performing approaches for presence-only species distribution models, providing it is appropriately parameterised to account for the class imbalance between presence and background samples [@ValaviEtAl2021; @ValaviEtAl2022].
+For our application, it also has the advantage of requiring little to no manual parameter tuning to achieve good predictive results, which makes it easier to model a larger numbers of taxa.
+
+```{r niche-model}
 # Random Forest Niche Model
@@
 niche_workflow <- workflow(niche_recipe, niche_model)
-```
 
-We follow @ValaviEtAl2021a in using down-sampling and their recommended hyperparameters to avoid overfitting.
-
-<!-- TODO: OR:
-Hyperparameter tuning is important to avoid over-fitting [@CITE].
-We tuned each model separately to maximise ROC AUC.
--->
-
-```{r tune-niche-models}
 # Tune model (separately for each taxon)
-# TODO: would be nice but is extremely slow
+# TODO: questioning - would be nice but is extremely slow
 # flora |>
@@
 
+For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our X<!--TODO--> predictor variables (@tbl-predictors).
+We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
+To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
+Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (3/4) and test (1/4) partitions.
+
+<!-- TODO: OR:
+Hyperparameter tuning is important to avoid over-fitting [@CITE].
+We tuned each model separately to maximise ROC AUC.
+-->
+
 ```{r fit-niche-models}
@@
 
+<!-- TODO: cut? -->
 The output of the model is probabilistic.
@@
 To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
-We select the threshold value for each model individually, using MaxSSS (as recommended by [@LuiEtAl2013])
+We select the threshold value for each model individually, using MaxSSS [as recommended by @LiuEtAl2013].
 This also makes it possible to analyses the predictions together as an assemblage.
@@
 
-## Results
+# Results
 
-### Model assessment
+## Model assessment
 
@@
 
-### Modelled ecological niches
+* Modelled ecological niches on current data
 
-* On current data
-* Sensitivity to climate fluctuations
+## Hindcasting
 
-### Hindcasting
+Sensitivity to climate fluctuations?
 
@@
 
-### Comparison to archaeological occurrences
-
-## Discussion
+* Comparison to archaeological occurrences
 
@@
 
+# Discussion
+
+* General trends:
+  * Quantified sensitivity of plant ranges to climate change
+  * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
+  * How could are reconstructed ranges at predicting archaeological assemblages? (+Implications)
+* Interesting individual case studies:
+  * Wheat progenitors
+  * ???
+
 ```{r fig-pred-observed-eh}
@@
 
-## Acknowledgments {.appendix}
+# Conclusion
+
+# References {.appendix}

### 2025-01-03 analysis/paper.tex
Refactor as manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  review]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Open Quaternary}

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources during
the transition to agriculture in West Asia, 21--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

\affiliation[1]{organization={University of
Bern},country={Switzerland},countrysep={,},postcodesep={}}
\affiliation[2]{organization={University of
Copenhagen},country={Denmark},countrysep={,},postcodesep={}}
\affiliation[3]{organization={University of the Basque
Country},country={Spain},countrysep={,},postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of {[}X{]} plant species known to have been gathered
or cultivated by early agricultural societies in Southwest Asia,
including the progenitors of the first crops. We used machine learning
to train an ecological niche model (ENM) of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using hindcast data from palaeoclimate simulations.
Species ranges were on average {[}X\%{]} {[}larger\textbar smaller{]} in
the Early Holocene compared to today, indicating {[}\ldots{]}. The
modelled ranges predict the observed occurrence of these species on
archaeological sites with {[}low\textbar medium\textbar high{]}
accuracy. The regional ubiquity of species in the archaeological record
is {[}not{]} correlated with the predicted size of its range and the
diversity of archaeobotanical assemblages is {[}not{]} correlated with
the predicted diversity of its environs. This indicates that trends in
taxonomic composition of the archaeobotanical record is {[}not{]} likely
to be influenced by environmental change and species turnover, not just,
as is often assumed, human economic choices.
\end{abstract}

\end{frontmatter}

\begin{itemize}
\tightlist
\item[$\square$]
  Finish analysis:

  \begin{itemize}
  \tightlist
  \item[$\square$]
    Add terrain, soil
  \item[$\square$]
    Find a resolution that's workable
  \item[$\square$]
    Do thresholding?
  \end{itemize}
\item[$\square$]
  First draft:

  \begin{itemize}
  \tightlist
  \item[$\square$]
    Introduction
  \item[$\square$]
    Background

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\square$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials

    \begin{itemize}
    \tightlist
    \item[$\boxtimes$]
      Occurrence data
    \item[$\boxtimes$]
      Predictor data
    \item[$\boxtimes$]
      Random Forest model
    \end{itemize}
  \item[$\square$]
    Results

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Model assessment
    \item[$\square$]
      Hindcasting
    \end{itemize}
  \item[$\square$]
    Discussion

    \begin{itemize}
    \tightlist
    \item[$\square$]
      General trends
    \item[$\square$]
      Case studies
    \end{itemize}
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-climate-periods
  \end{itemize}
\item[$\square$]
  Appendix with all hindcast predictions
\item[$\square$]
  References \& copyediting
\item[$\square$]
  Final proofread
\end{itemize}

\section{Introduction}\label{introduction}

\begin{itemize}
\tightlist
\item
  The first farming societies had an ecological context
\item
  Subsistence is understood (largely) through archaeobotany and
  zooarchaeology; ecological context from environmental archaeology,
  palynology, palaeoclimate records, etc.

  \begin{itemize}
  \tightlist
  \item
    But these have a variety of biases (human selection, taphonomy,
    etc.)
  \item
    And at the end of the day only represent specific places --
    interpolation to the entire region is not straightforward
  \item
    Overlap makes it difficult to see where human choices depart from
    the environmental background (cf. Martin et al., 2016)
  \item
    Or to fully contextualise subsistence strategies and shifts in
    strategies in response to environmental shifts (e.g. Yaworsky et
    al., 2023).
  \end{itemize}
\item
  Here we present an alternative approach using ENM

  \begin{itemize}
  \tightlist
  \item
    Whole region, at multiple climate snapshots
  \item
    Independent of archaeobot. and pal. clim. data, so can verify and
    compare
  \end{itemize}
\end{itemize}

\section{Background}\label{background}

\begin{itemize}
\tightlist
\item
  The transition to agriculture in West Asia was\ldots{}
\end{itemize}

\subsection{Biogeography and agricultural
origins}\label{biogeography-and-agricultural-origins}

\begin{itemize}
\tightlist
\item
  Has always been important in study of agricultural origins

  \begin{itemize}
  \tightlist
  \item
    Historically: Vavilov, Pumpelly \& Childe
  \item
    Genetic studies tell us origin points, but not ranges
  \end{itemize}
\item
  Epipal./Neo. plant-based economies were diverse

  \begin{itemize}
  \tightlist
  \item
    More than the ``founder crops'';
  \item
    More than food
  \item
    (In archaebot., not all intentionally collected)
  \item
    Regionally and temporally diverse
  \item
    \ldots so we model lots of species!
  \end{itemize}
\item
  Regional ecological reconstructions generally rely on the `expert
  interpolation' (or what do they call it with isoscapes?) method

  \begin{itemize}
  \tightlist
  \item
    See CSEAS (AEA-prep) presentation
  \item
    Figure: comparisons
  \end{itemize}
\end{itemize}

\subsection{Ecological niche modelling in
archaeology}\label{ecological-niche-modelling-in-archaeology}

\begin{itemize}
\tightlist
\item
  Review: (Franklin et al., 2015)
\item
  In SW Asia: (Conolly et al., 2012) (first use ever??), (Collins et
  al., 2018)
\item
  Varying uses: to reconstruct human niches (esp.~in Palaeolithic), to
  reconstruct spread of (esp.~domestic) species from archaeological
  record, or to reconstruct wild ranges from modern occurrences and
  palaeoclimate (our approach)
\item
  Caveats

  \begin{itemize}
  \tightlist
  \item
    In ENM literature, many sceptical about fore- and hind-casting
  \item
    Archaeo data a golden opportunity here
  \end{itemize}
\end{itemize}

\section{Methods and materials}\label{methods-and-materials}

\subsection{Occurrence data}\label{occurrence-data}

We consider X distinct taxa (Table~\ref{tbl-occ-count}) - all the
identifiable species known to be present at more than three Neolithic
sites in West Asia, according to our previous study (Arranz-Otaegui and
Roe, 2023).

Taxonomic names were resolves to the canonical form specified in the
GBIF Backbone Taxonomy (\textbf{GBIFSecretariat2023?}). So for example
occurrences for ``Bolboschoenus'' included all species and subspecies,
including specimens described as \emph{Bolboschoenus sp.},
\emph{Bolboschoenus maritimus} and \emph{Scirpus maritimus}. Domestic
species meeting our inclusion criteria were substituted with their wild
progenitor(s), where known.

\subsection{Occurrence data}\label{occurrence-data-1}

Occurrence data was obtained from the Global Biodiversity Information
Facility (GBIF) using via its application programming interface and the
R package rgbif (Chamberlain et al., 2024; Chamberlain and Boettiger,
2017). Although ENMs have reasonable predictive power even with small
training samples (Hernandez et al., 2006; Stockwell and Peterson, 2002;
Wisz et al., 2008), we excluded X taxa with less than fifty recorded
occurrences in our study region. We also excluded one taxon (\emph{Avena
sterilis}) with over 47,000 occurrences, as this would have been
computationally prohibitive and we were uncertain what account for such
a disproportionately high number of records.

Occurrence data only tells us where a species is present; there is
rarely definitive information on where the species is \emph{not} found.
We therefore need to generate random background points or
``pseudo-absences'' to feed to the model. There are several ways to do
this. We follow the advice of (Barbet-Massin et al., 2012) for
regression-based species distribution models and use a large (:10000)
random sample of points, weighted equally against the presences in the
regression. (Valavi et al., 2022) also recommend using a very large
background sample for random forest models.

\subsection{Predictor data}\label{predictor-data}

We modelled the occurrence of species as a function of X spatial
predictor variables (\textbf{?@tbl-predictors}). These included:

\begin{itemize}
\tightlist
\item
  Sixteen `bioclimatic' variables derived from monthly temperature and
  precipitation values, following standard practice for species
  distribution models (Hijmans et al., 2005). Contemporary bioclimatic
  predictor data for West Asia was extracted from the global CHELSA
  dataset (Karger et al., 2017), which predicts temperature and
  precipitation from downscaled general circulation model output at 1 km
  resolution.
\end{itemize}

\begin{itemize}
\tightlist
\item
  Terrain aspect and slope, which at high resolution perform well as
  proxies for solar radiation when modelling plant occurrence (Austin
  and Van Niel, 2011; Leempoel et al., 2015); and the topographic
  wetness index (TWI), which serves as a proxy for soil moisture and is
  particularly important in modelling arid environments (Campos et al.,
  2016; Di Virgilio et al., 2018; Kopecký and Čížková, 2010). All three
  were derived from the SRTM15+ digital elevation model using algorithms
  from WhiteboxTools (Lindsay, 2016).
\end{itemize}

\begin{itemize}
\tightlist
\item
  Edaphic data from SoilGrids (Hengl et al., 2017, 2014), which improves
  model performance for plants (Dubuis et al., 2013; Mod et al., 2016;
  Velazco et al., 2017). Based on a recent assessment of the reliability
  of SoilGrids data for species distribution modelling (Miller et al.,
  2024), we used a subset of four variables relating to soil texture
  (clay, silt, sand) and pH at the surface (0-5 cm depth).
\end{itemize}

Predictor data was transformed to the same projection system (WGS84 /
UTM 37 N) and a common resolution of X km.

For hindcasting, we used reconstructed bioclimatic data for key periods
(\textbf{?@tbl-climate-periods}) generated from downscaled paleoclimate
simulations from the HadCM3 general circulation model (Brown et al.,
2018). Terrain and soil predictors were held constant, since
reconstructions of these variables in the past are not available at
sufficient scale. It is not likely that either macroscale topography or
soil characteristics have altered significantly over the period of time
considered here, so we assume that this does not degrade model
performance, and may in fact benefit it by providing `anchoring'
predictors that are independent of climate change.

\subsection{Random Forest}\label{random-forest}

Ecological niche modelling is a classification problem that can be
approached with a wide range of statistical methods. A substantial
literature exists on the relatively performance of these approaches and
their respective parameterisations (reviewed in Valavi et al., 2022).
Random Forest, a widely-used machine learning algorithm, is amongst the
best performing approaches for presence-only species distribution
models, providing it is appropriately parameterised to account for the
class imbalance between presence and background samples (Valavi et al.,
2022, 2021). For our application, it also has the advantage of requiring
little to no manual parameter tuning to achieve good predictive results,
which makes it easier to model a larger numbers of taxa.

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our X predictor variables
(\textbf{?@tbl-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' (Wright and Ziegler, 2017) and the
`tidymodels' (\textbf{tidymodels?}) framework for data preprocessing and
model selection. To avoid overfitting, we follow (Valavi et al., 2021)
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (3/4) and test (1/4) partitions.

The output of the model is probabilistic. However, this should not be
understood as an actual probability of occurrence (\textbf{CITE?}), but
more akin to an estimate of habitat suitability. To simplify
interpretation, we can convert these predictions into binary
presence/absence maps, a process called ``thresholding''. We select the
threshold value for each model individually, using MaxSSS (as
recommended by Liu et al., 2013). This also makes it possible to
analyses the predictions together as an assemblage.

\section{Results}\label{results}

\subsection{Model assessment}\label{model-assessment}

\begin{itemize}
\tightlist
\item
  Modelled ecological niches on current data
\end{itemize}

\subsection{Hindcasting}\label{hindcasting}

Sensitivity to climate fluctuations?

\begin{itemize}
\tightlist
\item
  Comparison to archaeological occurrences
\end{itemize}

\section{Discussion}\label{discussion}

\begin{itemize}
\tightlist
\item
  General trends:

  \begin{itemize}
  \tightlist
  \item
    Quantified sensitivity of plant ranges to climate change
  \item
    Crop progenitors saw range contractions just before the onset of
    agriculture? (Moreso than other wild resources??)
  \item
    How could are reconstructed ranges at predicting archaeological
    assemblages? (+Implications)
  \end{itemize}
\item
  Interesting individual case studies:

  \begin{itemize}
  \tightlist
  \item
    Wheat progenitors
  \item
    ???
  \end{itemize}
\end{itemize}

\section{Conclusion}\label{conclusion}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-ArranzOtaeguiRoe2023}
Arranz-Otaegui, A., Roe, J., 2023. Revisiting the concept of the
{``{Neolithic Founder Crops}''} in southwest {Asia}. Vegetation History
and Archaeobotany. \url{https://doi.org/10.1007/s00334-023-00917-1}

\bibitem[\citeproctext]{ref-AustinVanNiel2011}
Austin, M.P., Van Niel, K.P., 2011. Improving species distribution
models for climate change studies: Variable selection and scale. Journal
of Biogeography 38, 1--8.
\url{https://doi.org/10.1111/j.1365-2699.2010.02416.x}

\bibitem[\citeproctext]{ref-BarbetMassinEtAl2012}
Barbet-Massin, M., Jiguet, F., Albert, C.H., Thuiller, W., 2012.
Selecting pseudo-absences for species distribution models: How, where
and how many?: {How} to use pseudo-absences in niche modelling? Methods
Ecol. Evol. 3, 327--338.
\url{https://doi.org/10.1111/j.2041-210X.2011.00172.x}

\bibitem[\citeproctext]{ref-BrownEtAl2018}
Brown, J.L., Hill, D.J., Dolan, A.M., Carnaval, A.C., Haywood, A.M.,
2018. {PaleoClim}, high spatial resolution paleoclimate surfaces for
global land areas. Sci Data 5, 180254.
\url{https://doi.org/10.1038/sdata.2018.254}

\bibitem[\citeproctext]{ref-CamposEtAl2016}
Campos, V.E., Cappa, F.M., Viviana, F.M., Giannoni, S.M., 2016. Using
remotely sensed data to model suitable habitats for tree species in a
desert environment. Journal of Vegetation Science 27, 200--210.
\url{https://doi.org/10.1111/jvs.12328}

\bibitem[\citeproctext]{ref-rgbif}
Chamberlain, S., Barve, V., Mcglinn, D., Oldoni, D., Desmet, P.,
Geffert, L., Ram, K., 2024.
\href{https://CRAN.R-project.org/package=rgbif}{Rgbif: Interface to the
global biodiversity information facility API}.

\bibitem[\citeproctext]{ref-ChamberlainBoettiger2017}
Chamberlain, S., Boettiger, C., 2017.
\href{https://doi.org/10.7287/peerj.preprints.3304v1}{R python, and ruby
clients for GBIF species occurrence data}. PeerJ PrePrints.

\bibitem[\citeproctext]{ref-CollinsEtAl2018}
Collins, C., Asouti, E., Grove, M., Kabukcu, C., Bradley, L.,
Chiverrell, R., 2018. Understanding resource choice at the transition
from foraging to farming: {An} application of palaeodistribution
modelling to the {Neolithic} of the {Konya Plain}, south-central
{Anatolia}, {Turkey}. J. Archaeol. Sci. 96, 57--72.
\url{https://doi.org/10.1016/j.jas.2018.02.003}

\bibitem[\citeproctext]{ref-ConollyEtAl2012}
Conolly, J., Manning, K., Colledge, S., Dobney, K., Shennan, S., 2012.
Species distribution modelling of ancient cattle from early {Neolithic}
sites in {SW Asia} and {Europe}. Holocene 22, 997--1010.
\url{https://doi.org/10.1177/0959683612437871}

\bibitem[\citeproctext]{ref-DiVirgilioEtAl2018}
Di Virgilio, G., Wardell-Johnson, G.W., Robinson, T.P., Temple-Smith,
D., Hesford, J., 2018. Characterising fine-scale variation in plant
species richness and endemism across topographically complex, semi-arid
landscapes. Journal of Arid Environments 156, 59--68.
\url{https://doi.org/10.1016/j.jaridenv.2018.04.005}

\bibitem[\citeproctext]{ref-DubuisEtAl2013}
Dubuis, A., Giovanettina, S., Pellissier, L., Pottier, J., Vittoz, P.,
Guisan, A., 2013. Improving the prediction of plant species distribution
and community composition by adding edaphic to topo-climatic variables.
Journal of Vegetation Science 24, 593--606.
\url{https://doi.org/10.1111/jvs.12002}

\bibitem[\citeproctext]{ref-FranklinEtAl2015}
Franklin, J., Potts, A.J., Fisher, E.C., Cowling, R.M., Marean, C.W.,
2015. Paleodistribution modeling in archaeology and paleoanthropology.
Quat. Sci. Rev. 110, 1--14.
\url{https://doi.org/10.1016/j.quascirev.2014.12.015}

\bibitem[\citeproctext]{ref-HenglEtAl2014}
Hengl, T., de Jesus, J.M., MacMillan, R.A., Batjes, N.H., Heuvelink,
G.B.M., Ribeiro, E., Samuel-Rosa, A., Kempen, B., Leenaars, J.G.B.,
Walsh, M.G., Gonzalez, M.R., 2014. {SoilGrids1km--global} soil
information based on automated mapping. PLoS One 9, e105992.
\url{https://doi.org/10.1371/journal.pone.0105992}

\bibitem[\citeproctext]{ref-HenglEtAl2017}
Hengl, T., Mendes de Jesus, J., Heuvelink, G.B.M., Ruiperez Gonzalez,
M., Kilibarda, M., Blagotić, A., Shangguan, W., Wright, M.N., Geng, X.,
Bauer-Marschallinger, B., Guevara, M.A., Vargas, R., MacMillan, R.A.,
Batjes, N.H., Leenaars, J.G.B., Ribeiro, E., Wheeler, I., Mantel, S.,
Kempen, B., 2017. {SoilGrids250m}: {Global} gridded soil information
based on machine learning. PLoS One 12, e0169748.
\url{https://doi.org/10.1371/journal.pone.0169748}

\bibitem[\citeproctext]{ref-HernandezEtAl2006}
Hernandez, P.A., Graham, C.H., Master, L.L., Albert, D.L., 2006. The
effect of sample size and species characteristics on performance of
different species distribution modeling methods. Ecography 29, 773--785.
\url{https://doi.org/10.1111/j.0906-7590.2006.04700.x}

\bibitem[\citeproctext]{ref-HijmansEtAl2005}
Hijmans, R.J., Cameron, S.E., Parra, J.L., Jones, P.G., Jarvis, A.,
2005. Very high resolution interpolated climate surfaces for global land
areas. Int. J. Climatol. 25, 1965--1978.
\url{https://doi.org/10.1002/joc.1276}

\bibitem[\citeproctext]{ref-KargerEtAl2017}
Karger, D.N., Conrad, O., Böhner, J., Kawohl, T., Kreft, H., Soria-Auza,
R.W., Zimmermann, N.E., Linder, H.P., Kessler, M., 2017. Climatologies
at high resolution for the earth's land surface areas. Sci Data 4,
170122. \url{https://doi.org/10.1038/sdata.2017.122}

\bibitem[\citeproctext]{ref-KopeckyCizkova2010}
Kopecký, M., Čížková, Š., 2010. Using topographic wetness index in
vegetation ecology: Does the algorithm matter? Appl. Veg. Sci. 13,
450--459. \url{https://doi.org/10.1111/j.1654-109X.2010.01083.x}

\bibitem[\citeproctext]{ref-LeempoelEtAl2015}
Leempoel, K., Parisod, C., Geiser, C., Daprà, L., Vittoz, P., Joost, S.,
2015. Very high-resolution digital elevation models: Are multi-scale
derived variables ecologically relevant? Methods in Ecology and
Evolution 6, 1373--1383. \url{https://doi.org/10.1111/2041-210X.12427}

\bibitem[\citeproctext]{ref-Lindsay2016}
Lindsay, J.B., 2016. Whitebox {GAT}: {A} case study in geomorphometric
analysis. Comput. Geosci. 95, 75--84.
\url{https://doi.org/10.1016/j.cageo.2016.07.003}

\bibitem[\citeproctext]{ref-LiuEtAl2013}
Liu, C., White, M., Newell, G., 2013. Selecting thresholds for the
prediction of species occurrence with presence-only data. Journal of
Biogeography 40, 778--789. \url{https://doi.org/10.1111/jbi.12058}

\bibitem[\citeproctext]{ref-MartinEtAl2016}
Martin, L., Edwards, Y.H., Roe, J., Garrard, A., 2016. Faunal turnover
in the {Azraq Basin}, eastern {Jordan} 28,000 to 9,000 cal {BP},
signalling climate change and human impact. Quaternary Research 86,
200--219. \url{https://doi.org/10.1016/j.yqres.2016.07.001}

\bibitem[\citeproctext]{ref-MillerEtAl2024}
Miller, T., Blackwood, C.B., Case, A.L., 2024. Assessing the utility of
{SoilGrids250} for biogeographic inference of plant populations. Ecology
and Evolution 14, e10986. \url{https://doi.org/10.1002/ece3.10986}

\bibitem[\citeproctext]{ref-ModEtAl2016}
Mod, H.K., Scherrer, D., Luoto, M., Guisan, A., 2016. What we use is not
what we know: Environmental predictors in plant distribution models.
Journal of Vegetation Science 27, 1308--1322.
\url{https://doi.org/10.1111/jvs.12444}

\bibitem[\citeproctext]{ref-StockwellPeterson2002}
Stockwell, D.R.B., Peterson, A.T., 2002. Effects of sample size on
accuracy of species distribution models. Ecological Modelling 148,
1--13. \url{https://doi.org/10.1016/S0304-3800(01)00388-X}

\bibitem[\citeproctext]{ref-ValaviEtAl2021}
Valavi, R., Elith, J., Lahoz-Monfort, J.J., Guillera-Arroita, G., 2021.
Modelling species presence-only data with random forests. Ecography 44,
1731--1742. \url{https://doi.org/10.1111/ecog.05615}

\bibitem[\citeproctext]{ref-ValaviEtAl2022}
Valavi, R., Guillera-Arroita, G., Lahoz-Monfort, J.J., Elith, J., 2022.
Predictive performance of presence-only species distribution models: A
benchmark study with reproducible code. Ecological Monographs 92,
e01486. \url{https://doi.org/10.1002/ecm.1486}

\bibitem[\citeproctext]{ref-VelazcoEtAl2017}
Velazco, S.J.E., Galvão, F., Villalobos, F., De Marco Júnior, P., 2017.
Using worldwide edaphic data to model plant species niches: {An}
assessment at a continental extent. PLoS One 12, e0186025.
\url{https://doi.org/10.1371/journal.pone.0186025}

\bibitem[\citeproctext]{ref-WiszEtAl2008}
Wisz, M.S., Hijmans, R.J., Li, J., Peterson, A.T., Graham, C.H., Guisan,
A., Group, N.P.S.D.W., 2008. Effects of sample size on the performance
of species distribution models. Diversity and Distributions 14,
763--773. \url{https://doi.org/10.1111/j.1472-4642.2008.00482.x}

\bibitem[\citeproctext]{ref-WrightZiegler2017}
Wright, M.N., Ziegler, A., 2017. Ranger: {A Fast Implementation} of
{Random Forests} for {High Dimensional Data} in {C}++ and {R}. Journal
of Statistical Software 77, 1--17.
\url{https://doi.org/10.18637/jss.v077.i01}

\bibitem[\citeproctext]{ref-YaworskyEtAl2023}
Yaworsky, P.M., Hussain, S.T., Riede, F., 2023. Climate-driven habitat
shifts of high-ranked prey species structure {Late Upper Paleolithic}
hunting. Scientific Reports 13, 4238.
\url{https://doi.org/10.1038/s41598-023-31085-x}

\end{CSLReferences}

\end{document}

### 2025-01-03 analysis/paper.qmd
Add ENM background

@@
   * Genetic studies tell us origin points, but not ranges
+* Important to e.g.
+  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2017; @MartinEtAl2025]
+  * Reconstruct sequences of domestication [@YeomansEtAl2017]
 * Epipal./Neo. plant-based economies were diverse
@@
 
-* Review: @FranklinEtAl2015
-* In SW Asia: @ConollyEtAl2012 (first use ever??), @CollinsEtAl2018
-* Varying uses: to reconstruct human niches (esp.
-in Palaeolithic), to reconstruct spread of (esp.
-domestic) species from archaeological record, or to reconstruct wild ranges from modern occurrences and palaeoclimate (our approach)
-* Caveats
-  * In ENM literature, many sceptical about fore- and hind-casting
-  * Archaeo data a golden opportunity here
+<!-- TODO: more citations? -->
+Ecological niche modelling (ENM) or species distribution modelling (SDM) is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors.
+Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
+This model can then be used to predict the range of the organism in question either in the same or a different environment.
+@CITE suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
+Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.
+
+Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
+In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g.
+@BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
+This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
+Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
+These applications can be distinguihed from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.
+
+@FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
+One relevant early example is @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites in Europe and West Asia to map the evolving niche of cattle over the Pleistocene–Holocene transition.
+It has been used to model the availability of fauna exploited by humans at wider scales [e.g.
+@deAndresHerreroEtAl2018; @YaworksyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik?
+--> [@CollinsEtAl2018].
+Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].
+
+In the majority of studies to date (palaeo)ecological niche modelling has been applied to archaeological data in an 'inductive' fashion, i.e.
+faunal and botanical remains from ancient sites are used as the occurrence dataset for training a model using either past or present environmental data.
+However, both the zooarchaeological and archaeobotanical records are sparse and subject to a complex array of depositional, taphonomic and recovery biases factors that , many of which are not fully understood and/or cannot be corrected for.
+This means that while the archaeological attestation of the presence of a species might generally be relied upon, it is highly unlikely that its absence is representative of true past distributions.
+
+The alternative approach is to train the model using contemporary occurrence and environmental data and then use palaeoenvironmental data to 'hindcast' its predictions backwards in time.
+Like @FranklinEtAl2015, we view the hindcasting approach as more promising, because training datasets for both occurrences and environment are far more readily available, complete and reliable for the present than the past.
+There is some scepticism in the ecological niche modelling literature about the ability of such models to make accurate predictions in unknown environments (like the past) [@CITES?], 
+but here the hindcasting approach also presents an opportunity:
+it reserves archaeological occurrence data as an independent dataset that can be used to assess the retrodictive performance of the model.
+This possibily was suggested by @FranklinEtAl2015 but to our knowledge our study represents the first attempt to actually do so.
+
+The major practical limitation of the hindcasting approach is that it relies on spatially explicit, high resolution palaeoenvironmental surfaces with continuous coverage of the region and periods of interest.
+Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @YaworskyEtAl2023].
+In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].
+<!-- TODO: mention that GBIF, CHELSA, SoilGrids, WorldClim, DEMs, etc. are also great? -->
 

### 2025-01-03 analysis/paper.tex
Add ENM background

@@
 \item
+  Important to e.g.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Distinguish environmental from potentially anthropogenic change
+    (\textbf{MartinEtAl2017?}; \textbf{MartinEtAl2025?})
+  \item
+    Reconstruct sequences of domestication (\textbf{YeomansEtAl2017?})
+  \end{itemize}
+\item
   Epipal./Neo. plant-based economies were diverse
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Review: (Franklin et al., 2015)
-\item
-  In SW Asia: (Conolly et al., 2012) (first use ever??), (Collins et
-  al., 2018)
-\item
-  Varying uses: to reconstruct human niches (esp.~in Palaeolithic), to
-  reconstruct spread of (esp.~domestic) species from archaeological
-  record, or to reconstruct wild ranges from modern occurrences and
-  palaeoclimate (our approach)
-\item
-  Caveats
+Ecological niche modelling (ENM) or species distribution modelling (SDM)
+is widely used by ecologists to predict the geographic range of a
+species based on a set of environmental predictors. Essentially, it
+involves combining records of where an organism has been observed with
+environmental data (climate, topography, etc.) for those locations to
+model the range of environmental values at which that species -- its
+environmental niche. This model can then be used to predict the range of
+the organism in question either in the same or a different environment.
+(\textbf{CITE?}) suggests reserving the term `species distribution
+modelling' for when the method is used to recover the verifiable range
+of a species in a real and existing environment, and using `ecological
+niche modelling' as the broader term covering hypothetical or predictive
+applications -- a convention we follow here when referring to predictive
+or `hindcast' models of past ranges. Within this overarching framework,
+ecological niche modelling encompasses a wide range of applications and
+a variety of potential environmental predictors, modelling approaches,
+and methodologies, which we will not attempt to review here.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    In ENM literature, many sceptical about fore- and hind-casting
-  \item
-    Archaeo data a golden opportunity here
-  \end{itemize}
-\end{itemize}
+Ecological niche modelling has long been of interest to archaeologists
+as both a means of exploring the biological niche of humans and for
+reconstructing the past environments they inhabited (Franklin et al.,
+2015; \textbf{DavidPollyEronen2011?}). In the first sense, it has been
+used most extensively to model the range of humans and other hominin
+species (e.g. \textbf{BenitoEtAl2017?}; \textbf{YousefiEtAl2020?};
+\textbf{BanksEtAl2021?}; \textbf{YaworskyEtAl2024a?};
+\textbf{YaworskyEtAl2024b?}; \textbf{GuranEtAl2024?}), especially in the
+Palaeolithic. This overlaps with what archaeologists usually call
+generically `predictive modelling'
+(\textbf{VerhagenWhitley2020?})---more precisely `site distribution
+modelling'---which is essentially the same approach as (and often
+borrows methodologies from) ecological niche modelling but applied to
+the occurrence of archaeological sites. Here what is modelled is not
+strictly a biological niche alone, but also aspects of human geography,
+taphonomy, and archaeological visibility. These applications can be
+distinguihed from `palaeoecological niche modelling', where the object
+of model remains, as in ecology, a non-human biological niche.
+
+(Franklin et al., 2015) review palaeoecological niche modelling and
+advocate for its greater adoption in environmental archaeology. One
+relevant early example is (Conolly et al., 2012) used the occurrence of
+wild and domestic \emph{Bos} remains at prehistoric archaeological sites
+in Europe and West Asia to map the evolving niche of cattle over the
+Pleistocene--Holocene transition. It has been used to model the
+availability of fauna exploited by humans at wider scales (e.g.
+\textbf{deAndresHerreroEtAl2018?}; \textbf{YaworksyEtAl2023?}) and, in a
+West Asian context, of foraged plant resources in the landscape around
+the Neolithic site of XX (Collins et al., 2018). Modelling the spread of
+crops has been another significant archaeological application
+(\textbf{CremaEtAl?}).
+
+In the majority of studies to date (palaeo)ecological niche modelling
+has been applied to archaeological data in an `inductive' fashion,
+i.e.~faunal and botanical remains from ancient sites are used as the
+occurrence dataset for training a model using either past or present
+environmental data. However, both the zooarchaeological and
+archaeobotanical records are sparse and subject to a complex array of
+depositional, taphonomic and recovery biases factors that , many of
+which are not fully understood and/or cannot be corrected for. This
+means that while the archaeological attestation of the presence of a
+species might generally be relied upon, it is highly unlikely that its
+absence is representative of true past distributions.
+
+The alternative approach is to train the model using contemporary
+occurrence and environmental data and then use palaeoenvironmental data
+to `hindcast' its predictions backwards in time. Like (Franklin et al.,
+2015), we view the hindcasting approach as more promising, because
+training datasets for both occurrences and environment are far more
+readily available, complete and reliable for the present than the past.
+There is some scepticism in the ecological niche modelling literature
+about the ability of such models to make accurate predictions in unknown
+environments (like the past) (\textbf{CITES?}), but here the hindcasting
+approach also presents an opportunity: it reserves archaeological
+occurrence data as an independent dataset that can be used to assess the
+retrodictive performance of the model. This possibily was suggested by
+(Franklin et al., 2015) but to our knowledge our study represents the
+first attempt to actually do so.
+
+The major practical limitation of the hindcasting approach is that it
+relies on spatially explicit, high resolution palaeoenvironmental
+surfaces with continuous coverage of the region and periods of interest.
+Until recently, this has not been widely available for most
+applications, which is perhaps why only a minority of studies use it
+(cf. Yaworsky et al., 2023). In this study, we are able to take
+advantage of the increasing availability of high resolution, global
+palaeoclimate data derived from simulation experiments with general
+circulation models of climate (Brown et al., 2018;
+\textbf{BrownEtAl2020?}; \textbf{KargerEtAl2023?}).
 

### 2025-01-04 analysis/paper.qmd
Optimise model code

@@
 ---
-title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21–8.3 ka
+title: Biogeography of crop progenitors and wild plant resources in West Asia, 21–8.3 ka
 date: today
@@
   This paper presents the first comprehensive reconstructions of the 
-  palaeodistributions of [X] plant species known to have been gathered or 
-  cultivated by early agricultural societies in Southwest Asia, including the 
+  palaeodistributions of 56 plant species found regularly in association with
+  early agricultural archaeological sites in West Asia, including the 
   progenitors of the first crops. We used machine learning to train an 
-  ecological niche model (ENM) of each species based on its present-day 
-  distribution in relation to climate and environmental variables. Predictions 
-  of the potential ranges of these species at key stages of the 
-  Pleistocene–Holocene transition could then be derived from these models using 
-  hindcast data from palaeoclimate simulations.
-  Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
-  compared to today, indicating [...]. The modelled ranges predict the
-  observed occurrence of these species on archaeological sites with [low|medium|high]
-  accuracy. The regional ubiquity of species in the archaeological record is [not] 
-  correlated with the predicted size of its range and the diversity of
-  archaeobotanical assemblages is [not] correlated with the predicted diversity
-  of its environs. This indicates that trends in taxonomic composition of
-  the archaeobotanical record is [not] likely to be influenced by environmental 
-  change and species turnover, not just, as is often assumed, human economic 
-  choices.
+  ecological niche model of each species based on its present-day distribution 
+  in relation to climate and environmental variables. Predictions of the 
+  potential ranges of these species at key stages of the Pleistocene–Holocene 
+  transition could then be derived from these models using hindcast data from 
+  palaeoclimate simulations.
+  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
+  the Early Holocene compared to present conditions, indicating [...]. 
+  The modelled ranges predict the observed occurrence of these species on 
+  archaeological sites with [low|medium|high] accuracy. The regional ubiquity of 
+  species in the archaeological record is [not] correlated with the predicted 
+  size of its range and the diversity of archaeobotanical assemblages is [not] 
+  correlated with the predicted diversity of its environs. This indicates that 
+  trends in taxonomic composition of the archaeobotanical record is [not] likely 
+  to have been influenced by environmental change and species turnover, in
+  addition to human economic choices.
 format: 
@@
   echo: false
-  eval: false
 ---
 
-* [ ] Finish analysis:
-  - [ ] Add terrain, soil
-  - [ ] Find a resolution that's workable
-  - [ ] Do thresholding?
+* [x] Finish modelling
 - [ ] First draft:
@@
     - [ ] Biogeography
-    - [ ] ENM
+    - [x] ENM
   - [x] Methods & Materials
-    - [x] Occurrence data
-    - [x] Predictor data
-    - [x] Random Forest model
   - [ ] Results
@@
 - [ ] Figures
+  - [ ] tbl-occ-count
   - [ ] tbl-predictors
   - [ ] tbl-climate-periods
-* [ ] Appendix with all hindcast predictions
+- [ ] Clean up package code
+- [ ] Appendix with all hindcast predictions
 - [ ] References & copyediting
@@
 
-```{r dependencies, eval=TRUE, message=FALSE}
+```{r dependencies, message=FALSE}
 library("cowplot")
@@
 
-```{r setup-whitebox}
-```
-
 ```{r define-regions}
 latlong <- 4326
-utm37n <- 32637
+w_asia_albers <- "+proj=aea +lat_1=22.5 +lat_2=42.5 +lon_0=40"
 
@@
 
-```{r define-periods, eval=TRUE}
+```{r define-periods}
 archaeo_periods <- tribble(
@@
 
-climate_periods <- fct_inorder(c(
-  lgm = "LGM (21 ka)",
-  hs1 = "HS1 (17.0-14.7 ka)",
-  ba  = "BA (14.7-12.9 ka)",
-  yds = "YDS (12.9-11.7 ka)",
-  eh  = "EH (11.7-8.3 ka)",
-  cur = "current"
-))
+climate_periods <- tibble(
+  period = c("Last Glacial Maximum", "Heinrich Stadial 1", "Bølling-Allerød",
+             "Younger Dryas", "Early Holocene", "Current"),
+  code = c("lgm", "hs1", "ba", "yds", "eh", "cur"),
+  label = fct_inorder(c("LGM (21 ka)", "HS1 (17.0-14.7 ka)", "BA (14.7-12.9 ka)", 
+            "YDS (12.9-11.7 ka)", "EH (11.7-8.3 ka)", "current")),
+  start_bp = c(21000, 17000, 14700, 12900, 11700, NA),
+  end_bp = c(17000, 14700, 12900, 11700, 8300, NA)
+)
+```
+
+```{r define-parameters}
+# Minimum number of GBIF occurrences for modelling
+min_n_occ <- 50
+
+# Number of background samples per taxon
+n_background <- 10000
+
+# Raster resolution for prediction (meters)
+pred_res <- 5000
 ```
@@
 
-```{r read-basemap}
+```{r data-basemap, message=FALSE}
 # Turn off s2 while we deal with Natural Earth's dodgy geometries
@@
 
+```{r fig-region}
+#| fig-cap: Study region
+ggplot() + 
+  annotation_spatial(ne_land, fill = "#ffffff") +
+  layer_spatial(w_asia, fill = NA) +
+  coord_sf(crs = w_asia_albers)
+```
+
+<!-- TODO define study region and period -->
+
 ## Occurrence data
 
-```{r data-flora}
+```{r data-archaeo-flora}
 # Flora present at archaeological sites
-archaeo_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv")) |>
+archaeo_flora <- read_tsv(
+  raw_data("swasia_neolithic_flora.tsv"),
+  col_types = cols(
+    source = col_character(),
+    source_id = col_character(),
+    source_site_name = col_character(),
+    site_name = col_character(),
+    latitude = col_double(),
+    longitude = col_double(),
+    phase = col_character(),
+    phase_description = col_character(),
+    phase_code = col_character(),
+    age_start = col_integer(),
+    age_end = col_integer(),
+    taxon_source = col_character(),
+    n = col_double(),
+    prop = col_double(),
+    reference = col_character(),
+    taxon_detail = col_character(),
+    taxon = col_character(),
+    genus = col_character(),
+    family = col_character(),
+    category = col_character(),
+    founder_crop = col_character(),
+    edibility = col_character(),
+    grass_type = col_character(),
+    legume_type = col_character()
+  )
+) |>
   rename(taxon_group = "taxon", taxon = "taxon_detail") |>
   # Assign to archaeological periods
-  # TODO: refactor as function
-  # TODO: assign climate periods too
   mutate(
     age_mid = age_end + ((age_start - age_end) / 2),
-    period = cut(-age_mid, 
+    period = cut(
+      -age_mid, 
       breaks = -c(archaeo_periods$start_bp, 0), 
       labels = archaeo_periods$period,
-                 ordered_result = TRUE)
+      ordered_result = TRUE
+    ),
+    climate_period = cut(
+      -age_mid, 
+      breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
+      labels = climate_periods[climate_periods$code != "cur",]$code,
+      ordered_result = TRUE
+    )
   ) ->
@@
   nrow()
+```
 
-# Flora to be modelled
+We consider X<!-- TODO --> distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic sites in West Asia, according to our previous study [@ArranzOtaeguiRoe2023].
+<!-- TODO: something about not distinguishing between source of plant remains? -->
+
+```{r data-flora}
+if (!file_exists(derived_data("flora.tsv"))) {
+  message("Regenerating flora...")
+  # Subset of flora to be modelled
   archaeo_flora |>
@@
     flora
-```
 
-We consider X<!-- TODO --> distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic sites in West Asia, according to our previous study [@ArranzOtaeguiRoe2023].
-<!-- TODO: something about not distinguishing between source of plant remains? -->
-
-```{r data-flora-taxonomy}
-# - Normalise species names and match to the GBIF Backbone Taxonomy
-# - Recode domesticates as their wild progenitors.
-# - Add other wild progenitors of wheats
+  # Standardise taxonomy
   flora <- flora |>
+    # Normalise species names and match to the GBIF Backbone Taxonomy
     mutate(
@@
       taxon_original = taxon,
-    taxon = map_chr(taxon, gbif_accepted_name),
+      taxon = map_chr(taxon, gbif_accepted_name)
+    ) |>
+    # Recode domesticates as their wild progenitors.
+    mutate(
       taxon = case_match(
@@
     ) |>
-  bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
-                             "Aegilops tauschii")))
+    # Add wild progenitors of wheats
+    bind_rows(
+      tibble(
+        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
+        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
+      )
+    ) |>
+    # Deduplicate by canonical taxon
+    summarise(
+      taxon_detail = paste0("*", taxon_original[taxon_original != taxon], "*", collapse = ", "),
+      taxon_detail = na_if(taxon_detail, "**"),
+      n_present = max(n_present),
+      p_present = max(p_present),
+      .by = taxon
+    )
+
+  write_tsv(flora, derived_data("flora.tsv"))
+}
+
+flora <- read_tsv(
+  derived_data("flora.tsv"),
+  col_types = cols(
+    taxon = col_character(),
+    taxon_detail = col_character(),
+    n_present = col_double(),
+    p_present = col_double()
+  )
+)
 ```
@@
 Taxonomic names were resolves to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
-<!-- TODO: still relevent?
-aren't we modelling species?
--->So for example occurrences for "Bolboschoenus" included all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.
+So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
 Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.
@@
 ```{r data-occ-count}
-flora |>
-  group_by(taxon) |>
-  summarise(
-    taxon_original = list(c(taxon_original)),
-    n_present = max(n_present), # TODO: be more precise?
-    p_present = max(p_present)
-    ) |>
-  mutate(n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
+if (!file_exists(derived_data("flora_occ_count.tsv"))) {
+  message("Reacquiring GBIF occurrence counts...")
+  flora <- flora |>
+    mutate(
+      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                             hasCoordinate = TRUE, 
                                             hasGeospatialIssue = FALSE,
-                                               geometry = bbox_wkt(w_asia)))) ->
-  flora
+                                            geometry = bbox_wkt(w_asia)))
+    )
+  write_tsv(flora, derived_data("flora_occ_count.tsv"))
+}
+flora <- read_tsv(derived_data("flora_occ_count.tsv"))
 ```
@@
 ```{r tbl-occ-count}
-#| tbl-cap: "Species modelled in this study."
-# TODO: add to caption:
-# - with presence in Neolithic sites in West Asia, out of the X sites included in Y
-# - with number of reliably-georeferenced occurrences in West Asia recorded in GBIF
-# TODO: add count of species included?
+#| tbl-cap: "Recorded occurrence of flora considered in this study in Neolithic and contemporary West Asia"
 flora |>
   mutate(
-    taxon_detail = map_chr(taxon_original, paste, collapse = ", "),
-    taxon_detail = if_else(taxon == taxon_detail, NA_character_, taxon_detail),
-    #include = n_occ > 50
+    exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis"),
+    label = if_else(
+      !is.na(taxon_detail),
+      glue::glue("*{taxon}*\n\n<small>incl. {taxon_detail}</small>"),
+      paste0("*", taxon, "*")
+    )
   ) |>
-  select(-taxon_original) |>
-  # group_by(include) |>
-  filter(n_occ > 50) |>
+  select(label, n_present, p_present, n_occ, exclude) |>
   arrange(-n_present) |>
   gt() |>
-  # TODO: move synonyms to footnotes?
-  cols_merge(
-    c(taxon, taxon_detail),
-    rows = !is.na(taxon_detail),
-    pattern = "{1}<br><small>(incl. {2})</small>"
+  cols_label(
+    label = "Taxon",
+    n_present = "Neolithic",
+    n_occ = "Present"
   ) |>
+  tab_spanner("Occurrences", columns = starts_with("n_")) |>
+  cols_merge_n_pct(n_present, p_present) |>
+  cols_hide(exclude) |>
+  tab_footnote(
+    "Excluded from modelling due to sample size", 
+    cells_body(columns = c(n_occ), rows = exclude),
+    placement = "right"
+  ) |>
+  cols_align("left", c(label)) |>
   fmt_percent(p_present, decimals = 0) |>
-  cols_merge_n_pct(n_present, p_present) |>
-  tab_style(cell_text(style = "italic"), cells_body(taxon))
+  fmt_markdown(label) |>
+  sub_missing() |>
+  tab_options(footnotes.marks = "*")
 ```
 
+Georeferenced occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
+GBIF occurrences marked has having imprecise or duplicate coordinates were excluded from the training dataset, as were fossil records.
+Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` occurrences in West Asia, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
+We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.
+
 ```{r filter-flora}
-flora <- filter(flora, n_occ > 50)
-# Why are there 47k occurrences of Avena sterilis??
+flora <- filter(flora, n_occ > min_n_occ)
+# Why are there 50k occurrences of Avena sterilis??
 flora <- filter(flora, taxon != "Avena sterilis")
@@
 
-Occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package rgbif [@ChamberlainBoettiger2017; @rgbif].
-Although ENMs have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded X<!-- TODO --> taxa with less than fifty recorded occurrences in our study region.
-We also excluded one taxon (*Avena sterilis*) with over 47,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.
-
 ```{r fetch-occ}
-# TODO: Refactor
-gbif_data <- function(scientificName, ...) {
-  gbif_data <- rgbif::occ_data(scientificName = scientificName, ...)
-  gbif_data$data
-}
-
-flora |>
-  mutate(
-    occ = map(taxon, gbif_data, 
-              hasCoordinate = TRUE, hasGeospatialIssue = FALSE, 
-              geometry = w_asia, limit = 100000,
-              .progress = TRUE)
-  ) ->
-  flora
-```
-
-```{r tidy-occ}
 # TODO: further tidying. Ideas:
@@
 # - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
-flora |>
-  mutate(
+if (!file_exists(derived_data("flora_occ.tsv"))) {
+  message("Reacquiring GBIF occurrences...")
+  flora <- mutate(
+    flora,
+    occ = map(
+      taxon, gbif_data, 
+      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
+      limit = 100000, .progress = TRUE
+    ),
     occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
@@
     })
-  ) ->
-  flora
+  )
+  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
+}
+flora <- nest(read_tsv(derived_data("flora_occ.tsv")), occ = gbif_key:latitude)
 ```
@@
 ```{r fig-occ-map}
-flora |>
-  unnest(occ) |>
-  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
-  ggplot() + 
-  facet_wrap(vars(taxon)) +
-  layer_spatial(ne_land, fill = "#ffffff") +
-  geom_sf(shape = ".") +
-  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
-           ylim = w_asia[c("ymin", "ymax")],
-           crs = utm37n,
-           default_crs = latlong)
+# TODO: Can this be made presentable?
+#flora |>
+#  unnest(occ) |>
+#  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
+#  ggplot() + 
+#  facet_wrap(vars(taxon)) +
+#  layer_spatial(ne_land, fill = "#ffffff") +
+#  geom_sf(shape = ".") +
+#  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
+#           ylim = w_asia[c("ymin", "ymax")],
+#           crs = w_asia_albers,
+#           default_crs = latlong)
 ```
 
-<!-- TODO: occurrence tidying -->
-
-```{r generate-background}
-# TODO: maybe defer this to just before the model is actually fit ?
-#   (and discard the background sample)
-# using a custom recipe step even?
-n_background <- 10000
-
+```{r data-background}
+# TODO: slow - cache
 flora <- flora |>
@@
     occ = map(occ, mutate, present = TRUE, weight = 1),
-    bg_data = map(occ, function(x, region, n) { 
+    bg = map(occ, function(x, region, n) { 
       background_sample(
@@
     }, region = w_asia_land, n = n_background),
-    occ = map2(occ, bg_data, bind_rows)
+    # occ_bg = map2(occ, bg, bind_rows)
+    occ = map2(occ, bg, bind_rows)
   ) |>
-  select(-bg_data)
+  select(-bg)
 ```
 
-<!-- TODO: fix informal tone -->
+<!-- TODO: rewrite -->
 Occurrence data only tells us where a species is present;
@@
     "paleoclim", 
-    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
+    paste0("paleoclim_", climate_periods$code, "_2_5m_w_asia.tif")
   ),
@@
 
-names(climate) <- names(climate_periods)
+names(climate) <- climate_periods$code
 
@@
 
-# TODO: derived terrain products are two big to commit
-# TODO: does TWI actually make sense at this resolution? Could start with (much)
-# higher res, then upscale?
 terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
@@
 ```{r data-environment}
-# For time-invariant predictors, collapse attributes into dimension and 
-# duplicate to create a dummy period dimension
-# TODO: maybe need to downsample first?
-terrain <- st_redimension(terrain, name = "variable")
-terrain <- do.call(c, rep(list(terrain), length(climate_periods)))
-names(terrain) <- names(climate_periods)
-terrain <- st_redimension(terrain, name = "period") # TODO: why *so* slow?
+# Prepare environmental data cube for prediction (not training)
+# Duplicate time-invariant predictors to create dummy period dimension
+terrain_cube <- list(terrain) |>
+  rep(nrow(climate_periods)) |>
+  set_names(climate_periods$code) |>
+  c(list(along = "period")) |>
+  do.call(what = c)
 
-soil <- st_redimension(soil, name = "variable")
-soil <- do.call(c, rep(list(soil), length(climate_periods)))
-names(soil) <- names(climate_periods)
-soil <- st_redimension(soil, name = "period")
+soil_cube <- list(soil) |>
+  rep(nrow(climate_periods)) |>
+  set_names(climate_periods$code) |>
+  c(list(along = "period")) |>
+  do.call(what = c)
 
-# Resample to common grid
-terrain <- st_warp(terrain, climate, use_gdal = TRUE, method = "cubicspline")
-soil <- st_warp(soil, climate, use_gdal = TRUE, method = "cubicspline")
+# Resample predictor data to common projection and resolution
+climate_cube <- st_warp(climate, crs = w_asia_albers, 
+                        cellsize = c(pred_res, pred_res))
+terrain_cube <- st_warp(terrain_cube, climate_cube)
+soil_cube <- st_warp(soil_cube, climate_cube)
 
-# TODO: combine and split variable
-# TODO: fill NAs?
+# Combine into prediction cube
+environment <- c(climate_cube, terrain_cube, soil_cube)
+
+# Save some memory
+rm(climate_cube, terrain_cube, soil_cube)
 ```
@@
 ```{r extract-predictor-data}
-flora |> 
+# TODO: factor out
+bind_extract <- function(x, y) {
+  extract <- st_extract(y, st_transform(x, st_crs(y)))
+  cbind(x, select(tibble(extract), -geometry))
+}
+
+flora <- flora |> 
   mutate(
-    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x)))),
-    # TODO: where is the extra geometry column coming from?
-    occ = map(occ, \(x) as_tibble(select(x, -geometry))),
-    # TODO: can this happen earlier?
-    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE))))
-  ) ->
-  flora
+    occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, remove = FALSE),
+    occ = map(occ, bind_extract, slice(climate, "period", "cur")),
+    occ = map(occ, bind_extract, terrain),
+    occ = map(occ, bind_extract, soil)
+  )
 ```
@@
   mutate(
+    # De-sf
+    occ = map(occ, \(x) select(tibble(x), -geometry)),
+    # TODO: can this happen earlier?
+    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
     occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
-    occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
+    # occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
     occ_split = map(occ, \(x) initial_split(x, strata = present)),
@@
   update_role(starts_with("bio"), new_role = "predictor") |>
+  update_role(c(elevation, slope, aspect, wetness, clay, phh2o, sand, silt),
+              new_role = "predictor") |>
+  # Impute missing values based on spatial autocorrelation
+  step_impute_linear(all_numeric_predictors(), 
+                     impute_with = imp_vars(latitude, longitude)) |>
+  # Drop redundant predictors
   step_nzv(all_predictors())
@@
 
-# TODO: Or... tune parameters? But this is very compute-intensive.
-# niche_model <- rand_forest(
-#   mode = "classification", 
-#   mtry = tune(),
-#   trees = tune(),
-#   min_n = tune()
-# ) |>
-#   set_engine("ranger", probability = TRUE)
-
 niche_workflow <- workflow(niche_recipe, niche_model)
-
-# Tune model (separately for each taxon)
-# TODO: questioning - would be nice but is extremely slow
-# flora |>
-#   slice_sample(n = 10) |>
-#   mutate(
-#     niche_tuning_grid = map(occ_split, function(data, recipe) {
-#       grid_regular(
-#         finalize(mtry(), bake(prep(recipe), training(data), all_predictors())),
-#         trees(),
-#         min_n(),
-#         levels = 5
-#       )
-#     }, recipe = niche_recipe, .progress = TRUE),
-#     niche_tuning = map2(niche_tuning_grid, occ_folds, function(grid, folds, wflow) {
-#       tune_grid(wflow, folds, grid = grid)
-#     }, wflow = niche_workflow, .progress = TRUE),
-#     niche_hparams = map(niche_tuning, select_best, metric = "roc_auc"),
-#     niche_model = map(niche_hparams, function(params, wflow) {
-#       finalize_workflow(wflow, params)
-#     }, wflow = niche_workflow)
-#   )
 ```
@@
 
-<!-- TODO: OR:
-Hyperparameter tuning is important to avoid over-fitting [@CITE].
-We tuned each model separately to maximise ROC AUC.
--->
-
 ```{r fit-niche-models}
 # Fit best model to training data
-flora <- flora |>
-  mutate(
+if (!file_exists(derived_data("flora_niche_fit.Rdata"))) {
+  flora <- mutate(
+    flora,
     niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
   )
+  saveRDS(flora, derived_data("flora_niche_fit.Rdata"))
+}
+flora <- readRDS(derived_data("flora_niche_fit.Rdata"))
 ```
 
-<!-- TODO: cut? -->
+<!-- TODO: rewrite to remove thresholding -->
 The output of the model is probabilistic.
@@
 This also makes it possible to analyses the predictions together as an assemblage.
-
-```{r thresholding}
-# TODO: thresholding?
-# - Pick a threshold
-# - Reclassify pred raster
-# - Extract predicted classification (for cur) for occ
-# - ...or predict back onto the occurrence data? do we actually need the rasters for this?
-# - Calculate sensitivity and with yardstick
-# - Sum (SSS)
-# - (Plot?)
-# - Pick max SSS
-#
-# Or just stick with probabilities?
-```
 
@@
 ```{r predict}
-# TODO: structure this way from the beginning?
-# Thanks to https://stackoverflow.com/questions/65183306/stars-package-how-to-define-additional-dimensions-based-on-an-attribute-filena
-climate_pred <- do.call(c, map(climate, st_redimension, name = "variable")) |>
-  st_redimension(name = "period") |>
-  split("variable")
-
 # NOTE: very high memory usage
+if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
   flora <- flora |>
@@
       prediction = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
-                     obj = climate_pred)
+                       obj = environment, .progress = TRUE)
     )
+  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
+}
+flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
 ```
 
-```{r fig-niche-prediction}
-niche_prediction <- filter(flora, taxon == "Linum bienne")$prediction[[1]]
+```{r fig-niche-prediction, eval=FALSE}
+niche_taxon <- "Linum bienne"
+niche_prediction <- filter(flora, taxon == niche_taxon)$prediction[[1]]
 
 # Order and relabel periods
-climate_periods <- niche_prediction |>
-  st_get_dimension_values("period") |>
-  ordered(levels = c("LGM", "HS1", "BA", "YDS", "EH", "cur")) |>
-  recode(
-    "LGM" = "LGM (21 ka)",
-    "HS1" = "HS1 (17.0-14.7 ka)",
-    "BA"  = "BA (14.7-12.9 ka)",
-    "YDS" = "YDS (12.9-11.7 ka)",
-    "EH"  = "EH (11.7-8.3 ka)",
-    "cur" = "current"
-  )
-
-niche_prediction <- st_set_dimensions(niche_prediction, "period", values = climate_periods)
-
 ggplot() + 
-  geom_stars(data = select(niche_prediction, prediction.TRUE.) > 0.5) +
   facet_wrap(~period, ncol = 3) +
-  # scale_fill_batlow(reverse = TRUE) +
-  coord_sf(crs = latlong) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(niche_prediction, prediction.TRUE.),
+             na.action = na.omit) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(crs = w_asia_albers) +
   labs(
-    title = "Linum bienne",
+    title = niche_taxon,
     fill = NULL,
@@
   ) +
-  theme_map() +
   theme(
@@
 
-```{r sum-pred-eh}
+```{r sum-pred-eh, eval=FALSE}
 flora |>
@@
 
-```{r fig-sum-pred-eh}
+```{r fig-sum-pred-eh, eval=FALSE}
 #| fig-cap: predicted species richness for the early Holocene
@@
 
-```{r fig-pred-observed-eh}
-
-```
-
 # Conclusion

### 2025-01-04 analysis/paper.tex
Optimise model code

@@
 
+\usepackage{booktabs}
+\usepackage{caption}
+\usepackage{longtable}
+\usepackage{colortbl}
+\usepackage{array}
+\usepackage{anyfontsize}
+\usepackage{multirow}
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21--8.3 ka},
+  pdftitle={Biogeography of crop progenitors and wild plant resources in West Asia, 21--8.3 ka},
   pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
@@
 \begin{frontmatter}
-\title{Biogeography of crop progenitors and wild plant resources during
-the transition to agriculture in West Asia, 21--8.3 ka}
+\title{Biogeography of crop progenitors and wild plant resources in West
+Asia, 21--8.3 ka}
 \author[1,2]{Joe Roe%
@@
 This paper presents the first comprehensive reconstructions of the
-palaeodistributions of {[}X{]} plant species known to have been gathered
-or cultivated by early agricultural societies in Southwest Asia,
-including the progenitors of the first crops. We used machine learning
-to train an ecological niche model (ENM) of each species based on its
-present-day distribution in relation to climate and environmental
-variables. Predictions of the potential ranges of these species at key
-stages of the Pleistocene--Holocene transition could then be derived
-from these models using hindcast data from palaeoclimate simulations.
-Species ranges were on average {[}X\%{]} {[}larger\textbar smaller{]} in
-the Early Holocene compared to today, indicating {[}\ldots{]}. The
-modelled ranges predict the observed occurrence of these species on
-archaeological sites with {[}low\textbar medium\textbar high{]}
-accuracy. The regional ubiquity of species in the archaeological record
-is {[}not{]} correlated with the predicted size of its range and the
-diversity of archaeobotanical assemblages is {[}not{]} correlated with
-the predicted diversity of its environs. This indicates that trends in
-taxonomic composition of the archaeobotanical record is {[}not{]} likely
-to be influenced by environmental change and species turnover, not just,
-as is often assumed, human economic choices.
+palaeodistributions of 56 plant species found regularly in association
+with early agricultural archaeological sites in West Asia, including the
+progenitors of the first crops. We used machine learning to train an
+ecological niche model of each species based on its present-day
+distribution in relation to climate and environmental variables.
+Predictions of the potential ranges of these species at key stages of
+the Pleistocene--Holocene transition could then be derived from these
+models using hindcast data from palaeoclimate simulations. Species
+ranges are predicted to have been on average {[}X\%{]}
+{[}larger\textbar smaller{]} in the Early Holocene compared to present
+conditions, indicating {[}\ldots{]}. The modelled ranges predict the
+observed occurrence of these species on archaeological sites with
+{[}low\textbar medium\textbar high{]} accuracy. The regional ubiquity of
+species in the archaeological record is {[}not{]} correlated with the
+predicted size of its range and the diversity of archaeobotanical
+assemblages is {[}not{]} correlated with the predicted diversity of its
+environs. This indicates that trends in taxonomic composition of the
+archaeobotanical record is {[}not{]} likely to have been influenced by
+environmental change and species turnover, in addition to human economic
+choices.
 \end{abstract}
@@
 \tightlist
-\item[$\square$]
-  Finish analysis:
-
-  \begin{itemize}
-  \tightlist
-  \item[$\square$]
-    Add terrain, soil
-  \item[$\square$]
-    Find a resolution that's workable
-  \item[$\square$]
-    Do thresholding?
-  \end{itemize}
+\item[$\boxtimes$]
+  Finish modelling
 \item[$\square$]
@@
       Biogeography
-    \item[$\square$]
+    \item[$\boxtimes$]
       ENM
@@
     Methods \& Materials
-
-    \begin{itemize}
-    \tightlist
-    \item[$\boxtimes$]
-      Occurrence data
-    \item[$\boxtimes$]
-      Predictor data
-    \item[$\boxtimes$]
-      Random Forest model
-    \end{itemize}
   \item[$\square$]
@@
   \item[$\square$]
+    tbl-occ-count
+  \item[$\square$]
     tbl-predictors
@@
   \end{itemize}
+\item[$\square$]
+  Clean up package code
 \item[$\square$]
@@
 
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-region-1.pdf}}
+
+}
+
+\caption{\label{fig-region}Study region}
+
+\end{figure}%
+
 \subsection{Occurrence data}\label{occurrence-data}
@@
 GBIF Backbone Taxonomy (\textbf{GBIFSecretariat2023?}). So for example
-occurrences for ``Bolboschoenus'' included all species and subspecies,
-including specimens described as \emph{Bolboschoenus sp.},
-\emph{Bolboschoenus maritimus} and \emph{Scirpus maritimus}. Domestic
-species meeting our inclusion criteria were substituted with their wild
-progenitor(s), where known.
+occurrences for \emph{Bolboschoenus maritimus} also include those
+recorded under the older nomenclature \emph{Scirpus maritimus} (see
+Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
+criteria were substituted with their wild progenitor(s), where known.
 
@@
 
-Occurrence data was obtained from the Global Biodiversity Information
-Facility (GBIF) using via its application programming interface and the
-R package rgbif (Chamberlain et al., 2024; Chamberlain and Boettiger,
-2017). Although ENMs have reasonable predictive power even with small
-training samples (Hernandez et al., 2006; Stockwell and Peterson, 2002;
-Wisz et al., 2008), we excluded X taxa with less than fifty recorded
-occurrences in our study region. We also excluded one taxon (\emph{Avena
-sterilis}) with over 47,000 occurrences, as this would have been
+\begin{verbatim}
+Rows: 62 Columns: 5
+-- Column specification --------------------------------------------------------
+Delimiter: "\t"
+chr (2): taxon, taxon_detail
+dbl (3): n_present, p_present, n_occ
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+\end{verbatim}
+
+\begin{verbatim}
+Warning: HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+HTML tags found, and they will be removed.
+* Set `options(gt.html_tag_check = FALSE)` to disable this check.
+\end{verbatim}
+
+\begin{table}
+
+\caption{\label{tbl-occ-count}Recorded occurrence of flora considered in
+this study in Neolithic and contemporary West Asia}
+
+\centering{
+
+\fontsize{12.0pt}{14.4pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrr}
+\toprule
+ & \multicolumn{2}{c}{Occurrences} \\ 
+\cmidrule(lr){2-3}
+Taxon & Neolithic & Present \\ 
+\midrule\addlinespace[2.5pt]
+\emph{Triticum turgidum dicoccum}
+
+incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 (44\%) & 200 \\ 
+\emph{Hordeum spontaneum}
+
+incl. \emph{Hordeum vulgare} & 62 (43\%) & 6912 \\ 
+\emph{Triticum monococcum aegilopoides}
+
+incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 (30\%) & 3213 \\ 
+\emph{Bolboschoenus maritimus}
+
+incl. \emph{Scirpus maritimus} & 33 (23\%) & 364 \\ 
+\emph{Vicia ervilia} & 33 (23\%) & 858 \\ 
+\emph{Buglossoides tenuiflora} & 31 (21\%) & 159 \\ 
+\emph{Arnebia decumbens} & 25 (17\%) & 240 \\ 
+\emph{Buglossoides arvensis} & 25 (17\%) & 251 \\ 
+\emph{Medicago radiata} & 21 (14\%) & 883 \\ 
+\emph{Androsace maxima} & 20 (14\%) & 130 \\ 
+\emph{Vicia orientalis}
+
+incl. \emph{Lens culinaris} & 17 (12\%) & 441 \\ 
+\emph{Medicago astroites}
+
+incl. \emph{Trigonella astroites} & 16 (11\%) & 130 \\ 
+\emph{Arnebia linearifolia} & 14 (10\%) & 158 \\ 
+\emph{Linum bienne}
+
+incl. \emph{Linum usitatissimum} & 11 (8\%) & 274 \\ 
+\emph{Gypsophila vaccaria}
+
+incl. \emph{Vaccaria pyramidata} & 11 (8\%) & 326 \\ 
+\emph{Carex divisa} & 10 (7\%) & 330 \\ 
+\emph{Ficus carica} & 10 (7\%) & 4265 \\ 
+\emph{Lathyrus oleraceus}
+
+incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 (6\%) & 2269 \\ 
+\emph{Vicia faba} & 8 (6\%) & 2578 \\ 
+\emph{Aizoanthemopsis hispanica}
+
+incl. \emph{Aizoon hispanicum} & 7 (5\%) & 303 \\ 
+\emph{Bolboschoenus glaucus} & 7 (5\%) & 27\textsuperscript{\textit{*}} \\ 
+\emph{Pistacia atlantica} & 7 (5\%) & 928 \\ 
+\emph{Polygonum arenarium arenarium}
+
+incl. \emph{Polygonum venantianum} & 7 (5\%) & 5\textsuperscript{\textit{*}} \\ 
+\emph{Prosopis farcta} & 7 (5\%) & 1205 \\ 
+\emph{Rumex pulcher} & 7 (5\%) & 794 \\ 
+\emph{Ammi majus} & 6 (4\%) & 461 \\ 
+\emph{Cicer reticulatum}
+
+incl. \emph{Cicer arietinum} & 6 (4\%) & 193 \\ 
+\emph{Hordeum bulbosum} & 6 (4\%) & 2036 \\ 
+\emph{Polygonum corrigioloides} & 6 (4\%) & 4\textsuperscript{\textit{*}} \\ 
+\emph{Salsola kali} & 6 (4\%) & 369 \\ 
+\emph{Taeniatherum caput-medusae} & 6 (4\%) & 235 \\ 
+\emph{Capparis spinosa} & 5 (3\%) & 4031 \\ 
+\emph{Chenopodium album} & 5 (3\%) & 357 \\ 
+\emph{Lolium temulentum} & 5 (3\%) & 148 \\ 
+\emph{Poa bulbosa} & 5 (3\%) & 1401 \\ 
+\emph{Aegilops crassa} & 4 (3\%) & 170 \\ 
+\emph{Atriplex prostrata} & 4 (3\%) & 112 \\ 
+\emph{Brachypodium distachyon} & 4 (3\%) & 1581 \\ 
+\emph{Fumaria densiflora} & 4 (3\%) & 194 \\ 
+\emph{Gypsophila pilosa} & 4 (3\%) & 79 \\ 
+\emph{Hordeum murinum} & 4 (3\%) & 1342 \\ 
+\emph{Lathyrus aphaca} & 4 (3\%) & 2085 \\ 
+\emph{Lathyrus sativus} & 4 (3\%) & 297 \\ 
+\emph{Secale cereale} & 4 (3\%) & 1153 \\ 
+\emph{Triticum durum} & 4 (3\%) & 219 \\ 
+\emph{Vitis sylvestris} & 4 (3\%) & 3\textsuperscript{\textit{*}} \\ 
+\emph{Adonis flammea} & 3 (2\%) & 79 \\ 
+\emph{Avena sterilis} & 3 (2\%) & 50001\textsuperscript{\textit{*}} \\ 
+\emph{Bassia arabica} & 3 (2\%) & 230 \\ 
+\emph{Bromus sterilis} & 3 (2\%) & 385 \\ 
+\emph{Cephalaria syriaca} & 3 (2\%) & 144 \\ 
+\emph{Glaucium aleppicum} & 3 (2\%) & 22\textsuperscript{\textit{*}} \\ 
+\emph{Helianthemum ledifolium} & 3 (2\%) & 668 \\ 
+\emph{Lolium rigidum} & 3 (2\%) & 1464 \\ 
+\emph{Phalaris paradoxa} & 3 (2\%) & 370 \\ 
+\emph{Quercus ithaburensis} & 3 (2\%) & 1431 \\ 
+\emph{Triticum aestivum compactum} & 3 (2\%) & 209 \\ 
+\emph{Verbena officinalis} & 3 (2\%) & 1046 \\ 
+\emph{Vicia narbonensis} & 3 (2\%) & 1529 \\ 
+\emph{Triticum urartu} & — & 2033 \\ 
+\emph{Aegilops speltoides} & — & 1429 \\ 
+\emph{Aegilops tauschii} & — & 1395 \\ 
+\bottomrule
+\end{tabular*}
+\begin{minipage}{\linewidth}
+\textsuperscript{\textit{*}}Excluded from modelling due to sample size\\
+\end{minipage}
+
+}
+
+\end{table}%
+
+Georeferenced occurrence data was obtained from the Global Biodiversity
+Information Facility (GBIF) using via its application programming
+interface and the R package `rgbif' (Chamberlain et al., 2024;
+Chamberlain and Boettiger, 2017). GBIF occurrences marked has having
+imprecise or duplicate coordinates were excluded from the training
+dataset, as were fossil records. Although niche models have reasonable
+predictive power even with small training samples (Hernandez et al.,
+2006; Stockwell and Peterson, 2002; Wisz et al., 2008), we excluded 57
+taxa with less than 50 occurrences in West Asia, following
+recommendations for niche models generally and Random Forest-based
+models specifically (Stockwell and Peterson, 2002;
+\textbf{LuanEtAl2020?}). We also excluded one taxon (\emph{Avena
+sterilis}) with over 50,000 occurrences, as this would have been
 computationally prohibitive and we were uncertain what account for such
 a disproportionately high number of records.
+
+\begin{verbatim}
+Rows: 35208 Columns: 11
+-- Column specification --------------------------------------------------------
+Delimiter: "\t"
+chr (5): taxon, taxon_detail, scientific_name, genus, species
+dbl (6): n_present, p_present, n_occ, gbif_key, longitude, latitude
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+\end{verbatim}
 
@@
 
-For hindcasting, we used reconstructed bioclimatic data for key periods
-(\textbf{?@tbl-climate-periods}) generated from downscaled paleoclimate
-simulations from the HadCM3 general circulation model (Brown et al.,
-2018). Terrain and soil predictors were held constant, since
-reconstructions of these variables in the past are not available at
-sufficient scale. It is not likely that either macroscale topography or
-soil characteristics have altered significantly over the period of time
-considered here, so we assume that this does not degrade model
+For hindcasting, we used reconstructed bioclimatic data for 5 key
+periods (\textbf{?@tbl-climate-periods}) generated from downscaled
+paleoclimate simulations from the HadCM3 general circulation model
+(Brown et al., 2018). Terrain and soil predictors were held constant,
+since reconstructions of these variables in the past are not available
+at sufficient scale. It is not likely that either macroscale topography
+or soil characteristics have altered significantly over the period of
+time considered here, so we assume that this does not degrade model
 performance, and may in fact benefit it by providing `anchoring'
@@
 \subsection{Model assessment}\label{model-assessment}
+
+\begin{verbatim}
+Warning: Removed 51 rows containing missing values or values outside the scale range
+(`geom_text_repel()`).
+\end{verbatim}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-model-accuracy-vs-n_occ-1.pdf}}
+
+}
+
+\caption{\label{fig-model-accuracy-vs-n_occ}Number of taxon occurrences
+and model accuracy on test dataset}
+
+\end{figure}%
 

### 2025-01-04 R/enm.R

@@ -8,6 +8,8 @@
 #'                an object with an [sf::st_sample()] method, or an object
 #'                coercible with [sf::st_bbox] (for much faster sampling from
 #'                a rectangular bounding box).
+#' @param coord_x Name of the column holding the x coordinate.
+#' @param coord_y Name of the column holding the y coordinate.
 #' @param ...     Name-value pairs of expressions to be added as fixed attributes
 #'                of the pseudo-absences. See [dplyr::mutate()].
 #' @param N       integer. Number of pseudo-absence points to generate.
@@ -17,8 +19,6 @@
 #' specified in `...`.
 #'
 #' @export
-#'
-#' @examples
 background_sample <- function(region, N, coord_x = "x", coord_y = "y", ...) {
   if (inherits(region, c("sf", "sfc", "sfg"))) {
     bg_sample <- sample_sf(region, N, ...)
@@ -34,7 +34,7 @@ background_sample <- function(region, N, coord_x = "x", coord_y = "y", ...) {
     colnames(bg_sample)[colnames(bg_sample) == "Y"] <- coord_y
   }
 
-  bg_sample
+  dplyr::select(tibble::tibble(bg_sample), -geometry)
 }
 
 #' @noRd

### 2025-01-04 R/gbif.R

@@ -20,6 +20,14 @@ gbif_accepted_name <- function(name) {
   else gbif$canonicalName
 }
 
+#' Acquire GBIF occurrence data for a given taxon
+#'
+#' @export
+gbif_data <- function(scientificName, ...) {
+  gbif_data <- rgbif::occ_data(scientificName = scientificName, ...)
+  gbif_data$data
+}
+
 #' Clean GBIF data
 #'
 #' A collection of functions for cleaning occurrence data from GBIF.

### 2025-01-04 analysis/paper.qmd
Improve PDF formatting

@@
 ---
-title: Biogeography of crop progenitors and wild plant resources in West Asia, 21–8.3 ka
+title: Biogeography of crop progenitors and wild plant resources in Late Pleistocene–Early Holocene West Asia, 21–8.3 ka
 date: today
@@
     keep-tex: true
-    cite-method: citeproc
+    # cite-method: citeproc
     journal:
@@
       formatting: review
+      cite-style: authoryear
 bibliography: references.bib
@@
 - [ ] Figures
-  - [ ] tbl-occ-count
+  - [ ] tbl-occ-count (needs pdf cleanup)
   - [ ] tbl-predictors
@@
 @FranklinEtAl2015 review palaeoecological niche modelling and advocate for its greater adoption in environmental archaeology.
-One relevant early example is @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites in Europe and West Asia to map the evolving niche of cattle over the Pleistocene–Holocene transition.
+In an early application to West Asia, @ConollyEtAl2012 used the occurrence of wild and domestic *Bos* remains at prehistoric archaeological sites to map the evolving niche of cattle over the Pleistocene–Holocene transition.
 It has been used to model the availability of fauna exploited by humans at wider scales [e.g.
@@
 
-## Occurrence data
-
 ```{r data-archaeo-flora}
@@
 ```
-
-We consider X<!-- TODO --> distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic sites in West Asia, according to our previous study [@ArranzOtaeguiRoe2023].
-<!-- TODO: something about not distinguishing between source of plant remains? -->
 
@@
 
-Taxonomic names were resolves to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
+We began by considering `r nrow(flora)` distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic (c.
+11.7–6.5 ka) sites in West Asia, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
+<!-- TODO: something about not distinguishing between source of plant remains? -->
+Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
 So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
@@
 }
-flora <- read_tsv(derived_data("flora_occ_count.tsv"))
+flora <- read_tsv(
+  derived_data("flora_occ_count.tsv"),
+  col_types = cols(
+    taxon = col_character(),
+    taxon_detail = col_character(),
+    n_present = col_double(),
+    p_present = col_double(),
+    n_occ = col_double()
+  )
+)
 ```
@@
 }
-flora <- nest(read_tsv(derived_data("flora_occ.tsv")), occ = gbif_key:latitude)
+flora <- read_tsv(
+  derived_data("flora_occ.tsv"),
+  col_types = cols(
+    taxon = col_character(),
+    taxon_detail = col_character(),
+    n_present = col_double(),
+    p_present = col_double(),
+    n_occ = col_double(),
+    gbif_key = col_double(),
+    scientific_name = col_character(),
+    genus = col_character(),
+    species = col_character(),
+    longitude = col_double(),
+    latitude = col_double()
+  )
+) |>
+  nest(occ = gbif_key:latitude)
 ```
@@
 
-Predictor data was transformed to the same projection system (WGS84 / UTM 37 N) and a common resolution of X<!-- TODO --> km.
+Predictor data was transformed to common equal-area projection and resolution of `r pred_res / 1000` km.
+
+```{r tbl-climate-periods}
+#| fig-cap: Palaeoclimatic periods used for hindcasting, after @BrownEtAl2018
+# TODO: add refs (as footnotes?)
+climate_periods |>
+  mutate(
+    code = toupper(code),
+    age_start = start_bp / 1000,
+    age_end = end_bp / 1000
+  ) |>
+  select(period, code, age_start, age_end) |>
+  gt() |>
+  cols_label(period = "Period", age_start = "Age") |>
+  cols_merge(
+    c(period, code),
+    pattern = "{1} ({2})"
+  ) |>
+  cols_merge_range(age_start, age_end) |>
+  cols_units(age_start = "ka") |>
+  sub_missing()
+```
 
@@
 To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
-Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (3/4) and test (1/4) partitions.
+Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (¾) and test (¼) partitions.
 
@@
   filter(.metric == "accuracy") |> # TODO: or roc_auc?
-  mutate(label = if_else(.estimate < 0.8, taxon, NA_character_)) |>
+  mutate(label = if_else(.estimate < 0.75, taxon, NA_character_)) |>
   ggplot(aes(n_occ, .estimate)) + 

### 2025-01-04 analysis/paper.tex
Improve PDF formatting

@@
 \documentclass[
-  number,
+  authoryear,
   review]{elsarticle}
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-harv}
 \usepackage{bookmark}
@@
 \hypersetup{
-  pdftitle={Biogeography of crop progenitors and wild plant resources in West Asia, 21--8.3 ka},
+  pdftitle={Biogeography of crop progenitors and wild plant resources in Late Pleistocene--Early Holocene West Asia, 21--8.3 ka},
   pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
@@
 \begin{frontmatter}
-\title{Biogeography of crop progenitors and wild plant resources in West
-Asia, 21--8.3 ka}
+\title{Biogeography of crop progenitors and wild plant resources in Late
+Pleistocene--Early Holocene West Asia, 21--8.3 ka}
 \author[1,2]{Joe Roe%
@@
   \item[$\square$]
-    tbl-occ-count
+    tbl-occ-count (needs pdf cleanup)
   \item[$\square$]
@@
     Overlap makes it difficult to see where human choices depart from
-    the environmental background (cf. Martin et al., 2016)
+    the environmental background \citep[cf.][]{MartinEtAl2016}
   \item
     Or to fully contextualise subsistence strategies and shifts in
-    strategies in response to environmental shifts (e.g. Yaworsky et
-    al., 2023).
+    strategies in response to environmental shifts
+    \citep[e.g.][]{YaworskyEtAl2023}.
   \end{itemize}
@@
     Distinguish environmental from potentially anthropogenic change
-    (\textbf{MartinEtAl2017?}; \textbf{MartinEtAl2025?})
+    \citep{MartinEtAl2017, MartinEtAl2025}
   \item
-    Reconstruct sequences of domestication (\textbf{YeomansEtAl2017?})
+    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
   \end{itemize}
@@
 the organism in question either in the same or a different environment.
-(\textbf{CITE?}) suggests reserving the term `species distribution
+\citet{CITE} suggests reserving the term `species distribution
 modelling' for when the method is used to recover the verifiable range
@@
 as both a means of exploring the biological niche of humans and for
-reconstructing the past environments they inhabited (Franklin et al.,
-2015; \textbf{DavidPollyEronen2011?}). In the first sense, it has been
-used most extensively to model the range of humans and other hominin
-species (e.g. \textbf{BenitoEtAl2017?}; \textbf{YousefiEtAl2020?};
-\textbf{BanksEtAl2021?}; \textbf{YaworskyEtAl2024a?};
-\textbf{YaworskyEtAl2024b?}; \textbf{GuranEtAl2024?}), especially in the
-Palaeolithic. This overlaps with what archaeologists usually call
-generically `predictive modelling'
-(\textbf{VerhagenWhitley2020?})---more precisely `site distribution
+reconstructing the past environments they inhabited
+\citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
+has been used most extensively to model the range of humans and other
+hominin species
+\citep[e.g.][]{BenitoEtAl2017, YousefiEtAl2020, BanksEtAl2021, YaworskyEtAl2024a, YaworskyEtAl2024b, GuranEtAl2024},
+especially in the Palaeolithic. This overlaps with what archaeologists
+usually call generically `predictive modelling'
+\citep{VerhagenWhitley2020}---more precisely `site distribution
 modelling'---which is essentially the same approach as (and often
@@
 
-(Franklin et al., 2015) review palaeoecological niche modelling and
-advocate for its greater adoption in environmental archaeology. One
-relevant early example is (Conolly et al., 2012) used the occurrence of
-wild and domestic \emph{Bos} remains at prehistoric archaeological sites
-in Europe and West Asia to map the evolving niche of cattle over the
+\citet{FranklinEtAl2015} review palaeoecological niche modelling and
+advocate for its greater adoption in environmental archaeology. In an
+early application to West Asia, \citet{ConollyEtAl2012} used the
+occurrence of wild and domestic \emph{Bos} remains at prehistoric
+archaeological sites to map the evolving niche of cattle over the
 Pleistocene--Holocene transition. It has been used to model the
-availability of fauna exploited by humans at wider scales (e.g.
-\textbf{deAndresHerreroEtAl2018?}; \textbf{YaworksyEtAl2023?}) and, in a
-West Asian context, of foraged plant resources in the landscape around
-the Neolithic site of XX (Collins et al., 2018). Modelling the spread of
+availability of fauna exploited by humans at wider scales
+\citep[e.g.][]{deAndresHerreroEtAl2018, YaworksyEtAl2023} and, in a West
+Asian context, of foraged plant resources in the landscape around the
+Neolithic site of XX \citep{CollinsEtAl2018}. Modelling the spread of
 crops has been another significant archaeological application
-(\textbf{CremaEtAl?}).
+\citep{CremaEtAl}.
 
@@
 occurrence and environmental data and then use palaeoenvironmental data
-to `hindcast' its predictions backwards in time. Like (Franklin et al.,
-2015), we view the hindcasting approach as more promising, because
-training datasets for both occurrences and environment are far more
-readily available, complete and reliable for the present than the past.
-There is some scepticism in the ecological niche modelling literature
-about the ability of such models to make accurate predictions in unknown
-environments (like the past) (\textbf{CITES?}), but here the hindcasting
-approach also presents an opportunity: it reserves archaeological
-occurrence data as an independent dataset that can be used to assess the
-retrodictive performance of the model. This possibily was suggested by
-(Franklin et al., 2015) but to our knowledge our study represents the
+to `hindcast' its predictions backwards in time. Like
+\citet{FranklinEtAl2015}, we view the hindcasting approach as more
+promising, because training datasets for both occurrences and
+environment are far more readily available, complete and reliable for
+the present than the past. There is some scepticism in the ecological
+niche modelling literature about the ability of such models to make
+accurate predictions in unknown environments (like the past)
+\citep{CITES}, but here the hindcasting approach also presents an
+opportunity: it reserves archaeological occurrence data as an
+independent dataset that can be used to assess the retrodictive
+performance of the model. This possibily was suggested by
+\citet{FranklinEtAl2015} but to our knowledge our study represents the
 first attempt to actually do so.
@@
 applications, which is perhaps why only a minority of studies use it
-(cf. Yaworsky et al., 2023). In this study, we are able to take
+\citep[cf.][]{YaworskyEtAl2023}. In this study, we are able to take
 advantage of the increasing availability of high resolution, global
 palaeoclimate data derived from simulation experiments with general
-circulation models of climate (Brown et al., 2018;
-\textbf{BrownEtAl2020?}; \textbf{KargerEtAl2023?}).
+circulation models of climate
+\citep{BrownEtAl2018, BrownEtAl2020, KargerEtAl2023}.
 
@@
 
-\subsection{Occurrence data}\label{occurrence-data}
-
-We consider X distinct taxa (Table~\ref{tbl-occ-count}) - all the
-identifiable species known to be present at more than three Neolithic
-sites in West Asia, according to our previous study (Arranz-Otaegui and
-Roe, 2023).
-
-Taxonomic names were resolves to the canonical form specified in the
-GBIF Backbone Taxonomy (\textbf{GBIFSecretariat2023?}). So for example
-occurrences for \emph{Bolboschoenus maritimus} also include those
-recorded under the older nomenclature \emph{Scirpus maritimus} (see
+We began by considering 62 distinct taxa (Table~\ref{tbl-occ-count}) -
+all the identifiable species known to be present at more than three
+Neolithic (c.~11.7--6.5 ka) sites in West Asia, according to our
+previous study of the regional archaeobotanical data
+\citep{ArranzOtaeguiRoe2023}. Taxonomic names were resolved to the
+canonical form specified in the GBIF Backbone Taxonomy
+\citep{GBIFSecretariat2023}. So for example occurrences for
+\emph{Bolboschoenus maritimus} also include those recorded under the
+older nomenclature \emph{Scirpus maritimus} (see
 Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
@@
 
-\subsection{Occurrence data}\label{occurrence-data-1}
-
-\begin{verbatim}
-Rows: 62 Columns: 5
--- Column specification --------------------------------------------------------
-Delimiter: "\t"
-chr (2): taxon, taxon_detail
-dbl (3): n_present, p_present, n_occ
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-\end{verbatim}
+\subsection{Occurrence data}\label{occurrence-data}
 
@@
 Information Facility (GBIF) using via its application programming
-interface and the R package `rgbif' (Chamberlain et al., 2024;
-Chamberlain and Boettiger, 2017). GBIF occurrences marked has having
-imprecise or duplicate coordinates were excluded from the training
-dataset, as were fossil records. Although niche models have reasonable
-predictive power even with small training samples (Hernandez et al.,
-2006; Stockwell and Peterson, 2002; Wisz et al., 2008), we excluded 57
-taxa with less than 50 occurrences in West Asia, following
+interface and the R package `rgbif'
+\citep{ChamberlainBoettiger2017, rgbif}. GBIF occurrences marked has
+having imprecise or duplicate coordinates were excluded from the
+training dataset, as were fossil records. Although niche models have
+reasonable predictive power even with small training samples
+\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
+excluded 57 taxa with less than 50 occurrences in West Asia, following
 recommendations for niche models generally and Random Forest-based
-models specifically (Stockwell and Peterson, 2002;
-\textbf{LuanEtAl2020?}). We also excluded one taxon (\emph{Avena
-sterilis}) with over 50,000 occurrences, as this would have been
-computationally prohibitive and we were uncertain what account for such
-a disproportionately high number of records.
-
-\begin{verbatim}
-Rows: 35208 Columns: 11
--- Column specification --------------------------------------------------------
-Delimiter: "\t"
-chr (5): taxon, taxon_detail, scientific_name, genus, species
-dbl (6): n_present, p_present, n_occ, gbif_key, longitude, latitude
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-\end{verbatim}
+models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
+excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
+as this would have been computationally prohibitive and we were
+uncertain what account for such a disproportionately high number of
+records.
 
@@
 ``pseudo-absences'' to feed to the model. There are several ways to do
-this. We follow the advice of (Barbet-Massin et al., 2012) for
+this. We follow the advice of \citet{BarbetMassinEtAl2012} for
 regression-based species distribution models and use a large (:10000)
 random sample of points, weighted equally against the presences in the
-regression. (Valavi et al., 2022) also recommend using a very large
+regression. \citet{ValaviEtAl2022} also recommend using a very large
 background sample for random forest models.
@@
   precipitation values, following standard practice for species
-  distribution models (Hijmans et al., 2005). Contemporary bioclimatic
+  distribution models \citep{HijmansEtAl2005}. Contemporary bioclimatic
   predictor data for West Asia was extracted from the global CHELSA
-  dataset (Karger et al., 2017), which predicts temperature and
+  dataset \citep{KargerEtAl2017}, which predicts temperature and
   precipitation from downscaled general circulation model output at 1 km
@@
   Terrain aspect and slope, which at high resolution perform well as
-  proxies for solar radiation when modelling plant occurrence (Austin
-  and Van Niel, 2011; Leempoel et al., 2015); and the topographic
+  proxies for solar radiation when modelling plant occurrence
+  \citep{AustinVanNiel2011, LeempoelEtAl2015}; and the topographic
   wetness index (TWI), which serves as a proxy for soil moisture and is
-  particularly important in modelling arid environments (Campos et al.,
-  2016; Di Virgilio et al., 2018; Kopecký and Čížková, 2010). All three
-  were derived from the SRTM15+ digital elevation model using algorithms
-  from WhiteboxTools (Lindsay, 2016).
+  particularly important in modelling arid environments
+  \citep{KopeckyCizkova2010, CamposEtAl2016, DiVirgilioEtAl2018}. All
+  three were derived from the SRTM15+ digital elevation model using
+  algorithms from WhiteboxTools \citep{Lindsay2016}.
 \end{itemize}
@@
 \item
-  Edaphic data from SoilGrids (Hengl et al., 2017, 2014), which improves
-  model performance for plants (Dubuis et al., 2013; Mod et al., 2016;
-  Velazco et al., 2017). Based on a recent assessment of the reliability
-  of SoilGrids data for species distribution modelling (Miller et al.,
-  2024), we used a subset of four variables relating to soil texture
-  (clay, silt, sand) and pH at the surface (0-5 cm depth).
+  Edaphic data from SoilGrids \citep{HenglEtAl2014, HenglEtAl2017},
+  which improves model performance for plants
+  \citep{DubuisEtAl2013, ModEtAl2016, VelazcoEtAl2017}. Based on a
+  recent assessment of the reliability of SoilGrids data for species
+  distribution modelling \citep{MillerEtAl2024}, we used a subset of
+  four variables relating to soil texture (clay, silt, sand) and pH at
+  the surface (0-5 cm depth).
 \end{itemize}
 
-Predictor data was transformed to the same projection system (WGS84 /
-UTM 37 N) and a common resolution of X km.
+Predictor data was transformed to common equal-area projection and
+resolution of 5 km.
+
+\begin{table}
+
+\caption{\label{tbl-climate-periods}}
+
+\centering{
+
+\fontsize{12.0pt}{14.4pt}\selectfont
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr}
+\toprule
+Period & Age, ka \\ 
+\midrule\addlinespace[2.5pt]
+Last Glacial Maximum (LGM) & 21.0–17.0 \\ 
+Heinrich Stadial 1 (HS1) & 17.0–14.7 \\ 
+Bølling-Allerød (BA) & 14.7–12.9 \\ 
+Younger Dryas (YDS) & 12.9–11.7 \\ 
+Early Holocene (EH) & 11.7–8.3 \\ 
+Current (CUR) & — \\ 
+\bottomrule
+\end{tabular*}
+
+Palaeoclimatic periods used for hindcasting, after \citet{BrownEtAl2018}
+
+}
+
+\end{table}%
 
 For hindcasting, we used reconstructed bioclimatic data for 5 key
-periods (\textbf{?@tbl-climate-periods}) generated from downscaled
+periods (Table~\ref{tbl-climate-periods}) generated from downscaled
 paleoclimate simulations from the HadCM3 general circulation model
-(Brown et al., 2018). Terrain and soil predictors were held constant,
+\citep{BrownEtAl2018}. Terrain and soil predictors were held constant,
 since reconstructions of these variables in the past are not available
@@
 literature exists on the relatively performance of these approaches and
-their respective parameterisations (reviewed in Valavi et al., 2022).
-Random Forest, a widely-used machine learning algorithm, is amongst the
-best performing approaches for presence-only species distribution
-models, providing it is appropriately parameterised to account for the
-class imbalance between presence and background samples (Valavi et al.,
-2022, 2021). For our application, it also has the advantage of requiring
-little to no manual parameter tuning to achieve good predictive results,
-which makes it easier to model a larger numbers of taxa.
+their respective parameterisations \citep[reviewed
+in][]{ValaviEtAl2022}. Random Forest, a widely-used machine learning
+algorithm, is amongst the best performing approaches for presence-only
+species distribution models, providing it is appropriately parameterised
+to account for the class imbalance between presence and background
+samples \citep{ValaviEtAl2021, ValaviEtAl2022}. For our application, it
+also has the advantage of requiring little to no manual parameter tuning
+to achieve good predictive results, which makes it easier to model a
+larger numbers of taxa.
 
@@
 (\textbf{?@tbl-predictors}). We used the Random Forest algorithm
-implemented in the R package `ranger' (Wright and Ziegler, 2017) and the
-`tidymodels' (\textbf{tidymodels?}) framework for data preprocessing and
-model selection. To avoid overfitting, we follow (Valavi et al., 2021)
+implemented in the R package `ranger' \citep{WrightZiegler2017} and the
+`tidymodels' \citep{tidymodels} framework for data preprocessing and
+model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
 in their recommended hyperparameters and use of down-sampling to balance
@@
 independently, with redundant zero-variance predictors excluded, and
-assessed based on balanced training (3/4) and test (1/4) partitions.
+assessed based on balanced training (¾) and test (¼) partitions.
 
 The output of the model is probabilistic. However, this should not be
-understood as an actual probability of occurrence (\textbf{CITE?}), but
-more akin to an estimate of habitat suitability. To simplify
-interpretation, we can convert these predictions into binary
-presence/absence maps, a process called ``thresholding''. We select the
-threshold value for each model individually, using MaxSSS (as
-recommended by Liu et al., 2013). This also makes it possible to
-analyses the predictions together as an assemblage.
+understood as an actual probability of occurrence \citep{CITE}, but more
+akin to an estimate of habitat suitability. To simplify interpretation,
+we can convert these predictions into binary presence/absence maps, a
+process called ``thresholding''. We select the threshold value for each
+model individually, using MaxSSS \citep[as recommended
+by][]{LiuEtAl2013}. This also makes it possible to analyses the
+predictions together as an assemblage.
 
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-ArranzOtaeguiRoe2023}
-Arranz-Otaegui, A., Roe, J., 2023. Revisiting the concept of the
-{``{Neolithic Founder Crops}''} in southwest {Asia}. Vegetation History
-and Archaeobotany. \url{https://doi.org/10.1007/s00334-023-00917-1}
-
-\bibitem[\citeproctext]{ref-AustinVanNiel2011}
-Austin, M.P., Van Niel, K.P., 2011. Improving species distribution
-models for climate change studies: Variable selection and scale. Journal
-of Biogeography 38, 1--8.
-\url{https://doi.org/10.1111/j.1365-2699.2010.02416.x}
-
-\bibitem[\citeproctext]{ref-BarbetMassinEtAl2012}
-Barbet-Massin, M., Jiguet, F., Albert, C.H., Thuiller, W., 2012.
-Selecting pseudo-absences for species distribution models: How, where
-and how many?: {How} to use pseudo-absences in niche modelling? Methods
-Ecol. Evol. 3, 327--338.
-\url{https://doi.org/10.1111/j.2041-210X.2011.00172.x}
-
-\bibitem[\citeproctext]{ref-BrownEtAl2018}
-Brown, J.L., Hill, D.J., Dolan, A.M., Carnaval, A.C., Haywood, A.M.,
-2018. {PaleoClim}, high spatial resolution paleoclimate surfaces for
-global land areas. Sci Data 5, 180254.
-\url{https://doi.org/10.1038/sdata.2018.254}
-
-\bibitem[\citeproctext]{ref-CamposEtAl2016}
-Campos, V.E., Cappa, F.M., Viviana, F.M., Giannoni, S.M., 2016. Using
-remotely sensed data to model suitable habitats for tree species in a
-desert environment. Journal of Vegetation Science 27, 200--210.
-\url{https://doi.org/10.1111/jvs.12328}
-
-\bibitem[\citeproctext]{ref-rgbif}
-Chamberlain, S., Barve, V., Mcglinn, D., Oldoni, D., Desmet, P.,
-Geffert, L., Ram, K., 2024.
-\href{https://CRAN.R-project.org/package=rgbif}{Rgbif: Interface to the
-global biodiversity information facility API}.
-
-\bibitem[\citeproctext]{ref-ChamberlainBoettiger2017}
-Chamberlain, S., Boettiger, C., 2017.
-\href{https://doi.org/10.7287/peerj.preprints.3304v1}{R python, and ruby
-clients for GBIF species occurrence data}. PeerJ PrePrints.
-
-\bibitem[\citeproctext]{ref-CollinsEtAl2018}
-Collins, C., Asouti, E., Grove, M., Kabukcu, C., Bradley, L.,
-Chiverrell, R., 2018. Understanding resource choice at the transition
-from foraging to farming: {An} application of palaeodistribution
-modelling to the {Neolithic} of the {Konya Plain}, south-central
-{Anatolia}, {Turkey}. J. Archaeol. Sci. 96, 57--72.
-\url{https://doi.org/10.1016/j.jas.2018.02.003}
-
-\bibitem[\citeproctext]{ref-ConollyEtAl2012}
-Conolly, J., Manning, K., Colledge, S., Dobney, K., Shennan, S., 2012.
-Species distribution modelling of ancient cattle from early {Neolithic}
-sites in {SW Asia} and {Europe}. Holocene 22, 997--1010.
-\url{https://doi.org/10.1177/0959683612437871}
-
-\bibitem[\citeproctext]{ref-DiVirgilioEtAl2018}
-Di Virgilio, G., Wardell-Johnson, G.W., Robinson, T.P., Temple-Smith,
-D., Hesford, J., 2018. Characterising fine-scale variation in plant
-species richness and endemism across topographically complex, semi-arid
-landscapes. Journal of Arid Environments 156, 59--68.
-\url{https://doi.org/10.1016/j.jaridenv.2018.04.005}
-
-\bibitem[\citeproctext]{ref-DubuisEtAl2013}
-Dubuis, A., Giovanettina, S., Pellissier, L., Pottier, J., Vittoz, P.,
-Guisan, A., 2013. Improving the prediction of plant species distribution
-and community composition by adding edaphic to topo-climatic variables.
-Journal of Vegetation Science 24, 593--606.
-\url{https://doi.org/10.1111/jvs.12002}
-
-\bibitem[\citeproctext]{ref-FranklinEtAl2015}
-Franklin, J., Potts, A.J., Fisher, E.C., Cowling, R.M., Marean, C.W.,
-2015. Paleodistribution modeling in archaeology and paleoanthropology.
-Quat. Sci. Rev. 110, 1--14.
-\url{https://doi.org/10.1016/j.quascirev.2014.12.015}
-
-\bibitem[\citeproctext]{ref-HenglEtAl2014}
-Hengl, T., de Jesus, J.M., MacMillan, R.A., Batjes, N.H., Heuvelink,
-G.B.M., Ribeiro, E., Samuel-Rosa, A., Kempen, B., Leenaars, J.G.B.,
-Walsh, M.G., Gonzalez, M.R., 2014. {SoilGrids1km--global} soil
-information based on automated mapping. PLoS One 9, e105992.
-\url{https://doi.org/10.1371/journal.pone.0105992}
-
-\bibitem[\citeproctext]{ref-HenglEtAl2017}
-Hengl, T., Mendes de Jesus, J., Heuvelink, G.B.M., Ruiperez Gonzalez,
-M., Kilibarda, M., Blagotić, A., Shangguan, W., Wright, M.N., Geng, X.,
-Bauer-Marschallinger, B., Guevara, M.A., Vargas, R., MacMillan, R.A.,
-Batjes, N.H., Leenaars, J.G.B., Ribeiro, E., Wheeler, I., Mantel, S.,
-Kempen, B., 2017. {SoilGrids250m}: {Global} gridded soil information
-based on machine learning. PLoS One 12, e0169748.
-\url{https://doi.org/10.1371/journal.pone.0169748}
-
-\bibitem[\citeproctext]{ref-HernandezEtAl2006}
-Hernandez, P.A., Graham, C.H., Master, L.L., Albert, D.L., 2006. The
-effect of sample size and species characteristics on performance of
-different species distribution modeling methods. Ecography 29, 773--785.
-\url{https://doi.org/10.1111/j.0906-7590.2006.04700.x}
-
-\bibitem[\citeproctext]{ref-HijmansEtAl2005}
-Hijmans, R.J., Cameron, S.E., Parra, J.L., Jones, P.G., Jarvis, A.,
-2005. Very high resolution interpolated climate surfaces for global land
-areas. Int. J. Climatol. 25, 1965--1978.
-\url{https://doi.org/10.1002/joc.1276}
-
-\bibitem[\citeproctext]{ref-KargerEtAl2017}
-Karger, D.N., Conrad, O., Böhner, J., Kawohl, T., Kreft, H., Soria-Auza,
-R.W., Zimmermann, N.E., Linder, H.P., Kessler, M., 2017. Climatologies
-at high resolution for the earth's land surface areas. Sci Data 4,
-170122. \url{https://doi.org/10.1038/sdata.2017.122}
-
-\bibitem[\citeproctext]{ref-KopeckyCizkova2010}
-Kopecký, M., Čížková, Š., 2010. Using topographic wetness index in
-vegetation ecology: Does the algorithm matter? Appl. Veg. Sci. 13,
-450--459. \url{https://doi.org/10.1111/j.1654-109X.2010.01083.x}
-
-\bibitem[\citeproctext]{ref-LeempoelEtAl2015}
-Leempoel, K., Parisod, C., Geiser, C., Daprà, L., Vittoz, P., Joost, S.,
-2015. Very high-resolution digital elevation models: Are multi-scale
-derived variables ecologically relevant? Methods in Ecology and
-Evolution 6, 1373--1383. \url{https://doi.org/10.1111/2041-210X.12427}
-
-\bibitem[\citeproctext]{ref-Lindsay2016}
-Lindsay, J.B., 2016. Whitebox {GAT}: {A} case study in geomorphometric
-analysis. Comput. Geosci. 95, 75--84.
-\url{https://doi.org/10.1016/j.cageo.2016.07.003}
-
-\bibitem[\citeproctext]{ref-LiuEtAl2013}
-Liu, C., White, M., Newell, G., 2013. Selecting thresholds for the
-prediction of species occurrence with presence-only data. Journal of
-Biogeography 40, 778--789. \url{https://doi.org/10.1111/jbi.12058}
-
-\bibitem[\citeproctext]{ref-MartinEtAl2016}
-Martin, L., Edwards, Y.H., Roe, J., Garrard, A., 2016. Faunal turnover
-in the {Azraq Basin}, eastern {Jordan} 28,000 to 9,000 cal {BP},
-signalling climate change and human impact. Quaternary Research 86,
-200--219. \url{https://doi.org/10.1016/j.yqres.2016.07.001}
-
-\bibitem[\citeproctext]{ref-MillerEtAl2024}
-Miller, T., Blackwood, C.B., Case, A.L., 2024. Assessing the utility of
-{SoilGrids250} for biogeographic inference of plant populations. Ecology
-and Evolution 14, e10986. \url{https://doi.org/10.1002/ece3.10986}
-
-\bibitem[\citeproctext]{ref-ModEtAl2016}
-Mod, H.K., Scherrer, D., Luoto, M., Guisan, A., 2016. What we use is not
-what we know: Environmental predictors in plant distribution models.
-Journal of Vegetation Science 27, 1308--1322.
-\url{https://doi.org/10.1111/jvs.12444}
-
-\bibitem[\citeproctext]{ref-StockwellPeterson2002}
-Stockwell, D.R.B., Peterson, A.T., 2002. Effects of sample size on
-accuracy of species distribution models. Ecological Modelling 148,
-1--13. \url{https://doi.org/10.1016/S0304-3800(01)00388-X}
-
-\bibitem[\citeproctext]{ref-ValaviEtAl2021}
-Valavi, R., Elith, J., Lahoz-Monfort, J.J., Guillera-Arroita, G., 2021.
-Modelling species presence-only data with random forests. Ecography 44,
-1731--1742. \url{https://doi.org/10.1111/ecog.05615}
-
-\bibitem[\citeproctext]{ref-ValaviEtAl2022}
-Valavi, R., Guillera-Arroita, G., Lahoz-Monfort, J.J., Elith, J., 2022.
-Predictive performance of presence-only species distribution models: A
-benchmark study with reproducible code. Ecological Monographs 92,
-e01486. \url{https://doi.org/10.1002/ecm.1486}
-
-\bibitem[\citeproctext]{ref-VelazcoEtAl2017}
-Velazco, S.J.E., Galvão, F., Villalobos, F., De Marco Júnior, P., 2017.
-Using worldwide edaphic data to model plant species niches: {An}
-assessment at a continental extent. PLoS One 12, e0186025.
-\url{https://doi.org/10.1371/journal.pone.0186025}
-
-\bibitem[\citeproctext]{ref-WiszEtAl2008}
-Wisz, M.S., Hijmans, R.J., Li, J., Peterson, A.T., Graham, C.H., Guisan,
-A., Group, N.P.S.D.W., 2008. Effects of sample size on the performance
-of species distribution models. Diversity and Distributions 14,
-763--773. \url{https://doi.org/10.1111/j.1472-4642.2008.00482.x}
-
-\bibitem[\citeproctext]{ref-WrightZiegler2017}
-Wright, M.N., Ziegler, A., 2017. Ranger: {A Fast Implementation} of
-{Random Forests} for {High Dimensional Data} in {C}++ and {R}. Journal
-of Statistical Software 77, 1--17.
-\url{https://doi.org/10.18637/jss.v077.i01}
-
-\bibitem[\citeproctext]{ref-YaworskyEtAl2023}
-Yaworsky, P.M., Hussain, S.T., Riede, F., 2023. Climate-driven habitat
-shifts of high-ranked prey species structure {Late Upper Paleolithic}
-hunting. Scientific Reports 13, 4238.
-\url{https://doi.org/10.1038/s41598-023-31085-x}
-
-\end{CSLReferences}
+\renewcommand\refname{References}
+  \bibliography{references.bib}
 

### 2025-01-05 analysis/paper.qmd
Add results

@@
 ---
-title: Biogeography of crop progenitors and wild plant resources in Late Pleistocene–Early Holocene West Asia, 21–8.3 ka
+title: Biogeography of crop progenitors and wild plant resources in the Late Pleistocene and Early Holocene of West Asia, 21–8.3 ka
 date: today
@@
   Species ranges are predicted to have been on average [X%] [larger|smaller] in 
-  the Early Holocene compared to present conditions, indicating [...]. 
-  The modelled ranges predict the observed occurrence of these species on 
-  archaeological sites with [low|medium|high] accuracy. The regional ubiquity of 
-  species in the archaeological record is [not] correlated with the predicted 
-  size of its range and the diversity of archaeobotanical assemblages is [not] 
+  the Early Holocene compared to present conditions, but did [not] change as
+  drastically between the Last Glacial Maximum and the Early Holocene. 
+  However, the modelled ranges do not predict the observed occurrence of these 
+  species on archaeological sites with any degree of accuracy. The regional 
+  ubiquity of species in the archaeological record is [not] correlated with the 
+  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
   correlated with the predicted diversity of its environs. This indicates that 
@@
     - [ ] Model assessment
-    - [ ] Hindcasting
   - [ ] Discussion
@@
 
+# TODO: drop LGM and HS1 since there are no sites 
+# (can then maybe increase prediction resolution?)
 climate_periods <- tibble(
@@
 
+ne_rivers <- read_sf(raw_data("ne"), "ne_10m_rivers_lake_centerlines_scale_rank")
+
 ne_lakes <- read_sf(raw_data("ne"), "ne_10m_lakes") |> 
@@
 ```
-
-```{r fig-region}
-#| fig-cap: Study region
-ggplot() + 
-  annotation_spatial(ne_land, fill = "#ffffff") +
-  layer_spatial(w_asia, fill = NA) +
-  coord_sf(crs = w_asia_albers)
-```
-
-<!-- TODO define study region and period -->
 
@@
 ```
+
+```{r data-archaeo-sites}
+archaeo_sites <- summarise(
+  archaeo_flora, 
+  latitude = first(latitude), 
+  longitude = first(longitude), 
+  .by = c(site_name, period)
+) |>
+  st_as_sf(coords = c("longitude", "latitude"), crs = latlong, remove = FALSE)
+```
+
+```{r fig-region}
+#| fig-cap: Study region (red) and archaeological sites used to generate modelled flora. Solid circles indicate Neolithic sites.
+# TODO: better pallette
+ggplot() + 
+  annotation_spatial(ne_land, fill = "#ffffff") +
+  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "blue") +
+  annotation_spatial(ne_lakes, colour = "blue", fill = "lightblue") +
+  layer_spatial(w_asia, colour = "red", fill = NA) +
+  layer_spatial(archaeo_sites, shape = 1) +
+  layer_spatial(filter(archaeo_sites, period %in% neolithic), shape = 16) +
+  scale_linewidth_identity() +
+  coord_sf(crs = w_asia_albers)
+```
+
+<!-- TODO define study region and period -->
 
@@
 climate_cube <- st_warp(climate, crs = w_asia_albers, 
-                        cellsize = c(pred_res, pred_res))
-terrain_cube <- st_warp(terrain_cube, climate_cube)
-soil_cube <- st_warp(soil_cube, climate_cube)
+                        cellsize = c(pred_res, pred_res), segments = 1000)
+terrain_cube <- st_warp(terrain_cube, climate_cube, segments = 1000)
+soil_cube <- st_warp(soil_cube, climate_cube, segments = 1000)
 
@@
 
-```{r extract-predictor-data}
+```{r data-flora-predictors}
 # TODO: factor out
@@
     sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
-    probability = TRUE
+    # probability = TRUE
   )
@@
 
-<!-- TODO: rewrite to remove thresholding -->
-The output of the model is probabilistic.
-However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to an estimate of habitat suitability.
-To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
-We select the threshold value for each model individually, using MaxSSS [as recommended by @LiuEtAl2013].
-This also makes it possible to analyses the predictions together as an assemblage.
+<!-- TODO: prediction, hindcasting and hindcasting assessment methodology -->
 
 # Results
-
-## Model assessment
-
-```{r fig-model-accuracy-vs-n_occ}
-#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
-flora |>
-  mutate(niche_test_fit = map(niche, collect_metrics)) |>
-  unnest(niche_test_fit) |>
-  filter(.metric == "accuracy") |> # TODO: or roc_auc?
-  mutate(label = if_else(.estimate < 0.75, taxon, NA_character_)) |>
-  ggplot(aes(n_occ, .estimate)) + 
-  geom_smooth(method = "glm", formula = y ~ log(x)) +
-  geom_point() + 
-  geom_text_repel(aes(label = label), fontface = "italic") +
-  scale_x_log10(name = "Occurrences") +
-  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
-  theme_half_open() 
-```
-
-* Modelled ecological niches on current data
-
-## Hindcasting
-
-Sensitivity to climate fluctuations?
 
@@
 ```
+
+We trained Random Forest models for `r nrow(flora)` taxa using contemporary occurrence data from GBIF, a random sample of background points, and the predictor variables described in @tbl-predictors.
+Substituting the "current" climate predictors for those derived from palaeoclimatic simulations [@BrownEtAl2018], we could then generate hindcast predictions for reconstructed past environments in `r english(nrow(climate_periods))` key climate periods – a total of `r nrow(flora) * nrow(climate_periods)` modelled palaeodistributions.
+Predicted distributions for individual taxa are presented in the appendix and accompanying material.
+
+## Model assessment
+
+```{r data-model-assessment}
+flora <- mutate(
+  flora,
+  niche_test_metrics = map(niche, collect_metrics)
+)
+
+niche_test_metrics <- flora |>
+  unnest(niche_test_metrics) |>
+  summarise(
+    min = min(.estimate),
+    max = max(.estimate),
+    mean = mean(.estimate),
+    sd = sd(.estimate),
+    .by = c(.metric)
+  )
+```
+
+```{r fig-model-accuracy-vs-n_occ}
+#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
+flora |>
+  unnest(niche_test_metrics) |>
+  filter(.metric == "accuracy") |>
+  mutate(label = if_else(.estimate < 0.80, taxon, NA_character_)) |>
+  ggplot(aes(n_occ, .estimate)) + 
+  geom_smooth(method = "glm", formula = y ~ log(x)) +
+  geom_point() + 
+  geom_text_repel(aes(label = label), fontface = "italic") +
+  scale_x_log10(name = "Occurrences") +
+  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
+  theme_minimal_grid() 
+```
+
+We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
+Model accuracy (i.e.
+proportion of correctly classified test samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
+The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_metrics, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_metrics, .metric == "accuracy")$sd, 3)`
+Model accuracy is loosely correlated with the number of occurrences available for training (@fig-model-accuracy-vs-n_occ), though the worst-performing models also included some species with relatively large sample sizes: <!-- TODO e.g.
+-->.
+Test metrics and ROC curves for the individual models are included in the appendix.
+
+```{r niche-archaeo}
+# TODO: factor out
+# TODO: what about negatives?
+archaeo_flora_predictions <- archaeo_flora |>
+  filter(taxon %in% flora$taxon) |>
+  st_as_sf(coords = c("longitude", "latitude"), crs = latlong) |>
+  mutate(
+    prediction = map(taxon, \(x) filter(flora, taxon == x)$prediction[[1]]),
+    prediction = map2(prediction, geometry, function(pred, geom) {
+      sf_point <- st_transform(st_sf(st_sfc(geom, crs = latlong)), w_asia_albers)
+      st_extract(pred, sf_point)
+    }),
+    prediction = map2(prediction, climate_period, function(pred, per) {
+      pred |>
+        as_tibble() |>
+        filter(as.character(period) == per) |>
+        transmute(
+          p_absent = prediction.FALSE.,
+          p_present = prediction.TRUE.,
+          pred_present = p_present > p_absent
+        )
+    })
+  ) |>
+  unnest(c(prediction))
+
+p_archaeo_pred_correct <- sum(archaeo_flora_predictions$pred_present, na.rm = TRUE) / length(!is.na(archaeo_flora_predictions$pred_present))
+
+archaeo_flora_pred_summary <- archaeo_flora_predictions |>
+  as_tibble() |>
+  summarise(
+    n_present = sum(!is.na(pred_present)),
+    n_predicted = sum(pred_present, na.rm = TRUE),
+    p_predicted = n_predicted / n_present,
+    .by = c(climate_period, site_name, phase_code, age_mid)
+  )
+```
+
+The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is significantly worse, with only `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted.
+<!-- TODO: negative -->
+
+# Discussion
+
+* General trends:
+  * Quantified sensitivity of plant ranges to climate change
+  * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
+  * How could are reconstructed ranges at predicting archaeological assemblages? (+Implications)
+* Interesting individual case studies:
+  * Wheat progenitors
+  * ???
+
+# Conclusion
+
+# References {.appendix}
+
+----
 
@@
 ```
-
-* Comparison to archaeological occurrences
 
@@
 ```
-
-# Discussion
-
-* General trends:
-  * Quantified sensitivity of plant ranges to climate change
-  * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
-  * How could are reconstructed ranges at predicting archaeological assemblages? (+Implications)
-* Interesting individual case studies:
-  * Wheat progenitors
-  * ???
-
-# Conclusion
-
-# References {.appendix}

### 2025-01-05 analysis/paper.tex
Add results

@@
 \hypersetup{
-  pdftitle={Biogeography of crop progenitors and wild plant resources in Late Pleistocene--Early Holocene West Asia, 21--8.3 ka},
+  pdftitle={Biogeography of crop progenitors and wild plant resources in the Late Pleistocene and Early Holocene of West Asia, 21--8.3 ka},
   pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
@@
 \begin{frontmatter}
-\title{Biogeography of crop progenitors and wild plant resources in Late
-Pleistocene--Early Holocene West Asia, 21--8.3 ka}
+\title{Biogeography of crop progenitors and wild plant resources in the
+Late Pleistocene and Early Holocene of West Asia, 21--8.3 ka}
 \author[1,2]{Joe Roe%
@@
 {[}larger\textbar smaller{]} in the Early Holocene compared to present
-conditions, indicating {[}\ldots{]}. The modelled ranges predict the
+conditions, but did {[}not{]} change as drastically between the Last
+Glacial Maximum and the Early Holocene. The modelled ranges predict the
 observed occurrence of these species on archaeological sites with
@@
 \begin{verbatim}
-Warning: Removed 51 rows containing missing values or values outside the scale range
+Warning: Removed 55 rows containing missing values or values outside the scale range
 (`geom_text_repel()`).
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Modelled ecological niches on current data
-\end{itemize}
+We assessed the predictive performance of the fitted niche models in the
+contemporary environment based on the reserved test partition. Model
+accuracy (i.e.~proportion of correctly classified test samples) ranged
+between 73\% and 98\%, with an average of 91\%. The area under the
+models' receiver operating characteristic curves (ROC-AUC) was in the
+range of 0.9734227±0.0596337 Model accuracy is loosely correlated with
+the number of occurrences available for training
+(Figure~\ref{fig-model-accuracy-vs-n_occ}), though the worst-performing
+models also included some species with relatively large sample sizes: .
+Test metrics and ROC curves for the individual models are included in
+the appendix.
+
+Overall we assessed all models as performing adequately enough to make
+reliable predictions on the desired scale, and that in general the
+majority of the models had high predictive power. Their predictive
+capability when applied to the past is a seperate question, which we
+address in the next section.
 
@@
 \item
-  Comparison to archaeological occurrences
+  Ability to predict archaeological occurrences
 \end{itemize}

### 2025-01-05 analysis/data/data-soilgrids.R

@@ -1,9 +1,15 @@
 # data-soilgrids.R
 # Retrieve SoilGrids <https://www.isric.org/explore/soilgrids> data for W. Asia
 # Based on https://git.wur.nl/isric/soilgrids/soilgrids.notebooks/-/blob/master/markdown/webdav_from_R_terra.md
-# NOTE: relies on an external API (last run 2024-11-28)
+# NOTE: relies on an external API (last run 2024-01-05)
+library("fs")
+library("purrr")
+library("sf")
 library("terra")
+library("whitebox")
+library("BadiaPaleoFloraENM")
 
+w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)
 soilgrids_vars <- c("clay", "sand", "silt", "phh2o")
 
 soilgrids <- map(soilgrids_vars, function(var, region) {
@@ -19,11 +25,19 @@ soilgrids <- map(soilgrids_vars, function(var, region) {
 
 names(soilgrids) <- soilgrids_vars
 
+# Write
+temp_path <- dir_create(path_temp("soilgrids"))
 walk2(
   soilgrids,
-  derived_data("soilgrids", paste0("soilgrids_", names(soilgrids), "_0-5cm_mean_w_asia.tif")),
+  path(temp_path, paste0("soilgrids_", names(soilgrids), "_0-5cm_mean_w_asia.tif")),
   writeRaster,
   datatype = "FLT4S",
   overwrite = TRUE
 )
 
+# Fill NA holes (mostly modern reservoirs) using IDW
+walk2(
+  dir_ls(temp_path),
+  derived_data("soilgrids", path_file(dir_ls(temp_path))),
+  wbt_fill_missing_data
+)

### 2025-01-05 analysis/paper.qmd
Add discussion of comparison with arch data
Removing LGM and HS1 as we don't have sites from them

@@
 ---
-title: Biogeography of crop progenitors and wild plant resources in the Late Pleistocene and Early Holocene of West Asia, 21–8.3 ka
+title: Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7–8.3 ka
 date: today
@@
   - [x] Methods & Materials
-  - [ ] Results
-    - [ ] Model assessment
+  - [x] Results
   - [ ] Discussion
@@
     - [ ] Case studies
+    - [x] Comparison with archaeobot
   - [ ] Conclusion
 - [ ] Figures
+  - [x] fig-region
   - [ ] tbl-occ-count (needs pdf cleanup)
   - [ ] tbl-predictors
-  - [ ] tbl-climate-periods
+- [ ] Outstanding TODOs
+- [ ] Appendix with all hindcast predictions
+- [ ] Copyedit and finalise references
 - [ ] Clean up package code
-- [ ] Appendix with all hindcast predictions
-- [ ] References & copyediting
 - [ ] Final proofread
@@
 
-# TODO: drop LGM and HS1 since there are no sites 
-# (can then maybe increase prediction resolution?)
 climate_periods <- tibble(
-  period = c("Last Glacial Maximum", "Heinrich Stadial 1", "Bølling-Allerød",
-             "Younger Dryas", "Early Holocene", "Current"),
-  code = c("lgm", "hs1", "ba", "yds", "eh", "cur"),
-  label = fct_inorder(c("LGM (21 ka)", "HS1 (17.0-14.7 ka)", "BA (14.7-12.9 ka)", 
-            "YDS (12.9-11.7 ka)", "EH (11.7-8.3 ka)", "current")),
-  start_bp = c(21000, 17000, 14700, 12900, 11700, NA),
-  end_bp = c(17000, 14700, 12900, 11700, 8300, NA)
+  period = c("Bølling-Allerød", "Younger Dryas", "Early Holocene", "Current"),
+  code = c("ba", "yds", "eh", "cur"),
+  label = fct_inorder(c("BA (14.7-12.9 ka)", "YDS (12.9-11.7 ka)", 
+                      "EH (11.7-8.3 ka)", "current")),
+  start_bp = c(14700, 12900, 11700, NA),
+  end_bp = c(12900, 11700, 8300, NA)
 )
@@
   * (In archaebot., not all intentionally collected)
-  * Regionally and temporally diverse
+  * Geographically and temporally diverse
   * ...so we model lots of species!
@@
   st_difference(ne_ocean) |>
-  st_difference(ne_lakes) |>
+  # st_difference(ne_lakes) |>
   st_union()
@@
 #| fig-cap: Study region (red) and archaeological sites used to generate modelled flora. Solid circles indicate Neolithic sites.
-# TODO: better pallette
 ggplot() + 
   annotation_spatial(ne_land, fill = "#ffffff") +
-  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "blue") +
-  annotation_spatial(ne_lakes, colour = "blue", fill = "lightblue") +
+  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
   layer_spatial(w_asia, colour = "red", fill = NA) +
@@
 
-```{r filter-flora}
+```{r data-filter-flora}
 flora <- filter(flora, n_occ > min_n_occ)
@@
 
-```{r fetch-occ}
+```{r data-occ}
 # TODO: further tidying. Ideas:
@@
 #| fig-cap: Palaeoclimatic periods used for hindcasting, after @BrownEtAl2018
-# TODO: add refs (as footnotes?)
+# TODO: integrate into text
 climate_periods |>
@@
 
-```{r fit-niche-models}
+```{r niche-fit}
 # Fit best model to training data
@@
 
-```{r predict}
-# NOTE: very high memory usage
+```{r niche-predict}
+# NOTE: high memory usage
 if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
@@
   flora,
-  niche_test_metrics = map(niche, collect_metrics)
-)
+  pred = map(niche, collect_predictions),
+  niche_test_metrics = map(niche, collect_metrics),
+  sensitivity = map(pred, sens, truth = present, estimate = .pred_class,
+                    event_level = "second"), # reference present=TRUE
+  niche_test_metrics = map2(niche_test_metrics, sensitivity, bind_rows)
+) |>
+  select(-sensitivity)
 
@@
 
-```{r fig-model-accuracy-vs-n_occ}
-#| fig-cap: Number of taxon occurrences and model accuracy on test dataset
+```{r fig-model-sensitivity-vs-n_occ}
+#| fig-cap: Number of taxon occurrences and model sensitivity on test dataset
 flora |>
   unnest(niche_test_metrics) |>
-  filter(.metric == "accuracy") |>
-  mutate(label = if_else(.estimate < 0.80, taxon, NA_character_)) |>
+  filter(.metric == "sens") |>
+  mutate(label = if_else(.estimate < 0.85, taxon, NA_character_)) |>
   ggplot(aes(n_occ, .estimate)) + 
@@
   scale_x_log10(name = "Occurrences") +
-  scale_y_continuous(name = "Model accuracy", labels = label_percent()) +
+  scale_y_continuous(name = "Sensitivity", labels = label_percent()) +
   theme_minimal_grid() 
@@
 We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
-Model accuracy (i.e.
-proportion of correctly classified test samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
+Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
+Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_metrics, .metric == "sens")$min)` and `r percent(filter(niche_test_metrics, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "sens")$mean)`.
 The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_metrics, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_metrics, .metric == "accuracy")$sd, 3)`
-Model accuracy is loosely correlated with the number of occurrences available for training (@fig-model-accuracy-vs-n_occ), though the worst-performing models also included some species with relatively large sample sizes: <!-- TODO e.g.
--->.
+Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
 Test metrics and ROC curves for the individual models are included in the appendix.
@@
   * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
-  * How could are reconstructed ranges at predicting archaeological assemblages? (+Implications)
 * Interesting individual case studies:
@@
 
+## Comparison with archaeobotanical data
+
+The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations.
+Since they do accurately predict the test dataset <!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
+This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
+Another obvious flaw in our methodology is that the time slices used for palaeoclimatic reconstruction are very broad—each covering around two millennia—and therefore potentially unrepresentative of the environment around sites at the specific time at which they were occupied.
+The variable quality of the archaeological test dataset, especially in terms of chronology, is also a plausible factor.
+
+At the same time, we cannot rule out more substantive reasons for the discrepency between predicted and observed archaeological occurrences.
+The niches of the modelled species could have changed since the Early Holocene, which would not be captured in a model trained purely on modern specimens.
+Human economic choices—mobility, foraging strategies, cultivation, etc.—could also produce archaeobotanical assemblages whose composition depart significantly from that of the surrounding local flora.
+Further refinement of the methodology for hindcast palaeoecological niche models, for example using more finely resolved palaeoclimate sequences [e.g.
+@KargerEtAl2023], hyperparameter tuning to avoid overfitting, and improved archaeological datasets, would help disentangle these potential explanations.
+
+<!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
+
 # Conclusion
+
+* We present the first comprehensive reconstructions of the palaeodistributions of 56 plant species found regularly in association with early agricultural archaeological sites in West Asia
+  * R
+* Modelling at scale using random forest, modern occurrences, and hindcasting represents a significant advance in pENM methodology
+  * Relies on recent open ecological and climatic datasets
+  * ...open archaeological datasets still lacking!
+* First (? - check Yaworsky) attempt to verify hindcast models with archaeo. compositional data
+  * Results not to promising, but this doesn't mean the models are useless!
+  * Discrepencies suggests several areas for further research and methodological development
 

### 2025-01-05 analysis/paper.tex
Add discussion of comparison with arch data
Removing LGM and HS1 as we don't have sites from them

--- 2025-01-04 analysis/paper.tex
@@
 \hypersetup{
-  pdftitle={Biogeography of crop progenitors and wild plant resources in Late Pleistocene--Early Holocene West Asia, 21--8.3 ka},
+  pdftitle={Biogeography of crop progenitors and wild plant resources over the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka},
   pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
@@
 \begin{frontmatter}
-\title{Biogeography of crop progenitors and wild plant resources in Late
-Pleistocene--Early Holocene West Asia, 21--8.3 ka}
+\title{Biogeography of crop progenitors and wild plant resources over
+the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka}
 \author[1,2]{Joe Roe%
@@
 {[}larger\textbar smaller{]} in the Early Holocene compared to present
-conditions, indicating {[}\ldots{]}. The modelled ranges predict the
-observed occurrence of these species on archaeological sites with
-{[}low\textbar medium\textbar high{]} accuracy. The regional ubiquity of
-species in the archaeological record is {[}not{]} correlated with the
-predicted size of its range and the diversity of archaeobotanical
-assemblages is {[}not{]} correlated with the predicted diversity of its
-environs. This indicates that trends in taxonomic composition of the
-archaeobotanical record is {[}not{]} likely to have been influenced by
-environmental change and species turnover, in addition to human economic
-choices.
+conditions, but did {[}not{]} change as drastically between the Last
+Glacial Maximum and the Early Holocene. However, the modelled ranges do
+not predict the observed occurrence of these species on archaeological
+sites with any degree of accuracy. The regional ubiquity of species in
+the archaeological record is {[}not{]} correlated with the predicted
+size of its range and the diversity of archaeobotanical assemblages is
+{[}not{]} correlated with the predicted diversity of its environs. This
+indicates that trends in taxonomic composition of the archaeobotanical
+record is {[}not{]} likely to have been influenced by environmental
+change and species turnover, in addition to human economic choices.
 \end{abstract}
@@
       Model assessment
-    \item[$\square$]
-      Hindcasting
     \end{itemize}
@@
 
-\caption{\label{fig-region}Study region}
+\caption{\label{fig-region}Study region (red) and archaeological sites
+used to generate modelled flora. Solid circles indicate Neolithic
+sites.}
 
@@
 \midrule\addlinespace[2.5pt]
-Last Glacial Maximum (LGM) & 21.0–17.0 \\ 
-Heinrich Stadial 1 (HS1) & 17.0–14.7 \\ 
 Bølling-Allerød (BA) & 14.7–12.9 \\ 
@@
 
-For hindcasting, we used reconstructed bioclimatic data for 5 key
+For hindcasting, we used reconstructed bioclimatic data for 3 key
 periods (Table~\ref{tbl-climate-periods}) generated from downscaled
@@
 
-The output of the model is probabilistic. However, this should not be
-understood as an actual probability of occurrence \citep{CITE}, but more
-akin to an estimate of habitat suitability. To simplify interpretation,
-we can convert these predictions into binary presence/absence maps, a
-process called ``thresholding''. We select the threshold value for each
-model individually, using MaxSSS \citep[as recommended
-by][]{LiuEtAl2013}. This also makes it possible to analyses the
-predictions together as an assemblage.
+\section{Results}\label{results}
 
-\section{Results}\label{results}
+We trained Random Forest models for 56 taxa using contemporary
+occurrence data from GBIF, a random sample of background points, and the
+predictor variables described in \textbf{?@tbl-predictors}. Substituting
+the ``current'' climate predictors for those derived from palaeoclimatic
+simulations \citep{BrownEtAl2018}, we could then generate hindcast
+predictions for reconstructed past environments in 4 key climate periods
+-- a total of 224 modelled palaeodistributions. Predicted distributions
+for individual taxa are presented in the appendix and accompanying
+material.
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-model-accuracy-vs-n_occ-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-model-sensitivity-vs-n_occ-1.pdf}}
 
@@
 
-\caption{\label{fig-model-accuracy-vs-n_occ}Number of taxon occurrences
-and model accuracy on test dataset}
+\caption{\label{fig-model-sensitivity-vs-n_occ}Number of taxon
+occurrences and model sensitivity on test dataset}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Modelled ecological niches on current data
-\end{itemize}
+We assessed the predictive performance of the fitted niche models in the
+contemporary environment based on the reserved test partition. Model
+accuracy (proportion of correctly classified presence and background
+samples) ranged between 72\% and 98\%, with an average of 91\%.
+Sensitivity (proportion of correctly classified presence samples) ranged
+between 80\% and 100\%, with an average of 93\%. The area under the
+models' receiver operating characteristic curves (ROC-AUC) was in the
+range of 0.973±0.058 Model sensitivity is loosely correlated with the
+number of occurrences available for training
+(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
+models all having less than 300 recorded occurrences: \emph{Buglossoides
+tenuiflora}, \emph{Cephalaria syriaca}, \emph{Gypsophila pilosa},
+\emph{Lathyrus sativus}, \emph{Triticum turgidum dicoccum}. Test metrics
+and ROC curves for the individual models are included in the appendix.
 
-\subsection{Hindcasting}\label{hindcasting}
-
-Sensitivity to climate fluctuations?
-
-\begin{itemize}
-\tightlist
-\item
-  Comparison to archaeological occurrences
-\end{itemize}
+The ability of the hindcast models to predict the occurrence of specific
+species at archaeological sites is significantly worse, with only 12\%
+of presences in archaeobotanical assemblages successfully predicted.
 
@@
     agriculture? (Moreso than other wild resources??)
-  \item
-    How could are reconstructed ranges at predicting archaeological
-    assemblages? (+Implications)
   \end{itemize}
@@
 
+\subsection{Comparison with archaeobotanical
+data}\label{comparison-with-archaeobotanical-data}
+
+The failure of our hindcast models to predict the occurrence of species
+in archaeobotanical assemblages has several possible explanations. Since
+they do accurately predict the test dataset , a likely culprit is
+overfitting of the models to the present environment. This implies that
+the modelled palaeodistributions should be seen as conservative
+estimates or a minimal range. Another obvious flaw in our methodology is
+that the time slices used for palaeoclimatic reconstruction are very
+broad---each covering around two millennia---and therefore potentially
+unrepresentative of the environment around sites at the specific time at
+which they were occupied. The variable quality of the archaeological
+test dataset, especially in terms of chronology, is also a plausible
+factor.
+
+At the same time, we cannot rule out more substantive reasons for the
+discrepency between predicted and observed archaeological occurrences.
+The niches of the modelled species could have changed since the Early
+Holocene, which would not be captured in a model trained purely on
+modern specimens. Human economic choices---mobility, foraging
+strategies, cultivation, etc.---could also produce archaeobotanical
+assemblages whose composition depart significantly from that of the
+surrounding local flora. Further refinement of the methodology for
+hindcast palaeoecological niche models, for example using more finely
+resolved palaeoclimate sequences \citep{KargerEtAl2023}, hyperparameter
+tuning to avoid overfitting, and improved archaeological datasets, would
+help disentangle these potential explanations.
+
 \section{Conclusion}\label{conclusion}
 
-\renewcommand\refname{References}
+\section{References}\label{references}
+
+\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+
   \bibliography{references.bib}

### 2025-01-05 analysis/paper.qmd
Add discussion

@@
 abstract: |
-  This paper presents the first comprehensive reconstructions of the 
-  palaeodistributions of 56 plant species found regularly in association with
-  early agricultural archaeological sites in West Asia, including the 
+  This paper presents the first continuous, spatially-explicit reconstructions 
+  of the palaeodistributions of 56 plant species found regularly in association 
+  with early agricultural archaeological sites in West Asia, including the 
   progenitors of the first crops. We used machine learning to train an 
@@
   potential ranges of these species at key stages of the Pleistocene–Holocene 
-  transition could then be derived from these models using hindcast data from 
+  transition could then be derived from these models using downsampled data from 
   palaeoclimate simulations.
-  Species ranges are predicted to have been on average [X%] [larger|smaller] in 
-  the Early Holocene compared to present conditions, but did [not] change as
-  drastically between the Last Glacial Maximum and the Early Holocene. 
-  However, the modelled ranges do not predict the observed occurrence of these 
-  species on archaeological sites with any degree of accuracy. The regional 
+  Species ranges are predicted to have been on significantly smaller in the 
+  Early Holocene compared to present conditions, with many species that are 
+  found throughout the region's 'hilly flanks' today indicated to have much more
+  restricted distributions centered on the Levant, Cyprus and Western Anatolia. 
+  Ranges shrunk by on average c. 10% from the terminal Pleistocene to the Early 
+  Holocene, with a further c. 10% reduction for the duration of the Younger 
+  Dryas. However, modelled ranges do not reliably predict the observed 
+  occurrence of specific species at archaeological sites. [...]
+  The regional 
   ubiquity of species in the archaeological record is [not] correlated with the 
@@
 
-* [x] Finish modelling
+* [x] Stop modelling
 - [ ] First draft:
@@
   - [x] fig-region
+  - [x] fig-occ
   - [ ] tbl-occ-count (needs pdf cleanup)
@@
 - [ ] Outstanding TODOs
-- [ ] Appendix with all hindcast predictions
-- [ ] Copyedit and finalise references
+- [ ] Add missing references
+- [ ] Copyedit
+- [ ] Appendix with all hindcast predictions (HTML?)
 - [ ] Clean up package code
@@
 So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
-Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.
+Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.
 
@@
 ```{r fig-occ-map}
-# TODO: Can this be made presentable?
-#flora |>
-#  unnest(occ) |>
-#  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
-#  ggplot() + 
-#  facet_wrap(vars(taxon)) +
-#  layer_spatial(ne_land, fill = "#ffffff") +
-#  geom_sf(shape = ".") +
-#  coord_sf(xlim = w_asia[c("xmin", "xmax")], 
-#           ylim = w_asia[c("ymin", "ymax")],
-#           crs = w_asia_albers,
-#           default_crs = latlong)
+#| fig-cap: Georeferenced occurrence records from West Asia for the `r nrow(flora)` modelled taxa (N=`r nrow(unnest(flora, occ))`)
+flora |>
+ unnest(occ) |>
+ distinct(latitude, longitude) |>
+ st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
+ ggplot() +
+ annotation_spatial(ne_land, fill = "#ffffff") +
+ geom_sf(shape = ".") +
+ layer_spatial(w_asia, fill = NA) +
+ coord_sf(crs = w_asia_albers, default_crs = latlong)
 ```
@@
 A substantial literature exists on the relatively performance of these approaches and their respective parameterisations [reviewed in @ValaviEtAl2022].
-Random Forest, a widely-used machine learning algorithm, is amongst the best performing approaches for presence-only species distribution models, providing it is appropriately parameterised to account for the class imbalance between presence and background samples [@ValaviEtAl2021; @ValaviEtAl2022].
+Random Forest, a widely-used machine learning algorithm, is amongst the best performing methods for presence-only species distribution models, providing it is appropriately parameterised to account for the class imbalance between presence and background samples [@ValaviEtAl2021; @ValaviEtAl2022].
 For our application, it also has the advantage of requiring little to no manual parameter tuning to achieve good predictive results, which makes it easier to model a larger numbers of taxa.
@@
 
+```{r niche-pred-range}
+area_present <- function(pred, res = pred_res) {
+  st_apply(pred, "period", sum, na.rm = TRUE)["pred_present"] |>
+    data.frame() |>
+    mutate(area_present = pred_present * (pred_res)^2) |>
+    select(period, area_present)
+}
+
+flora <- flora |>
+  mutate(
+    # Binarize
+    prediction = map(prediction, mutate, pred_present = prediction.TRUE. > 0.5),
+    area_present = map(prediction, area_present)
+  )
+```
+
 # Discussion
 
-* General trends:
-  * Quantified sensitivity of plant ranges to climate change
-  * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
-* Interesting individual case studies:
-  * Wheat progenitors
-  * ???
-
-## Comparison with archaeobotanical data
+## Reduction in range sizes over the Pleistocene/Holocene boundary
+
+```{r fig-range-density}
+#| fig-cap: Distribution of predicted species ranges by period. Dashed lines indicate the median range.
+flora_pred_area_present_summary <- flora |>
+  unnest(area_present) |>
+  summarise(med_area_present = median(area_present) / 1e6, .by = period) |>
+  mutate(
+    period_label = map(period, \(x) filter(climate_periods, code == x)$label),
+    period_label = unlist(period_label),
+  )
+
+flora |>
+  unnest(area_present) |> 
+  mutate(
+    period_label = map(period, \(x) filter(climate_periods, code == x)$label),
+    period_label = unlist(period_label),
+    area_present = area_present / 1e6 # m2 to km2
+  ) |> 
+  ggplot() + 
+  facet_wrap(vars(period_label), ncol = 1, strip.position = "left") + 
+  geom_density(aes(area_present)) +
+  geom_vline(aes(xintercept = med_area_present), flora_pred_area_present_summary) +
+  scale_x_log10(name = "Predicted range km²", labels = label_number_auto()) + 
+  scale_y_continuous(name = NULL, labels = NULL) +
+  theme_minimal_vgrid() +
+  theme(
+    strip.text.y.left = element_text(angle = 0, hjust = 0),
+    axis.line.y = element_blank(),
+    axis.ticks.y = element_blank()
+  )
+```
+
+Species ranges are predicted to have been signicantly smaller in the terminal Pleistocene and Early Holocene that under current conditions [@fig-range-density], though the magnitude of this change is likely also to reflect a degree of overfitting in the model (see below).
+Fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable.
+The average range of modelled species shrank by `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` over the Pleistocene–Holocene boundary, with a noticeable dip in the Younger Dryas.
+This perhaps indicates that although this period is considered one of climatatic amelioration globally [@JonesEtAl2019], the colder conditions of the Pleistocene may have supported more extensive plant-based economies in West Asia specifically.
+
+<!-- TODO: map of sum species abundance? -->
+
+Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene.
+These include *Ficus carica* (fig);
+*Hordeum* spp. (wild barleys);
+*Lathyrus aphaca* and *L. sativus* (both marginally edible legumes);
+*Triticum aestivum compactum* (in the N. Levant), *T. monococcum aegilopoides*, *T. durum*, and *Triticum urartu* (but not the other wheat progenitor, *T. turgidum dicoccum* – see below);
+*Aegilops speltoides*, but not *Aegilops tauschii* (goatgrasses);
+and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
+Central Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
+leaving the Levant and to a lesser extent Western Anatolia Cyprus as refugia.
+
+Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @] and as a centre of origin of agriculture [@].
+It should also be noted that the selection of flora for this study (@tbl-occ-count) is biased towards Levantine fauna due to the history of archaeological research in the region,
+as is the occurrence dataset used to train the model [@fig-occ-map].
+Within the Levant, retreat from the Badia.
+
+Loss of the Northern Mesopotamia–Southeastern Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@KozlowskiAurenche2005], which puts this region at the centre of the development of agriculture.
+Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication.
+Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [@cite].
+But since these studies are based on modern genomes, if the wild range of these plants has, as our modelling suggests, shifted since the Early Holocene, the apparent centre of diversity may have shifted with them.
+Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant)
+and more broadly there need not be a link between the core *wild* range of a plant and the core area of its domestication. 
+A scenario in which cultivation emerged at the edges of the ranges of valuable plant resources—as a means of extend their natural niche—is also plausible.
+
+The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that domestication took place just as early in the eastern *Mashriq* as it did in the west [@Braidwood; @GanjDarehGoats; @HillyFlanksStuff].
+We consider that the most likely explanation for this is that our flora does not include the species that were most important to plant subsistence in the east.
+Archaeobotanical data on Neolithic sites in the Zagros is limited (compared to the Levant in particular) due to a hiatus in field research there from the 1980s to early 2010s [@cite].
+Recent research [@Amaia?] indicates ... <!-- TODO: weird plants -->
+
+Cyprus and W. Anatolia not considered part of the core zone – maybe they should be?
+
+Exceptions to the trend of range reduction include *Cicer reticulatum* (wild chickpea), which has a relatively stable range centered on Northern Mesopotamia;
+and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to have two very restricted ranges centered around the Black Sea Coast of Anatolia and the Palmyra basin.
+In the latter case, neither of these areas are part of the predicted modern distribution of wild emmer (centered around the Caucasus and Northern Mesopotamia), but it would be consistent with archaeological evidence for early cultivation at Tell Abu Hureyra and other sites in the Upper Euphrates.
+On the other hand, the model for *Triticum turgidum dicoccum* has the lowest predictive performance of all our fitted models.
+
+## Biogeography of crop progenitors
+
+Several crop progenitors reduced in range before cultivation?
+Reduction in range from BA to EH very evident in Hordeum, Pistacia atlantica.
+
+The wheat story.
+
+Flax has a very restricted distribution (consistent with low occurrence in founders paper?).
+As does Pistacia atlantica, Bolboschoenus maritimus
+
+*Secale cereale* = an Anatolian boy
+
+## Hindcast models do not predict archaeobotanical composition
 
@@
 
-* We present the first comprehensive reconstructions of the palaeodistributions of 56 plant species found regularly in association with early agricultural archaeological sites in West Asia
-  * R
+* We present the first continuous, spatially explicit models of the palaeodistributions of 56 plant species found regularly in association with early agricultural archaeological sites in West Asia
+  * A new line of evidence on archaeoecology
+  * Complementary to archaeobot/pollen/etc. because it is independent from it
+  * All models are wrong... but it's easier to see how these are wrong than lines on maps
 * Modelling at scale using random forest, modern occurrences, and hindcasting represents a significant advance in pENM methodology
@@
 ```{r fig-niche-prediction, eval=FALSE}
-niche_taxon <- "Linum bienne"
+# "Adonis flammea"                   "Aegilops crassa"                  "Aizoanthemopsis hispanica"       
+# "Ammi majus"                       "Androsace maxima"                 "Arnebia decumbens"               
+# "Arnebia linearifolia"             "Atriplex prostrata"               "Bassia arabica"                  
+# "Bolboschoenus maritimus"          "Brachypodium distachyon"          "Bromus sterilis"                 
+# "Buglossoides arvensis"            "Buglossoides tenuiflora"          "Capparis spinosa"                
+# "Carex divisa"                     "Cephalaria syriaca"               "Chenopodium album"               
+# "Cicer reticulatum"                "Ficus carica"                     "Fumaria densiflora"              
+# "Gypsophila pilosa"                "Helianthemum ledifolium"          "Hordeum bulbosum"                
+# "Hordeum murinum"                  "Hordeum spontaneum"               "Lathyrus aphaca"                 
+# "Lathyrus sativus"                 "Vicia orientalis"                 "Linum bienne"                    
+# "Lolium rigidum"                   "Lolium temulentum"                "Medicago radiata"                
+# "Phalaris paradoxa"                "Pistacia atlantica"               "Lathyrus oleraceus"              
+# "Poa bulbosa"                      "Prosopis farcta"                  "Quercus ithaburensis"            
+# "Rumex pulcher"                    "Salsola kali"                     "Secale cereale"                  
+# "Taeniatherum caput-medusae"       "Medicago astroites"               "Triticum aestivum compactum"     
+# "Triticum turgidum dicoccum"       "Triticum monococcum aegilopoides" "Triticum durum"                  
+# "Gypsophila vaccaria"              "Verbena officinalis"              "Vicia ervilia"                   
+# "Vicia faba"                       "Vicia narbonensis"                "Triticum urartu"                 
+# "Aegilops speltoides"              "Aegilops tauschii"
+
+niche_taxon <- "Aegilops tauschii"
 niche_prediction <- filter(flora, taxon == niche_taxon)$prediction[[1]]
@@
 ggplot() + 
-  facet_wrap(~period, ncol = 3) +
+  facet_wrap(~period, ncol = 2) +
   annotation_spatial(ne_land, fill = "white") +
@@
              na.action = na.omit) +
+  annotation_spatial(ne_land, fill = NA) +
   annotation_spatial(w_asia, fill = NA) +

### 2025-01-05 analysis/paper.tex
Add discussion

@@
 \hypersetup{
-  pdftitle={Biogeography of crop progenitors and wild plant resources over the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka},
+  pdftitle={Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3 ka},
   pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
@@
 \begin{frontmatter}
-\title{Biogeography of crop progenitors and wild plant resources over
-the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka}
+\title{Biogeography of crop progenitors and wild plant resources over in
+terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3
+ka}
 \author[1,2]{Joe Roe%
@@
 \begin{abstract}
-This paper presents the first comprehensive reconstructions of the
-palaeodistributions of 56 plant species found regularly in association
-with early agricultural archaeological sites in West Asia, including the
-progenitors of the first crops. We used machine learning to train an
-ecological niche model of each species based on its present-day
-distribution in relation to climate and environmental variables.
-Predictions of the potential ranges of these species at key stages of
-the Pleistocene--Holocene transition could then be derived from these
-models using hindcast data from palaeoclimate simulations. Species
-ranges are predicted to have been on average {[}X\%{]}
-{[}larger\textbar smaller{]} in the Early Holocene compared to present
-conditions, but did {[}not{]} change as drastically between the Last
-Glacial Maximum and the Early Holocene. However, the modelled ranges do
-not predict the observed occurrence of these species on archaeological
-sites with any degree of accuracy. The regional ubiquity of species in
-the archaeological record is {[}not{]} correlated with the predicted
-size of its range and the diversity of archaeobotanical assemblages is
-{[}not{]} correlated with the predicted diversity of its environs. This
-indicates that trends in taxonomic composition of the archaeobotanical
-record is {[}not{]} likely to have been influenced by environmental
-change and species turnover, in addition to human economic choices.
+This paper presents the first continuous, spatially-explicit
+reconstructions of the palaeodistributions of 56 plant species found
+regularly in association with early agricultural archaeological sites in
+West Asia, including the progenitors of the first crops. We used machine
+learning to train an ecological niche model of each species based on its
+present-day distribution in relation to climate and environmental
+variables. Predictions of the potential ranges of these species at key
+stages of the Pleistocene--Holocene transition could then be derived
+from these models using downsampled data from palaeoclimate simulations.
+Species ranges are predicted to have been on significantly smaller in
+the Early Holocene compared to present conditions, with many species
+that are found throughout the region's `hilly flanks' today indicated to
+have much more restricted distributions centered on the Levant, Cyprus
+and Western Anatolia. Ranges shrunk by on average c.~10\% from the
+terminal Pleistocene to the Early Holocene, with a further c.~10\%
+reduction for the duration of the Younger Dryas. However, modelled
+ranges do not reliably predict the observed occurrence of specific
+species at archaeological sites. {[}\ldots{]} The regional ubiquity of
+species in the archaeological record is {[}not{]} correlated with the
+predicted size of its range and the diversity of archaeobotanical
+assemblages is {[}not{]} correlated with the predicted diversity of its
+environs. This indicates that trends in taxonomic composition of the
+archaeobotanical record is {[}not{]} likely to have been influenced by
+environmental change and species turnover, in addition to human economic
+choices.
 \end{abstract}
@@
 \item[$\boxtimes$]
-  Finish modelling
+  Stop modelling
 \item[$\square$]
@@
     Methods \& Materials
-  \item[$\square$]
+  \item[$\boxtimes$]
     Results
-
-    \begin{itemize}
-    \tightlist
-    \item[$\square$]
-      Model assessment
-    \end{itemize}
   \item[$\square$]
@@
       Case studies
+    \item[$\boxtimes$]
+      Comparison with archaeobot
     \end{itemize}
@@
   \tightlist
+  \item[$\boxtimes$]
+    fig-region
+  \item[$\boxtimes$]
+    fig-occ
   \item[$\square$]
@@
     tbl-predictors
-  \item[$\square$]
-    tbl-climate-periods
   \end{itemize}
 \item[$\square$]
+  Outstanding TODOs
+\item[$\square$]
+  Add missing references
+\item[$\square$]
+  Copyedit
+\item[$\square$]
+  Appendix with all hindcast predictions (HTML?)
+\item[$\square$]
   Clean up package code
-\item[$\square$]
-  Appendix with all hindcast predictions
-\item[$\square$]
-  References \& copyediting
 \item[$\square$]
@@
   \item
-    Regionally and temporally diverse
+    Geographically and temporally diverse
   \item
@@
 Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
-criteria were substituted with their wild progenitor(s), where known.
+criteria were substituted with their wild progenitor(s), where
+different.
 
@@
 
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-occ-map-1.pdf}}
+
+}
+
+\caption{\label{fig-occ-map}Georeferenced occurrence records from West
+Asia for the \texttt{r\ nrow(flora)} modelled taxa
+(N=\texttt{r\ nrow(unnest(flora,\ occ))})}
+
+\end{figure}%
+
 Occurrence data only tells us where a species is present; there is
@@
 in][]{ValaviEtAl2022}. Random Forest, a widely-used machine learning
-algorithm, is amongst the best performing approaches for presence-only
+algorithm, is amongst the best performing methods for presence-only
 species distribution models, providing it is appropriately parameterised
@@
 tenuiflora}, \emph{Cephalaria syriaca}, \emph{Gypsophila pilosa},
-\emph{Lathyrus sativus}, \emph{Triticum turgidum dicoccum}. Test metrics
-and ROC curves for the individual models are included in the appendix.
+\emph{Lathyrus sativus}, and \emph{Triticum turgidum dicoccum}. Test
+metrics and ROC curves for the individual models are included in the
+appendix.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  General trends:
+\subsection{Reduction in range sizes over the Pleistocene/Holocene
+boundary}\label{reduction-in-range-sizes-over-the-pleistoceneholocene-boundary}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Quantified sensitivity of plant ranges to climate change
-  \item
-    Crop progenitors saw range contractions just before the onset of
-    agriculture? (Moreso than other wild resources??)
-  \end{itemize}
-\item
-  Interesting individual case studies:
+\begin{figure}
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Wheat progenitors
-  \item
-    ???
-  \end{itemize}
-\end{itemize}
+\centering{
 
-\subsection{Comparison with archaeobotanical
-data}\label{comparison-with-archaeobotanical-data}
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-range-density-1.pdf}}
+
+}
+
+\caption{\label{fig-range-density}Distribution of predicted species
+ranges by period. Dashed lines indicate the median range.}
+
+\end{figure}%
+
+Species ranges are predicted to have been signicantly smaller in the
+terminal Pleistocene and Early Holocene that under current conditions
+Figure~\ref{fig-range-density}, though the magnitude of this change is
+likely also to reflect a degree of overfitting in the model (see below).
+Fluctuations in modelled range size between the Bølling-Allerød
+(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
+(11.7--8.3 ka) are more directly comparable. The average range of
+modelled species shrank by 12\% over the Pleistocene--Holocene boundary,
+with a noticeable dip in the Younger Dryas. This perhaps indicates that
+although this period is considered one of climatatic amelioration
+globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
+may have supported more extensive plant-based economies in West Asia
+specifically.
+
+Many taxa that occur (or are predicted to occur) across the `hilly
+flanks' today---including most crop progenitors---are reconstructed to
+have had a significantly more restricted distribution in the terminal
+Pleistocene/Early Holocene. These include \emph{Ficus carica} (fig);
+\emph{Hordeum} spp. (wild barleys); \emph{Lathyrus aphaca} and \emph{L.
+sativus} (both marginally edible legumes); \emph{Triticum aestivum
+compactum} (in the N. Levant), \emph{T. monococcum aegilopoides},
+\emph{T. durum}, and \emph{Triticum urartu} (but not the other wheat
+progenitor, \emph{T. turgidum dicoccum} -- see below); \emph{Aegilops
+speltoides}, but not \emph{Aegilops tauschii} (goatgrasses); and
+\emph{Vicia} spp. (vetches), including \emph{Vicia faba} (broad beans).
+Central Anatolia, Northern Mesopotamia, and the Zagros Mountains in
+particular disappear from the predicted ranges of these species, leaving
+the Levant and to a lesser extent Western Anatolia Cyprus as refugia.
+
+Our results for the Levant are consistent with the current understanding
+of this region as developing early intensive foraging economies {[}the
+Natufian culture, @{]} and as a centre of origin of agriculture {[}@{]}.
+It should also be noted that the selection of flora for this study
+(Table~\ref{tbl-occ-count}) is biased towards Levantine fauna due to the
+history of archaeological research in the region, as is the occurrence
+dataset used to train the model Figure~\ref{fig-occ-map}. Within the
+Levant, retreat from the Badia.
+
+Loss of the Northern Mesopotamia--Southeastern Anatolia region from the
+predicted ranges of crop progenitors is interesting in light of the
+`golden triangle' hypothesis \citep{KozlowskiAurenche2005}, which puts
+this region at the centre of the development of agriculture. Multiple
+lines of archaeological evidence have emerged that point away from this
+hypothesis and towards a more geographically diverse origin
+\citep{Asouti2006, cites}, but it remains the area with some of the
+earliest clear evidence of domestication. Comparative genetics also
+points to the Northern Mesopotamia region as the centre of diversity of
+many crops \citep{cite}. But since these studies are based on modern
+genomes, if the wild range of these plants has, as our modelling
+suggests, shifted since the Early Holocene, the apparent centre of
+diversity may have shifted with them. Our reconstructions are consistent
+with the late arrival of intensive plant-based foraging economies in
+this region (cf.~the Natufian of the Levant) and more broadly there need
+not be a link between the core \emph{wild} range of a plant and the core
+area of its domestication. A scenario in which cultivation emerged at
+the edges of the ranges of valuable plant resources---as a means of
+extend their natural niche---is also plausible.
+
+The near-absence of the Zagros in any predicted ranges is also
+surprising, given mounting evidence that domestication took place just
+as early in the eastern \emph{Mashriq} as it did in the west
+\citep{Braidwood, GanjDarehGoats, HillyFlanksStuff}. We consider that
+the most likely explanation for this is that our flora does not include
+the species that were most important to plant subsistence in the east.
+Archaeobotanical data on Neolithic sites in the Zagros is limited
+(compared to the Levant in particular) due to a hiatus in field research
+there from the 1980s to early 2010s \citep{cite}. Recent research
+\citep{Amaia} indicates \ldots{}
+
+Cyprus and W. Anatolia not considered part of the core zone -- maybe
+they should be?
+
+Exceptions to the trend of range reduction include \emph{Cicer
+reticulatum} (wild chickpea), which has a relatively stable range
+centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
+(wild emmer wheat), which is predicted to have two very restricted
+ranges centered around the Black Sea Coast of Anatolia and the Palmyra
+basin. In the latter case, neither of these areas are part of the
+predicted modern distribution of wild emmer (centered around the
+Caucasus and Northern Mesopotamia), but it would be consistent with
+archaeological evidence for early cultivation at Tell Abu Hureyra and
+other sites in the Upper Euphrates. On the other hand, the model for
+\emph{Triticum turgidum dicoccum} has the lowest predictive performance
+of all our fitted models.
+
+\subsection{Biogeography of crop
+progenitors}\label{biogeography-of-crop-progenitors}
+
+Several crop progenitors reduced in range before cultivation? Reduction
+in range from BA to EH very evident in Hordeum, Pistacia atlantica.
+
+The wheat story.
+
+Flax has a very restricted distribution (consistent with low occurrence
+in founders paper?). As does Pistacia atlantica, Bolboschoenus maritimus
+
+\emph{Secale cereale} = an Anatolian boy
+
+\subsection{Hindcast models do not predict archaeobotanical
+composition}\label{hindcast-models-do-not-predict-archaeobotanical-composition}
 
@@
 hindcast palaeoecological niche models, for example using more finely
-resolved palaeoclimate sequences \citep{KargerEtAl2023}, hyperparameter
-tuning to avoid overfitting, and improved archaeological datasets, would
-help disentangle these potential explanations.
+resolved palaeoclimate sequences \citep[e.g.][]{KargerEtAl2023},
+hyperparameter tuning to avoid overfitting, and improved archaeological
+datasets, would help disentangle these potential explanations.
 
 \section{Conclusion}\label{conclusion}
+
+\begin{itemize}
+\tightlist
+\item
+  We present the first continuous, spatially explicit models of the
+  palaeodistributions of 56 plant species found regularly in association
+  with early agricultural archaeological sites in West Asia
+
+  \begin{itemize}
+  \tightlist
+  \item
+    A new line of evidence on archaeoecology
+  \item
+    Complementary to archaeobot/pollen/etc. because it is independent
+    from it
+  \item
+    All models are wrong\ldots{} but it's easier to see how these are
+    wrong than lines on maps
+  \end{itemize}
+\item
+  Modelling at scale using random forest, modern occurrences, and
+  hindcasting represents a significant advance in pENM methodology
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Relies on recent open ecological and climatic datasets
+  \item
+    \ldots open archaeological datasets still lacking!
+  \end{itemize}
+\item
+  First (? - check Yaworsky) attempt to verify hindcast models with
+  archaeo. compositional data
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Results not to promising, but this doesn't mean the models are
+    useless!
+  \item
+    Discrepencies suggests several areas for further research and
+    methodological development
+  \end{itemize}
+\end{itemize}
 

### 2025-01-07 analysis/paper.qmd
Add introduction and expand discussion

@@
 ---
-title: Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7–8.3 ka
+title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
 date: today
@@
       name: Open Quaternary
-      formatting: review
+      formatting: preprint
       cite-style: authoryear
@@
 - [ ] First draft:
-  - [ ] Introduction
+  - [x] Introduction
   - [ ] Background
@@
   - [ ] Discussion
-    - [ ] General trends
+    - [x] General trends
     - [ ] Case studies
@@
   - [ ] tbl-predictors
+  - [ ] tbl-model-metrics
 - [ ] Outstanding TODOs
@@
 w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)
+
+levant <- st_bbox(c(xmin = 33, xmax = 39, ymin = 30, ymax = 38))
 s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
@@
 
-* The first farming societies had an ecological context
-* Subsistence is understood (largely) through archaeobotany and zooarchaeology; ecological context from environmental archaeology, palynology, palaeoclimate records, etc.
-  * But these have a variety of biases (human selection, taphonomy, etc.)
-  * And at the end of the day only represent specific places – interpolation to the entire region is not straightforward
-  * Overlap makes it difficult to see where human choices depart from the environmental background [cf. @MartinEtAl2016]
-  * Or to fully contextualise subsistence strategies and shifts in strategies in response to environmental shifts [e.g. @YaworskyEtAl2023].
-* Here we present an alternative approach using ENM
-  * Whole region, at multiple climate snapshots
-  * Independent of archaeobot. and pal. clim. data, so can verify and compare
+The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
+West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@cite].
+Even before the end of the Ice Age, these species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
+As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.
+
+Decades of research in archaeobotany and zooarchaeology have reconstructed the subsistence economies of Late Epipalaeolithic and Neolithic sites in great detail [@cites].
+Together with studies of charcoal [@cite], pollen [@cite], soil [@cite], and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
+However, each of these sources of evidence is subject to the wide variety of taphonomic and recovery biases that are inherent in any direct record of the past.
+They are also, by definition, records of the (human) environment at particular times and places.
+Interpolating these snapshots to give a holistic picture of the regional ecologies is not straightforward – to date, it has tended to rely on non-explicit, inductive modelling.
+The majority are also filtered through human action, producing a mixed single that makes it difficult to disentangle anthropic effects from the background of environmental change in this period of rapid climatic alteration.
+
+In this paper we present a complementary, deductive approach based on ecological niche modelling.
+Rather than inferring environmental conditions from preserved physical evidence, 
+we predict the ranges of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
+Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
+<!-- TODO which we can use to ... --->
+Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple past climatologies.
 
@@
 * Important to e.g.
-  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2017; @MartinEtAl2025]
+  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2016; @MartinEtAl2025]
   * Reconstruct sequences of domestication [@YeomansEtAl2017]
@@
 It has been used to model the availability of fauna exploited by humans at wider scales [e.g.
-@deAndresHerreroEtAl2018; @YaworksyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik?
+@deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik?
 --> [@CollinsEtAl2018].
@@
 
-# Methods and materials
+# Data and model
 
@@
     n_present = "Neolithic",
+    p_present = "(%)",
     n_occ = "Present"
   ) |>
-  tab_spanner("Occurrences", columns = starts_with("n_")) |>
-  cols_merge_n_pct(n_present, p_present) |>
+  tab_spanner("Occurrences", columns = c(n_present, p_present, n_occ)) |>
+  # cols_merge_n_pct(n_present, p_present) |>
   cols_hide(exclude) |>
@@
   cols_align("left", c(label)) |>
+  cols_align("right", c(n_present, p_present, n_occ)) |>
+  cols_width(label ~ pct(50), n_present ~ 10, p_present ~ 10, n_occ ~ 20) |>
   fmt_percent(p_present, decimals = 0) |>
@@
   sub_missing() |>
-  tab_options(footnotes.marks = "*")
+  tab_options(
+    footnotes.marks = "*",
+    table.font.size = 11,
+    table.width = pct(100),
+    latex.use_longtable = TRUE
+  )
 ```
@@
 
-```{r fig-occ-map}
-#| fig-cap: Georeferenced occurrence records from West Asia for the `r nrow(flora)` modelled taxa (N=`r nrow(unnest(flora, occ))`)
+```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia for the ", nrow(flora), " modelled taxa (N=", nrow(unnest(flora, occ)))}
 flora |>
@@
 
-# Results
+# Model assessment
 
@@
     mutate(
-      prediction = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
-                       obj = environment, .progress = TRUE)
+      pred = map(niche, collect_predictions),
+      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
+                       obj = environment, .progress = TRUE),
+      # Binarize
+      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
+      palaeodist = map(palaeodist, mutate, absent = prediction.FALSE. > 0.5)
     )
@@
 Predicted distributions for individual taxa are presented in the appendix and accompanying material.
-
-## Model assessment
 
@@
 
-```{r fig-model-sensitivity-vs-n_occ}
-#| fig-cap: Number of taxon occurrences and model sensitivity on test dataset
+```{r fig-model-sensitivity-vs-n_occ, warning=FALSE}
+#| fig-cap: Model sensitivity by number of training occurrences
 flora |>
@@
   geom_point() + 
-  geom_text_repel(aes(label = label), fontface = "italic") +
+  geom_text_repel(aes(label = label), fontface = "italic", size = 10/.pt) +
   scale_x_log10(name = "Occurrences") +
@@
 # TODO: what about negatives?
-archaeo_flora_predictions <- archaeo_flora |>
+archaeo_flora_predictions <-
+
+archaeo_flora |>
   filter(taxon %in% flora$taxon) |>
@@
   mutate(
-    prediction = map(taxon, \(x) filter(flora, taxon == x)$prediction[[1]]),
+    prediction = map(taxon, \(x) filter(flora, taxon == x)$palaeodist[[1]]),
     prediction = map2(prediction, geometry, function(pred, geom) {
@@
     }),
+    prediction = map(prediction, as_tibble),
     prediction = map2(prediction, climate_period, function(pred, per) {
-      pred |>
-        as_tibble() |>
-        filter(as.character(period) == per) |>
-        transmute(
-          p_absent = prediction.FALSE.,
-          p_present = prediction.TRUE.,
-          pred_present = p_present > p_absent
-        )
-    })
+      filter(pred, as.character(period) == per)
+    }),
+    prediction = map(prediction, select,
+                     pred_present = present, pred_absent = absent)
   ) |>
@@
 area_present <- function(pred, res = pred_res) {
-  st_apply(pred, "period", sum, na.rm = TRUE)["pred_present"] |>
+  st_apply(pred, "period", sum, na.rm = TRUE)["present"] |>
     data.frame() |>
-    mutate(area_present = pred_present * (pred_res)^2) |>
+    mutate(area_present = present * (pred_res)^2) |>
     select(period, area_present)
@@
 
-flora <- flora |>
-  mutate(
-    # Binarize
-    prediction = map(prediction, mutate, pred_present = prediction.TRUE. > 0.5),
-    area_present = map(prediction, area_present)
-  )
-```
-
-# Discussion
+flora <- mutate(flora, area_present = map(palaeodist, area_present))
+```
+
+# Results
 
@@
 and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
-Central Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
-leaving the Levant and to a lesser extent Western Anatolia Cyprus as refugia.
+most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
+leaving the Levant and to a lesser extent the Aegean and Cyprus as refugia.
 
 Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @] and as a centre of origin of agriculture [@].
-It should also be noted that the selection of flora for this study (@tbl-occ-count) is biased towards Levantine fauna due to the history of archaeological research in the region,
-as is the occurrence dataset used to train the model [@fig-occ-map].
 Within the Levant, retreat from the Badia.
 
-Loss of the Northern Mesopotamia–Southeastern Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@KozlowskiAurenche2005], which puts this region at the centre of the development of agriculture.
-Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication.
-Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [@cite].
+Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture.
+Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyHopfWeiss2012; @KabukcuEtAl2021; @UlasEtAl2024].
+Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [e.g. @HaasEtAl2019].
 But since these studies are based on modern genomes, if the wild range of these plants has, as our modelling suggests, shifted since the Early Holocene, the apparent centre of diversity may have shifted with them.
-Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant)
+Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant),
 and more broadly there need not be a link between the core *wild* range of a plant and the core area of its domestication. 
@@
 
-Cyprus and W. Anatolia not considered part of the core zone – maybe they should be?
-
-Exceptions to the trend of range reduction include *Cicer reticulatum* (wild chickpea), which has a relatively stable range centered on Northern Mesopotamia;
-and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to have two very restricted ranges centered around the Black Sea Coast of Anatolia and the Palmyra basin.
-In the latter case, neither of these areas are part of the predicted modern distribution of wild emmer (centered around the Caucasus and Northern Mesopotamia), but it would be consistent with archaeological evidence for early cultivation at Tell Abu Hureyra and other sites in the Upper Euphrates.
-On the other hand, the model for *Triticum turgidum dicoccum* has the lowest predictive performance of all our fitted models.
+Cyprus and the Aegean are not conventionally considered part of the primary zone of domestication but rather amongst the first regions that acquired agriculture from West Asia.
+Our analysis complicates this picture, as it indicates that the wild ranges of many crop progenitors included these regions.
+Early examples of several domesticates are recorded at sites on Cyprus, Western Anatolia and Greece [@ArranzOtaeguiRoe2023],
+and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
+Were these area part of the same broader 'interaction sphere' [@cite] that produce Neolithic agriculture in West Asia?
+
+Exceptions to the dominant trend of range reduction include *Cicer reticulatum* (wild chickpea), which has a relatively stable range centered on Northern Mesopotamia;
+and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to have two limited ranges centered around the Black Sea Coast of Anatolia and the Palmyra basin.
+In the latter case, neither of these areas are part of the predicted modern distribution of wild emmer (centered around the Caucasus and Northern Mesopotamia), but it would be consistent with archaeological evidence for early cultivation at sites in the Upper Euphrates [@Willcox2024].
 
@@
 
-Several crop progenitors reduced in range before cultivation?
-Reduction in range from BA to EH very evident in Hordeum, Pistacia atlantica.
+* Most cereal and legume crops predicted to be Levantine
+* Cereals show diverging southern/northern ranges (legumes not so much)
+* Not rye (Anatolia), chickpea (N. Mesopotamia)
+
+```{r fig-hindcast-barely-levant, eval=FALSE, cache=TRUE}
+#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
+palaeodist_hordeum <- do.call(c, c(
+  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
+  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
+))
+
+# TODO: expand period facet labels
+ggplot() +
+  facet_grid(taxon ~ period) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.), 
+             na.action = na.omit) +
+  annotation_spatial(ne_land, fill = NA) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_x_continuous(breaks = c(34, 36, 38)) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(
+    xlim = c(33, 39), ylim = c(30, 38),
+    crs = w_asia_albers, default_crs = latlong
+  ) +
+  labs(fill = "P(present)", x = NULL, y = NULL) +
+  theme(
+    strip.background = element_blank(),
+    strip.text.y = element_text(face = "italic")
+  )
+```
+
+```{r fig-hindcast-wheat-levant, eval=FALSE, cache=TRUE}
+#| fig-cap: Predicted palaeodistribution of wild wheat in the Levant
+palaeodist_triticum <- do.call(c, c(
+  filter(flora, str_starts(taxon, "Triticum"))$palaeodist,
+  list(along = list(taxon = filter(flora, str_starts(taxon, "Triticum"))$taxon))
+))
+
+# TODO: expand period facet labels
+ggplot() +
+  facet_grid(taxon ~ period) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(palaeodist_triticum, prediction.TRUE.)) + 
+  annotation_spatial(ne_land, fill = NA) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_x_continuous(breaks = c(34, 36, 38)) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(
+    xlim = c(33, 39), ylim = c(30, 38),
+    crs = w_asia_albers, default_crs = latlong
+  ) +
+  labs(fill = "P(present)", x = NULL, y = NULL) +
+  theme(
+    strip.background = element_blank(),
+    strip.text.y = element_text(face = "italic")
+  )
+```
+
+<!-- TODO binarized range figure other founder crops? -->
+
+Wild barley [*Hordeum spontaneum*, @fig-hindcast-barely-levant] and pistachio [*Pistacia atlantica*, @fig] show a contraction of their predicted ranges from the Pleistocene to the Holocene,
+concurrent with them being brought into cultivation.
+Plausible that this prompted cultivation?
+These two species [@ArranzOtaeguiRoe2023] also see marked declines in the archaeobotanical record from the Early PPNA/Early PPNB (where they were amongst the most common taxa) to the Late PPNB and Late Neolithic.
+However, this trend is not seen in most other crop progenitors.
+
+Cereals / other progenitors (not all) reduced in range before cultivation?
+Reduction in range from BA to EH very evident in Hordeum, less so in ...
+
+<!-- TODO: composite wheat progentor figure -->
 
 The wheat story.
+
+<!-- TODO: figure with flax, Pistacia, Bolboschoenus -->
 
@@
 As does Pistacia atlantica, Bolboschoenus maritimus
+
+<!-- TODO: figure with rye -->
 
@@
 <!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
-
-# Conclusion
+Is inconsistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
+
+# Discussion
 
@@
 
-niche_taxon <- "Aegilops tauschii"
-niche_prediction <- filter(flora, taxon == niche_taxon)$prediction[[1]]
+niche_taxon <- "Lathyrus oleraceus"
+niche_prediction <- filter(flora, taxon == niche_taxon)$palaeodist[[1]]
 

### 2025-01-07 analysis/paper.tex
Add introduction and expand discussion

@@
   authoryear,
-  review]{elsarticle}
+  preprint]{elsarticle}
 
@@
 \hypersetup{
-  pdftitle={Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3 ka},
+  pdftitle={Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka},
   pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
@@
 \begin{frontmatter}
-\title{Biogeography of crop progenitors and wild plant resources over in
-terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3
-ka}
+\title{Biogeography of crop progenitors and wild plant resources in the
+terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka}
 \author[1,2]{Joe Roe%
@@
     \tightlist
-    \item[$\square$]
+    \item[$\boxtimes$]
       General trends
@@
     tbl-predictors
+  \item[$\square$]
+    tbl-model-metrics
   \end{itemize}
@@
 
-\begin{itemize}
-\tightlist
-\item
-  The first farming societies had an ecological context
-\item
-  Subsistence is understood (largely) through archaeobotany and
-  zooarchaeology; ecological context from environmental archaeology,
-  palynology, palaeoclimate records, etc.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    But these have a variety of biases (human selection, taphonomy,
-    etc.)
-  \item
-    And at the end of the day only represent specific places --
-    interpolation to the entire region is not straightforward
-  \item
-    Overlap makes it difficult to see where human choices depart from
-    the environmental background \citep[cf.][]{MartinEtAl2016}
-  \item
-    Or to fully contextualise subsistence strategies and shifts in
-    strategies in response to environmental shifts
-    \citep[e.g.][]{YaworskyEtAl2023}.
-  \end{itemize}
-\item
-  Here we present an alternative approach using ENM
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Whole region, at multiple climate snapshots
-  \item
-    Independent of archaeobot. and pal. clim. data, so can verify and
-    compare
-  \end{itemize}
-\end{itemize}
+The Pleistocene--Holocene transition in West Asia marked a turning point
+in global environmental history, as humans brought the first plants
+under cultivation and began modifying surrounding ecosystems to support
+their own subsistence. West Asia is part of the native range of a
+remarkable number of domesticable plant species, including wild
+relatives of wheat, barley, peas, lentils, and other crops of global
+importance \citep{cite}. Even before the end of the Ice Age, these
+species supported uniquely dense and complex Late Epipalaeolithic
+(15--11.7 ka) societies based on intensive foraging \citep{cite} and
+eventually pre-domestication cultivation \citep{cite}. As they were
+brought under domestication in the Pre-Pottery Neolithic period
+(11.7--8.5 ka), the world's first agriculture was shaped by the
+ecosystems from which it emerged and was embedded in.
+
+Decades of research in archaeobotany and zooarchaeology have
+reconstructed the subsistence economies of Late Epipalaeolithic and
+Neolithic sites in great detail \citep{cites}. Together with studies of
+charcoal \citep{cite}, pollen \citep{cite}, soil \citep{cite}, and a
+variety of palaeoclimate archives \citep{JonesEtAl2019}, they also tell
+us much about the environments surrounding these settlements. However,
+each of these sources of evidence is subject to the wide variety of
+taphonomic and recovery biases that are inherent in any direct record of
+the past. They are also, by definition, records of the (human)
+environment at particular times and places. Interpolating these
+snapshots to give a holistic picture of the regional ecologies is not
+straightforward -- to date, it has tended to rely on non-explicit,
+inductive modelling. The majority are also filtered through human
+action, producing a mixed single that makes it difficult to disentangle
+anthropic effects from the background of environmental change in this
+period of rapid climatic alteration.
+
+In this paper we present a complementary, deductive approach based on
+ecological niche modelling. Rather than inferring environmental
+conditions from preserved physical evidence, we predict the ranges of
+individual species relevant to human subsistence based on a model of
+their current environmental niche and simulations of past palaeoclimate.
+Though hypothetical, this gives us an independent line of evidence on
+past ecologies that is independent of the environmental archaeological
+and palaeoclimatic records. Our computational approach is also readily
+scaled up, allowing us to model spatially-explicit palaeodistributions
+for a large number of species, for the whole region, under multiple past
+climatologies.
 
@@
     Distinguish environmental from potentially anthropogenic change
-    \citep{MartinEtAl2017, MartinEtAl2025}
+    \citep{MartinEtAl2016, MartinEtAl2025}
   \item
@@
 availability of fauna exploited by humans at wider scales
-\citep[e.g.][]{deAndresHerreroEtAl2018, YaworksyEtAl2023} and, in a West
+\citep[e.g.][]{deAndresHerreroEtAl2018, YaworskyEtAl2023} and, in a West
 Asian context, of foraged plant resources in the landscape around the
@@
 
-\section{Methods and materials}\label{methods-and-materials}
+\section{Data and model}\label{data-and-model}
 
@@
 
-\begin{table}
+\begingroup
+\setlength\LTleft{0\linewidth}
+\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont
+\setlength{\LTpost}{0mm}
+
+\begin{longtable}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.50\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 15.00pt -2\tabcolsep-1.5\arrayrulewidth}}
 
@@
 
-\centering{
-
-\fontsize{12.0pt}{14.4pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lrr}
+\tabularnewline
+
 \toprule
- & \multicolumn{2}{c}{Occurrences} \\ 
-\cmidrule(lr){2-3}
-Taxon & Neolithic & Present \\ 
+ & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 30.00pt -2\tabcolsep-1.5\arrayrulewidth}}{Occurrences} \\ 
+\cmidrule(lr){2-4}
+Taxon & Neolithic & (\%) & Present \\ 
 \midrule\addlinespace[2.5pt]
@@
 
-incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 (44\%) & 200 \\ 
+incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 & 44\% & 200 \\ 
 \emph{Hordeum spontaneum}
 
-incl. \emph{Hordeum vulgare} & 62 (43\%) & 6912 \\ 
+incl. \emph{Hordeum vulgare} & 62 & 43\% & 6912 \\ 
 \emph{Triticum monococcum aegilopoides}
 
-incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 (30\%) & 3213 \\ 
+incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 & 30\% & 3213 \\ 
 \emph{Bolboschoenus maritimus}
 
-incl. \emph{Scirpus maritimus} & 33 (23\%) & 364 \\ 
-\emph{Vicia ervilia} & 33 (23\%) & 858 \\ 
-\emph{Buglossoides tenuiflora} & 31 (21\%) & 159 \\ 
-\emph{Arnebia decumbens} & 25 (17\%) & 240 \\ 
-\emph{Buglossoides arvensis} & 25 (17\%) & 251 \\ 
-\emph{Medicago radiata} & 21 (14\%) & 883 \\ 
-\emph{Androsace maxima} & 20 (14\%) & 130 \\ 
+incl. \emph{Scirpus maritimus} & 33 & 23\% & 364 \\ 
+\emph{Vicia ervilia} & 33 & 23\% & 858 \\ 
+\emph{Buglossoides tenuiflora} & 31 & 21\% & 159 \\ 
+\emph{Arnebia decumbens} & 25 & 17\% & 240 \\ 
+\emph{Buglossoides arvensis} & 25 & 17\% & 251 \\ 
+\emph{Medicago radiata} & 21 & 14\% & 883 \\ 
+\emph{Androsace maxima} & 20 & 14\% & 130 \\ 
 \emph{Vicia orientalis}
 
-incl. \emph{Lens culinaris} & 17 (12\%) & 441 \\ 
+incl. \emph{Lens culinaris} & 17 & 12\% & 441 \\ 
 \emph{Medicago astroites}
 
-incl. \emph{Trigonella astroites} & 16 (11\%) & 130 \\ 
-\emph{Arnebia linearifolia} & 14 (10\%) & 158 \\ 
+incl. \emph{Trigonella astroites} & 16 & 11\% & 130 \\ 
+\emph{Arnebia linearifolia} & 14 & 10\% & 158 \\ 
 \emph{Linum bienne}
 
-incl. \emph{Linum usitatissimum} & 11 (8\%) & 274 \\ 
+incl. \emph{Linum usitatissimum} & 11 & 8\% & 274 \\ 
 \emph{Gypsophila vaccaria}
 
-incl. \emph{Vaccaria pyramidata} & 11 (8\%) & 326 \\ 
-\emph{Carex divisa} & 10 (7\%) & 330 \\ 
-\emph{Ficus carica} & 10 (7\%) & 4265 \\ 
+incl. \emph{Vaccaria pyramidata} & 11 & 8\% & 326 \\ 
+\emph{Carex divisa} & 10 & 7\% & 330 \\ 
+\emph{Ficus carica} & 10 & 7\% & 4265 \\ 
 \emph{Lathyrus oleraceus}
 
-incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 (6\%) & 2269 \\ 
-\emph{Vicia faba} & 8 (6\%) & 2578 \\ 
+incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 & 6\% & 2269 \\ 
+\emph{Vicia faba} & 8 & 6\% & 2578 \\ 
 \emph{Aizoanthemopsis hispanica}
 
-incl. \emph{Aizoon hispanicum} & 7 (5\%) & 303 \\ 
-\emph{Bolboschoenus glaucus} & 7 (5\%) & 27\textsuperscript{\textit{*}} \\ 
-\emph{Pistacia atlantica} & 7 (5\%) & 928 \\ 
+incl. \emph{Aizoon hispanicum} & 7 & 5\% & 303 \\ 
+\emph{Bolboschoenus glaucus} & 7 & 5\% & 27\textsuperscript{\textit{*}} \\ 
+\emph{Pistacia atlantica} & 7 & 5\% & 928 \\ 
 \emph{Polygonum arenarium arenarium}
 
-incl. \emph{Polygonum venantianum} & 7 (5\%) & 5\textsuperscript{\textit{*}} \\ 
-\emph{Prosopis farcta} & 7 (5\%) & 1205 \\ 
-\emph{Rumex pulcher} & 7 (5\%) & 794 \\ 
-\emph{Ammi majus} & 6 (4\%) & 461 \\ 
+incl. \emph{Polygonum venantianum} & 7 & 5\% & 5\textsuperscript{\textit{*}} \\ 
+\emph{Prosopis farcta} & 7 & 5\% & 1205 \\ 
+\emph{Rumex pulcher} & 7 & 5\% & 794 \\ 
+\emph{Ammi majus} & 6 & 4\% & 461 \\ 
 \emph{Cicer reticulatum}
 
-incl. \emph{Cicer arietinum} & 6 (4\%) & 193 \\ 
-\emph{Hordeum bulbosum} & 6 (4\%) & 2036 \\ 
-\emph{Polygonum corrigioloides} & 6 (4\%) & 4\textsuperscript{\textit{*}} \\ 
-\emph{Salsola kali} & 6 (4\%) & 369 \\ 
-\emph{Taeniatherum caput-medusae} & 6 (4\%) & 235 \\ 
-\emph{Capparis spinosa} & 5 (3\%) & 4031 \\ 
-\emph{Chenopodium album} & 5 (3\%) & 357 \\ 
-\emph{Lolium temulentum} & 5 (3\%) & 148 \\ 
-\emph{Poa bulbosa} & 5 (3\%) & 1401 \\ 
-\emph{Aegilops crassa} & 4 (3\%) & 170 \\ 
-\emph{Atriplex prostrata} & 4 (3\%) & 112 \\ 
-\emph{Brachypodium distachyon} & 4 (3\%) & 1581 \\ 
-\emph{Fumaria densiflora} & 4 (3\%) & 194 \\ 
-\emph{Gypsophila pilosa} & 4 (3\%) & 79 \\ 
-\emph{Hordeum murinum} & 4 (3\%) & 1342 \\ 
-\emph{Lathyrus aphaca} & 4 (3\%) & 2085 \\ 
-\emph{Lathyrus sativus} & 4 (3\%) & 297 \\ 
-\emph{Secale cereale} & 4 (3\%) & 1153 \\ 
-\emph{Triticum durum} & 4 (3\%) & 219 \\ 
-\emph{Vitis sylvestris} & 4 (3\%) & 3\textsuperscript{\textit{*}} \\ 
-\emph{Adonis flammea} & 3 (2\%) & 79 \\ 
-\emph{Avena sterilis} & 3 (2\%) & 50001\textsuperscript{\textit{*}} \\ 
-\emph{Bassia arabica} & 3 (2\%) & 230 \\ 
-\emph{Bromus sterilis} & 3 (2\%) & 385 \\ 
-\emph{Cephalaria syriaca} & 3 (2\%) & 144 \\ 
-\emph{Glaucium aleppicum} & 3 (2\%) & 22\textsuperscript{\textit{*}} \\ 
-\emph{Helianthemum ledifolium} & 3 (2\%) & 668 \\ 
-\emph{Lolium rigidum} & 3 (2\%) & 1464 \\ 
-\emph{Phalaris paradoxa} & 3 (2\%) & 370 \\ 
-\emph{Quercus ithaburensis} & 3 (2\%) & 1431 \\ 
-\emph{Triticum aestivum compactum} & 3 (2\%) & 209 \\ 
-\emph{Verbena officinalis} & 3 (2\%) & 1046 \\ 
-\emph{Vicia narbonensis} & 3 (2\%) & 1529 \\ 
-\emph{Triticum urartu} & — & 2033 \\ 
-\emph{Aegilops speltoides} & — & 1429 \\ 
-\emph{Aegilops tauschii} & — & 1395 \\ 
+incl. \emph{Cicer arietinum} & 6 & 4\% & 193 \\ 
+\emph{Hordeum bulbosum} & 6 & 4\% & 2036 \\ 
+\emph{Polygonum corrigioloides} & 6 & 4\% & 4\textsuperscript{\textit{*}} \\ 
+\emph{Salsola kali} & 6 & 4\% & 369 \\ 
+\emph{Taeniatherum caput-medusae} & 6 & 4\% & 235 \\ 
+\emph{Capparis spinosa} & 5 & 3\% & 4031 \\ 
+\emph{Chenopodium album} & 5 & 3\% & 357 \\ 
+\emph{Lolium temulentum} & 5 & 3\% & 148 \\ 
+\emph{Poa bulbosa} & 5 & 3\% & 1401 \\ 
+\emph{Aegilops crassa} & 4 & 3\% & 170 \\ 
+\emph{Atriplex prostrata} & 4 & 3\% & 112 \\ 
+\emph{Brachypodium distachyon} & 4 & 3\% & 1581 \\ 
+\emph{Fumaria densiflora} & 4 & 3\% & 194 \\ 
+\emph{Gypsophila pilosa} & 4 & 3\% & 79 \\ 
+\emph{Hordeum murinum} & 4 & 3\% & 1342 \\ 
+\emph{Lathyrus aphaca} & 4 & 3\% & 2085 \\ 
+\emph{Lathyrus sativus} & 4 & 3\% & 297 \\ 
+\emph{Secale cereale} & 4 & 3\% & 1153 \\ 
+\emph{Triticum durum} & 4 & 3\% & 219 \\ 
+\emph{Vitis sylvestris} & 4 & 3\% & 3\textsuperscript{\textit{*}} \\ 
+\emph{Adonis flammea} & 3 & 2\% & 79 \\ 
+\emph{Avena sterilis} & 3 & 2\% & 50001\textsuperscript{\textit{*}} \\ 
+\emph{Bassia arabica} & 3 & 2\% & 230 \\ 
+\emph{Bromus sterilis} & 3 & 2\% & 385 \\ 
+\emph{Cephalaria syriaca} & 3 & 2\% & 144 \\ 
+\emph{Glaucium aleppicum} & 3 & 2\% & 22\textsuperscript{\textit{*}} \\ 
+\emph{Helianthemum ledifolium} & 3 & 2\% & 668 \\ 
+\emph{Lolium rigidum} & 3 & 2\% & 1464 \\ 
+\emph{Phalaris paradoxa} & 3 & 2\% & 370 \\ 
+\emph{Quercus ithaburensis} & 3 & 2\% & 1431 \\ 
+\emph{Triticum aestivum compactum} & 3 & 2\% & 209 \\ 
+\emph{Verbena officinalis} & 3 & 2\% & 1046 \\ 
+\emph{Vicia narbonensis} & 3 & 2\% & 1529 \\ 
+\emph{Triticum urartu} & — & — & 2033 \\ 
+\emph{Aegilops speltoides} & — & — & 1429 \\ 
+\emph{Aegilops tauschii} & — & — & 1395 \\ 
 \bottomrule
-\end{tabular*}
+
+\end{longtable}
+
 \begin{minipage}{\linewidth}
@@
 \end{minipage}
-
-}
-
-\end{table}%
+\endgroup
 
@@
 \caption{\label{fig-occ-map}Georeferenced occurrence records from West
-Asia for the \texttt{r\ nrow(flora)} modelled taxa
-(N=\texttt{r\ nrow(unnest(flora,\ occ))})}
+Asia for the 56 modelled taxa (N=35208}
 
@@
 
-\section{Results}\label{results}
+\section{Model assessment}\label{model-assessment}
 
@@
 
-\subsection{Model assessment}\label{model-assessment}
-
-\begin{verbatim}
-Warning: Removed 51 rows containing missing values or values outside the scale range
-(`geom_text_repel()`).
-\end{verbatim}
-
 \begin{figure}
@@
 
-\caption{\label{fig-model-sensitivity-vs-n_occ}Number of taxon
-occurrences and model sensitivity on test dataset}
+\caption{\label{fig-model-sensitivity-vs-n_occ}Model sensitivity by
+number of training occurrences}
 
@@
 
-\section{Discussion}\label{discussion}
+\section{Results}\label{results}
 
@@
 \emph{Vicia} spp. (vetches), including \emph{Vicia faba} (broad beans).
-Central Anatolia, Northern Mesopotamia, and the Zagros Mountains in
+most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in
 particular disappear from the predicted ranges of these species, leaving
-the Levant and to a lesser extent Western Anatolia Cyprus as refugia.
+the Levant and to a lesser extent the Aegean and Cyprus as refugia.
 
@@
 Natufian culture, @{]} and as a centre of origin of agriculture {[}@{]}.
-It should also be noted that the selection of flora for this study
-(Table~\ref{tbl-occ-count}) is biased towards Levantine fauna due to the
-history of archaeological research in the region, as is the occurrence
-dataset used to train the model Figure~\ref{fig-occ-map}. Within the
-Levant, retreat from the Badia.
-
-Loss of the Northern Mesopotamia--Southeastern Anatolia region from the
-predicted ranges of crop progenitors is interesting in light of the
-`golden triangle' hypothesis \citep{KozlowskiAurenche2005}, which puts
-this region at the centre of the development of agriculture. Multiple
-lines of archaeological evidence have emerged that point away from this
-hypothesis and towards a more geographically diverse origin
+Within the Levant, retreat from the Badia.
+
+Loss of the Northern Mesopotamia--Anatolia region from the predicted
+ranges of crop progenitors is interesting in light of the `golden
+triangle' hypothesis
+\citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
+puts this region at the centre of the development of agriculture.
+Multiple lines of archaeological evidence have emerged that point away
+from this hypothesis and towards a more geographically diverse origin
 \citep{Asouti2006, cites}, but it remains the area with some of the
-earliest clear evidence of domestication. Comparative genetics also
-points to the Northern Mesopotamia region as the centre of diversity of
-many crops \citep{cite}. But since these studies are based on modern
-genomes, if the wild range of these plants has, as our modelling
-suggests, shifted since the Early Holocene, the apparent centre of
-diversity may have shifted with them. Our reconstructions are consistent
-with the late arrival of intensive plant-based foraging economies in
-this region (cf.~the Natufian of the Levant) and more broadly there need
-not be a link between the core \emph{wild} range of a plant and the core
-area of its domestication. A scenario in which cultivation emerged at
-the edges of the ranges of valuable plant resources---as a means of
-extend their natural niche---is also plausible.
+earliest clear evidence of domestication
+\citep{ZoharyHopfWeiss2012, KabukcuEtAl2021, UlasEtAl2024}. Comparative
+genetics also points to the Northern Mesopotamia region as the centre of
+diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But since these
+studies are based on modern genomes, if the wild range of these plants
+has, as our modelling suggests, shifted since the Early Holocene, the
+apparent centre of diversity may have shifted with them. Our
+reconstructions are consistent with the late arrival of intensive
+plant-based foraging economies in this region (cf.~the Natufian of the
+Levant), and more broadly there need not be a link between the core
+\emph{wild} range of a plant and the core area of its domestication. A
+scenario in which cultivation emerged at the edges of the ranges of
+valuable plant resources---as a means of extend their natural niche---is
+also plausible.
 
@@
 
-Cyprus and W. Anatolia not considered part of the core zone -- maybe
-they should be?
-
-Exceptions to the trend of range reduction include \emph{Cicer
+Cyprus and the Aegean are not conventionally considered part of the
+primary zone of domestication but rather amongst the first regions that
+acquired agriculture from West Asia. Our analysis complicates this
+picture, as it indicates that the wild ranges of many crop progenitors
+included these regions. Early examples of several domesticates are
+recorded at sites on Cyprus, Western Anatolia and Greece
+\citep{ArranzOtaeguiRoe2023}, and the Aegean region was probably
+connected to West Asia by a land bridge via Anatolia until the Early
+Holocene \citep{AksuHiscott2022}. Were these area part of the same
+broader `interaction sphere' \citep{cite} that produce Neolithic
+agriculture in West Asia?
+
+Exceptions to the dominant trend of range reduction include \emph{Cicer
 reticulatum} (wild chickpea), which has a relatively stable range
 centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
-(wild emmer wheat), which is predicted to have two very restricted
-ranges centered around the Black Sea Coast of Anatolia and the Palmyra
-basin. In the latter case, neither of these areas are part of the
-predicted modern distribution of wild emmer (centered around the
-Caucasus and Northern Mesopotamia), but it would be consistent with
-archaeological evidence for early cultivation at Tell Abu Hureyra and
-other sites in the Upper Euphrates. On the other hand, the model for
-\emph{Triticum turgidum dicoccum} has the lowest predictive performance
-of all our fitted models.
+(wild emmer wheat), which is predicted to have two limited ranges
+centered around the Black Sea Coast of Anatolia and the Palmyra basin.
+In the latter case, neither of these areas are part of the predicted
+modern distribution of wild emmer (centered around the Caucasus and
+Northern Mesopotamia), but it would be consistent with archaeological
+evidence for early cultivation at sites in the Upper Euphrates
+\citep{Willcox2024}.
 
@@
 
-Several crop progenitors reduced in range before cultivation? Reduction
-in range from BA to EH very evident in Hordeum, Pistacia atlantica.
+\begin{itemize}
+\tightlist
+\item
+  Most cereal and legume crops predicted to be Levantine
+\item
+  Cereals show diverging southern/northern ranges (legumes not so much)
+\item
+  Not rye (Anatolia), chickpea (N. Mesopotamia)
+\end{itemize}
+
+Wild barley \textbf{?@fig-hindcast-barely-levant} and pistachio
+\citep[\emph{Pistacia atlantica},][]{fig} show a contraction of their
+predicted ranges from the Pleistocene to the Holocene, concurrent with
+them being brought into cultivation. Plausible that this prompted
+cultivation? These two species \citep{ArranzOtaeguiRoe2023} also see
+marked declines in the archaeobotanical record from the Early PPNA/Early
+PPNB (where they were amongst the most common taxa) to the Late PPNB and
+Late Neolithic. However, this trend is not seen in most other crop
+progenitors.
+
+Cereals / other progenitors (not all) reduced in range before
+cultivation? Reduction in range from BA to EH very evident in Hordeum,
+less so in \ldots{}
 
@@
 
-\section{Conclusion}\label{conclusion}
+Is inconsistent with more ``macro'' trends such as the reduced range of
+Hordeum and Pistacia correlating with its reduced abundance in the
+archaeobotanical record.
+
+\section{Discussion}\label{discussion}
 

### 2025-01-08 analysis/appendix.qmd
Add appendix

---
title: "Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka"
subtitle: "Fitted model summaries and predicted palaeodistributions"
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This appendix describes ecological niche models of 56<!-- TODO keep updated -->
  plant species relevant to the subsistence of Late Epipalaeolithic (15–11.7 ka) 
  and Neolithic (11.7–8.2 ka) societies in West Asia. It includes summaries of 
  the training data and performance metrics of the fitted models together with 
  predicted palaeodistributions of each species for three simulated 
  climatologies over the Pleistocene/Holocene boundary (the Bølling–Allerød, 
  14.7–12.9 ka; Younger Dryas, 12.9–11.7 ka; and Early Holocene, 11.7–8.3 ka) as
  well as current conditions.
  For a full description of the methodology, references, and discussion of the 
  results, please see the main text. The data and R code used to produce the
  models, as well as the full predicted palaeodistributions in raster format,
  are achived with Zenodo at <https://doi.org/> <!-- TODO insert Zenodo DOI -->
format:
  html:
    toc: true
execute:
  echo: false
---

```{r dependencies, message=FALSE}
library("cowplot")
library("dplyr")
library("here")
library("forcats")
library("ggplot2")
library("ggspatial")
library("gt")
library("khroma")
library("patchwork")
library("purrr")
library("readr")
library("sf")
library("stars")
library("tidyr")
library("tune")
library("yardstick")
library("vip")

library("BadiaPaleoFloraENM") # this package, install with devtools::install()

options(knitr.duplicate.label = "allow")
```

```{r data, message=FALSE}
# Model
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
archaeo_flora <- read_tsv(
  derived_data("archaeo_flora_predictions.tsv"),
  col_types = cols(
    source = col_character(),
    source_id = col_character(),
    source_site_name = col_character(),
    site_name = col_character(),
    latitude = col_double(),
    longitude = col_double(),
    phase = col_character(),
    phase_description = col_character(),
    phase_code = col_character(),
    age_start = col_double(),
    age_end = col_double(),
    taxon_source = col_character(),
    n = col_double(),
    prop = col_double(),
    reference = col_character(),
    taxon = col_character(),
    taxon_group = col_character(),
    genus = col_character(),
    family = col_character(),
    category = col_character(),
    founder_crop = col_character(),
    edibility = col_character(),
    grass_type = col_character(),
    legume_type = col_character(),
    age_mid = col_double(),
    period = col_character(),
    climate_period = col_character(),
    pred_present = col_logical(),
    pred_absent = col_logical()
  )
)

# Chronology
climate_periods <- tibble(
  period = c("Bølling-Allerød", "Younger Dryas", "Early Holocene", "Current"),
  code = c("ba", "yds", "eh", "cur"),
  date_range = c("(14.7-12.9 ka)", "(12.9-11.7 ka)", "(11.7-8.3 ka)", ""),
  label = fct_inorder(paste(toupper(code), date_range)),
  long_label = fct_inorder(paste(period, date_range)),
  start_bp = c(14700, 12900, 11700, NA),
  end_bp = c(12900, 11700, 8300, NA)
)

# Map
latlong <- 4326
w_asia_albers <- "+proj=aea +lat_1=22.5 +lat_2=42.5 +lon_0=40"

w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)

# Turn off s2 while we deal with Natural Earth's funky geometries
use_s2 <- sf_use_s2(FALSE)

ne_countries <- read_sf(raw_data("ne"), "ne_10m_admin_0_countries")

ne_rivers <- read_sf(raw_data("ne"), "ne_10m_rivers_lake_centerlines_scale_rank")

ne_ocean <- read_sf(raw_data("ne"), "ne_10m_ocean") |>
  st_geometry() |>
  st_crop(buffer_bbox(w_asia, 10)) |>
  st_union()

ne_land <- read_sf(raw_data("ne"), "ne_10m_land") |> 
  st_geometry() |>
  st_crop(buffer_bbox(w_asia, 10)) |>
  st_difference(ne_ocean) |>
  st_union()

# Restore previous setting of s2
sf_use_s2(use_s2)
```

```{r content, message=FALSE, results="asis"}
flora <- arrange(flora, taxon)
content <- pmap(
  flora, 
  function(taxon, taxon_detail, occ, niche, palaeodist, ...) {
    archaeo <- archaeo_flora[archaeo_flora$taxon == taxon,]
    knitr::knit_child(
      here("analysis", "appendix_section.qmd"),
      envir = environment(),
      quiet = TRUE
    )
  },
  archaeo_flora = archaeo_flora, 
)

cat(unlist(content), sep = "\n")
```

### 2025-01-08 analysis/appendix_section.qmd
Add appendix

# *`r taxon`*

```{r taxon-detail, eval=!is.na(taxon_detail), results = "asis"}
cat("Including ", knitr::combine_words(taxon_detail), ".", sep = "")
```

```{r fig-model, fig.cap=paste0("Fitted model summary for *", taxon, "*")}
fig_occ <- occ |>
  filter(as.logical(present)) |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) |>
  ggplot() +
  annotation_spatial(ne_land, fill = "#ffffff") +
  annotation_spatial(ne_rivers, aes(linewidth = strokeweig),
                     colour = "lightblue") +
  geom_sf(size = 0.5) +
  layer_spatial(w_asia, fill = NA) +
  scale_linewidth_identity() +
  coord_sf(crs = w_asia_albers) +
  labs(
    title = "GBIF occurrences",
    subtitle = paste0("N=", nrow(occ))
  )

fig_roc_curve <- niche |>
  collect_predictions() |>
  roc_curve(.pred_TRUE, truth = present, event_level = "second") |>
  ggplot() +
  geom_abline(intercept = c(0, 0), slope = 1, linetype = "dashed") +
  geom_line(aes(1 - specificity, sensitivity)) +
  labs(title = "ROC curve") +
  theme_minimal_grid() +
  theme(
    plot.title = element_text(size = 12, face = "plain"),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 10)
  )

fig_vip <- niche |>
  extract_fit_engine() |>
  vip() +
  labs(
    title = "Variable importance",
    subtitle = "Top 10 (impurity)"
  ) +
  theme_minimal_grid() +
  theme(
    plot.title = element_text(size = 12, face = "plain"),
    plot.subtitle = element_text(size = 10),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 10)
  )

fig_occ + (fig_roc_curve / fig_vip) +
  plot_layout(widths = c(3, 1))
```

```{r fig-palaeodist, fig.height=16, fig.cap=paste0("Predicted palaeodistributions of *", taxon, "*")}
ggplot() +
  facet_wrap(
    vars(
      factor(period, levels = climate_periods$code, 
             labels = climate_periods$long_label)
    ),
    ncol = 1
  ) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(
    data = select(palaeodist, prediction.TRUE.), 
    na.action = na.omit
  ) +
  geom_sf(
    aes(shape = pred_present),
    data = archaeo |>
      mutate(
        period = climate_period, 
        pred_present = replace_na(pred_present, FALSE)
      ) |>
      distinct(longitude, latitude, .keep_all = TRUE) |>
      st_as_sf(coords = c("longitude", "latitude"), crs = latlong)
  ) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE) +
  scale_shape_manual(
    name = "Archaeological\noccurrences", 
    values = c(1, 16),
    labels = c("Not predicted", "Predicted")
  ) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "P(present)", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

```{r tbl-archaeo}
# TODO shouldn't be necessary
if (nrow(archaeo) > 0) {
  archaeo |>
    drop_na(climate_period) |>
    summarise(
      n_assemb = n(),
      mean_p = mean(prop),
      age_start = max(age_start),
      age_end = min(age_end),
      pred_present = all(pred_present),
      .by = c(source, climate_period, site_name)
    ) |>
    mutate(
      climate_period = map(climate_period, function(x) {
        climate_periods[climate_periods$code == x,]$long_label[[1]]
      }),
      climate_period = unlist(climate_period),
      age_start = age_start / 1000,
      age_end = age_end / 1000
    )|>
    group_by(climate_period) |>
    arrange(desc(age_start)) |>
    select(climate_period, site_name, age_start, age_end, n_assemb, mean_p, source,
           pred_present) |>
    gt(caption = paste0("Archaeological occurrences of *", taxon, "*")) |>
    cols_label(site_name = "Site", n_assemb = "N assemblages", 
               mean_p = "Average prop.", source = "Source", 
               age_start = "Age range", pred_present = "Predicted?") |>
    cols_merge(c(age_start, age_end), pattern = "{1}–{2} ka") |>
    fmt_number(c(age_start, age_end), decimals = 1) |>
    fmt_percent(c(mean_p)) |>
    fmt_tf(c(pred_present), tf_style = "yes-no") |>
    sub_missing()
}
```

### 2025-01-08 analysis/appendix.qmd
Revise species list

@@
 abstract: |
-  This appendix describes ecological niche models of 56<!-- TODO keep updated -->
+  This appendix describes ecological niche models of 81<!-- TODO keep updated -->
   plant species relevant to the subsistence of Late Epipalaeolithic (15–11.7 ka) 

### 2025-01-08 analysis/paper.qmd
Revise species list

@@
   email: joeroe@hey.com
+  url: https://joeroe.io
   affiliations:
@@
   This paper presents the first continuous, spatially-explicit reconstructions 
-  of the palaeodistributions of 56 plant species found regularly in association 
+  of the palaeodistributions of 81<!-- TODO update --> plant species found regularly in association 
   with early agricultural archaeological sites in West Asia, including the 
@@
   palaeoclimate simulations.
-  Species ranges are predicted to have been on significantly smaller in the 
-  Early Holocene compared to present conditions, with many species that are 
-  found throughout the region's 'hilly flanks' today indicated to have much more
-  restricted distributions centered on the Levant, Cyprus and Western Anatolia. 
-  Ranges shrunk by on average c. 10% from the terminal Pleistocene to the Early 
-  Holocene, with a further c. 10% reduction for the duration of the Younger 
-  Dryas. However, modelled ranges do not reliably predict the observed 
-  occurrence of specific species at archaeological sites. [...]
-  The regional 
-  ubiquity of species in the archaeological record is [not] correlated with the 
+  The models predict significant reductions and/or shifts in species ranges in
+  the terminal Pleistocene and Early Holocene compared to present conditions. 
+  Many species that are found throughout the region's 'hilly flanks' today 
+  are indicated to have much more restricted distributions centered on the 
+  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
+  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
+  Modelled ranges do not reliably predict the observed  occurrence of specific 
+  species at archaeological sites,
+  however [...]
+  The regional ubiquity of species in the archaeological record is [not] correlated with the 
   predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
@@
 - [ ] Copyedit
-- [ ] Appendix with all hindcast predictions (HTML?)
+- [x] Appendix with all hindcast predictions (HTML?)
 - [ ] Clean up package code
@@
 
-neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")
+epipal <- c("Late Epipal.")
+ppn <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C")
 
@@
   code = c("ba", "yds", "eh", "cur"),
-  label = fct_inorder(c("BA (14.7-12.9 ka)", "YDS (12.9-11.7 ka)", 
-                      "EH (11.7-8.3 ka)", "current")),
+  date_range = c("(14.7-12.9 ka)", "(12.9-11.7 ka)", "(11.7-8.3 ka)", ""),
+  label = fct_inorder(paste(toupper(code), date_range)),
+  long_label = fct_inorder(paste(period, date_range)),
   start_bp = c(14700, 12900, 11700, NA),
@@
 ```{r define-parameters}
-# Minimum number of GBIF occurrences for modelling
+# Minimum number archaeological occurrences for modelling
+min_n_archaeo <- 1
+
+# Minimum number of present occurrences for modelling
 min_n_occ <- 50
@@
     )
-  ) ->
-  archaeo_flora
-
-n_assemb_neolithic <- archaeo_flora |> 
-  distinct(site_name, phase_code, period) |> 
-  filter(period %in% neolithic) |> 
-  nrow()
-
-n_site_neolithic <- archaeo_flora |> 
-  filter(period %in% neolithic) |> 
-  distinct(site_name) |> 
-  nrow()
+  ) |>
+  # Filter to Epipal/Neo
+  filter(period %in% epipal | period %in% ppn)
+
+n_archaeo_assemb <- nrow(distinct(archaeo_flora, site_name, phase_code))
+n_archaeo_site <- nrow(distinct(archaeo_flora, site_name))
 ```
@@
 ```{r fig-region}
-#| fig-cap: Study region (red) and archaeological sites used to generate modelled flora. Solid circles indicate Neolithic sites.
+#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
 ggplot() + 
@@
   annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
-  layer_spatial(w_asia, colour = "red", fill = NA) +
-  layer_spatial(archaeo_sites, shape = 1) +
-  layer_spatial(filter(archaeo_sites, period %in% neolithic), shape = 16) +
+  layer_spatial(w_asia, fill = NA) +
+  layer_spatial(archaeo_sites, size = 0.5) +
   scale_linewidth_identity() +
@@
 
-<!-- TODO define study region and period -->
-
 ```{r data-flora}
+# TODO: nest archaeo flora in flora to make it easier to keep track of taxon transformations
 if (!file_exists(derived_data("flora.tsv"))) {
@@
   archaeo_flora |>
-    filter(period %in% neolithic) |>
     group_by(taxon) |>
@@
       n_present = length(unique(phase_code)),
-      p_present = n_present / n_assemb_neolithic,
+      p_present = n_present / n_archaeo_assemb,
       .groups = "drop"
     ) |>
-    filter(n_present > 2) |>
+    filter(n_present > min_n_archaeo) |>
     # Exclude taxa not identified to species
@@
 
-We began by considering `r nrow(flora)` distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic (c.
-11.7–6.5 ka) sites in West Asia, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
-<!-- TODO: something about not distinguishing between source of plant remains? -->
+The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
+Our starting point was a list of `r nrow(flora)` taxa (@tbl-taxa) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
+This was based on dataset of `r n_archaeo_assemb` assemblages collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
+We did not attempt to distinguish between the source of the remains;
+archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
+However, we assume that there presence at a site of human settlement at least implies that they were part of the wider ecosystem that supported habitation there.
+
 Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
-So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-occ-count).
+So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-taxa).
 Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.
@@
 
-```{r data-occ-count}
-if (!file_exists(derived_data("flora_occ_count.tsv"))) {
+```{r data-n-occ} 
+if (!file_exists(derived_data("flora_n_occ.tsv"))) {
   message("Reacquiring GBIF occurrence counts...")
@@
     )
-  write_tsv(flora, derived_data("flora_occ_count.tsv"))
+  write_tsv(flora, derived_data("flora_n_occ.tsv"))
 }
+
 flora <- read_tsv(
-  derived_data("flora_occ_count.tsv"),
+  derived_data("flora_n_occ.tsv"),
   col_types = cols(
@@
     p_present = col_double(),
-    n_occ = col_double()
-  )
-)
-```
-
-```{r tbl-occ-count}
-#| tbl-cap: "Recorded occurrence of flora considered in this study in Neolithic and contemporary West Asia"
-flora |>
-  mutate(
-    exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis"),
-    label = if_else(
-      !is.na(taxon_detail),
-      glue::glue("*{taxon}*\n\n<small>incl. {taxon_detail}</small>"),
-      paste0("*", taxon, "*")
-    )
-  ) |>
-  select(label, n_present, p_present, n_occ, exclude) |>
-  arrange(-n_present) |>
-  gt() |>
-  cols_label(
-    label = "Taxon",
-    n_present = "Neolithic",
-    p_present = "(%)",
-    n_occ = "Present"
-  ) |>
-  tab_spanner("Occurrences", columns = c(n_present, p_present, n_occ)) |>
-  # cols_merge_n_pct(n_present, p_present) |>
-  cols_hide(exclude) |>
-  tab_footnote(
-    "Excluded from modelling due to sample size", 
-    cells_body(columns = c(n_occ), rows = exclude),
-    placement = "right"
-  ) |>
-  cols_align("left", c(label)) |>
-  cols_align("right", c(n_present, p_present, n_occ)) |>
-  cols_width(label ~ pct(50), n_present ~ 10, p_present ~ 10, n_occ ~ 20) |>
-  fmt_percent(p_present, decimals = 0) |>
-  fmt_markdown(label) |>
-  sub_missing() |>
-  tab_options(
-    footnotes.marks = "*",
-    table.font.size = 11,
-    table.width = pct(100),
-    latex.use_longtable = TRUE
-  )
-```
-
-Georeferenced occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
-GBIF occurrences marked has having imprecise or duplicate coordinates were excluded from the training dataset, as were fossil records.
-Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` occurrences in West Asia, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
-We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.
-
-```{r data-filter-flora}
-flora <- filter(flora, n_occ > min_n_occ)
-# Why are there 50k occurrences of Avena sterilis??
-flora <- filter(flora, taxon != "Avena sterilis")
+    n_occ = col_integer()
+  )
+)
+
+# Filter small samples
+# ...and one very large one (why are there 50k occurrences of Avena sterilis?)
+flora_all <- flora
+flora <- flora |>
+  filter(n_occ > min_n_occ) |>
+  filter(taxon != "Avena sterilis")
 ```
@@
 ```{r data-occ}
-# TODO: further tidying. Ideas:
+# Cleaning pipeline based on:
 # - https://data-blog.gbif.org/post/gbif-filtering-guide/
-# - https://cran.r-project.org/web/packages/CoordinateCleaner/vignettes/Cleaning_GBIF_data_with_CoordinateCleaner.html
-# - https://www.r-bloggers.com/2021/03/downloading-and-cleaning-gbif-data-with-r/
 if (!file_exists(derived_data("flora_occ.tsv"))) {
@@
     ),
-    occ = map(occ, gbif_drop_imprecise_coords, threshold = 5000),
+    occ = map(occ, filter, occurrenceStatus  == "PRESENT"),
+    occ = map(occ, gbif_drop_imprecise_coords, threshold = 1000),
     occ = map(occ, gbif_drop_fossils),
@@
              latitude)
-    })
-  )
+    }),
+    # Update occ count after cleaning
+    n_occ = map_int(occ, nrow)
+  )
+
+  # Refilter based on cleaned records
+  flora <- filter(flora, n_occ > min_n_occ)
+
   write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
 }
+
 flora <- read_tsv(
@@
     p_present = col_double(),
-    n_occ = col_double(),
+    n_occ = col_integer(),
     gbif_key = col_double(),
@@
 
-```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia for the ", nrow(flora), " modelled taxa (N=", nrow(unnest(flora, occ)))}
+Georeferenced occurrence data for West Asia was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
+GBIF was cleaned to removed fossil occurrences, recorded absences, and records with missing, imprecise (>1 km uncertainty), or invalid coordinates.
+Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
+We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what accounted for such a disproportionately high number of records.
+
+```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia used to train models (N=", nrow(unnest(flora, occ)))}
 flora |>
@@
 There are several ways to do this. 
-We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (:10000) random sample of points, weighted equally against the presences in the regression.
-@ValaviEtAl2022 also recommend using a very large background sample for random forest models.
+We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) random sample of points, weighted equally against the presences in the regression.
+@ValaviEtAl2022 also recommend using a large background sample for random forest models.
 
 ## Predictor data
+
+```{r tbl-predictors}
+# TODO
+```
 
@@
     sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
-    # probability = TRUE
+    importance = "impurity"
   )
@@
   flora,
-  pred = map(niche, collect_predictions),
   niche_test_metrics = map(niche, collect_metrics),
@@
 
-niche_test_metrics <- flora |>
+niche_test_summary <- flora |>
   unnest(niche_test_metrics) |>
@@
 We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
-Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_metrics, .metric == "accuracy")$min)` and `r percent(filter(niche_test_metrics, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "accuracy")$mean)`.
-Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_metrics, .metric == "sens")$min)` and `r percent(filter(niche_test_metrics, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_metrics, .metric == "sens")$mean)`.
-The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_metrics, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_metrics, .metric == "accuracy")$sd, 3)`
+Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_summary, .metric == "accuracy")$min)` and `r percent(filter(niche_test_summary, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "accuracy")$mean)`.
+Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_summary, .metric == "sens")$min)` and `r percent(filter(niche_test_summary, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "sens")$mean)`.
+The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`
 Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
@@
 ```{r niche-archaeo}
+# TODO: need to deal with taxon variants / GBIF normalisation
 # TODO: factor out
 # TODO: what about negatives?
-archaeo_flora_predictions <-
-
-archaeo_flora |>
+archaeo_flora_predictions <-  archaeo_flora |>
   filter(taxon %in% flora$taxon) |>
-  st_as_sf(coords = c("longitude", "latitude"), crs = latlong) |>
+  st_as_sf(coords = c("longitude", "latitude"), crs = latlong, remove = FALSE) |>
   mutate(
@@
   ) |>
-  unnest(c(prediction))
+  unnest(c(prediction)) |>
+  as_tibble() |>
+  select(-geometry)
+
+write_tsv(archaeo_flora_predictions, derived_data("archaeo_flora_predictions.tsv"))
 
@@
 
-# Results
+```{r tbl-taxa}
+#| tbl-cap: "Summary of species modelled"
+# TODO missing excluded species
+# TODO: redo with:
+# - alt. taxons as footnotes
+# - check boxes of which (climate?) periods observed in ?
+flora |>
+  # Summarise archaeo occurrences by period
+
+  # Extract model metrics
+  unnest(c(niche_test_metrics)) |>
+  select(-.estimator, -.config) |>
+  pivot_wider(names_from = .metric, values_from = .estimate) |>
+  # Excluded samples
+  mutate(exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis")) |>
+  # Prepare table
+  select(taxon, n_present, n_occ, exclude, accuracy, roc_auc, sens) |>
+  arrange(taxon) |>
+  # Table
+  gt() |>
+  cols_label_with(fn = str_to_sentence) |>
+  cols_label(
+    n_present = "Archaeological",
+    n_occ = "Present (GBIF)",
+    roc_auc = "ROC-AUC",
+    sens = "Sensitivity"
+  ) |>
+  tab_spanner("Occurrences", columns = c(n_present, n_occ)) |>
+  tab_spanner("Model", columns = accuracy:sens) |>
+  cols_hide(exclude) |>
+  tab_footnote(
+    "Excluded from modelling due to sample size", 
+    cells_body(columns = c(n_occ), rows = exclude),
+    placement = "right"
+  ) |>
+  cols_align("right") |>
+  cols_align("left", c(taxon)) |>
+  fmt_number(accuracy:sens) |>
+  sub_missing() |>
+  tab_style(cell_text(style = "italic"), cells_body(taxon)) |>
+  tab_options(
+    footnotes.marks = "*",
+    table.font.size = 11,
+    table.width = pct(100),
+    latex.use_longtable = TRUE
+  )
+```
+
+# Discussion
 
@@
 
-```{r fig-range-density}
-#| fig-cap: Distribution of predicted species ranges by period. Dashed lines indicate the median range.
+```{r data-pred-range-summary}
 flora_pred_area_present_summary <- flora |>
@@
 
+n_reduced_range <- flora |>
+  unnest(area_present) |>
+  pivot_wider(names_from = period, values_from = area_present) |>
+  filter(eh < ba) |>
+  nrow()
+
+n_reduced_range_10 <- flora |>
+  unnest(area_present) |>
+  pivot_wider(names_from = period, values_from = area_present) |>
+  filter(eh <= ba * 0.9) |>
+  nrow()
+```
+
+```{r fig-pred-range-density}
+#| fig-cap: Distribution of predicted species ranges by period. Dashed lines indicate the median range.
 flora |>
@@
 
-Species ranges are predicted to have been signicantly smaller in the terminal Pleistocene and Early Holocene that under current conditions [@fig-range-density], though the magnitude of this change is likely also to reflect a degree of overfitting in the model (see below).
-Fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable.
-The average range of modelled species shrank by `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` over the Pleistocene–Holocene boundary, with a noticeable dip in the Younger Dryas.
+Our reconstructed palaeodistributions (shown in full in the appendix) indicate that the majority of species have experienced a significant reduction and/or shift in their ranges since the Early Holocene.
+`r n_reduced_range` of `r nrow(flora)` species have reduced ranges; `r n_reduced_range_10` of more than 10% or more.
+Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further below), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable [@fig-range-density].
+The average range of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` during the Younger Dryas (i.e.
+ranges recovered slightly between the Younger Dryas and Early Holocene).
 This perhaps indicates that although this period is considered one of climatatic amelioration globally [@JonesEtAl2019], the colder conditions of the Pleistocene may have supported more extensive plant-based economies in West Asia specifically.
 
-<!-- TODO: map of sum species abundance? -->
-
-Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene.
+```{r data-palaeodist-sum}
+palaeodist_sum <- do.call(c, c(flora$palaeodist, list(along = "taxon"))) |>
+  st_apply(c("x", "y", "period"), sum, na.rm = FALSE)
+```
+
+```{r fig-palaeodist-sum}
+#| fig-cap: Predicted species richness (sum of predicted ranges) by period
+# TODO: expand period facet labels
+ggplot() +
+  facet_wrap(vars(factor(period, levels = climate_periods$code, labels = climate_periods$label))) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(palaeodist_sum, prediction.TRUE.), 
+             na.action = na.omit) +
+  annotation_spatial(ne_land, fill = NA) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(crs = w_asia_albers) +
+  labs(fill = "Richness", x = NULL, y = NULL) +
+  theme(
+    strip.background = element_blank(),
+    strip.text.y = element_text(face = "italic")
+  )
+```
+
+Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene (@fig-sum-palaeodist).
 These include *Ficus carica* (fig);
@@
 
-```{r fig-hindcast-barely-levant, eval=FALSE, cache=TRUE}
-#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
+```{r fig.height=10, fig-palaeodist-barely-pistachio-levant, eval=FALSE, cache=TRUE}
+#| fig-cap: Contraction of reconstructed range of of wild barleys and pistachio in the Levant
 palaeodist_hordeum <- do.call(c, c(
-  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
-  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
+  filter(flora, str_starts(taxon, "Hordeum") | taxon == "Pistacia atlantica")$palaeodist,
+  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum") | taxon == "Pistacia atlantica")$taxon))
 ))
 
-# TODO: expand period facet labels
 ggplot() +
-  facet_grid(taxon ~ period) +
+  facet_grid(taxon ~ fct_relabel(period, toupper)) +
   annotation_spatial(ne_land, fill = "white") +
@@
   annotation_spatial(ne_land, fill = NA) +
-  annotation_spatial(w_asia, fill = NA) +
   scale_x_continuous(breaks = c(34, 36, 38)) +
@@
 
-```{r fig-hindcast-wheat-levant, eval=FALSE, cache=TRUE}
-#| fig-cap: Predicted palaeodistribution of wild wheat in the Levant
-palaeodist_triticum <- do.call(c, c(
-  filter(flora, str_starts(taxon, "Triticum"))$palaeodist,
-  list(along = list(taxon = filter(flora, str_starts(taxon, "Triticum"))$taxon))
-))
-
-# TODO: expand period facet labels
-ggplot() +
-  facet_grid(taxon ~ period) +
-  annotation_spatial(ne_land, fill = "white") +
-  geom_stars(data = select(palaeodist_triticum, prediction.TRUE.)) + 
-  annotation_spatial(ne_land, fill = NA) +
-  annotation_spatial(w_asia, fill = NA) +
-  scale_x_continuous(breaks = c(34, 36, 38)) +
-  scale_fill_batlow(reverse = TRUE) +
-  coord_sf(
-    xlim = c(33, 39), ylim = c(30, 38),
-    crs = w_asia_albers, default_crs = latlong
-  ) +
-  labs(fill = "P(present)", x = NULL, y = NULL) +
-  theme(
-    strip.background = element_blank(),
-    strip.text.y = element_text(face = "italic")
-  )
-```
-
-<!-- TODO binarized range figure other founder crops? -->
-
-Wild barley [*Hordeum spontaneum*, @fig-hindcast-barely-levant] and pistachio [*Pistacia atlantica*, @fig] show a contraction of their predicted ranges from the Pleistocene to the Holocene,
-concurrent with them being brought into cultivation.
-Plausible that this prompted cultivation?
-These two species [@ArranzOtaeguiRoe2023] also see marked declines in the archaeobotanical record from the Early PPNA/Early PPNB (where they were amongst the most common taxa) to the Late PPNB and Late Neolithic.
-However, this trend is not seen in most other crop progenitors.
-
-Cereals / other progenitors (not all) reduced in range before cultivation?
-Reduction in range from BA to EH very evident in Hordeum, less so in ...
-
-<!-- TODO: composite wheat progentor figure -->
+Wild barley (*Hordeum spontaneum*), its relatives, and pistachio (*Pistacia atlantica*) show a contraction of their predicted ranges from the Pleistocene to the Holocene,
+concurrent with them being brought into cultivation [@fig-palaeodist-barely-pistachio-levant].
+These two species also see marked declines in the archaeobotanical record from the Early PPNA/Early PPNB (where they were amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
+Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end.
+We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
+the decreasing availability of barley and pistachio in the terminal Pleistocene may have prompted cultivation as a strategy to retain access to them,
+though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
+However, these dynamics are not seen in the majority of crop progenitors.
 
@@
 <!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
-Is inconsistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
-
-# Discussion
-
-* We present the first continuous, spatially explicit models of the palaeodistributions of 56 plant species found regularly in association with early agricultural archaeological sites in West Asia
+Is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
+
+# Conclusion
+
+* We present the first continuous, spatially explicit models of the palaeodistributions of N plant species found regularly in association with early agricultural archaeological sites in West Asia
   * A new line of evidence on archaeoecology
@@
 # References {.appendix}
-
-----
-
-```{r fig-niche-prediction, eval=FALSE}
-# "Adonis flammea"                   "Aegilops crassa"                  "Aizoanthemopsis hispanica"       
-# "Ammi majus"                       "Androsace maxima"                 "Arnebia decumbens"               
-# "Arnebia linearifolia"             "Atriplex prostrata"               "Bassia arabica"                  
-# "Bolboschoenus maritimus"          "Brachypodium distachyon"          "Bromus sterilis"                 
-# "Buglossoides arvensis"            "Buglossoides tenuiflora"          "Capparis spinosa"                
-# "Carex divisa"                     "Cephalaria syriaca"               "Chenopodium album"               
-# "Cicer reticulatum"                "Ficus carica"                     "Fumaria densiflora"              
-# "Gypsophila pilosa"                "Helianthemum ledifolium"          "Hordeum bulbosum"                
-# "Hordeum murinum"                  "Hordeum spontaneum"               "Lathyrus aphaca"                 
-# "Lathyrus sativus"                 "Vicia orientalis"                 "Linum bienne"                    
-# "Lolium rigidum"                   "Lolium temulentum"                "Medicago radiata"                
-# "Phalaris paradoxa"                "Pistacia atlantica"               "Lathyrus oleraceus"              
-# "Poa bulbosa"                      "Prosopis farcta"                  "Quercus ithaburensis"            
-# "Rumex pulcher"                    "Salsola kali"                     "Secale cereale"                  
-# "Taeniatherum caput-medusae"       "Medicago astroites"               "Triticum aestivum compactum"     
-# "Triticum turgidum dicoccum"       "Triticum monococcum aegilopoides" "Triticum durum"                  
-# "Gypsophila vaccaria"              "Verbena officinalis"              "Vicia ervilia"                   
-# "Vicia faba"                       "Vicia narbonensis"                "Triticum urartu"                 
-# "Aegilops speltoides"              "Aegilops tauschii"
-
-niche_taxon <- "Lathyrus oleraceus"
-niche_prediction <- filter(flora, taxon == niche_taxon)$palaeodist[[1]]
-
-# Order and relabel periods
-ggplot() + 
-  facet_wrap(~period, ncol = 2) +
-  annotation_spatial(ne_land, fill = "white") +
-  geom_stars(data = select(niche_prediction, prediction.TRUE.),
-             na.action = na.omit) +
-  annotation_spatial(ne_land, fill = NA) +
-  annotation_spatial(w_asia, fill = NA) +
-  scale_fill_batlow(reverse = TRUE) +
-  coord_sf(crs = w_asia_albers) +
-  labs(
-    title = niche_taxon,
-    fill = NULL,
-    x = NULL,
-    y = NULL
-  ) +
-  theme(
-    plot.title = element_text(face = "italic")
-  )
-```
-
-```{r sum-pred-eh, eval=FALSE}
-flora |>
-  mutate(pred_eh = map(fit, \(x) predict(climate$EH, model = x, type = "prob"))) ->
-  x
-
-pred_eh <- x$pred_eh
-names(pred_eh) <- x$taxon
-pred_eh <- do.call(c, map(pred_eh, select, .pred_TRUE))
-pred_eh <- merge(pred_eh, name = "taxon")
-sum_pred_eh <- st_apply(pred_eh, c("x", "y"), sum, na.rm = TRUE)
-```
-
-```{r fig-sum-pred-eh, eval=FALSE}
-#| fig-cap: predicted species richness for the early Holocene
-ggplot() + 
-  geom_stars(data = sum_pred_eh) +
-  scale_fill_batlow(reverse = FALSE) +
-  coord_sf(crs = latlong) +
-  labs(
-    title = "Predicted species richness",
-    subtitle = "Early Holocene (11.7–8.326 ka)",
-    fill = NULL,
-    x = NULL,
-    y = NULL
-  )
-
-saveRDS(pred_eh, "pred_eh.Rdata")
-```

### 2025-01-08 analysis/paper.tex
Revise species list

@@
 This paper presents the first continuous, spatially-explicit
-reconstructions of the palaeodistributions of 56 plant species found
+reconstructions of the palaeodistributions of 81 plant species found
 regularly in association with early agricultural archaeological sites in
@@
 from these models using downsampled data from palaeoclimate simulations.
-Species ranges are predicted to have been on significantly smaller in
-the Early Holocene compared to present conditions, with many species
-that are found throughout the region's `hilly flanks' today indicated to
-have much more restricted distributions centered on the Levant, Cyprus
-and Western Anatolia. Ranges shrunk by on average c.~10\% from the
-terminal Pleistocene to the Early Holocene, with a further c.~10\%
-reduction for the duration of the Younger Dryas. However, modelled
-ranges do not reliably predict the observed occurrence of specific
-species at archaeological sites. {[}\ldots{]} The regional ubiquity of
+The models predict significant reductions and/or shifts in species
+ranges in the terminal Pleistocene and Early Holocene compared to
+present conditions. Many species that are found throughout the region's
+`hilly flanks' today are indicated to have much more restricted
+distributions centered on the Levant, Cyprus and Western Anatolia. In
+addition, overall ranges shrunk by an average of c.~25\% from the
+terminal Pleistocene to the Early Holocene. Modelled ranges do not
+reliably predict the observed occurrence of specific species at
+archaeological sites, however {[}\ldots{]} The regional ubiquity of
 species in the archaeological record is {[}not{]} correlated with the
@@
   \tightlist
-  \item[$\square$]
+  \item[$\boxtimes$]
     Introduction
@@
   Copyedit
-\item[$\square$]
+\item[$\boxtimes$]
   Appendix with all hindcast predictions (HTML?)
@@
 
-\caption{\label{fig-region}Study region (red) and archaeological sites
-used to generate modelled flora. Solid circles indicate Neolithic
-sites.}
+\caption{\label{fig-region}Late Epipalaeolithic \& Pre-Pottery Neolithic
+archaeological sites used to generate modelled flora}
 
@@
 
-We began by considering 62 distinct taxa (Table~\ref{tbl-occ-count}) -
-all the identifiable species known to be present at more than three
-Neolithic (c.~11.7--6.5 ka) sites in West Asia, according to our
-previous study of the regional archaeobotanical data
-\citep{ArranzOtaeguiRoe2023}. Taxonomic names were resolved to the
-canonical form specified in the GBIF Backbone Taxonomy
-\citep{GBIFSecretariat2023}. So for example occurrences for
-\emph{Bolboschoenus maritimus} also include those recorded under the
-older nomenclature \emph{Scirpus maritimus} (see
-Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
-criteria were substituted with their wild progenitor(s), where
-different.
+The aim of our study was to model the biogeography of species relevant
+to human subsistence economies in West Asia (excluding the Southern
+Arabian peninsula, see Figure~\ref{fig-region}) during the
+archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
+Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 96
+taxa (Table~\ref{tbl-taxa}) comprising the identifiable species observed
+at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites,
+according to our previous study of the regional archaeobotanical data
+\citep{ArranzOtaeguiRoe2023}. This was based on dataset of 117
+assemblages collated from three previously published regional
+archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
+\citep{ORIGINS}, and COMPAG \citep{COMPAG}. We did not attempt to
+distinguish between the source of the remains; archaeobotanical
+assemblages are subject to a variety of preservational and recovery
+biases, so by no means were all the species on our list consumed or even
+deliberately collected by people. However, we assume that there presence
+at a site of human settlement at least implies that they were part of
+the wider ecosystem that supported habitation there.
+
+Taxonomic names were resolved to the canonical form specified in the
+GBIF Backbone Taxonomy \citep{GBIFSecretariat2023}. So for example
+occurrences for \emph{Bolboschoenus maritimus} also include those
+recorded under the older nomenclature \emph{Scirpus maritimus} (see
+Table~\ref{tbl-taxa}). Domestic species meeting our inclusion criteria
+were substituted with their wild progenitor(s), where different.
 
@@
 
-\begin{verbatim}
-Warning: HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-HTML tags found, and they will be removed.
-* Set `options(gt.html_tag_check = FALSE)` to disable this check.
-\end{verbatim}
-
-\begingroup
-\setlength\LTleft{0\linewidth}
-\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont
-\setlength{\LTpost}{0mm}
-
-\begin{longtable}{@{\extracolsep{\fill}}>{\raggedright\arraybackslash}p{\dimexpr 0.50\linewidth -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 7.50pt -2\tabcolsep-1.5\arrayrulewidth}>{\raggedleft\arraybackslash}p{\dimexpr 15.00pt -2\tabcolsep-1.5\arrayrulewidth}}
-
-\caption{\label{tbl-occ-count}Recorded occurrence of flora considered in
-this study in Neolithic and contemporary West Asia}
-
-\tabularnewline
-
-\toprule
- & \multicolumn{3}{>{\centering\arraybackslash}m{\dimexpr 30.00pt -2\tabcolsep-1.5\arrayrulewidth}}{Occurrences} \\ 
-\cmidrule(lr){2-4}
-Taxon & Neolithic & (\%) & Present \\ 
-\midrule\addlinespace[2.5pt]
-\emph{Triticum turgidum dicoccum}
-
-incl. \emph{Triticum aestivum}, \emph{Triticum dicoccoides}, \emph{Triticum dicoccum} & 64 & 44\% & 200 \\ 
-\emph{Hordeum spontaneum}
-
-incl. \emph{Hordeum vulgare} & 62 & 43\% & 6912 \\ 
-\emph{Triticum monococcum aegilopoides}
-
-incl. \emph{Triticum boeoticum}, \emph{Triticum monococcum} & 44 & 30\% & 3213 \\ 
-\emph{Bolboschoenus maritimus}
-
-incl. \emph{Scirpus maritimus} & 33 & 23\% & 364 \\ 
-\emph{Vicia ervilia} & 33 & 23\% & 858 \\ 
-\emph{Buglossoides tenuiflora} & 31 & 21\% & 159 \\ 
-\emph{Arnebia decumbens} & 25 & 17\% & 240 \\ 
-\emph{Buglossoides arvensis} & 25 & 17\% & 251 \\ 
-\emph{Medicago radiata} & 21 & 14\% & 883 \\ 
-\emph{Androsace maxima} & 20 & 14\% & 130 \\ 
-\emph{Vicia orientalis}
-
-incl. \emph{Lens culinaris} & 17 & 12\% & 441 \\ 
-\emph{Medicago astroites}
-
-incl. \emph{Trigonella astroites} & 16 & 11\% & 130 \\ 
-\emph{Arnebia linearifolia} & 14 & 10\% & 158 \\ 
-\emph{Linum bienne}
-
-incl. \emph{Linum usitatissimum} & 11 & 8\% & 274 \\ 
-\emph{Gypsophila vaccaria}
-
-incl. \emph{Vaccaria pyramidata} & 11 & 8\% & 326 \\ 
-\emph{Carex divisa} & 10 & 7\% & 330 \\ 
-\emph{Ficus carica} & 10 & 7\% & 4265 \\ 
-\emph{Lathyrus oleraceus}
-
-incl. \emph{Pisum elatius}, \emph{Pisum sativum} & 8 & 6\% & 2269 \\ 
-\emph{Vicia faba} & 8 & 6\% & 2578 \\ 
-\emph{Aizoanthemopsis hispanica}
-
-incl. \emph{Aizoon hispanicum} & 7 & 5\% & 303 \\ 
-\emph{Bolboschoenus glaucus} & 7 & 5\% & 27\textsuperscript{\textit{*}} \\ 
-\emph{Pistacia atlantica} & 7 & 5\% & 928 \\ 
-\emph{Polygonum arenarium arenarium}
-
-incl. \emph{Polygonum venantianum} & 7 & 5\% & 5\textsuperscript{\textit{*}} \\ 
-\emph{Prosopis farcta} & 7 & 5\% & 1205 \\ 
-\emph{Rumex pulcher} & 7 & 5\% & 794 \\ 
-\emph{Ammi majus} & 6 & 4\% & 461 \\ 
-\emph{Cicer reticulatum}
-
-incl. \emph{Cicer arietinum} & 6 & 4\% & 193 \\ 
-\emph{Hordeum bulbosum} & 6 & 4\% & 2036 \\ 
-\emph{Polygonum corrigioloides} & 6 & 4\% & 4\textsuperscript{\textit{*}} \\ 
-\emph{Salsola kali} & 6 & 4\% & 369 \\ 
-\emph{Taeniatherum caput-medusae} & 6 & 4\% & 235 \\ 
-\emph{Capparis spinosa} & 5 & 3\% & 4031 \\ 
-\emph{Chenopodium album} & 5 & 3\% & 357 \\ 
-\emph{Lolium temulentum} & 5 & 3\% & 148 \\ 
-\emph{Poa bulbosa} & 5 & 3\% & 1401 \\ 
-\emph{Aegilops crassa} & 4 & 3\% & 170 \\ 
-\emph{Atriplex prostrata} & 4 & 3\% & 112 \\ 
-\emph{Brachypodium distachyon} & 4 & 3\% & 1581 \\ 
-\emph{Fumaria densiflora} & 4 & 3\% & 194 \\ 
-\emph{Gypsophila pilosa} & 4 & 3\% & 79 \\ 
-\emph{Hordeum murinum} & 4 & 3\% & 1342 \\ 
-\emph{Lathyrus aphaca} & 4 & 3\% & 2085 \\ 
-\emph{Lathyrus sativus} & 4 & 3\% & 297 \\ 
-\emph{Secale cereale} & 4 & 3\% & 1153 \\ 
-\emph{Triticum durum} & 4 & 3\% & 219 \\ 
-\emph{Vitis sylvestris} & 4 & 3\% & 3\textsuperscript{\textit{*}} \\ 
-\emph{Adonis flammea} & 3 & 2\% & 79 \\ 
-\emph{Avena sterilis} & 3 & 2\% & 50001\textsuperscript{\textit{*}} \\ 
-\emph{Bassia arabica} & 3 & 2\% & 230 \\ 
-\emph{Bromus sterilis} & 3 & 2\% & 385 \\ 
-\emph{Cephalaria syriaca} & 3 & 2\% & 144 \\ 
-\emph{Glaucium aleppicum} & 3 & 2\% & 22\textsuperscript{\textit{*}} \\ 
-\emph{Helianthemum ledifolium} & 3 & 2\% & 668 \\ 
-\emph{Lolium rigidum} & 3 & 2\% & 1464 \\ 
-\emph{Phalaris paradoxa} & 3 & 2\% & 370 \\ 
-\emph{Quercus ithaburensis} & 3 & 2\% & 1431 \\ 
-\emph{Triticum aestivum compactum} & 3 & 2\% & 209 \\ 
-\emph{Verbena officinalis} & 3 & 2\% & 1046 \\ 
-\emph{Vicia narbonensis} & 3 & 2\% & 1529 \\ 
-\emph{Triticum urartu} & — & — & 2033 \\ 
-\emph{Aegilops speltoides} & — & — & 1429 \\ 
-\emph{Aegilops tauschii} & — & — & 1395 \\ 
-\bottomrule
-
-\end{longtable}
-
-\begin{minipage}{\linewidth}
-\textsuperscript{\textit{*}}Excluded from modelling due to sample size\\
-\end{minipage}
-\endgroup
-
-Georeferenced occurrence data was obtained from the Global Biodiversity
-Information Facility (GBIF) using via its application programming
-interface and the R package `rgbif'
-\citep{ChamberlainBoettiger2017, rgbif}. GBIF occurrences marked has
-having imprecise or duplicate coordinates were excluded from the
-training dataset, as were fossil records. Although niche models have
-reasonable predictive power even with small training samples
+Georeferenced occurrence data for West Asia was obtained from the Global
+Biodiversity Information Facility (GBIF) using via its application
+programming interface and the R package `rgbif'
+\citep{ChamberlainBoettiger2017, rgbif}. GBIF was cleaned to removed
+fossil occurrences, recorded absences, and records with missing,
+imprecise (\textgreater1 km uncertainty), or invalid coordinates.
+Although niche models have reasonable predictive power even with small
+training samples
 \citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
-excluded 57 taxa with less than 50 occurrences in West Asia, following
+excluded 81 taxa with less than 50 usable occurrences, following
 recommendations for niche models generally and Random Forest-based
@@
 as this would have been computationally prohibitive and we were
-uncertain what account for such a disproportionately high number of
+uncertain what accounted for such a disproportionately high number of
 records.
@@
 \caption{\label{fig-occ-map}Georeferenced occurrence records from West
-Asia for the 56 modelled taxa (N=35208}
+Asia used to train models (N=78222}
 
@@
 this. We follow the advice of \citet{BarbetMassinEtAl2012} for
-regression-based species distribution models and use a large (:10000)
+regression-based species distribution models and use a large (≈10000)
 random sample of points, weighted equally against the presences in the
-regression. \citet{ValaviEtAl2022} also recommend using a very large
+regression. \citet{ValaviEtAl2022} also recommend using a large
 background sample for random forest models.
@@
 We modelled the occurrence of species as a function of X spatial
-predictor variables (\textbf{?@tbl-predictors}). These included:
+predictor variables (Table~\ref{tbl-predictors}). These included:
 
@@
 (presence or absence/background) based on our X predictor variables
-(\textbf{?@tbl-predictors}). We used the Random Forest algorithm
+(Table~\ref{tbl-predictors}). We used the Random Forest algorithm
 implemented in the R package `ranger' \citep{WrightZiegler2017} and the
@@
 
-We trained Random Forest models for 56 taxa using contemporary
+We trained Random Forest models for 81 taxa using contemporary
 occurrence data from GBIF, a random sample of background points, and the
-predictor variables described in \textbf{?@tbl-predictors}. Substituting
-the ``current'' climate predictors for those derived from palaeoclimatic
-simulations \citep{BrownEtAl2018}, we could then generate hindcast
-predictions for reconstructed past environments in 4 key climate periods
--- a total of 224 modelled palaeodistributions. Predicted distributions
-for individual taxa are presented in the appendix and accompanying
-material.
+predictor variables described in Table~\ref{tbl-predictors}.
+Substituting the ``current'' climate predictors for those derived from
+palaeoclimatic simulations \citep{BrownEtAl2018}, we could then generate
+hindcast predictions for reconstructed past environments in 4 key
+climate periods -- a total of 324 modelled palaeodistributions.
+Predicted distributions for individual taxa are presented in the
+appendix and accompanying material.
 
@@
 accuracy (proportion of correctly classified presence and background
-samples) ranged between 72\% and 98\%, with an average of 91\%.
+samples) ranged between 72\% and 99\%, with an average of 92\%.
 Sensitivity (proportion of correctly classified presence samples) ranged
-between 80\% and 100\%, with an average of 93\%. The area under the
+between 79\% and 100\%, with an average of 94\%. The area under the
 models' receiver operating characteristic curves (ROC-AUC) was in the
-range of 0.973±0.058 Model sensitivity is loosely correlated with the
+range of 0.978±0.061 Model sensitivity is loosely correlated with the
 number of occurrences available for training
 (Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
-models all having less than 300 recorded occurrences: \emph{Buglossoides
-tenuiflora}, \emph{Cephalaria syriaca}, \emph{Gypsophila pilosa},
-\emph{Lathyrus sativus}, and \emph{Triticum turgidum dicoccum}. Test
-metrics and ROC curves for the individual models are included in the
-appendix.
+models all having less than 300 recorded occurrences: \emph{Avena
+fatua}, \emph{Lepidium perfoliatum}, \emph{Triticum turgidum dicoccum},
+and \emph{Zygophyllum fabago}. Test metrics and ROC curves for the
+individual models are included in the appendix.
 
 The ability of the hindcast models to predict the occurrence of specific
-species at archaeological sites is significantly worse, with only 12\%
-of presences in archaeobotanical assemblages successfully predicted.
-
-\section{Results}\label{results}
+species at archaeological sites is significantly worse, with only 5\% of
+presences in archaeobotanical assemblages successfully predicted.
+
+\section{Discussion}\label{discussion}
+
+\begingroup
+\setlength\LTleft{0\linewidth}
+\setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont
+
+\begin{longtable}{@{\extracolsep{\fill}}lrrrrr}
+
+\caption{\label{tbl-taxa}Summary of species modelled}
+
+\tabularnewline
+
+\toprule
+ & \multicolumn{2}{c}{Occurrences} & \multicolumn{3}{c}{Model} \\ 
+\cmidrule(lr){2-3} \cmidrule(lr){4-6}
+Taxon & Archaeological & Present (GBIF) & Accuracy & ROC-AUC & Sensitivity \\ 
+\midrule\addlinespace[2.5pt]
+{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.99 \\ 
+{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.89 \\ 
+{\itshape Aegilops speltoides} & — & 701 & 0.93 & 0.98 & 0.94 \\ 
+{\itshape Aegilops tauschii} & — & 725 & 0.90 & 0.98 & 0.94 \\ 
+{\itshape Aizoanthemopsis hispanica} & 4 & 515 & 0.98 & 0.99 & 0.96 \\ 
+{\itshape Ammi majus} & 4 & 900 & 0.99 & 0.99 & 0.98 \\ 
+{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.96 & 0.88 \\ 
+{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.85 & 0.94 & 0.91 \\ 
+{\itshape Arnebia decumbens} & 21 & 324 & 0.94 & 0.98 & 0.93 \\ 
+{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 1.00 & 0.98 \\ 
+{\itshape Asphodelus aestivus} & 2 & 76 & 0.96 & 0.97 & 0.96 \\ 
+{\itshape Atriplex prostrata} & 4 & 266 & 0.97 & 1.00 & 0.97 \\ 
+{\itshape Avena fatua} & 2 & 579 & 0.85 & 0.93 & 0.85 \\ 
+{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 0.98 \\ 
+{\itshape Bolboschoenus maritimus} & 29 & 1040 & 0.98 & 1.00 & 0.99 \\ 
+{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 \\ 
+{\itshape Bromus sterilis} & 3 & 759 & 0.96 & 1.00 & 0.98 \\ 
+{\itshape Buglossoides arvensis} & 23 & 260 & 0.86 & 0.96 & 0.90 \\ 
+{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.94 \\ 
+{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 \\ 
+{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 \\ 
+{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.93 \\ 
+{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 \\ 
+{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.92 \\ 
+{\itshape Cicer reticulatum} & 3 & 52 & 0.96 & 1.00 & 0.92 \\ 
+{\itshape Citrullus colocynthis} & 4 & 475 & 0.89 & 0.95 & 0.89 \\ 
+{\itshape Crithopsis delileana} & 3 & 406 & 0.97 & 0.99 & 0.97 \\ 
+{\itshape Erodium ciconium} & 2 & 359 & 0.91 & 0.99 & 0.99 \\ 
+{\itshape Ficus carica} & 8 & 3495 & 0.93 & 0.98 & 0.93 \\ 
+{\itshape Fumaria densiflora} & 4 & 413 & 0.98 & 0.99 & 0.98 \\ 
+{\itshape Gypsophila elegans} & 3 & 76 & 0.93 & 0.96 & 0.85 \\ 
+{\itshape Gypsophila pilosa} & 5 & 173 & 0.93 & 0.99 & 0.95 \\ 
+{\itshape Gypsophila vaccaria} & 8 & 511 & 0.96 & 0.99 & 0.93 \\ 
+{\itshape Halothamnus hierochunticus} & 2 & 57 & 0.99 & 1.00 & 1.00 \\ 
+{\itshape Helianthemum ledifolium} & 2 & 697 & 0.97 & 0.98 & 0.96 \\ 
+{\itshape Helianthemum salicifolium} & 2 & 1400 & 0.98 & 0.99 & 0.97 \\ 
+{\itshape Hordeum bulbosum} & 4 & 2850 & 0.97 & 0.99 & 0.96 \\ 
+{\itshape Hordeum murinum} & 5 & 1949 & 0.95 & 0.99 & 0.96 \\ 
+{\itshape Hordeum spontaneum} & 51 & 4656 & 0.96 & 0.99 & 0.97 \\ 
+{\itshape Juglans regia} & 2 & 821 & 0.85 & 0.96 & 0.94 \\ 
+{\itshape Lathyrus aphaca} & 4 & 2079 & 0.95 & 0.99 & 0.96 \\ 
+{\itshape Lathyrus cicera} & 2 & 591 & 0.90 & 0.98 & 0.94 \\ 
+{\itshape Lathyrus oleraceus} & 6 & 1580 & 0.88 & 0.97 & 0.94 \\ 
+{\itshape Lathyrus sativus} & 4 & 196 & 0.80 & 0.94 & 0.87 \\ 
+{\itshape Lepidium perfoliatum} & 3 & 83 & 0.80 & 0.92 & 0.83 \\ 
+{\itshape Linum bienne} & 9 & 376 & 0.96 & 1.00 & 0.98 \\ 
+{\itshape Lolium rigidum} & 5 & 2529 & 0.98 & 1.00 & 0.98 \\ 
+{\itshape Lolium temulentum} & 3 & 159 & 0.93 & 0.95 & 0.92 \\ 
+{\itshape Medicago astroites} & 15 & 143 & 0.90 & 0.97 & 0.88 \\ 
+{\itshape Medicago minima} & 2 & 1379 & 0.92 & 0.99 & 0.98 \\ 
+{\itshape Medicago radiata} & 20 & 799 & 0.93 & 0.99 & 0.97 \\ 
+{\itshape Moltkia coerulea} & 2 & 71 & 0.86 & 0.99 & 1.00 \\ 
+{\itshape Neotorularia torulosa} & 2 & 539 & 0.98 & 0.99 & 0.96 \\ 
+{\itshape Peganum harmala} & 2 & 917 & 0.91 & 0.98 & 0.95 \\ 
+{\itshape Phalaris paradoxa} & 3 & 1191 & 0.99 & 0.99 & 0.98 \\ 
+{\itshape Phragmites australis} & 4 & 3412 & 0.97 & 0.99 & 0.97 \\ 
+{\itshape Pistacia atlantica} & 6 & 1874 & 0.98 & 1.00 & 0.98 \\ 
+{\itshape Plantago lagopus} & 2 & 1730 & 0.98 & 1.00 & 0.98 \\ 
+{\itshape Poa bulbosa} & 5 & 1880 & 0.95 & 0.98 & 0.90 \\ 
+{\itshape Prosopis farcta} & 5 & 2263 & 0.97 & 1.00 & 0.97 \\ 
+{\itshape Quercus ithaburensis} & 2 & 2216 & 0.98 & 1.00 & 0.99 \\ 
+{\itshape Rumex pulcher} & 6 & 1394 & 0.98 & 1.00 & 0.99 \\ 
+{\itshape Salsola kali} & 6 & 537 & 0.98 & 1.00 & 1.00 \\ 
+{\itshape Salvia absconditiflora} & 3 & 105 & 0.92 & 0.99 & 0.91 \\ 
+{\itshape Secale cereale} & 4 & 268 & 0.79 & 0.95 & 0.92 \\ 
+{\itshape Secale strictum} & 3 & 124 & 0.74 & 0.96 & 0.92 \\ 
+{\itshape Stipa dregeana} & 2 & 79 & 0.93 & 0.99 & 0.97 \\ 
+{\itshape Suaeda fruticosa} & 3 & 238 & 0.99 & 1.00 & 1.00 \\ 
+{\itshape Taeniatherum caput-medusae} & 4 & 262 & 0.82 & 0.96 & 0.92 \\ 
+{\itshape Tribulus terrestris} & 2 & 1091 & 0.91 & 0.97 & 0.93 \\ 
+{\itshape Triticum aestivum compactum} & 2 & 147 & 0.80 & 0.96 & 0.92 \\ 
+{\itshape Triticum durum} & 3 & 95 & 0.90 & 0.94 & 0.88 \\ 
+{\itshape Triticum monococcum aegilopoides} & 36 & 1176 & 0.88 & 0.98 & 0.96 \\ 
+{\itshape Triticum turgidum dicoccum} & 47 & 60 & 0.72 & 0.85 & 0.79 \\ 
+{\itshape Triticum urartu} & — & 424 & 0.93 & 1.00 & 1.00 \\ 
+{\itshape Verbena officinalis} & 3 & 934 & 0.95 & 0.99 & 0.96 \\ 
+{\itshape Vicia ervilia} & 26 & 688 & 0.88 & 0.97 & 0.93 \\ 
+{\itshape Vicia faba} & 7 & 1597 & 0.91 & 0.98 & 0.95 \\ 
+{\itshape Vicia narbonensis} & 3 & 1451 & 0.93 & 0.98 & 0.94 \\ 
+{\itshape Vicia orientalis} & 14 & 353 & 0.84 & 0.95 & 0.88 \\ 
+{\itshape Zygophyllum fabago} & 3 & 263 & 0.89 & 0.95 & 0.84 \\ 
+\bottomrule
+
+\end{longtable}
+
+\endgroup
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-range-density-1.pdf}}
-
-}
-
-\caption{\label{fig-range-density}Distribution of predicted species
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-pred-range-density-1.pdf}}
+
+}
+
+\caption{\label{fig-pred-range-density}Distribution of predicted species
 ranges by period. Dashed lines indicate the median range.}
@@
 
-Species ranges are predicted to have been signicantly smaller in the
-terminal Pleistocene and Early Holocene that under current conditions
-Figure~\ref{fig-range-density}, though the magnitude of this change is
-likely also to reflect a degree of overfitting in the model (see below).
-Fluctuations in modelled range size between the Bølling-Allerød
+Our reconstructed palaeodistributions (shown in full in the appendix)
+indicate that the majority of species have experienced a significant
+reduction and/or shift in their ranges since the Early Holocene. 75 of
+81 species have reduced ranges; 65 of more than 10\% or more. Though the
+magnitude of the reduction from prehistory to the present likely
+reflects a degree of overfitting in the model (discussed further below),
+fluctuations in modelled range size between the Bølling-Allerød
 (14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
-(11.7--8.3 ka) are more directly comparable. The average range of
-modelled species shrank by 12\% over the Pleistocene--Holocene boundary,
-with a noticeable dip in the Younger Dryas. This perhaps indicates that
-although this period is considered one of climatatic amelioration
-globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
-may have supported more extensive plant-based economies in West Asia
+(11.7--8.3 ka) are more directly comparable
+\textbf{?@fig-range-density}. The average range of modelled species was
+25\% in the Early Holocene compared to the Bølling--Allerød, and 32\%
+during the Younger Dryas (i.e.~ranges recovered slightly between the
+Younger Dryas and Early Holocene). This perhaps indicates that although
+this period is considered one of climatatic amelioration globally
+\citep{JonesEtAl2019}, the colder conditions of the Pleistocene may have
+supported more extensive plant-based economies in West Asia
 specifically.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-sum-1.pdf}}
+
+}
+
+\caption{\label{fig-palaeodist-sum}Predicted species richness (sum of
+predicted ranges) by period}
+
+\end{figure}%
 
@@
 have had a significantly more restricted distribution in the terminal
-Pleistocene/Early Holocene. These include \emph{Ficus carica} (fig);
-\emph{Hordeum} spp. (wild barleys); \emph{Lathyrus aphaca} and \emph{L.
-sativus} (both marginally edible legumes); \emph{Triticum aestivum
-compactum} (in the N. Levant), \emph{T. monococcum aegilopoides},
-\emph{T. durum}, and \emph{Triticum urartu} (but not the other wheat
-progenitor, \emph{T. turgidum dicoccum} -- see below); \emph{Aegilops
-speltoides}, but not \emph{Aegilops tauschii} (goatgrasses); and
-\emph{Vicia} spp. (vetches), including \emph{Vicia faba} (broad beans).
-most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in
-particular disappear from the predicted ranges of these species, leaving
-the Levant and to a lesser extent the Aegean and Cyprus as refugia.
+Pleistocene/Early Holocene (\textbf{?@fig-sum-palaeodist}). These
+include \emph{Ficus carica} (fig); \emph{Hordeum} spp. (wild barleys);
+\emph{Lathyrus aphaca} and \emph{L. sativus} (both marginally edible
+legumes); \emph{Triticum aestivum compactum} (in the N. Levant),
+\emph{T. monococcum aegilopoides}, \emph{T. durum}, and \emph{Triticum
+urartu} (but not the other wheat progenitor, \emph{T. turgidum dicoccum}
+-- see below); \emph{Aegilops speltoides}, but not \emph{Aegilops
+tauschii} (goatgrasses); and \emph{Vicia} spp. (vetches), including
+\emph{Vicia faba} (broad beans). most of Anatolia, Northern Mesopotamia,
+and the Zagros Mountains in particular disappear from the predicted
+ranges of these species, leaving the Levant and to a lesser extent the
+Aegean and Cyprus as refugia.
 
@@
 
-Wild barley \textbf{?@fig-hindcast-barely-levant} and pistachio
-\citep[\emph{Pistacia atlantica},][]{fig} show a contraction of their
-predicted ranges from the Pleistocene to the Holocene, concurrent with
-them being brought into cultivation. Plausible that this prompted
-cultivation? These two species \citep{ArranzOtaeguiRoe2023} also see
-marked declines in the archaeobotanical record from the Early PPNA/Early
-PPNB (where they were amongst the most common taxa) to the Late PPNB and
-Late Neolithic. However, this trend is not seen in most other crop
+Wild barley (\emph{Hordeum spontaneum}), its relatives, and pistachio
+(\emph{Pistacia atlantica}) show a contraction of their predicted ranges
+from the Pleistocene to the Holocene, concurrent with them being brought
+into cultivation \textbf{?@fig-palaeodist-barely-pistachio-levant}.
+These two species also see marked declines in the archaeobotanical
+record from the Early PPNA/Early PPNB (where they were amongst the most
+common taxa) to the Late PPNB and Late Neolithic
+\citep{ArranzOtaeguiRoe2023}. Conversely, the various wild wheat species
+native to West Asia show almost no response to Pleistocene/Holocene
+climate change, even within the Levant, and in the archaeobotanical
+record wheat displays the opposite trend to barley and pistachio --
+becoming gradually more abundant through the course of the Neolithic and
+dominant by its end. We hypothesise that climate-linked range shifts
+were a factor driving these changes in the apparent economic important
+of different crops: the decreasing availability of barley and pistachio
+in the terminal Pleistocene may have prompted cultivation as a strategy
+to retain access to them, though in the long run it made them less
+attractive as staple crops compared to the more resilient wheat.
+However, these dynamics are not seen in the majority of crop
 progenitors.
-
-Cereals / other progenitors (not all) reduced in range before
-cultivation? Reduction in range from BA to EH very evident in Hordeum,
-less so in \ldots{}
 
@@
 
-Is inconsistent with more ``macro'' trends such as the reduced range of
+Is consistent with more ``macro'' trends such as the reduced range of
 Hordeum and Pistacia correlating with its reduced abundance in the
@@
 
-\section{Discussion}\label{discussion}
+\section{Conclusion}\label{conclusion}
 
@@
   We present the first continuous, spatially explicit models of the
-  palaeodistributions of 56 plant species found regularly in association
+  palaeodistributions of N plant species found regularly in association
   with early agricultural archaeological sites in West Asia
@@
 
-\section{References}\label{references}
-
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
-
+\renewcommand\refname{References}
   \bibliography{references.bib}

### 2025-01-08 analysis/paper.qmd
Add crop progenitors discussion

@@
   - [x] Results
-  - [ ] Discussion
-    - [x] General trends
-    - [ ] Case studies
-    - [x] Comparison with archaeobot
+  - [x] Discussion
   - [ ] Conclusion
@@
   - [ ] tbl-predictors
-  - [ ] tbl-model-metrics
 - [ ] Outstanding TODOs
@@
 
-<!-- TODO: more citations? -->
-Ecological niche modelling (ENM) or species distribution modelling (SDM) is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors.
+Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
 Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
@@
 In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].
-<!-- TODO: mention that GBIF, CHELSA, SoilGrids, WorldClim, DEMs, etc. are also great? -->
 
@@
 
-# Model assessment
-
 ```{r niche-predict}
@@
 flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
-```
+
+# TODO write TIFs
+```
+
+# Model assessment
 
@@
   annotation_spatial(ne_land, fill = "white") +
-  geom_stars(data = select(palaeodist_sum, prediction.TRUE.), 
+  geom_stars(data = select(palaeodist_sum, present), 
              na.action = na.omit) +
@@
 
-* Most cereal and legume crops predicted to be Levantine
-* Cereals show diverging southern/northern ranges (legumes not so much)
-* Not rye (Anatolia), chickpea (N. Mesopotamia)
-
-```{r fig.height=10, fig-palaeodist-barely-pistachio-levant, eval=FALSE, cache=TRUE}
-#| fig-cap: Contraction of reconstructed range of of wild barleys and pistachio in the Levant
+```{r fig-palaeodist-chickpea-rye}
+#| fig-cap: Predicted palaeodistribution of wild chickpea and rye in the Early Holocene (11.7–8.2 ka)
+chickpea_rye <- c("Secale cereale", "Cicer reticulatum")
+palaeodist_chickpea_rye <- do.call(c, c(
+  filter(flora, taxon %in% chickpea_rye)$palaeodist,
+  list(along = list(taxon = filter(flora, taxon %in% chickpea_rye)$taxon))
+))
+
+ggplot() +
+  facet_wrap(vars(taxon)) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(palaeodist_chickpea_rye, prediction.TRUE.) |>
+               filter(period == "eh"), 
+             na.action = na.omit) +
+  annotation_spatial(ne_land, fill = NA) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(crs = w_asia_albers) +
+  labs(fill = "P(present)", x = NULL, y = NULL) +
+  theme(
+    strip.background = element_blank(),
+    strip.text = element_text(face = "italic")
+  )
+```
+
+Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
+Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
+One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-chickpea-rye).
+Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian [@fig-palaeodist-chickpea-rye].
+This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].
+
+```{r fig-palaeodist-locals}
+#| fig-cap: Predicted palaeodistribution of flax, pistachio and clubrush in the Early Holocene (11.7–8.2 ka)
+locals <- c("Linum bienne", "Pistacia atlantica", "Bolboschoenus maritimus")
+palaeodist_locals <- do.call(c, c(
+  filter(flora, taxon %in% locals)$palaeodist,
+  list(along = list(taxon = filter(flora, taxon %in% locals)$taxon))
+))
+
+ggplot() +
+  facet_wrap(vars(taxon), nrow = 2) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(palaeodist_locals, prediction.TRUE.) |>
+               filter(period == "eh"), 
+             na.action = na.omit) +
+  annotation_spatial(ne_land, fill = NA) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(crs = w_asia_albers) +
+  labs(fill = "P(present)", x = NULL, y = NULL) +
+  theme(
+    strip.background = element_blank(),
+    strip.text = element_text(face = "italic")
+  )
+```
+
+Flax (*Linum bienne*) is predicted to have had a highly concentrated distribution in Cyprus and along the Mediterranean coast of the southern Levant.
+This is consistent with its low ubiquity in archaeobotanical assemblages [@ArranzOtaeguiRoe2023], despite conventionally being considered a 'founder crop', and presumably implies that its domestication was similarly geographically constrained.
+It is the only unambiguous crop progenitor with such a restricted range, though pistachio (*Pistacia atlantica*) and clubrush (*Bolboschoenus maritimus*) are similarly constrained to the Mediterranean coast (and North Africa, in the case of pistachio).
+
+```{r fig.height=10, fig-palaeodist-barely-levant, eval=FALSE, cache=TRUE}
+#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
 palaeodist_hordeum <- do.call(c, c(
-  filter(flora, str_starts(taxon, "Hordeum") | taxon == "Pistacia atlantica")$palaeodist,
-  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum") | taxon == "Pistacia atlantica")$taxon))
+  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
+  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
 ))
@@
 
-Wild barley (*Hordeum spontaneum*), its relatives, and pistachio (*Pistacia atlantica*) show a contraction of their predicted ranges from the Pleistocene to the Holocene,
-concurrent with them being brought into cultivation [@fig-palaeodist-barely-pistachio-levant].
-These two species also see marked declines in the archaeobotanical record from the Early PPNA/Early PPNB (where they were amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
+Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
+concurrent with it being brought into cultivation [@fig-palaeodist-barely-levant].
+It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
+Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was cultivated in the Neolithic.
 Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end.
@@
 
-The wheat story.
-
-<!-- TODO: figure with flax, Pistacia, Bolboschoenus -->
-
-Flax has a very restricted distribution (consistent with low occurrence in founders paper?).
-As does Pistacia atlantica, Bolboschoenus maritimus
-
-<!-- TODO: figure with rye -->
-
-*Secale cereale* = an Anatolian boy
+```{r fig-palaeodist-bread}
+#| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
+bread <- c("Aegilops tauschii", "Aegilops speltoides", 
+           "Triticum turgidum dicoccum", "Triticum urartu")
+palaeodist_bread <- do.call(c, c(
+  filter(flora, taxon %in% bread)$palaeodist,
+  list(along = list(taxon = filter(flora, taxon %in% bread)$taxon))
+)) |>
+  st_apply(c("x", "y", "period"), sum, na.rm = FALSE) |>
+  mutate(present = factor(present))
+
+ggplot() +
+  facet_wrap(vars(fct_relabel(period, toupper))) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(palaeodist_bread, present), 
+             na.action = na.omit) +
+  annotation_spatial(ne_land, fill = NA) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_fill_batlow(reverse = TRUE, discrete = TRUE) +
+  coord_sf(crs = w_asia_albers) +
+  labs(fill = "N present", x = NULL, y = NULL) +
+  theme(
+    strip.background = element_blank(),
+    strip.text.y = element_text(face = "italic")
+  )
+```
+
+Bread wheat (*Triticum aestivum*), the most common wheat cultivar today, has a complex ancestry that involves two recent hybridisation events [@LevyFeldman2022]:
+most recently between emmer (*Triticum turgidum dicoccum*) and a goatgrass (*Aegilops tauschii*) c.
+9 ka, and before that, in emmer, between wild red einkorn (*T.
+urartu*) and another goatgrass (*Aegilops speltoides*).
+Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to eachother in the time frame of domestication is a narrow area around the Orontes river in the northern Levant [@fig-palaeodist-bread].
+This area is c.
+200 km from the only two sites in our archaeobotanical database where either both species of wheat or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
+Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is likely to occur in this vicinity.
 
@@
 
+Is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
 <!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
-Is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
 

### 2025-01-08 analysis/paper.tex
Add crop progenitors discussion

@@
     Results
-  \item[$\square$]
+  \item[$\boxtimes$]
     Discussion
-
-    \begin{itemize}
-    \tightlist
-    \item[$\boxtimes$]
-      General trends
-    \item[$\square$]
-      Case studies
-    \item[$\boxtimes$]
-      Comparison with archaeobot
-    \end{itemize}
   \item[$\square$]
@@
     tbl-predictors
-  \item[$\square$]
-    tbl-model-metrics
   \end{itemize}
@@
 
-Ecological niche modelling (ENM) or species distribution modelling (SDM)
-is widely used by ecologists to predict the geographic range of a
-species based on a set of environmental predictors. Essentially, it
-involves combining records of where an organism has been observed with
-environmental data (climate, topography, etc.) for those locations to
-model the range of environmental values at which that species -- its
-environmental niche. This model can then be used to predict the range of
-the organism in question either in the same or a different environment.
-\citet{CITE} suggests reserving the term `species distribution
-modelling' for when the method is used to recover the verifiable range
-of a species in a real and existing environment, and using `ecological
-niche modelling' as the broader term covering hypothetical or predictive
-applications -- a convention we follow here when referring to predictive
-or `hindcast' models of past ranges. Within this overarching framework,
-ecological niche modelling encompasses a wide range of applications and
-a variety of potential environmental predictors, modelling approaches,
-and methodologies, which we will not attempt to review here.
+Ecological niche modelling or species distribution modelling is widely
+used by ecologists to predict the geographic range of a species based on
+a set of environmental predictors \citep{FranklinMiller2009}.
+Essentially, it involves combining records of where an organism has been
+observed with environmental data (climate, topography, etc.) for those
+locations to model the range of environmental values at which that
+species -- its environmental niche. This model can then be used to
+predict the range of the organism in question either in the same or a
+different environment. \citet{CITE} suggests reserving the term `species
+distribution modelling' for when the method is used to recover the
+verifiable range of a species in a real and existing environment, and
+using `ecological niche modelling' as the broader term covering
+hypothetical or predictive applications -- a convention we follow here
+when referring to predictive or `hindcast' models of past ranges. Within
+this overarching framework, ecological niche modelling encompasses a
+wide range of applications and a variety of potential environmental
+predictors, modelling approaches, and methodologies, which we will not
+attempt to review here.
 
@@
 presences in archaeobotanical assemblages successfully predicted.
-
-\section{Discussion}\label{discussion}
 
@@
 
+\section{Discussion}\label{discussion}
+
 \subsection{Reduction in range sizes over the Pleistocene/Holocene
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Most cereal and legume crops predicted to be Levantine
-\item
-  Cereals show diverging southern/northern ranges (legumes not so much)
-\item
-  Not rye (Anatolia), chickpea (N. Mesopotamia)
-\end{itemize}
-
-Wild barley (\emph{Hordeum spontaneum}), its relatives, and pistachio
-(\emph{Pistacia atlantica}) show a contraction of their predicted ranges
-from the Pleistocene to the Holocene, concurrent with them being brought
-into cultivation \textbf{?@fig-palaeodist-barely-pistachio-levant}.
-These two species also see marked declines in the archaeobotanical
-record from the Early PPNA/Early PPNB (where they were amongst the most
-common taxa) to the Late PPNB and Late Neolithic
-\citep{ArranzOtaeguiRoe2023}. Conversely, the various wild wheat species
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-chickpea-rye-1.pdf}}
+
+}
+
+\caption{\label{fig-palaeodist-chickpea-rye}Predicted palaeodistribution
+of wild chickpea and rye in the Early Holocene (11.7--8.2 ka)}
+
+\end{figure}%
+
+Almost all the cereal and legume crop progenitors we modelled are
+predicted to have only been found in the Levant during the terminal
+Pleistocene and Early Holocene (see appendix). Part of this may be do
+with the fact that both our initial flora and training occurrence
+dataset have a strong bias towards the southern Levant, but strikingly
+the modelled \emph{current} ranges of these plants do tend to include
+Anatolia and the Zagros, so this cannot be the only factor. One notable
+exception is the wild ancestor of chickpea (\emph{Cicer reticulatum}),
+which is predicted to have a distribution centered on Northern
+Mesopotamia but encompassing much of the the `hilly flanks' (except the
+southern Levant, \textbf{?@fig-chickpea-rye}). Another is rye
+(\emph{Secale cereale}), which is inferred to be primarily Western
+Anatolian Figure~\ref{fig-palaeodist-chickpea-rye}. This is perhaps
+relevant to rye's unusual domestication history, as a crop of West Asian
+origin that was apparently not first cultivated until much later than
+the `founder crops', in Europe \citep{SchreiberEtAl2021}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-locals-1.pdf}}
+
+}
+
+\caption{\label{fig-palaeodist-locals}Predicted palaeodistribution of
+flax, pistachio and clubrush in the Early Holocene (11.7--8.2 ka)}
+
+\end{figure}%
+
+Flax (\emph{Linum bienne}) is predicted to have had a highly
+concentrated distribution in Cyprus and along the Mediterranean coast of
+the southern Levant. This is consistent with its low ubiquity in
+archaeobotanical assemblages \citep{ArranzOtaeguiRoe2023}, despite
+conventionally being considered a `founder crop', and presumably implies
+that its domestication was similarly geographically constrained. It is
+the only unambiguous crop progenitor with such a restricted range,
+though pistachio (\emph{Pistacia atlantica}) and clubrush
+(\emph{Bolboschoenus maritimus}) are similarly constrained to the
+Mediterranean coast (and North Africa, in the case of pistachio).
+
+Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
+of their predicted ranges from the Pleistocene to the Holocene,
+concurrent with it being brought into cultivation
+\textbf{?@fig-palaeodist-barely-levant}. It also sees a marked decline
+in the archaeobotanical record from the Early PPNA/Early PPNB (where it
+was amongst the most common taxa) to the Late PPNB and Late Neolithic
+\citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia atlantica})
+shows similar trends, but it is less certain that this species was
+cultivated in the Neolithic. Conversely, the various wild wheat species
 native to West Asia show almost no response to Pleistocene/Holocene
@@
 
-The wheat story.
-
-Flax has a very restricted distribution (consistent with low occurrence
-in founders paper?). As does Pistacia atlantica, Bolboschoenus maritimus
-
-\emph{Secale cereale} = an Anatolian boy
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-bread-1.pdf}}
+
+}
+
+\caption{\label{fig-palaeodist-bread}Combined predicted
+palaeodistributions of bread wheat progenitors}
+
+\end{figure}%
+
+Bread wheat (\emph{Triticum aestivum}), the most common wheat cultivar
+today, has a complex ancestry that involves two recent hybridisation
+events \citep{LevyFeldman2022}: most recently between emmer
+(\emph{Triticum turgidum dicoccum}) and a goatgrass (\emph{Aegilops
+tauschii}) c.~9 ka, and before that, in emmer, between wild red einkorn
+(\emph{T. urartu}) and another goatgrass (\emph{Aegilops speltoides}).
+Although there is fairly wide overlap between these species today,
+according to our modelling the only place all four (or even just three)
+are predicted to have coincided or been found in close proximity to
+eachother in the time frame of domestication is a narrow area around the
+Orontes river in the northern Levant Figure~\ref{fig-palaeodist-bread}.
+This area is c.~200 km from the only two sites in our archaeobotanical
+database where either both species of wheat or both species of wheat and
+the goatgrasses are found together: Tell Abu Hureyra and El Kowm II
+\citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to Benjamin Nowak
+  for pointing this out to us.} Taken together this suggests an origin
+of domesticated bread wheat, otherwise only loosely geographically
+constrained to the Levant--Upper Euphrates corridor
+\citep{LevyFeldman2022}, is likely to occur in this vicinity.
 

### 2025-01-09 analysis/appendix.qmd
Correct and refactor archaeo_pred

@@
     toc: true
+    # embed-resources: true
 execute:
@@
 flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))
-archaeo_flora <- read_tsv(
-  derived_data("archaeo_flora_predictions.tsv"),
-  col_types = cols(
-    source = col_character(),
-    source_id = col_character(),
-    source_site_name = col_character(),
-    site_name = col_character(),
-    latitude = col_double(),
-    longitude = col_double(),
-    phase = col_character(),
-    phase_description = col_character(),
-    phase_code = col_character(),
-    age_start = col_double(),
-    age_end = col_double(),
-    taxon_source = col_character(),
-    n = col_double(),
-    prop = col_double(),
-    reference = col_character(),
-    taxon = col_character(),
-    taxon_group = col_character(),
-    genus = col_character(),
-    family = col_character(),
-    category = col_character(),
-    founder_crop = col_character(),
-    edibility = col_character(),
-    grass_type = col_character(),
-    legume_type = col_character(),
-    age_mid = col_double(),
-    period = col_character(),
-    climate_period = col_character(),
-    pred_present = col_logical(),
-    pred_absent = col_logical()
-  )
-)
+archaeo_pred <- readRDS(derived_data("archaeo_pred.Rdata"))
+flora <- left_join(flora, archaeo_pred)
 
@@
   flora, 
-  function(taxon, taxon_detail, occ, niche, palaeodist, ...) {
-    archaeo <- archaeo_flora[archaeo_flora$taxon == taxon,]
+  function(taxon, taxon_aliases, archaeo_pred, occ, niche, palaeodist, ...) {
     knitr::knit_child(
@@
     )
-  },
-  archaeo_flora = archaeo_flora, 
+  }
 )

### 2025-01-09 analysis/appendix_section.qmd
Correct and refactor archaeo_pred

@@
 
-```{r taxon-detail, eval=!is.na(taxon_detail), results = "asis"}
-cat("Including ", knitr::combine_words(taxon_detail), ".", sep = "")
+```{r taxon-detail, eval=(length(taxon_aliases) > 0), results = "asis"}
+cat("Including ", knitr::combine_words(taxon_aliases), ".", sep = "")
 ```
@@
 ```{r fig-palaeodist, fig.height=16, fig.cap=paste0("Predicted palaeodistributions of *", taxon, "*")}
+if (length(archaeo_pred) > 0) {
+  archaeo_geom <- geom_sf(
+    aes(shape = .pred_class),
+    data = archaeo_pred |>
+      mutate(period = climate_period) |>
+      distinct(longitude, latitude, period, .keep_all = TRUE) |>
+      st_as_sf(coords = c("longitude", "latitude"), crs = latlong)
+  )
+} else {
+  archaeo_geom <- NULL
+}
+
 ggplot() +
@@
   ) +
-  geom_sf(
-    aes(shape = pred_present),
-    data = archaeo |>
-      mutate(
-        period = climate_period, 
-        pred_present = replace_na(pred_present, FALSE)
-      ) |>
-      distinct(longitude, latitude, .keep_all = TRUE) |>
-      st_as_sf(coords = c("longitude", "latitude"), crs = latlong)
-  ) +
+  archaeo_geom +
   annotation_spatial(ne_land, fill = NA) +
@@
 
-```{r tbl-archaeo}
-# TODO shouldn't be necessary
-if (nrow(archaeo) > 0) {
-  archaeo |>
-    drop_na(climate_period) |>
+```{r tbl-archaeo, eval=(length(archaeo_pred) > 0)}
+archaeo_pred |>
+  mutate(taxon = taxon) |>
   summarise(
@@
     age_end = min(age_end),
-      pred_present = all(pred_present),
+    pred_present = all(as.logical(.pred_class)),
     .by = c(source, climate_period, site_name)
@@
          pred_present) |>
-    gt(caption = paste0("Archaeological occurrences of *", taxon, "*")) |>
+  gt(caption = md(paste0("Archaeological occurrences of *", taxon, "*"))) |>
   cols_label(site_name = "Site", n_assemb = "N assemblages",
@@
   sub_missing()
-}
 ```

### 2025-01-09 analysis/paper.qmd
Correct and refactor archaeo_pred

@@
   This paper presents the first continuous, spatially-explicit reconstructions 
-  of the palaeodistributions of 81<!-- TODO update --> plant species found regularly in association 
+  of the palaeodistributions of 82<!-- TODO update --> plant species found regularly in association 
   with early agricultural archaeological sites in West Asia, including the 
@@
 
+<!--
+
 * [x] Stop modelling
@@
   - [x] Discussion
-  - [ ] Conclusion
+  - [x] Conclusion
 - [ ] Figures
@@
   - [x] fig-occ
-  - [ ] tbl-occ-count (needs pdf cleanup)
+  - [x] tbl-results-summary
   - [ ] tbl-predictors
+  - [x] palaeodist figs
 - [ ] Outstanding TODOs
@@
 - [ ] Copyedit
+  - [ ] Replace "see below" etc. with section links
 - [x] Appendix with all hindcast predictions (HTML?)
@@
 - [ ] Final proofread
+
+-->
 
@@
 West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@cite].
-Even before the end of the Ice Age, these species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
+These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
 As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.
@@
 @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
-This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
+This overlaps with what archaeologists usually call generically 'predictive modelling' [@VerhagenWhitley2020]—or more precisely 'site distribution modelling'—which is essentially the same approach as (and often borrows methodologies from) ecological niche modelling but applied to the occurrence of archaeological sites.
 Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
@@
 It has been used to model the availability of fauna exploited by humans at wider scales [e.g.
-@deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic site of XX<!-- TODO: Kortik?
---> [@CollinsEtAl2018].
-Modelling the spread of crops has been another significant archaeological application [@CremaEtAl].
+@deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic sites on the Konya Plain [@CollinsEtAl2018].
+Modelling the spread of crops has been another significant archaeological application [e.g. @CremaEtAl], though not as yet applied to West Asia.
 
@@
 Like @FranklinEtAl2015, we view the hindcasting approach as more promising, because training datasets for both occurrences and environment are far more readily available, complete and reliable for the present than the past.
-There is some scepticism in the ecological niche modelling literature about the ability of such models to make accurate predictions in unknown environments (like the past) [@CITES?], 
+There is some scepticism in the ecological niche modelling literature about the ability of such models to make accurate predictions in unknown environments [like the past, @FranklinEtAl2015], 
 but here the hindcasting approach also presents an opportunity:
@@
 
-```{r data-archaeo-flora}
-# Flora present at archaeological sites
-archaeo_flora <- read_tsv(
+```{r data-flora}
+# Start with flora present at archaeological sites
+if (!file_exists(derived_data("flora.Rdata"))) {
+  flora <- read_tsv(
     raw_data("swasia_neolithic_flora.tsv"),
@@
     # Filter to Epipal/Neo
-  filter(period %in% epipal | period %in% ppn)
-
-n_archaeo_assemb <- nrow(distinct(archaeo_flora, site_name, phase_code))
-n_archaeo_site <- nrow(distinct(archaeo_flora, site_name))
-```
-
-```{r data-archaeo-sites}
-archaeo_sites <- summarise(
-  archaeo_flora, 
-  latitude = first(latitude), 
-  longitude = first(longitude), 
-  .by = c(site_name, period)
-) |>
-  st_as_sf(coords = c("longitude", "latitude"), crs = latlong, remove = FALSE)
-```
-
-```{r fig-region}
-#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
-ggplot() + 
-  annotation_spatial(ne_land, fill = "#ffffff") +
-  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
-  layer_spatial(w_asia, fill = NA) +
-  layer_spatial(archaeo_sites, size = 0.5) +
-  scale_linewidth_identity() +
-  coord_sf(crs = w_asia_albers)
-```
-
-```{r data-flora}
-# TODO: nest archaeo flora in flora to make it easier to keep track of taxon transformations
-if (!file_exists(derived_data("flora.tsv"))) {
-  message("Regenerating flora...")
-  # Subset of flora to be modelled
-  archaeo_flora |>
-    group_by(taxon) |>
-    summarise(
-      n_present = length(unique(phase_code)),
-      p_present = n_present / n_archaeo_assemb,
-      .groups = "drop"
-    ) |>
-    filter(n_present > min_n_archaeo) |>
-    # Exclude taxa not identified to species
-    filter(!str_ends(taxon, "sp.")) ->
-    flora
-
-  # Standardise taxonomy
-  flora <- flora |>
+    filter(period %in% epipal | period %in% ppn) |>
+    # Group by taxon
+    drop_na(taxon) |>
+    nest(.by = c(taxon), .key = "archaeo") |>
     # Normalise species names and match to the GBIF Backbone Taxonomy
     mutate(
-      taxon = case_match(taxon,
+      # Spot corrections
+      taxon = case_match(
+        taxon,
         "Vicia narbonense" ~ "Vicia narbonensis", # N=3
@@
         "Stipa capensis" ~ "Stipa dregeana", # N=2
-                         .default = taxon),
+        .default = taxon
+      ),
       taxon_original = taxon,
-      taxon = map_chr(taxon, gbif_accepted_name)
-    ) |>
+      taxon = map_chr(taxon, gbif_accepted_name),
+      .before = archaeo
+    ) |>
+    # Drop unmatched (mostly not identified to species)
+    drop_na(taxon) |>
     # Recode domesticates as their wild progenitors.
@@
     ) |>
+    # Regroup by canonical taxon
+    summarise(
+      taxon_aliases = list(taxon_original[taxon_original != taxon]),
+      archaeo = list(bind_rows(archaeo)),
+      .by = taxon
+    ) |>
+    mutate(
+      n_archaeo = map_int(archaeo, nrow)
+    ) |>
+    filter(n_archaeo > min_n_archaeo) |>
     # Add wild progenitors of wheats
@@
       )
-    ) |>
-    # Deduplicate by canonical taxon
+    )
+
+  saveRDS(flora, derived_data("flora.Rdata"))
+}
+flora <- readRDS(derived_data("flora.Rdata"))
+```
+
+```{r data-archaeo-sites}
+archaeo_sites <- flora |>
+  unnest(archaeo) |>
   summarise(
-      taxon_detail = paste0("*", taxon_original[taxon_original != taxon], "*", collapse = ", "),
-      taxon_detail = na_if(taxon_detail, "**"),
-      n_present = max(n_present),
-      p_present = max(p_present),
-      .by = taxon
-    )
-
-  write_tsv(flora, derived_data("flora.tsv"))
-}
-
-flora <- read_tsv(
-  derived_data("flora.tsv"),
-  col_types = cols(
-    taxon = col_character(),
-    taxon_detail = col_character(),
-    n_present = col_double(),
-    p_present = col_double()
-  )
-)
+    latitude = first(latitude), 
+    longitude = first(longitude), 
+    .by = c(site_name, period)
+  ) |>
+  st_as_sf(coords = c("longitude", "latitude"), crs = latlong)
+```
+
+```{r fig-region}
+#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
+ggplot() + 
+  annotation_spatial(ne_land, fill = "#ffffff") +
+  annotation_spatial(ne_rivers, aes(linewidth = strokeweig), colour = "lightblue") +
+  layer_spatial(w_asia, fill = NA) +
+  layer_spatial(archaeo_sites, size = 0.5) +
+  scale_linewidth_identity() +
+  coord_sf(crs = w_asia_albers)
 ```
@@
 The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
-Our starting point was a list of `r nrow(flora)` taxa (@tbl-taxa) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
-This was based on dataset of `r n_archaeo_assemb` assemblages collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
-We did not attempt to distinguish between the source of the remains;
+Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
+This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
+We did not attempt to distinguish between the source of the remains [cf. @WallaceEtAl2018];
 archaeobotanical assemblages are subject to a variety of preservational and recovery biases, so by no means were all the species on our list consumed or even deliberately collected by people.
@@
 Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
-So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-taxa).
+So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
 Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.
@@
 
+<!-- TODO generate GBIF occurrence citation:
+https://docs.ropensci.org/rgbif/articles/getting_occurrence_data.html -->
+
 ```{r data-n-occ} 
-if (!file_exists(derived_data("flora_n_occ.tsv"))) {
+if (!file_exists(derived_data("flora_n_occ.Rdata"))) {
   message("Reacquiring GBIF occurrence counts...")
@@
     )
-  write_tsv(flora, derived_data("flora_n_occ.tsv"))
+  saveRDS(flora, derived_data("flora_n_occ.Rdata"))
 }
-
-flora <- read_tsv(
-  derived_data("flora_n_occ.tsv"),
-  col_types = cols(
-    taxon = col_character(),
-    taxon_detail = col_character(),
-    n_present = col_double(),
-    p_present = col_double(),
-    n_occ = col_integer()
-  )
-)
+flora <- readRDS(derived_data("flora_n_occ.Rdata"))
 
@@
 # - https://data-blog.gbif.org/post/gbif-filtering-guide/
-if (!file_exists(derived_data("flora_occ.tsv"))) {
+if (!file_exists(derived_data("flora_occ.Rdata"))) {
   message("Reacquiring GBIF occurrences...")
@@
 
-  write_tsv(unnest(flora, cols = c(occ)), derived_data("flora_occ.tsv"))
+  saveRDS(flora, derived_data("flora_occ.Rdata"))
 }
 
-flora <- read_tsv(
-  derived_data("flora_occ.tsv"),
-  col_types = cols(
-    taxon = col_character(),
-    taxon_detail = col_character(),
-    n_present = col_double(),
-    p_present = col_double(),
-    n_occ = col_integer(),
-    gbif_key = col_double(),
-    scientific_name = col_character(),
-    genus = col_character(),
-    species = col_character(),
-    longitude = col_double(),
-    latitude = col_double()
-  )
-) |>
-  nest(occ = gbif_key:latitude)
+flora <- readRDS(derived_data("flora_occ.Rdata"))
 ```
@@
 GBIF was cleaned to removed fossil occurrences, recorded absences, and records with missing, imprecise (>1 km uncertainty), or invalid coordinates.
-Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora, n_occ > min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
+Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
 We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what accounted for such a disproportionately high number of records.
@@
 
-## Predictor data
-
-```{r tbl-predictors}
-# TODO
-```
-
-We modelled the occurrence of species as a function of <!-- TODO -->X spatial predictor variables (@tbl-predictors).
+## Predictor data {#sec-predictors}
+
+We modelled the occurrence of species as a function of 24 geospatial predictor variables.
 These included:
@@
 
-Predictor data was transformed to common equal-area projection and resolution of `r pred_res / 1000` km.
-
-```{r tbl-climate-periods}
-#| fig-cap: Palaeoclimatic periods used for hindcasting, after @BrownEtAl2018
-# TODO: integrate into text
-climate_periods |>
-  mutate(
-    code = toupper(code),
-    age_start = start_bp / 1000,
-    age_end = end_bp / 1000
-  ) |>
-  select(period, code, age_start, age_end) |>
-  gt() |>
-  cols_label(period = "Period", age_start = "Age") |>
-  cols_merge(
-    c(period, code),
-    pattern = "{1} ({2})"
-  ) |>
-  cols_merge_range(age_start, age_end) |>
-  cols_units(age_start = "ka") |>
-  sub_missing()
-```
-
-For hindcasting, we used reconstructed bioclimatic data for `r english(nrow(climate_periods) - 1)` key periods (@tbl-climate-periods) generated from downscaled paleoclimate simulations from the HadCM3 general circulation model [@BrownEtAl2018].
+For hindcasting, we used reconstructed bioclimatic data for three key climatologies generated from downscaled paleoclimate simulations from the HadCM3 general circulation model [@FordhamEtAl2017; @BrownEtAl2018]:
+the Bølling–Allerød (c. 14.7–12.9 ka), the Younger Dryas (c. 12.9–11.7 ka), and the Early Holocene (11.7–8.3 ka).
 Terrain and soil predictors were held constant, since reconstructions of these variables in the past are not available at sufficient scale.
-It is not likely that either macroscale topography or soil characteristics have altered significantly over the period of time considered here, so we assume that this does not degrade model performance, and may in fact benefit it by providing 'anchoring' predictors that are independent of climate change.
+It is unlikely that either macroscale topography or soil characteristics have altered significantly over the period of time considered here, so we assume that this does not degrade model performance, and may in fact benefit it by providing 'anchoring' predictors that are independent of climate change.
+
+For training, test predictions, and archaeological predictions predictor data was left in its native projection and resolution.
+For hindcast palaeodistributions, it was transformed to common equal-area projection and resolution of `r pred_res / 1000` km.
 
@@
     occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
-    # occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
     occ_split = map(occ, \(x) initial_split(x, strata = present)),
-    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
   )
@@
 
-For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our X<!--TODO--> predictor variables (@tbl-predictors).
+For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our 24 predictor variables (@sec-predictors).
 We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
@@
 
-We trained Random Forest models for `r nrow(flora)` taxa using contemporary occurrence data from GBIF, a random sample of background points, and the predictor variables described in @tbl-predictors.
+We trained Random Forest models for `r nrow(flora)` taxa using contemporary occurrence data from GBIF, a random sample of background points, and the predictor variables described in @sec-predictors.
 Substituting the "current" climate predictors for those derived from palaeoclimatic simulations [@BrownEtAl2018], we could then generate hindcast predictions for reconstructed past environments in `r english(nrow(climate_periods))` key climate periods – a total of `r nrow(flora) * nrow(climate_periods)` modelled palaeodistributions.
@@
 
-We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition.
+We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition [@tbl-results-summary].
 Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_summary, .metric == "accuracy")$min)` and `r percent(filter(niche_test_summary, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "accuracy")$mean)`.
 Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_summary, .metric == "sens")$min)` and `r percent(filter(niche_test_summary, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "sens")$mean)`.
-The area under the models' receiver operating characteristic curves (ROC-AUC) was in the range of `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`
+The area under the models' receiver operating characteristic curves (ROC-AUC) was on average `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`
 Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
@@
 
-```{r niche-archaeo}
-# TODO: need to deal with taxon variants / GBIF normalisation
-# TODO: factor out
+```{r niche-predict-archaeo}
+# Extract predictors and make predictions for archaeobotanical occurrences
 # TODO: what about negatives?
-archaeo_flora_predictions <-  archaeo_flora |>
-  filter(taxon %in% flora$taxon) |>
-  st_as_sf(coords = c("longitude", "latitude"), crs = latlong, remove = FALSE) |>
+if (!file.exists(derived_data("archaeo_pred.Rdata"))) {
+  archaeo_pred <- flora |>
+    unnest(archaeo) |>
+    nest(.by = c(climate_period), .key = "archaeo") |>
     mutate(
-    prediction = map(taxon, \(x) filter(flora, taxon == x)$palaeodist[[1]]),
-    prediction = map2(prediction, geometry, function(pred, geom) {
-      sf_point <- st_transform(st_sf(st_sfc(geom, crs = latlong)), w_asia_albers)
-      st_extract(pred, sf_point)
+      archaeo = map(archaeo, st_as_sf, coords = c("longitude", "latitude"),
+                    crs = latlong, remove = FALSE),
+      archaeo = map2(archaeo, climate_period, function(x, period) {
+        bind_extract(x, slice(climate, "period", as.character(period)))
       }),
-    prediction = map(prediction, as_tibble),
-    prediction = map2(prediction, climate_period, function(pred, per) {
-      filter(pred, as.character(period) == per)
-    }),
-    prediction = map(prediction, select,
-                     pred_present = present, pred_absent = absent)
-  ) |>
-  unnest(c(prediction)) |>
-  as_tibble() |>
-  select(-geometry)
-
-write_tsv(archaeo_flora_predictions, derived_data("archaeo_flora_predictions.tsv"))
-
-p_archaeo_pred_correct <- sum(archaeo_flora_predictions$pred_present, na.rm = TRUE) / length(!is.na(archaeo_flora_predictions$pred_present))
-
-archaeo_flora_pred_summary <- archaeo_flora_predictions |>
-  as_tibble() |>
+      archaeo = map(archaeo, bind_extract, terrain),
+      archaeo = map(archaeo, bind_extract, soil)
+    ) |>
+    unnest(archaeo) |>
+    nest(.by = any_of(colnames(flora)), .key = "archaeo") |> # Restore original nesting
+    mutate(
+      archaeo_pred = map2(niche, archaeo, function(x, y) {
+        # TODO: shouldn't be necessary - remove from workflow
+        y <- mutate(y, gbif_key = NA, scientific_name = NA, species = NA)
+        augment(extract_workflow(x), new_data = y)
+      })
+    ) |>
+    select(taxon, archaeo_pred)
+
+  saveRDS(archaeo_pred, derived_data("archaeo_pred.Rdata"))
+}
+archaeo_pred <- readRDS(derived_data("archaeo_pred.Rdata"))
+
+n_archaeo_pred_correct <- sum(as.logical(unnest(archaeo_pred, archaeo_pred)$.pred_class))
+p_archaeo_pred_correct <- n_archaeo_pred_correct / nrow(unnest(archaeo_pred, archaeo_pred))
+
+archaeo_pred_summary <- archaeo_pred |>
+  unnest(archaeo_pred) |>
   summarise(
-    n_present = sum(!is.na(pred_present)),
-    n_predicted = sum(pred_present, na.rm = TRUE),
-    p_predicted = n_predicted / n_present,
-    .by = c(climate_period, site_name, phase_code, age_mid)
-  )
-```
-
-The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is significantly worse, with only `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted.
+    n_predicted = sum(as.logical(.pred_class)),
+    p_predicted = n_predicted / n(),
+    .by = c(taxon)
+  )
+
+flora <- left_join(flora, archaeo_pred)
+```
+
+The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is worse, with `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted.
+Model sensitivity (proportion of corrected predicted presenses) in relation to the archaeological data is on average `r round(mean(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)`±`r round(sd(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)` [@tbl-results-summary].
+A full assessment of the hindcasting performance of the individual models can be found in the appendix.
 <!-- TODO: negative -->
@@
 
-```{r tbl-taxa}
+```{r tbl-results-summary}
 #| tbl-cap: "Summary of species modelled"
-# TODO missing excluded species
-# TODO: redo with:
+add_taxon_alias_footnotes <- function(tab, taxon_aliases) {
+  rows <- which(map_lgl(taxon_aliases, \(x) length(x) > 0))
+  notes <- map_chr(taxon_aliases[rows], function(aliases) {
+    paste(
+      "Including",
+      knitr::combine_words(paste0("*", aliases, "*"))
+    )
+  })
+  for (i in 1:length(rows)) {
+    tab <- tab_footnote(tab, md(notes[i]), cells_body(c(taxon), rows = rows[i]))
+  }
+  tab
+}
+
+# TODO:
 # - alt. taxons as footnotes
 # - check boxes of which (climate?) periods observed in ?
-flora |>
-  # Summarise archaeo occurrences by period
-
+results_summary <- flora_all |>
+  left_join(flora, join_by(taxon, taxon_aliases, archaeo, n_archaeo)) |>
+  left_join(archaeo_pred_summary, join_by(taxon)) |>
+  mutate(n_occ = coalesce(n_occ.y, n_occ.x), .before = n_occ.x) |>
   # Extract model metrics
-  unnest(c(niche_test_metrics)) |>
+  unnest(c(niche_test_metrics), keep_empty = TRUE) |>
   select(-.estimator, -.config) |>
@@
   # Prepare table
-  select(taxon, n_present, n_occ, exclude, accuracy, roc_auc, sens) |>
-  arrange(taxon) |>
-  # Table
+  select(taxon, taxon_aliases, n_archaeo, n_occ, exclude, accuracy, roc_auc, 
+         sens, p_predicted) |>
+  arrange(taxon)
+
+results_summary |>
+  select(-taxon_aliases) |>
   gt() |>
@@
   cols_label(
-    n_present = "Archaeological",
-    n_occ = "Present (GBIF)",
+    n_archaeo = "Arch.",
+    n_occ = "Cur.",
+    accuracy = "Acc.",
     roc_auc = "ROC-AUC",
-    sens = "Sensitivity"
-  ) |>
-  tab_spanner("Occurrences", columns = c(n_present, n_occ)) |>
-  tab_spanner("Model", columns = accuracy:sens) |>
+    sens = "Sens.",
+    p_predicted = "Sens. (arch.)"
+  ) |>
+  tab_spanner("Occurrences", columns = c(n_archaeo, n_occ)) |>
+  tab_spanner("Model", columns = accuracy:p_predicted) |>
   cols_hide(exclude) |>
@@
   ) |>
+  add_taxon_alias_footnotes(results_summary$taxon_aliases) |>
   cols_align("right") |>
   cols_align("left", c(taxon)) |>
-  fmt_number(accuracy:sens) |>
+  fmt_number(accuracy:p_predicted) |>
   sub_missing() |>
@@
   tab_options(
-    footnotes.marks = "*",
+    # footnotes.marks = "*",
     table.font.size = 11,
@@
 
-```{r fig-pred-range-density}
+```{r fig-pred-range-density, warning=FALSE}
 #| fig-cap: Distribution of predicted species ranges by period. Dashed lines indicate the median range.
@@
 `r n_reduced_range` of `r nrow(flora)` species have reduced ranges; `r n_reduced_range_10` of more than 10% or more.
-Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further below), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable [@fig-range-density].
+Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further below), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
 The average range of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` during the Younger Dryas (i.e.
@@
 
-Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene (@fig-sum-palaeodist).
+Many taxa that occur (or are predicted to occur) across the 'hilly flanks' today—including most crop progenitors—are reconstructed to have had a significantly more restricted distribution in the terminal Pleistocene/Early Holocene (@fig-palaeodist-sum).
 These include *Ficus carica* (fig);
@@
 
-Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @] and as a centre of origin of agriculture [@].
-Within the Levant, retreat from the Badia.
+<!-- TODO expand/rewrite -->
+Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @cite] and as a centre of origin of agriculture [@cite].
+Within the Levant, many species show moderate reductions in ranges over the Pleistocene/Holocence boundary, retreating from the Badia/transjordan region [e.g. @fig-palaeodist-barely-levant].
 
 Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture.
-Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyHopfWeiss2012; @KabukcuEtAl2021; @UlasEtAl2024].
+Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyEtAl2012; @KabukcuEtAl2021; @UlasEtAl2024].
 Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [e.g. @HaasEtAl2019].
@@
 Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
-One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-chickpea-rye).
+One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye).
 Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian [@fig-palaeodist-chickpea-rye].
@@
 
-```{r fig.height=10, fig-palaeodist-barely-levant, eval=FALSE, cache=TRUE}
+```{r fig.height=10, fig-palaeodist-barely-levant, cache=TRUE}
 #| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
+# TODO: why so slow?
 palaeodist_hordeum <- do.call(c, c(
@@
 urartu*) and another goatgrass (*Aegilops speltoides*).
-Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to eachother in the time frame of domestication is a narrow area around the Orontes river in the northern Levant [@fig-palaeodist-bread].
+Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to eachother in the time frame of domestication is a narrow area around the Orontes river in the northern Levant (@fig-palaeodist-bread).
 This area is c.
@@
 
-Is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
+It is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
 <!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
@@
 
-* We present the first continuous, spatially explicit models of the palaeodistributions of N plant species found regularly in association with early agricultural archaeological sites in West Asia
-  * A new line of evidence on archaeoecology
-  * Complementary to archaeobot/pollen/etc. because it is independent from it
-  * All models are wrong... but it's easier to see how these are wrong than lines on maps
-* Modelling at scale using random forest, modern occurrences, and hindcasting represents a significant advance in pENM methodology
-  * Relies on recent open ecological and climatic datasets
-  * ...open archaeological datasets still lacking!
-* First (? - check Yaworsky) attempt to verify hindcast models with archaeo. compositional data
-  * Results not to promising, but this doesn't mean the models are useless!
+We present the first continuous, spatially explicit models of the palaeodistributions of `r nrow(flora)` plant species found regularly in association with Late Epipalaeolithic and Early Neolithic sites in West Asia.
+This deductive approach—modelling the niche of a species based on its occurrence in relation to environmental factors today, and using this together with palaeoclimate simulations to infer its past distribution—represents a new line of evidence on the archaeoecology of the world's first agricultural societies.
+It provides a complementary picture to that gleaned from environmental archaeology and climatological archives because it is independent of the taphonomic, anthropic, and recovery-related processes that affect these records.
+
+"All models are wrong" [@cite], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
+
+* Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
 * Discrepencies suggests several areas for further research and methodological development
 
+Modelling a large number of species using machine learning, the substantial occurrence datasets available for the present day, and a hindcasting approach to past distributions also represents a significant advance in the methodology of palaeoecological niche modelling.
+This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@@KargerEtAl2017; @BrownEtAl2018].
+Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
+Though we have benefitted from the relatively good state of publication of archaeobotanical compilations in our region of study [@ADEMNES; @COMPAG; @ORIGINS], 
+further development of open, comprehensive and up-to-date 'backbone' datasets on site locations and chronologies is needed to advance archaeoecological modelling to the same level.
+
 # References {.appendix}

### 2025-01-09 analysis/paper.tex
Correct and refactor archaeo_pred

@@
 This paper presents the first continuous, spatially-explicit
-reconstructions of the palaeodistributions of 81 plant species found
+reconstructions of the palaeodistributions of 82 plant species found
 regularly in association with early agricultural archaeological sites in
@@
 
-\begin{itemize}
-\tightlist
-\item[$\boxtimes$]
-  Stop modelling
-\item[$\square$]
-  First draft:
-
-  \begin{itemize}
-  \tightlist
-  \item[$\boxtimes$]
-    Introduction
-  \item[$\square$]
-    Background
-
-    \begin{itemize}
-    \tightlist
-    \item[$\square$]
-      Biogeography
-    \item[$\boxtimes$]
-      ENM
-    \end{itemize}
-  \item[$\boxtimes$]
-    Methods \& Materials
-  \item[$\boxtimes$]
-    Results
-  \item[$\boxtimes$]
-    Discussion
-  \item[$\square$]
-    Conclusion
-  \end{itemize}
-\item[$\square$]
-  Figures
-
-  \begin{itemize}
-  \tightlist
-  \item[$\boxtimes$]
-    fig-region
-  \item[$\boxtimes$]
-    fig-occ
-  \item[$\square$]
-    tbl-occ-count (needs pdf cleanup)
-  \item[$\square$]
-    tbl-predictors
-  \end{itemize}
-\item[$\square$]
-  Outstanding TODOs
-\item[$\square$]
-  Add missing references
-\item[$\square$]
-  Copyedit
-\item[$\boxtimes$]
-  Appendix with all hindcast predictions (HTML?)
-\item[$\square$]
-  Clean up package code
-\item[$\square$]
-  Final proofread
-\end{itemize}
-
 \section{Introduction}\label{introduction}
@@
 relatives of wheat, barley, peas, lentils, and other crops of global
-importance \citep{cite}. Even before the end of the Ice Age, these
-species supported uniquely dense and complex Late Epipalaeolithic
-(15--11.7 ka) societies based on intensive foraging \citep{cite} and
-eventually pre-domestication cultivation \citep{cite}. As they were
-brought under domestication in the Pre-Pottery Neolithic period
-(11.7--8.5 ka), the world's first agriculture was shaped by the
-ecosystems from which it emerged and was embedded in.
+importance \citep{cite}. These species supported uniquely dense and
+complex Late Epipalaeolithic (15--11.7 ka) societies based on intensive
+foraging \citep{cite} and eventually pre-domestication cultivation
+\citep{cite}. As they were brought under domestication in the
+Pre-Pottery Neolithic period (11.7--8.5 ka), the world's first
+agriculture was shaped by the ecosystems from which it emerged and was
+embedded in.
 
@@
 usually call generically `predictive modelling'
-\citep{VerhagenWhitley2020}---more precisely `site distribution
+\citep{VerhagenWhitley2020}---or more precisely `site distribution
 modelling'---which is essentially the same approach as (and often
@@
 Asian context, of foraged plant resources in the landscape around the
-Neolithic site of XX \citep{CollinsEtAl2018}. Modelling the spread of
-crops has been another significant archaeological application
-\citep{CremaEtAl}.
+Neolithic sites on the Konya Plain \citep{CollinsEtAl2018}. Modelling
+the spread of crops has been another significant archaeological
+application \citep[e.g.][]{CremaEtAl}, though not as yet applied to West
+Asia.
 
@@
 niche modelling literature about the ability of such models to make
-accurate predictions in unknown environments (like the past)
-\citep{CITES}, but here the hindcasting approach also presents an
-opportunity: it reserves archaeological occurrence data as an
-independent dataset that can be used to assess the retrodictive
+accurate predictions in unknown environments \citep[like the
+past,][]{FranklinEtAl2015}, but here the hindcasting approach also
+presents an opportunity: it reserves archaeological occurrence data as
+an independent dataset that can be used to assess the retrodictive
 performance of the model. This possibily was suggested by
@@
 archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
-Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 96
-taxa (Table~\ref{tbl-taxa}) comprising the identifiable species observed
-at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites,
-according to our previous study of the regional archaeobotanical data
-\citep{ArranzOtaeguiRoe2023}. This was based on dataset of 117
-assemblages collated from three previously published regional
+Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 97
+taxa (Table~\ref{tbl-results-summary}) comprising the identifiable
+species observed at at least two Late Epipalaeolithic/Pre-Pottery
+Neolithic sites, according to our previous study of the regional
+archaeobotanical data \citep{ArranzOtaeguiRoe2023}. This was based on
+dataset collated from three previously published regional
 archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
 \citep{ORIGINS}, and COMPAG \citep{COMPAG}. We did not attempt to
-distinguish between the source of the remains; archaeobotanical
-assemblages are subject to a variety of preservational and recovery
-biases, so by no means were all the species on our list consumed or even
-deliberately collected by people. However, we assume that there presence
-at a site of human settlement at least implies that they were part of
-the wider ecosystem that supported habitation there.
+distinguish between the source of the remains
+\citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
+to a variety of preservational and recovery biases, so by no means were
+all the species on our list consumed or even deliberately collected by
+people. However, we assume that there presence at a site of human
+settlement at least implies that they were part of the wider ecosystem
+that supported habitation there.
 
@@
 recorded under the older nomenclature \emph{Scirpus maritimus} (see
-Table~\ref{tbl-taxa}). Domestic species meeting our inclusion criteria
-were substituted with their wild progenitor(s), where different.
+Table~\ref{tbl-results-summary}). Domestic species meeting our inclusion
+criteria were substituted with their wild progenitor(s), where
+different.
 
@@
 \citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
-excluded 81 taxa with less than 50 usable occurrences, following
+excluded 14 taxa with less than 50 usable occurrences, following
 recommendations for niche models generally and Random Forest-based
@@
 \caption{\label{fig-occ-map}Georeferenced occurrence records from West
-Asia used to train models (N=78222}
+Asia used to train models (N=81641}
 
@@
 
-\subsection{Predictor data}\label{predictor-data}
-
-We modelled the occurrence of species as a function of X spatial
-predictor variables (Table~\ref{tbl-predictors}). These included:
+\subsection{Predictor data}\label{sec-predictors}
+
+We modelled the occurrence of species as a function of 24 geospatial
+predictor variables. These included:
 
@@
 
-\begin{table}
-
-\caption{\label{tbl-climate-periods}}
-
-\centering{
-
-\fontsize{12.0pt}{14.4pt}\selectfont
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr}
-\toprule
-Period & Age, ka \\ 
-\midrule\addlinespace[2.5pt]
-Bølling-Allerød (BA) & 14.7–12.9 \\ 
-Younger Dryas (YDS) & 12.9–11.7 \\ 
-Early Holocene (EH) & 11.7–8.3 \\ 
-Current (CUR) & — \\ 
-\bottomrule
-\end{tabular*}
-
-Palaeoclimatic periods used for hindcasting, after \citet{BrownEtAl2018}
-
-}
-
-\end{table}%
-
-For hindcasting, we used reconstructed bioclimatic data for 3 key
-periods (Table~\ref{tbl-climate-periods}) generated from downscaled
-paleoclimate simulations from the HadCM3 general circulation model
-\citep{BrownEtAl2018}. Terrain and soil predictors were held constant,
+For hindcasting, we used reconstructed bioclimatic data for three key
+climatologies generated from downscaled paleoclimate simulations from
+the HadCM3 general circulation model
+\citep{FordhamEtAl2017, BrownEtAl2018}: the Bølling--Allerød
+(c.~14.7--12.9 ka), the Younger Dryas (c.~12.9--11.7 ka), and the Early
+Holocene (11.7--8.3 ka). Terrain and soil predictors were held constant,
 since reconstructions of these variables in the past are not available
-at sufficient scale. It is not likely that either macroscale topography
-or soil characteristics have altered significantly over the period of
-time considered here, so we assume that this does not degrade model
+at sufficient scale. It is unlikely that either macroscale topography or
+soil characteristics have altered significantly over the period of time
+considered here, so we assume that this does not degrade model
 performance, and may in fact benefit it by providing `anchoring'
@@
 For each taxon we trained a classification model to predict occurrence
-(presence or absence/background) based on our X predictor variables
-(Table~\ref{tbl-predictors}). We used the Random Forest algorithm
+(presence or absence/background) based on our 24 predictor variables
+(Section~\ref{sec-predictors}). We used the Random Forest algorithm
 implemented in the R package `ranger' \citep{WrightZiegler2017} and the
@@
 
-We trained Random Forest models for 81 taxa using contemporary
+We trained Random Forest models for 82 taxa using contemporary
 occurrence data from GBIF, a random sample of background points, and the
-predictor variables described in Table~\ref{tbl-predictors}.
+predictor variables described in Section~\ref{sec-predictors}.
 Substituting the ``current'' climate predictors for those derived from
@@
 hindcast predictions for reconstructed past environments in 4 key
-climate periods -- a total of 324 modelled palaeodistributions.
+climate periods -- a total of 328 modelled palaeodistributions.
 Predicted distributions for individual taxa are presented in the
@@
 We assessed the predictive performance of the fitted niche models in the
-contemporary environment based on the reserved test partition. Model
-accuracy (proportion of correctly classified presence and background
-samples) ranged between 72\% and 99\%, with an average of 92\%.
-Sensitivity (proportion of correctly classified presence samples) ranged
-between 79\% and 100\%, with an average of 94\%. The area under the
-models' receiver operating characteristic curves (ROC-AUC) was in the
-range of 0.978±0.061 Model sensitivity is loosely correlated with the
-number of occurrences available for training
-(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
-models all having less than 300 recorded occurrences: \emph{Avena
-fatua}, \emph{Lepidium perfoliatum}, \emph{Triticum turgidum dicoccum},
-and \emph{Zygophyllum fabago}. Test metrics and ROC curves for the
-individual models are included in the appendix.
+contemporary environment based on the reserved test partition
+Table~\ref{tbl-results-summary}. Model accuracy (proportion of correctly
+classified presence and background samples) ranged between 73\% and
+99\%, with an average of 93\%. Sensitivity (proportion of correctly
+classified presence samples) ranged between 79\% and 100\%, with an
+average of 94\%. The area under the models' receiver operating
+characteristic curves (ROC-AUC) was on average 0.978±0.061 Model
+sensitivity is loosely correlated with the number of occurrences
+available for training (Figure~\ref{fig-model-sensitivity-vs-n_occ}),
+with the worst-performing models all having less than 300 recorded
+occurrences: \emph{Asphodelus aestivus}, \emph{Triticum turgidum
+dicoccum}, \emph{Zygophyllum fabago}, and \emph{Stipa dregeana}. Test
+metrics and ROC curves for the individual models are included in the
+appendix.
+
+\begin{verbatim}
+Joining with `by = join_by(taxon)`
+\end{verbatim}
 
 The ability of the hindcast models to predict the occurrence of specific
-species at archaeological sites is significantly worse, with only 5\% of
-presences in archaeobotanical assemblages successfully predicted.
+species at archaeological sites is worse, with 10\% of presences in
+archaeobotanical assemblages successfully predicted. Model sensitivity
+(proportion of corrected predicted presenses) in relation to the
+archaeological data is on average 0.06±0.13
+Table~\ref{tbl-results-summary}. A full assessment of the hindcasting
+performance of the individual models can be found in the appendix.
 
@@
 \setlength\LTright{0\linewidth}\fontsize{8.2pt}{9.9pt}\selectfont
-
-\begin{longtable}{@{\extracolsep{\fill}}lrrrrr}
-
-\caption{\label{tbl-taxa}Summary of species modelled}
+\setlength{\LTpost}{0mm}
+
+\begin{longtable}{@{\extracolsep{\fill}}lrrrrrr}
+
+\caption{\label{tbl-results-summary}Summary of species modelled}
 
@@
 \toprule
- & \multicolumn{2}{c}{Occurrences} & \multicolumn{3}{c}{Model} \\ 
-\cmidrule(lr){2-3} \cmidrule(lr){4-6}
-Taxon & Archaeological & Present (GBIF) & Accuracy & ROC-AUC & Sensitivity \\ 
+ & \multicolumn{2}{c}{Occurrences} & \multicolumn{4}{c}{Model} \\ 
+\cmidrule(lr){2-3} \cmidrule(lr){4-7}
+Taxon & Arch. & Cur. & Acc. & ROC-AUC & Sens. & Sens. (arch.) \\ 
 \midrule\addlinespace[2.5pt]
-{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.99 \\ 
-{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.89 \\ 
-{\itshape Aegilops speltoides} & — & 701 & 0.93 & 0.98 & 0.94 \\ 
-{\itshape Aegilops tauschii} & — & 725 & 0.90 & 0.98 & 0.94 \\ 
-{\itshape Aizoanthemopsis hispanica} & 4 & 515 & 0.98 & 0.99 & 0.96 \\ 
-{\itshape Ammi majus} & 4 & 900 & 0.99 & 0.99 & 0.98 \\ 
-{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.96 & 0.88 \\ 
-{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.85 & 0.94 & 0.91 \\ 
-{\itshape Arnebia decumbens} & 21 & 324 & 0.94 & 0.98 & 0.93 \\ 
-{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 1.00 & 0.98 \\ 
-{\itshape Asphodelus aestivus} & 2 & 76 & 0.96 & 0.97 & 0.96 \\ 
-{\itshape Atriplex prostrata} & 4 & 266 & 0.97 & 1.00 & 0.97 \\ 
-{\itshape Avena fatua} & 2 & 579 & 0.85 & 0.93 & 0.85 \\ 
-{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 0.98 \\ 
-{\itshape Bolboschoenus maritimus} & 29 & 1040 & 0.98 & 1.00 & 0.99 \\ 
-{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 \\ 
-{\itshape Bromus sterilis} & 3 & 759 & 0.96 & 1.00 & 0.98 \\ 
-{\itshape Buglossoides arvensis} & 23 & 260 & 0.86 & 0.96 & 0.90 \\ 
-{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.94 \\ 
-{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 \\ 
-{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 \\ 
-{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.93 \\ 
-{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 \\ 
-{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.92 \\ 
-{\itshape Cicer reticulatum} & 3 & 52 & 0.96 & 1.00 & 0.92 \\ 
-{\itshape Citrullus colocynthis} & 4 & 475 & 0.89 & 0.95 & 0.89 \\ 
-{\itshape Crithopsis delileana} & 3 & 406 & 0.97 & 0.99 & 0.97 \\ 
-{\itshape Erodium ciconium} & 2 & 359 & 0.91 & 0.99 & 0.99 \\ 
-{\itshape Ficus carica} & 8 & 3495 & 0.93 & 0.98 & 0.93 \\ 
-{\itshape Fumaria densiflora} & 4 & 413 & 0.98 & 0.99 & 0.98 \\ 
-{\itshape Gypsophila elegans} & 3 & 76 & 0.93 & 0.96 & 0.85 \\ 
-{\itshape Gypsophila pilosa} & 5 & 173 & 0.93 & 0.99 & 0.95 \\ 
-{\itshape Gypsophila vaccaria} & 8 & 511 & 0.96 & 0.99 & 0.93 \\ 
-{\itshape Halothamnus hierochunticus} & 2 & 57 & 0.99 & 1.00 & 1.00 \\ 
-{\itshape Helianthemum ledifolium} & 2 & 697 & 0.97 & 0.98 & 0.96 \\ 
-{\itshape Helianthemum salicifolium} & 2 & 1400 & 0.98 & 0.99 & 0.97 \\ 
-{\itshape Hordeum bulbosum} & 4 & 2850 & 0.97 & 0.99 & 0.96 \\ 
-{\itshape Hordeum murinum} & 5 & 1949 & 0.95 & 0.99 & 0.96 \\ 
-{\itshape Hordeum spontaneum} & 51 & 4656 & 0.96 & 0.99 & 0.97 \\ 
-{\itshape Juglans regia} & 2 & 821 & 0.85 & 0.96 & 0.94 \\ 
-{\itshape Lathyrus aphaca} & 4 & 2079 & 0.95 & 0.99 & 0.96 \\ 
-{\itshape Lathyrus cicera} & 2 & 591 & 0.90 & 0.98 & 0.94 \\ 
-{\itshape Lathyrus oleraceus} & 6 & 1580 & 0.88 & 0.97 & 0.94 \\ 
-{\itshape Lathyrus sativus} & 4 & 196 & 0.80 & 0.94 & 0.87 \\ 
-{\itshape Lepidium perfoliatum} & 3 & 83 & 0.80 & 0.92 & 0.83 \\ 
-{\itshape Linum bienne} & 9 & 376 & 0.96 & 1.00 & 0.98 \\ 
-{\itshape Lolium rigidum} & 5 & 2529 & 0.98 & 1.00 & 0.98 \\ 
-{\itshape Lolium temulentum} & 3 & 159 & 0.93 & 0.95 & 0.92 \\ 
-{\itshape Medicago astroites} & 15 & 143 & 0.90 & 0.97 & 0.88 \\ 
-{\itshape Medicago minima} & 2 & 1379 & 0.92 & 0.99 & 0.98 \\ 
-{\itshape Medicago radiata} & 20 & 799 & 0.93 & 0.99 & 0.97 \\ 
-{\itshape Moltkia coerulea} & 2 & 71 & 0.86 & 0.99 & 1.00 \\ 
-{\itshape Neotorularia torulosa} & 2 & 539 & 0.98 & 0.99 & 0.96 \\ 
-{\itshape Peganum harmala} & 2 & 917 & 0.91 & 0.98 & 0.95 \\ 
-{\itshape Phalaris paradoxa} & 3 & 1191 & 0.99 & 0.99 & 0.98 \\ 
-{\itshape Phragmites australis} & 4 & 3412 & 0.97 & 0.99 & 0.97 \\ 
-{\itshape Pistacia atlantica} & 6 & 1874 & 0.98 & 1.00 & 0.98 \\ 
-{\itshape Plantago lagopus} & 2 & 1730 & 0.98 & 1.00 & 0.98 \\ 
-{\itshape Poa bulbosa} & 5 & 1880 & 0.95 & 0.98 & 0.90 \\ 
-{\itshape Prosopis farcta} & 5 & 2263 & 0.97 & 1.00 & 0.97 \\ 
-{\itshape Quercus ithaburensis} & 2 & 2216 & 0.98 & 1.00 & 0.99 \\ 
-{\itshape Rumex pulcher} & 6 & 1394 & 0.98 & 1.00 & 0.99 \\ 
-{\itshape Salsola kali} & 6 & 537 & 0.98 & 1.00 & 1.00 \\ 
-{\itshape Salvia absconditiflora} & 3 & 105 & 0.92 & 0.99 & 0.91 \\ 
-{\itshape Secale cereale} & 4 & 268 & 0.79 & 0.95 & 0.92 \\ 
-{\itshape Secale strictum} & 3 & 124 & 0.74 & 0.96 & 0.92 \\ 
-{\itshape Stipa dregeana} & 2 & 79 & 0.93 & 0.99 & 0.97 \\ 
-{\itshape Suaeda fruticosa} & 3 & 238 & 0.99 & 1.00 & 1.00 \\ 
-{\itshape Taeniatherum caput-medusae} & 4 & 262 & 0.82 & 0.96 & 0.92 \\ 
-{\itshape Tribulus terrestris} & 2 & 1091 & 0.91 & 0.97 & 0.93 \\ 
-{\itshape Triticum aestivum compactum} & 2 & 147 & 0.80 & 0.96 & 0.92 \\ 
-{\itshape Triticum durum} & 3 & 95 & 0.90 & 0.94 & 0.88 \\ 
-{\itshape Triticum monococcum aegilopoides} & 36 & 1176 & 0.88 & 0.98 & 0.96 \\ 
-{\itshape Triticum turgidum dicoccum} & 47 & 60 & 0.72 & 0.85 & 0.79 \\ 
-{\itshape Triticum urartu} & — & 424 & 0.93 & 1.00 & 1.00 \\ 
-{\itshape Verbena officinalis} & 3 & 934 & 0.95 & 0.99 & 0.96 \\ 
-{\itshape Vicia ervilia} & 26 & 688 & 0.88 & 0.97 & 0.93 \\ 
-{\itshape Vicia faba} & 7 & 1597 & 0.91 & 0.98 & 0.95 \\ 
-{\itshape Vicia narbonensis} & 3 & 1451 & 0.93 & 0.98 & 0.94 \\ 
-{\itshape Vicia orientalis} & 14 & 353 & 0.84 & 0.95 & 0.88 \\ 
-{\itshape Zygophyllum fabago} & 3 & 263 & 0.89 & 0.95 & 0.84 \\ 
+{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
+{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.86 & 0.00 \\ 
+{\itshape Aegilops speltoides} & — & 701 & 0.92 & 0.98 & 0.95 & — \\ 
+{\itshape Aegilops tauschii} & — & 725 & 0.89 & 0.98 & 0.94 & — \\ 
+{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 515 & 0.98 & 1.00 & 0.96 & 0.00 \\ 
+{\itshape Ammi majus} & 4 & 900 & 0.98 & 0.99 & 0.97 & 0.00 \\ 
+{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.95 & 0.88 & 0.12 \\ 
+{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.84 & 0.95 & 0.91 & 0.00 \\ 
+{\itshape Arnebia decumbens} & 21 & 324 & 0.95 & 0.95 & 0.89 & 0.00 \\ 
+{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 0.99 & 0.94 & 0.00 \\ 
+{\itshape Asphodelus aestivus} & 2 & 76 & 0.97 & 0.97 & 0.83 & 0.00 \\ 
+{\itshape Atriplex prostrata} & 4 & 266 & 0.98 & 1.00 & 0.94 & 0.00 \\ 
+{\itshape Avena fatua} & 2 & 579 & 0.86 & 0.93 & 0.86 & 0.00 \\ 
+{\itshape Avena sterilis} & 5 & 53244\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
+{\itshape Bassia prostrata} & 2 & 46\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Bolboschoenus glaucus} & 5 & 27\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{3}} & 31 & 1040 & 0.98 & 0.99 & 0.97 & 0.03 \\ 
+{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 & 0.00 \\ 
+{\itshape Bromus sterilis} & 3 & 759 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
+{\itshape Buglossoides arvensis} & 23 & 260 & 0.87 & 0.97 & 0.95 & 0.09 \\ 
+{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.95 & 0.08 \\ 
+{\itshape Camphorosma monspeliaca} & 2 & 38\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 & 0.00 \\ 
+{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 & 0.00 \\ 
+{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.94 & 0.00 \\ 
+{\itshape Celtis tournefortii} & 2 & 14\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 & 0.50 \\ 
+{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.94 & 0.00 \\ 
+{\itshape Cicer reticulatum}\textsuperscript{\textit{4}} & 3 & 52 & 0.97 & 1.00 & 1.00 & 0.00 \\ 
+{\itshape Citrullus colocynthis} & 4 & 475 & 0.89 & 0.96 & 0.90 & 0.00 \\ 
+{\itshape Crithopsis delileana} & 3 & 406 & 0.98 & 1.00 & 0.97 & 0.00 \\ 
+{\itshape Erodium ciconium} & 2 & 359 & 0.92 & 0.99 & 0.94 & 0.00 \\ 
+{\itshape Euclidium syriacum} & 3 & 32\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Ficus carica} & 8 & 3495 & 0.94 & 0.99 & 0.95 & 0.25 \\ 
+{\itshape Fumaria densiflora} & 4 & 413 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
+{\itshape Gypsophila elegans} & 3 & 76 & 0.92 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Gypsophila pilosa} & 5 & 173 & 0.93 & 0.97 & 0.90 & 0.00 \\ 
+{\itshape Gypsophila vaccaria}\textsuperscript{\textit{5}} & 8 & 511 & 0.96 & 0.99 & 0.93 & 0.00 \\ 
+{\itshape Halothamnus hierochunticus} & 2 & 57 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
+{\itshape Helianthemum ledifolium} & 2 & 697 & 0.97 & 0.99 & 0.97 & 0.00 \\ 
+{\itshape Helianthemum salicifolium} & 2 & 1400 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
+{\itshape Henrardia pubescens} & 3 & 1\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Hordeum bulbosum} & 4 & 2850 & 0.97 & 0.99 & 0.96 & 0.00 \\ 
+{\itshape Hordeum murinum} & 5 & 1949 & 0.94 & 0.98 & 0.93 & 0.00 \\ 
+{\itshape Hordeum spontaneum}\textsuperscript{\textit{6}} & 110 & 4656 & 0.96 & 0.99 & 0.97 & 0.20 \\ 
+{\itshape Juglans regia} & 2 & 821 & 0.85 & 0.96 & 0.95 & 0.00 \\ 
+{\itshape Krascheninnikovia ceratoides} & 2 & 42\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Lathyrus aphaca} & 4 & 2079 & 0.94 & 0.99 & 0.97 & 0.00 \\ 
+{\itshape Lathyrus cicera} & 2 & 591 & 0.89 & 0.98 & 0.94 & 0.50 \\ 
+{\itshape Lathyrus oleraceus}\textsuperscript{\textit{7}} & 8 & 1580 & 0.89 & 0.97 & 0.93 & 0.12 \\ 
+{\itshape Lathyrus sativus} & 4 & 196 & 0.79 & 0.95 & 0.92 & 0.00 \\ 
+{\itshape Lepidium perfoliatum} & 3 & 83 & 0.77 & 0.94 & 0.91 & 0.00 \\ 
+{\itshape Lepidium ruderale} & 2 & 37\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Linum bienne}\textsuperscript{\textit{8}} & 14 & 376 & 0.96 & 1.00 & 0.99 & 0.00 \\ 
+{\itshape Lolium rigidum} & 5 & 2529 & 0.98 & 1.00 & 0.98 & 0.00 \\ 
+{\itshape Lolium temulentum} & 3 & 159 & 0.94 & 0.96 & 0.90 & 0.00 \\ 
+{\itshape Medicago astroites}\textsuperscript{\textit{9}} & 15 & 143 & 0.89 & 0.96 & 0.88 & 0.33 \\ 
+{\itshape Medicago minima} & 2 & 1379 & 0.92 & 0.99 & 0.97 & 0.50 \\ 
+{\itshape Medicago radiata} & 20 & 799 & 0.93 & 0.99 & 0.98 & 0.20 \\ 
+{\itshape Minuartia rubella} & 2 & 0\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Moltkia angustifolia} & 2 & 11\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Moltkia coerulea} & 2 & 71 & 0.86 & 0.97 & 0.88 & 0.00 \\ 
+{\itshape Neotorularia torulosa}\textsuperscript{\textit{10}} & 2 & 539 & 0.98 & 0.98 & 0.93 & 0.00 \\ 
+{\itshape Peganum harmala} & 2 & 917 & 0.92 & 0.97 & 0.93 & 0.00 \\ 
+{\itshape Phalaris paradoxa} & 3 & 1191 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
+{\itshape Phragmites australis} & 4 & 3412 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
+{\itshape Pistacia atlantica} & 6 & 1874 & 0.98 & 1.00 & 0.97 & 0.17 \\ 
+{\itshape Pistacia terebinthus}\textsuperscript{\textit{11}} & 2 & 3419 & 0.98 & 1.00 & 0.98 & 0.50 \\ 
+{\itshape Plantago lagopus} & 2 & 1730 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
+{\itshape Poa bulbosa} & 5 & 1880 & 0.95 & 0.98 & 0.92 & 0.00 \\ 
+{\itshape Polygonum arenarium arenarium}\textsuperscript{\textit{12}} & 7 & 5\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Polygonum corrigioloides} & 6 & 4\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Prosopis farcta} & 5 & 2263 & 0.98 & 0.99 & 0.96 & 0.00 \\ 
+{\itshape Quercus ithaburensis} & 2 & 2216 & 0.99 & 1.00 & 0.97 & 0.00 \\ 
+{\itshape Rumex pulcher} & 6 & 1394 & 0.98 & 1.00 & 0.98 & 0.00 \\ 
+{\itshape Salsola kali} & 6 & 537 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
+{\itshape Salvia absconditiflora}\textsuperscript{\textit{13}} & 3 & 105 & 0.94 & 0.98 & 0.92 & 0.00 \\ 
+{\itshape Secale cereale} & 7 & 268 & 0.81 & 0.94 & 0.90 & 0.00 \\ 
+{\itshape Secale strictum}\textsuperscript{\textit{14}} & 3 & 124 & 0.76 & 0.94 & 0.97 & 0.00 \\ 
+{\itshape Stipa dregeana} & 2 & 79 & 0.95 & 0.96 & 0.85 & 0.00 \\ 
+{\itshape Suaeda fruticosa} & 3 & 238 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
+{\itshape Taeniatherum caput-medusae} & 5 & 262 & 0.81 & 0.96 & 0.95 & 0.00 \\ 
+{\itshape Tribulus terrestris} & 2 & 1091 & 0.90 & 0.96 & 0.89 & 0.00 \\ 
+{\itshape Triticum aestivum compactum} & 2 & 147 & 0.82 & 0.96 & 0.92 & 0.00 \\ 
+{\itshape Triticum aestivum spelta}\textsuperscript{\textit{15}} & 2 & 16\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Triticum durum} & 3 & 95 & 0.91 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Triticum monococcum aegilopoides}\textsuperscript{\textit{16}} & 62 & 1176 & 0.88 & 0.99 & 0.99 & 0.08 \\ 
+{\itshape Triticum turgidum dicoccum}\textsuperscript{\textit{17}} & 60 & 60 & 0.73 & 0.86 & 0.79 & 0.18 \\ 
+{\itshape Triticum urartu} & — & 424 & 0.93 & 0.99 & 0.95 & — \\ 
+{\itshape Verbena officinalis} & 3 & 934 & 0.96 & 0.99 & 0.97 & 0.00 \\ 
+{\itshape Vicia ervilia} & 26 & 688 & 0.87 & 0.96 & 0.94 & 0.12 \\ 
+{\itshape Vicia faba} & 7 & 1597 & 0.90 & 0.98 & 0.94 & 0.43 \\ 
+{\itshape Vicia narbonensis} & 3 & 1451 & 0.93 & 0.99 & 0.98 & 0.33 \\ 
+{\itshape Vicia orientalis}\textsuperscript{\textit{18}} & 16 & 353 & 0.84 & 0.95 & 0.92 & 0.19 \\ 
+{\itshape Vitis sylvestris} & 3 & 12\textsuperscript{\textit{2}} & — & — & — & — \\ 
+{\itshape Zygophyllum fabago} & 3 & 263 & 0.89 & 0.93 & 0.81 & 0.00 \\ 
 \bottomrule
@@
 
+\begin{minipage}{\linewidth}
+\textsuperscript{\textit{1}}Including \emph{Aizoon hispanicum}\\
+\textsuperscript{\textit{2}}Excluded from modelling due to sample size\\
+\textsuperscript{\textit{3}}Including \emph{Scirpus maritimus}\\
+\textsuperscript{\textit{4}}Including \emph{Cicer arietinum}\\
+\textsuperscript{\textit{5}}Including \emph{Vaccaria pyramidata}\\
+\textsuperscript{\textit{6}}Including \emph{Hordeum vulgare}\\
+\textsuperscript{\textit{7}}Including \emph{Pisum sativum} and \emph{Pisum elatius}\\
+\textsuperscript{\textit{8}}Including \emph{Linum usitatissimum}\\
+\textsuperscript{\textit{9}}Including \emph{Trigonella astroites}\\
+\textsuperscript{\textit{10}}Including \emph{Torularia torulosa}\\
+\textsuperscript{\textit{11}}Including \emph{Pistacia palaestina}\\
+\textsuperscript{\textit{12}}Including \emph{Polygonum venantianum}\\
+\textsuperscript{\textit{13}}Including \emph{Salvia cryptantha}\\
+\textsuperscript{\textit{14}}Including \emph{Secale montanum}\\
+\textsuperscript{\textit{15}}Including \emph{Triticum spelta}\\
+\textsuperscript{\textit{16}}Including \emph{Triticum monococcum} and \emph{Triticum boeoticum}\\
+\textsuperscript{\textit{17}}Including \emph{Triticum aestivum}, \emph{Triticum dicoccum}, and \emph{Triticum dicoccoides}\\
+\textsuperscript{\textit{18}}Including \emph{Lens culinaris}\\
+\end{minipage}
 \endgroup
@@
 reduction and/or shift in their ranges since the Early Holocene. 75 of
-81 species have reduced ranges; 65 of more than 10\% or more. Though the
+82 species have reduced ranges; 66 of more than 10\% or more. Though the
 magnitude of the reduction from prehistory to the present likely
@@
 (11.7--8.3 ka) are more directly comparable
-\textbf{?@fig-range-density}. The average range of modelled species was
-25\% in the Early Holocene compared to the Bølling--Allerød, and 32\%
-during the Younger Dryas (i.e.~ranges recovered slightly between the
-Younger Dryas and Early Holocene). This perhaps indicates that although
-this period is considered one of climatatic amelioration globally
-\citep{JonesEtAl2019}, the colder conditions of the Pleistocene may have
-supported more extensive plant-based economies in West Asia
+(Figure~\ref{fig-pred-range-density}). The average range of modelled
+species was 16\% in the Early Holocene compared to the Bølling--Allerød,
+and 25\% during the Younger Dryas (i.e.~ranges recovered slightly
+between the Younger Dryas and Early Holocene). This perhaps indicates
+that although this period is considered one of climatatic amelioration
+globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
+may have supported more extensive plant-based economies in West Asia
 specifically.
@@
 have had a significantly more restricted distribution in the terminal
-Pleistocene/Early Holocene (\textbf{?@fig-sum-palaeodist}). These
+Pleistocene/Early Holocene (Figure~\ref{fig-palaeodist-sum}). These
 include \emph{Ficus carica} (fig); \emph{Hordeum} spp. (wild barleys);
@@
 Our results for the Levant are consistent with the current understanding
-of this region as developing early intensive foraging economies {[}the
-Natufian culture, @{]} and as a centre of origin of agriculture {[}@{]}.
-Within the Levant, retreat from the Badia.
+of this region as developing early intensive foraging economies
+\citep[the Natufian culture,][]{cite} and as a centre of origin of
+agriculture \citep{cite}. Within the Levant, many species show moderate
+reductions in ranges over the Pleistocene/Holocence boundary, retreating
+from the Badia/transjordan region
+e.g.~\ref{fig-palaeodist-barely-levant}.
 
@@
 earliest clear evidence of domestication
-\citep{ZoharyHopfWeiss2012, KabukcuEtAl2021, UlasEtAl2024}. Comparative
+\citep{ZoharyEtAl2012, KabukcuEtAl2021, UlasEtAl2024}. Comparative
 genetics also points to the Northern Mesopotamia region as the centre of
@@
 Mesopotamia but encompassing much of the the `hilly flanks' (except the
-southern Levant, \textbf{?@fig-chickpea-rye}). Another is rye
-(\emph{Secale cereale}), which is inferred to be primarily Western
+southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}). Another is
+rye (\emph{Secale cereale}), which is inferred to be primarily Western
 Anatolian Figure~\ref{fig-palaeodist-chickpea-rye}. This is perhaps
@@
 
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-barely-levant-1.pdf}}
+
+}
+
+\caption{\label{fig-palaeodist-barely-levant}Predicted
+palaeodistribution of wild barleys in the Levant}
+
+\end{figure}%
+
 Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
@@
 concurrent with it being brought into cultivation
-\textbf{?@fig-palaeodist-barely-levant}. It also sees a marked decline
+Figure~\ref{fig-palaeodist-barely-levant}. It also sees a marked decline
 in the archaeobotanical record from the Early PPNA/Early PPNB (where it
@@
 eachother in the time frame of domestication is a narrow area around the
-Orontes river in the northern Levant Figure~\ref{fig-palaeodist-bread}.
-This area is c.~200 km from the only two sites in our archaeobotanical
-database where either both species of wheat or both species of wheat and
-the goatgrasses are found together: Tell Abu Hureyra and El Kowm II
+Orontes river in the northern Levant
+(Figure~\ref{fig-palaeodist-bread}). This area is c.~200 km from the
+only two sites in our archaeobotanical database where either both
+species of wheat or both species of wheat and the goatgrasses are found
+together: Tell Abu Hureyra and El Kowm II
 \citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to Benjamin Nowak
@@
 
-Is consistent with more ``macro'' trends such as the reduced range of
+It is consistent with more ``macro'' trends such as the reduced range of
 Hordeum and Pistacia correlating with its reduced abundance in the
@@
 
+We present the first continuous, spatially explicit models of the
+palaeodistributions of 82 plant species found regularly in association
+with Late Epipalaeolithic and Early Neolithic sites in West Asia. This
+deductive approach---modelling the niche of a species based on its
+occurrence in relation to environmental factors today, and using this
+together with palaeoclimate simulations to infer its past
+distribution---represents a new line of evidence on the archaeoecology
+of the world's first agricultural societies. It provides a complementary
+picture to that gleaned from environmental archaeology and
+climatological archives because it is independent of the taphonomic,
+anthropic, and recovery-related processes that affect these records.
+
+``All models are wrong'' \citep{cite}, but the performance of our models
+on independent test datasets give confidence in its predictions, which
+are plausible and consistent with broad-scale patterns in the
+archaeological record.
+
 \begin{itemize}
@@
 \item
-  We present the first continuous, spatially explicit models of the
-  palaeodistributions of N plant species found regularly in association
-  with early agricultural archaeological sites in West Asia
-
-  \begin{itemize}
-  \tightlist
-  \item
-    A new line of evidence on archaeoecology
-  \item
-    Complementary to archaeobot/pollen/etc. because it is independent
-    from it
-  \item
-    All models are wrong\ldots{} but it's easier to see how these are
-    wrong than lines on maps
-  \end{itemize}
-\item
-  Modelling at scale using random forest, modern occurrences, and
-  hindcasting represents a significant advance in pENM methodology
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Relies on recent open ecological and climatic datasets
-  \item
-    \ldots open archaeological datasets still lacking!
-  \end{itemize}
-\item
-  First (? - check Yaworsky) attempt to verify hindcast models with
-  archaeo. compositional data
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Results not to promising, but this doesn't mean the models are
-    useless!
+  Results w.r.t. to specific archaeologically-attested species not to
+  promising, but this doesn't mean the models are useless!
 \item
@@
 \end{itemize}
-\end{itemize}
+
+Modelling a large number of species using machine learning, the
+substantial occurrence datasets available for the present day, and a
+hindcasting approach to past distributions also represents a significant
+advance in the methodology of palaeoecological niche modelling. This
+approach is enabled by the availability of high quality, global open
+datasets in ecology \citep{GBIF2025, GBIFSecretariat2023}, earth science
+\citep{SRTM}, and climatology
+\citep[@][]{KargerEtAl2017, BrownEtAl2018}. Unfortunately in relation to
+these fields open data in archaeology lags conspiciously behind. Though
+we have benefitted from the relatively good state of publication of
+archaeobotanical compilations in our region of study
+\citep{ADEMNES, COMPAG, ORIGINS}, further development of open,
+comprehensive and up-to-date `backbone' datasets on site locations and
+chronologies is needed to advance archaeoecological modelling to the
+same level.
 

### 2025-01-10 analysis/appendix_section.qmd
Prepare draft

@@
 
-```{r fig-palaeodist, fig.height=16, fig.cap=paste0("Predicted palaeodistributions of *", taxon, "*")}
+```{r fig-palaeodist, warning=FALSE, fig.height=16, fig.cap=paste0("Predicted palaeodistributions of *", taxon, "*")}
 if (length(archaeo_pred) > 0) {

### 2025-01-10 analysis/paper.qmd
Prepare draft

@@
   Modelled ranges do not reliably predict the observed  occurrence of specific 
-  species at archaeological sites,
+  species at archaeological sites, but are aligned with more general trends in
+  the relative abundance of species.<!--
   however [...]
@@
   to have been influenced by environmental change and species turnover, in
-  addition to human economic choices.
+  addition to human economic choices.-->
 format: 
@@
 The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
-West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@cite].
-These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@cite] and eventually pre-domestication cultivation [@cite].
+West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@Diamond2002].
+These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@BarYosef1998; @MaherEtAl2012] and eventually plant management and pre-domestication cultivation [@Harris2007].
 As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.
 
-Decades of research in archaeobotany and zooarchaeology have reconstructed the subsistence economies of Late Epipalaeolithic and Neolithic sites in great detail [@cites].
-Together with studies of charcoal [@cite], pollen [@cite], soil [@cite], and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
+Decades of research in archaeobotany and zooarchaeology have reconstructed the subsistence economies of Late Epipalaeolithic and Neolithic sites in great detail.
+Together with studies of charcoal, pollen, soil, and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
 However, each of these sources of evidence is subject to the wide variety of taphonomic and recovery biases that are inherent in any direct record of the past.
@@
 # Background
-
+<!--
 * The transition to agriculture in West Asia was...
@@
 ## Ecological niche modelling in archaeology
-
+-->
 Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
@@
 This model can then be used to predict the range of the organism in question either in the same or a different environment.
-@CITE suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
+@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
 Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.
@@
 @deAndresHerreroEtAl2018; @YaworskyEtAl2023] and, in a West Asian context, of foraged plant resources in the landscape around the Neolithic sites on the Konya Plain [@CollinsEtAl2018].
-Modelling the spread of crops has been another significant archaeological application [e.g. @CremaEtAl], though not as yet applied to West Asia.
+Modelling the spread of crops has been another significant archaeological application [e.g. @KrzyzanskaEtAl2022; @Krzyzanska2023], though not as yet applied to West Asia.
 
@@
 The major practical limitation of the hindcasting approach is that it relies on spatially explicit, high resolution palaeoenvironmental surfaces with continuous coverage of the region and periods of interest.
-Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @YaworskyEtAl2023].
+Until recently, this has not been widely available for most applications, which is perhaps why only a minority of studies use it [cf. @KrzyzanskaEtAl2022; @YaworskyEtAl2023].
 In this study, we are able to take advantage of the increasing availability of high resolution, global palaeoclimate data derived from simulation experiments with general circulation models of climate [@BrownEtAl2018; @BrownEtAl2020; @KargerEtAl2023].
@@
 Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
-This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@COMPAG].
+This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@LucasFuller2018; @FullerEtAl2018; based on @ColledgeEtAl2004; @ShennanConolly2007].
 We did not attempt to distinguish between the source of the remains [cf. @WallaceEtAl2018];
@@
 
-flora <- left_join(flora, archaeo_pred)
+flora <- left_join(flora, archaeo_pred, join_by(taxon))
 ```
@@
 <!-- TODO expand/rewrite -->
-Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @cite] and as a centre of origin of agriculture [@cite].
-Within the Levant, many species show moderate reductions in ranges over the Pleistocene/Holocence boundary, retreating from the Badia/transjordan region [e.g. @fig-palaeodist-barely-levant].
+Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @BarYosef1998] and as a centre of origin of agriculture [@Zeder2011a].
+Within the Levant, many species show moderate reductions in ranges over the Pleistocene/Holocence boundary, retreating from the Badia/transjordan region (e.g.
+barley, fig.
+@fig-palaeodist-barely-levant).
 
 Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture.
-Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @cites], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyEtAl2012; @KabukcuEtAl2021; @UlasEtAl2024].
+Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @FullerEtAl2011; @ArranzOtaeguiEtAl2016], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyEtAl2012; @KabukcuEtAl2021; @UlasEtAl2024].
 Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [e.g. @HaasEtAl2019].
@@
 
-The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that domestication took place just as early in the eastern *Mashriq* as it did in the west [@Braidwood; @GanjDarehGoats; @HillyFlanksStuff].
+The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that domestication took place just as early in the eastern *Mashriq* as it did in the west [@Zeder2024].
 We consider that the most likely explanation for this is that our flora does not include the species that were most important to plant subsistence in the east.
-Archaeobotanical data on Neolithic sites in the Zagros is limited (compared to the Levant in particular) due to a hiatus in field research there from the 1980s to early 2010s [@cite].
-Recent research [@Amaia?] indicates ... <!-- TODO: weird plants -->
+Archaeobotanical data on Neolithic sites in the Zagros is limited (compared to the Levant in particular) due to a hiatus in field research there from the 1980s to early 2010s [@Zeder2024].
+Recent research [@WhitlamEtAl2018; @GonzalezCarreteroEtAl2023] indicates that plant subsistence in this region was based on a distinct set of species, compared to the Levant and Anatolia.
 
@@
 and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
-Were these area part of the same broader 'interaction sphere' [@cite] that produce Neolithic agriculture in West Asia?
+Were these area part of the same broader 'interaction sphere' that produce Neolithic agriculture in West Asia?
 
@@
 One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye).
-Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian [@fig-palaeodist-chickpea-rye].
+Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
 This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].
@@
 Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
-concurrent with it being brought into cultivation [@fig-palaeodist-barely-levant].
+concurrent with it being brought into cultivation (@fig-palaeodist-barely-levant).
 It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
@@
 
-"All models are wrong" [@cite], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
-
+"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
+<!--
 * Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
 * Discrepencies suggests several areas for further research and methodological development
-
+ --> 
 Modelling a large number of species using machine learning, the substantial occurrence datasets available for the present day, and a hindcasting approach to past distributions also represents a significant advance in the methodology of palaeoecological niche modelling.
-This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@@KargerEtAl2017; @BrownEtAl2018].
+This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@KargerEtAl2017; @BrownEtAl2018].
 Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
-Though we have benefitted from the relatively good state of publication of archaeobotanical compilations in our region of study [@ADEMNES; @COMPAG; @ORIGINS], 
+Though we have benefitted from the relatively long tradition of compiling archaeobotanical data in our region of study [@ColledgeEtAl2004; @ShennanConolly2007; @ADEMNES; @LucasFuller2018; @FullerEtAl2018; @WallaceEtAl2018; @ORIGINS],
 further development of open, comprehensive and up-to-date 'backbone' datasets on site locations and chronologies is needed to advance archaeoecological modelling to the same level.

### 2025-01-10 analysis/paper.tex
Prepare draft

@@
 reliably predict the observed occurrence of specific species at
-archaeological sites, however {[}\ldots{]} The regional ubiquity of
-species in the archaeological record is {[}not{]} correlated with the
-predicted size of its range and the diversity of archaeobotanical
-assemblages is {[}not{]} correlated with the predicted diversity of its
-environs. This indicates that trends in taxonomic composition of the
-archaeobotanical record is {[}not{]} likely to have been influenced by
-environmental change and species turnover, in addition to human economic
-choices.
+archaeological sites, but are aligned with more general trends in the
+relative abundance of species.
 \end{abstract}
@@
 relatives of wheat, barley, peas, lentils, and other crops of global
-importance \citep{cite}. These species supported uniquely dense and
-complex Late Epipalaeolithic (15--11.7 ka) societies based on intensive
-foraging \citep{cite} and eventually pre-domestication cultivation
-\citep{cite}. As they were brought under domestication in the
-Pre-Pottery Neolithic period (11.7--8.5 ka), the world's first
-agriculture was shaped by the ecosystems from which it emerged and was
-embedded in.
+importance \citep{Diamond2002}. These species supported uniquely dense
+and complex Late Epipalaeolithic (15--11.7 ka) societies based on
+intensive foraging \citep{BarYosef1998, MaherEtAl2012} and eventually
+plant management and pre-domestication cultivation \citep{Harris2007}.
+As they were brought under domestication in the Pre-Pottery Neolithic
+period (11.7--8.5 ka), the world's first agriculture was shaped by the
+ecosystems from which it emerged and was embedded in.
 
@@
 reconstructed the subsistence economies of Late Epipalaeolithic and
-Neolithic sites in great detail \citep{cites}. Together with studies of
-charcoal \citep{cite}, pollen \citep{cite}, soil \citep{cite}, and a
-variety of palaeoclimate archives \citep{JonesEtAl2019}, they also tell
-us much about the environments surrounding these settlements. However,
-each of these sources of evidence is subject to the wide variety of
-taphonomic and recovery biases that are inherent in any direct record of
-the past. They are also, by definition, records of the (human)
-environment at particular times and places. Interpolating these
-snapshots to give a holistic picture of the regional ecologies is not
-straightforward -- to date, it has tended to rely on non-explicit,
-inductive modelling. The majority are also filtered through human
-action, producing a mixed single that makes it difficult to disentangle
-anthropic effects from the background of environmental change in this
-period of rapid climatic alteration.
+Neolithic sites in great detail. Together with studies of charcoal,
+pollen, soil, and a variety of palaeoclimate archives
+\citep{JonesEtAl2019}, they also tell us much about the environments
+surrounding these settlements. However, each of these sources of
+evidence is subject to the wide variety of taphonomic and recovery
+biases that are inherent in any direct record of the past. They are
+also, by definition, records of the (human) environment at particular
+times and places. Interpolating these snapshots to give a holistic
+picture of the regional ecologies is not straightforward -- to date, it
+has tended to rely on non-explicit, inductive modelling. The majority
+are also filtered through human action, producing a mixed single that
+makes it difficult to disentangle anthropic effects from the background
+of environmental change in this period of rapid climatic alteration.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  The transition to agriculture in West Asia was\ldots{}
-\end{itemize}
-
-\subsection{Biogeography and agricultural
-origins}\label{biogeography-and-agricultural-origins}
-
-\begin{itemize}
-\tightlist
-\item
-  Has always been important in study of agricultural origins
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Historically: Vavilov, Pumpelly \& Childe
-  \item
-    Genetic studies tell us origin points, but not ranges
-  \end{itemize}
-\item
-  Important to e.g.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Distinguish environmental from potentially anthropogenic change
-    \citep{MartinEtAl2016, MartinEtAl2025}
-  \item
-    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
-  \end{itemize}
-\item
-  Epipal./Neo. plant-based economies were diverse
-
-  \begin{itemize}
-  \tightlist
-  \item
-    More than the ``founder crops'';
-  \item
-    More than food
-  \item
-    (In archaebot., not all intentionally collected)
-  \item
-    Geographically and temporally diverse
-  \item
-    \ldots so we model lots of species!
-  \end{itemize}
-\item
-  Regional ecological reconstructions generally rely on the `expert
-  interpolation' (or what do they call it with isoscapes?) method
-
-  \begin{itemize}
-  \tightlist
-  \item
-    See CSEAS (AEA-prep) presentation
-  \item
-    Figure: comparisons
-  \end{itemize}
-\end{itemize}
-
-\subsection{Ecological niche modelling in
-archaeology}\label{ecological-niche-modelling-in-archaeology}
-
 Ecological niche modelling or species distribution modelling is widely
@@
 predict the range of the organism in question either in the same or a
-different environment. \citet{CITE} suggests reserving the term `species
-distribution modelling' for when the method is used to recover the
-verifiable range of a species in a real and existing environment, and
-using `ecological niche modelling' as the broader term covering
-hypothetical or predictive applications -- a convention we follow here
-when referring to predictive or `hindcast' models of past ranges. Within
-this overarching framework, ecological niche modelling encompasses a
-wide range of applications and a variety of potential environmental
-predictors, modelling approaches, and methodologies, which we will not
-attempt to review here.
+different environment. \citet{TownsendPetersonSoberon2012} suggests
+reserving the term `species distribution modelling' for when the method
+is used to recover the verifiable range of a species in a real and
+existing environment, and using `ecological niche modelling' as the
+broader term covering hypothetical or predictive applications -- a
+convention we follow here when referring to predictive or `hindcast'
+models of past ranges. Within this overarching framework, ecological
+niche modelling encompasses a wide range of applications and a variety
+of potential environmental predictors, modelling approaches, and
+methodologies, which we will not attempt to review here.
 
@@
 the spread of crops has been another significant archaeological
-application \citep[e.g.][]{CremaEtAl}, though not as yet applied to West
-Asia.
+application \citep[e.g.][]{KrzyzanskaEtAl2022, Krzyzanska2023}, though
+not as yet applied to West Asia.
 
@@
 applications, which is perhaps why only a minority of studies use it
-\citep[cf.][]{YaworskyEtAl2023}. In this study, we are able to take
-advantage of the increasing availability of high resolution, global
-palaeoclimate data derived from simulation experiments with general
-circulation models of climate
+\citep[cf.][]{KrzyzanskaEtAl2022, YaworskyEtAl2023}. In this study, we
+are able to take advantage of the increasing availability of high
+resolution, global palaeoclimate data derived from simulation
+experiments with general circulation models of climate
 \citep{BrownEtAl2018, BrownEtAl2020, KargerEtAl2023}.
@@
 archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
-\citep{ORIGINS}, and COMPAG \citep{COMPAG}. We did not attempt to
-distinguish between the source of the remains
+\citep{ORIGINS}, and COMPAG
+\citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
+on][]{ColledgeEtAl2004}; \citealp{ShennanConolly2007}}. We did not
+attempt to distinguish between the source of the remains
 \citep[cf.][]{WallaceEtAl2018}; archaeobotanical assemblages are subject
@@
 
-Predictor data was transformed to common equal-area projection and
-resolution of 5 km.
-
 For hindcasting, we used reconstructed bioclimatic data for three key
@@
 predictors that are independent of climate change.
+
+For training, test predictions, and archaeological predictions predictor
+data was left in its native projection and resolution. For hindcast
+palaeodistributions, it was transformed to common equal-area projection
+and resolution of 5 km.
 
@@
 appendix.
-
-\begin{verbatim}
-Joining with `by = join_by(taxon)`
-\end{verbatim}
 
@@
 of this region as developing early intensive foraging economies
-\citep[the Natufian culture,][]{cite} and as a centre of origin of
-agriculture \citep{cite}. Within the Levant, many species show moderate
-reductions in ranges over the Pleistocene/Holocence boundary, retreating
-from the Badia/transjordan region
-e.g.~\ref{fig-palaeodist-barely-levant}.
+\citep[the Natufian culture,][]{BarYosef1998} and as a centre of origin
+of agriculture \citep{Zeder2011a}. Within the Levant, many species show
+moderate reductions in ranges over the Pleistocene/Holocence boundary,
+retreating from the Badia/transjordan region (e.g.~barley, fig.
+Figure~\ref{fig-palaeodist-barely-levant}).
 
@@
 from this hypothesis and towards a more geographically diverse origin
-\citep{Asouti2006, cites}, but it remains the area with some of the
-earliest clear evidence of domestication
-\citep{ZoharyEtAl2012, KabukcuEtAl2021, UlasEtAl2024}. Comparative
-genetics also points to the Northern Mesopotamia region as the centre of
-diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But since these
-studies are based on modern genomes, if the wild range of these plants
-has, as our modelling suggests, shifted since the Early Holocene, the
-apparent centre of diversity may have shifted with them. Our
-reconstructions are consistent with the late arrival of intensive
+\citep{Asouti2006, FullerEtAl2011, ArranzOtaeguiEtAl2016}, but it
+remains the area with some of the earliest clear evidence of
+domestication \citep{ZoharyEtAl2012, KabukcuEtAl2021, UlasEtAl2024}.
+Comparative genetics also points to the Northern Mesopotamia region as
+the centre of diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But
+since these studies are based on modern genomes, if the wild range of
+these plants has, as our modelling suggests, shifted since the Early
+Holocene, the apparent centre of diversity may have shifted with them.
+Our reconstructions are consistent with the late arrival of intensive
 plant-based foraging economies in this region (cf.~the Natufian of the
@@
 as early in the eastern \emph{Mashriq} as it did in the west
-\citep{Braidwood, GanjDarehGoats, HillyFlanksStuff}. We consider that
-the most likely explanation for this is that our flora does not include
-the species that were most important to plant subsistence in the east.
-Archaeobotanical data on Neolithic sites in the Zagros is limited
-(compared to the Levant in particular) due to a hiatus in field research
-there from the 1980s to early 2010s \citep{cite}. Recent research
-\citep{Amaia} indicates \ldots{}
+\citep{Zeder2024}. We consider that the most likely explanation for this
+is that our flora does not include the species that were most important
+to plant subsistence in the east. Archaeobotanical data on Neolithic
+sites in the Zagros is limited (compared to the Levant in particular)
+due to a hiatus in field research there from the 1980s to early 2010s
+\citep{Zeder2024}. Recent research
+\citep{WhitlamEtAl2018, GonzalezCarreteroEtAl2023} indicates that plant
+subsistence in this region was based on a distinct set of species,
+compared to the Levant and Anatolia.
 
@@
 Holocene \citep{AksuHiscott2022}. Were these area part of the same
-broader `interaction sphere' \citep{cite} that produce Neolithic
-agriculture in West Asia?
+broader `interaction sphere' that produce Neolithic agriculture in West
+Asia?
 
@@
 rye (\emph{Secale cereale}), which is inferred to be primarily Western
-Anatolian Figure~\ref{fig-palaeodist-chickpea-rye}. This is perhaps
+Anatolian (Figure~\ref{fig-palaeodist-chickpea-rye}). This is perhaps
 relevant to rye's unusual domestication history, as a crop of West Asian
@@
 concurrent with it being brought into cultivation
-Figure~\ref{fig-palaeodist-barely-levant}. It also sees a marked decline
-in the archaeobotanical record from the Early PPNA/Early PPNB (where it
-was amongst the most common taxa) to the Late PPNB and Late Neolithic
-\citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia atlantica})
-shows similar trends, but it is less certain that this species was
-cultivated in the Neolithic. Conversely, the various wild wheat species
-native to West Asia show almost no response to Pleistocene/Holocene
-climate change, even within the Levant, and in the archaeobotanical
-record wheat displays the opposite trend to barley and pistachio --
-becoming gradually more abundant through the course of the Neolithic and
-dominant by its end. We hypothesise that climate-linked range shifts
-were a factor driving these changes in the apparent economic important
-of different crops: the decreasing availability of barley and pistachio
-in the terminal Pleistocene may have prompted cultivation as a strategy
-to retain access to them, though in the long run it made them less
-attractive as staple crops compared to the more resilient wheat.
-However, these dynamics are not seen in the majority of crop
-progenitors.
+(Figure~\ref{fig-palaeodist-barely-levant}). It also sees a marked
+decline in the archaeobotanical record from the Early PPNA/Early PPNB
+(where it was amongst the most common taxa) to the Late PPNB and Late
+Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
+atlantica}) shows similar trends, but it is less certain that this
+species was cultivated in the Neolithic. Conversely, the various wild
+wheat species native to West Asia show almost no response to
+Pleistocene/Holocene climate change, even within the Levant, and in the
+archaeobotanical record wheat displays the opposite trend to barley and
+pistachio -- becoming gradually more abundant through the course of the
+Neolithic and dominant by its end. We hypothesise that climate-linked
+range shifts were a factor driving these changes in the apparent
+economic important of different crops: the decreasing availability of
+barley and pistachio in the terminal Pleistocene may have prompted
+cultivation as a strategy to retain access to them, though in the long
+run it made them less attractive as staple crops compared to the more
+resilient wheat. However, these dynamics are not seen in the majority of
+crop progenitors.
 
@@
 
-``All models are wrong'' \citep{cite}, but the performance of our models
-on independent test datasets give confidence in its predictions, which
-are plausible and consistent with broad-scale patterns in the
-archaeological record.
-
-\begin{itemize}
-\tightlist
-\item
-  Results w.r.t. to specific archaeologically-attested species not to
-  promising, but this doesn't mean the models are useless!
-\item
-  Discrepencies suggests several areas for further research and
-  methodological development
-\end{itemize}
-
-Modelling a large number of species using machine learning, the
-substantial occurrence datasets available for the present day, and a
-hindcasting approach to past distributions also represents a significant
-advance in the methodology of palaeoecological niche modelling. This
-approach is enabled by the availability of high quality, global open
-datasets in ecology \citep{GBIF2025, GBIFSecretariat2023}, earth science
-\citep{SRTM}, and climatology
-\citep[@][]{KargerEtAl2017, BrownEtAl2018}. Unfortunately in relation to
+``All models are wrong'' \citep{Box1976}, but the performance of our
+models on independent test datasets give confidence in its predictions,
+which are plausible and consistent with broad-scale patterns in the
+archaeological record. Modelling a large number of species using machine
+learning, the substantial occurrence datasets available for the present
+day, and a hindcasting approach to past distributions also represents a
+significant advance in the methodology of palaeoecological niche
+modelling. This approach is enabled by the availability of high quality,
+global open datasets in ecology \citep{GBIF2025, GBIFSecretariat2023},
+earth science \citep{SRTM}, and climatology
+\citep{KargerEtAl2017, BrownEtAl2018}. Unfortunately in relation to
 these fields open data in archaeology lags conspiciously behind. Though
-we have benefitted from the relatively good state of publication of
-archaeobotanical compilations in our region of study
-\citep{ADEMNES, COMPAG, ORIGINS}, further development of open,
-comprehensive and up-to-date `backbone' datasets on site locations and
-chronologies is needed to advance archaeoecological modelling to the
-same level.
+we have benefitted from the relatively long tradition of compiling
+archaeobotanical data in our region of study
+\citep{ColledgeEtAl2004, ShennanConolly2007, ADEMNES, LucasFuller2018, FullerEtAl2018, WallaceEtAl2018, ORIGINS},
+further development of open, comprehensive and up-to-date `backbone'
+datasets on site locations and chronologies is needed to advance
+archaeoecological modelling to the same level.
 

### 2025-01-10 analysis/appendix.qmd
Add Zenodo links

@@
   models, as well as the full predicted palaeodistributions in raster format,
-  are achived with Zenodo at <https://doi.org/> <!-- TODO insert Zenodo DOI -->
+  are achived with Zenodo at <https://doi.org/10.5281/zenodo.14629984>.
 format:

### 2025-03-06 analysis/paper.qmd
Add Zenodo links
Resolve some TODOs

@@
 * [x] Stop modelling
-- [ ] First draft:
-  - [x] Introduction
-  - [ ] Background
-    - [ ] Biogeography
-    - [x] ENM
-  - [x] Methods & Materials
-  - [x] Results
-  - [x] Discussion
-  - [x] Conclusion
-- [ ] Figures
-  - [x] fig-region
-  - [x] fig-occ
-  - [x] tbl-results-summary
-  - [ ] tbl-predictors
-  - [x] palaeodist figs
+- [x] First draft:
+- [x] Figures
+- [ ] Biogeography background section
+- [ ] Revise discussion of archaeo verification
 - [ ] Outstanding TODOs
@@
 - [ ] Copyedit
-  - [ ] Replace "see below" etc. with section links
 - [x] Appendix with all hindcast predictions (HTML?)
 - [ ] Clean up package code
+- [ ] Zenodo release
 - [ ] Final proofread
@@
 Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
-<!-- TODO which we can use to ... --->
+This means that the ancient data can be reserved for assessing the model's ability to 'hindcast' past conditions.
+In this sense, discrepencies between the two records are perhaps the most interesting result, as they indicate processes affecting one or both records that are not fully accounted for and therefore generate new questions.
 Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple past climatologies.
@@
 # Background
-<!--
+
 * The transition to agriculture in West Asia was...
@@
 ## Ecological niche modelling in archaeology
--->
+
 Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
@@
 <!-- TODO: rewrite -->
-Occurrence data only tells us where a species is present;
-there is rarely definitive information on where the species is *not* found.
-We therefore need to generate random background points or "pseudo-absences" to feed to the model.
-There are several ways to do this. 
-We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) random sample of points, weighted equally against the presences in the regression.
-@ValaviEtAl2022 also recommend using a large background sample for random forest models.
+Random Forest is a presence–absence approach to niche modelling and therefore requires not just data on where a species is present, but where it is definitely not present.
+However 'absence' data is rarely available because it requires exhaustive survey.
+In practice, most applications of niche modelling are 'presence-only' and, where absence data is required (as for Random Forest), it is supplied as a random background sample of 'pseudo-absence' points.
+The purpose of this sample is to inform the model about the nature of the underlying environment. 
+The stochastic generation process means that some 'pseudo-absence' points will overlap or fall close to presences, so ensuring the model is not overly influenced by background samples is critical to its predictive importance [@ValaviEtAl2022].
+Here we follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) uniform sample of points from across the land area of the study region.
+These points are then weighted equally against the presences in the regression to produce a 'balanced Random Forest' [@ValaviEtAl2022].
 
@@
 `r n_reduced_range` of `r nrow(flora)` species have reduced ranges; `r n_reduced_range_10` of more than 10% or more.
-Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further below), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
+Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further in @sec-discuss-hindcasting), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
 The average range of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` during the Younger Dryas (i.e.
@@
 *Lathyrus aphaca* and *L. sativus* (both marginally edible legumes);
-*Triticum aestivum compactum* (in the N. Levant), *T. monococcum aegilopoides*, *T. durum*, and *Triticum urartu* (but not the other wheat progenitor, *T. turgidum dicoccum* – see below);
+*Triticum aestivum compactum* (in the N. Levant), *T. monococcum aegilopoides*, *T. durum*, and *Triticum urartu* (but not the other wheat progenitor, *T. turgidum dicoccum* – see @sec-discuss-crops);
 *Aegilops speltoides*, but not *Aegilops tauschii* (goatgrasses);
@@
 
-## Biogeography of crop progenitors
+## Biogeography of crop progenitors {#sec-discuss-crops}
 
@@
 
-## Hindcast models do not predict archaeobotanical composition
+## Hindcast models do not predict archaeobotanical composition {#sec-discuss-hindcasting}
 
@@
 "All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
-<!--
+
 * Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
-* Discrepencies suggests several areas for further research and methodological development
- --> 
+* Discrepancies suggests several areas for further research and methodological development
+
 Modelling a large number of species using machine learning, the substantial occurrence datasets available for the present day, and a hindcasting approach to past distributions also represents a significant advance in the methodology of palaeoecological niche modelling.
@@
 Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
-Though we have benefitted from the relatively long tradition of compiling archaeobotanical data in our region of study [@ColledgeEtAl2004; @ShennanConolly2007; @ADEMNES; @LucasFuller2018; @FullerEtAl2018; @WallaceEtAl2018; @ORIGINS],
+Though we have benefited from the relatively long tradition of compiling archaeobotanical data in our region of study [@ColledgeEtAl2004; @ShennanConolly2007; @ADEMNES; @LucasFuller2018; @FullerEtAl2018; @WallaceEtAl2018; @ORIGINS],
 further development of open, comprehensive and up-to-date 'backbone' datasets on site locations and chronologies is needed to advance archaeoecological modelling to the same level.
 
+# Data availability
+
+The data and R code used to produced is archived with Zenodo at <https://doi.org/10.5281/zenodo.14629984>
+
 # References {.appendix}

### 2025-03-06 analysis/paper.tex
Add Zenodo links
Resolve some TODOs

@@
 past ecologies that is independent of the environmental archaeological
-and palaeoclimatic records. Our computational approach is also readily
-scaled up, allowing us to model spatially-explicit palaeodistributions
-for a large number of species, for the whole region, under multiple past
-climatologies.
+and palaeoclimatic records. This means that the ancient data can be
+reserved for assessing the model's ability to `hindcast' past
+conditions. In this sense, discrepencies between the two records are
+perhaps the most interesting result, as they indicate processes
+affecting one or both records that are not fully accounted for and
+therefore generate new questions. Our computational approach is also
+readily scaled up, allowing us to model spatially-explicit
+palaeodistributions for a large number of species, for the whole region,
+under multiple past climatologies.
 
 \section{Background}\label{background}
+
+\begin{itemize}
+\tightlist
+\item
+  The transition to agriculture in West Asia was\ldots{}
+\end{itemize}
+
+\subsection{Biogeography and agricultural
+origins}\label{biogeography-and-agricultural-origins}
+
+\begin{itemize}
+\tightlist
+\item
+  Has always been important in study of agricultural origins
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Historically: Vavilov, Pumpelly \& Childe
+  \item
+    Genetic studies tell us origin points, but not ranges
+  \end{itemize}
+\item
+  Important to e.g.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Distinguish environmental from potentially anthropogenic change
+    \citep{MartinEtAl2016, MartinEtAl2025}
+  \item
+    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
+  \end{itemize}
+\item
+  Epipal./Neo. plant-based economies were diverse
+
+  \begin{itemize}
+  \tightlist
+  \item
+    More than the ``founder crops'';
+  \item
+    More than food
+  \item
+    (In archaebot., not all intentionally collected)
+  \item
+    Geographically and temporally diverse
+  \item
+    \ldots so we model lots of species!
+  \end{itemize}
+\item
+  Regional ecological reconstructions generally rely on the `expert
+  interpolation' (or what do they call it with isoscapes?) method
+
+  \begin{itemize}
+  \tightlist
+  \item
+    See CSEAS (AEA-prep) presentation
+  \item
+    Figure: comparisons
+  \end{itemize}
+\end{itemize}
+
+\subsection{Ecological niche modelling in
+archaeology}\label{ecological-niche-modelling-in-archaeology}
 
@@
 
-Occurrence data only tells us where a species is present; there is
-rarely definitive information on where the species is \emph{not} found.
-We therefore need to generate random background points or
-``pseudo-absences'' to feed to the model. There are several ways to do
-this. We follow the advice of \citet{BarbetMassinEtAl2012} for
-regression-based species distribution models and use a large (≈10000)
-random sample of points, weighted equally against the presences in the
-regression. \citet{ValaviEtAl2022} also recommend using a large
-background sample for random forest models.
+Random Forest is a presence--absence approach to niche modelling and
+therefore requires not just data on where a species is present, but
+where it is definitely not present. However `absence' data is rarely
+available because it requires exhaustive survey. In practice, most
+applications of niche modelling are `presence-only' and, where absence
+data is required (as for Random Forest), it is supplied as a random
+background sample of `pseudo-absence' points. The purpose of this sample
+is to inform the model about the nature of the underlying environment.
+The stochastic generation process means that some `pseudo-absence'
+points will overlap or fall close to presences, so ensuring the model is
+not overly influenced by background samples is critical to its
+predictive importance \citep{ValaviEtAl2022}. Here we follow the advice
+of \citet{BarbetMassinEtAl2012} for regression-based species
+distribution models and use a large (≈10000) uniform sample of points
+from across the land area of the study region. These points are then
+weighted equally against the presences in the regression to produce a
+`balanced Random Forest' \citep{ValaviEtAl2022}.
 
@@
 magnitude of the reduction from prehistory to the present likely
-reflects a degree of overfitting in the model (discussed further below),
-fluctuations in modelled range size between the Bølling-Allerød
-(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
-(11.7--8.3 ka) are more directly comparable
-(Figure~\ref{fig-pred-range-density}). The average range of modelled
-species was 16\% in the Early Holocene compared to the Bølling--Allerød,
-and 25\% during the Younger Dryas (i.e.~ranges recovered slightly
-between the Younger Dryas and Early Holocene). This perhaps indicates
-that although this period is considered one of climatatic amelioration
-globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
-may have supported more extensive plant-based economies in West Asia
-specifically.
+reflects a degree of overfitting in the model (discussed further in
+Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled range
+size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
+(12.9--11.7 ka), and Early Holocene (11.7--8.3 ka) are more directly
+comparable (Figure~\ref{fig-pred-range-density}). The average range of
+modelled species was 16\% in the Early Holocene compared to the
+Bølling--Allerød, and 25\% during the Younger Dryas (i.e.~ranges
+recovered slightly between the Younger Dryas and Early Holocene). This
+perhaps indicates that although this period is considered one of
+climatatic amelioration globally \citep{JonesEtAl2019}, the colder
+conditions of the Pleistocene may have supported more extensive
+plant-based economies in West Asia specifically.
 
@@
 urartu} (but not the other wheat progenitor, \emph{T. turgidum dicoccum}
--- see below); \emph{Aegilops speltoides}, but not \emph{Aegilops
-tauschii} (goatgrasses); and \emph{Vicia} spp. (vetches), including
-\emph{Vicia faba} (broad beans). most of Anatolia, Northern Mesopotamia,
-and the Zagros Mountains in particular disappear from the predicted
-ranges of these species, leaving the Levant and to a lesser extent the
-Aegean and Cyprus as refugia.
+-- see Section~\ref{sec-discuss-crops}); \emph{Aegilops speltoides}, but
+not \emph{Aegilops tauschii} (goatgrasses); and \emph{Vicia} spp.
+(vetches), including \emph{Vicia faba} (broad beans). most of Anatolia,
+Northern Mesopotamia, and the Zagros Mountains in particular disappear
+from the predicted ranges of these species, leaving the Levant and to a
+lesser extent the Aegean and Cyprus as refugia.
 
@@
 
-\subsection{Biogeography of crop
-progenitors}\label{biogeography-of-crop-progenitors}
+\subsection{Biogeography of crop progenitors}\label{sec-discuss-crops}
 
@@
 \subsection{Hindcast models do not predict archaeobotanical
-composition}\label{hindcast-models-do-not-predict-archaeobotanical-composition}
+composition}\label{sec-discuss-hindcasting}
 
@@
 which are plausible and consistent with broad-scale patterns in the
-archaeological record. Modelling a large number of species using machine
-learning, the substantial occurrence datasets available for the present
-day, and a hindcasting approach to past distributions also represents a
-significant advance in the methodology of palaeoecological niche
-modelling. This approach is enabled by the availability of high quality,
-global open datasets in ecology \citep{GBIF2025, GBIFSecretariat2023},
-earth science \citep{SRTM}, and climatology
-\citep{KargerEtAl2017, BrownEtAl2018}. Unfortunately in relation to
-these fields open data in archaeology lags conspiciously behind. Though
-we have benefitted from the relatively long tradition of compiling
-archaeobotanical data in our region of study
+archaeological record.
+
+\begin{itemize}
+\tightlist
+\item
+  Results w.r.t. to specific archaeologically-attested species not to
+  promising, but this doesn't mean the models are useless!
+\item
+  Discrepancies suggests several areas for further research and
+  methodological development
+\end{itemize}
+
+Modelling a large number of species using machine learning, the
+substantial occurrence datasets available for the present day, and a
+hindcasting approach to past distributions also represents a significant
+advance in the methodology of palaeoecological niche modelling. This
+approach is enabled by the availability of high quality, global open
+datasets in ecology \citep{GBIF2025, GBIFSecretariat2023}, earth science
+\citep{SRTM}, and climatology \citep{KargerEtAl2017, BrownEtAl2018}.
+Unfortunately in relation to these fields open data in archaeology lags
+conspiciously behind. Though we have benefited from the relatively long
+tradition of compiling archaeobotanical data in our region of study
 \citep{ColledgeEtAl2004, ShennanConolly2007, ADEMNES, LucasFuller2018, FullerEtAl2018, WallaceEtAl2018, ORIGINS},
@@
 
+\section{Data availability}\label{data-availability}
+
+The data and R code used to produced is archived with Zenodo at
+\url{https://doi.org/10.5281/zenodo.14629984}
+
 \renewcommand\refname{References}

### 2025-03-31 analysis/paper.qmd
Add background text from Amaia

@@
 
-# Background
-
-* The transition to agriculture in West Asia was...
+# Background {#sec-bg}
+
+The transition to agriculture represents one of the most fundamental changes in human history.
+West Asia is one of the regions where this process has been studied in the most detail:
+decades of research have traced the gradual development of a Neolithic way of life and the changes that occurred in the plant species and their geographical distribution as a result.
+Although archaeobotanical assemblages can be biased due to issues of preservation, sampling, recovery techniques, and lab procedures [Dennel 1976, Hastorf and Popper 1988] —and although they include not just food remains but plant resources that were used for other purposes or arrived at the site accidentally [@HastorfPopper1988]—large-scale studies have still revealed coherent patterns in the exploitation of plants over time [Colledge et al.
+2004, Arranz-Otaegui et al.
+2016].
+
+The possibility of an abrupt, geographically-constrained process of plant domestication was proposed in the 1990s [@HillmanDavies1990; @HillmanDavies1992; @HeunEtAl1997; @OzkanEtAl2002] and developed as an explanatory model in the 2000s [@LevYadunEtAl2000; @GopherEtAl2001; @AbboEtAl2005].
+As part of this model, some authors [Lev-Yadun, Gopher and Abbo 2000, Gopher, Abbo and Lev-Yadun 2001, Abbo, Lev-Yadun and Gopher 2010, 2012] argued that eight plant species, collectively referred to as 'founder crops' or the 'Neolithic crop package' [Zohary and Hopf 1988] were selected and domesticated once, without any phase of pre-domestication cultivation [Abbo, Lev-Yadun and Gopher 2011, p.
+177)].
+This process could have been rapid under strong artificial selection [Hillman and Davies 1990, 1992] and may have occurred in a single region or 'core area' – generally located in southeast Turkey [Ladizinsky and Adler 1976, Heun et al.
+1997, Özkan et al.
+2002, 2005, Mori et al.
+2003, Luo et al.
+2007].
+From this single point of origin, it was supposed that domesticated or semi-domesticated plants radiated outwards to other regions [Abbo et al. 2006, Kilian et al. 2007, Özkan et al. 2011].
+
+The 'short gestation' paradigm was challenged by others [Helbaek 1969, Harris 1989, Kislev 1989, Colledge 2001, Weiss et al. 2004, Willcox et al. 2008, Fuller et al. 2018]. 
+@Helbaek1966 argued that before the appearance of domesticated plants, a phase of cultivation of wild seeds must have taken place. 
+The existence of a phase of cultivation of morphologically wild cereals or 'pre-domestication cultivation' was identified in the archaeological record through the study of plant domestication traits such as grain size, shattering v.
+non-shattering rachises.
+The archaeobotanical evidence shows that during the Pre-Pottery Neolithic A (PPNA) cereals exhibited sizes similar to those recorded in domestic species, but their dispersal mechanism was still the same as the one present in morphologically-wild species [i.e.
+shattering, see Helbaek 1966, Kislev 1989, Hillman et al.
+2000, Colledge 2001, Willcox et al.
+2008].
+This evidence suggested that wild cereal stands could have been cultivated for as much as a thousand years before non-shattering domestic forms became prevalent in the archaeological record [Tanno and Willcox 2006, 2012, Arranz-Otaegui et al.
+2016)].
+Additional archaeobotanical [Colledge 2001, Willcox et al.
+2008, 2009, Riehl et al.
+2013, Arranz-Otaegui et al.
+2016, Weide et al.
+2018, Douché and Willcox 2018, Whitlam et al.
+2018] and genetic data [Badr et al.
+2000, Molina-Cano et al.
+2005, Kilian et al.
+2007, Özkan et al.
+2011, Lob et al.
+2023 VHA] in recent years has further challenged the short-gestation model to explain the origins of plant domestication and agriculture in West Asia.
+
+Similarly, the concept of a limited set of eight 'founder crops' [@ZoharyHopf1988] that were the first species cultivated, domesticated and then spread as the basis of Neolithic agricultural systems, is not supported by the latest evidence.
+Our previous analyses of the composition of available archaeobotanical datasets shows that these crops were of marginal importance during the Epipalaeolithic period [@ArranzOtaeguiEtAl2018] and that Neolithic subsistence did not rely either solely or primarily on the exploitation of these species [Arranz-OtaeguiRoe2023].
+Instead, multiple species of grasses, legumes, fruits, nuts, and other plants were exploited over the Late Pleistocene-Early Holocene transition in southwest Asia.
 
@@
 
-* Has always been important in study of agricultural origins
-  * Historically: Vavilov, Pumpelly & Childe
-  * Genetic studies tell us origin points, but not ranges
-* Important to e.g.
-  * Distinguish environmental from potentially anthropogenic change [@MartinEtAl2016; @MartinEtAl2025]
-  * Reconstruct sequences of domestication [@YeomansEtAl2017]
-* Epipal./Neo. plant-based economies were diverse
-  * More than the "founder crops"; 
-  * More than food
-  * (In archaebot., not all intentionally collected)
-  * Geographically and temporally diverse
-  * ...so we model lots of species!
-* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
-  * See CSEAS (AEA-prep) presentation
-  * Figure: comparisons
+The study of the natural distribution of the progenitors of domesticated crops has been a central part of discussions on the origins of agriculture and plant domestication from the beginning.
+@Humboldt1807 acknowledged the importance of the natural distribution of wild species to explain the origin and domestication of crops like spelt and rye.
+@Candolle1820 integrated the study of plant ecology and biogeography and influenced @Darwin1859, who later reflected in detail about the geographical distribution of plants and species diversity. 
+At that time, there was intense debate about whether there were single or multiple "centres of creation" of species. 
+Researchers aimed to evaluate whether plant and animal species emerged in the same locations where they were currently distributed.
+
+After Darwin, this early interest in crop origins evolved into more specific discussions about the "centres of plant domestication".
+Vavilov [@Vavilov1926; @Vavilov1950] can be considered one of the first authors that sought to determine the number of regions in which plants had been independently domesticated. 
+His main method was 'differential phytogeography' [@Vavilov1950]: 
+he classified the variation within a crop and established the regions of maximum diversity, to locate the geographic regions in which crops originated. 
+Using this method, Vavilov suggested that there were at least eight centres of origin. 
+However, his work was later criticised by @Harlan1971, who argued that 'centres of origin' and 'centres of diversity' had to be separated. 
+For Harlan, a 'centre' was as an "area in which things originate and out of which things are dispersed" [@Harlan1971, p.
+468], and he suggested that three main centres of origin of domesticated crops existed.
+He further indicated that Vavilov's approach to the question was simplistic and that more data proxies had to be considered (e.g.
+archaeology, history, geology…), an approach more in the tradition of @Candolle1909.
+Indeed, the inclusion of archaeobotany and genetics in the last decades, together with the study of wild relative distributions has been fundamental in characterising the origins of agriculture [@FullerColledge2008].
+As a result of modern interdisciplinary studies, the number of recognised centres of plant domestication has increased considerably, from the three centres suggested by Harlan in 1971 to the six to eight centres argued for in the 1990s [@Smith1995] and up to as much as 24 potential centres reported in 2009 [@PuruggananFuller2009; see also @Fuller2010].
+
+Biogeographic research into the centres of origin and/or domestication of crops has directly informed broader understanding of the process of agricultural origins.
+In particular, Childe's *Man Makes Himself* [-@Childe1936] both set the agenda for much of the archaeological work that took place between the end of the Second World War and the 80s [e.g.
+@BraidwoodBraidwood1950; @Kenyon1952; @Kenyon1960; @Braidwood1981; @HoleFlannery1968; @Moore1975] and stimulated theoretical discussions on the subject [e.g.
+@Sauer1952; @Harris1969; @Harris1977; @Harris1989; @Flannery1973; @Hawkes1969].
+Childe (unlike for example @Pumpelly before him) correctly located the centre of origin of European agriculture in West Asia, but not based on the region's prehistoric archaeological record, which at that point was barely known.
+Instead he was guided to the region by biogeographic work by Vavilov and @PeakeFleure1927;
+only later was this prediction validated by archaeological work on the Epipalaeolithic and Neolithic of Palestine [@Boyd2018].
+
+Contemporary research on crop origins was pioneered by Harlan and Zohary, who compared the current distribution of the wild progenitors of domesticated plants in southwest Asia [@HarlanZohary1966; @Zohary1969; @Zohary1973; @ZoharyHopf1973] to the rapidly-expanding archaeobotanical record [@Harlan1971; @Harlan1977; see also @ZoharyHopf1988; @HarlanZohary1966].
+They both were interested in evaluating which were the wild ancestors of domesticated crops and studying their natural distribution to understand their domestication process [Zohary 1969, Zohary 1973, 1992, Zohary and Spiegel-Roy 1975].
+Indeed, the natural distribution of the wild relatives of domestic plant species was later used as a criterion to infer 'pre-domestic cultivation' in the archaeological record. 
+For example, the presence of seeds of chickpea at Jericho [Hopf 1983, see also Hopf 1969] led the author to interpret the remains as cultivars, as the natural distribution of the wild form of chickpea was located further to the north.
+The same rationale was applied to the einkorn remains found at several Pre-Pottery Neolithic sites in the southern Levant [Hopf 1969, Colledge 2001], as the wild progenitors of this species was thought to be restricted to the northern Levantine area [Heun et al.
+1997, 2009, Zohary and Hopf 2000].
+The same idea—presence of plants outside their natural range—has been repeated in the literature more recently by several other authors [Tanno and Willcox 2006, Willcox et al.
+2008, Hilman et al.
+2000].
+
+Despite the importance of biogeography in the development and validation of hypotheses regarding the origins of agriculture, there have been few studies of the wild range of specific crop progenitors or other relevant plant species [cf.
+for domestic animals, e.g.
+@YeomansEtAl2017].
+Observations regarding translocation or range expansion must therefore rely on a relatively rough and ahistoric notion of a species' 'natural distribution' – that is, one based primarily on contemporary or recent-historic occurrences.
+Yet we know there has been considerable climatic and environmental change in West Asia since the terminal Pleistocene [@JonesEtAl2019], so it is very unlikely that these ranges were in fact static.
+Reconstructions of broader environments have attempted to trace their fluctuations through time, either for the entire region [e.g.
+@VanZeistBottema1991; Hillman2000] or parts of it [e.g.
+@Cordova2007], but these are at the level of the vegetation zone rather than individual species.
+They also invariably rely on what might be called 'expert interpolation' (where the author composes a map based on his or her own knowledge of the relevant data) rather than an explicit modelling process.
+This makes it difficult, if not impossible, for users of such reconstructions to understand exactly how they were derived or what could explain, for example, the significant discrepencies between the predictions of different experts.
+
+<!-- TODO: figure highlighting discrepencies between 'expert interpolation' -->
 
@@
 The aim of our study was to model the biogeography of species relevant to human subsistence economies in West Asia (excluding the Southern Arabian peninsula, see @fig-region) during the archaeological Late Epipalaeolithic (15–11.7 ka) and Pre-Pottery Neolithic (11.7–8.3 ka) periods.
+Based on current understandings, we assume that plant-based subsistence during this period was broad, geographically- and temporally-varied, and reflects a gradual, geographically decentred, and nonlinear tranistion to greater reliance on cultivars (i.e.
+agriculture, see @sec-bg).
 Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
@@
 
-<!-- TODO: rewrite -->
 Random Forest is a presence–absence approach to niche modelling and therefore requires not just data on where a species is present, but where it is definitely not present.

### 2025-03-31 analysis/paper.qmd
Incorporate Amaia's suggestions

@@
   This paper presents the first continuous, spatially-explicit reconstructions 
-  of the palaeodistributions of 82<!-- TODO update --> plant species found regularly in association 
+  of the palaeodistributions of 82 plant species found regularly in association 
   with early agricultural archaeological sites in West Asia, including the 
@@
 The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
-West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@Diamond2002].
-These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies based on intensive foraging [@BarYosef1998; @MaherEtAl2012] and eventually plant management and pre-domestication cultivation [@Harris2007].
-As they were brought under domestication in the Pre-Pottery Neolithic period (11.7–8.5 ka), the world's first agriculture was shaped by the ecosystems from which it emerged and was embedded in.
+West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@HarlanZohary1966; @ZoharyHopf2000; @Diamond2002].
+These species supported uniquely dense and complex Late Epipalaeolithic [15–11.7 ka, @BarYosef1998; @MaherEtAl2012] societies based on foraging [@HillmanEtAl1989; @Colledge2001; @WeissEtAl2004] and eventually plant management and pre-domestication cultivation [@Colledge2001; @WeissEtAl2006; @Harris2007; @WillcoxEtAl2008].
+The first agro-ecosystems emerged as these plants were domesticated in the Pre-Pottery Neolithic period (11.7–8.5 ka), shaped by the broader ecosystem in which it was embedded.
 
 Decades of research in archaeobotany and zooarchaeology have reconstructed the subsistence economies of Late Epipalaeolithic and Neolithic sites in great detail.
-Together with studies of charcoal, pollen, soil, and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
+Together with studies of other environmental archaeological records and a variety of palaeoclimate archives [@JonesEtAl2019], they also tell us much about the environments surrounding these settlements.
 However, each of these sources of evidence is subject to the wide variety of taphonomic and recovery biases that are inherent in any direct record of the past.
@@
 `r n_reduced_range` of `r nrow(flora)` species have reduced ranges; `r n_reduced_range_10` of more than 10% or more.
-Though the magnitude of the reduction from prehistory to the present likely reflects a degree of overfitting in the model (discussed further in @sec-discuss-hindcasting), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
+Though the magnitude of the change in range size from prehistory to the present likely reflects a degree of overfitting in the model (discussed further in @sec-discuss-hindcasting), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
 The average range of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` during the Younger Dryas (i.e.
@@
 and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
-most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
+Most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
 leaving the Levant and to a lesser extent the Aegean and Cyprus as refugia.
@@
 
-Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture.
-Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @FullerEtAl2011; @ArranzOtaeguiEtAl2016], but it remains the area with some of the earliest clear evidence of domestication [@ZoharyEtAl2012; @KabukcuEtAl2021; @UlasEtAl2024].
-Comparative genetics also points to the Northern Mesopotamia region as the centre of diversity of many crops [e.g. @HaasEtAl2019].
-But since these studies are based on modern genomes, if the wild range of these plants has, as our modelling suggests, shifted since the Early Holocene, the apparent centre of diversity may have shifted with them.
-Our reconstructions are consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant),
-and more broadly there need not be a link between the core *wild* range of a plant and the core area of its domestication. 
-A scenario in which cultivation emerged at the edges of the ranges of valuable plant resources—as a means of extend their natural niche—is also plausible.
-
-The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that domestication took place just as early in the eastern *Mashriq* as it did in the west [@Zeder2024].
+Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture and plant domestication.
+Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @FullerEtAl2011; @ArranzOtaeguiEtAl2016], 
+and our reconstructions are also consistent with the late arrival of intensive plant-based foraging economies in this region (cf. the Natufian of the Levant).
+
+The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that animal domestication took place just as early in the eastern *Mashriq* as it did in the west [@Zeder2024].
 We consider that the most likely explanation for this is that our flora does not include the species that were most important to plant subsistence in the east.
 Archaeobotanical data on Neolithic sites in the Zagros is limited (compared to the Levant in particular) due to a hiatus in field research there from the 1980s to early 2010s [@Zeder2024].
-Recent research [@WhitlamEtAl2018; @GonzalezCarreteroEtAl2023] indicates that plant subsistence in this region was based on a distinct set of species, compared to the Levant and Anatolia.
+Recent research [@RiehlEtAl2013; @WeideEtAl2017; @WeideEtAl2018; @WhitlamEtAl2018; @GonzalezCarreteroEtAl2023] indicates that plant subsistence in this region was based on a distinct set of species, compared to the Levant and Anatolia.
 
@@
 and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to have two limited ranges centered around the Black Sea Coast of Anatolia and the Palmyra basin.
-In the latter case, neither of these areas are part of the predicted modern distribution of wild emmer (centered around the Caucasus and Northern Mesopotamia), but it would be consistent with archaeological evidence for early cultivation at sites in the Upper Euphrates [@Willcox2024].
+In the latter case, neither of these areas are part of the predicted (by our model) modern distribution of wild emmer, which is centered around the Caucasus and Northern Mesopotamia.
+But it would be consistent with archaeological evidence for early cultivation at sites in the Upper Euphrates [@Willcox2024].
 
@@
 Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
-This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].
+This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2000; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].
 
@@
 It is the only unambiguous crop progenitor with such a restricted range, though pistachio (*Pistacia atlantica*) and clubrush (*Bolboschoenus maritimus*) are similarly constrained to the Mediterranean coast (and North Africa, in the case of pistachio).
+This is despite the fact that they are well-attested in the archaeotanical record from across West Asia.
 
@@
 It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
-Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was cultivated in the Neolithic.
-Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end.
+Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was managed in the Neolithic.
+Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end [@ArranzOtaeguiRoe2023].
 We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
@@
 though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
-However, these dynamics are not seen in the majority of crop progenitors.
+However, these dynamics are not seen in the majority of crop progenitors and may also have involved other factors, such as changes in cultural food preferences.
 
@@
 Bread wheat (*Triticum aestivum*), the most common wheat cultivar today, has a complex ancestry that involves two recent hybridisation events [@LevyFeldman2022]:
-most recently between emmer (*Triticum turgidum dicoccum*) and a goatgrass (*Aegilops tauschii*) c.
+most recently between domestic emmer (*Triticum turgidum dicoccum*) and a goatgrass (*Aegilops tauschii*) c.
 9 ka, and before that, in emmer, between wild red einkorn (*T.
 urartu*) and another goatgrass (*Aegilops speltoides*).
-Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to eachother in the time frame of domestication is a narrow area around the Orontes river in the northern Levant (@fig-palaeodist-bread).
+Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to each other in the time frame of domestication is a narrow area around the Orontes river in the northern Levant (@fig-palaeodist-bread).
 This area is c.
-200 km from the only two sites in our archaeobotanical database where either both species of wheat or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
+200 km from the only two sites in our archaeobotanical database where either both species of wheat (*T.
+urartu* and *T.
+turgidum dicoccum*) or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
 Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is likely to occur in this vicinity.
@@
 
+<!--
 It is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
-<!-- TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
+TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
 
@@
 
+# Acknowledgements
+
+This work was partially funded by the European Research Council (ERC-2021-STG project 101039060, *PalaeOrigins: Tracing the Epipalaeolithic origins of plant management in southwest Asia*).
+
 # References {.appendix}

### 2025-03-31 analysis/lay_summary.qmd
Add lay summary

---
title: "Modelling vegetation around the world's first farms"
subtitle: "Lay summary of 'Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka'"
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
format: pdf
---

We reconstructed the past ranges of 82 plant species that were used by prehistoric societies in West Asia around 10,000 years ago, including the ancestors of the first crops.
Our results suggest that under the cold, arid conditions of that time these were much less widespread than they are now, and that they became even scarcer just before the emergence of agriculture.
The Mediterranean coast of the Levant and, to our surprise, Cyprus and Western Anatolia appear to have been refugia for many economically important plants.
Our data introduces an as-yet unexplored new line of evidence on early plant domestication and has implications for how archaeologists interpret the botanical remains they recover from sites, as for example the recovery of a plant from outside its 'native range' is often taken as evidence that it was being managed or cultivated.
It also gives insights into the broader ecosystems from which the world's first agricultural systems emerged.

Our approach was based on 'ecological niche modelling', a technique widely used by ecologists to model the distribution of species now or predict where they will be in the future (for example, under climate change) but which can also be combined with palaeoclimate data to 'retro-dict' where they were in the past.
In our case, we used palaeoclimate data from the 'CHELSA-TraCE21k' experiment, which took one of the global models used to predict future climate change and essentially turned it backwards.
Combining this with large open datasets of where plants have been observed growing today, in relation to the same set of environmental variables, we were able to model a large number of species which we know are found at early agricultural sites in West Asia.
We also used data on the species observed at archaeological sites to test the predictions of our models, though this had mixed results – something we intend to probe further in future research.

### 2025-03-31 analysis/paper.qmd
Add lay summary

@@
 <!--
-
-* [x] Stop modelling
-- [x] First draft:
-- [x] Figures
-- [ ] Biogeography background section
+TODO before publication:
 - [ ] Revise discussion of archaeo verification
@@
 - [ ] Add missing references
-- [ ] Copyedit
 - [x] Appendix with all hindcast predictions (HTML?)
@@
 - [ ] Final proofread
-
 -->

### 2025-03-31 analysis/lay_summary.qmd
Update bibliography

@@
 
-We reconstructed the past ranges of 82 plant species that were used by prehistoric societies in West Asia around 10,000 years ago, including the ancestors of the first crops.
+We used machine learning to reconstruct the past ranges of 82 plant species that were used by prehistoric societies in West Asia around 10,000 years ago, including the ancestors of the first crops.
 Our results suggest that under the cold, arid conditions of that time these were much less widespread than they are now, and that they became even scarcer just before the emergence of agriculture.

### 2025-04-01 analysis/paper.qmd
Add references

@@
   average of c. 25% from the terminal Pleistocene to the Early Holocene. 
-  Modelled ranges do not reliably predict the observed  occurrence of specific 
-  species at archaeological sites, but are aligned with more general trends in
-  the relative abundance of species.<!--
+  The models performance in predicting the occurrence of specific species at 
+  archaeological sites is highly variable, but in aggregate the predictions are
+  coherent and align with broad-scale trends in the ubiquity and relative 
+  abundance of species in the archaeological record.<!--
   however [...]
@@
 The Pleistocene–Holocene transition in West Asia marked a turning point in global environmental history, as humans brought the first plants under cultivation and began modifying surrounding ecosystems to support their own subsistence.
-West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@HarlanZohary1966; @ZoharyHopf2000; @Diamond2002].
-These species supported uniquely dense and complex Late Epipalaeolithic [15–11.7 ka, @BarYosef1998; @MaherEtAl2012] societies based on foraging [@HillmanEtAl1989; @Colledge2001; @WeissEtAl2004] and eventually plant management and pre-domestication cultivation [@Colledge2001; @WeissEtAl2006; @Harris2007; @WillcoxEtAl2008].
-The first agro-ecosystems emerged as these plants were domesticated in the Pre-Pottery Neolithic period (11.7–8.5 ka), shaped by the broader ecosystem in which it was embedded.
+West Asia is part of the native range of a remarkable number of domesticable plant species, including wild relatives of wheat, barley, peas, lentils, and other crops of global importance [@HarlanZohary1966; @Diamond2002; @ZoharyEtAl2012].
+These species supported uniquely dense and complex Late Epipalaeolithic (15–11.7 ka) societies [@BarYosef1998; @MaherEtAl2012] based on foraging [@HarrisHillman1989; @Colledge2001; @WeissEtAl2004] and eventually plant management and pre-domestication cultivation [@Colledge2001; @WeissEtAl2006; @Harris2007; @WillcoxEtAl2008].
+The first agro-ecosystems emerged as these plants were domesticated in the Pre-Pottery Neolithic period (11.7–8.5 ka) and were shaped by the broader ecosystem in which it was embedded.
 
@@
 decades of research have traced the gradual development of a Neolithic way of life and the changes that occurred in the plant species and their geographical distribution as a result.
-Although archaeobotanical assemblages can be biased due to issues of preservation, sampling, recovery techniques, and lab procedures [Dennel 1976, Hastorf and Popper 1988] —and although they include not just food remains but plant resources that were used for other purposes or arrived at the site accidentally [@HastorfPopper1988]—large-scale studies have still revealed coherent patterns in the exploitation of plants over time [Colledge et al.
-2004, Arranz-Otaegui et al.
-2016].
+Although archaeobotanical assemblages can be biased due to issues of preservation, sampling, recovery techniques, and lab procedures [@Dennel1976; @HastorfPopper1988]—and although they include not just food remains but plant resources that were used for other purposes or arrived at the site accidentally [@HastorfPopper1988]—large-scale studies have still revealed coherent patterns in the exploitation of plants over time [@ColledgeEtAl2004; @ArranzOtaeguiEtAl2016].
 
 The possibility of an abrupt, geographically-constrained process of plant domestication was proposed in the 1990s [@HillmanDavies1990; @HillmanDavies1992; @HeunEtAl1997; @OzkanEtAl2002] and developed as an explanatory model in the 2000s [@LevYadunEtAl2000; @GopherEtAl2001; @AbboEtAl2005].
-As part of this model, some authors [Lev-Yadun, Gopher and Abbo 2000, Gopher, Abbo and Lev-Yadun 2001, Abbo, Lev-Yadun and Gopher 2010, 2012] argued that eight plant species, collectively referred to as 'founder crops' or the 'Neolithic crop package' [Zohary and Hopf 1988] were selected and domesticated once, without any phase of pre-domestication cultivation [Abbo, Lev-Yadun and Gopher 2011, p.
-177)].
-This process could have been rapid under strong artificial selection [Hillman and Davies 1990, 1992] and may have occurred in a single region or 'core area' – generally located in southeast Turkey [Ladizinsky and Adler 1976, Heun et al.
-1997, Özkan et al.
-2002, 2005, Mori et al.
-2003, Luo et al.
-2007].
-From this single point of origin, it was supposed that domesticated or semi-domesticated plants radiated outwards to other regions [Abbo et al. 2006, Kilian et al. 2007, Özkan et al. 2011].
-
-The 'short gestation' paradigm was challenged by others [Helbaek 1969, Harris 1989, Kislev 1989, Colledge 2001, Weiss et al. 2004, Willcox et al. 2008, Fuller et al. 2018]. 
-@Helbaek1966 argued that before the appearance of domesticated plants, a phase of cultivation of wild seeds must have taken place. 
+As part of this model, some authors [@LevYadunEtAl2000; @GopherEtAl2001; @AbboEtAl2010; @AbboEtAl2012] argued that eight plant species, collectively referred to as 'founder crops' or the 'Neolithic crop package' [@ZoharyHopf1988] were selected and domesticated once, without any phase of pre-domestication cultivation [@AbboEtAl2011, p.
+177].
+This process could have been rapid under strong artificial selection [@HillmanDavies1990; @HillmanDavies1992] and may have occurred in a single region or 'core area' – generally located in southeast Turkey [@LadizinskyAdler1976; @HeunEtAl1997; @OzkanEtAl2002; @OzkanEtAl2005; @MoriEtAl2003; @LuoEtAl2007].
+From this single point of origin, it was supposed that domesticated or semi-domesticated plants radiated outwards to other regions [@AbboEtAl2006; @KilianEtAl2007; @OzkanEtAl2011].
+
+The 'short gestation' paradigm was challenged by others [@Helbaek1969; @Harris1989; @Kislev1989; @Colledge2001; @WeissEtAl2004; @WillcoxEtAl2008; @FullerEtAl2018]. 
+Helbæk [in @Kirkbride1966] argued that before the appearance of domesticated plants, a phase of cultivation of wild seeds must have taken place. 
 The existence of a phase of cultivation of morphologically wild cereals or 'pre-domestication cultivation' was identified in the archaeological record through the study of plant domestication traits such as grain size, shattering v.
@@
 The archaeobotanical evidence shows that during the Pre-Pottery Neolithic A (PPNA) cereals exhibited sizes similar to those recorded in domestic species, but their dispersal mechanism was still the same as the one present in morphologically-wild species [i.e.
-shattering, see Helbaek 1966, Kislev 1989, Hillman et al.
-2000, Colledge 2001, Willcox et al.
-2008].
-This evidence suggested that wild cereal stands could have been cultivated for as much as a thousand years before non-shattering domestic forms became prevalent in the archaeological record [Tanno and Willcox 2006, 2012, Arranz-Otaegui et al.
-2016)].
-Additional archaeobotanical [Colledge 2001, Willcox et al.
-2008, 2009, Riehl et al.
-2013, Arranz-Otaegui et al.
-2016, Weide et al.
-2018, Douché and Willcox 2018, Whitlam et al.
-2018] and genetic data [Badr et al.
-2000, Molina-Cano et al.
-2005, Kilian et al.
-2007, Özkan et al.
-2011, Lob et al.
-2023 VHA] in recent years has further challenged the short-gestation model to explain the origins of plant domestication and agriculture in West Asia.
+shattering, see @Kirkbride1966; @Kislev1989; @HillmanEtAl2001; @Colledge2001; @WillcoxEtAl2008].
+This evidence suggested that wild cereal stands could have been cultivated for as much as a thousand years before non-shattering domestic forms became prevalent in the archaeological record [@TannoWillcox2006; @TannoWillcox2012; @ArranzOtaeguiEtAl2016].
+Additional archaeobotanical [@Colledge2001; @WillcoxEtAl2008; @WillcoxEtAl2009; @RiehlEtAl2013; @ArranzOtaeguiEtAl2016; @WeideEtAl2018; @DoucheWillcox2018; @WhitlamEtAl2018] and genetic data [@BadrEtAl2000; @MolinaCanoEtAl2005; @KilianEtAl2007; @OzkanEtAl2011; @IobBotigue2023] in recent years has further challenged the short-gestation model to explain the origins of plant domestication and agriculture in West Asia.
 
 Similarly, the concept of a limited set of eight 'founder crops' [@ZoharyHopf1988] that were the first species cultivated, domesticated and then spread as the basis of Neolithic agricultural systems, is not supported by the latest evidence.
-Our previous analyses of the composition of available archaeobotanical datasets shows that these crops were of marginal importance during the Epipalaeolithic period [@ArranzOtaeguiEtAl2018] and that Neolithic subsistence did not rely either solely or primarily on the exploitation of these species [Arranz-OtaeguiRoe2023].
+Our previous analyses of the composition of available archaeobotanical datasets shows that these crops were of marginal importance during the Epipalaeolithic period [@ArranzOtaeguiEtAl2018] and that Neolithic subsistence did not rely either solely or primarily on the exploitation of these species [@ArranzOtaeguiRoe2023].
 Instead, multiple species of grasses, legumes, fruits, nuts, and other plants were exploited over the Late Pleistocene-Early Holocene transition in southwest Asia.
@@
 @Humboldt1807 acknowledged the importance of the natural distribution of wild species to explain the origin and domestication of crops like spelt and rye.
-@Candolle1820 integrated the study of plant ecology and biogeography and influenced @Darwin1859, who later reflected in detail about the geographical distribution of plants and species diversity. 
+@Candolle1886 integrated the study of plant ecology and biogeography and influenced @Darwin1859, who later reflected in detail about the geographical distribution of plants and species diversity. 
 At that time, there was intense debate about whether there were single or multiple "centres of creation" of species. 
@@
 After Darwin, this early interest in crop origins evolved into more specific discussions about the "centres of plant domestication".
-Vavilov [@Vavilov1926; @Vavilov1950] can be considered one of the first authors that sought to determine the number of regions in which plants had been independently domesticated. 
-His main method was 'differential phytogeography' [@Vavilov1950]: 
+@Vavilov1926 was among the first to seek to determine the number of regions in which plants had been independently domesticated [@Harris1990].
+His main method was 'differential phytogeography':
 he classified the variation within a crop and established the regions of maximum diversity, to locate the geographic regions in which crops originated. 
 Using this method, Vavilov suggested that there were at least eight centres of origin. 
-However, his work was later criticised by @Harlan1971, who argued that 'centres of origin' and 'centres of diversity' had to be separated. 
+His work was later criticised by @Harlan1971, who argued that 'centres of origin' and 'centres of diversity' had to be separated. 
 For Harlan, a 'centre' was as an "area in which things originate and out of which things are dispersed" [@Harlan1971, p.
@@
 He further indicated that Vavilov's approach to the question was simplistic and that more data proxies had to be considered (e.g.
-archaeology, history, geology…), an approach more in the tradition of @Candolle1909.
+archaeology, history, geology), an approach more in the tradition of @Candolle1886.
 Indeed, the inclusion of archaeobotany and genetics in the last decades, together with the study of wild relative distributions has been fundamental in characterising the origins of agriculture [@FullerColledge2008].
@@
 
-Biogeographic research into the centres of origin and/or domestication of crops has directly informed broader understanding of the process of agricultural origins.
-In particular, Childe's *Man Makes Himself* [-@Childe1936] both set the agenda for much of the archaeological work that took place between the end of the Second World War and the 80s [e.g.
-@BraidwoodBraidwood1950; @Kenyon1952; @Kenyon1960; @Braidwood1981; @HoleFlannery1968; @Moore1975] and stimulated theoretical discussions on the subject [e.g.
-@Sauer1952; @Harris1969; @Harris1977; @Harris1989; @Flannery1973; @Hawkes1969].
-Childe (unlike for example @Pumpelly before him) correctly located the centre of origin of European agriculture in West Asia, but not based on the region's prehistoric archaeological record, which at that point was barely known.
+Biogeographic research into the centres of origin and/or domestication of crops has also long informed broader understanding of the process of agricultural origins.
+In *Man Makes Himself*, @Childe1936 correctly located the centre of origin of European agriculture in the 'Fertile Crescent' of West Asia (unlike for example @Pumpelly1908 before him).
+This was not based on the region's prehistoric archaeological record, which at that point was barely known.
 Instead he was guided to the region by biogeographic work by Vavilov and @PeakeFleure1927;
 only later was this prediction validated by archaeological work on the Epipalaeolithic and Neolithic of Palestine [@Boyd2018].
+In subsequent decades, the search for more precise origin zones of specific domestic plants relied on the assumption that "the locus of domestication of a wild plant would presumably be within its area of original distribution in the wild state" [@Butzer1971; paraphrasing @Helbaek1959] – and that this "natural habitat" has not changed significantly over the last 12,000 years [@Butzer1971].
 
 Contemporary research on crop origins was pioneered by Harlan and Zohary, who compared the current distribution of the wild progenitors of domesticated plants in southwest Asia [@HarlanZohary1966; @Zohary1969; @Zohary1973; @ZoharyHopf1973] to the rapidly-expanding archaeobotanical record [@Harlan1971; @Harlan1977; see also @ZoharyHopf1988; @HarlanZohary1966].
-They both were interested in evaluating which were the wild ancestors of domesticated crops and studying their natural distribution to understand their domestication process [Zohary 1969, Zohary 1973, 1992, Zohary and Spiegel-Roy 1975].
+They both were interested in evaluating which were the wild ancestors of domesticated crops and studying their natural distribution to understand their domestication process [@Zohary1969; @Zohary1973; @ZoharySpiegelRoy1975].
 Indeed, the natural distribution of the wild relatives of domestic plant species was later used as a criterion to infer 'pre-domestic cultivation' in the archaeological record. 
-For example, the presence of seeds of chickpea at Jericho [Hopf 1983, see also Hopf 1969] led the author to interpret the remains as cultivars, as the natural distribution of the wild form of chickpea was located further to the north.
-The same rationale was applied to the einkorn remains found at several Pre-Pottery Neolithic sites in the southern Levant [Hopf 1969, Colledge 2001], as the wild progenitors of this species was thought to be restricted to the northern Levantine area [Heun et al.
-1997, 2009, Zohary and Hopf 2000].
-The same idea—presence of plants outside their natural range—has been repeated in the literature more recently by several other authors [Tanno and Willcox 2006, Willcox et al.
-2008, Hilman et al.
-2000].
+For example, the presence of seeds of chickpea at Jericho led @Hopf1986 to interpret the remains as cultivars, as the natural distribution of the wild form of chickpea was located further to the north.
+The same rationale was applied to the einkorn remains found at several Pre-Pottery Neolithic sites in the southern Levant [@Hopf1969; @Colledge2001], as the wild progenitors of this species was thought to be restricted to the northern Levantine area [@HeunEtAl1997; @ZoharyEtAl2012].
+The same idea—presence of plants outside their natural range—has been repeated in the literature more recently by several other authors [@TannoWillcox2006; @WillcoxEtAl2008; @HillmanEtAl2001].
 
@@
 Reconstructions of broader environments have attempted to trace their fluctuations through time, either for the entire region [e.g.
-@VanZeistBottema1991; Hillman2000] or parts of it [e.g.
+@VanZeistBottema1991; Hillman in @MooreEtAl2000] or parts of it [e.g.
 @Cordova2007], but these are at the level of the vegetation zone rather than individual species.
@@
 Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
-This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2000; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].
+This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2001; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].
 
@@
 The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations. 
-Since they do accurately predict the test dataset <!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
+Since they do accurately predict the test dataset<!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
 This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
@@
 "All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
-
-* Results w.r.t. to specific archaeologically-attested species not to promising, but this doesn't mean the models are useless!
-* Discrepancies suggests several areas for further research and methodological development
+Results with regard to specific archaeologically-attested species are not as promising, but this also reflects the incomplete and coarsely temporally-resolved nature of the archaeobotanical verification dataset.
+Discrepencies between the modelled distribution, whether on the broad scale (e.g.
+the restricted geographic range of most species compared to their attestation in the archaeological record), or relating to specific species (i.e.
+false positives and false negatives), suggest several avenues for future investigation.
 

### 2025-04-01 analysis/paper.tex
Add references

@@
 addition, overall ranges shrunk by an average of c.~25\% from the
-terminal Pleistocene to the Early Holocene. Modelled ranges do not
-reliably predict the observed occurrence of specific species at
-archaeological sites, but are aligned with more general trends in the
-relative abundance of species.
+terminal Pleistocene to the Early Holocene. The models performance in
+predicting the occurrence of specific species at archaeological sites is
+highly variable, but in aggregate the predictions are coherent and align
+with broad-scale trends in the ubiquity and relative abundance of
+species in the archaeological record.
 \end{abstract}
@@
 relatives of wheat, barley, peas, lentils, and other crops of global
-importance \citep{Diamond2002}. These species supported uniquely dense
-and complex Late Epipalaeolithic (15--11.7 ka) societies based on
-intensive foraging \citep{BarYosef1998, MaherEtAl2012} and eventually
-plant management and pre-domestication cultivation \citep{Harris2007}.
-As they were brought under domestication in the Pre-Pottery Neolithic
-period (11.7--8.5 ka), the world's first agriculture was shaped by the
-ecosystems from which it emerged and was embedded in.
+importance \citep{HarlanZohary1966, Diamond2002, ZoharyEtAl2012}. These
+species supported uniquely dense and complex Late Epipalaeolithic
+(15--11.7 ka) societies \citep{BarYosef1998, MaherEtAl2012} based on
+foraging \citep{HarrisHillman1989, Colledge2001, WeissEtAl2004} and
+eventually plant management and pre-domestication cultivation
+\citep{Colledge2001, WeissEtAl2006, Harris2007, WillcoxEtAl2008}. The
+first agro-ecosystems emerged as these plants were domesticated in the
+Pre-Pottery Neolithic period (11.7--8.5 ka) and were shaped by the
+broader ecosystem in which it was embedded.
 
@@
 reconstructed the subsistence economies of Late Epipalaeolithic and
-Neolithic sites in great detail. Together with studies of charcoal,
-pollen, soil, and a variety of palaeoclimate archives
-\citep{JonesEtAl2019}, they also tell us much about the environments
-surrounding these settlements. However, each of these sources of
-evidence is subject to the wide variety of taphonomic and recovery
-biases that are inherent in any direct record of the past. They are
-also, by definition, records of the (human) environment at particular
-times and places. Interpolating these snapshots to give a holistic
-picture of the regional ecologies is not straightforward -- to date, it
-has tended to rely on non-explicit, inductive modelling. The majority
-are also filtered through human action, producing a mixed single that
-makes it difficult to disentangle anthropic effects from the background
-of environmental change in this period of rapid climatic alteration.
+Neolithic sites in great detail. Together with studies of other
+environmental archaeological records and a variety of palaeoclimate
+archives \citep{JonesEtAl2019}, they also tell us much about the
+environments surrounding these settlements. However, each of these
+sources of evidence is subject to the wide variety of taphonomic and
+recovery biases that are inherent in any direct record of the past. They
+are also, by definition, records of the (human) environment at
+particular times and places. Interpolating these snapshots to give a
+holistic picture of the regional ecologies is not straightforward -- to
+date, it has tended to rely on non-explicit, inductive modelling. The
+majority are also filtered through human action, producing a mixed
+single that makes it difficult to disentangle anthropic effects from the
+background of environmental change in this period of rapid climatic
+alteration.
 
@@
 
-\section{Background}\label{background}
-
-\begin{itemize}
-\tightlist
-\item
-  The transition to agriculture in West Asia was\ldots{}
-\end{itemize}
+\section{Background}\label{sec-bg}
+
+The transition to agriculture represents one of the most fundamental
+changes in human history. West Asia is one of the regions where this
+process has been studied in the most detail: decades of research have
+traced the gradual development of a Neolithic way of life and the
+changes that occurred in the plant species and their geographical
+distribution as a result. Although archaeobotanical assemblages can be
+biased due to issues of preservation, sampling, recovery techniques, and
+lab procedures \citep{Dennel1976, HastorfPopper1988}---and although they
+include not just food remains but plant resources that were used for
+other purposes or arrived at the site accidentally
+\citep{HastorfPopper1988}---large-scale studies have still revealed
+coherent patterns in the exploitation of plants over time
+\citep{ColledgeEtAl2004, ArranzOtaeguiEtAl2016}.
+
+The possibility of an abrupt, geographically-constrained process of
+plant domestication was proposed in the 1990s
+\citep{HillmanDavies1990, HillmanDavies1992, HeunEtAl1997, OzkanEtAl2002}
+and developed as an explanatory model in the 2000s
+\citep{LevYadunEtAl2000, GopherEtAl2001, AbboEtAl2005}. As part of this
+model, some authors
+\citep{LevYadunEtAl2000, GopherEtAl2001, AbboEtAl2010, AbboEtAl2012}
+argued that eight plant species, collectively referred to as `founder
+crops' or the `Neolithic crop package' \citep{ZoharyHopf1988} were
+selected and domesticated once, without any phase of pre-domestication
+cultivation \citep[p.~177]{AbboEtAl2011}. This process could have been
+rapid under strong artificial selection
+\citep{HillmanDavies1990, HillmanDavies1992} and may have occurred in a
+single region or `core area' -- generally located in southeast Turkey
+\citep{LadizinskyAdler1976, HeunEtAl1997, OzkanEtAl2002, OzkanEtAl2005, MoriEtAl2003, LuoEtAl2007}.
+From this single point of origin, it was supposed that domesticated or
+semi-domesticated plants radiated outwards to other regions
+\citep{AbboEtAl2006, KilianEtAl2007, OzkanEtAl2011}.
+
+The `short gestation' paradigm was challenged by others
+\citep{Helbaek1969, Harris1989, Kislev1989, Colledge2001, WeissEtAl2004, WillcoxEtAl2008, FullerEtAl2018}.
+Helbæk \citep[in][]{Kirkbride1966} argued that before the appearance of
+domesticated plants, a phase of cultivation of wild seeds must have
+taken place. The existence of a phase of cultivation of morphologically
+wild cereals or `pre-domestication cultivation' was identified in the
+archaeological record through the study of plant domestication traits
+such as grain size, shattering v. non-shattering rachises. The
+archaeobotanical evidence shows that during the Pre-Pottery Neolithic A
+(PPNA) cereals exhibited sizes similar to those recorded in domestic
+species, but their dispersal mechanism was still the same as the one
+present in morphologically-wild species \citep[i.e.~shattering,
+see][]{Kirkbride1966, Kislev1989, HillmanEtAl2001, Colledge2001, WillcoxEtAl2008}.
+This evidence suggested that wild cereal stands could have been
+cultivated for as much as a thousand years before non-shattering
+domestic forms became prevalent in the archaeological record
+\citep{TannoWillcox2006, TannoWillcox2012, ArranzOtaeguiEtAl2016}.
+Additional archaeobotanical
+\citep{Colledge2001, WillcoxEtAl2008, WillcoxEtAl2009, RiehlEtAl2013, ArranzOtaeguiEtAl2016, WeideEtAl2018, DoucheWillcox2018, WhitlamEtAl2018}
+and genetic data
+\citep{BadrEtAl2000, MolinaCanoEtAl2005, KilianEtAl2007, OzkanEtAl2011, IobBotigue2023}
+in recent years has further challenged the short-gestation model to
+explain the origins of plant domestication and agriculture in West Asia.
+
+Similarly, the concept of a limited set of eight `founder crops'
+\citep{ZoharyHopf1988} that were the first species cultivated,
+domesticated and then spread as the basis of Neolithic agricultural
+systems, is not supported by the latest evidence. Our previous analyses
+of the composition of available archaeobotanical datasets shows that
+these crops were of marginal importance during the Epipalaeolithic
+period \citep{ArranzOtaeguiEtAl2018} and that Neolithic subsistence did
+not rely either solely or primarily on the exploitation of these species
+\citep{ArranzOtaeguiRoe2023}. Instead, multiple species of grasses,
+legumes, fruits, nuts, and other plants were exploited over the Late
+Pleistocene-Early Holocene transition in southwest Asia.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Has always been important in study of agricultural origins
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Historically: Vavilov, Pumpelly \& Childe
-  \item
-    Genetic studies tell us origin points, but not ranges
-  \end{itemize}
-\item
-  Important to e.g.
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Distinguish environmental from potentially anthropogenic change
-    \citep{MartinEtAl2016, MartinEtAl2025}
-  \item
-    Reconstruct sequences of domestication \citep{YeomansEtAl2017}
-  \end{itemize}
-\item
-  Epipal./Neo. plant-based economies were diverse
-
-  \begin{itemize}
-  \tightlist
-  \item
-    More than the ``founder crops'';
-  \item
-    More than food
-  \item
-    (In archaebot., not all intentionally collected)
-  \item
-    Geographically and temporally diverse
-  \item
-    \ldots so we model lots of species!
-  \end{itemize}
-\item
-  Regional ecological reconstructions generally rely on the `expert
-  interpolation' (or what do they call it with isoscapes?) method
-
-  \begin{itemize}
-  \tightlist
-  \item
-    See CSEAS (AEA-prep) presentation
-  \item
-    Figure: comparisons
-  \end{itemize}
-\end{itemize}
+The study of the natural distribution of the progenitors of domesticated
+crops has been a central part of discussions on the origins of
+agriculture and plant domestication from the beginning.
+\citet{Humboldt1807} acknowledged the importance of the natural
+distribution of wild species to explain the origin and domestication of
+crops like spelt and rye. \citet{Candolle1886} integrated the study of
+plant ecology and biogeography and influenced \citet{Darwin1859}, who
+later reflected in detail about the geographical distribution of plants
+and species diversity. At that time, there was intense debate about
+whether there were single or multiple ``centres of creation'' of
+species. Researchers aimed to evaluate whether plant and animal species
+emerged in the same locations where they were currently distributed.
+
+After Darwin, this early interest in crop origins evolved into more
+specific discussions about the ``centres of plant domestication''.
+\citet{Vavilov1926} was among the first to seek to determine the number
+of regions in which plants had been independently domesticated
+\citep{Harris1990}. His main method was `differential phytogeography':
+he classified the variation within a crop and established the regions of
+maximum diversity, to locate the geographic regions in which crops
+originated. Using this method, Vavilov suggested that there were at
+least eight centres of origin. His work was later criticised by
+\citet{Harlan1971}, who argued that `centres of origin' and `centres of
+diversity' had to be separated. For Harlan, a `centre' was as an ``area
+in which things originate and out of which things are dispersed''
+\citep[p.~468]{Harlan1971}, and he suggested that three main centres of
+origin of domesticated crops existed. He further indicated that
+Vavilov's approach to the question was simplistic and that more data
+proxies had to be considered (e.g.~archaeology, history, geology), an
+approach more in the tradition of \citet{Candolle1886}. Indeed, the
+inclusion of archaeobotany and genetics in the last decades, together
+with the study of wild relative distributions has been fundamental in
+characterising the origins of agriculture \citep{FullerColledge2008}. As
+a result of modern interdisciplinary studies, the number of recognised
+centres of plant domestication has increased considerably, from the
+three centres suggested by Harlan in 1971 to the six to eight centres
+argued for in the 1990s \citep{Smith1995} and up to as much as 24
+potential centres reported in 2009
+\citetext{\citealp{PuruggananFuller2009}; \citealp[see
+also][]{Fuller2010}}.
+
+Biogeographic research into the centres of origin and/or domestication
+of crops has also long informed broader understanding of the process of
+agricultural origins. In \emph{Man Makes Himself}, \citet{Childe1936}
+correctly located the centre of origin of European agriculture in the
+`Fertile Crescent' of West Asia (unlike for example \citet{Pumpelly1908}
+before him). This was not based on the region's prehistoric
+archaeological record, which at that point was barely known. Instead he
+was guided to the region by biogeographic work by Vavilov and
+\citet{PeakeFleure1927}; only later was this prediction validated by
+archaeological work on the Epipalaeolithic and Neolithic of Palestine
+\citep{Boyd2018}. In subsequent decades, the search for more precise
+origin zones of specific domestic plants relied on the assumption that
+``the locus of domestication of a wild plant would presumably be within
+its area of original distribution in the wild state''
+\citetext{\citealp{Butzer1971}; \citealp[paraphrasing][]{Helbaek1959}}
+-- and that this ``natural habitat'' has not changed significantly over
+the last 12,000 years \citep{Butzer1971}.
+
+Contemporary research on crop origins was pioneered by Harlan and
+Zohary, who compared the current distribution of the wild progenitors of
+domesticated plants in southwest Asia
+\citep{HarlanZohary1966, Zohary1969, Zohary1973, ZoharyHopf1973} to the
+rapidly-expanding archaeobotanical record
+\citetext{\citealp{Harlan1971}; \citealp{Harlan1977}; \citealp[see
+also][]{ZoharyHopf1988}; \citealp{HarlanZohary1966}}. They both were
+interested in evaluating which were the wild ancestors of domesticated
+crops and studying their natural distribution to understand their
+domestication process
+\citep{Zohary1969, Zohary1973, ZoharySpiegelRoy1975}. Indeed, the
+natural distribution of the wild relatives of domestic plant species was
+later used as a criterion to infer `pre-domestic cultivation' in the
+archaeological record. For example, the presence of seeds of chickpea at
+Jericho led \citet{Hopf1986} to interpret the remains as cultivars, as
+the natural distribution of the wild form of chickpea was located
+further to the north. The same rationale was applied to the einkorn
+remains found at several Pre-Pottery Neolithic sites in the southern
+Levant \citep{Hopf1969, Colledge2001}, as the wild progenitors of this
+species was thought to be restricted to the northern Levantine area
+\citep{HeunEtAl1997, ZoharyEtAl2012}. The same idea---presence of plants
+outside their natural range---has been repeated in the literature more
+recently by several other authors
+\citep{TannoWillcox2006, WillcoxEtAl2008, HillmanEtAl2001}.
+
+Despite the importance of biogeography in the development and validation
+of hypotheses regarding the origins of agriculture, there have been few
+studies of the wild range of specific crop progenitors or other relevant
+plant species \citep[cf.~for domestic animals, e.g.][]{YeomansEtAl2017}.
+Observations regarding translocation or range expansion must therefore
+rely on a relatively rough and ahistoric notion of a species' `natural
+distribution' -- that is, one based primarily on contemporary or
+recent-historic occurrences. Yet we know there has been considerable
+climatic and environmental change in West Asia since the terminal
+Pleistocene \citep{JonesEtAl2019}, so it is very unlikely that these
+ranges were in fact static. Reconstructions of broader environments have
+attempted to trace their fluctuations through time, either for the
+entire region
+\citetext{\citealp[e.g.][]{VanZeistBottema1991}; \citealp[Hillman
+in][]{MooreEtAl2000}} or parts of it \citep[e.g.][]{Cordova2007}, but
+these are at the level of the vegetation zone rather than individual
+species. They also invariably rely on what might be called `expert
+interpolation' (where the author composes a map based on his or her own
+knowledge of the relevant data) rather than an explicit modelling
+process. This makes it difficult, if not impossible, for users of such
+reconstructions to understand exactly how they were derived or what
+could explain, for example, the significant discrepencies between the
+predictions of different experts.
 
@@
 archaeological Late Epipalaeolithic (15--11.7 ka) and Pre-Pottery
-Neolithic (11.7--8.3 ka) periods. Our starting point was a list of 97
-taxa (Table~\ref{tbl-results-summary}) comprising the identifiable
-species observed at at least two Late Epipalaeolithic/Pre-Pottery
-Neolithic sites, according to our previous study of the regional
-archaeobotanical data \citep{ArranzOtaeguiRoe2023}. This was based on
-dataset collated from three previously published regional
-archaeobotanical databases: ADEMNES \citep{ADEMNES}, ORIGINS
-\citep{ORIGINS}, and COMPAG
+Neolithic (11.7--8.3 ka) periods. Based on current understandings, we
+assume that plant-based subsistence during this period was broad,
+geographically- and temporally-varied, and reflects a gradual,
+geographically decentred, and nonlinear tranistion to greater reliance
+on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
+point was a list of 97 taxa (Table~\ref{tbl-results-summary}) comprising
+the identifiable species observed at at least two Late
+Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous
+study of the regional archaeobotanical data
+\citep{ArranzOtaeguiRoe2023}. This was based on dataset collated from
+three previously published regional archaeobotanical databases: ADEMNES
+\citep{ADEMNES}, ORIGINS \citep{ORIGINS}, and COMPAG
 \citetext{\citealp{LucasFuller2018}; \citealp{FullerEtAl2018}; \citealp[based
@@
 82 species have reduced ranges; 66 of more than 10\% or more. Though the
-magnitude of the reduction from prehistory to the present likely
-reflects a degree of overfitting in the model (discussed further in
-Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled range
-size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
+magnitude of the change in range size from prehistory to the present
+likely reflects a degree of overfitting in the model (discussed further
+in Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled
+range size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
 (12.9--11.7 ka), and Early Holocene (11.7--8.3 ka) are more directly
@@
 not \emph{Aegilops tauschii} (goatgrasses); and \emph{Vicia} spp.
-(vetches), including \emph{Vicia faba} (broad beans). most of Anatolia,
+(vetches), including \emph{Vicia faba} (broad beans). Most of Anatolia,
 Northern Mesopotamia, and the Zagros Mountains in particular disappear
@@
 \citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
-puts this region at the centre of the development of agriculture.
-Multiple lines of archaeological evidence have emerged that point away
-from this hypothesis and towards a more geographically diverse origin
-\citep{Asouti2006, FullerEtAl2011, ArranzOtaeguiEtAl2016}, but it
-remains the area with some of the earliest clear evidence of
-domestication \citep{ZoharyEtAl2012, KabukcuEtAl2021, UlasEtAl2024}.
-Comparative genetics also points to the Northern Mesopotamia region as
-the centre of diversity of many crops \citep[e.g.][]{HaasEtAl2019}. But
-since these studies are based on modern genomes, if the wild range of
-these plants has, as our modelling suggests, shifted since the Early
-Holocene, the apparent centre of diversity may have shifted with them.
-Our reconstructions are consistent with the late arrival of intensive
+puts this region at the centre of the development of agriculture and
+plant domestication. Multiple lines of archaeological evidence have
+emerged that point away from this hypothesis and towards a more
+geographically diverse origin
+\citep{Asouti2006, FullerEtAl2011, ArranzOtaeguiEtAl2016}, and our
+reconstructions are also consistent with the late arrival of intensive
 plant-based foraging economies in this region (cf.~the Natufian of the
-Levant), and more broadly there need not be a link between the core
-\emph{wild} range of a plant and the core area of its domestication. A
-scenario in which cultivation emerged at the edges of the ranges of
-valuable plant resources---as a means of extend their natural niche---is
-also plausible.
+Levant).
 
 The near-absence of the Zagros in any predicted ranges is also
-surprising, given mounting evidence that domestication took place just
-as early in the eastern \emph{Mashriq} as it did in the west
+surprising, given mounting evidence that animal domestication took place
+just as early in the eastern \emph{Mashriq} as it did in the west
 \citep{Zeder2024}. We consider that the most likely explanation for this
@@
 \citep{Zeder2024}. Recent research
-\citep{WhitlamEtAl2018, GonzalezCarreteroEtAl2023} indicates that plant
-subsistence in this region was based on a distinct set of species,
-compared to the Levant and Anatolia.
+\citep{RiehlEtAl2013, WeideEtAl2017, WeideEtAl2018, WhitlamEtAl2018, GonzalezCarreteroEtAl2023}
+indicates that plant subsistence in this region was based on a distinct
+set of species, compared to the Levant and Anatolia.
 
@@
 centered around the Black Sea Coast of Anatolia and the Palmyra basin.
-In the latter case, neither of these areas are part of the predicted
-modern distribution of wild emmer (centered around the Caucasus and
-Northern Mesopotamia), but it would be consistent with archaeological
-evidence for early cultivation at sites in the Upper Euphrates
-\citep{Willcox2024}.
+In the latter case, neither of these areas are part of the predicted (by
+our model) modern distribution of wild emmer, which is centered around
+the Caucasus and Northern Mesopotamia. But it would be consistent with
+archaeological evidence for early cultivation at sites in the Upper
+Euphrates \citep{Willcox2024}.
 
@@
 relevant to rye's unusual domestication history, as a crop of West Asian
-origin that was apparently not first cultivated until much later than
-the `founder crops', in Europe \citep{SchreiberEtAl2021}.
+origin that was intensively exploited
+\citep{HillmanEtAl2001, DoucheWillcox2018} but apparently not first
+cultivated until much later than the `founder crops', in Europe
+\citep{SchreiberEtAl2021}.
 
@@
 (\emph{Bolboschoenus maritimus}) are similarly constrained to the
-Mediterranean coast (and North Africa, in the case of pistachio).
+Mediterranean coast (and North Africa, in the case of pistachio). This
+is despite the fact that they are well-attested in the archaeotanical
+record from across West Asia.
 
@@
 atlantica}) shows similar trends, but it is less certain that this
-species was cultivated in the Neolithic. Conversely, the various wild
-wheat species native to West Asia show almost no response to
+species was managed in the Neolithic. Conversely, the various wild wheat
+species native to West Asia show almost no response to
 Pleistocene/Holocene climate change, even within the Levant, and in the
@@
 pistachio -- becoming gradually more abundant through the course of the
-Neolithic and dominant by its end. We hypothesise that climate-linked
-range shifts were a factor driving these changes in the apparent
-economic important of different crops: the decreasing availability of
-barley and pistachio in the terminal Pleistocene may have prompted
-cultivation as a strategy to retain access to them, though in the long
-run it made them less attractive as staple crops compared to the more
-resilient wheat. However, these dynamics are not seen in the majority of
-crop progenitors.
+Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}. We
+hypothesise that climate-linked range shifts were a factor driving these
+changes in the apparent economic important of different crops: the
+decreasing availability of barley and pistachio in the terminal
+Pleistocene may have prompted cultivation as a strategy to retain access
+to them, though in the long run it made them less attractive as staple
+crops compared to the more resilient wheat. However, these dynamics are
+not seen in the majority of crop progenitors and may also have involved
+other factors, such as changes in cultural food preferences.
 
@@
 today, has a complex ancestry that involves two recent hybridisation
-events \citep{LevyFeldman2022}: most recently between emmer
+events \citep{LevyFeldman2022}: most recently between domestic emmer
 (\emph{Triticum turgidum dicoccum}) and a goatgrass (\emph{Aegilops
@@
 according to our modelling the only place all four (or even just three)
-are predicted to have coincided or been found in close proximity to
-eachother in the time frame of domestication is a narrow area around the
+are predicted to have coincided or been found in close proximity to each
+other in the time frame of domestication is a narrow area around the
 Orontes river in the northern Levant
@@
 only two sites in our archaeobotanical database where either both
-species of wheat or both species of wheat and the goatgrasses are found
-together: Tell Abu Hureyra and El Kowm II
-\citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to Benjamin Nowak
-  for pointing this out to us.} Taken together this suggests an origin
-of domesticated bread wheat, otherwise only loosely geographically
-constrained to the Levant--Upper Euphrates corridor
-\citep{LevyFeldman2022}, is likely to occur in this vicinity.
+species of wheat (\emph{T. urartu} and \emph{T. turgidum dicoccum}) or
+both species of wheat and the goatgrasses are found together: Tell Abu
+Hureyra and El Kowm II \citep{ArranzOtaeguiRoe2023}.\footnote{We are
+  grateful to Benjamin Nowak for pointing this out to us.} Taken
+together this suggests an origin of domesticated bread wheat, otherwise
+only loosely geographically constrained to the Levant--Upper Euphrates
+corridor \citep{LevyFeldman2022}, is likely to occur in this vicinity.
 
@@
 in archaeobotanical assemblages has several possible explanations. Since
-they do accurately predict the test dataset , a likely culprit is
+they do accurately predict the test dataset, a likely culprit is
 overfitting of the models to the present environment. This implies that
@@
 
-It is consistent with more ``macro'' trends such as the reduced range of
-Hordeum and Pistacia correlating with its reduced abundance in the
-archaeobotanical record.
-
 \section{Conclusion}\label{conclusion}
@@
 which are plausible and consistent with broad-scale patterns in the
-archaeological record.
-
-\begin{itemize}
-\tightlist
-\item
-  Results w.r.t. to specific archaeologically-attested species not to
-  promising, but this doesn't mean the models are useless!
-\item
-  Discrepancies suggests several areas for further research and
-  methodological development
-\end{itemize}
+archaeological record. Results with regard to specific
+archaeologically-attested species are not as promising, but this also
+reflects the incomplete and coarsely temporally-resolved nature of the
+archaeobotanical verification dataset. Discrepencies between the
+modelled distribution, whether on the broad scale (e.g.~the restricted
+geographic range of most species compared to their attestation in the
+archaeological record), or relating to specific species (i.e.~false
+positives and false negatives), suggest several avenues for future
+investigation.
 
@@
 
+\section{Acknowledgements}\label{acknowledgements}
+
+This work was partially funded by the European Research Council
+(ERC-2021-STG project 101039060, \emph{PalaeOrigins: Tracing the
+Epipalaeolithic origins of plant management in southwest Asia}).
+
 \renewcommand\refname{References}

### 2025-07-02 analysis/paper.qmd
Update in response to reviews

@@
         taxon,
+        c("Bolboschoenus glaucus", "Bolboschoenus maritimus") ~ "Bolboschoenus",
         "Vicia narbonense" ~ "Vicia narbonensis", # N=3
@@
 
-Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
-So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
+The taxonomic identifications of archaeobotanical material given in our source databases were controlled to ensure consistency between sources and to group taxa that cannot always be reliably distinguished [for details see @ArranzOtaeguiRoe2023].
+Taxonomic names were then matched to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023] so they could be related to modern occurrences.
+The aim in both cases was to resolve taxonomic ambiguity by using a more inclusive taxonomic unit.
+For example, the model discussed here as *Bolboschoenus* includes both archaeological occurrences of *B.
+maritimus* and *B.
+glaucus*, since these were not distinguished by archaeobotanists until recently [@WollstonecroftEtAl2011], as well as modern occurrences under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
 Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.
 
-## Occurrence data
+## Occurrence data {#sec-occ-data}
 
@@
 
+GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
+@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other regions.
+
 ```{r data-background}
@@
                      impute_with = imp_vars(latitude, longitude)) |>
+  # Drop correlated predictors
+  step_corr(all_predictors(), threshold = 0.7) |>
   # Drop redundant predictors
@@
 
-For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our 24 predictor variables (@sec-predictors).
+For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on up to 24 predictor variables (@sec-predictors).
+Highly correlated (Pearson's *r*>0.7) predictors were removed on a taxon-by-taxon basis, to mitigate issues of overfitting due to colinearity [@DormannEtAl2013], as were redundant predictors with zero variance.
 We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
@@
 Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end [@ArranzOtaeguiRoe2023].
-We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
-the decreasing availability of barley and pistachio in the terminal Pleistocene may have prompted cultivation as a strategy to retain access to them,
-though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
-However, these dynamics are not seen in the majority of crop progenitors and may also have involved other factors, such as changes in cultural food preferences.
 
@@
 #| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
-bread <- c("Aegilops tauschii", "Aegilops speltoides", 
-           "Triticum turgidum dicoccum", "Triticum urartu")
+bread <- c("Aegilops tauschii", "Triticum aestivum", 
+           "Triticum turgidum dicoccum")
 palaeodist_bread <- do.call(c, c(
@@
 This area is c.
-200 km from the only two sites in our archaeobotanical database where either both species of wheat (*T.
-urartu* and *T.
+200 km from the only two sites in our archaeobotanical database where bread wheat (*T.
+aestivum*) co-occurs with both of its recent progenitors (*A.
+tauschii* and *T.
 turgidum dicoccum*) or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
-Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is likely to occur in this vicinity.
+Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is possibly occurred in this vicinity.
 
@@
 
+For a variety of reasons, our models almost certainly underestimate the *fundamental* niche or potential distribution of the target taxon.
+These include:
+
+* The uneven coverage of the GBIF occurrence dataset (see @sec-occ-data)
+
 At the same time, we cannot rule out more substantive reasons for the discrepency between predicted and observed archaeological occurrences.

### 2025-07-02 analysis/paper.tex
Update in response to reviews

@@
 
-Taxonomic names were resolved to the canonical form specified in the
-GBIF Backbone Taxonomy \citep{GBIFSecretariat2023}. So for example
-occurrences for \emph{Bolboschoenus maritimus} also include those
-recorded under the older nomenclature \emph{Scirpus maritimus} (see
+The taxonomic identifications of archaeobotanical material given in our
+source databases were controlled to ensure consistency between sources
+and to group taxa that cannot always be reliably distinguished
+\citep[for details see][]{ArranzOtaeguiRoe2023}. Taxonomic names were
+then matched to the canonical form specified in the GBIF Backbone
+Taxonomy \citep{GBIFSecretariat2023} so they could be related to modern
+occurrences. The aim in both cases was to resolve taxonomic ambiguity by
+using a more inclusive taxonomic unit. For example, the model discussed
+here as \emph{Bolboschoenus} includes both archaeological occurrences of
+\emph{B. maritimus} and \emph{B. glaucus}, since these were not
+distinguished by archaeobotanists until recently
+\citep{WollstonecroftEtAl2011}, as well as modern occurrences under the
+older nomenclature \emph{Scirpus maritimus} (see
 Table~\ref{tbl-results-summary}). Domestic species meeting our inclusion
@@
 
-\subsection{Occurrence data}\label{occurrence-data}
+\subsection{Occurrence data}\label{sec-occ-data}
 
@@
 \end{figure}%
+
+GBIF is currently the best available general-purpose occurrence dataset
+for the West Asia region, its coverage is uneven both geographically and
+from species to species. Figure~\ref{fig-occ-map} shows clearly that the
+Southern Levant, and Israel specifically, is significantly more densely
+sampled than other regions.
 
@@
 For each taxon we trained a classification model to predict occurrence
-(presence or absence/background) based on our 24 predictor variables
-(Section~\ref{sec-predictors}). We used the Random Forest algorithm
-implemented in the R package `ranger' \citep{WrightZiegler2017} and the
-`tidymodels' \citep{tidymodels} framework for data preprocessing and
-model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
-in their recommended hyperparameters and use of down-sampling to balance
-presence and background samples. Models for each taxon were fit
-independently, with redundant zero-variance predictors excluded, and
-assessed based on balanced training (¾) and test (¼) partitions.
+(presence or absence/background) based on up to 24 predictor variables
+(Section~\ref{sec-predictors}). Highly correlated (Pearson's
+\emph{r}\textgreater0.7) predictors were removed on a taxon-by-taxon
+basis, to mitigate issues of overfitting due to colinearity
+\citep{DormannEtAl2013}, as were redundant predictors with zero
+variance. We used the Random Forest algorithm implemented in the R
+package `ranger' \citep{WrightZiegler2017} and the `tidymodels'
+\citep{tidymodels} framework for data preprocessing and model selection.
+To avoid overfitting, we follow \citet{ValaviEtAl2021} in their
+recommended hyperparameters and use of down-sampling to balance presence
+and background samples. Models for each taxon were fit independently,
+with redundant zero-variance predictors excluded, and assessed based on
+balanced training (¾) and test (¼) partitions.
 
@@
 pistachio -- becoming gradually more abundant through the course of the
-Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}. We
-hypothesise that climate-linked range shifts were a factor driving these
-changes in the apparent economic important of different crops: the
-decreasing availability of barley and pistachio in the terminal
-Pleistocene may have prompted cultivation as a strategy to retain access
-to them, though in the long run it made them less attractive as staple
-crops compared to the more resilient wheat. However, these dynamics are
-not seen in the majority of crop progenitors and may also have involved
-other factors, such as changes in cultural food preferences.
+Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}.
 
@@
 (Figure~\ref{fig-palaeodist-bread}). This area is c.~200 km from the
-only two sites in our archaeobotanical database where either both
-species of wheat (\emph{T. urartu} and \emph{T. turgidum dicoccum}) or
-both species of wheat and the goatgrasses are found together: Tell Abu
-Hureyra and El Kowm II \citep{ArranzOtaeguiRoe2023}.\footnote{We are
-  grateful to Benjamin Nowak for pointing this out to us.} Taken
-together this suggests an origin of domesticated bread wheat, otherwise
-only loosely geographically constrained to the Levant--Upper Euphrates
-corridor \citep{LevyFeldman2022}, is likely to occur in this vicinity.
+only two sites in our archaeobotanical database where bread wheat
+(\emph{T. aestivum}) co-occurs with both of its recent progenitors
+(\emph{A. tauschii} and \emph{T. turgidum dicoccum}) or both species of
+wheat and the goatgrasses are found together: Tell Abu Hureyra and El
+Kowm II \citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to
+  Benjamin Nowak for pointing this out to us.} Taken together this
+suggests an origin of domesticated bread wheat, otherwise only loosely
+geographically constrained to the Levant--Upper Euphrates corridor
+\citep{LevyFeldman2022}, is possibly occurred in this vicinity.
 
@@
 factor.
+
+For a variety of reasons, our models almost certainly underestimate the
+\emph{fundamental} niche or potential distribution of the target taxon.
+These include:
+
+\begin{itemize}
+\tightlist
+\item
+  The uneven coverage of the GBIF occurrence dataset (see
+  Section~\ref{sec-occ-data})
+\end{itemize}
 

### 2025-08-10 analysis/paper.qmd
Broaden training region
To West Eurasia, plus related changes suggested by reviewers.

@@
 - [x] Appendix with all hindcast predictions (HTML?)
+-   [] TIFs?
 - [ ] Clean up package code
@@
 
+w_eurasia <- st_bbox(c(xmin = -30, xmax = 90, ymin = 0, ymax = 60), crs = 4326)
+
 w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)
@@
 # Minimum number archaeological occurrences for modelling
-min_n_archaeo <- 1
+min_n_archaeo <- 3
 
 # Minimum number of present occurrences for modelling
-min_n_occ <- 50
+min_n_occ <- 40
 
@@
 In *Man Makes Himself*, @Childe1936 correctly located the centre of origin of European agriculture in the 'Fertile Crescent' of West Asia (unlike for example @Pumpelly1908 before him).
-This was not based on the region's prehistoric archaeological record, which at that point was barely known.
+This was not based on the region's prehistoric archaeological record, which at the time had only been cursorily explored.
 Instead he was guided to the region by biogeographic work by Vavilov and @PeakeFleure1927;
@@
 Here what is modelled is not strictly a biological niche alone, but also aspects of human geography, taphonomy, and archaeological visibility.
-These applications can be distinguihed from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.
+These applications can be distinguished from 'palaeoecological niche modelling', where the object of model remains, as in ecology, a non-human biological niche.
 
@@
   st_geometry() |>
-  st_crop(buffer_bbox(w_asia, 10)) |>
+  st_crop(buffer_bbox(w_eurasia, 10)) |>
   st_union()
@@
   st_geometry() |>
-  st_crop(buffer_bbox(w_asia, 10)) |>
+  st_crop(buffer_bbox(w_eurasia, 10)) |>
   st_difference(ne_ocean) |>
@@
 w_asia_land <- st_intersection(st_as_sfc(w_asia), ne_land)
+w_eurasia_land <- st_intersection(st_as_sfc(w_eurasia), ne_land)
 
@@
     nest(.by = c(taxon), .key = "archaeo") |>
-    # Normalise species names and match to the GBIF Backbone Taxonomy
+    # Add wild progenitors of wheats
+    bind_rows(
+      tibble(
+        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")      )
+    ) |>
+    # Normalise taxon names and match to the GBIF Backbone Taxonomy
     mutate(
+      taxon_original = taxon,
       # Spot corrections
@@
         taxon,
-        c("Bolboschoenus glaucus", "Bolboschoenus maritimus") ~ "Bolboschoenus",
+        # c("Bolboschoenus glaucus", "Bolboschoenus maritimus", "Scirpus maritimus") ~ "Bolboschoenus",
         "Vicia narbonense" ~ "Vicia narbonensis", # N=3
@@
       ),
-      taxon_original = taxon,
-      taxon = map_chr(taxon, gbif_accepted_name),
-      .before = archaeo
-    ) |>
-    # Drop unmatched (mostly not identified to species)
-    drop_na(taxon) |>
-    # Recode domesticates as their wild progenitors.
-    mutate(
+      # Recode domesticates as their wild progenitors
       taxon = case_match(
@@
         "Vicia lens" ~ "Vicia orientalis",
+        "Lens culinaris" ~ "Vicia orientalis",
         "Linum usitatissimum" ~ "Linum bienne",
@@
         .default = taxon
-      )
-    ) |>
+      ),
+      taxon_gbif = name_backbone_checklist(taxon, kingdom = "Plantae"),
+      .before = archaeo
+    ) |>
+    mutate(
+      # taxon = coalesce(taxon_gbif$species, taxon_gbif$genus),
+      taxon = taxon_gbif$species, # not interested in genus-only matches
+      gbif_key = coalesce(taxon_gbif$acceptedUsageKey, taxon_gbif$usageKey),
+      .before = archaeo
+    ) |>
+    # Drop taxons not matched
+    drop_na(taxon) |>
     # Regroup by canonical taxon
+    select(-taxon_gbif) |>
     summarise(
+      gbif_key = first(gbif_key),
       taxon_aliases = list(taxon_original[taxon_original != taxon]),
+      # TODO: this creates duplicate rows
       archaeo = list(bind_rows(archaeo)),
@@
     mutate(
+      # Discard duplicates arising from merging of taxonomic synonyms
+      archaeo = map(archaeo, function(x) {
+        if (ncol(x) > 0) distinct(x, phase_code, .keep_all = TRUE)
+        else data.frame()
+      }),
       n_archaeo = map_int(archaeo, nrow)
     ) |>
-    filter(n_archaeo > min_n_archaeo) |>
-    # Add wild progenitors of wheats
-    bind_rows(
-      tibble(
-        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
-        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
-      )
-    )
+    filter(n_archaeo >= min_n_archaeo)
 
@@
 #| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
+# TODO: Update to show Eurasia (with zoom-in for sites?)
 ggplot() + 
@@
 agriculture, see @sec-bg).
-Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identifiable species observed at at least two Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
+Our starting point was a list of `r nrow(flora)` taxa (@tbl-results-summary) comprising the identified species observed at at least `r english(min_n_archaeo)` Late Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous study of the regional archaeobotanical data [@ArranzOtaeguiRoe2023].
 This was based on dataset collated from three previously published regional archaeobotanical databases: ADEMNES [@ADEMNES], ORIGINS [@ORIGINS], and COMPAG [@LucasFuller2018; @FullerEtAl2018; based on @ColledgeEtAl2004; @ShennanConolly2007].
@@
 
-The taxonomic identifications of archaeobotanical material given in our source databases were controlled to ensure consistency between sources and to group taxa that cannot always be reliably distinguished [for details see @ArranzOtaeguiRoe2023].
+The taxonomic identifications of archaeobotanical material given in our source databases were previously controlled to ensure consistency between sources and to remove taxa that cannot be reliably distinguished [for details see @ArranzOtaeguiRoe2023].
 Taxonomic names were then matched to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023] so they could be related to modern occurrences.
-The aim in both cases was to resolve taxonomic ambiguity by using a more inclusive taxonomic unit.
-For example, the model discussed here as *Bolboschoenus* includes both archaeological occurrences of *B.
-maritimus* and *B.
-glaucus*, since these were not distinguished by archaeobotanists until recently [@WollstonecroftEtAl2011], as well as modern occurrences under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
-Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.
+Archaeologically-attested domestic species meeting our inclusion criteria were substituted for their wild progenitors (where different) when gathering occurrence data, since the domestic forms are now widespread and presumably uninformative of the species original niche.
 
@@
 
-```{r data-n-occ} 
-if (!file_exists(derived_data("flora_n_occ.Rdata"))) {
-  message("Reacquiring GBIF occurrence counts...")
-  flora <- flora |>
-    mutate(
-      n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
-                                            hasCoordinate = TRUE, 
-                                            hasGeospatialIssue = FALSE,
-                                            geometry = bbox_wkt(w_asia)))
-    )
-  saveRDS(flora, derived_data("flora_n_occ.Rdata"))
-}
-flora <- readRDS(derived_data("flora_n_occ.Rdata"))
+```{r data-occ}
+# Download occurrence data from GBIF, generating a citable derived dataset
+if (length(dir_ls(derived_data("gbif"))) == 0) {
+  # Requires GBIF credentials as environment variables, see:
+  # https://docs.ropensci.org/rgbif/articles/gbif_credentials.html
+  message("Reacquiring GBIF occurrences...")
+  gbif_download <- occ_download(
+    pred_in("taxonKey", flora$gbif_key),
+    pred_within(wk::as_wkt(w_eurasia)),
+    pred_default(), # Filter common data quality issues, see ?rgbif::pred_default() for details
+    # pred_lt("coordinateUncertaintyInMeters", 1000), # to strict for W. Asia!
+    format = "SIMPLE_CSV"
+  ) |>
+    occ_download_wait() # wait for dataset to be prepared
+
+  # Cache data
+  occ_download_get(gbif_download$key, path = derived_data("gbif"))
+
+  # Cache metadata
+  gbif_download |>
+    unclass() |>
+    list_flatten() |>
+    discard_at(\(x) str_starts(x, "request")) |>
+    as_tibble() |>
+    write_csv(derived_data(
+      "gbif",
+      paste0(gbif_download$key, "_metadata.csv")
+    ))
+}
+
+flora_occ <- dir_ls(derived_data("gbif"), glob = "*.zip")[1] |> # assume only one dataset present
+  as.download() |>
+  occ_download_import() |>
+  gbif_standardise_names() |>
+  select(taxon_key, gbif_id, latitude, longitude) |>
+  group_by(taxon_key) |>
+  distinct(latitude, longitude, .keep_all = TRUE)
+
+flora_occ_meta <- read_csv(dir_ls(derived_data("gbif"), glob = "*.csv")[1])
+
+# Join to archaeo data
+flora <- flora_occ |>
+  nest(.key = "occ") |>
+  ungroup() |>
+  right_join(flora, by = c("taxon_key" = "gbif_key")) |>
+  mutate(n_occ = map_int(occ, vctrs::vec_size))
 
 # Filter small samples
-# ...and one very large one (why are there 50k occurrences of Avena sterilis?)
 flora_all <- flora
 flora <- flora |>
-  filter(n_occ > min_n_occ) |>
-  filter(taxon != "Avena sterilis")
-```
-
-```{r data-occ}
-# Cleaning pipeline based on:
-# - https://data-blog.gbif.org/post/gbif-filtering-guide/
-if (!file_exists(derived_data("flora_occ.Rdata"))) {
-  message("Reacquiring GBIF occurrences...")
-  flora <- mutate(
-    flora,
-    occ = map(
-      taxon, gbif_data, 
-      hasCoordinate = TRUE, hasGeospatialIssue = FALSE, geometry = w_asia, 
-      limit = 100000, .progress = TRUE
-    ),
-    occ = map(occ, filter, occurrenceStatus  == "PRESENT"),
-    occ = map(occ, gbif_drop_imprecise_coords, threshold = 1000),
-    occ = map(occ, gbif_drop_fossils),
-    occ = map(occ, gbif_drop_duplicate_coords),
-    occ = map(occ, gbif_standardise_names),
-    occ = map(occ, \(x) {
-      select(x, gbif_key = key, scientific_name, genus, species, longitude, 
-             latitude)
-    }),
-    # Update occ count after cleaning
-    n_occ = map_int(occ, nrow)
-  )
-
-  # Refilter based on cleaned records
-  flora <- filter(flora, n_occ > min_n_occ)
-
-  saveRDS(flora, derived_data("flora_occ.Rdata"))
-}
-
-flora <- readRDS(derived_data("flora_occ.Rdata"))
-```
-
-Georeferenced occurrence data for West Asia was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
-GBIF was cleaned to removed fossil occurrences, recorded absences, and records with missing, imprecise (>1 km uncertainty), or invalid coordinates.
-Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
-We also excluded one taxon (*Avena sterilis*) with over 50,000 occurrences, as this would have been computationally prohibitive and we were uncertain what accounted for such a disproportionately high number of records.
+  filter(n_occ >= min_n_occ)
+```
+
+<!-- TODO: cite dataset -->
+Georeferenced occurrence data for West Eurasia between 0 and 60° of latitude was obtained from the Global Biodiversity Information Facility (GBIF) using via and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
+The GBIF dataset [@GBIFDerivedData] excluded fossil occurrences, recorded absences, and records with missing or dubious coordinates.
+Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we did not attempt to model `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
+Multiple records of the same taxon at the same coordinate were discarded because they do not impart information to the model.
+The resulting cleaned dataset used to train our niche models comprises `r flora_occ_meta$totalRecords` from `r flora_occ_meta$numberDatasets` constituent datasets.
 
 ```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia used to train models (N=", nrow(unnest(flora, occ)))}
+# TODO: Update, Eurasia
 flora |>
@@
 GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
-@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other regions.
+@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other parts of West Asia.
 
@@
       )
-    }, region = w_asia_land, n = n_background),
-    # occ_bg = map2(occ, bg, bind_rows)
+    }, region = w_eurasia_land, n = n_background),
     occ = map2(occ, bg, bind_rows)
@@
 However 'absence' data is rarely available because it requires exhaustive survey.
-In practice, most applications of niche modelling are 'presence-only' and, where absence data is required (as for Random Forest), it is supplied as a random background sample of 'pseudo-absence' points.
+In practice, most applications of niche modelling are 'presence-only' and, where absence data is required (as for Random Forest), it is supplied as a random background sample of points.^[We use a background sample that is representative of the entire study area, as opposed to attempting to construct a 'pseudo-absence' sample.
+See @SilleroEtAl2021 for the distinction.]
 The purpose of this sample is to inform the model about the nature of the underlying environment. 
-The stochastic generation process means that some 'pseudo-absence' points will overlap or fall close to presences, so ensuring the model is not overly influenced by background samples is critical to its predictive importance [@ValaviEtAl2022].
+The stochastic generation process means that some of these points will overlap or fall close to presences, so ensuring the model is not overly influenced by background samples is critical to its predictive importance [@ValaviEtAl2022].
 Here we follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (≈10000) uniform sample of points from across the land area of the study region.
@@
 ```{r data-climate}
-# Reacquire derived_data/paleoclim if needed:
+# Reacquire paleoclim data if needed:
+if (!file_exists(derived_data("climate.nc"))) {
   if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
     message("Redownloading Paleoclim data...")
-  source(here("data", "data-paleoclim.R"))
-}
-
+    source(here("analysis", "data", "data-paleoclim.R"))
+  }
+
+  # Store as multidim file so we can read back in as a virtual raster
+  message("Rebuilding palaeoclimate array...")
   climate <- read_stars(
@@
       "paleoclim", 
-    paste0("paleoclim_", climate_periods$code, "_2_5m_w_asia.tif")
+      paste0("paleoclim_", climate_periods$code, "_2_5m_w_eurasia.tif")
     ),
@@
     split("band")
+
+  write_mdim(climate, derived_data("climate.nc"))
+}
+
+climate <- read_stars(derived_data("climate.nc"), proxy = TRUE)
+# Recover dimension values lost in NC conversion
+climate <- st_set_dimensions(climate, "period", values = climate_periods$code)
 ```
@@
 # Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
-if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
+if (!file_exists(derived_data("dem", "srtm_15plus_w_eurasia.tif"))) {
   message("Redownloading DEM data...")
-  source(here("data", "data-dem.R"))
-}
-
-terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_asia.tif", 
-                                            "srtm_15plus_w_asia_slope.tif",
-                                            "srtm_15plus_w_asia_aspect.tif",
-                                            "srtm_15plus_w_asia_wetness.tif")))
+  source(here("analysis", "data", "data-dem.R"))
+}
+
+terrain <- read_stars(derived_data("dem", c("srtm_15plus_w_eurasia.tif", 
+                                            "srtm_15plus_w_eurasia_slope.tif",
+                                            "srtm_15plus_w_eurasia_aspect.tif",
+                                            "srtm_15plus_w_eurasia_wetness.tif")))
 names(terrain) <- c("elevation", "slope", "aspect", "wetness")
@@
 * Terrain aspect and slope, which at high resolution perform well as proxies for solar radiation when modelling plant occurrence [@AustinVanNiel2011; @LeempoelEtAl2015]; and the topographic wetness index (TWI), which serves as a proxy for soil moisture and is particularly important in modelling arid environments [@KopeckyCizkova2010; @CamposEtAl2016; @DiVirgilioEtAl2018].
-All three were derived from the SRTM15+ digital elevation model using algorithms from WhiteboxTools [@Lindsay2016].
+All three were derived from the SRTM30+ digital elevation model using algorithms from WhiteboxTools [@Lindsay2016].
 
@@
   message("Redownloading SoilGrids data...")
-  source(here("data", "data-soilgrids.R"))
+  source(here("analysis", "data", "data-soilgrids.R"))
 }
@@
 # Prepare environmental data cube for prediction (not training)
+if (!file_exists(derived_data("env_w_asia.nc"))) {
+  climate_w_asia <- st_crop(climate, w_asia)
+  terrain_w_asia <- st_crop(terrain, w_asia)
+  soil_w_asia <- st_crop(soil, st_transform(w_asia, soil))
+
   # Duplicate time-invariant predictors to create dummy period dimension
-terrain_cube <- list(terrain) |>
+  terrain_cube <- list(terrain_w_asia) |>
     rep(nrow(climate_periods)) |>
@@
 
-soil_cube <- list(soil) |>
+  soil_cube <- list(soil_w_asia) |>
     rep(nrow(climate_periods)) |>
@@
 
-# Resample predictor data to common projection and resolution
-climate_cube <- st_warp(climate, crs = w_asia_albers, 
+  # Resample to common projection and resolution
+  climate_cube <- st_warp(st_as_stars(climate_w_asia), crs = w_asia_albers, 
                           cellsize = c(pred_res, pred_res), segments = 1000)
-terrain_cube <- st_warp(terrain_cube, climate_cube, segments = 1000)
-soil_cube <- st_warp(soil_cube, climate_cube, segments = 1000)
+  terrain_cube <- st_warp(st_as_stars(terrain_cube), climate_cube, 
+                          segments = 1000)
+  soil_cube <- st_warp(st_as_stars(soil_cube), climate_cube, 
+                       segments = 1000)
 
   # Combine into prediction cube
-environment <- c(climate_cube, terrain_cube, soil_cube)
-
-# Save some memory
-rm(climate_cube, terrain_cube, soil_cube)
+  env_w_asia <- c(climate_cube, terrain_cube, soil_cube)
+  write_mdim(env_w_asia, derived_data("env_w_asia.nc"))
+}
+
+env_w_asia <- read_mdim(derived_data("env_w_asia.nc"))
 ```
@@
 
+if (!file_exists(derived_data("flora_background.Rdata"))) {
   flora <- flora |> 
@@
       occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, remove = FALSE),
-    occ = map(occ, bind_extract, slice(climate, "period", "cur")),
-    occ = map(occ, bind_extract, terrain),
-    occ = map(occ, bind_extract, soil)
-  )
+      occ = map(occ, bind_extract, st_as_stars(hack(climate, "period", "cur"))),
+      occ = map(occ, bind_extract, st_as_stars(terrain)),
+      occ = map(occ, bind_extract, st_as_stars(soil))
+    )
+  saveRDS(flora, derived_data("flora_background.Rdata"))
+}
+flora <- readRDS(derived_data("flora_background.Rdata"))
 ```
@@
 ```{r niche-data}
-# Split training, test (for model assessment) and cross-validation sets (for
-# tuning)
+# Split training and test sets for model assessment 
 flora <-  flora |>
@@
 niche_recipe <- recipe(training(flora$occ_split[[1]])) |>
-  update_role(gbif_key:latitude, new_role = "ID") |>
+  update_role(c(gbif_id, longitude, latitude), new_role = "ID") |>
   update_role(present, new_role = "outcome") |>
@@
               new_role = "predictor") |>
-  # Impute missing values based on spatial autocorrelation
-  step_impute_linear(all_numeric_predictors(), 
+  # Impute missing values assuming spatial autocorrelation:
+  #   can use regression or nearest neighbours
+  # step_impute_linear(all_numeric_predictors(),
+  #                    impute_with = imp_vars(latitude, longitude)) |>
+  step_impute_knn(all_numeric_predictors(),
                      impute_with = imp_vars(latitude, longitude)) |>
@@
     "ranger", 
-    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
+    # handled by hardhat::importance_weights
+    # sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
     importance = "impurity"
@@
 if (!file_exists(derived_data("flora_niche_fit.Rdata"))) {
-  flora <- mutate(
-    flora,
+  flora <- flora |>
+    mutate(
       niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
@@
 if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
-  flora <- flora |>
+  # flora <-
+  flora |>
     mutate(
-      pred = map(niche, collect_predictions),
+      pred = map(niche, collect_predictions)
+    )
+
       palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
-                       obj = environment, .progress = TRUE),
+                       obj = env_w_asia, .progress = TRUE),
       # Binarize
+      # TODO: use max SSS
       palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
@@
   # Excluded samples
-  mutate(exclude = (n_occ < min_n_occ) | (taxon == "Avena sterilis")) |>
+  mutate(exclude = (n_occ < min_n_occ)) |>
   # Prepare table
@@
 
-Our reconstructed palaeodistributions (shown in full in the appendix) indicate that the majority of species have experienced a significant reduction and/or shift in their ranges since the Early Holocene.
-`r n_reduced_range` of `r nrow(flora)` species have reduced ranges; `r n_reduced_range_10` of more than 10% or more.
+Our reconstructed palaeodistributions (shown in full in the appendix) indicate that the majority of species had significantly different geographic ranges in the Late Pleistocene/Early Holocene compared today.
+`r n_reduced_range` of `r nrow(flora)` species are predicted reduced ranges in the past; `r n_reduced_range_10` of more than 10% or more.
 Though the magnitude of the change in range size from prehistory to the present likely reflects a degree of overfitting in the model (discussed further in @sec-discuss-hindcasting), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
@@
 
+<!-- TODO not Bolboschoenus issue - another cause of potential discrepencies? -->
+
 <!--
@@
 
+We are grateful to Alex Weide and the second, anonymous reviewer of our manuscript, whose suggestions have greatly improved this study.
+
 This work was partially funded by the European Research Council (ERC-2021-STG project 101039060, *PalaeOrigins: Tracing the Epipalaeolithic origins of plant management in southwest Asia*).

### 2025-08-10 R/hack.R

@@ -1,0 +1,39 @@
+#' Workaround for broken slice.stars_proxy:
+#' https://github.com/r-spatial/stars/issues/751
+#' @export
+hack = function(.data, ...) {
+  collect_stars(.data, match.call(), "slice", ".data", env = parent.frame())
+}
+
+collect_stars = function(x, call, fn, args = "x", env, ...) {
+  call_list = attr(x, "call_list") %||% list()
+  dots = list(...)
+  nd = names(dots)
+  # I would say now to do
+  # env = as.environment(append(as.list(env), dots)) -> but that didn't work.
+  # so we iterate over ... :
+  for (i in seq_along(dots))
+    env[[ nd[i] ]] = dots[[i]]
+  args = c(args, nd)
+  # set function to call:
+  lst = as.list(call)
+  if (!missing(fn))
+    lst[[1]] = as.name(fn)
+  # set argument names:
+  if (!missing(fn) && fn == "[") {
+    lst[[2]] = as.name(args[1])
+    lst[[3]] = as.name(args[2])
+    for (i in seq_along(args)[-(1:2)]) {
+      if (!args[i] %in% names(lst))
+        lst[[ args[i] ]] = as.name(args[i]) # appends
+    }
+  } else {
+    for (i in seq_along(args)) {
+      lst[[i+1]] = as.name(args[i])
+      names(lst)[[i+1]] = args[i]
+    }
+  }
+  call = as.call(lst)
+  environment(call) = env
+  structure(x, call_list = c(call_list, call))
+}

### 2025-08-10 analysis/data/data-dem.R

@@ -4,21 +4,22 @@
 #   environment and system installation of WhiteBox Tools
 #   <https://www.whiteboxgeo.com/>
 #
-# NOTE: relies on an external API (last run 2024-11-28)
+# NOTE: relies on an external API (last run 2025-08-09)
 library("elevatr")
+library("terra")
 library("whitebox")
 
-srtm_15plus <- get_elev_raster(st_sf(st_as_sfc(w_asia)), prj = 4326, src = "srtm15plus")
-writeRaster(srtm_15plus, derived_data("dem", "srtm_15plus_w_asia.tif"))
+srtm_15plus <- get_elev_raster(st_sf(st_as_sfc(w_eurasia)), prj = 4326, src = "srtm15plus")
+writeRaster(srtm_15plus, derived_data("dem", "srtm_15plus_w_eurasia.tif"))
 
-dem <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia.tif"))
+dem <- wbt_file_path(derived_data("dem", "srtm_15plus_w_eurasia.tif"))
 
 # Slope
-slope <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia_slope.tif"))
+slope <- wbt_file_path(derived_data("dem", "srtm_15plus_w_eurasia_slope.tif"))
 wbt_slope(dem, slope, compress_rasters = TRUE)
 
 # Aspect
-aspect <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia_aspect.tif"))
+aspect <- wbt_file_path(derived_data("dem", "srtm_15plus_w_eurasia_aspect.tif"))
 wbt_aspect(dem, aspect, compress_rasters = TRUE)
 
 # Topographic wetness
@@ -32,5 +33,5 @@ wbt_flow_accumulation_full_workflow(
   compress_rasters = TRUE
 )
 
-wetness <- wbt_file_path(derived_data("dem", "srtm_15plus_w_asia_wetness.tif"))
+wetness <- wbt_file_path(derived_data("dem", "srtm_15plus_w_eurasia_wetness.tif"))
 wbt_wetness_index(sca, slope, wetness, compress_rasters = TRUE)

### 2025-08-10 analysis/data/data-paleoclim.R

@@ -1,26 +1,26 @@
 # data-paleoclim.R
 # Retrieve Paleoclim <http://www.paleoclim.org/> data for W. Asia
-# NOTE: relies on an external API (last run 2024-11-28)
+# NOTE: relies on an external API (last run 2025-08-08)
 library(rpaleoclim)
 
-paleoclim <- map(names(climate_periods), function(period, region) {
+paleoclim <- map(climate_periods$code, function(period, region) {
   paleoclim(
     period = period,
     resolution = "2_5m",
     region = region,
     cache_path = derived_data("paleoclim", "raw")
   )
-}, region = w_asia)
+}, region = w_eurasia)
 
 # Fix extent of CHELSA 'current' layer
-terra::ext(paleoclim[[6]]) <- terra::ext(paleoclim[[1]])
+terra::ext(paleoclim[[4]]) <- terra::ext(paleoclim[[1]])
 
-names(paleoclim) <- names(climate_periods)
+names(paleoclim) <- climate_periods$code
 
 # Write files
 walk2(
   paleoclim,
-  derived_data("paleoclim", paste0("paleoclim_", names(paleoclim), "_2_5m_w_asia.tif")),
+  derived_data("paleoclim", paste0("paleoclim_", names(paleoclim), "_2_5m_w_eurasia.tif")),
   terra::writeRaster,
   datatype = "FLT4S",
   overwrite = TRUE

### 2025-08-10 analysis/data/data-soilgrids.R

@@ -1,7 +1,7 @@
 # data-soilgrids.R
 # Retrieve SoilGrids <https://www.isric.org/explore/soilgrids> data for W. Asia
 # Based on https://git.wur.nl/isric/soilgrids/soilgrids.notebooks/-/blob/master/markdown/webdav_from_R_terra.md
-# NOTE: relies on an external API (last run 2024-01-05)
+# NOTE: relies on an external API (last run 2025-08-08)
 library("fs")
 library("purrr")
 library("sf")
@@ -9,7 +9,6 @@ library("terra")
 library("whitebox")
 library("BadiaPaleoFloraENM")
 
-w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)
 soilgrids_vars <- c("clay", "sand", "silt", "phh2o")
 
 soilgrids <- map(soilgrids_vars, function(var, region) {
@@ -21,7 +20,7 @@ soilgrids <- map(soilgrids_vars, function(var, region) {
   window(soilgrid) <- bbox
 
   return(soilgrid)
-}, region = w_asia)
+}, region = w_eurasia)
 
 names(soilgrids) <- soilgrids_vars
 
@@ -29,7 +28,7 @@ names(soilgrids) <- soilgrids_vars
 temp_path <- dir_create(path_temp("soilgrids"))
 walk2(
   soilgrids,
-  path(temp_path, paste0("soilgrids_", names(soilgrids), "_0-5cm_mean_w_asia.tif")),
+  path(temp_path, paste0("soilgrids_", names(soilgrids), "_0-5cm_mean_w_eurasia.tif")),
   writeRaster,
   datatype = "FLT4S",
   overwrite = TRUE

### 2025-08-14 analysis/appendix.qmd
Update results and discussion

@@
 abstract: |
-  This appendix describes ecological niche models of 81<!-- TODO keep updated -->
+  This appendix describes ecological niche models of 69<!-- TODO keep updated -->
   plant species relevant to the subsistence of Late Epipalaeolithic (15–11.7 ka) 

### 2025-08-14 analysis/appendix_section.qmd
Update results and discussion

@@
   scale_linewidth_identity() +
-  coord_sf(crs = w_asia_albers) +
+  coord_sf(xlim = w_asia[c(1,3)], ylim = w_asia[c(2,4)],
+           crs = w_asia_albers, default_crs = latlong) +
   labs(
     title = "GBIF occurrences",
-    subtitle = paste0("N=", nrow(occ))
+    subtitle = paste0("N=", nrow(occ)),
+    caption = "NB. occurrences outside of prediction region not shown"
   )
@@
     values = c(1, 16),
-    labels = c("Not predicted", "Predicted")
+    labels = c("Not predicted at\nP(present)>0.5", "Predicted at\nP(present)>0.5")
   ) +

### 2025-08-14 analysis/paper.qmd
Update results and discussion

@@
   palaeoclimate simulations.
+  <!-- TODO: rewrite following reviews. Highlight BA to EH decline vs. 'amelioration' concept -->
   The models predict significant reductions and/or shifts in species ranges in
@@
   echo: false
+  cache: true
 ---
@@
 
-```{r dependencies, message=FALSE}
+```{r dependencies, cache=FALSE, message=FALSE}
 library("cowplot")
@@
 library("khroma")
+library("probably")
 library("progressr")
@@
       tibble(
-        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")      )
+        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"))
     ) |>
@@
     ) |>
-    filter(n_archaeo >= min_n_archaeo)
+    filter(
+      n_archaeo >= min_n_archaeo | 
+        taxon %in% c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
+    )
 
@@
 # TODO: slow - cache
+if (!file_exists(derived_data("flora_background.Rdata"))) {
 flora <- flora |>
@@
   select(-bg)
+  saveRDS(flora, derived_data("flora_background.Rdata"))
+}
+flora <- readRDS(derived_data("flora_background.Rdata"))
 ```
@@
 
-if (!file_exists(derived_data("flora_background.Rdata"))) {
+# TODO: slow, cache
 flora <- flora |> 
   mutate(
-      occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, remove = FALSE),
+    occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, 
+              remove = FALSE),
     occ = map(occ, bind_extract, st_as_stars(hack(climate, "period", "cur"))),
@@
   )
-  saveRDS(flora, derived_data("flora_background.Rdata"))
-}
-flora <- readRDS(derived_data("flora_background.Rdata"))
 ```
@@
 flora <-  flora |>
+  # Manually impute some problematic cases
+  # mutate(
+  #   occ = case_when(
+  #     taxon_key %in% c(12225945, 7405743) ~ map(occ, function(x) {
+  #       mutate(x, across())
+  #     })
+  #   )
+  # ) |>
   mutate(
@@
 
-```{r niche-fit}
+```{r niche-fit, cache.lazy=FALSE}
+# Don't use lazy cache here: https://github.com/yihui/knitr/issues/572
 # Fit best model to training data
@@
 
-<!-- TODO: prediction, hindcasting and hindcasting assessment methodology -->
-
-```{r niche-predict}
+```{r niche-predict, cache.lazy=FALSE}
 # NOTE: high memory usage
 if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
-  # flora <-
-  flora |>
+  flora <- flora |>
     mutate(
-      pred = map(niche, collect_predictions)
-    )
-
+      pred = map(niche, collect_predictions),
       palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
@@
       # Binarize
-      # TODO: use max SSS
       palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
-      palaeodist = map(palaeodist, mutate, absent = prediction.FALSE. > 0.5)
     )
@@
 
-```{r data-model-assessment}
+```{r data-model-assessment, cache.lazy=FALSE}
 flora <- mutate(
@@
   filter(.metric == "sens") |>
-  mutate(label = if_else(.estimate < 0.85, taxon, NA_character_)) |>
+  mutate(label = if_else(.estimate < 0.50, taxon, NA_character_)) |>
   ggplot(aes(n_occ, .estimate)) + 
-  geom_smooth(method = "glm", formula = y ~ log(x)) +
+  geom_smooth(method = "glm", formula = y ~ log(x), fullrange = TRUE, 
+              se = FALSE) +
   geom_point() + 
   geom_text_repel(aes(label = label), fontface = "italic", size = 10/.pt) +
-  scale_x_log10(name = "Occurrences") +
-  scale_y_continuous(name = "Sensitivity", labels = label_percent()) +
+  scale_x_log10(name = "Occurrences", labels = label_number()) +
+  scale_y_continuous(name = "Sensitivity", labels = label_percent(),
+                     limits = c(0, 1)) +
   theme_minimal_grid() 
@@
 
-```{r niche-predict-archaeo}
+```{r niche-predict-archaeo, cache.lazy=FALSE}
 # Extract predictors and make predictions for archaeobotanical occurrences
@@
     unnest(archaeo) |>
+    # Add dummy GBIF id
+    mutate(gbif_id = 0) |>
     nest(.by = c(climate_period), .key = "archaeo") |>
@@
       archaeo = map2(archaeo, climate_period, function(x, period) {
-        bind_extract(x, slice(climate, "period", as.character(period)))
+        bind_extract(x, st_as_stars(hack(climate, "period", as.character(period))))
       }),
-      archaeo = map(archaeo, bind_extract, terrain),
-      archaeo = map(archaeo, bind_extract, soil)
+      archaeo = map(archaeo, bind_extract, st_as_stars(terrain)),
+      archaeo = map(archaeo, bind_extract, st_as_stars(soil))
     ) |>
@@
 
-```{r niche-pred-range}
+```{r niche-pred-range, cache.lazy=FALSE}
 area_present <- function(pred, res = pred_res) {
@@
   # Excluded samples
-  mutate(exclude = (n_occ < min_n_occ)) |>
+  mutate(exclude = (n_occ < min_n_occ) & 
+           !taxon %in% c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")) |>
   # Prepare table
@@
     roc_auc = "ROC-AUC",
-    sens = "Sens.",
-    p_predicted = "Sens. (arch.)"
+    sens = "Sens. cur.",
+    p_predicted = "Sens. arch."
   ) |>
@@
   coord_sf(crs = w_asia_albers) +
-  labs(fill = "Richness", x = NULL, y = NULL) +
+  labs(fill = "Predicted\nrichness", x = NULL, y = NULL) +
   theme(
@@
 and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
-Were these area part of the same broader 'interaction sphere' that produce Neolithic agriculture in West Asia?
+Were these area part of the same broader 'interaction sphere' that produced Neolithic agriculture in West Asia?
 
@@
 One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye).
-Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
+Another is rye (*Secale cereale*), which is inferred to be primarily Anatolian (@fig-palaeodist-chickpea-rye).
 This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2001; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].
@@
 
-```{r fig.height=10, fig-palaeodist-barely-levant, cache=TRUE}
+```{r fig-palaeodist-barely-levant, fig.height=10}
 #| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
-# TODO: why so slow?
-palaeodist_hordeum <- do.call(c, c(
-  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
-  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
-))
-
-ggplot() +
-  facet_grid(taxon ~ fct_relabel(period, toupper)) +
-  annotation_spatial(ne_land, fill = "white") +
-  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.), 
-             na.action = na.omit) +
-  annotation_spatial(ne_land, fill = NA) +
-  scale_x_continuous(breaks = c(34, 36, 38)) +
-  scale_fill_batlow(reverse = TRUE) +
-  coord_sf(
-    xlim = c(33, 39), ylim = c(30, 38),
-    crs = w_asia_albers, default_crs = latlong
-  ) +
-  labs(fill = "P(present)", x = NULL, y = NULL) +
-  theme(
-    strip.background = element_blank(),
-    strip.text.y = element_text(face = "italic")
-  )
+# TODO: why OOM?`
+#palaeodist_hordeum <- do.call(c, c(
+#  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
+#  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
+#))
+#
+#ggplot() +
+#  facet_grid(taxon ~ fct_relabel(period, toupper)) +
+#  annotation_spatial(ne_land, fill = "white") +
+#  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.), 
+#             na.action = na.omit) +
+#  annotation_spatial(ne_land, fill = NA) +
+#  scale_x_continuous(breaks = c(34, 36, 38)) +
+#  scale_fill_batlow(reverse = TRUE) +
+#  coord_sf(
+#    xlim = c(33, 39), ylim = c(30, 38),
+#    crs = w_asia_albers, default_crs = latlong
+#  ) +
+#  labs(fill = "P(present)", x = NULL, y = NULL) +
+#  theme(
+#    strip.background = element_blank(),
+#    strip.text.y = element_text(face = "italic")
+#  )
 ```
@@
 #| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
-bread <- c("Aegilops tauschii", "Triticum aestivum", 
-           "Triticum turgidum dicoccum")
+bread <- c("Aegilops tauschii", "Triticum aestivum", "Triticum turgidum")
 palaeodist_bread <- do.call(c, c(
@@
   annotation_spatial(ne_land, fill = "white") +
-  geom_stars(data = select(palaeodist_bread, present), 
-             na.action = na.omit) +
+  geom_stars(data = select(palaeodist_bread, present), na.action = na.omit) +
   annotation_spatial(ne_land, fill = NA) +
@@
 
-## Hindcast models do not predict archaeobotanical composition {#sec-discuss-hindcasting}
-
+## Disagreement between hindcast models and archaeobotanical composition {#sec-discuss-hindcasting}
+
+In general there is middling to low agreement between our hindcast predictions of species niches and observed archaeological occurrences of the same species [@tlb-results-summary].
+
+The majority of our models perform poorly as a binary classification with a threshold of `P(present)>0.5`.
+However, this is an arbitrary threshold; lowering it can improve sensitivity at the cost of specificity.
+<!-- TODO: so why didn't we threshold? Losing information - stick with the probabilities. -->
+In principle archaeological occurrences—as true presences—could be used to 'calibrate' the threshold, but there aren't enough archaeological attestations of most of the modelled species.
+
+<!-- TODO: What about not binary? Mean P(T) and P(F) for example? -->
+
+```{r fig-archaeo-sensitivity}
+#| fig-cap: Summary of hindcasting performance of models in relation to model performance and available archaeological test data
+results_summary |>
+  pivot_longer(c(n_archaeo, sens, n_occ), names_to = "metric") |>
+  mutate(
+    metric = case_match(
+      metric,
+      "n_archaeo" ~ "N occurrences (archaeological)",
+      "n_occ" ~ "N occurrences (current)",
+      "sens" ~ "Model sensitivity (current)"
+    )
+  ) |>
+  ggplot(aes(value, p_predicted)) +
+  facet_wrap(vars(metric), ncol = 1, scales = "free_x", 
+              strip.position = "bottom") +
+  geom_smooth(method = "lm", fullrange = TRUE, se = FALSE) +
+  geom_point() +
+  scale_x_continuous(labels = label_number()) +
+  scale_y_continuous(limits = c(0, 1), labels = label_percent()) +
+  labs(x = NULL, y = "Archaeological occurrences predicted") +
+  theme_minimal_grid() +
+  theme(strip.placement = "outside")
+```
+
+At this point we cannot explain precisely why this discrepency exists, though assessments of model performance offer some clues.
+
+The hindcasting performance of models is not strongly correlated with either the number of occurrences in the training dataset, the performance of the model on test data, or the number of attested archaeological occurrences [@fig-archaeo-sensitivity].
+This indicates against a single, simple explanation for the discrepancy, i.e. that one or the other signal is 'wrong'.
+
+<!-- TODO: core issue - fundamental and realised niches, different but related signals -->
+
+For a variety of reasons, our models probably underestimate the *fundamental* niche or potential distribution of the target taxon.
+These include:
+
+* The uneven coverage of the GBIF occurrence dataset (see @sec-occ-data)
+* Use of time-averaged climate slices
+
+At the same time, there are several reasons to believe that the archaeobotanical data (at least as it has been compiled for this study) *overestimates* the fundamental niche:
+
+* Time-averaged, sometimes quite severely
+* (Understudied) possibility for human transplanting and transport; we know this is the dawn of agriculture!
+* Generally small sample sizes, so individual misidentifications have a large impact
+
+So what can we do about it???
+
+<!-- TODO: rewrite following reviews -->
 The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations. 
@@
 
-For a variety of reasons, our models almost certainly underestimate the *fundamental* niche or potential distribution of the target taxon.
-These include:
-
-* The uneven coverage of the GBIF occurrence dataset (see @sec-occ-data)
-
 At the same time, we cannot rule out more substantive reasons for the discrepency between predicted and observed archaeological occurrences.
@@
 Results with regard to specific archaeologically-attested species are not as promising, but this also reflects the incomplete and coarsely temporally-resolved nature of the archaeobotanical verification dataset.
-Discrepencies between the modelled distribution, whether on the broad scale (e.g.
+Discrepancies between the modelled distribution, whether on the broad scale (e.g.
 the restricted geographic range of most species compared to their attestation in the archaeological record), or relating to specific species (i.e.

### 2025-08-14 analysis/paper.tex
Update results and discussion

@@
 from these models using downsampled data from palaeoclimate simulations.
+
 The models predict significant reductions and/or shifts in species
@@
 before him). This was not based on the region's prehistoric
-archaeological record, which at that point was barely known. Instead he
-was guided to the region by biogeographic work by Vavilov and
-\citet{PeakeFleure1927}; only later was this prediction validated by
-archaeological work on the Epipalaeolithic and Neolithic of Palestine
-\citep{Boyd2018}. In subsequent decades, the search for more precise
-origin zones of specific domestic plants relied on the assumption that
-``the locus of domestication of a wild plant would presumably be within
-its area of original distribution in the wild state''
+archaeological record, which at the time had only been cursorily
+explored. Instead he was guided to the region by biogeographic work by
+Vavilov and \citet{PeakeFleure1927}; only later was this prediction
+validated by archaeological work on the Epipalaeolithic and Neolithic of
+Palestine \citep{Boyd2018}. In subsequent decades, the search for more
+precise origin zones of specific domestic plants relied on the
+assumption that ``the locus of domestication of a wild plant would
+presumably be within its area of original distribution in the wild
+state''
 \citetext{\citealp{Butzer1971}; \citealp[paraphrasing][]{Helbaek1959}}
@@
 taphonomy, and archaeological visibility. These applications can be
-distinguihed from `palaeoecological niche modelling', where the object
+distinguished from `palaeoecological niche modelling', where the object
 of model remains, as in ecology, a non-human biological niche.
@@
 on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
-point was a list of 97 taxa (Table~\ref{tbl-results-summary}) comprising
-the identifiable species observed at at least two Late
+point was a list of 66 taxa (Table~\ref{tbl-results-summary}) comprising
+the identified species observed at at least 3 Late
 Epipalaeolithic/Pre-Pottery Neolithic sites, according to our previous
@@
 The taxonomic identifications of archaeobotanical material given in our
-source databases were controlled to ensure consistency between sources
-and to group taxa that cannot always be reliably distinguished
+source databases were previously controlled to ensure consistency
+between sources and to remove taxa that cannot be reliably distinguished
 \citep[for details see][]{ArranzOtaeguiRoe2023}. Taxonomic names were
@@
 Taxonomy \citep{GBIFSecretariat2023} so they could be related to modern
-occurrences. The aim in both cases was to resolve taxonomic ambiguity by
-using a more inclusive taxonomic unit. For example, the model discussed
-here as \emph{Bolboschoenus} includes both archaeological occurrences of
-\emph{B. maritimus} and \emph{B. glaucus}, since these were not
-distinguished by archaeobotanists until recently
-\citep{WollstonecroftEtAl2011}, as well as modern occurrences under the
-older nomenclature \emph{Scirpus maritimus} (see
-Table~\ref{tbl-results-summary}). Domestic species meeting our inclusion
-criteria were substituted with their wild progenitor(s), where
-different.
+occurrences. Archaeologically-attested domestic species meeting our
+inclusion criteria were substituted for their wild progenitors (where
+different) when gathering occurrence data, since the domestic forms are
+now widespread and presumably uninformative of the species original
+niche.
 
@@
 
-Georeferenced occurrence data for West Asia was obtained from the Global
-Biodiversity Information Facility (GBIF) using via its application
-programming interface and the R package `rgbif'
-\citep{ChamberlainBoettiger2017, rgbif}. GBIF was cleaned to removed
-fossil occurrences, recorded absences, and records with missing,
-imprecise (\textgreater1 km uncertainty), or invalid coordinates.
-Although niche models have reasonable predictive power even with small
-training samples
-\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we
-excluded 14 taxa with less than 50 usable occurrences, following
-recommendations for niche models generally and Random Forest-based
-models specifically \citep{StockwellPeterson2002, LuanEtAl2020}. We also
-excluded one taxon (\emph{Avena sterilis}) with over 50,000 occurrences,
-as this would have been computationally prohibitive and we were
-uncertain what accounted for such a disproportionately high number of
-records.
+\begin{verbatim}
+Rows: 1 Columns: 11
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (5): key, doi, license, status, downloadLink
+dbl  (3): size, totalRecords, numberDatasets
+dttm (3): created, modified, eraseAfter
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+\end{verbatim}
+
+Georeferenced occurrence data for West Eurasia between 0 and 60° of
+latitude was obtained from the Global Biodiversity Information Facility
+(GBIF) using via and the R package `rgbif'
+\citep{ChamberlainBoettiger2017, rgbif}. The GBIF dataset
+\citep{GBIFDerivedData} excluded fossil occurrences, recorded absences,
+and records with missing or dubious coordinates. Although niche models
+have reasonable predictive power even with small training samples
+\citep{StockwellPeterson2002, HernandezEtAl2006, WiszEtAl2008}, we did
+not attempt to model 3 taxa with less than 40 usable occurrences,
+following recommendations for niche models generally and Random
+Forest-based models specifically
+\citep{StockwellPeterson2002, LuanEtAl2020}. Multiple records of the
+same taxon at the same coordinate were discarded because they do not
+impart information to the model. The resulting cleaned dataset used to
+train our niche models comprises \ensuremath{3.387196\times 10^{6}} from
+4764 constituent datasets.
 
@@
 \caption{\label{fig-occ-map}Georeferenced occurrence records from West
-Asia used to train models (N=81641}
+Asia used to train models (N=1409939}
 
@@
 Southern Levant, and Israel specifically, is significantly more densely
-sampled than other regions.
+sampled than other parts of West Asia.
 
@@
 data is required (as for Random Forest), it is supplied as a random
-background sample of `pseudo-absence' points. The purpose of this sample
-is to inform the model about the nature of the underlying environment.
-The stochastic generation process means that some `pseudo-absence'
-points will overlap or fall close to presences, so ensuring the model is
-not overly influenced by background samples is critical to its
-predictive importance \citep{ValaviEtAl2022}. Here we follow the advice
-of \citet{BarbetMassinEtAl2012} for regression-based species
-distribution models and use a large (≈10000) uniform sample of points
-from across the land area of the study region. These points are then
-weighted equally against the presences in the regression to produce a
-`balanced Random Forest' \citep{ValaviEtAl2022}.
+background sample of points.\footnote{We use a background sample that is
+  representative of the entire study area, as opposed to attempting to
+  construct a `pseudo-absence' sample. See \citet{SilleroEtAl2021} for
+  the distinction.} The purpose of this sample is to inform the model
+about the nature of the underlying environment. The stochastic
+generation process means that some of these points will overlap or fall
+close to presences, so ensuring the model is not overly influenced by
+background samples is critical to its predictive importance
+\citep{ValaviEtAl2022}. Here we follow the advice of
+\citet{BarbetMassinEtAl2012} for regression-based species distribution
+models and use a large (≈10000) uniform sample of points from across the
+land area of the study region. These points are then weighted equally
+against the presences in the regression to produce a `balanced Random
+Forest' \citep{ValaviEtAl2022}.
 
@@
 predictor variables. These included:
+
+\begin{verbatim}
+bio_1, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+\end{verbatim}
+
+\begin{verbatim}
+Warning in parse_netcdf_meta(meta_data, get_names(x)): NAs introduced by
+coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_10, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_11, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_12, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_13, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_14, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_15, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_16, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_17, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_18, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_19, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_2, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_3, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_4, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_5, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_6, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_7, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_8, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
+
+\begin{verbatim}
+bio_9, 
+\end{verbatim}
+
+\begin{verbatim}
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
+Vertical dimension.
+Warning in CPL_read_gdal(as.character(x), as.character(options),
+as.character(driver), : NAs introduced by coercion
+\end{verbatim}
 
@@
   \citep{KopeckyCizkova2010, CamposEtAl2016, DiVirgilioEtAl2018}. All
-  three were derived from the SRTM15+ digital elevation model using
+  three were derived from the SRTM30+ digital elevation model using
   algorithms from WhiteboxTools \citep{Lindsay2016}.
@@
 and resolution of 5 km.
+
+\begin{verbatim}
+Warning: There were 57 warnings in `mutate()`.
+The first warning was:
+i In argument: `occ = map(occ, bind_extract, st_as_stars(hack(climate,
+  "period", "cur")))`.
+Caused by warning in `CPL_get_metadata()`:
+! GDAL Message 1: dimension #0 (period) is not a Time or Vertical dimension.
+i Run `dplyr::last_dplyr_warnings()` to see the 56 remaining warnings.
+\end{verbatim}
 
@@
 
-We trained Random Forest models for 82 taxa using contemporary
+We trained Random Forest models for 63 taxa using contemporary
 occurrence data from GBIF, a random sample of background points, and the
@@
 hindcast predictions for reconstructed past environments in 4 key
-climate periods -- a total of 328 modelled palaeodistributions.
+climate periods -- a total of 252 modelled palaeodistributions.
 Predicted distributions for individual taxa are presented in the
@@
 Table~\ref{tbl-results-summary}. Model accuracy (proportion of correctly
-classified presence and background samples) ranged between 73\% and
-99\%, with an average of 93\%. Sensitivity (proportion of correctly
-classified presence samples) ranged between 79\% and 100\%, with an
-average of 94\%. The area under the models' receiver operating
-characteristic curves (ROC-AUC) was on average 0.978±0.061 Model
+classified presence and background samples) ranged between 92\% and
+100\%, with an average of 97\%. Sensitivity (proportion of correctly
+classified presence samples) ranged between 12\% and 100\%, with an
+average of 82\%. The area under the models' receiver operating
+characteristic curves (ROC-AUC) was on average 0.987±0.018 Model
 sensitivity is loosely correlated with the number of occurrences
@@
 with the worst-performing models all having less than 300 recorded
-occurrences: \emph{Asphodelus aestivus}, \emph{Triticum turgidum
-dicoccum}, \emph{Zygophyllum fabago}, and \emph{Stipa dregeana}. Test
-metrics and ROC curves for the individual models are included in the
-appendix.
+occurrences: \emph{Lathyrus sativus}, \emph{Lepidium perfoliatum},
+\emph{Triticum aestivum}, \emph{Bolboschoenus glaucus}, \emph{Gypsophila
+vaccaria}, \emph{Euclidium syriacum}, \emph{Gypsophila elegans},
+\emph{Citrullus colocynthis}, \emph{Triticum monococcum}, \emph{Triticum
+turgidum}, \emph{Aegilops crassa}, \emph{Secale strictum}, \emph{Cicer
+reticulatum}, \emph{Medicago astroites}, \emph{Triticum durum},
+\emph{Gypsophila pilosa}, \emph{Arnebia decumbens}, \emph{Salvia
+absconditiflora}, \emph{Suaeda fruticosa}, \emph{Buglossoides
+tenuiflora}, \emph{Lathyrus oleraceus}, \emph{Vitis sylvestris}, and
+\emph{Vicia orientalis}. Test metrics and ROC curves for the individual
+models are included in the appendix.
 
@@
 (proportion of corrected predicted presenses) in relation to the
-archaeological data is on average 0.06±0.13
+archaeological data is on average 0.07±0.15
 Table~\ref{tbl-results-summary}. A full assessment of the hindcasting
@@
 \midrule\addlinespace[2.5pt]
-{\itshape Adonis dentata} & 2 & 715 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
-{\itshape Aegilops crassa} & 4 & 143 & 0.82 & 0.94 & 0.86 & 0.00 \\ 
-{\itshape Aegilops speltoides} & — & 701 & 0.92 & 0.98 & 0.95 & — \\ 
-{\itshape Aegilops tauschii} & — & 725 & 0.89 & 0.98 & 0.94 & — \\ 
-{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 515 & 0.98 & 1.00 & 0.96 & 0.00 \\ 
-{\itshape Ammi majus} & 4 & 900 & 0.98 & 0.99 & 0.97 & 0.00 \\ 
-{\itshape Androsace maxima} & 17 & 231 & 0.88 & 0.95 & 0.88 & 0.12 \\ 
-{\itshape Arenaria serpyllifolia} & 3 & 139 & 0.84 & 0.95 & 0.91 & 0.00 \\ 
-{\itshape Arnebia decumbens} & 21 & 324 & 0.95 & 0.95 & 0.89 & 0.00 \\ 
-{\itshape Arnebia linearifolia} & 13 & 225 & 0.98 & 0.99 & 0.94 & 0.00 \\ 
-{\itshape Asphodelus aestivus} & 2 & 76 & 0.97 & 0.97 & 0.83 & 0.00 \\ 
-{\itshape Atriplex prostrata} & 4 & 266 & 0.98 & 1.00 & 0.94 & 0.00 \\ 
-{\itshape Avena fatua} & 2 & 579 & 0.86 & 0.93 & 0.86 & 0.00 \\ 
-{\itshape Avena sterilis} & 5 & 53244\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Bassia arabica} & 4 & 264 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
-{\itshape Bassia prostrata} & 2 & 46\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Bolboschoenus glaucus} & 5 & 27\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{3}} & 31 & 1040 & 0.98 & 0.99 & 0.97 & 0.03 \\ 
-{\itshape Brachypodium distachyon} & 5 & 2270 & 0.98 & 0.99 & 0.98 & 0.00 \\ 
-{\itshape Bromus sterilis} & 3 & 759 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
-{\itshape Buglossoides arvensis} & 23 & 260 & 0.87 & 0.97 & 0.95 & 0.09 \\ 
-{\itshape Buglossoides tenuiflora} & 26 & 252 & 0.96 & 0.99 & 0.95 & 0.08 \\ 
-{\itshape Camphorosma monspeliaca} & 2 & 38\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Capparis spinosa} & 4 & 4938 & 0.97 & 0.99 & 0.97 & 0.00 \\ 
-{\itshape Capsella bursa-pastoris} & 2 & 1329 & 0.94 & 0.99 & 0.94 & 0.00 \\ 
-{\itshape Carex divisa} & 9 & 336 & 0.95 & 0.98 & 0.94 & 0.00 \\ 
-{\itshape Celtis tournefortii} & 2 & 14\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Ceratonia siliqua} & 2 & 5072 & 0.98 & 1.00 & 0.99 & 0.50 \\ 
-{\itshape Chenopodium album} & 5 & 480 & 0.93 & 0.98 & 0.94 & 0.00 \\ 
-{\itshape Cicer reticulatum}\textsuperscript{\textit{4}} & 3 & 52 & 0.97 & 1.00 & 1.00 & 0.00 \\ 
-{\itshape Citrullus colocynthis} & 4 & 475 & 0.89 & 0.96 & 0.90 & 0.00 \\ 
-{\itshape Crithopsis delileana} & 3 & 406 & 0.98 & 1.00 & 0.97 & 0.00 \\ 
-{\itshape Erodium ciconium} & 2 & 359 & 0.92 & 0.99 & 0.94 & 0.00 \\ 
-{\itshape Euclidium syriacum} & 3 & 32\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Ficus carica} & 8 & 3495 & 0.94 & 0.99 & 0.95 & 0.25 \\ 
-{\itshape Fumaria densiflora} & 4 & 413 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
-{\itshape Gypsophila elegans} & 3 & 76 & 0.92 & 0.99 & 1.00 & 0.00 \\ 
-{\itshape Gypsophila pilosa} & 5 & 173 & 0.93 & 0.97 & 0.90 & 0.00 \\ 
-{\itshape Gypsophila vaccaria}\textsuperscript{\textit{5}} & 8 & 511 & 0.96 & 0.99 & 0.93 & 0.00 \\ 
-{\itshape Halothamnus hierochunticus} & 2 & 57 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
-{\itshape Helianthemum ledifolium} & 2 & 697 & 0.97 & 0.99 & 0.97 & 0.00 \\ 
-{\itshape Helianthemum salicifolium} & 2 & 1400 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
-{\itshape Henrardia pubescens} & 3 & 1\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Hordeum bulbosum} & 4 & 2850 & 0.97 & 0.99 & 0.96 & 0.00 \\ 
-{\itshape Hordeum murinum} & 5 & 1949 & 0.94 & 0.98 & 0.93 & 0.00 \\ 
-{\itshape Hordeum spontaneum}\textsuperscript{\textit{6}} & 110 & 4656 & 0.96 & 0.99 & 0.97 & 0.20 \\ 
-{\itshape Juglans regia} & 2 & 821 & 0.85 & 0.96 & 0.95 & 0.00 \\ 
-{\itshape Krascheninnikovia ceratoides} & 2 & 42\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Lathyrus aphaca} & 4 & 2079 & 0.94 & 0.99 & 0.97 & 0.00 \\ 
-{\itshape Lathyrus cicera} & 2 & 591 & 0.89 & 0.98 & 0.94 & 0.50 \\ 
-{\itshape Lathyrus oleraceus}\textsuperscript{\textit{7}} & 8 & 1580 & 0.89 & 0.97 & 0.93 & 0.12 \\ 
-{\itshape Lathyrus sativus} & 4 & 196 & 0.79 & 0.95 & 0.92 & 0.00 \\ 
-{\itshape Lepidium perfoliatum} & 3 & 83 & 0.77 & 0.94 & 0.91 & 0.00 \\ 
-{\itshape Lepidium ruderale} & 2 & 37\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Linum bienne}\textsuperscript{\textit{8}} & 14 & 376 & 0.96 & 1.00 & 0.99 & 0.00 \\ 
-{\itshape Lolium rigidum} & 5 & 2529 & 0.98 & 1.00 & 0.98 & 0.00 \\ 
-{\itshape Lolium temulentum} & 3 & 159 & 0.94 & 0.96 & 0.90 & 0.00 \\ 
-{\itshape Medicago astroites}\textsuperscript{\textit{9}} & 15 & 143 & 0.89 & 0.96 & 0.88 & 0.33 \\ 
-{\itshape Medicago minima} & 2 & 1379 & 0.92 & 0.99 & 0.97 & 0.50 \\ 
-{\itshape Medicago radiata} & 20 & 799 & 0.93 & 0.99 & 0.98 & 0.20 \\ 
-{\itshape Minuartia rubella} & 2 & 0\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Moltkia angustifolia} & 2 & 11\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Moltkia coerulea} & 2 & 71 & 0.86 & 0.97 & 0.88 & 0.00 \\ 
-{\itshape Neotorularia torulosa}\textsuperscript{\textit{10}} & 2 & 539 & 0.98 & 0.98 & 0.93 & 0.00 \\ 
-{\itshape Peganum harmala} & 2 & 917 & 0.92 & 0.97 & 0.93 & 0.00 \\ 
-{\itshape Phalaris paradoxa} & 3 & 1191 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
-{\itshape Phragmites australis} & 4 & 3412 & 0.97 & 0.99 & 0.95 & 0.00 \\ 
-{\itshape Pistacia atlantica} & 6 & 1874 & 0.98 & 1.00 & 0.97 & 0.17 \\ 
-{\itshape Pistacia terebinthus}\textsuperscript{\textit{11}} & 2 & 3419 & 0.98 & 1.00 & 0.98 & 0.50 \\ 
-{\itshape Plantago lagopus} & 2 & 1730 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
-{\itshape Poa bulbosa} & 5 & 1880 & 0.95 & 0.98 & 0.92 & 0.00 \\ 
-{\itshape Polygonum arenarium arenarium}\textsuperscript{\textit{12}} & 7 & 5\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Polygonum corrigioloides} & 6 & 4\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Prosopis farcta} & 5 & 2263 & 0.98 & 0.99 & 0.96 & 0.00 \\ 
-{\itshape Quercus ithaburensis} & 2 & 2216 & 0.99 & 1.00 & 0.97 & 0.00 \\ 
-{\itshape Rumex pulcher} & 6 & 1394 & 0.98 & 1.00 & 0.98 & 0.00 \\ 
-{\itshape Salsola kali} & 6 & 537 & 0.98 & 1.00 & 1.00 & 0.00 \\ 
-{\itshape Salvia absconditiflora}\textsuperscript{\textit{13}} & 3 & 105 & 0.94 & 0.98 & 0.92 & 0.00 \\ 
-{\itshape Secale cereale} & 7 & 268 & 0.81 & 0.94 & 0.90 & 0.00 \\ 
-{\itshape Secale strictum}\textsuperscript{\textit{14}} & 3 & 124 & 0.76 & 0.94 & 0.97 & 0.00 \\ 
-{\itshape Stipa dregeana} & 2 & 79 & 0.95 & 0.96 & 0.85 & 0.00 \\ 
-{\itshape Suaeda fruticosa} & 3 & 238 & 0.99 & 1.00 & 0.98 & 0.00 \\ 
-{\itshape Taeniatherum caput-medusae} & 5 & 262 & 0.81 & 0.96 & 0.95 & 0.00 \\ 
-{\itshape Tribulus terrestris} & 2 & 1091 & 0.90 & 0.96 & 0.89 & 0.00 \\ 
-{\itshape Triticum aestivum compactum} & 2 & 147 & 0.82 & 0.96 & 0.92 & 0.00 \\ 
-{\itshape Triticum aestivum spelta}\textsuperscript{\textit{15}} & 2 & 16\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Triticum durum} & 3 & 95 & 0.91 & 0.99 & 1.00 & 0.00 \\ 
-{\itshape Triticum monococcum aegilopoides}\textsuperscript{\textit{16}} & 62 & 1176 & 0.88 & 0.99 & 0.99 & 0.08 \\ 
-{\itshape Triticum turgidum dicoccum}\textsuperscript{\textit{17}} & 60 & 60 & 0.73 & 0.86 & 0.79 & 0.18 \\ 
-{\itshape Triticum urartu} & — & 424 & 0.93 & 0.99 & 0.95 & — \\ 
-{\itshape Verbena officinalis} & 3 & 934 & 0.96 & 0.99 & 0.97 & 0.00 \\ 
-{\itshape Vicia ervilia} & 26 & 688 & 0.87 & 0.96 & 0.94 & 0.12 \\ 
-{\itshape Vicia faba} & 7 & 1597 & 0.90 & 0.98 & 0.94 & 0.43 \\ 
-{\itshape Vicia narbonensis} & 3 & 1451 & 0.93 & 0.99 & 0.98 & 0.33 \\ 
-{\itshape Vicia orientalis}\textsuperscript{\textit{18}} & 16 & 353 & 0.84 & 0.95 & 0.92 & 0.19 \\ 
-{\itshape Vitis sylvestris} & 3 & 12\textsuperscript{\textit{2}} & — & — & — & — \\ 
-{\itshape Zygophyllum fabago} & 3 & 263 & 0.89 & 0.93 & 0.81 & 0.00 \\ 
+{\itshape Aegilops crassa} & 4 & 223 & 0.98 & 0.98 & 0.43 & 0.00 \\ 
+{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 762 & 0.99 & 1.00 & 0.88 & 0.50 \\ 
+{\itshape Ammi majus} & 4 & 10311 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
+{\itshape Androsace maxima} & 17 & 4946 & 0.94 & 0.98 & 0.91 & 0.06 \\ 
+{\itshape Arenaria serpyllifolia} & 3 & 101612 & 0.98 & 0.98 & 1.00 & 0.00 \\ 
+{\itshape Arnebia decumbens} & 21 & 616 & 0.97 & 0.97 & 0.59 & 0.05 \\ 
+{\itshape Arnebia linearifolia} & 13 & 239 & 1.00 & 0.99 & 0.89 & 0.00 \\ 
+{\itshape Atriplex prostrata} & 4 & 80181 & 0.98 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Avena sterilis} & 5 & 15906 & 0.96 & 0.99 & 0.99 & 0.40 \\ 
+{\itshape Bassia arabica} & 4 & 262 & 1.00 & 1.00 & 0.95 & 0.00 \\ 
+{\itshape Bolboschoenus glaucus} & 5 & 418 & 0.98 & 0.97 & 0.53 & 0.00 \\ 
+{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{2}} & 31 & 53809 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
+{\itshape Brachypodium distachyon} & 5 & 13800 & 0.98 & 1.00 & 0.99 & 0.20 \\ 
+{\itshape Bromus sterilis} & 3 & 89764 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Buglossoides arvensis} & 23 & 23004 & 0.95 & 0.98 & 0.98 & 0.09 \\ 
+{\itshape Buglossoides tenuiflora} & 26 & 399 & 0.99 & 0.99 & 0.71 & 0.04 \\ 
+{\itshape Capparis spinosa} & 4 & 8617 & 0.97 & 0.99 & 0.96 & 0.00 \\ 
+{\itshape Carex divisa} & 9 & 7400 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
+{\itshape Chenopodium album} & 5 & 184403 & 0.99 & 0.98 & 1.00 & 0.00 \\ 
+{\itshape Cicer reticulatum}\textsuperscript{\textit{3}} & 3 & 52 & 1.00 & 1.00 & 0.69 & 0.00 \\ 
+{\itshape Citrullus colocynthis} & 4 & 1823 & 0.94 & 0.95 & 0.69 & 0.00 \\ 
+{\itshape Crithopsis delileana} & 3 & 430 & 1.00 & 1.00 & 0.93 & 0.00 \\ 
+{\itshape Euclidium syriacum} & 3 & 338 & 0.98 & 0.97 & 0.27 & 0.00 \\ 
+{\itshape Ficus carica} & 8 & 64424 & 0.98 & 0.99 & 0.99 & 0.38 \\ 
+{\itshape Fumaria densiflora} & 4 & 3704 & 0.96 & 0.99 & 0.94 & 0.00 \\ 
+{\itshape Gypsophila elegans} & 3 & 1603 & 0.95 & 0.98 & 0.81 & 0.00 \\ 
+{\itshape Gypsophila pilosa} & 5 & 348 & 0.99 & 0.99 & 0.67 & 0.00 \\ 
+{\itshape Gypsophila vaccaria}\textsuperscript{\textit{4}} & 8 & 1177 & 0.95 & 0.98 & 0.65 & 0.00 \\ 
+{\itshape Henrardia pubescens} & 3 & 1\textsuperscript{\textit{5}} & — & — & — & — \\ 
+{\itshape Hordeum bulbosum} & 4 & 4134 & 0.97 & 1.00 & 0.95 & 0.00 \\ 
+{\itshape Hordeum murinum} & 5 & 74770 & 0.98 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Hordeum spontaneum}\textsuperscript{\textit{6}} & 76 & 3098 & 0.99 & 1.00 & 0.97 & 0.20 \\ 
+{\itshape Lathyrus aphaca} & 4 & 18652 & 0.96 & 0.99 & 0.98 & 0.00 \\ 
+{\itshape Lathyrus oleraceus}\textsuperscript{\textit{7}} & 8 & 819 & 0.98 & 0.99 & 0.80 & 0.00 \\ 
+{\itshape Lathyrus sativus} & 4 & 2120 & 0.92 & 0.96 & 0.75 & 0.00 \\ 
+{\itshape Lepidium perfoliatum} & 3 & 2764 & 0.94 & 0.98 & 0.84 & 0.00 \\ 
+{\itshape Linum bienne}\textsuperscript{\textit{8}} & 14 & 12824 & 0.98 & 0.99 & 0.99 & 0.07 \\ 
+{\itshape Lolium rigidum} & 5 & 10620 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
+{\itshape Lolium temulentum} & 3 & 4463 & 0.94 & 0.98 & 0.93 & 0.00 \\ 
+{\itshape Medicago astroites}\textsuperscript{\textit{9}} & 15 & 104 & 1.00 & 0.99 & 0.64 & 0.00 \\ 
+{\itshape Medicago radiata} & 20 & 834 & 0.99 & 0.99 & 0.91 & 0.30 \\ 
+{\itshape Phalaris paradoxa} & 3 & 3287 & 0.97 & 0.99 & 0.94 & 0.00 \\ 
+{\itshape Phragmites australis} & 4 & 373041 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Pistacia atlantica} & 6 & 2419 & 0.99 & 1.00 & 0.94 & 0.33 \\ 
+{\itshape Poa bulbosa} & 5 & 30769 & 0.96 & 0.98 & 0.99 & 0.20 \\ 
+{\itshape Polygonum arenarium}\textsuperscript{\textit{10}} & 7 & 25\textsuperscript{\textit{5}} & — & — & — & — \\ 
+{\itshape Polygonum corrigioloides} & 6 & 12\textsuperscript{\textit{5}} & — & — & — & — \\ 
+{\itshape Prosopis farcta} & 5 & 2512 & 0.99 & 1.00 & 0.96 & 0.60 \\ 
+{\itshape Rumex pulcher} & 6 & 17606 & 0.97 & 0.99 & 0.99 & 0.00 \\ 
+{\itshape Salsola kali} & 6 & 14699 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
+{\itshape Salvia absconditiflora}\textsuperscript{\textit{11}} & 3 & 128 & 0.99 & 1.00 & 0.65 & 0.00 \\ 
+{\itshape Secale cereale} & 4 & 14803 & 0.95 & 0.98 & 0.97 & 0.00 \\ 
+{\itshape Secale strictum}\textsuperscript{\textit{12}} & 3 & 162 & 0.99 & 0.99 & 0.50 & 0.00 \\ 
+{\itshape Suaeda fruticosa} & 3 & 577 & 0.99 & 1.00 & 0.79 & 0.00 \\ 
+{\itshape Taeniatherum caput-medusae} & 4 & 1969 & 0.96 & 0.99 & 0.89 & 0.00 \\ 
+{\itshape Triticum aestivum}\textsuperscript{\textit{13}} & 4 & 217 & 0.99 & 0.98 & 0.59 & 0.00 \\ 
+{\itshape Triticum durum} & 3 & 64 & 1.00 & 0.99 & 0.50 & 0.00 \\ 
+{\itshape Triticum monococcum}\textsuperscript{\textit{14}} & 47 & 870 & 0.98 & 0.99 & 0.80 & 0.15 \\ 
+{\itshape Triticum turgidum}\textsuperscript{\textit{15}} & 53 & 345 & 0.97 & 0.93 & 0.22 & 0.00 \\ 
+{\itshape Verbena officinalis} & 3 & 89654 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Vicia ervilia} & 26 & 1924 & 0.95 & 0.98 & 0.86 & 0.35 \\ 
+{\itshape Vicia faba} & 7 & 42636 & 0.97 & 0.99 & 0.99 & 0.43 \\ 
+{\itshape Vicia narbonensis}\textsuperscript{\textit{16}} & 3 & 2826 & 0.95 & 0.99 & 0.89 & 0.33 \\ 
+{\itshape Vicia orientalis}\textsuperscript{\textit{17}} & 16 & 41 & 1.00 & 0.95 & 0.12 & 0.00 \\ 
+{\itshape Vitis sylvestris} & 3 & 63 & 1.00 & 0.99 & 0.36 & 0.00 \\ 
+{\itshape Zygophyllum fabago} & 3 & 3554 & 0.97 & 0.99 & 0.91 & 0.00 \\ 
 \bottomrule
@@
 \textsuperscript{\textit{1}}Including \emph{Aizoon hispanicum}\\
-\textsuperscript{\textit{2}}Excluded from modelling due to sample size\\
-\textsuperscript{\textit{3}}Including \emph{Scirpus maritimus}\\
-\textsuperscript{\textit{4}}Including \emph{Cicer arietinum}\\
-\textsuperscript{\textit{5}}Including \emph{Vaccaria pyramidata}\\
+\textsuperscript{\textit{2}}Including \emph{Scirpus maritimus}\\
+\textsuperscript{\textit{3}}Including \emph{Cicer arietinum}\\
+\textsuperscript{\textit{4}}Including \emph{Vaccaria pyramidata}\\
+\textsuperscript{\textit{5}}Excluded from modelling due to sample size\\
 \textsuperscript{\textit{6}}Including \emph{Hordeum vulgare}\\
@@
 \textsuperscript{\textit{9}}Including \emph{Trigonella astroites}\\
-\textsuperscript{\textit{10}}Including \emph{Torularia torulosa}\\
-\textsuperscript{\textit{11}}Including \emph{Pistacia palaestina}\\
-\textsuperscript{\textit{12}}Including \emph{Polygonum venantianum}\\
-\textsuperscript{\textit{13}}Including \emph{Salvia cryptantha}\\
-\textsuperscript{\textit{14}}Including \emph{Secale montanum}\\
-\textsuperscript{\textit{15}}Including \emph{Triticum spelta}\\
-\textsuperscript{\textit{16}}Including \emph{Triticum monococcum} and \emph{Triticum boeoticum}\\
-\textsuperscript{\textit{17}}Including \emph{Triticum aestivum}, \emph{Triticum dicoccum}, and \emph{Triticum dicoccoides}\\
-\textsuperscript{\textit{18}}Including \emph{Lens culinaris}\\
+\textsuperscript{\textit{10}}Including \emph{Polygonum venantianum}\\
+\textsuperscript{\textit{11}}Including \emph{Salvia cryptantha}\\
+\textsuperscript{\textit{12}}Including \emph{Secale montanum}\\
+\textsuperscript{\textit{13}}Including \emph{Triticum spelta} and \emph{Triticum aestivocompactum}\\
+\textsuperscript{\textit{14}}Including \emph{Triticum boeoticum}\\
+\textsuperscript{\textit{15}}Including \emph{Triticum aestivum}, \emph{Triticum dicoccum}, and \emph{Triticum dicoccoides}\\
+\textsuperscript{\textit{16}}Including \emph{Vicia narbonense}\\
+\textsuperscript{\textit{17}}Including \emph{Lens culinaris} and \emph{Lens orientalis}\\
 \end{minipage}
@@
 Our reconstructed palaeodistributions (shown in full in the appendix)
-indicate that the majority of species have experienced a significant
-reduction and/or shift in their ranges since the Early Holocene. 75 of
-82 species have reduced ranges; 66 of more than 10\% or more. Though the
-magnitude of the change in range size from prehistory to the present
-likely reflects a degree of overfitting in the model (discussed further
-in Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled
-range size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
-(12.9--11.7 ka), and Early Holocene (11.7--8.3 ka) are more directly
-comparable (Figure~\ref{fig-pred-range-density}). The average range of
-modelled species was 16\% in the Early Holocene compared to the
-Bølling--Allerød, and 25\% during the Younger Dryas (i.e.~ranges
-recovered slightly between the Younger Dryas and Early Holocene). This
-perhaps indicates that although this period is considered one of
-climatatic amelioration globally \citep{JonesEtAl2019}, the colder
-conditions of the Pleistocene may have supported more extensive
-plant-based economies in West Asia specifically.
+indicate that the majority of species had significantly different
+geographic ranges in the Late Pleistocene/Early Holocene compared today.
+55 of 63 species are predicted reduced ranges in the past; 54 of more
+than 10\% or more. Though the magnitude of the change in range size from
+prehistory to the present likely reflects a degree of overfitting in the
+model (discussed further in Section~\ref{sec-discuss-hindcasting}),
+fluctuations in modelled range size between the Bølling-Allerød
+(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
+(11.7--8.3 ka) are more directly comparable
+(Figure~\ref{fig-pred-range-density}). The average range of modelled
+species was 31\% in the Early Holocene compared to the Bølling--Allerød,
+and 28\% during the Younger Dryas (i.e.~ranges recovered slightly
+between the Younger Dryas and Early Holocene). This perhaps indicates
+that although this period is considered one of climatatic amelioration
+globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
+may have supported more extensive plant-based economies in West Asia
+specifically.
 
@@
 retreating from the Badia/transjordan region (e.g.~barley, fig.
-Figure~\ref{fig-palaeodist-barely-levant}).
+\textbf{?@fig-palaeodist-barely-levant}).
 
@@
 southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}). Another is
-rye (\emph{Secale cereale}), which is inferred to be primarily Western
-Anatolian (Figure~\ref{fig-palaeodist-chickpea-rye}). This is perhaps
-relevant to rye's unusual domestication history, as a crop of West Asian
-origin that was intensively exploited
-\citep{HillmanEtAl2001, DoucheWillcox2018} but apparently not first
-cultivated until much later than the `founder crops', in Europe
-\citep{SchreiberEtAl2021}.
+rye (\emph{Secale cereale}), which is inferred to be primarily Anatolian
+(Figure~\ref{fig-palaeodist-chickpea-rye}). This is perhaps relevant to
+rye's unusual domestication history, as a crop of West Asian origin that
+was intensively exploited \citep{HillmanEtAl2001, DoucheWillcox2018} but
+apparently not first cultivated until much later than the `founder
+crops', in Europe \citep{SchreiberEtAl2021}.
 
@@
 
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-barely-levant-1.pdf}}
-
-}
-
-\caption{\label{fig-palaeodist-barely-levant}Predicted
-palaeodistribution of wild barleys in the Levant}
-
-\end{figure}%
-
 Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
@@
 concurrent with it being brought into cultivation
-(Figure~\ref{fig-palaeodist-barely-levant}). It also sees a marked
-decline in the archaeobotanical record from the Early PPNA/Early PPNB
-(where it was amongst the most common taxa) to the Late PPNB and Late
-Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
-atlantica}) shows similar trends, but it is less certain that this
-species was managed in the Neolithic. Conversely, the various wild wheat
-species native to West Asia show almost no response to
-Pleistocene/Holocene climate change, even within the Levant, and in the
-archaeobotanical record wheat displays the opposite trend to barley and
-pistachio -- becoming gradually more abundant through the course of the
-Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}.
+(\textbf{?@fig-palaeodist-barely-levant}). It also sees a marked decline
+in the archaeobotanical record from the Early PPNA/Early PPNB (where it
+was amongst the most common taxa) to the Late PPNB and Late Neolithic
+\citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia atlantica})
+shows similar trends, but it is less certain that this species was
+managed in the Neolithic. Conversely, the various wild wheat species
+native to West Asia show almost no response to Pleistocene/Holocene
+climate change, even within the Levant, and in the archaeobotanical
+record wheat displays the opposite trend to barley and pistachio --
+becoming gradually more abundant through the course of the Neolithic and
+dominant by its end \citep{ArranzOtaeguiRoe2023}.
 
@@
 We present the first continuous, spatially explicit models of the
-palaeodistributions of 82 plant species found regularly in association
+palaeodistributions of 63 plant species found regularly in association
 with Late Epipalaeolithic and Early Neolithic sites in West Asia. This
@@
 reflects the incomplete and coarsely temporally-resolved nature of the
-archaeobotanical verification dataset. Discrepencies between the
+archaeobotanical verification dataset. Discrepancies between the
 modelled distribution, whether on the broad scale (e.g.~the restricted
@@
 
+We are grateful to Alex Weide and the second, anonymous reviewer of our
+manuscript, whose suggestions have greatly improved this study.
+
 This work was partially funded by the European Research Council

### 2025-08-15 analysis/appendix.qmd
Finalise resubmission

@@
 abstract: |
-  This appendix describes ecological niche models of 69<!-- TODO keep updated -->
+  This appendix describes ecological niche models of 65<!-- TODO keep updated -->
   plant species relevant to the subsistence of Late Epipalaeolithic (15–11.7 ka) 
@@
     toc: true
-    # embed-resources: true
+    embed-resources: true
 execute:

### 2025-08-15 analysis/lay_summary.qmd
Finalise resubmission

--- 2025-03-31 analysis/lay_summary.qmd
@@
 
-We reconstructed the past ranges of 82 plant species that were used by prehistoric societies in West Asia around 10,000 years ago, including the ancestors of the first crops.
+We used machine learning to reconstruct the past ranges of 65 plant species that were used by prehistoric societies in West Asia around 10,000 years ago, including the ancestors of the first crops.
 Our results suggest that under the cold, arid conditions of that time these were much less widespread than they are now, and that they became even scarcer just before the emergence of agriculture.

### 2025-08-15 analysis/paper.qmd
Finalise resubmission

@@
   orcid: 0000-0002-1011-1244
-  email: joeroe@hey.com
+  email: jar@hum.ku.dk
   url: https://joeroe.io
@@
   This paper presents the first continuous, spatially-explicit reconstructions 
-  of the palaeodistributions of 82 plant species found regularly in association 
+  of the palaeodistributions of 65 plant species found regularly in association 
   with early agricultural archaeological sites in West Asia, including the 
@@
   in relation to climate and environmental variables. Predictions of the 
-  potential ranges of these species at key stages of the Pleistocene–Holocene 
+  potential niches of these species at key stages of the Pleistocene–Holocene 
   transition could then be derived from these models using downsampled data from 
   palaeoclimate simulations.
-  <!-- TODO: rewrite following reviews. Highlight BA to EH decline vs. 'amelioration' concept -->
-  The models predict significant reductions and/or shifts in species ranges in
-  the terminal Pleistocene and Early Holocene compared to present conditions. 
-  Many species that are found throughout the region's 'hilly flanks' today 
-  are indicated to have much more restricted distributions centered on the 
-  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
-  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
-  The models performance in predicting the occurrence of specific species at 
-  archaeological sites is highly variable, but in aggregate the predictions are
-  coherent and align with broad-scale trends in the ubiquity and relative 
-  abundance of species in the archaeological record.<!--
-  however [...]
-  The regional ubiquity of species in the archaeological record is [not] correlated with the 
-  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
-  correlated with the predicted diversity of its environs. This indicates that 
-  trends in taxonomic composition of the archaeobotanical record is [not] likely 
-  to have been influenced by environmental change and species turnover, in
-  addition to human economic choices.-->
+  Our models performed well against independent contemporary test data, but 
+  their ability to predict the occurrence of specific species at 
+  archaeological sites was much more variable, probably reflecting a tendency of 
+  the method to underestimate the species' fundamental niche. Nevertheless, the 
+  majority of species are predicted to have had more restricted geographic 
+  distributions under past climate conditions compared to today. Crop 
+  progenitors and several wild food species modelled are modelled to have been 
+  concentrated in the Levant and, to a lesser extent, Cyprus and Western 
+  Anatolia, suggesting these regions may have served as glacial *refugia*. 
+  The average size of species' niche shrunk by an average of c. 25% from the 
+  terminal Pleistocene to the Early Holocene, indicating that economically
+  significant plants were adapted to cryo-arid conditions and did not, as often
+  assumed, initially respond positively to the 'ameliorated' climate of the 
+  Holocene.
 format: 
@@
       cite-style: authoryear
-bibliography: references.bib
+  docx:
+    toc: true
+    number-sections: true
+bibliography: 
+  - references.bib
+  - gbif.bib
 execute: 
@@
 TODO before publication:
-- [ ] Revise discussion of archaeo verification
-- [ ] Outstanding TODOs
-- [ ] Add missing references
-- [x] Appendix with all hindcast predictions (HTML?)
--   [] TIFs?
+- [ ] Zenodo release
 - [ ] Clean up package code
-- [ ] Zenodo release
-- [ ] Final proofread
 -->
@@
 library("khroma")
+library("patchwork")
 library("probably")
@@
 Rather than inferring environmental conditions from preserved physical evidence, 
-we predict the ranges of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
+we predict the potential niche of individual species relevant to human subsistence based on a model of their current environmental niche and simulations of past palaeoclimate. 
 Though hypothetical, this gives us an independent line of evidence on past ecologies that is independent of the environmental archaeological and palaeoclimatic records.
@@
 In this sense, discrepencies between the two records are perhaps the most interesting result, as they indicate processes affecting one or both records that are not fully accounted for and therefore generate new questions.
-Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple past climatologies.
+Our computational approach is also readily scaled up, allowing us to model spatially-explicit palaeodistributions for a large number of species, for the whole region, under multiple reconstructed climate scenarios.
 
@@
 
-<!-- TODO: figure highlighting discrepencies between 'expert interpolation' -->
-
 ## Ecological niche modelling in archaeology
 
-Ecological niche modelling or species distribution modelling is widely used by ecologists to predict the geographic range of a species based on a set of environmental predictors [@FranklinMiller2009].
+Ecological niche modelling or species distribution modelling is widely used by ecologists to mathematically predict the geographic distribution of a species from a sample of occurrences [@FranklinMiller2009; @SilleroEtAl2021].
 Essentially, it involves combining records of where an organism has been observed with environmental data (climate, topography, etc.) for those locations to model the range of environmental values at which that species  – its environmental niche.
-This model can then be used to predict the range of the organism in question either in the same or a different environment.
-@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past ranges.
+This model can then be used to predict the potential niche of the organism in question either in the same or a different environment.
+@TownsendPetersonSoberon2012 suggests reserving the term 'species distribution modelling' for when the method is used to recover the verifiable range of a species in a real and existing environment, and using 'ecological niche modelling' as the broader term covering hypothetical or predictive applications – a convention we follow here when referring to predictive or 'hindcast' models of past niches.
 Within this overarching framework, ecological niche modelling encompasses a wide range of applications and a variety of potential environmental predictors, modelling approaches, and methodologies, which we will not attempt to review here.
@@
 Ecological niche modelling has long been of interest to archaeologists as both a means of exploring the biological niche of humans and for reconstructing the past environments they inhabited [@DavidPollyEronen2011; @FranklinEtAl2015].
-In the first sense, it has been used most extensively to model the range of humans and other hominin species [e.g.
+In the first sense, it has been used most extensively to model the niche of humans and other hominin species [e.g.
 @BenitoEtAl2017; @YousefiEtAl2020; @BanksEtAl2021; @YaworskyEtAl2024a; @YaworskyEtAl2024b; @GuranEtAl2024], especially in the Palaeolithic.
@@
   st_geometry() |>
-  st_crop(buffer_bbox(w_eurasia, 10)) |>
+#  st_crop(buffer_bbox(w_eurasia, 10)) |>
   st_difference(ne_ocean) |>
@@
     drop_na(taxon) |>
+    filter(taxon != "Medicago astroites") |> # Not reliably identifiable
     # Regroup by canonical taxon
@@
       taxon_aliases = list(taxon_original[taxon_original != taxon]),
-      # TODO: this creates duplicate rows
       archaeo = list(bind_rows(archaeo)),
@@
 ```{r fig-region}
-#| fig-cap: Late Epipalaeolithic & Pre-Pottery Neolithic archaeological sites used to generate modelled flora
-# TODO: Update to show Eurasia (with zoom-in for sites?)
+#| fig-cap: Map of the study region (West Asia, grey box) with locations of Late Epipalaeolithic and Pre-Pottery Neolithic archaeobotanical assemblages.
 ggplot() + 
@@
 ## Occurrence data {#sec-occ-data}
-
-<!-- TODO generate GBIF occurrence citation:
-https://docs.ropensci.org/rgbif/articles/getting_occurrence_data.html -->
 
@@
     pred_default(), # Filter common data quality issues, see ?rgbif::pred_default() for details
-    # pred_lt("coordinateUncertaintyInMeters", 1000), # to strict for W. Asia!
+    # pred_lt("coordinateUncertaintyInMeters", 1000), # too strict for W. Asia!
     format = "SIMPLE_CSV"
@@
 
-flora_occ_meta <- read_csv(dir_ls(derived_data("gbif"), glob = "*.csv")[1])
+flora_occ_meta <- read_csv(dir_ls(derived_data("gbif"), glob = "*.csv")[1],
+                           show_col_types = FALSE)
+
+# Generate GBIF citation
+writeLines(paste0(r"(
+@misc{GBIFDataDownload,
+  title = {GBIF {Occurrence} {Download}},
+  author = {{GBIF}},
+	publisher = {The Global Biodiversity Information Facility},
+  date = {2025-08-08},
+  year = {2025},
+  url = {https://doi.org/)", flora_occ_meta$doi, r"(}
+})"), here("analysis/gbif.bib"))
 
@@
 
-<!-- TODO: cite dataset -->
 Georeferenced occurrence data for West Eurasia between 0 and 60° of latitude was obtained from the Global Biodiversity Information Facility (GBIF) using via and the R package 'rgbif' [@ChamberlainBoettiger2017; @rgbif].
-The GBIF dataset [@GBIFDerivedData] excluded fossil occurrences, recorded absences, and records with missing or dubious coordinates.
+The GBIF dataset [@GBIFDataDownload] excluded fossil occurrences, recorded absences, and records with missing or dubious coordinates.
 Although niche models have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we did not attempt to model `r english(nrow(filter(flora_all, n_occ < min_n_occ)))` taxa with less than `r min_n_occ` usable occurrences, following recommendations for niche models generally and Random Forest-based models specifically [@StockwellPeterson2002; @LuanEtAl2020].
 Multiple records of the same taxon at the same coordinate were discarded because they do not impart information to the model.
-The resulting cleaned dataset used to train our niche models comprises `r flora_occ_meta$totalRecords` from `r flora_occ_meta$numberDatasets` constituent datasets.
-
-```{r fig-occ-map, fig.cap=paste0("Georeferenced occurrence records from West Asia used to train models (N=", nrow(unnest(flora, occ)))}
-# TODO: Update, Eurasia
-flora |>
+The resulting cleaned dataset used to train our niche models comprises `r format(flora_occ_meta$totalRecords, scientific = FALSE)` occurrences from `r flora_occ_meta$numberDatasets` constituent datasets.
+
+```{r fig-occ-map, message=FALSE, fig.cap=paste0("Georeferenced occurrence records from West Eurasia used to train models (N=", nrow(unnest(flora, occ)), ").
+Inset, right: prediction region (West Asia).")}
+fig_occ_map <- flora |>
  unnest(occ) |>
@@
  geom_sf(shape = ".") +
- layer_spatial(w_asia, fill = NA) +
  coord_sf(crs = w_asia_albers, default_crs = latlong)
+
+fig_occ_map_inset <- fig_occ_map +
+  coord_sf(
+    crs = w_asia_albers, default_crs = latlong,
+    xlim = w_asia[c(1,3)], ylim = w_asia[c(2,4)], expand = FALSE
+  )
+
+fig_occ_map <- fig_occ_map +
+ layer_spatial(w_asia, fill = NA)
+
+fig_occ_map + fig_occ_map_inset
 ```
@@
 GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
-@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other parts of West Asia.
+The Southern Levant, and Israel specifically, is significantly more densely sampled than other parts of West Asia (@fig-occ-map).
 
 ```{r data-background}
-# TODO: slow - cache
 if (!file_exists(derived_data("flora_background.Rdata"))) {
@@
 
-```{r data-climate}
+```{r data-climate, warning=FALSE, message=FALSE}
 # Reacquire paleoclim data if needed:
@@
 
-climate <- read_stars(derived_data("climate.nc"), proxy = TRUE)
+climate <- read_stars(derived_data("climate.nc"), proxy = TRUE, quiet = TRUE)
 # Recover dimension values lost in NC conversion
@@
 
-```{r data-flora-predictors}
-# TODO: factor out
+```{r data-flora-predictors, warning = FALSE}
 bind_extract <- function(x, y) {
@@
 
-# TODO: slow, cache
 flora <- flora |> 
@@
 flora <-  flora |>
-  # Manually impute some problematic cases
-  # mutate(
-  #   occ = case_when(
-  #     taxon_key %in% c(12225945, 7405743) ~ map(occ, function(x) {
-  #       mutate(x, across())
-  #     })
-  #   )
-  # ) |>
   mutate(
@@
     occ = map(occ, \(x) select(tibble(x), -geometry)),
-    # TODO: can this happen earlier?
     occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
@@
   saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
+
+  # Write predictions in TIF format
+  palaeodist_dir <- dir_create(here("analysis", "palaeodist"))
+  walk2(flora$taxon, flora$palaeodist, function(taxon, palaeodist) {
+    filename <- paste0(str_replace(taxon, " ", "_"), ".tif")
+    write_stars(
+      select(palaeodist, "prediction.TRUE."), 
+      path(palaeodist_dir, filename)
+    )
+  })
 }
@@
 
-# TODO write TIFs
 ```
@@
 
-We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition [@tbl-results-summary].
+We assessed the predictive performance of the fitted niche models in the contemporary environment based on the reserved test partition (@tbl-results-summary).
 Model accuracy (proportion of correctly classified presence and background samples) ranged between `r percent(filter(niche_test_summary, .metric == "accuracy")$min)` and `r percent(filter(niche_test_summary, .metric == "accuracy")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "accuracy")$mean)`.
 Sensitivity (proportion of correctly classified presence samples) ranged between `r percent(filter(niche_test_summary, .metric == "sens")$min)` and `r percent(filter(niche_test_summary, .metric == "sens")$max)`, with an average of `r percent(filter(niche_test_summary, .metric == "sens")$mean)`.
-The area under the models' receiver operating characteristic curves (ROC-AUC) was on average `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`
-Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 300 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.85)$taxon, "*"))`.
+The area under the models' receiver operating characteristic curves (ROC-AUC) was on average `r round(filter(niche_test_summary, .metric == "roc_auc")$mean, 3)`±`r round(filter(niche_test_summary, .metric == "accuracy")$sd, 3)`.
+Model sensitivity is loosely correlated with the number of occurrences available for training (@fig-model-sensitivity-vs-n_occ), with the worst-performing models all having less than 1000 recorded occurrences: `r knitr::combine_words(paste0("*", filter(unnest(flora, niche_test_metrics), .metric == "sens" & .estimate < 0.50)$taxon, "*"))`.
 Test metrics and ROC curves for the individual models are included in the appendix.
@@
 # Extract predictors and make predictions for archaeobotanical occurrences
-# TODO: what about negatives?
 if (!file.exists(derived_data("archaeo_pred.Rdata"))) {
@@
       archaeo_pred = map2(niche, archaeo, function(x, y) {
-        # TODO: shouldn't be necessary - remove from workflow
         y <- mutate(y, gbif_key = NA, scientific_name = NA, species = NA)
@@
 
-The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is worse, with `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted.
-Model sensitivity (proportion of corrected predicted presenses) in relation to the archaeological data is on average `r round(mean(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)`±`r round(sd(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)` [@tbl-results-summary].
+The ability of the hindcast models to predict the occurrence of specific species at archaeological sites is worse, with just `r percent(p_archaeo_pred_correct)` of presences in archaeobotanical assemblages successfully predicted at a threshold of 0.5.
+Model sensitivity (proportion of corrected predicted presenses) in relation to the archaeological data is on average `r round(mean(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)`±`r round(sd(archaeo_pred_summary$p_predicted, na.rm=TRUE), 2)` (@tbl-results-summary).
 A full assessment of the hindcasting performance of the individual models can be found in the appendix.
-<!-- TODO: negative -->
+
+In summary, the majority of our models perform poorly as simple binary classifiers with a threshold of `P(present)>0.5`.
+However, this is an arbitrary threshold; lowering it can improve sensitivity (proportion of occurrences correctly predicted) at the cost of specificity (proportion of non-occurrences incorrectly predicted).
+There are several approaches to selecting informative thresholds for binary classification in the ENM literature [@LiuEtAl2005; @LiuEtAl2013; @LiuEtAl2015], with most aiming for an optimum compromise between sensitivity and specificity.
+We were unable to find a thresholding technique that significantly improved the hindcast models' performance as binary classifiers.
+By definition, binarizing model output also discards information from the underlying probabilistic prediction, so we have opted to present results probabilistically as far as possible.
+In principle the archaeological occurrences themselves could be used to 'calibrate' the hindcast models and select an optimum threshold, but there aren't enough archaeological attestations of most of the modelled species for this to be practicable at this time.
 
@@
 
-# TODO:
-# - alt. taxons as footnotes
-# - check boxes of which (climate?) periods observed in ?
 results_summary <- flora_all |>
@@
 
-## Reduction in range sizes over the Pleistocene/Holocene boundary
+## Reduction in potential niche sizes over the Pleistocene/Holocene boundary
 
@@
 ```{r fig-pred-range-density, warning=FALSE}
-#| fig-cap: Distribution of predicted species ranges by period. Dashed lines indicate the median range.
+#| fig-cap: Distribution of predicted species niche size by period (logarithmic scale). Dashed lines indicate the median niche size.
 flora |>
@@
   geom_vline(aes(xintercept = med_area_present), flora_pred_area_present_summary) +
-  scale_x_log10(name = "Predicted range km²", labels = label_number_auto()) + 
+  scale_x_log10(name = "Potential niche km²", labels = label_number_auto()) + 
   scale_y_continuous(name = NULL, labels = NULL) +
@@
 
-Our reconstructed palaeodistributions (shown in full in the appendix) indicate that the majority of species had significantly different geographic ranges in the Late Pleistocene/Early Holocene compared today.
-`r n_reduced_range` of `r nrow(flora)` species are predicted reduced ranges in the past; `r n_reduced_range_10` of more than 10% or more.
-Though the magnitude of the change in range size from prehistory to the present likely reflects a degree of overfitting in the model (discussed further in @sec-discuss-hindcasting), fluctuations in modelled range size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
-The average range of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` during the Younger Dryas (i.e.
-ranges recovered slightly between the Younger Dryas and Early Holocene).
+Our reconstructed palaeodistributions (shown in full in the appendix) indicate that the majority of species had significantly different potential geographic niches in the Late Pleistocene/Early Holocene compared today.
+`r n_reduced_range` of `r nrow(flora)` species are predicted reduced niches in the past; `r n_reduced_range_10` of more than 10% or more.
+Though the magnitude of the change in potential niche size from prehistory to the present likely reflects a degree of overfitting in the model (discussed further in @sec-discuss-hindcasting), fluctuations in modelled niche size between the Bølling-Allerød (14.7–12.9 ka), Younger Dryas (12.9–11.7 ka), and Early Holocene (11.7–8.3 ka) are more directly comparable (@fig-pred-range-density).
+The average potential niche size of modelled species was `r percent(1 - (filter(flora_pred_area_present_summary, period == "eh")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` smaller in the Early Holocene compared to the Bølling–Allerød, and `r percent(1 - (filter(flora_pred_area_present_summary, period == "yds")$med_area_present / filter(flora_pred_area_present_summary, period == "ba")$med_area_present))` smaller during the Younger Dryas.
 This perhaps indicates that although this period is considered one of climatatic amelioration globally [@JonesEtAl2019], the colder conditions of the Pleistocene may have supported more extensive plant-based economies in West Asia specifically.
@@
 #| fig-cap: Predicted species richness (sum of predicted ranges) by period
-# TODO: expand period facet labels
 ggplot() +
@@
 and *Vicia* spp. (vetches), including *Vicia faba* (broad beans).
-Most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted ranges of these species,
+Most of Anatolia, Northern Mesopotamia, and the Zagros Mountains in particular disappear from the predicted niches of these species,
 leaving the Levant and to a lesser extent the Aegean and Cyprus as refugia.
 
-<!-- TODO expand/rewrite -->
 Our results for the Levant are consistent with the current understanding of this region as developing early intensive foraging economies [the Natufian culture, @BarYosef1998] and as a centre of origin of agriculture [@Zeder2011a].
-Within the Levant, many species show moderate reductions in ranges over the Pleistocene/Holocence boundary, retreating from the Badia/transjordan region (e.g.
+Within the Levant, many species show moderate reductions in potential niche size over the Pleistocene/Holocene boundary, retreating from the Badia/transjordan region (e.g.
 barley, fig.
-@fig-palaeodist-barely-levant).
-
-Loss of the Northern Mesopotamia–Anatolia region from the predicted ranges of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture and plant domestication.
+@fig-palaeodist-barley-levant).
+
+Loss of the Northern Mesopotamia–Anatolia region from the predicted potential niches of crop progenitors is interesting in light of the 'golden triangle' hypothesis [@Lev-YadunEtAl2000; @KozlowskiAurenche2005; @AbboEtAl2010], which puts this region at the centre of the development of agriculture and plant domestication.
 Multiple lines of archaeological evidence have emerged that point away from this hypothesis and towards a more geographically diverse origin [@Asouti2006; @FullerEtAl2011; @ArranzOtaeguiEtAl2016], 
@@
 
-The near-absence of the Zagros in any predicted ranges is also surprising, given mounting evidence that animal domestication took place just as early in the eastern *Mashriq* as it did in the west [@Zeder2024].
+The near-absence of the Zagros in any predicted niches is also surprising, given mounting evidence that animal domestication took place just as early in the eastern *Mashriq* as it did in the west [@Zeder2024].
 We consider that the most likely explanation for this is that our flora does not include the species that were most important to plant subsistence in the east.
 Archaeobotanical data on Neolithic sites in the Zagros is limited (compared to the Levant in particular) due to a hiatus in field research there from the 1980s to early 2010s [@Zeder2024].
-Recent research [@RiehlEtAl2013; @WeideEtAl2017; @WeideEtAl2018; @WhitlamEtAl2018; @GonzalezCarreteroEtAl2023] indicates that plant subsistence in this region was based on a distinct set of species, compared to the Levant and Anatolia.
+Recent research [@RiehlEtAl2013; @WeideEtAl2017; @WeideEtAl2018; @WhitlamEtAl2018; @GonzalezCarreteroEtAl2023] indicates that plant subsistence in this region was based on a distinct set of species than that of the Levant and Anatolia.
 
 Cyprus and the Aegean are not conventionally considered part of the primary zone of domestication but rather amongst the first regions that acquired agriculture from West Asia.
-Our analysis complicates this picture, as it indicates that the wild ranges of many crop progenitors included these regions.
+Our analysis complicates this picture, as it indicates that the wild distribution of many crop progenitors included these regions.
 Early examples of several domesticates are recorded at sites on Cyprus, Western Anatolia and Greece [@ArranzOtaeguiRoe2023],
@@
 
-Exceptions to the dominant trend of range reduction include *Cicer reticulatum* (wild chickpea), which has a relatively stable range centered on Northern Mesopotamia;
-and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to have two limited ranges centered around the Black Sea Coast of Anatolia and the Palmyra basin.
+Exceptions to the dominant trend of niche size reduction include *Cicer reticulatum* (wild chickpea), which has a relatively stable potential niche centered on Northern Mesopotamia;
+and *Triticum turgidum dicoccum* (wild emmer wheat), which is predicted to occur in two limited zones centered around the Black Sea Coast of Anatolia and the Palmyra basin.
 In the latter case, neither of these areas are part of the predicted (by our model) modern distribution of wild emmer, which is centered around the Caucasus and Northern Mesopotamia.
@@
 Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
-Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
+Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but the modelled *current* potential niche of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
+It has previously been proposed that the Southern Levant was an "important glacial refuge area for wild cereals" [@RobertsEtAl2018], which accords with these results.
+
 One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye).
@@
 This is consistent with its low ubiquity in archaeobotanical assemblages [@ArranzOtaeguiRoe2023], despite conventionally being considered a 'founder crop', and presumably implies that its domestication was similarly geographically constrained.
-It is the only unambiguous crop progenitor with such a restricted range, though pistachio (*Pistacia atlantica*) and clubrush (*Bolboschoenus maritimus*) are similarly constrained to the Mediterranean coast (and North Africa, in the case of pistachio).
+It is the only unambiguous crop progenitor with such a restricted potential niche, though pistachio (*Pistacia atlantica*) and clubrush (*Bolboschoenus maritimus*) are similarly constrained to the Mediterranean coast (and North Africa, in the case of pistachio).
 This is despite the fact that they are well-attested in the archaeotanical record from across West Asia.
 
-```{r fig-palaeodist-barely-levant, fig.height=10}
-#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
-# TODO: why OOM?`
-#palaeodist_hordeum <- do.call(c, c(
-#  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
-#  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
-#))
-#
-#ggplot() +
-#  facet_grid(taxon ~ fct_relabel(period, toupper)) +
-#  annotation_spatial(ne_land, fill = "white") +
-#  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.), 
-#             na.action = na.omit) +
-#  annotation_spatial(ne_land, fill = NA) +
-#  scale_x_continuous(breaks = c(34, 36, 38)) +
-#  scale_fill_batlow(reverse = TRUE) +
-#  coord_sf(
-#    xlim = c(33, 39), ylim = c(30, 38),
-#    crs = w_asia_albers, default_crs = latlong
-#  ) +
-#  labs(fill = "P(present)", x = NULL, y = NULL) +
-#  theme(
-#    strip.background = element_blank(),
-#    strip.text.y = element_text(face = "italic")
-#  )
-```
-
-Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
-concurrent with it being brought into cultivation (@fig-palaeodist-barely-levant).
+```{r fig-palaeodist-barley-levant, fig.height=10}
+#| fig-cap: Predicted palaeodistribution of wild barleys
+palaeodist_hordeum <- do.call(c, c(
+  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
+  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
+))
+
+ggplot() +
+  facet_grid(taxon ~ fct_relabel(period, toupper)) +
+  annotation_spatial(ne_land, fill = "white") +
+  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.),
+             na.action = na.omit) +
+  annotation_spatial(ne_land, fill = NA) +
+  annotation_spatial(w_asia, fill = NA) +
+  scale_fill_batlow(reverse = TRUE) +
+  coord_sf(crs = w_asia_albers) +
+  labs(fill = "P(present)", x = NULL, y = NULL) +
+  theme(
+    strip.background = element_blank(),
+    strip.text = element_text(face = "italic")
+  )
+```
+
+Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted potential niche from the Pleistocene to the Holocene,
+concurrent with it being brought into cultivation (@fig-palaeodist-barley-levant).
 It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
@@
 #| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
-bread <- c("Aegilops tauschii", "Triticum aestivum", "Triticum turgidum")
+bread <- c("Aegilops tauschii", "Triticum turgidum")
 palaeodist_bread <- do.call(c, c(
@@
 )) |>
+  mutate(present2 = prediction.TRUE. > 0.2) |> # Need lower threshold for wheat
   st_apply(c("x", "y", "period"), sum, na.rm = FALSE) |>
-  mutate(present = factor(present))
+  mutate(present2 = factor(present2))
 
@@
   annotation_spatial(ne_land, fill = "white") +
-  geom_stars(data = select(palaeodist_bread, present), na.action = na.omit) +
+  geom_stars(data = select(palaeodist_bread, present2), na.action = na.omit) +
   annotation_spatial(ne_land, fill = NA) +
@@
 urartu*) and another goatgrass (*Aegilops speltoides*).
-Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to each other in the time frame of domestication is a narrow area around the Orontes river in the northern Levant (@fig-palaeodist-bread).
-This area is c.
-200 km from the only two sites in our archaeobotanical database where bread wheat (*T.
-aestivum*) co-occurs with both of its recent progenitors (*A.
+The predicted potential niches of the two recent progenitors—*T. turgidum dicoccum* and *A. tauschii*—have relateively limited areas of overlap.
+These include the Syrian Jazira (@fig-palaeodist-bread), close to the only two sites in our archaeobotanical database where bread wheat (*T.
+aestivum*) co-occurs with both of the recent progenitors (*A.
 tauschii* and *T.
-turgidum dicoccum*) or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
-Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is possibly occurred in this vicinity.
+turgidum dicoccum*): Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
+Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], in this vicinity.
 
@@
 
-The majority of our models perform poorly as a binary classification with a threshold of `P(present)>0.5`.
-However, this is an arbitrary threshold; lowering it can improve sensitivity at the cost of specificity.
-<!-- TODO: so why didn't we threshold? Losing information - stick with the probabilities. -->
-In principle archaeological occurrences—as true presences—could be used to 'calibrate' the threshold, but there aren't enough archaeological attestations of most of the modelled species.
-
-<!-- TODO: What about not binary? Mean P(T) and P(F) for example? -->
-
-```{r fig-archaeo-sensitivity}
+```{r fig-archaeo-sensitivity, message=FALSE, warning=FALSE}
 #| fig-cap: Summary of hindcasting performance of models in relation to model performance and available archaeological test data
+#| fig-height: 10
 results_summary |>
@@
 
-At this point we cannot explain precisely why this discrepency exists, though assessments of model performance offer some clues.
-
-The hindcasting performance of models is not strongly correlated with either the number of occurrences in the training dataset, the performance of the model on test data, or the number of attested archaeological occurrences [@fig-archaeo-sensitivity].
+The hindcasting performance of models is not strongly correlated with either the number of occurrences in the training dataset, the performance of the model on test data, or the number of attested archaeological occurrences (@fig-archaeo-sensitivity).
 This indicates against a single, simple explanation for the discrepancy, i.e. that one or the other signal is 'wrong'.
 
-<!-- TODO: core issue - fundamental and realised niches, different but related signals -->
-
-For a variety of reasons, our models probably underestimate the *fundamental* niche or potential distribution of the target taxon.
-These include:
-
-* The uneven coverage of the GBIF occurrence dataset (see @sec-occ-data)
-* Use of time-averaged climate slices
+The core issue is that the two sets of evidence discussed here—modelled palaeodistributions and archaeologically-attested palaeooccurrences—are signals of related but distinct phenomena.
+The hindcast niche models are projections of the species' present realised niche into a reconstructed past climatology.
+The archaeobotanical record is an incomplete sample of a past realised niche passed through various anthropic and taphonomic filters.
+Neither can be assumed to be fully accurate representations of the ultimate aim of this study, which is the *fundamental* niche of the species in relation to reconstructable climate and environmental factors – though both carry information about it.
+
+For each specific species there are a variety of potential explanations for why the signals differ, but without further lines of evidence we are not in a position to distinguish them.
+In general, our models probably underestimate the fundamental niche or potential distribution of the target taxon for a variety of reasons:
+
+* The inherent tendency of machine learning models to overfit to the training scenario, despite our efforts to mitigate this;
+* Uneven spatial sampling intensity of the GBIF occurrence dataset (see @sec-occ-data), meaning that some parts of the species' niche are probably either over- and under-represented in the training data;
+* Use of fairly coarse (2000 year) time-averaged climate slices, which capture a general trend but are probably unrepresentative of the environment around sites at the specific time at which they were occupied.
 
@@
 
-* Time-averaged, sometimes quite severely
-* (Understudied) possibility for human transplanting and transport; we know this is the dawn of agriculture!
-* Generally small sample sizes, so individual misidentifications have a large impact
-
-So what can we do about it???
-
-<!-- TODO: rewrite following reviews -->
-The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations. 
-Since they do accurately predict the test dataset<!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
-This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
-Another obvious flaw in our methodology is that the time slices used for palaeoclimatic reconstruction are very broad—each covering around two millennia—and therefore potentially unrepresentative of the environment around sites at the specific time at which they were occupied.
-The variable quality of the archaeological test dataset, especially in terms of chronology, is also a plausible factor.
+* Archaeological deposits are also time-averaged signals, in some cases over quite large spans of time;
+* The chronology attached to our archaeobotanical data is imprecise, e.g. primarily based on point estimates of radiocarbon dates [@Michczynski2007], producing a further averaging effect;
+* Archaeological occurences could be the result of human transplanting, range expansion and/or transport, especially considering this is the period where agriculture began to be practiced;
+* Most archaeobotanical samples in our archaeological test dataset were small, so individual errors and misidentifications have a correspondingly large impact on quantification of model performance.
 
@@
 Human economic choices—mobility, foraging strategies, cultivation, etc.—could also produce archaeobotanical assemblages whose composition depart significantly from that of the surrounding local flora.
-Further refinement of the methodology for hindcast palaeoecological niche models, for example using more finely resolved palaeoclimate sequences [e.g.
-@KargerEtAl2023], hyperparameter tuning to avoid overfitting, and improved archaeological datasets, would help disentangle these potential explanations.
-
-<!-- TODO not Bolboschoenus issue - another cause of potential discrepencies? -->
-
-<!--
-It is consistent with more "macro" trends such as the reduced range of Hordeum and Pistacia correlating with its reduced abundance in the archaeobotanical record.
-TODO: do model predictions match more aggregated stats e.g. biodiversity? -->
+
+What are the implications of this assessment?
+In the immediate term, the modelled palaeodistributions presented here should be taken as conservative estimates – a minimal likely potential niche of the species.
+Closer inspection of discrepencies between the modelled and attested distributions of particular species in relation to specific environmental, taphonomic and cultural factors could yield further insights.
+Future applications of hindcast palaeoecological niche models in archaeology could refine the methodology in several ways: using more finely resolved palaeoclimate sequences [e.g.
+@KargerEtAl2023]; building more refined archaeological chronologies; performing hyperparameter tuning to further mitigate overfitting;^[We did not perform hyperparameter tuning in this study due to computational limitations.] and compiling expanded archaeological datasets that could be use to calibrate binary classification.
 
@@
 
-"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
-Results with regard to specific archaeologically-attested species are not as promising, but this also reflects the incomplete and coarsely temporally-resolved nature of the archaeobotanical verification dataset.
-Discrepancies between the modelled distribution, whether on the broad scale (e.g.
+The modelled palaeodistributions of the species presented here (see appendix) represent plausible minimum potential niche under the average climatic conditions of the Bølling-Allerød, Younger Dryas, and Early Holocene periods.
+The models' generally high performance as assessed against independent contemporary test datasets lend confidence to these predictions.
+Their application as a binary predictor of archaeological presence or absence is significantly less promising, which likely reflects a combination of methodological limitations of our modelling approach, the incomplete and coarsely temporally-resolved nature of the archaeobotanical test dataset, and genuine discrepencies between the species' fundamental niche and its occurrence in the archaeological record.
+Whether on the broad scale (e.g.
 the restricted geographic range of most species compared to their attestation in the archaeological record), or relating to specific species (i.e.
-false positives and false negatives), suggest several avenues for future investigation.
+false positives and false negatives), these discrepencies suggest several avenues for future investigation.
 
@@
 This approach is enabled by the availability of high quality, global open datasets in ecology [@GBIF2025; @GBIFSecretariat2023], earth science [@SRTM], and climatology [@KargerEtAl2017; @BrownEtAl2018].
-Unfortunately in relation to these fields open data in archaeology lags conspiciously behind.
+However, several areas of methodological improvement are evident.
+Notably, the state of open data availability in archaeology lags conspiciously behind the fields mentioned above.
 Though we have benefited from the relatively long tradition of compiling archaeobotanical data in our region of study [@ColledgeEtAl2004; @ShennanConolly2007; @ADEMNES; @LucasFuller2018; @FullerEtAl2018; @WallaceEtAl2018; @ORIGINS],
@@
 
-The data and R code used to produced is archived with Zenodo at <https://doi.org/10.5281/zenodo.14629984>
+The data and R code used to produce this study is archived with Zenodo at <https://doi.org/10.5281/zenodo.14629984>.
+Modelled palaeodistributions in raster format (.TIF) can be found in the same repository.
 

### 2025-08-15 analysis/paper.tex
Finalise resubmission

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   preprint]{elsarticle}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-harv}
 
@@
 \journal{Open Quaternary}
-
-\usepackage[]{natbib}
-\bibliographystyle{elsarticle-harv}
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 }
- \ead{joeroe@hey.com} 
+ \ead{jar@hum.ku.dk} 
 \author[3]{Amaia Arranz-Otaegui%
@@
 This paper presents the first continuous, spatially-explicit
-reconstructions of the palaeodistributions of 82 plant species found
+reconstructions of the palaeodistributions of 65 plant species found
 regularly in association with early agricultural archaeological sites in
@@
 present-day distribution in relation to climate and environmental
-variables. Predictions of the potential ranges of these species at key
+variables. Predictions of the potential niches of these species at key
 stages of the Pleistocene--Holocene transition could then be derived
 from these models using downsampled data from palaeoclimate simulations.
-
-The models predict significant reductions and/or shifts in species
-ranges in the terminal Pleistocene and Early Holocene compared to
-present conditions. Many species that are found throughout the region's
-`hilly flanks' today are indicated to have much more restricted
-distributions centered on the Levant, Cyprus and Western Anatolia. In
-addition, overall ranges shrunk by an average of c.~25\% from the
-terminal Pleistocene to the Early Holocene. The models performance in
-predicting the occurrence of specific species at archaeological sites is
-highly variable, but in aggregate the predictions are coherent and align
-with broad-scale trends in the ubiquity and relative abundance of
-species in the archaeological record.
+Our models performed well against independent contemporary test data,
+but their ability to predict the occurrence of specific species at
+archaeological sites was much more variable, probably reflecting a
+tendency of the method to underestimate the species' fundamental niche.
+Nevertheless, the majority of species are predicted to have had more
+restricted geographic distributions under past climate conditions
+compared to today. Crop progenitors and several wild food species
+modelled are modelled to have been concentrated in the Levant and, to a
+lesser extent, Cyprus and Western Anatolia, suggesting these regions may
+have served as glacial \emph{refugia}. The average size of species'
+niche shrunk by an average of c.~25\% from the terminal Pleistocene to
+the Early Holocene, indicating that economically significant plants were
+adapted to cryo-arid conditions and did not, as often assumed, initially
+respond positively to the `ameliorated' climate of the Holocene.
 \end{abstract}
@@
 ecological niche modelling. Rather than inferring environmental
-conditions from preserved physical evidence, we predict the ranges of
-individual species relevant to human subsistence based on a model of
-their current environmental niche and simulations of past palaeoclimate.
-Though hypothetical, this gives us an independent line of evidence on
-past ecologies that is independent of the environmental archaeological
-and palaeoclimatic records. This means that the ancient data can be
-reserved for assessing the model's ability to `hindcast' past
-conditions. In this sense, discrepencies between the two records are
-perhaps the most interesting result, as they indicate processes
+conditions from preserved physical evidence, we predict the potential
+niche of individual species relevant to human subsistence based on a
+model of their current environmental niche and simulations of past
+palaeoclimate. Though hypothetical, this gives us an independent line of
+evidence on past ecologies that is independent of the environmental
+archaeological and palaeoclimatic records. This means that the ancient
+data can be reserved for assessing the model's ability to `hindcast'
+past conditions. In this sense, discrepencies between the two records
+are perhaps the most interesting result, as they indicate processes
 affecting one or both records that are not fully accounted for and
@@
 palaeodistributions for a large number of species, for the whole region,
-under multiple past climatologies.
+under multiple reconstructed climate scenarios.
 
@@
 Ecological niche modelling or species distribution modelling is widely
-used by ecologists to predict the geographic range of a species based on
-a set of environmental predictors \citep{FranklinMiller2009}.
-Essentially, it involves combining records of where an organism has been
-observed with environmental data (climate, topography, etc.) for those
-locations to model the range of environmental values at which that
-species -- its environmental niche. This model can then be used to
-predict the range of the organism in question either in the same or a
+used by ecologists to mathematically predict the geographic distribution
+of a species from a sample of occurrences
+\citep{FranklinMiller2009, SilleroEtAl2021}. Essentially, it involves
+combining records of where an organism has been observed with
+environmental data (climate, topography, etc.) for those locations to
+model the range of environmental values at which that species -- its
+environmental niche. This model can then be used to predict the
+potential niche of the organism in question either in the same or a
 different environment. \citet{TownsendPetersonSoberon2012} suggests
@@
 convention we follow here when referring to predictive or `hindcast'
-models of past ranges. Within this overarching framework, ecological
+models of past niches. Within this overarching framework, ecological
 niche modelling encompasses a wide range of applications and a variety
@@
 \citep{DavidPollyEronen2011, FranklinEtAl2015}. In the first sense, it
-has been used most extensively to model the range of humans and other
+has been used most extensively to model the niche of humans and other
 hominin species
@@
 
-\caption{\label{fig-region}Late Epipalaeolithic \& Pre-Pottery Neolithic
-archaeological sites used to generate modelled flora}
+\caption{\label{fig-region}Map of the study region (West Asia, grey box)
+with locations of Late Epipalaeolithic and Pre-Pottery Neolithic
+archaeobotanical assemblages.}
 
@@
 on cultivars (i.e.~agriculture, see Section~\ref{sec-bg}). Our starting
-point was a list of 66 taxa (Table~\ref{tbl-results-summary}) comprising
+point was a list of 68 taxa (Table~\ref{tbl-results-summary}) comprising
 the identified species observed at at least 3 Late
@@
 
-\begin{verbatim}
-Rows: 1 Columns: 11
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (5): key, doi, license, status, downloadLink
-dbl  (3): size, totalRecords, numberDatasets
-dttm (3): created, modified, eraseAfter
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-\end{verbatim}
-
 Georeferenced occurrence data for West Eurasia between 0 and 60° of
@@
 \citep{ChamberlainBoettiger2017, rgbif}. The GBIF dataset
-\citep{GBIFDerivedData} excluded fossil occurrences, recorded absences,
+\citep{GBIFDataDownload} excluded fossil occurrences, recorded absences,
 and records with missing or dubious coordinates. Although niche models
@@
 impart information to the model. The resulting cleaned dataset used to
-train our niche models comprises \ensuremath{3.387196\times 10^{6}} from
-4764 constituent datasets.
+train our niche models comprises 3392920 occurrences from 4769
+constituent datasets.
 
@@
 \caption{\label{fig-occ-map}Georeferenced occurrence records from West
-Asia used to train models (N=1409939}
+Eurasia used to train models (N=1412083). Inset, right: prediction
+region (West Asia).}
 
@@
 for the West Asia region, its coverage is uneven both geographically and
-from species to species. Figure~\ref{fig-occ-map} shows clearly that the
-Southern Levant, and Israel specifically, is significantly more densely
-sampled than other parts of West Asia.
+from species to species. The Southern Levant, and Israel specifically,
+is significantly more densely sampled than other parts of West Asia
+(Figure~\ref{fig-occ-map}).
 
@@
 
-\begin{verbatim}
-bio_1, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-\end{verbatim}
-
-\begin{verbatim}
-Warning in parse_netcdf_meta(meta_data, get_names(x)): NAs introduced by
-coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_10, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_11, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_12, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_13, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_14, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_15, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_16, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_17, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_18, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_19, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_2, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_3, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_4, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_5, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_6, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_7, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_8, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
-\begin{verbatim}
-bio_9, 
-\end{verbatim}
-
-\begin{verbatim}
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : GDAL Message 1: dimension #0 (period) is not a Time or
-Vertical dimension.
-Warning in CPL_read_gdal(as.character(x), as.character(options),
-as.character(driver), : NAs introduced by coercion
-\end{verbatim}
-
 \begin{itemize}
@@
 and resolution of 5 km.
-
-\begin{verbatim}
-Warning: There were 57 warnings in `mutate()`.
-The first warning was:
-i In argument: `occ = map(occ, bind_extract, st_as_stars(hack(climate,
-  "period", "cur")))`.
-Caused by warning in `CPL_get_metadata()`:
-! GDAL Message 1: dimension #0 (period) is not a Time or Vertical dimension.
-i Run `dplyr::last_dplyr_warnings()` to see the 56 remaining warnings.
-\end{verbatim}
 
@@
 
-We trained Random Forest models for 63 taxa using contemporary
+We trained Random Forest models for 65 taxa using contemporary
 occurrence data from GBIF, a random sample of background points, and the
@@
 hindcast predictions for reconstructed past environments in 4 key
-climate periods -- a total of 252 modelled palaeodistributions.
+climate periods -- a total of 260 modelled palaeodistributions.
 Predicted distributions for individual taxa are presented in the
@@
 contemporary environment based on the reserved test partition
-Table~\ref{tbl-results-summary}. Model accuracy (proportion of correctly
-classified presence and background samples) ranged between 92\% and
-100\%, with an average of 97\%. Sensitivity (proportion of correctly
-classified presence samples) ranged between 12\% and 100\%, with an
-average of 82\%. The area under the models' receiver operating
-characteristic curves (ROC-AUC) was on average 0.987±0.018 Model
+(Table~\ref{tbl-results-summary}). Model accuracy (proportion of
+correctly classified presence and background samples) ranged between
+93\% and 100\%, with an average of 98\%. Sensitivity (proportion of
+correctly classified presence samples) ranged between 11\% and 100\%,
+with an average of 83\%. The area under the models' receiver operating
+characteristic curves (ROC-AUC) was on average 0.986±0.017. Model
 sensitivity is loosely correlated with the number of occurrences
 available for training (Figure~\ref{fig-model-sensitivity-vs-n_occ}),
-with the worst-performing models all having less than 300 recorded
-occurrences: \emph{Lathyrus sativus}, \emph{Lepidium perfoliatum},
-\emph{Triticum aestivum}, \emph{Bolboschoenus glaucus}, \emph{Gypsophila
-vaccaria}, \emph{Euclidium syriacum}, \emph{Gypsophila elegans},
-\emph{Citrullus colocynthis}, \emph{Triticum monococcum}, \emph{Triticum
-turgidum}, \emph{Aegilops crassa}, \emph{Secale strictum}, \emph{Cicer
-reticulatum}, \emph{Medicago astroites}, \emph{Triticum durum},
-\emph{Gypsophila pilosa}, \emph{Arnebia decumbens}, \emph{Salvia
-absconditiflora}, \emph{Suaeda fruticosa}, \emph{Buglossoides
-tenuiflora}, \emph{Lathyrus oleraceus}, \emph{Vitis sylvestris}, and
-\emph{Vicia orientalis}. Test metrics and ROC curves for the individual
-models are included in the appendix.
+with the worst-performing models all having less than 1000 recorded
+occurrences: \emph{Triticum durum}, \emph{Euclidium syriacum},
+\emph{Triticum turgidum}, \emph{Aegilops crassa}, and \emph{Vicia
+orientalis}. Test metrics and ROC curves for the individual models are
+included in the appendix.
 
 The ability of the hindcast models to predict the occurrence of specific
-species at archaeological sites is worse, with 10\% of presences in
-archaeobotanical assemblages successfully predicted. Model sensitivity
-(proportion of corrected predicted presenses) in relation to the
-archaeological data is on average 0.07±0.15
-Table~\ref{tbl-results-summary}. A full assessment of the hindcasting
+species at archaeological sites is worse, with just 8\% of presences in
+archaeobotanical assemblages successfully predicted at a threshold of
+0.5. Model sensitivity (proportion of corrected predicted presenses) in
+relation to the archaeological data is on average 0.06±0.12
+(Table~\ref{tbl-results-summary}). A full assessment of the hindcasting
 performance of the individual models can be found in the appendix.
+
+In summary, the majority of our models perform poorly as simple binary
+classifiers with a threshold of \texttt{P(present)\textgreater{}0.5}.
+However, this is an arbitrary threshold; lowering it can improve
+sensitivity (proportion of occurrences correctly predicted) at the cost
+of specificity (proportion of non-occurrences incorrectly predicted).
+There are several approaches to selecting informative thresholds for
+binary classification in the ENM literature
+\citep{LiuEtAl2005, LiuEtAl2013, LiuEtAl2015}, with most aiming for an
+optimum compromise between sensitivity and specificity. We were unable
+to find a thresholding technique that significantly improved the
+hindcast models' performance as binary classifiers. By definition,
+binarizing model output also discards information from the underlying
+probabilistic prediction, so we have opted to present results
+probabilistically as far as possible. In principle the archaeological
+occurrences themselves could be used to `calibrate' the hindcast models
+and select an optimum threshold, but there aren't enough archaeological
+attestations of most of the modelled species for this to be practicable
+at this time.
 
@@
 \cmidrule(lr){2-3} \cmidrule(lr){4-7}
-Taxon & Arch. & Cur. & Acc. & ROC-AUC & Sens. & Sens. (arch.) \\ 
+Taxon & Arch. & Cur. & Acc. & ROC-AUC & Sens. cur. & Sens. arch. \\ 
 \midrule\addlinespace[2.5pt]
-{\itshape Aegilops crassa} & 4 & 223 & 0.98 & 0.98 & 0.43 & 0.00 \\ 
-{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 762 & 0.99 & 1.00 & 0.88 & 0.50 \\ 
-{\itshape Ammi majus} & 4 & 10311 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
-{\itshape Androsace maxima} & 17 & 4946 & 0.94 & 0.98 & 0.91 & 0.06 \\ 
-{\itshape Arenaria serpyllifolia} & 3 & 101612 & 0.98 & 0.98 & 1.00 & 0.00 \\ 
-{\itshape Arnebia decumbens} & 21 & 616 & 0.97 & 0.97 & 0.59 & 0.05 \\ 
-{\itshape Arnebia linearifolia} & 13 & 239 & 1.00 & 0.99 & 0.89 & 0.00 \\ 
-{\itshape Atriplex prostrata} & 4 & 80181 & 0.98 & 0.99 & 1.00 & 0.00 \\ 
-{\itshape Avena sterilis} & 5 & 15906 & 0.96 & 0.99 & 0.99 & 0.40 \\ 
-{\itshape Bassia arabica} & 4 & 262 & 1.00 & 1.00 & 0.95 & 0.00 \\ 
-{\itshape Bolboschoenus glaucus} & 5 & 418 & 0.98 & 0.97 & 0.53 & 0.00 \\ 
-{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{2}} & 31 & 53809 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
-{\itshape Brachypodium distachyon} & 5 & 13800 & 0.98 & 1.00 & 0.99 & 0.20 \\ 
+{\itshape Aegilops crassa} & 4 & 223 & 0.98 & 0.98 & 0.26 & 0.00 \\ 
+{\itshape Aegilops speltoides} & 0 & 448 & 0.99 & 0.99 & 0.77 & — \\ 
+{\itshape Aegilops tauschii} & 0 & 1257 & 0.97 & 0.99 & 0.84 & — \\ 
+{\itshape Aizoanthemopsis hispanica}\textsuperscript{\textit{1}} & 4 & 762 & 0.99 & 1.00 & 0.87 & 0.50 \\ 
+{\itshape Ammi majus} & 4 & 10337 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
+{\itshape Androsace maxima} & 17 & 4946 & 0.94 & 0.98 & 0.90 & 0.00 \\ 
+{\itshape Arenaria serpyllifolia} & 3 & 101617 & 0.98 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Arnebia decumbens} & 21 & 617 & 0.97 & 0.97 & 0.59 & 0.05 \\ 
+{\itshape Arnebia linearifolia} & 13 & 239 & 1.00 & 1.00 & 0.83 & 0.00 \\ 
+{\itshape Atriplex prostrata} & 4 & 80186 & 0.99 & 0.99 & 0.99 & 0.00 \\ 
+{\itshape Avena sterilis} & 5 & 15906 & 0.96 & 0.99 & 0.98 & 0.40 \\ 
+{\itshape Bassia arabica} & 4 & 262 & 1.00 & 0.99 & 0.88 & 0.00 \\ 
+{\itshape Bolboschoenus glaucus} & 5 & 418 & 0.98 & 0.97 & 0.63 & 0.00 \\ 
+{\itshape Bolboschoenus maritimus}\textsuperscript{\textit{2}} & 31 & 53813 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
+{\itshape Brachypodium distachyon} & 5 & 13800 & 0.98 & 0.99 & 0.99 & 0.20 \\ 
 {\itshape Bromus sterilis} & 3 & 89764 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
 {\itshape Buglossoides arvensis} & 23 & 23004 & 0.95 & 0.98 & 0.98 & 0.09 \\ 
-{\itshape Buglossoides tenuiflora} & 26 & 399 & 0.99 & 0.99 & 0.71 & 0.04 \\ 
-{\itshape Capparis spinosa} & 4 & 8617 & 0.97 & 0.99 & 0.96 & 0.00 \\ 
-{\itshape Carex divisa} & 9 & 7400 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
-{\itshape Chenopodium album} & 5 & 184403 & 0.99 & 0.98 & 1.00 & 0.00 \\ 
-{\itshape Cicer reticulatum}\textsuperscript{\textit{3}} & 3 & 52 & 1.00 & 1.00 & 0.69 & 0.00 \\ 
-{\itshape Citrullus colocynthis} & 4 & 1823 & 0.94 & 0.95 & 0.69 & 0.00 \\ 
+{\itshape Buglossoides tenuiflora} & 26 & 399 & 0.99 & 0.98 & 0.79 & 0.04 \\ 
+{\itshape Capparis spinosa} & 4 & 8617 & 0.96 & 0.99 & 0.95 & 0.00 \\ 
+{\itshape Carex divisa} & 9 & 7400 & 0.96 & 0.99 & 0.97 & 0.00 \\ 
+{\itshape Chenopodium album} & 5 & 184412 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Cicer reticulatum}\textsuperscript{\textit{3}} & 3 & 52 & 1.00 & 1.00 & 1.00 & 0.00 \\ 
+{\itshape Citrullus colocynthis} & 4 & 1823 & 0.94 & 0.96 & 0.71 & 0.00 \\ 
 {\itshape Crithopsis delileana} & 3 & 430 & 1.00 & 1.00 & 0.93 & 0.00 \\ 
-{\itshape Euclidium syriacum} & 3 & 338 & 0.98 & 0.97 & 0.27 & 0.00 \\ 
-{\itshape Ficus carica} & 8 & 64424 & 0.98 & 0.99 & 0.99 & 0.38 \\ 
-{\itshape Fumaria densiflora} & 4 & 3704 & 0.96 & 0.99 & 0.94 & 0.00 \\ 
-{\itshape Gypsophila elegans} & 3 & 1603 & 0.95 & 0.98 & 0.81 & 0.00 \\ 
-{\itshape Gypsophila pilosa} & 5 & 348 & 0.99 & 0.99 & 0.67 & 0.00 \\ 
-{\itshape Gypsophila vaccaria}\textsuperscript{\textit{4}} & 8 & 1177 & 0.95 & 0.98 & 0.65 & 0.00 \\ 
+{\itshape Euclidium syriacum} & 3 & 338 & 0.98 & 0.96 & 0.31 & 0.00 \\ 
+{\itshape Ficus carica} & 8 & 64440 & 0.98 & 0.99 & 0.99 & 0.38 \\ 
+{\itshape Fumaria densiflora} & 4 & 3704 & 0.96 & 0.99 & 0.95 & 0.00 \\ 
+{\itshape Gypsophila elegans} & 3 & 1613 & 0.96 & 0.99 & 0.83 & 0.00 \\ 
+{\itshape Gypsophila pilosa} & 5 & 348 & 0.99 & 0.99 & 0.69 & 0.00 \\ 
+{\itshape Gypsophila vaccaria}\textsuperscript{\textit{4}} & 8 & 1177 & 0.95 & 0.96 & 0.69 & 0.00 \\ 
 {\itshape Henrardia pubescens} & 3 & 1\textsuperscript{\textit{5}} & — & — & — & — \\ 
-{\itshape Hordeum bulbosum} & 4 & 4134 & 0.97 & 1.00 & 0.95 & 0.00 \\ 
+{\itshape Hordeum bulbosum} & 4 & 4134 & 0.97 & 0.99 & 0.94 & 0.00 \\ 
 {\itshape Hordeum murinum} & 5 & 74770 & 0.98 & 0.99 & 1.00 & 0.00 \\ 
-{\itshape Hordeum spontaneum}\textsuperscript{\textit{6}} & 76 & 3098 & 0.99 & 1.00 & 0.97 & 0.20 \\ 
+{\itshape Hordeum spontaneum}\textsuperscript{\textit{6}} & 76 & 3098 & 0.99 & 1.00 & 0.98 & 0.21 \\ 
 {\itshape Lathyrus aphaca} & 4 & 18652 & 0.96 & 0.99 & 0.98 & 0.00 \\ 
-{\itshape Lathyrus oleraceus}\textsuperscript{\textit{7}} & 8 & 819 & 0.98 & 0.99 & 0.80 & 0.00 \\ 
-{\itshape Lathyrus sativus} & 4 & 2120 & 0.92 & 0.96 & 0.75 & 0.00 \\ 
+{\itshape Lathyrus oleraceus}\textsuperscript{\textit{7}} & 8 & 819 & 0.98 & 0.99 & 0.78 & 0.00 \\ 
+{\itshape Lathyrus sativus} & 4 & 2120 & 0.93 & 0.97 & 0.77 & 0.00 \\ 
 {\itshape Lepidium perfoliatum} & 3 & 2764 & 0.94 & 0.98 & 0.84 & 0.00 \\ 
@@
 {\itshape Lolium rigidum} & 5 & 10620 & 0.97 & 0.99 & 0.98 & 0.00 \\ 
-{\itshape Lolium temulentum} & 3 & 4463 & 0.94 & 0.98 & 0.93 & 0.00 \\ 
-{\itshape Medicago astroites}\textsuperscript{\textit{9}} & 15 & 104 & 1.00 & 0.99 & 0.64 & 0.00 \\ 
-{\itshape Medicago radiata} & 20 & 834 & 0.99 & 0.99 & 0.91 & 0.30 \\ 
-{\itshape Phalaris paradoxa} & 3 & 3287 & 0.97 & 0.99 & 0.94 & 0.00 \\ 
-{\itshape Phragmites australis} & 4 & 373041 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
-{\itshape Pistacia atlantica} & 6 & 2419 & 0.99 & 1.00 & 0.94 & 0.33 \\ 
+{\itshape Lolium temulentum} & 3 & 4463 & 0.94 & 0.98 & 0.94 & 0.00 \\ 
+{\itshape Medicago radiata} & 20 & 834 & 0.98 & 1.00 & 0.90 & 0.30 \\ 
+{\itshape Phalaris paradoxa} & 3 & 3287 & 0.96 & 0.99 & 0.94 & 0.00 \\ 
+{\itshape Phragmites australis} & 4 & 373066 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Pistacia atlantica} & 6 & 2419 & 0.98 & 1.00 & 0.93 & 0.33 \\ 
 {\itshape Poa bulbosa} & 5 & 30769 & 0.96 & 0.98 & 0.99 & 0.20 \\ 
-{\itshape Polygonum arenarium}\textsuperscript{\textit{10}} & 7 & 25\textsuperscript{\textit{5}} & — & — & — & — \\ 
+{\itshape Polygonum arenarium}\textsuperscript{\textit{9}} & 7 & 25\textsuperscript{\textit{5}} & — & — & — & — \\ 
 {\itshape Polygonum corrigioloides} & 6 & 12\textsuperscript{\textit{5}} & — & — & — & — \\ 
-{\itshape Prosopis farcta} & 5 & 2512 & 0.99 & 1.00 & 0.96 & 0.60 \\ 
+{\itshape Prosopis farcta} & 5 & 2512 & 0.99 & 1.00 & 0.96 & 0.00 \\ 
 {\itshape Rumex pulcher} & 6 & 17606 & 0.97 & 0.99 & 0.99 & 0.00 \\ 
-{\itshape Salsola kali} & 6 & 14699 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
-{\itshape Salvia absconditiflora}\textsuperscript{\textit{11}} & 3 & 128 & 0.99 & 1.00 & 0.65 & 0.00 \\ 
-{\itshape Secale cereale} & 4 & 14803 & 0.95 & 0.98 & 0.97 & 0.00 \\ 
-{\itshape Secale strictum}\textsuperscript{\textit{12}} & 3 & 162 & 0.99 & 0.99 & 0.50 & 0.00 \\ 
-{\itshape Suaeda fruticosa} & 3 & 577 & 0.99 & 1.00 & 0.79 & 0.00 \\ 
-{\itshape Taeniatherum caput-medusae} & 4 & 1969 & 0.96 & 0.99 & 0.89 & 0.00 \\ 
-{\itshape Triticum aestivum}\textsuperscript{\textit{13}} & 4 & 217 & 0.99 & 0.98 & 0.59 & 0.00 \\ 
-{\itshape Triticum durum} & 3 & 64 & 1.00 & 0.99 & 0.50 & 0.00 \\ 
-{\itshape Triticum monococcum}\textsuperscript{\textit{14}} & 47 & 870 & 0.98 & 0.99 & 0.80 & 0.15 \\ 
-{\itshape Triticum turgidum}\textsuperscript{\textit{15}} & 53 & 345 & 0.97 & 0.93 & 0.22 & 0.00 \\ 
-{\itshape Verbena officinalis} & 3 & 89654 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
-{\itshape Vicia ervilia} & 26 & 1924 & 0.95 & 0.98 & 0.86 & 0.35 \\ 
-{\itshape Vicia faba} & 7 & 42636 & 0.97 & 0.99 & 0.99 & 0.43 \\ 
-{\itshape Vicia narbonensis}\textsuperscript{\textit{16}} & 3 & 2826 & 0.95 & 0.99 & 0.89 & 0.33 \\ 
-{\itshape Vicia orientalis}\textsuperscript{\textit{17}} & 16 & 41 & 1.00 & 0.95 & 0.12 & 0.00 \\ 
-{\itshape Vitis sylvestris} & 3 & 63 & 1.00 & 0.99 & 0.36 & 0.00 \\ 
-{\itshape Zygophyllum fabago} & 3 & 3554 & 0.97 & 0.99 & 0.91 & 0.00 \\ 
+{\itshape Salsola kali} & 6 & 14701 & 0.98 & 1.00 & 0.99 & 0.00 \\ 
+{\itshape Salvia absconditiflora}\textsuperscript{\textit{10}} & 3 & 128 & 1.00 & 1.00 & 0.81 & 0.00 \\ 
+{\itshape Secale cereale} & 4 & 14813 & 0.95 & 0.98 & 0.98 & 0.00 \\ 
+{\itshape Secale strictum}\textsuperscript{\textit{11}} & 3 & 162 & 0.99 & 0.98 & 0.62 & 0.00 \\ 
+{\itshape Suaeda fruticosa} & 3 & 577 & 0.99 & 0.99 & 0.76 & 0.00 \\ 
+{\itshape Taeniatherum caput-medusae} & 4 & 1969 & 0.97 & 0.99 & 0.90 & 0.00 \\ 
+{\itshape Triticum aestivum}\textsuperscript{\textit{12}} & 4 & 217 & 0.99 & 0.96 & 0.50 & 0.00 \\ 
+{\itshape Triticum durum} & 3 & 65 & 0.99 & 0.95 & 0.11 & 0.00 \\ 
+{\itshape Triticum monococcum}\textsuperscript{\textit{13}} & 47 & 870 & 0.97 & 0.99 & 0.80 & 0.04 \\ 
+{\itshape Triticum turgidum}\textsuperscript{\textit{14}} & 53 & 345 & 0.98 & 0.91 & 0.25 & 0.00 \\ 
+{\itshape Triticum urartu} & 0 & 420 & 1.00 & 1.00 & 0.95 & — \\ 
+{\itshape Verbena officinalis} & 3 & 89655 & 0.99 & 0.99 & 1.00 & 0.00 \\ 
+{\itshape Vicia ervilia} & 26 & 1924 & 0.95 & 0.98 & 0.84 & 0.27 \\ 
+{\itshape Vicia faba} & 7 & 42644 & 0.97 & 0.99 & 0.99 & 0.29 \\ 
+{\itshape Vicia narbonensis}\textsuperscript{\textit{15}} & 3 & 2826 & 0.95 & 0.99 & 0.90 & 0.33 \\ 
+{\itshape Vicia orientalis}\textsuperscript{\textit{16}} & 16 & 41 & 1.00 & 0.97 & 0.14 & 0.00 \\ 
+{\itshape Vitis sylvestris} & 3 & 63 & 1.00 & 1.00 & 0.50 & 0.00 \\ 
+{\itshape Zygophyllum fabago} & 3 & 3554 & 0.97 & 0.99 & 0.94 & 0.00 \\ 
 \bottomrule
@@
 \textsuperscript{\textit{8}}Including \emph{Linum usitatissimum}\\
-\textsuperscript{\textit{9}}Including \emph{Trigonella astroites}\\
-\textsuperscript{\textit{10}}Including \emph{Polygonum venantianum}\\
-\textsuperscript{\textit{11}}Including \emph{Salvia cryptantha}\\
-\textsuperscript{\textit{12}}Including \emph{Secale montanum}\\
-\textsuperscript{\textit{13}}Including \emph{Triticum spelta} and \emph{Triticum aestivocompactum}\\
-\textsuperscript{\textit{14}}Including \emph{Triticum boeoticum}\\
-\textsuperscript{\textit{15}}Including \emph{Triticum aestivum}, \emph{Triticum dicoccum}, and \emph{Triticum dicoccoides}\\
-\textsuperscript{\textit{16}}Including \emph{Vicia narbonense}\\
-\textsuperscript{\textit{17}}Including \emph{Lens culinaris} and \emph{Lens orientalis}\\
+\textsuperscript{\textit{9}}Including \emph{Polygonum venantianum}\\
+\textsuperscript{\textit{10}}Including \emph{Salvia cryptantha}\\
+\textsuperscript{\textit{11}}Including \emph{Secale montanum}\\
+\textsuperscript{\textit{12}}Including \emph{Triticum spelta} and \emph{Triticum aestivocompactum}\\
+\textsuperscript{\textit{13}}Including \emph{Triticum boeoticum}\\
+\textsuperscript{\textit{14}}Including \emph{Triticum aestivum}, \emph{Triticum dicoccum}, and \emph{Triticum dicoccoides}\\
+\textsuperscript{\textit{15}}Including \emph{Vicia narbonense}\\
+\textsuperscript{\textit{16}}Including \emph{Lens culinaris} and \emph{Lens orientalis}\\
 \end{minipage}
@@
 
-\subsection{Reduction in range sizes over the Pleistocene/Holocene
-boundary}\label{reduction-in-range-sizes-over-the-pleistoceneholocene-boundary}
+\subsection{Reduction in potential niche sizes over the
+Pleistocene/Holocene
+boundary}\label{reduction-in-potential-niche-sizes-over-the-pleistoceneholocene-boundary}
 
@@
 \caption{\label{fig-pred-range-density}Distribution of predicted species
-ranges by period. Dashed lines indicate the median range.}
+niche size by period (logarithmic scale). Dashed lines indicate the
+median niche size.}
 
@@
 indicate that the majority of species had significantly different
-geographic ranges in the Late Pleistocene/Early Holocene compared today.
-55 of 63 species are predicted reduced ranges in the past; 54 of more
-than 10\% or more. Though the magnitude of the change in range size from
-prehistory to the present likely reflects a degree of overfitting in the
-model (discussed further in Section~\ref{sec-discuss-hindcasting}),
-fluctuations in modelled range size between the Bølling-Allerød
-(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
-(11.7--8.3 ka) are more directly comparable
-(Figure~\ref{fig-pred-range-density}). The average range of modelled
-species was 31\% in the Early Holocene compared to the Bølling--Allerød,
-and 28\% during the Younger Dryas (i.e.~ranges recovered slightly
-between the Younger Dryas and Early Holocene). This perhaps indicates
-that although this period is considered one of climatatic amelioration
-globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
-may have supported more extensive plant-based economies in West Asia
-specifically.
+potential geographic niches in the Late Pleistocene/Early Holocene
+compared today. 55 of 65 species are predicted reduced niches in the
+past; 52 of more than 10\% or more. Though the magnitude of the change
+in potential niche size from prehistory to the present likely reflects a
+degree of overfitting in the model (discussed further in
+Section~\ref{sec-discuss-hindcasting}), fluctuations in modelled niche
+size between the Bølling-Allerød (14.7--12.9 ka), Younger Dryas
+(12.9--11.7 ka), and Early Holocene (11.7--8.3 ka) are more directly
+comparable (Figure~\ref{fig-pred-range-density}). The average potential
+niche size of modelled species was 35\% smaller in the Early Holocene
+compared to the Bølling--Allerød, and 22\% smaller during the Younger
+Dryas. This perhaps indicates that although this period is considered
+one of climatatic amelioration globally \citep{JonesEtAl2019}, the
+colder conditions of the Pleistocene may have supported more extensive
+plant-based economies in West Asia specifically.
 
@@
 Northern Mesopotamia, and the Zagros Mountains in particular disappear
-from the predicted ranges of these species, leaving the Levant and to a
+from the predicted niches of these species, leaving the Levant and to a
 lesser extent the Aegean and Cyprus as refugia.
@@
 of agriculture \citep{Zeder2011a}. Within the Levant, many species show
-moderate reductions in ranges over the Pleistocene/Holocence boundary,
-retreating from the Badia/transjordan region (e.g.~barley, fig.
-\textbf{?@fig-palaeodist-barely-levant}).
+moderate reductions in potential niche size over the
+Pleistocene/Holocene boundary, retreating from the Badia/transjordan
+region (e.g.~barley, fig. Figure~\ref{fig-palaeodist-barley-levant}).
 
 Loss of the Northern Mesopotamia--Anatolia region from the predicted
-ranges of crop progenitors is interesting in light of the `golden
-triangle' hypothesis
+potential niches of crop progenitors is interesting in light of the
+`golden triangle' hypothesis
 \citep{Lev-YadunEtAl2000, KozlowskiAurenche2005, AbboEtAl2010}, which
@@
 
-The near-absence of the Zagros in any predicted ranges is also
+The near-absence of the Zagros in any predicted niches is also
 surprising, given mounting evidence that animal domestication took place
@@
 indicates that plant subsistence in this region was based on a distinct
-set of species, compared to the Levant and Anatolia.
+set of species than that of the Levant and Anatolia.
 
@@
 acquired agriculture from West Asia. Our analysis complicates this
-picture, as it indicates that the wild ranges of many crop progenitors
-included these regions. Early examples of several domesticates are
-recorded at sites on Cyprus, Western Anatolia and Greece
-\citep{ArranzOtaeguiRoe2023}, and the Aegean region was probably
+picture, as it indicates that the wild distribution of many crop
+progenitors included these regions. Early examples of several
+domesticates are recorded at sites on Cyprus, Western Anatolia and
+Greece \citep{ArranzOtaeguiRoe2023}, and the Aegean region was probably
 connected to West Asia by a land bridge via Anatolia until the Early
 Holocene \citep{AksuHiscott2022}. Were these area part of the same
-broader `interaction sphere' that produce Neolithic agriculture in West
+broader `interaction sphere' that produced Neolithic agriculture in West
 Asia?
 
-Exceptions to the dominant trend of range reduction include \emph{Cicer
-reticulatum} (wild chickpea), which has a relatively stable range
-centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
-(wild emmer wheat), which is predicted to have two limited ranges
-centered around the Black Sea Coast of Anatolia and the Palmyra basin.
-In the latter case, neither of these areas are part of the predicted (by
-our model) modern distribution of wild emmer, which is centered around
-the Caucasus and Northern Mesopotamia. But it would be consistent with
-archaeological evidence for early cultivation at sites in the Upper
-Euphrates \citep{Willcox2024}.
+Exceptions to the dominant trend of niche size reduction include
+\emph{Cicer reticulatum} (wild chickpea), which has a relatively stable
+potential niche centered on Northern Mesopotamia; and \emph{Triticum
+turgidum dicoccum} (wild emmer wheat), which is predicted to occur in
+two limited zones centered around the Black Sea Coast of Anatolia and
+the Palmyra basin. In the latter case, neither of these areas are part
+of the predicted (by our model) modern distribution of wild emmer, which
+is centered around the Caucasus and Northern Mesopotamia. But it would
+be consistent with archaeological evidence for early cultivation at
+sites in the Upper Euphrates \citep{Willcox2024}.
 
@@
 with the fact that both our initial flora and training occurrence
-dataset have a strong bias towards the southern Levant, but strikingly
-the modelled \emph{current} ranges of these plants do tend to include
-Anatolia and the Zagros, so this cannot be the only factor. One notable
-exception is the wild ancestor of chickpea (\emph{Cicer reticulatum}),
-which is predicted to have a distribution centered on Northern
-Mesopotamia but encompassing much of the the `hilly flanks' (except the
-southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}). Another is
-rye (\emph{Secale cereale}), which is inferred to be primarily Anatolian
-(Figure~\ref{fig-palaeodist-chickpea-rye}). This is perhaps relevant to
-rye's unusual domestication history, as a crop of West Asian origin that
-was intensively exploited \citep{HillmanEtAl2001, DoucheWillcox2018} but
-apparently not first cultivated until much later than the `founder
-crops', in Europe \citep{SchreiberEtAl2021}.
+dataset have a strong bias towards the southern Levant, but the modelled
+\emph{current} potential niche of these plants do tend to include
+Anatolia and the Zagros, so this cannot be the only factor. It has
+previously been proposed that the Southern Levant was an ``important
+glacial refuge area for wild cereals'' \citep{RobertsEtAl2018}, which
+accords with these results.
+
+One notable exception is the wild ancestor of chickpea (\emph{Cicer
+reticulatum}), which is predicted to have a distribution centered on
+Northern Mesopotamia but encompassing much of the the `hilly flanks'
+(except the southern Levant, Figure~\ref{fig-palaeodist-chickpea-rye}).
+Another is rye (\emph{Secale cereale}), which is inferred to be
+primarily Anatolian (Figure~\ref{fig-palaeodist-chickpea-rye}). This is
+perhaps relevant to rye's unusual domestication history, as a crop of
+West Asian origin that was intensively exploited
+\citep{HillmanEtAl2001, DoucheWillcox2018} but apparently not first
+cultivated until much later than the `founder crops', in Europe
+\citep{SchreiberEtAl2021}.
 
@@
 that its domestication was similarly geographically constrained. It is
-the only unambiguous crop progenitor with such a restricted range,
-though pistachio (\emph{Pistacia atlantica}) and clubrush
+the only unambiguous crop progenitor with such a restricted potential
+niche, though pistachio (\emph{Pistacia atlantica}) and clubrush
 (\emph{Bolboschoenus maritimus}) are similarly constrained to the
@@
 
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-palaeodist-barley-levant-1.pdf}}
+
+}
+
+\caption{\label{fig-palaeodist-barley-levant}Predicted
+palaeodistribution of wild barleys}
+
+\end{figure}%
+
 Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
-of their predicted ranges from the Pleistocene to the Holocene,
+of their predicted potential niche from the Pleistocene to the Holocene,
 concurrent with it being brought into cultivation
-(\textbf{?@fig-palaeodist-barely-levant}). It also sees a marked decline
-in the archaeobotanical record from the Early PPNA/Early PPNB (where it
-was amongst the most common taxa) to the Late PPNB and Late Neolithic
-\citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia atlantica})
-shows similar trends, but it is less certain that this species was
-managed in the Neolithic. Conversely, the various wild wheat species
-native to West Asia show almost no response to Pleistocene/Holocene
-climate change, even within the Levant, and in the archaeobotanical
-record wheat displays the opposite trend to barley and pistachio --
-becoming gradually more abundant through the course of the Neolithic and
-dominant by its end \citep{ArranzOtaeguiRoe2023}.
+(Figure~\ref{fig-palaeodist-barley-levant}). It also sees a marked
+decline in the archaeobotanical record from the Early PPNA/Early PPNB
+(where it was amongst the most common taxa) to the Late PPNB and Late
+Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
+atlantica}) shows similar trends, but it is less certain that this
+species was managed in the Neolithic. Conversely, the various wild wheat
+species native to West Asia show almost no response to
+Pleistocene/Holocene climate change, even within the Levant, and in the
+archaeobotanical record wheat displays the opposite trend to barley and
+pistachio -- becoming gradually more abundant through the course of the
+Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}.
 
@@
 (\emph{T. urartu}) and another goatgrass (\emph{Aegilops speltoides}).
-Although there is fairly wide overlap between these species today,
-according to our modelling the only place all four (or even just three)
-are predicted to have coincided or been found in close proximity to each
-other in the time frame of domestication is a narrow area around the
-Orontes river in the northern Levant
-(Figure~\ref{fig-palaeodist-bread}). This area is c.~200 km from the
-only two sites in our archaeobotanical database where bread wheat
-(\emph{T. aestivum}) co-occurs with both of its recent progenitors
-(\emph{A. tauschii} and \emph{T. turgidum dicoccum}) or both species of
-wheat and the goatgrasses are found together: Tell Abu Hureyra and El
-Kowm II \citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to
-  Benjamin Nowak for pointing this out to us.} Taken together this
-suggests an origin of domesticated bread wheat, otherwise only loosely
-geographically constrained to the Levant--Upper Euphrates corridor
-\citep{LevyFeldman2022}, is possibly occurred in this vicinity.
-
-\subsection{Hindcast models do not predict archaeobotanical
+The predicted potential niches of the two recent progenitors---\emph{T.
+turgidum dicoccum} and \emph{A. tauschii}---have relateively limited
+areas of overlap. These include the Syrian Jazira
+(Figure~\ref{fig-palaeodist-bread}), close to the only two sites in our
+archaeobotanical database where bread wheat (\emph{T. aestivum})
+co-occurs with both of the recent progenitors (\emph{A. tauschii} and
+\emph{T. turgidum dicoccum}): Tell Abu Hureyra and El Kowm II
+\citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to Benjamin Nowak
+  for pointing this out to us.} Taken together this suggests an origin
+of domesticated bread wheat, otherwise only loosely geographically
+constrained to the Levant--Upper Euphrates corridor
+\citep{LevyFeldman2022}, in this vicinity.
+
+\subsection{Disagreement between hindcast models and archaeobotanical
 composition}\label{sec-discuss-hindcasting}
 
-The failure of our hindcast models to predict the occurrence of species
-in archaeobotanical assemblages has several possible explanations. Since
-they do accurately predict the test dataset, a likely culprit is
-overfitting of the models to the present environment. This implies that
-the modelled palaeodistributions should be seen as conservative
-estimates or a minimal range. Another obvious flaw in our methodology is
-that the time slices used for palaeoclimatic reconstruction are very
-broad---each covering around two millennia---and therefore potentially
-unrepresentative of the environment around sites at the specific time at
-which they were occupied. The variable quality of the archaeological
-test dataset, especially in terms of chronology, is also a plausible
-factor.
-
-For a variety of reasons, our models almost certainly underestimate the
-\emph{fundamental} niche or potential distribution of the target taxon.
-These include:
+In general there is middling to low agreement between our hindcast
+predictions of species niches and observed archaeological occurrences of
+the same species \citep{tlb-results-summary}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-archaeo-sensitivity-1.pdf}}
+
+}
+
+\caption{\label{fig-archaeo-sensitivity}Summary of hindcasting
+performance of models in relation to model performance and available
+archaeological test data}
+
+\end{figure}%
+
+The hindcasting performance of models is not strongly correlated with
+either the number of occurrences in the training dataset, the
+performance of the model on test data, or the number of attested
+archaeological occurrences (Figure~\ref{fig-archaeo-sensitivity}). This
+indicates against a single, simple explanation for the discrepancy,
+i.e.~that one or the other signal is `wrong'.
+
+The core issue is that the two sets of evidence discussed
+here---modelled palaeodistributions and archaeologically-attested
+palaeooccurrences---are signals of related but distinct phenomena. The
+hindcast niche models are projections of the species' present realised
+niche into a reconstructed past climatology. The archaeobotanical record
+is an incomplete sample of a past realised niche passed through various
+anthropic and taphonomic filters. Neither can be assumed to be fully
+accurate representations of the ultimate aim of this study, which is the
+\emph{fundamental} niche of the species in relation to reconstructable
+climate and environmental factors -- though both carry information about
+it.
+
+For each specific species there are a variety of potential explanations
+for why the signals differ, but without further lines of evidence we are
+not in a position to distinguish them. In general, our models probably
+underestimate the fundamental niche or potential distribution of the
+target taxon for a variety of reasons:
 
@@
 \item
-  The uneven coverage of the GBIF occurrence dataset (see
-  Section~\ref{sec-occ-data})
+  The inherent tendency of machine learning models to overfit to the
+  training scenario, despite our efforts to mitigate this;
+\item
+  Uneven spatial sampling intensity of the GBIF occurrence dataset (see
+  Section~\ref{sec-occ-data}), meaning that some parts of the species'
+  niche are probably either over- and under-represented in the training
+  data;
+\item
+  Use of fairly coarse (2000 year) time-averaged climate slices, which
+  capture a general trend but are probably unrepresentative of the
+  environment around sites at the specific time at which they were
+  occupied.
+\end{itemize}
+
+At the same time, there are several reasons to believe that the
+archaeobotanical data (at least as it has been compiled for this study)
+\emph{overestimates} the fundamental niche:
+
+\begin{itemize}
+\tightlist
+\item
+  Archaeological deposits are also time-averaged signals, in some cases
+  over quite large spans of time;
+\item
+  The chronology attached to our archaeobotanical data is imprecise,
+  e.g.~primarily based on point estimates of radiocarbon dates
+  \citep{Michczynski2007}, producing a further averaging effect;
+\item
+  Archaeological occurences could be the result of human transplanting,
+  range expansion and/or transport, especially considering this is the
+  period where agriculture began to be practiced;
+\item
+  Most archaeobotanical samples in our archaeological test dataset were
+  small, so individual errors and misidentifications have a
+  correspondingly large impact on quantification of model performance.
 \end{itemize}
@@
 assemblages whose composition depart significantly from that of the
-surrounding local flora. Further refinement of the methodology for
-hindcast palaeoecological niche models, for example using more finely
-resolved palaeoclimate sequences \citep[e.g.][]{KargerEtAl2023},
-hyperparameter tuning to avoid overfitting, and improved archaeological
-datasets, would help disentangle these potential explanations.
+surrounding local flora.
+
+What are the implications of this assessment? In the immediate term, the
+modelled palaeodistributions presented here should be taken as
+conservative estimates -- a minimal likely potential niche of the
+species. Closer inspection of discrepencies between the modelled and
+attested distributions of particular species in relation to specific
+environmental, taphonomic and cultural factors could yield further
+insights. Future applications of hindcast palaeoecological niche models
+in archaeology could refine the methodology in several ways: using more
+finely resolved palaeoclimate sequences \citep[e.g.][]{KargerEtAl2023};
+building more refined archaeological chronologies; performing
+hyperparameter tuning to further mitigate overfitting;\footnote{We did
+  not perform hyperparameter tuning in this study due to computational
+  limitations.} and compiling expanded archaeological datasets that
+could be use to calibrate binary classification.
 
@@
 We present the first continuous, spatially explicit models of the
-palaeodistributions of 63 plant species found regularly in association
+palaeodistributions of 65 plant species found regularly in association
 with Late Epipalaeolithic and Early Neolithic sites in West Asia. This
@@
 
-``All models are wrong'' \citep{Box1976}, but the performance of our
-models on independent test datasets give confidence in its predictions,
-which are plausible and consistent with broad-scale patterns in the
-archaeological record. Results with regard to specific
-archaeologically-attested species are not as promising, but this also
-reflects the incomplete and coarsely temporally-resolved nature of the
-archaeobotanical verification dataset. Discrepancies between the
-modelled distribution, whether on the broad scale (e.g.~the restricted
-geographic range of most species compared to their attestation in the
-archaeological record), or relating to specific species (i.e.~false
-positives and false negatives), suggest several avenues for future
+The modelled palaeodistributions of the species presented here (see
+appendix) represent plausible minimum potential niche under the average
+climatic conditions of the Bølling-Allerød, Younger Dryas, and Early
+Holocene periods. The models' generally high performance as assessed
+against independent contemporary test datasets lend confidence to these
+predictions. Their application as a binary predictor of archaeological
+presence or absence is significantly less promising, which likely
+reflects a combination of methodological limitations of our modelling
+approach, the incomplete and coarsely temporally-resolved nature of the
+archaeobotanical test dataset, and genuine discrepencies between the
+species' fundamental niche and its occurrence in the archaeological
+record. Whether on the broad scale (e.g.~the restricted geographic range
+of most species compared to their attestation in the archaeological
+record), or relating to specific species (i.e.~false positives and false
+negatives), these discrepencies suggest several avenues for future
 investigation.
@@
 \citep{SRTM}, and climatology \citep{KargerEtAl2017, BrownEtAl2018}.
-Unfortunately in relation to these fields open data in archaeology lags
-conspiciously behind. Though we have benefited from the relatively long
-tradition of compiling archaeobotanical data in our region of study
+However, several areas of methodological improvement are evident.
+Notably, the state of open data availability in archaeology lags
+conspiciously behind the fields mentioned above. Though we have
+benefited from the relatively long tradition of compiling
+archaeobotanical data in our region of study
 \citep{ColledgeEtAl2004, ShennanConolly2007, ADEMNES, LucasFuller2018, FullerEtAl2018, WallaceEtAl2018, ORIGINS},
@@
 
-The data and R code used to produced is archived with Zenodo at
-\url{https://doi.org/10.5281/zenodo.14629984}
+The data and R code used to produce this study is archived with Zenodo
+at \url{https://doi.org/10.5281/zenodo.14629984}. Modelled
+palaeodistributions in raster format (.TIF) can be found in the same
+repository.
 
@@
 \renewcommand\refname{References}
-  \bibliography{references.bib}
+\bibliography{references.bib,gbif.bib}
 

### 2025-08-15 analysis/appendix.qmd
Finish renaming package

@@
   orcid: 0000-0002-1011-1244
-  email: joeroe@hey.com
+  email: jar@hum.ku.dk
   affiliations:
@@
 
-library("BadiaPaleoFloraENM") # this package, install with devtools::install()
+library("WAsiaPaleoFloraENM") # this package, install with devtools::install()
 

### 2025-10-20 analysis/paper.qmd
Finish renaming package
Fix citations in docx

@@
     number-sections: true
+    citeproc: true
+    csl: jcaa.csl
 bibliography: 
@@
 ---
-
-<!--
-TODO before publication:
-- [ ] Zenodo release
-- [ ] Clean up package code
--->
 
@@
 
-library("BadiaPaleoFloraENM") # this package, install with devtools::install()
+library("WAsiaPaleoFloraENM") # this package, install with devtools::install()
 ```
@@
 
-In general there is middling to low agreement between our hindcast predictions of species niches and observed archaeological occurrences of the same species [@tlb-results-summary].
+In general there is middling to low agreement between our hindcast predictions of species niches and observed archaeological occurrences of the same species [@tbl-results-summary].
 
